### 2019-05-01 paper.tex
first commit

\documentclass[11pt, a4paper]{article}

\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage[UKenglish]{babel}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{float}
\usepackage{graphicx}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{todonotes}

\newcommand{\nni}{\mathrm{NNI}}
\newcommand{\rnni}{\mathrm{RNNI}}
\newcommand{\rnniu}{\mathrm{RNNIu}}

\newtheorem{definition}{Definition}
\newtheorem{theorem}[definition]{Theorem}
\newtheorem{conjecture}[definition]{Conjecture}
\newtheorem{lemma}[definition]{Lemma}
\newtheorem{corollary}[definition]{Corollary}

\graphicspath{{figures/}}

\title{The Ranked Nearest Neighbour Interchange space}
\date{\today}
\author{Lena Collienne, Alex Gavryushkin}

\begin{document}

\maketitle

\begin{abstract}

\end{abstract}

\section{Introduction}

\section{Definitions}

%Do we restrict everything to RNNI? If not, make sure that we correctly distinguish between RNNI and RNNIu (for instance in the algorithm section)

%MRCA

%cluster

\begin{definition}
	Let $A$ be a pending subtree of the tree $T$.
	We call the set $c(A)$ of taxa of $A$ the \emph{cluster} induced by $A$.
\end{definition}

\section{Caterpillar Trees}

\begin{figure}[H]
	\centering
	\includegraphics[width=\textwidth]{NNI_vs_RNNI}
	\caption{Paths between caterpillar trees $T_1$ and $T_2$: black -- $\nni$ path of length $5$; red -- $\rnni$ path of length $6$; blue -- $\rnni$ path of length $8$.}
	\label{NNI_vs_RNNI}
\end{figure}

\begin{conjecture}
	The set of caterpillar trees is convex in $\rnni$ space.
\end{conjecture}

This conjecture is computationally proven for up to $8$ taxa (maximum $\rnni$ graph we can compute).
For a more restricted subset of trees in $\rnni$, namely caterpillar trees that have one cherry leaf in common, one can show convexity.

\begin{lemma}
	Let $T_1$ and $T_2$ be two caterpillar trees that both have a leaf labelled by $a$ as part of their cherry.
	There is a shortest paths from $T_1$ to $T_2$ that is a caterpillar path.
	\label{lemma:caterpillar_subset_convex}
\end{lemma}

\begin{figure}[h]
	\centering
	\includegraphics[width=0.3\textwidth]{label_internal_nodes}
	\caption{The labelling of the caterpillar tree as introduced in the proof of Lemma~\ref{lemma:caterpillar_subset_convex} is unique due to its definition.
	}
	\label{label_internal_nodes}
\end{figure}

\begin{proof}
	Let $T_1$ and $T_2$ be two caterpillar trees, both having a leaf labelled by $a$ as part of their cherry.
	Let $p$ be a shortest path from $T_1$ to $T_2$.
	We introduce a labelling of internal nodes of all trees on $p$ by first defining it for $T_1$ and $T_2$ and then explaining how the labelling is changed by $\rnni$ moves.

	In $T_1$ and $T_2$ the internal node of the cherry is labelled by the label of its child that is not $a$.
	All other internal nodes shall have the same label as their leaf children.
	From now on we will identify the leaf labelled by $x$ with $x$ and the internal node labelled by $x$ will be called $int_x$.
	An example for a caterpillar tree with internal leaf labels is depicted in Figure~\ref{label_internal_nodes}.
	In the following, changes of the internal labelling following $\rnni$ moves are defined:

	\textbf{Rank swaps:}
	Within rank swaps the labels of internal nodes do not change.
	For an example consider Figure~\ref{fig:rank_swap_internal_label}.

	\begin{figure}[H]
		\centering
		\includegraphics[width=0.8\textwidth]{rank_swap_internal_label}
		\caption{The labelling of the internal nodes that swap ranks (red) do not change within the rank swap.
		}
		\label{fig:rank_swap_internal_label}
	\end{figure}

	\textbf{NNI move:}
	Let there be an $\nni$ move on an edge $(int_x, int_y)$ where $rank(int_x) < rank(int_y)$ as in Figure~\ref{fig:nni_moves}.

	\begin{figure}[H]
		\centering
		\includegraphics[width=0.7\textwidth]{nni_moves_internal_label}
		\caption{Two possible $\nni$ moves on edge $(int_x, int_y)$ where $x \in A, y \in B$.
		}
		\label{fig:nni_moves}
	\end{figure}

	The nodes incident to the edge after the $\nni$ move are labelled by $int_x$ and $int_y$ as well.
	By default, the node that has lower rank than the other one is labelled by $int_x$, the other one by $int_y$.
	But if $int_x$ is no ancestor of $x$ or $int_y$ is no ancestor of $y$, we swap the labels of the two new nodes, such that $rank(int_x) > rank(int_y)$.
	An example of an $\nni$ move and the new labelling is provided in Figure~\ref{fig:nni_example_internal_label}.

	\begin{figure}[H]
		\centering
		\includegraphics[width=0.8\textwidth]{nni_example_internal_label}
		\caption{The first labelling after the $\nni$ move preserves the ranks of $int_4$ and $int_6$.
		But then $int_6$ is not an ancestor of $6$ any more.
		Therefore, $int_4$ and $int_6$ have to change their labels.
		}
		\label{fig:nni_example_internal_label}
	\end{figure}

	We proceed to show that $int_x$ and $int_y$ are ancestors of $x$ and $y$ after an $\nni$ move as described above, respectively.
	Let $A$ and $B$ be the subtrees that are children of $int_x$ and $C$ the subtree that is child of $int_y$ and does not contain $A$ and $B$, as depicted on the left of .
	Without loss of generality we can assume that $x \in A$, because $int_x$ is an ancestor of $x$ before the $\nni$ move.
	If otherwise $x\in B$, we just need to exchange $A$ and $B$ in the following argumentation.

	If $A$ is a child of the lower of both new nodes after the $\nni$ move, as in the upper tree on the right of Figure~\ref{fig:nni_example_internal_label}, we can keep the rank order of $int_x$ and $int_y$.
	So we assign $int_x$ to the lower node that is parent of $A$ and $C$ and $int_y$ to the upper node that is parent of $B$.
	Because $x \in A$ and $y \in A \cup B \cup C$, $int_x$ and $int_y$ are ancestors of $x$ and $y$, respectively.

	Let us now consider the $\nni$ move that results in a tree where $B$ and $C$ are the children of the lower of the two new nodes and $A$ is children of the upper new node.
	Because $x \in A$, $int_x$ needs to be assigned to the upper new node.
	It remains to prove that $y \in B \cup C$ to make sure that $int_y$ is an ancestor of $y$ after the $\nni$ move.

	In the following we consider the tree after the $\nni$ move.
	We know that $y \in A \cup B \cup C$, because $int_y$ is an ancestor of $y$ before the $\nni$ move.
	Therefore, it is sufficient to prove $y \notin A$.
	Let us consider the subtree $A$ and the labels of its internal nodes.
	Let $|A| = k$ be the number of leaves in $A$.
	There are $k-1$ internal nodes in $A$ that need to be labelled by these leaves.
	The parent of the root of $A$ is $int_x$, as it is labelled by $x$.
	Because $x \in A$, there are $k - 1$ leaves remaining in $A$ that must label some internal nodes.
	As the $k-1$ internal nodes of $A$ must be labelled by leafs that are descendants of these nodes, these labels need to be the $k-1$ leafs of $A$.
	Therefore, it holds:
	$int_z \in A \iff z \in A$.
	And as we know that $int_y \notin A$, $y$ can not be a leaf of $A$.
	Therefore, it is $y \in B \cup C$ and thus, $int_y$ is an ancestor of $y$ after the $\nni$ move.

	Concluding we see that our labelling is well defined.
	Another thing we should point out is that this labelling method leads to the labelling of $T_2$ as it is defined for caterpillar trees.
	This follows directly from the fact that $int_z$ is an ancestor of the leaf labelled by $z$.
	If one considers the internal node in $T_2$ with lowest rank, it can only have the label of the leaf in the cherry that is not labelled by $a$, because $int_z$ is an ancestor of $z$ for all leaf labels $z$.
	Now we can stepwise consider all internal nodes with rank one higher than the previously considered node and receive the same result.
	Hence, our way to change the labels of internal nodes on $p$ is consistent and ends in the correct labelling of $T_2$.

	Now we want to use this labelling to prove that there is always a caterpillar path that is at least as short as the given shortest path $p$.
	Therefore, we use the following observation:
	Within each $\rnni$ move the ranks of at most two internal nodes change by at most one each.
	This is true for both, rank swaps and $\nni$ moves.
	Within a rank swap two nodes simply exchange their ranks.
	And within an $\nni$ move two new nodes arise and are labelled by the former two nodes that bounded the edge on that the $\nni$ move is performed.
	But still, the ranks of these two nodes $int_x$ and $int_y$ both change by $1$.

	On a caterpillar path the only possible moves are $\nni$ moves.
	However, each of these $\nni$ move change the ranks of two internal nodes by at most one each.
	Therefore, we can construct a caterpillar path $r$ out of $p$:
	every time two internla nodes $int_x$ and $int_y$ exchange ranks, they do so on the caterpillar path $r$ as well.
	Since there are moves on $p$ that do not change the ranks of internal nodes, this caterpillar path has length $|r| \leq |p|$.
	Hence, there is always a shortest path between $T_1$ and $T_2$ that is a caterpillar path.
\end{proof}

\section{Split Theorem}

When analysing the complexity of the $\rnni$ space, the comparison to $\nni$ is an important step.
The proof of NP-completeness of $\nni$ in~\cite{jiang2000} is based on the fact that following theorem holds in $\nni$.

\begin{theorem}
	There are trees $T_1,T_2$ in $\nni$ sharing a partition which is not shared by any intermediate tree on a shortest path from $T_1$ to $T_2$.
	\label{thm:split_nni}
\end{theorem}

\begin{proof}
	See~\cite{Li1996}.
\end{proof}

The proof presented in~\cite{Li1996} for Theorem~\ref{thm:split_nni} does not work for $\rnni$. \todo{because?}
Therefore, the following conjecture was raised~\cite{Gavryushkin2017}.

\begin{conjecture}[Split Theorem]
	For the $\rnni$ graph the following statement holds:
	If a partition of leaves given by an edge is present in two trees $T$ and $R$ then the partition is presented in every tree on every shortest path between $T$ and $R$.
	\label{split_theorem}
\end{conjecture}

However, we found a counterexample to this conjecture:

\begin{figure}[H]
	\centering
	\includegraphics[width=0.8\textwidth]{split_theorem_counterexample}
	\caption{The split $123|45$ is present in $T_1$ and $T_2$, but in none of the trees on this shortest caterpillar path between them.}
	\label{split_theorem_counterexample}
\end{figure}

Since this counterexample clearly shows that the version of the Split Theorem stated in~\cite{Gavryushkin2017} does not hold, we will now claim an alternative to this conjecture.
The alternative formulation is based on the observation that there is a shortest path preserving all splits between the trees of the counterexample of the Split Theorem.
This path is shown in Figure~\ref{split_theorem_no_counterexample}.

\begin{figure}[H]
	\centering
	\includegraphics[width=0.8\textwidth]{split_theorem_no_counterexample}
	\caption{The Split $123|45$ is present in $T_1$ and $T_2$ and in every tree on this shortest path between them.}
	\label{split_theorem_no_counterexample}
\end{figure}

\begin{conjecture}[Weak Split Theorem]
	For $\rnni$ graph the following statement holds:
	if a partition of leaves given by an edge is present in two trees $T$ and $R$ then there exists a shortest path between $T$ and $R$ where this partition is present in every tree on this path.
	\label{split_theorem_weak}
\end{conjecture}

% The weak Split Theorem (in the version as in Lena's thesis) is computationally proven to hold for trees with up to $6$ taxa ($\sim 38$ min running time on laptop).
% We might want to consider the Cluster Theorem as well.
% The weak split theorem is stronger than the cluster theorem, since it also makes a statement about some trees that do not share a cluster

%A further and even weaker alternative formulation of the Split Theorem is the \emph{Cluster Theorem}, which considers clusters instead of splits.
%One can see that the trees of our counterexample~\ref{split_theorem_counterexample} do not fulfil the assumptions of the cluster theorem.
%
%\begin{conjecture}[Cluster Theorem]
%	For the $\rnni$ graph the following statement holds:
%	if two trees $T$ and $R$ contain the same cluster $C$, then $C$ is present as cluster in every tree on every shortest path between $T$ and $R$.
%	\label{cluster_theorem}
%\end{conjecture}

\section{Approximating Distances}

Following algorithm computes paths between two trees in $\rnni$:

\begin{algorithm}[H]
\caption{FIND\_PATH($T_1,T_2$)}
\label{alg:find_path}
\begin{algorithmic}[1]
	\STATE $T := T_1$
	\STATE $p := (T_1)$
	\FOR {$r = 1, \dots, n-2$}
		\STATE Let $S$ be the cluster associated with the node $v$ in $T_2$ with $rank(v) = r$
		\WHILE {$rank(MRCA_T(S))>r$}
			\STATE Update $T$: Decrease $rank(MRCA_T(S))$ by an $\rnni$ move \label{alg:line:move_set_down}
			\STATE $p = p+T$
		\ENDWHILE
	\ENDFOR
	\RETURN p
\end{algorithmic}
\end{algorithm}

One can easily prove that there is always exactly one possible $\rnni$ move that decreases the rank of $MRCA_T(S)$ as in line~\ref{alg:line:move_set_down} of Algorithm~\ref{alg:find_path}.

The proof of Theorem 7 in \cite{Gavryushkin2017} gives an algorithm to compute paths between trees in $\rnni$.

% Compare the two algorithms by sampling trees and check which algorithm gives shorter paths.

% Also mention the exact algortihm for computing the whole RNNI space for up to 7 (?) taxa and check that Algorithm find_path gives correct distances there.
%Maybe it's worth trying the other algorithm in these trees as well.

Algorithm~\ref{alg:find_path} provides an upper bound for the diameter of $\rnniu$ space.
It improves the bound $n^2 - 3n - \frac{5}{8}$ given in \cite{Gavryushkin2017}.

\begin{theorem}
	For $n \geq 3$, $\Delta(\rnniu) \leq \frac{1}{2}n^2-\frac{3}{2}n+1$
\end{theorem}

\begin{proof}
	This result directly follows from Algorithm~\ref{alg:find_path}.
	In the worst case each $r = 1, \dots, n-2$ requires $n-1-r$ $\rnni$ moves for moving the corresponding $MRCA$ down.
	Hence, there is for each pair of trees a path of length less or equal to $\sum\limits_{i = 1}^{n-2} i = \frac{(n-2)(n-1)}{2} = \frac{1}{2}n^2-\frac{3}{2}n+1$.
\end{proof}

% Is this boundary tight? Idea: there are always caterpillar trees that have this distance (we can already check this computationally, if we were to prove the caterpillar convexity, this follows directly)

\begin{lemma}
	There are two caterpillar trees with distance $\frac{1}{2}n^2-\frac{3}{2}n+1$ if the set of caterpillar trees is convex.
	\label{lemma:caterpillar_diameter_convex}
\end{lemma}

\begin{proof}
	If the set of caterpillar trees is convex, Bubble Sort proves that there is a caterpillar path between the caterpillar trees
	$(( \dots (1,2)3)\dots)n)$ and $(( \dots (n,n-1)n-2)\dots)1)$ with length $\frac{1}{2}n^2-\frac{3}{2}n+1$. 
	\todo{Explain how to use Bubble Sort}
\end{proof}

\begin{conjecture}
	There are two caterpillar trees in $\rnniu$ with distance $\frac{1}{2}n^2-\frac{3}{2}n+1$.
	\label{conj:caterpillar_diameter}
\end{conjecture}

% This can be computationally proven for up to $8$ taxa.

\begin{proof}
	Consider trees $(( \dots (n,1)2)\dots)n-1)$ and $(( \dots (n,n-1)n-2)\dots)1)$ and apply Lemma~\ref{lemma:caterpillar_subset_convex} and use Bubble Sort.
\end{proof}

\begin{corollary}
	If Conjecture~\ref{conj:caterpillar_diameter} holds, it is $\Delta(\rnniu) = \frac{1}{2}n^2-\frac{3}{2}n+1$.
\end{corollary}

\section{Partition lattice}

% Define Partition Lattice $\Pi_n$ and max chains in that lattice!

\begin{theorem}
	The $\rnni$ space on $n$ taxa is isomorphic to the space of maximum chains of the partition lattice $\Pi_n$ where two maximum chains are connected by an edge if, and only if, they differ by exactly one partition.
\end{theorem}

\bibliographystyle{abbrv}
\bibliography{lit}

\end{document}

### 2019-05-01 paper.tex
update proof convexity of caterpillars with shared cherry taxon

@@
 
+%leaf or taxon?
+
 \begin{definition}
@@
 	Let $p$ be a shortest path from $T_1$ to $T_2$.
-	We introduce a labelling of internal nodes of all trees on $p$ by first defining it for $T_1$ and $T_2$ and then explaining how the labelling is changed by $\rnni$ moves.
+	We introduce a labelling of internal nodes of all trees on $p$ by first defining it for the caterpillar trees $T_1$ and $T_2$ and then explaining how the labelling is changed by $\rnni$ moves.
 
@@
 	We proceed to show that $int_x$ and $int_y$ are ancestors of $x$ and $y$ after an $\nni$ move as described above, respectively.
-	Let $A$ and $B$ be the subtrees that are children of $int_x$ and $C$ the subtree that is child of $int_y$ and does not contain $A$ and $B$, as depicted on the left of .
-	Without loss of generality we can assume that $x \in A$, because $int_x$ is an ancestor of $x$ before the $\nni$ move.
-	If otherwise $x\in B$, we just need to exchange $A$ and $B$ in the following argumentation.
+	If $x$ stays in the subtree rooted in the lower node of $e$, this does obviously hold.
+	We now turn to the case that $x$ is not in the subtree rooted in the lower node of $e$ after the $\nni$ move.
+	Then the labels of the two internal nodes change such that $rank(int_x) > rank(int_y)$.
+	This makes sure that $int_x$ is an ancestor of $x$.
+	Thus, it remains to prove that $int_y$ is ancestor of $y$.
+	In particular, we need to show that $y$ is not in the same subtree $A$ as $x$, whose root has the node labelled by $int_x$ as parent before and after the $\nni$ move.
+	Let us consider the tree before the $\nni$ move.
+	$A$ has $|A|$ leaves and $|A| - 1$ internal nodes.
+	These internal nodes are labelled by some leaves of $A$, because all internal nodes must be labelled by a descendant leaf.
+	We know that $x \in A$, and the parent of the root of $A$ is labelled by $x$.
+	All remaining $|A|-1$ leaf labels of $A$ appear as labels of the $|A|-1$ internal nodes of $A$.
+	From the fact that the internal node labelled by $y$ is not inside $A$ we can conclude that $y$ cannot be a leaf of $A$.
+	This proves that the labelling of the internal nodes is is well defined and allows a unique identification of internal nodes by their labels.
 
-	If $A$ is a child of the lower of both new nodes after the $\nni$ move, as in the upper tree on the right of Figure~\ref{fig:nni_example_internal_label}, we can keep the rank order of $int_x$ and $int_y$.
-	So we assign $int_x$ to the lower node that is parent of $A$ and $C$ and $int_y$ to the upper node that is parent of $B$.
-	Because $x \in A$ and $y \in A \cup B \cup C$, $int_x$ and $int_y$ are ancestors of $x$ and $y$, respectively.
-
-	Let us now consider the $\nni$ move that results in a tree where $B$ and $C$ are the children of the lower of the two new nodes and $A$ is children of the upper new node.
-	Because $x \in A$, $int_x$ needs to be assigned to the upper new node.
-	It remains to prove that $y \in B \cup C$ to make sure that $int_y$ is an ancestor of $y$ after the $\nni$ move.
-
-	In the following we consider the tree after the $\nni$ move.
-	We know that $y \in A \cup B \cup C$, because $int_y$ is an ancestor of $y$ before the $\nni$ move.
-	Therefore, it is sufficient to prove $y \notin A$.
-	Let us consider the subtree $A$ and the labels of its internal nodes.
-	Let $|A| = k$ be the number of leaves in $A$.
-	There are $k-1$ internal nodes in $A$ that need to be labelled by these leaves.
-	The parent of the root of $A$ is $int_x$, as it is labelled by $x$.
-	Because $x \in A$, there are $k - 1$ leaves remaining in $A$ that must label some internal nodes.
-	As the $k-1$ internal nodes of $A$ must be labelled by leafs that are descendants of these nodes, these labels need to be the $k-1$ leafs of $A$.
-	Therefore, it holds:
-	$int_z \in A \iff z \in A$.
-	And as we know that $int_y \notin A$, $y$ can not be a leaf of $A$.
-	Therefore, it is $y \in B \cup C$ and thus, $int_y$ is an ancestor of $y$ after the $\nni$ move.
-
-	Concluding we see that our labelling is well defined.
-	Another thing we should point out is that this labelling method leads to the labelling of $T_2$ as it is defined for caterpillar trees.
-	This follows directly from the fact that $int_z$ is an ancestor of the leaf labelled by $z$.
-	If one considers the internal node in $T_2$ with lowest rank, it can only have the label of the leaf in the cherry that is not labelled by $a$, because $int_z$ is an ancestor of $z$ for all leaf labels $z$.
-	Now we can stepwise consider all internal nodes with rank one higher than the previously considered node and receive the same result.
-	Hence, our way to change the labels of internal nodes on $p$ is consistent and ends in the correct labelling of $T_2$.
-
-	Now we want to use this labelling to prove that there is always a caterpillar path that is at least as short as the given shortest path $p$.
+	The task is now to prove that there is always a caterpillar path at least as short as the given path $p$.
 	Therefore, we use the following observation:
 	Within each $\rnni$ move the ranks of at most two internal nodes change by at most one each.
-	This is true for both, rank swaps and $\nni$ moves.
-	Within a rank swap two nodes simply exchange their ranks.
-	And within an $\nni$ move two new nodes arise and are labelled by the former two nodes that bounded the edge on that the $\nni$ move is performed.
-	But still, the ranks of these two nodes $int_x$ and $int_y$ both change by $1$.
+	This is true for both rank swaps and $\nni$ moves, according to our definition of the labelling:
+	Within a rank swap two nodes exchange their ranks, while there are not necessarily changes of ranks when $\nni$ moves are performed.
 
-	On a caterpillar path the only possible moves are $\nni$ moves.
-	However, each of these $\nni$ move change the ranks of two internal nodes by at most one each.
+	On a caterpillar path the only possible moves are $\nni$ moves that change the ranks of two internal nodes by one each.
 	Therefore, we can construct a caterpillar path $r$ out of $p$:
-	every time two internla nodes $int_x$ and $int_y$ exchange ranks, they do so on the caterpillar path $r$ as well.
+	every time two internal nodes $int_x$ and $int_y$ exchange ranks on $p$, they do so on the caterpillar path $r$ as well.
 	Since there are moves on $p$ that do not change the ranks of internal nodes, this caterpillar path has length $|r| \leq |p|$.

### 2019-05-01 paper.tex
prove diameter RNNI

@@
 
-\begin{conjecture}
+\begin{lemma}
 	There are two caterpillar trees in $\rnniu$ with distance $\frac{1}{2}n^2-\frac{3}{2}n+1$.
 	\label{conj:caterpillar_diameter}
-\end{conjecture}
-
-% This can be computationally proven for up to $8$ taxa.
+\end{lemma}
 
 \begin{proof}
-	Consider trees $(( \dots (n,1)2)\dots)n-1)$ and $(( \dots (n,n-1)n-2)\dots)1)$ and apply Lemma~\ref{lemma:caterpillar_subset_convex} and use Bubble Sort.
+    For this proof we will give two caterpillar trees $T_1$ and $T_2$ on $n$ taxa with $d_{RNNI}(T_1,T_2) = \frac{1}{2}n^2-\frac{3}{2}n+1$.
+    \todo{$d$ or $d_{RNNI}$?}
+	Let $T_1 = (( \dots (n,1)2)\dots)n-1)$ and $T_2 = (( \dots (n,n-1)n-2)\dots)1)$.
+    According to Lemma~\ref{lemma:caterpillar_subset_convex}, there is a shortest path between $T_1$ and $T_2$ that only consists of caterpillar trees.
+    We can compute such a shortest caterpillar path by using a modified version of Bubble Sort.
+    \todo{Explain how to use Bubble Sort}
+    When running this algorithm on $T_1$ and $T_2$ there are $n-1-i$ swaps of taxon $i$ with its right neighbour.
+    Therefore, the total number of these swaps, and therefore $d_{RNNI}(T_1,T_2)$, is $\sum\limits_{i=1}^{n-1}(n-1-i) = \sum\limits_{i=0}^{n-2}i = \frac{1}{2}n^2-\frac{3}{2}n+1$
 \end{proof}

### 2019-05-02 paper.tex
why split counterexample in nni doesn't work in rnni

@@
 
-The proof presented in~\cite{Li1996} for Theorem~\ref{thm:split_nni} does not work for $\rnni$. \todo{because?}
+The proof presented in~\cite{Li1996} for Theorem~\ref{thm:split_nni} does not work for $\rnni$.
+
+%This part is mainly for myself, not sure whether it is important to have it in the paper
+The main reason that the counterexample presented in~\ref{Li1996} works in $\nni$ is following:
+Let $T_1$ and $T_2$ be two caterpillar trees.
+If each taxon $a$ is repaced by a cherry $(a_1,a_2)$ in both trees $T_1$ and $T_2$, the resulting trees $T_1'$ and $T_2'$ have distance $d_{\nni}(T_1',T_2') = d_{\nni}(T_1,T_2)$ in the $\nni$ graph.
+However, this does not work in $\rnni$.
+Introducing new cherries means introducing new internal nodes with new ranks.
+And a shortest path from $T_1$ to $T_2$ does not give a shortest path from $T_1'$ to $T_2'$, there will be some additional rank swaps needed for the internal nodes of the cherries.
+% How do we assign the ranks to the new cherry nodes?
+% We could give a precise example here, on 3 or 4 taxa (4 is the max, we can compute RNNI space for up to 8 taxa), because we know how the caterpillar trees with max distance look like.
+
 Therefore, the following conjecture was raised~\cite{Gavryushkin2017}.

### 2019-05-03 paper.tex
how to use Bubble Sort

@@
 
-%Do we restrict everything to RNNI? If not, make sure that we correctly distinguish between RNNI and RNNIu (for instance in the algorithm section)
+%Do we restrict everything to RNNI? Make sure that we correctly distinguish between RNNI and RNNIu (for instance in the algorithm section)
 
@@
 
-\section{Approximating Distances}
+\section{Algorithms}
+
+We are now going to introduce different algorithms that computes paths in $\rnni$.
+\cite{Gavryushkin2017} gave an algorithm for computing the whole $\rnni$ space.
+We implemented this algorithm and are able to use it on trees with up to $8$ taxa.
+Therefore, we can compare the paths given by the algorithms presented in this chapter with the exact shortest path for trees with up to $8$ taxa.
+
+% Shall we mention the data structure we use (list of sets) - this might be useful to see the connection to the partition lattice.
+
+A modified version of the classic Bubble Sort \todo{cite} algorithm returns a caterpillar path that is shortest among all caterpillar paths between given caterpillar trees $T_1, T_2$.
+Let $T_1 = (\ldots((1,2)3) \ldots )n), T_2 = (\ldots((\sigma_1,\sigma_2)\sigma_3) \ldots )\sigma_n)$ be two caterpillar trees.
+They can be seen as sequences $(1,2,\ldots,n)$ and $(\sigma_1, \sigma_2, \ldots, \sigma_n)$ that need to be sorted.
+$\nni$ moves between caterpillar trees correspond to exchanging two neighboured elements in a sequence.
+The only exception of this are $\nni$ moves that change the cherry of a tree.
+The corresponding moves on sequences are exchanged of either the second and third element or the first and third element.
+On the other hand, there is no difference between two the trees corresponding to two sequences where only the first two taxa are exchanged.
+Since all pairs of taxa whose order in $T_1$ is different from the one in $T_2$ need to be exchanged by $\nni$ moves on a caterpillar path, Bubble Sort will give a shortest path among all caterpillar paths.
+Just the moves for the first three positions of the sequence must be adapted.
+Let us consider the first three elements $(a,b,c,\ldots)$ of a sequence representing a caterpillar tree on a caterpillar path between $T_1$ and $T_2$
+If the order of these elements is the same in $T_2$, there are no swaps needed for these taxa.
+If just one of the two taxa $a$ and $b$ is on the other side of $c$ in $T_2$, this one will change wih $c$.
+But if both $a$ and $b$ are on the other side of $c$ in $T_2$, the one which is the righmost of all three taxa in $T_2$ swaps with $c$ at first.
+This makes sure that there is no additional swap of $a$ and $b$ needed on the path to $T_2$.
+\todo{This paragraph is just a draft. Is it needed at all?}
 

### 2019-05-03 paper.tex
delete caterpillar convexity => diameter, we have a real proof

@@
 \begin{lemma}
-	There are two caterpillar trees with distance $\frac{1}{2}n^2-\frac{3}{2}n+1$ if the set of caterpillar trees is convex.
-	\label{lemma:caterpillar_diameter_convex}
-\end{lemma}
-
-\begin{proof}
-	If the set of caterpillar trees is convex, Bubble Sort proves that there is a caterpillar path between the caterpillar trees
-	$(( \dots (1,2)3)\dots)n)$ and $(( \dots (n,n-1)n-2)\dots)1)$ with length $\frac{1}{2}n^2-\frac{3}{2}n+1$.
-	\todo{Explain how to use Bubble Sort}
-\end{proof}
-
-\begin{lemma}
 	There are two caterpillar trees in $\rnniu$ with distance $\frac{1}{2}n^2-\frac{3}{2}n+1$.

### 2019-05-06 paper.tex
corrections on why split thm counterexample doesn't work in RNNI

@@
 %This part is mainly for myself, not sure whether it is important to have it in the paper
-The main reason that the counterexample presented in~\ref{Li1996} works in $\nni$ is following:
+The counterexample presented in~\cite{Li1996} works in $\nni$ is based on following observation:
 Let $T_1$ and $T_2$ be two caterpillar trees.
-If each taxon $a$ is repaced by a cherry $(a_1,a_2)$ in both trees $T_1$ and $T_2$, the resulting trees $T_1'$ and $T_2'$ have distance $d_{\nni}(T_1',T_2') = d_{\nni}(T_1,T_2)$ in the $\nni$ graph.
+If each taxon $a$ is replaced by a cherry $(a_1,a_2)$ in both trees $T_1$ and $T_2$, the resulting trees $T_1'$ and $T_2'$ have distance $d_{\nni}(T_1',T_2') = d_{\nni}(T_1,T_2)$ in the $\nni$ graph.
+When proving~\ref{thm:split_nni},~\cite{Li1996} give an example of two trees $T$ and $R$ on $2n$ taxa that consist of two caterpillar trees each that share the root as parent.
+The split given by the edges incident to the root is $1, \ldots, n | n+1, \ldots, 2n$ in both trees $T$ and $R$.
+For the given example there is a shortest path in $\nni$ where at first $T$ is transformed to a tree $T'$ with $n$ cherries where each cherry contains one taxon of $\{1, \ldots, n\}$ and one taxon of $\{n+1, \ldots, 2n\}$.
+Afterwards, $T'$ is tranformed to a tree $R'$ that contains the same cherries as $T'$.
+Finally, all cherries of $R'$ are resolved in order to end up in tree $R$.
+For proving that this path is shorter than any path preserving the split $1, \ldots, n | n+1, \ldots, 2n$, the fact that $d_{\nni}(T',R') = d_{\nni}(T|_{\{1, \ldots, n\}}, R|_{\{n+1, \ldots, 2n\}})$ is being used.
 However, this does not work in $\rnni$.
 Introducing new cherries means introducing new internal nodes with new ranks.
-And a shortest path from $T_1$ to $T_2$ does not give a shortest path from $T_1'$ to $T_2'$, there will be some additional rank swaps needed for the internal nodes of the cherries.
+Since there are some rank swaps needed for an $\rnni$ path similar to the shortest $\nni$ path presented above, the path in $\rnni$ that breaks the split $1, \ldots, n | n+1, \ldots, 2n$ is no shortest path.
 % How do we assign the ranks to the new cherry nodes?
-% We could give a precise example here, on 3 or 4 taxa (4 is the max, we can compute RNNI space for up to 8 taxa), because we know how the caterpillar trees with max distance look like.
+% We could either create a path where $T'$ and $R'$ have distance equal to $d_{\rnni}(T|_{\{1, \ldots, n\}}, R|_{\{n+1, \ldots, 2n\}})$ and have a lot of additional rank swaps before and after that part OR
+% Build $T'$ and $R'$ that contain all the cherries within the least number of moves possible (3n-3) and having a longer path between $T'$ and $R'$.
+% Also anything in between these two options is possible. Are we able to disprove every option??
 

### 2019-05-06 paper.tex
typo

@@
 Finally, all cherries of $R'$ are resolved in order to end up in tree $R$.
-For proving that this path is shorter than any path preserving the split $1, \ldots, n | n+1, \ldots, 2n$, the fact that $d_{\nni}(T',R') = d_{\nni}(T|_{\{1, \ldots, n\}}, R|_{\{n+1, \ldots, 2n\}})$ is being used.
+For proving that this path is shorter than any path preserving the split $1, \ldots, n | n+1, \ldots, 2n$, the fact that $d_{\nni}(T',R') = d_{\nni}(T|_{\{1, \ldots, n\}}, R|_{\{1, \ldots, n\}})$ is being used.
 However, this does not work in $\rnni$.

### 2019-05-15 paper.tex
split + cluster thm

@@
 	\centering
-	\includegraphics[width=0.8\textwidth]{split_theorem_counterexample}
-	\caption{The split $123|45$ is present in $T_1$ and $T_2$, but in none of the trees on this shortest caterpillar path between them.}
-	\label{split_theorem_counterexample}
+	\includegraphics[width=\textwidth]{splitthm_counterexample}
+	\caption{The split $123|45$ is present in $T_1$ and $T_2$, but the path at the top is a shortest path where none of the trees contains this split.
+    However, this split is maintained on the path at the bottom that is a shortest path as well.}
+	\label{splitthm_counterexample}
 \end{figure}
 
-Since this counterexample clearly shows that the version of the Split Theorem stated in~\cite{Gavryushkin2017} does not hold, we will now claim an alternative to this conjecture.
-The alternative formulation is based on the observation that there is a shortest path preserving all splits between the trees of the counterexample of the Split Theorem.
-This path is shown in Figure~\ref{split_theorem_no_counterexample}.
-
-\begin{figure}[H]
-	\centering
-	\includegraphics[width=0.8\textwidth]{split_theorem_no_counterexample}
-	\caption{The Split $123|45$ is present in $T_1$ and $T_2$ and in every tree on this shortest path between them.}
-	\label{split_theorem_no_counterexample}
-\end{figure}
+Since the counterexample in Figure~\ref{splitthm_counterexample} clearly shows that the version of the Split Theorem stated in~\cite{Gavryushkin2017} does not hold, we will now claim an alternative to this conjecture.
+The alternative formulation is based on the observation that there is a shortest path preserving all splits between the trees of the counterexample of the Split Theorem, which is presented at the bottom of Figure~\ref{splitthm_counterexample}.
 
@@
 
+\todo{Do we want to keep the weak version of the split thm? It is stronger than the Cluster Thm!}
+
 % The weak Split Theorem (in the version as in Lena's thesis) is computationally proven to hold for trees with up to $6$ taxa ($\sim 38$ min running time on laptop).
-% We might want to consider the Cluster Theorem as well.
-% The weak split theorem is stronger than the cluster theorem, since it also makes a statement about some trees that do not share a cluster
 
-%A further and even weaker alternative formulation of the Split Theorem is the \emph{Cluster Theorem}, which considers clusters instead of splits.
-%One can see that the trees of our counterexample~\ref{split_theorem_counterexample} do not fulfil the assumptions of the cluster theorem.
-%
-%\begin{conjecture}[Cluster Theorem]
-%	For the $\rnni$ graph the following statement holds:
-%	if two trees $T$ and $R$ contain the same cluster $C$, then $C$ is present as cluster in every tree on every shortest path between $T$ and $R$.
-%	\label{cluster_theorem}
-%\end{conjecture}
+Another alternative formulation of the Split Theorem is the \emph{Cluster Theorem}, which considers clusters instead of splits.
+Considering clusters rather than splits is more reasonable since two rooted trees have the same set of clusters if, and only if, they are isomorphic~\cite{steel2016}.
+The trees $T_1, T_2$ of counterexample~\ref{splitthm_counterexample} share the same set of induced splits, but have distinct sets of clusters.
+
+\begin{conjecture}[Cluster Theorem]
+	For the $\rnni$ graph the following statement holds:
+	if two trees $T$ and $R$ contain the same cluster $C$, then $C$ is present as cluster in every tree on every shortest path between $T$ and $R$.
+	\label{cluster_theorem}
+\end{conjecture}
 
@@
 % Also mention the exact algortihm for computing the whole RNNI space for up to 7 (?) taxa and check that Algorithm find_path gives correct distances there.
-%Maybe it's worth trying the other algorithm in these trees as well.
+% Maybe it's worth trying the other algorithm in these trees as well.
 
@@
 \begin{theorem}
-	For $n \geq 3$, $\Delta(\rnniu) \leq \frac{1}{2}n^2-\frac{3}{2}n+1$
+	It is $\Delta(\rnniu) \leq \frac{1}{2}n^2-\frac{3}{2}n+1$
 \end{theorem}
@@
 \end{proof}
-
-% Is this boundary tight?
-Idea: there are always caterpillar trees that have this distance (we can already check this computationally, if we were to prove the caterpillar convexity, this follows directly)
 
@@
     We can compute such a shortest caterpillar path by using a modified version of Bubble Sort.
-    \todo{Explain how to use Bubble Sort}
     When running this algorithm on $T_1$ and $T_2$ there are $n-1-i$ swaps of taxon $i$ with its right neighbour.
@@
 \begin{corollary}
-	If Conjecture~\ref{conj:caterpillar_diameter} holds, it is $\Delta(\rnniu) = \frac{1}{2}n^2-\frac{3}{2}n+1$.
+It is $\Delta(\rnniu) = \frac{1}{2}n^2-\frac{3}{2}n+1$.
 \end{corollary}

### 2019-05-20 paper.tex
labels of caterpillars with distance=diameter

@@
 \usepackage{todonotes}
+\usepackage{enumitem}
 
@@
 
+\begin{lemma}
+    Let $T$ be the caterpillar tree $(\ldots ((1,2),3), \ldots, n)$.\\
+    A caterpillar tree $T'$ has distance $d(T,T') = \Delta(\rnniu) = \frac{1}{2}n^2-\frac{3}{2}n+1$ to $T$ if, and only if, $T' = ( \ldots ((n,t),n-1),n-2), \ldots, t+1), t-1) \ldots, 1)$
+\end{lemma}
+
+\begin{figure}[H]
+    \centering
+    \includegraphics[width=0.6\textwidth]{caterpillar_max_dist}
+    \caption{Caterpillar trees $T$ and $T'$ with distance $d(T,T') = \Delta(\rnniu)$}
+    \label{fig:caterpillar_max_distance}
+\end{figure}
+
+\begin{proof}
+    Let $T$ be the caterpillar tree $(\ldots ((1,2),3), \ldots, n)$.\\
+    If $T'$ has distance $d(T,T') = \Delta(\rnniu)$ to $T$, is must be $T' = ( \ldots ((n,t),n-1),n-2), \ldots, t+1), t-1) \ldots, 1)$:
+
+    Bubble Sort computes a shortest caterpillar path between two given trees.
+    The maximum distance $\frac{(n-1)(n-2)}{2}$ can be reached by this algorithm, if in every loop every neighboured pair of taxa needs to swap.
+    Starting at the tree $T$, this means that one of the taxa $1$ or $2$ moves up to the top of the tree in the first loop.
+    In every consecutive loop one taxon in the cherry of the current tree that moves up in the tree and ends up directly below the taxon that was moved up in the previously loop.
+    It is not hard to see that exchanging every neighboured pair of taxa as in Bubble Sort starting at $T$ leads to a tree $T'$ as described in the Lemma.
+    The taxon $t$ is an arbitrary taxon of the set $\{1, \ldots, n-1\}$.
+    If it is $1$, the uppermost taxon in $T'$ is $2$.
+    These trees $T'$ are the only ones that need this maximum number of $\rnni$ moves starting at $T$.
+    Therefore, it follows that $T' = ( \ldots ((n,t),n-1),n-2), \ldots, t+1), t-1) \ldots, 1)$ for every tree $T'$ with distance $d(T,T') = \Delta(\rnniu)$
+
+    It remains to show that if $T' = ( \ldots ((n,t),n-1),n-2), \ldots, t+1), t-1) \ldots, 1)$, it follows that $d(T,T') = \Delta(\rnniu)$ to $T$.
+    This can be proven by induction on the number of taxa $n$:
+
+    Induction basis: $n=3$\\
+    All trees have distance $1$.
+
+    Induction step: $n \to n+1$\\
+    Induction hypothesis: For trees $T_1 = (\ldots ((1,2),3), \ldots, n)$, $T_2 = ( \ldots ((n,t),n-1),n-2), \ldots, t+1), t-1) \ldots, 1)$ on $n$ taxa it is $d(T_1, T_2) = \frac{(n-1)(n-2)}{2} = \Delta(\rnniu_n)$.\\
+    Let $T, T'$ be trees on $n+1$ taxa as described in the Lemma and a shortest path $p$ from $T$ to $T'$.
+    Consider the trees $T|_n$ and $T'|_n$ that result from $T$ and $T'$ by deleting taxon $n+1$, respectively.
+    Deleting taxon $n+1$ from all trees on $p$, and deleting all trees that appear more than once in a row, results in a path $p'$ from $T|_n$ to $T'|_n$.
+    Applying the induction hypothesis onto $T|_n$ and $T'|_n$ shows that this path has length $\frac{(n-1)(n-2)}{2}$.
+    We can now compare this new path $p'$ with $p$.
+    The only steps that appear on $p$ but not on $p'$ are those involving the additional internal node.
+    We can label this node by $n+1$ and define a bijection between the internal nodes in all trees on $p'$ and the corresponding nodes in the trees in $p'$.
+    The additional internal node $n+1$ has rank $n$ in $T$ and ran $1$ in $T'$.
+    Each $\rnni$ move can decrease the rank of a node by at most $1$.
+    Therefore, there are at least $n-1$ $\rnni$ moves needed on $p$, which do not appear on $p'$, for moving the internal node $n+1$ down.
+    Hence, it is $d(T,T') = |p| \geq n-1 + |p'| = n-1 + \frac{(n-1)(n-2)}{2} = \frac{n(n-1)}{2} = \Delta(\rnniu)$.
+
+\end{proof}
+
 \section{Partition lattice}

### 2019-05-20 paper.tex
proof caterpillar convexity (?)

@@
     A caterpillar tree $T'$ has distance $d(T,T') = \Delta(\rnniu) = \frac{1}{2}n^2-\frac{3}{2}n+1$ to $T$ if, and only if, $T' = ( \ldots ((n,t),n-1),n-2), \ldots, t+1), t-1) \ldots, 1)$
+    \label{lemma:caterpillar_dist=diameter}
 \end{lemma}
+
+\todo{At the top of $T'$ it could also be $2,1$, or $t=1$}
 
@@
     Hence, it is $d(T,T') = |p| \geq n-1 + |p'| = n-1 + \frac{(n-1)(n-2)}{2} = \frac{n(n-1)}{2} = \Delta(\rnniu)$.
+    Thus, there is a shortest path from $T$ to $T'$ that only consists of caterpillar trees.
+
+\end{proof}
+
+\begin{theorem}
+    The set of caterpillar trees is convex.
+\end{theorem}
+
+\begin{proof}
+    We prove the Theorem by induction on the length $l$ of a shortest caterpillar path between two caterpillar trees $T_1$ and $T_2$.
+
+    Induction basis: $l = \Delta(\rnniu) = \frac{(n-1)(n-2)}{2}$\\
+    With Lemma~\ref{lemma:caterpillar_dist=diameter} one can see that for each pair of caterpillar trees $T_1$, $T_2$ with distance $d(T_1,T_2) = \Delta(\rnniu)$ there is a caterpillar path from $T_1$ to $T_2$ between these trees.
+
+    Induction step: $l+1 \to l$\\
+    Induction hypothesis: For each pait of caterpillar trees with distance $l+1$ there is a shortest path between these trees that only consists of caterpillar trees.
+    Let $T_1$ and $T_2$ be caterpillar trees with caterpillar distance $l$ \todo{define 'caterpillar distance'}.
+    If $l \neq \Delta(\rnniu)$, there must be a pair of taxa $x,y$ that are neighboured in $T_2$ and have the same order in $T_1$ and $T_2$, which means that $rank(p_x) > rank(p_y)$ or $rank(p_x) < rank(p_y)$ holds in both trees $T_1$ and $T_2$.
+    Then it follows that the tree $T_2'$ that equals $T_2$, but with exchanged taxa $x$ and $y$, has distance $l+1$ to $T_1$.
+    So there must be a shortest path from $T_2'$ to $T_1$ that only consists of caterpillar trees.
+    Since it does not matter in which order the pairs of taxa that need to exchange their position do exchange on a path between two caterpillar trees, $T_2$ is on a shortest caterpillar path from $T_2'$ to $T_1$.
+    Therefore, it must be $d(T_1, T_2) = l$.
 

### 2019-05-23 paper.tex
update proof lemma:caterpillar_dist=diameter (induction base convexity)

--- 2019-05-15 paper.tex
@@
 
-% Also mention the exact algortihm for computing the whole RNNI space for up to 7 (?) taxa and check that Algorithm find_path gives correct distances there.
+% Also mention the exact algorithm for computing the whole RNNI space for up to 7 (?) taxa and check that Algorithm find_path gives correct distances there.
 % Maybe it's worth trying the other algorithm in these trees as well.
@@
 
+\begin{lemma}
+    Let $T$ and $T'$ be two caterpillar trees with distance $d(T,T') = \Delta(\rnniu) = \frac{(n-1)(n-2)}{2}$.
+    There is a shortest path between $T$ and $T'$ that is a caterpillar path.
+    \label{lemma:caterpillar_dist=diameter}
+\end{lemma}
+
+    % \todo{At the top of $T'$ it could also be $2,1$, or $t=1$}
+    %
+    % \begin{figure}[H]
+    %     \centering
+    %     \includegraphics[width=0.6\textwidth]{caterpillar_max_dist}
+    %     \caption{Caterpillar trees $T$ and $T'$ with distance $d(T,T') = \Delta(\rnniu)$}
+    %     \label{fig:caterpillar_max_distance}
+    % \end{figure}
+
+\begin{proof}
+    Let $d_c(T,T')$ denote the length of a shortest caterpillar path from $T$ to $T'$, that is a caterpillar path that is shortest among all caterpillar paths from $T$ to $T'$.
+    We need to prove that $d(T,T') = \Delta(\rnniu)$ if, and only if, $d_c(T,T') = \Delta(\rnniu)$.
+
+    Let $T$ and $T'$ have distance $d(T,T') = \Delta(\rnniu)$.
+    We can compute a  shortest caterpillar path from $T$ to $T'$ using Bubble Sort.
+    In each loop $k=1, \ldots, n-2$ of the algorithm there can be at most $n-2-k$ exchanges of taxa.
+    It follows that the maximum length of a path computed by Bubble Sort is $\sum\limits_{k=1}^{n-2} k = \frac{(n-1)(n-2)}{2}$.
+    Since there cannot be a caterpillar path shorter than $\frac{(n-1)(n-2)}{2}$, if the distance is $d(T,T') \frac{(n-1)(n-2)}{2}$, it follows $d_c(T,T') = \Delta(\rnniu)$.
+
+    Let now $T$ and $T'$ have caterpillar distance $d_c(T,T') = \Delta(\rnniu)$.
+    We prove $d(T,T') = d_c(T,T')$ by induction on the number of taxa $n$ of $T$ ad $T'$:
+    We assume without loss of generality that $T$ is the caterpillar tree $(\ldots ((1,2),3), \ldots, n)$.
+    Otherwise, we re-label $T$ and $T'$.
+    Let $T'$ be a caterpillar tree with distance $\Delta(\rnniu)$ to $T$.
+
+    Induction basis: $n=3$\\
+    All trees have distance $1$ and are caterpillar trees.
+
+    Induction step: $n \to n+1$\\
+    Induction hypothesis: For trees $T$, $T'$ on $n$ taxa it is $d_c(T,T') = d(T, T') = \frac{(n-1)(n-2)}{2} = \Delta(\rnniu_n)$.\\
+    Let $T, T'$ be trees on $n+1$ taxa with caterpillar distance $\Delta(\rnniu)$.
+    Since we can reach $T'$ from $T$ by using Bubble Sort where each pair of taxa that is compared in that algorithm exchanges, we can conclude that taxon $n+1$ must be in the cherry of $T'$.
+    It follows that there are on every path (not only caterpillar paths) from $T$ to $T'$ there are $n-1$ $\rnni$ moves needed to move the parent of $n+1$ down in the tree, since each $\rnni$ move can decrease the rank of an internal node by at most one.
+    Let us consider the trees $T|_n$ to $T'|_n$ that result from $T$ and $T'$ by deleting taxon $n+1$.
+    When running Bubble Sort to compute a path from $T|_n$ to $T'|_n$ one can observe the same exchanges of taxa as on the caterpillar path from $T$ to $T'$, except for the $n-1$ ones that involve taxon $n+1$.
+    Hence, the caterpillar distance $d_c(T|_n, T'|_n)$ is $\frac{n(n-1)}{2} - (n-1)$ = $\frac{(n-1)(n-2)}{2}$.
+    So we can apply the induction hypothesis on $T|_n$ and $T'|_n$ and know that $d(T|_n,T'|_n) = \frac{(n-1)(n-2)}{2}$.
+
+    In general, deleting taxon $n+1$ and its parent in every tree on any path $p$ from $T$ to $T'$ results in a path $p'$ from $T|_n$ to $T'|n$.
+    Such a path $p'$ from $T|_n$ to $T'|n$ does not contain the $n-1$ $\rnni$ moves for moving the parent of $n+1$ down from the root to the cherry, so it is at least $n-1$ $\rnni$ moves shorter than $p$.
+    If we assume that there is a path from $T$ to $T'$ of length less than $\Delta(\rnniu_{n+1}) = \frac{n(n-1)}{2}$, we could delete taxon $n+1$ and the $n-1$ moves of it's parent and receive a path from $T|_n$ to $T'|_n$ of length less than $\frac{n(n-1)}{2} - (n-1)$ = $\frac{(n-1)(n-2)}{2}$.
+    So it would be $d(T|_n, T'|_n) < \Delta(\rnniu_n)$.
+    Since this is a contradiction to $d(T|_n, T'|_n) = \Delta(\rnniu_n)$, it must be $d(T,T') = \Delta(\rnniu_{n+1}) = \frac{n(n-1)}{2} $.
+
+\end{proof}
+
+\begin{theorem}
+    The set of caterpillar trees is convex.
+\end{theorem}
+
+\begin{proof}
+    We prove the Theorem by induction on the length $l$ of a shortest caterpillar path between two caterpillar trees $T_1$ and $T_2$.
+
+    Induction basis: $l = \Delta(\rnniu) = \frac{(n-1)(n-2)}{2}$\\
+    With Lemma~\ref{lemma:caterpillar_dist=diameter} one can see that for each pair of caterpillar trees $T_1$, $T_2$ with distance $d(T_1,T_2) = \Delta(\rnniu)$ there is a caterpillar path from $T_1$ to $T_2$ between these trees.
+
+    Induction step: $l+1 \to l$\\
+    Induction hypothesis: For each pait of caterpillar trees with distance $l+1$ there is a shortest path between these trees that only consists of caterpillar trees.
+    Let $T_1$ and $T_2$ be caterpillar trees with caterpillar distance $l$ \todo{define 'caterpillar distance'}.
+    If $l \neq \Delta(\rnniu)$, there must be a pair of taxa $x,y$ that are neighboured in $T_2$ and have the same order in $T_1$ and $T_2$, which means that $rank(p_x) > rank(p_y)$ or $rank(p_x) < rank(p_y)$ holds in both trees $T_1$ and $T_2$.
+    Then it follows that the tree $T_2'$ that equals $T_2$, but with exchanged taxa $x$ and $y$, has distance $l+1$ to $T_1$. \todo{really??}
+    So there must be a shortest path from $T_2'$ to $T_1$ that only consists of caterpillar trees.
+    Since it does not matter in which order the pairs of taxa that need to exchange their position do exchange on a path between two caterpillar trees, $T_2$ is on a shortest caterpillar path from $T_2'$ to $T_1$.
+    Therefore, it must be $d(T_1, T_2) = l$.
+
+\end{proof}
+
 \section{Partition lattice}

### 2019-05-23 paper.tex
update proof convexity - there's still a bit missing

@@
 %leaf or taxon?
+
+%caterpillar path, shortest caterpillar path, caterpillar distance (def currently in proof of lemma:caterpillar_dist=diameter)
 
@@
 			\STATE $p = p+T$
-		\ENDWHILE
+		\ENDWHILE between these trees
 	\ENDFOR
@@
 
-\begin{lemma}
+\begin{lemma} between these trees
 	There are two caterpillar trees in $\rnniu$ with distance $\frac{1}{2}n^2-\frac{3}{2}n+1$.
@@
     Such a path $p'$ from $T|_n$ to $T'|n$ does not contain the $n-1$ $\rnni$ moves for moving the parent of $n+1$ down from the root to the cherry, so it is at least $n-1$ $\rnni$ moves shorter than $p$.
-    If we assume that there is a path from $T$ to $T'$ of length less than $\Delta(\rnniu_{n+1}) = \frac{n(n-1)}{2}$, we could delete taxon $n+1$ and the $n-1$ moves of it's parent and receive a path from $T|_n$ to $T'|_n$ of length less than $\frac{n(n-1)}{2} - (n-1)$ = $\frac{(n-1)(n-2)}{2}$.
+    If we assume that there is a path from $T$ to $T'$ of length less than $\Delta(\rnniu_{n+1}) = \frac{n(n-1)}{2}$, we could delete between these trees taxon $n+1$ and the $n-1$ moves of it's parent and receive a path from $T|_n$ to $T'|_n$ of length less than $\frac{n(n-1)}{2} - (n-1)$ = $\frac{(n-1)(n-2)}{2}$.
     So it would be $d(T|_n, T'|_n) < \Delta(\rnniu_n)$.
@@
 \begin{proof}
-    We prove the Theorem by induction on the length $l$ of a shortest caterpillar path between two caterpillar trees $T_1$ and $T_2$.
+    We prove the Theorem by backwards induction on the length $l$ of a shortest caterpillar path between two caterpillar trees $T$ and $T'$.
 
     Induction basis: $l = \Delta(\rnniu) = \frac{(n-1)(n-2)}{2}$\\
-    With Lemma~\ref{lemma:caterpillar_dist=diameter} one can see that for each pair of caterpillar trees $T_1$, $T_2$ with distance $d(T_1,T_2) = \Delta(\rnniu)$ there is a caterpillar path from $T_1$ to $T_2$ between these trees.
+    With Lemma~\ref{lemma:caterpillar_dist=diameter} one can see that for each pair of caterpillar trees $T$, $T'$ with distance $d(T,T') = \Delta(\rnniu)$ there is a caterpillar path that is a shortest path from $T$ to $T'$.
 
     Induction step: $l+1 \to l$\\
-    Induction hypothesis: For each pait of caterpillar trees with distance $l+1$ there is a shortest path between these trees that only consists of caterpillar trees.
-    Let $T_1$ and $T_2$ be caterpillar trees with caterpillar distance $l$ \todo{define 'caterpillar distance'}.
-    If $l \neq \Delta(\rnniu)$, there must be a pair of taxa $x,y$ that are neighboured in $T_2$ and have the same order in $T_1$ and $T_2$, which means that $rank(p_x) > rank(p_y)$ or $rank(p_x) < rank(p_y)$ holds in both trees $T_1$ and $T_2$.
-    Then it follows that the tree $T_2'$ that equals $T_2$, but with exchanged taxa $x$ and $y$, has distance $l+1$ to $T_1$. \todo{really??}
-    So there must be a shortest path from $T_2'$ to $T_1$ that only consists of caterpillar trees.
-    Since it does not matter in which order the pairs of taxa that need to exchange their position do exchange on a path between two caterpillar trees, $T_2$ is on a shortest caterpillar path from $T_2'$ to $T_1$.
-    Therefore, it must be $d(T_1, T_2) = l$.
+    Induction hypothesis: For each pair of caterpillar trees with distance $l+1$ there is a shortest path between these trees that only consists of caterpillar trees.
+    Let $T$ and $T'$ be caterpillar trees with caterpillar distance $l$.
+    Since $l < \Delta(\rnniu) = \frac{(n-1)(n-2)}{2}$, there is one loop $k$ in the Bubble Sort algorithm from $T'$ to $T$ where there are less than $n-2-k$ swaps of neighboured taxa.
+    This means that the taxon $a$ that has parent with rank $n-2-k$ in $T'$ does not need to move all the way up starting at the cherry.
+    Hence, there is a taxon $b$ that has parent with rank one less than the parent of taxon $a$ in $T'$ that does not exchange with $a$ on the path to $p$.
+\todo{$a$ and $b$ do not need to be neighbours in $T'$!}
+    Let $T''$ be the caterpillar tree resulting from $T'$ by exchanging $a$ and $b$.
+    It follows that $T'$ is on a shortest caterpillar path from $T''$ to $T$ and therefore, $d_c(T,T'') = d_c(T,T') + 1 = l+1$.
+    We can apply the induction hypothesis on $T$ and $T''$ and know that $d(T,T'') = l+1$.
+    If it was $d(T,T') < l$, it would hold $d(T,T'') \leq d(T,T') + d(T',T'') < l + 1$ which contradicts the induction hypothesis.
+    We can conclude that $d(T,T') = l$.
 

### 2019-05-23 paper.tex
fix proof caterpillar convexity

@@
     Since $l < \Delta(\rnniu) = \frac{(n-1)(n-2)}{2}$, there is one loop $k$ in the Bubble Sort algorithm from $T'$ to $T$ where there are less than $n-2-k$ swaps of neighboured taxa.
-    This means that the taxon $a$ that has parent with rank $n-2-k$ in $T'$ does not need to move all the way up starting at the cherry.
-    Hence, there is a taxon $b$ that has parent with rank one less than the parent of taxon $a$ in $T'$ that does not exchange with $a$ on the path to $p$.
-\todo{$a$ and $b$ do not need to be neighbours in $T'$!}
+    This means that there is a taxon $a$ that has parent with rank $n-2-k$ in $T'$ and $a$ does not need to move all the way up starting at the cherry in loop $k$.
+    Therefore, there must be a taxon $b$ which has parent with rank one less than the parent of $a$ in the tree at the beginning of loop number $k$.
+    If we assume that loop $k$ is the first loop where a single taxon is not moved by $n-k-2$ $\rnni$ moves, it follows that $a$ and $b$ must be neighboured taxa in $T'$.
     Let $T''$ be the caterpillar tree resulting from $T'$ by exchanging $a$ and $b$.
-    It follows that $T'$ is on a shortest caterpillar path from $T''$ to $T$ and therefore, $d_c(T,T'') = d_c(T,T') + 1 = l+1$.
+    It follows that $T'$ has caterpillar distance $d_c(T,T'') = d_c(T,T') + 1 = l+1$ to $T$ and $d(T',T'') = 1$.
     We can apply the induction hypothesis on $T$ and $T''$ and know that $d(T,T'') = l+1$.

### 2019-05-23 paper.tex
minor

@@
 			\STATE $p = p+T$
-		\ENDWHILE between these trees
+		\ENDWHILE
 	\ENDFOR
@@
 
-\begin{lemma} between these trees
+\begin{lemma}
 	There are two caterpillar trees in $\rnniu$ with distance $\frac{1}{2}n^2-\frac{3}{2}n+1$.
@@
     Such a path $p'$ from $T|_n$ to $T'|n$ does not contain the $n-1$ $\rnni$ moves for moving the parent of $n+1$ down from the root to the cherry, so it is at least $n-1$ $\rnni$ moves shorter than $p$.
-    If we assume that there is a path from $T$ to $T'$ of length less than $\Delta(\rnniu_{n+1}) = \frac{n(n-1)}{2}$, we could delete between these trees taxon $n+1$ and the $n-1$ moves of it's parent and receive a path from $T|_n$ to $T'|_n$ of length less than $\frac{n(n-1)}{2} - (n-1)$ = $\frac{(n-1)(n-2)}{2}$.
+    If we assume that there is a path from $T$ to $T'$ of length less than $\Delta(\rnniu_{n+1}) = \frac{n(n-1)}{2}$, we could delete taxon $n+1$ and the $n-1$ moves of it's parent and receive a path from $T|_n$ to $T'|_n$ of length less than $\frac{n(n-1)}{2} - (n-1)$ = $\frac{(n-1)(n-2)}{2}$.
     So it would be $d(T|_n, T'|_n) < \Delta(\rnniu_n)$.

### 2019-05-26 paper.tex
change formulation lemma:caterpillar_dist=diameter

@@
 \begin{lemma}
-    Let $T$ and $T'$ be two caterpillar trees with distance $d(T,T') = \Delta(\rnniu) = \frac{(n-1)(n-2)}{2}$.
-    There is a shortest path between $T$ and $T'$ that is a caterpillar path.
+    Let $T$ and $T'$ be two caterpillar trees.
+    $T$ and $T'$ have distance $d(T,T') = \Delta(\rnniu) = \frac{(n-1)(n-2)}{2}$ if, and only if, they have caterpillar distance $d_c(T,T') = \Delta(\rnniu) = \frac{(n-1)(n-2)}{2}$.
+\todo{define $d_c$}
     \label{lemma:caterpillar_dist=diameter}
@@
     It follows that the maximum length of a path computed by Bubble Sort is $\sum\limits_{k=1}^{n-2} k = \frac{(n-1)(n-2)}{2}$.
-    Since there cannot be a caterpillar path shorter than $\frac{(n-1)(n-2)}{2}$, if the distance is $d(T,T') \frac{(n-1)(n-2)}{2}$, it follows $d_c(T,T') = \Delta(\rnniu)$.
+    Since there cannot be a caterpillar path shorter than $\frac{(n-1)(n-2)}{2}$ if it is $d(T,T') = \frac{(n-1)(n-2)}{2}$, it follows $d_c(T,T') = \Delta(\rnniu)$.
 

### 2019-05-26 paper.tex
minor

@@
     Let now $T$ and $T'$ have caterpillar distance $d_c(T,T') = \Delta(\rnniu)$.
-    We prove $d(T,T') = d_c(T,T')$ by induction on the number of taxa $n$ of $T$ ad $T'$:
+    We prove $d(T,T') = d_c(T,T')$ by induction on the number of taxa $n$ of $T$ and $T'$:
     We assume without loss of generality that $T$ is the caterpillar tree $(\ldots ((1,2),3), \ldots, n)$.
@@
     Induction hypothesis: For trees $T$, $T'$ on $n$ taxa it is $d_c(T,T') = d(T, T') = \frac{(n-1)(n-2)}{2} = \Delta(\rnniu_n)$.\\
-    Let $T, T'$ be trees on $n+1$ taxa with caterpillar distance $\Delta(\rnniu)$.
+    Let $T, T'$ be trees on $n+1$ taxa with caterpillar distance $\Delta(\rnniu_{n+1})$.
     Since we can reach $T'$ from $T$ by using Bubble Sort where each pair of taxa that is compared in that algorithm exchanges, we can conclude that taxon $n+1$ must be in the cherry of $T'$.
-    It follows that there are on every path (not only caterpillar paths) from $T$ to $T'$ there are $n-1$ $\rnni$ moves needed to move the parent of $n+1$ down in the tree, since each $\rnni$ move can decrease the rank of an internal node by at most one.
+    It follows that there are on every path (not only caterpillar paths) from $T$ to $T'$ $n-1$ $\rnni$ moves needed to move the parent of $n+1$ down in the tree, since each $\rnni$ move can decrease the rank of an internal node by at most one.
     Let us consider the trees $T|_n$ to $T'|_n$ that result from $T$ and $T'$ by deleting taxon $n+1$.
     When running Bubble Sort to compute a path from $T|_n$ to $T'|_n$ one can observe the same exchanges of taxa as on the caterpillar path from $T$ to $T'$, except for the $n-1$ ones that involve taxon $n+1$.
-    Hence, the caterpillar distance $d_c(T|_n, T'|_n)$ is $\frac{n(n-1)}{2} - (n-1)$ = $\frac{(n-1)(n-2)}{2}$.
+    Hence, the caterpillar distance is $d_c(T|_n, T'|_n) = \frac{n(n-1)}{2} - (n-1)$ = $\frac{(n-1)(n-2)}{2}$.
     So we can apply the induction hypothesis on $T|_n$ and $T'|_n$ and know that $d(T|_n,T'|_n) = \frac{(n-1)(n-2)}{2}$.

### 2019-05-26 paper.tex
few details in proof of caterpillar convexity

@@
     Induction basis: $l = \Delta(\rnniu) = \frac{(n-1)(n-2)}{2}$\\
-    With Lemma~\ref{lemma:caterpillar_dist=diameter} one can see that for each pair of caterpillar trees $T$, $T'$ with distance $d(T,T') = \Delta(\rnniu)$ there is a caterpillar path that is a shortest path from $T$ to $T'$.
+    With Lemma~\ref{lemma:caterpillar_dist=diameter} one can see that for each pair of caterpillar trees $T$, $T'$ with caterpillar distance $d_c(T,T') = \Delta(\rnniu)$, it is $d(T,T') = d_c(T,T')$, which means that there is a shortest path between $T$ and $T'$ that is a caterpillar path.
 
     Induction step: $l+1 \to l$\\
-    Induction hypothesis: For each pair of caterpillar trees with distance $l+1$ there is a shortest path between these trees that only consists of caterpillar trees.
+    Induction hypothesis: For each pair of caterpillar trees with caterpillar distance $l+1$ there is a shortest path between these trees that only consists of caterpillar trees, which means that the distance between these trees is $l+1$.
     Let $T$ and $T'$ be caterpillar trees with caterpillar distance $l$.
     Since $l < \Delta(\rnniu) = \frac{(n-1)(n-2)}{2}$, there is one loop $k$ in the Bubble Sort algorithm from $T'$ to $T$ where there are less than $n-2-k$ swaps of neighboured taxa.
-    This means that there is a taxon $a$ that has parent with rank $n-2-k$ in $T'$ and $a$ does not need to move all the way up starting at the cherry in loop $k$.
+    This means that there is a taxon $a$ that has parent with rank $n-2-k$ in $T$ and that does not need to move all the way up starting at the cherry in loop $k$.
     Therefore, there must be a taxon $b$ which has parent with rank one less than the parent of $a$ in the tree at the beginning of loop number $k$.

### 2019-06-05 paper.tex
max caterpillar distance <=> no shared triplet

@@
     In general, deleting taxon $n+1$ and its parent in every tree on any path $p$ from $T$ to $T'$ results in a path $p'$ from $T|_n$ to $T'|n$.
-    Such a path $p'$ from $T|_n$ to $T'|n$ does not contain the $n-1$ $\rnni$ moves for moving the parent of $n+1$ down from the root to the cherry, so it is at least $n-1$ $\rnni$ moves shorter than $p$.
-    If we assume that there is a path from $T$ to $T'$ of length less than $\Delta(\rnniu_{n+1}) = \frac{n(n-1)}{2}$, we could delete taxon $n+1$ and the $n-1$ moves of it's parent and receive a path from $T|_n$ to $T'|_n$ of length less than $\frac{n(n-1)}{2} - (n-1)$ = $\frac{(n-1)(n-2)}{2}$.
+    Such a path $p'$ from $T|_n$ to $T'|n$ does not contain moves involving the parent of $n+1$.
+    Since the parent of $n+1$ need to move from the root of $T_1$ to the cherry of $T_2$ and each $\rnni$ move can change the rank of a node by at most one, $p'$ is at least $n-1$ $\rnni$ moves shorter than $p$.
+\todo{Really?}
+    If we assume that there is a path from $T$ to $T'$ of length less than $\Delta(\rnniu_{n+1}) = \frac{n(n-1)}{2}$, we could delete taxon $n+1$ and at least $n-1$ moves of it's parent and receive a path from $T|_n$ to $T'|_n$ of length less than $\frac{n(n-1)}{2} - (n-1)$ = $\frac{(n-1)(n-2)}{2}$.
     So it would be $d(T|_n, T'|_n) < \Delta(\rnniu_n)$.
@@
 
+% idea for description of max distance caterpillar trees:
+
+\begin{lemma}
+    Let $T_1$ and $T_2$ be caterpillar trees.
+    It is $d_c(T_1,T_2) = \Delta(\rnniu)$ if, and only if, $T_1$ and $T_2$ share no induced triplet.
+    \todo{define induced triplets}
+\end{lemma}
+
+This Lemma does only hold for caterpillar trees and not for trees with more than one cherry.
+Furthermore, triplets do not represent ranked trees uniquely as they do not contain information about the ranks of internal nodes (but they can uniquely define caterpillar trees as these only have one cherry)
+
+\begin{proof}
+    % Insert the proof here
+    Fix $T_1 = (\ldots(1,2) \ldots ,n)$, use induction on $n$ and bubble sort (taxon $n$ must be in cherry of $T_2$ if distance is max).
+\end{proof}
+
 \section{Partition lattice}

### 2019-06-10 paper.tex
conj for alg computing trees with max distance from given tree

@@
 
+\begin{conjecture}
+    Algorithm~\ref{alg:max_dist_tree} computes a tree with maximum distance $\Delta(\rnniu)$ from a given tree $T$.
+\end{conjecture}
+
+According to Algorithm~\ref{alg:max_dist_tree}, the number of trees with maximum distance to a given tree $T$ depends on the number of cherries of $T$.
+For caterpillar trees it is $(n-1)!$.
+
+\begin{algorithm}[H]
+\caption{MAX\_DISTANCE\_TREE($T$)}
+\label{alg:max_dist_tree}
+\begin{algorithmic}[1]
+	\STATE $R$ only contains cherry taxa of $T$ and has internal node with rank $n-1$
+    \STATE In list $L$ all taxa $\{1,\ldots,n\}$ are sorted such that the ranks of their parents in $T$ do not decrease
+	\FOR {taxon $a$ in $L$}
+		\STATE Add an internal node as parent of $a$ on any edge in $R$ such that the new internal node has rank one less than the previously lowest internal node
+	\ENDFOR
+	\RETURN $R$
+\end{algorithmic}
+\end{algorithm}
+
 \section{Partition lattice}

### 2019-06-10 paper.tex
conj: max dist(T,R) <=> alg computes R given T

@@
 \begin{conjecture}
-    Algorithm~\ref{alg:max_dist_tree} computes a tree with maximum distance $\Delta(\rnniu)$ from a given tree $T$.
+    Let $T_1$ and $T_2$ be trees on $n$ taxa.
+    It is $d(T_1,T_2) = \Delta(\rnniu)$ if, and only if, $T_2$ can be produced from $T_1$ by applying Algorithm~\ref{alg:max_dist_tree}.
 \end{conjecture}
 
-According to Algorithm~\ref{alg:max_dist_tree}, the number of trees with maximum distance to a given tree $T$ depends on the number of cherries of $T$.
+According to Algorithm~\ref{alg:max_dist_tree}, the number of trees with maximum distance to a given tree $T$ depends on the number of cherries of $T$ and the ranks of their internal nodes.
 For caterpillar trees it is $(n-1)!$.

### 2019-06-11 paper.tex
connection alg:find_path and alg:max_dist_tree

@@
 
-The proof of Theorem 7 in \cite{Gavryushkin2017} gives an algorithm to compute paths between trees in $\rnni$.
+The proof of Theorem 7 in \cite{Gavryushkin2017} gives an algorithm to compute paths between trees in $\rnni$. %Do we want to mention that algorithm here?
 
@@
 \begin{corollary}
-It is $\Delta(\rnniu) = \frac{1}{2}n^2-\frac{3}{2}n+1$.
+It is $\Delta(\rnniu) = \frac{(n-1)(n-2)}{2} = \frac{1}{2}n^2-\frac{3}{2}n+1$.
 \end{corollary}
@@
 \begin{algorithmic}[1]
-	\STATE $R$ only contains cherry taxa of $T$ and has internal node with rank $n-1$
-    \STATE In list $L$ all taxa $\{1,\ldots,n\}$ are sorted such that the ranks of their parents in $T$ do not decrease
+	\STATE $R$ only contains cherry taxa $(a,b)$ of $T$ and has internal node with rank $n-1$
+    \STATE In list $L$ all taxa $\{1,\ldots,n\}\setminus\{a,b\}$ are sorted such that the ranks of their parents in $T$ do not decrease
 	\FOR {taxon $a$ in $L$}
@@
 \end{algorithm}
+
+\begin{lemma}
+    Let $T$ and $R$ be trees.
+    Algorithm~\ref{alg:find_path} gives a path of length $\frac{(n-1)(n-2)}{2}$ between $T$ and $R$ if, and only if, $R$ can be generated from $T$ by applying Algorithm~\ref{alg:max_dist_tree}.
+\end{lemma}
 

### 2019-06-13 paper.tex
bound for length of path when deleting one taxon from all trees

@@
 
+%This lemma might be central  for induction proofs in RNNI space
+%Is it correct and if it is, do we need to prove the statement, or is it rather obvious?
+%The lemma implies another algorithm to compute distances (it would not give shortest paths explicitly, unlike alg:find_path)
+\begin{lemma}
+    Let $T_1$ and $T_2$ be two trees on $n+1$ taxa.
+    Let $T_1|_n$ and $T_2|_n$ be these trees restricted to $n$ taxa, which means that the same taxon $n+1$ is deleted from both trees $T_1$ and $T_2$.
+    Then it is $d(T_1|_n, T_2|_n) \leq d(T_1,T_2) - |r_{T_1}(p_{n+1}) - r_{T_2}(p_{n+1})|$ where $r_{T_i}(p_j)$ determines the rank of the parent of taxon $i$ in tree $T_i$.
+\end{lemma}
+
+\begin{proof}
+    Induction on $k = d(T_1,T_2)$:
+
+    Basis($k = 1$):
+    If rank of $p_{n+1}$ changes between $T_1$ and $T_2$, it is $d(T_1|_n, T_2|_n) = 0$ and $|r_{T_1}(p_{n+1}) - r_{T_2}(p_{n+1})| = 1$ and therefore, $d(T_1|_n, T_2|_n) = 0 \leq 1 - 1 = d(T_1,T_2) - |r_{T_1}(p_{n+1}) - r_{T_2}(p_{n+1})|$.
+    If on the other hand the rank of $p_{n+1}$ does not change between $T_1$ and $T_2$, it is $d(T_1|_n, T_2|_n) = d(T_1,T_2) = 1$ and $|r_{T_1}(p_{n+1}) - r_{T_2}(p_{n+1})| = 1$ and therefore, $d(T_1|_n, T_2|_n) = 1 \leq 1 - 0 = d(T_1,T_2) - |r_{T_1}(p_{n+1}) - r_{T_2}(p_{n+1})|$.
+
+    Step($k \to k+1$):
+    Let $T_1$ ad $T_2$ be trees on $n+1$ taxa with distance $d(T_1,T_2) = k$.
+    Let $T_2'$ be the last tree on a shortest path from $T_1$ to $T_2$, i.e. $d(T_1,T_2') = k$.
+    We can apply the induction hypothesis on $T_1$ and $T_2'$ and see that $d(T_1|_n, T_2'|_n) \leq d(T_1, T_2') - |r_{T_1}(p_{n+1}) - r_{T_2'}(p_{n+1})| := k - m$.
+    Now we differ two cases:
+    \begin{enumerate}
+        \item $p_{n+1}$ changes its rank between $T_2$ and $T_2'$.
+
+        Then the trees $T_2|_n$ and $T_2'|_n$ are equal. \todo{Is this obivous?}
+        Therefore, it is $d(T_1|_n,T_2|_n) = d(T_1|_n, T_2'|_n) \leq k - m$.
+        Since the rank of $p_{n+1}$ changes by one between $T_2$ and $T_2'$ it is either $|r_{T_1}(p_{n+1}) - r_{T_2}(p_{n+1})| = m+1$ or $|r_{T_1}(p_{n+1}) - r_{T_2}(p_{n+1})| = m-1$.
+        In either case, we have $d(T_1|_n,T_2|_n) = d(T_1|_n, T_2'|_n) \leq k - m = (k+1) - (m-1) \leq d(T_1, T_2) - |r_{T_1}(p_{n+1}) - r_{T_2}(p_{n+1})|$.
+
+        \item $p_{n+1}$ has same rank in $T_2$ and $T_2'$.
+
+        In this case it is $d(T_1|_n,T_2|_n) = d(T_1|_n, T_2'|_n) + 1 \leq (k-m)+ 1$.
+        It also follows that $|r_{T_1}(p_{n+1}) - r_{T_2'}(p_{n+1})| = |r_{T_1}(p_{n+1}) - r_{T_2}(p_{n+1})|$ and therefore, we can conclude $d(T_1|_n,T_2|_n) \leq k-m+1 = k+1-m = d(T_1,T_2) -  |r_{T_1}(p_{n+1}) - r_{T_2}(p_{n+1})|$.
+    \end{enumerate}
+\end{proof}
+
 \section{Partition lattice}

### 2019-06-13 paper.tex
minor

@@
     Let $T_1$ and $T_2$ be two trees on $n+1$ taxa.
-    Let $T_1|_n$ and $T_2|_n$ be these trees restricted to $n$ taxa, which means that the same taxon $n+1$ is deleted from both trees $T_1$ and $T_2$.
-    Then it is $d(T_1|_n, T_2|_n) \leq d(T_1,T_2) - |r_{T_1}(p_{n+1}) - r_{T_2}(p_{n+1})|$ where $r_{T_i}(p_j)$ determines the rank of the parent of taxon $i$ in tree $T_i$.
+    Let $T_1|_n$ and $T_2|_n$ be these trees restricted to $n$ taxa, i.e. the same taxon $n+1$ is deleted from both trees $T_1$ and $T_2$.
+    Then it is $d(T_1|_n, T_2|_n) \leq d(T_1,T_2) - |r_{T_1}(p_{n+1}) - r_{T_2}(p_{n+1})|$ where $r_{T_i}(p_j)$ determines the rank of the parent of taxon $j$ in tree $T_i$.
 \end{lemma}
@@
     Step($k \to k+1$):
-    Let $T_1$ ad $T_2$ be trees on $n+1$ taxa with distance $d(T_1,T_2) = k$.
+    Let $T_1$ and $T_2$ be trees on $n+1$ taxa with distance $d(T_1,T_2) = k$.
     Let $T_2'$ be the last tree on a shortest path from $T_1$ to $T_2$, i.e. $d(T_1,T_2') = k$.
@@
 
-        In this case it is $d(T_1|_n,T_2|_n) = d(T_1|_n, T_2'|_n) + 1 \leq (k-m)+ 1$.
+        In this case it is $d(T_1|_n,T_2|_n) \leq d(T_1|_n,T_2'|_n) + d(T_2'|_n, T_2|_n) \leq d(T_1|_n, T_2'|_n) + 1 \leq (k-m)+ 1$.
         It also follows that $|r_{T_1}(p_{n+1}) - r_{T_2'}(p_{n+1})| = |r_{T_1}(p_{n+1}) - r_{T_2}(p_{n+1})|$ and therefore, we can conclude $d(T_1|_n,T_2|_n) \leq k-m+1 = k+1-m = d(T_1,T_2) -  |r_{T_1}(p_{n+1}) - r_{T_2}(p_{n+1})|$.

### 2019-06-13 paper.tex
minor

@@
     \begin{enumerate}
-        \item $p_{n+1}$ changes its rank between $T_2$ and $T_2'$.
+        \item $p_{n+1}$ changes its rank between $T_2$ and $T_2'$ or is affected by the $\nni$ move is performed between $T_2$ and $T_2'$.
 
@@
         Therefore, it is $d(T_1|_n,T_2|_n) = d(T_1|_n, T_2'|_n) \leq k - m$.
-        Since the rank of $p_{n+1}$ changes by one between $T_2$ and $T_2'$ it is either $|r_{T_1}(p_{n+1}) - r_{T_2}(p_{n+1})| = m+1$ or $|r_{T_1}(p_{n+1}) - r_{T_2}(p_{n+1})| = m-1$.
-        In either case, we have $d(T_1|_n,T_2|_n) = d(T_1|_n, T_2'|_n) \leq k - m = (k+1) - (m-1) \leq d(T_1, T_2) - |r_{T_1}(p_{n+1}) - r_{T_2}(p_{n+1})|$.
+        It is either $|r_{T_1}(p_{n+1}) - r_{T_2}(p_{n+1})| = m+1$, $|r_{T_1}(p_{n+1}) - r_{T_2}(p_{n+1})| = m$, or $|r_{T_1}(p_{n+1}) - r_{T_2}(p_{n+1})| = m-1$.
+        In either case, we have $d(T_1|_n,T_2|_n) = d(T_1|_n, T_2'|_n) \leq k - m = (k+1) - (m+1) \leq d(T_1, T_2) - |r_{T_1}(p_{n+1}) - r_{T_2}(p_{n+1})|$.
 

### 2019-06-13 paper.tex
restructure + minor edits on caterpillar convexity

@@
 
+%ranks of internal node shall be from 1 to n, leaves have rank 0
+
 %MRCA
@@
 %caterpillar path, shortest caterpillar path, caterpillar distance (def currently in proof of lemma:caterpillar_dist=diameter)
+
+%convexity (there is a shortest path within that set, not: all shortest path are within this set)
 
@@
 
+%This lemma might be central  for induction proofs in RNNI space
+%Is it correct and if it is, do we need to prove the statement, or is it rather obvious?
+%The lemma implies another algorithm to compute distances (it would not give shortest paths explicitly, unlike alg:find_path)
+\begin{lemma}
+    Let $T_1$ and $T_2$ be two trees on $n+1$ taxa.
+    Let $T_1|_n$ and $T_2|_n$ be these trees restricted to $n$ taxa, i.e. the same taxon $n+1$ is deleted from both trees $T_1$ and $T_2$.
+    Then it is $d(T_1|_n, T_2|_n) \leq d(T_1,T_2) - |r_{T_1}(p_{n+1}) - r_{T_2}(p_{n+1})|$ where $r_{T_i}(p_j)$ determines the rank of the parent of taxon $j$ in tree $T_i$.
+    \label{lemma:distance_delete_taxon}
+\end{lemma}
+
+\begin{proof}
+    Induction on $k = d(T_1,T_2)$:
+
+    Basis($k = 1$):
+    If rank of $p_{n+1}$ changes between $T_1$ and $T_2$, it is $d(T_1|_n, T_2|_n) = 0$ and $|r_{T_1}(p_{n+1}) - r_{T_2}(p_{n+1})| = 1$ and therefore, $d(T_1|_n, T_2|_n) = 0 \leq 1 - 1 = d(T_1,T_2) - |r_{T_1}(p_{n+1}) - r_{T_2}(p_{n+1})|$.
+    If on the other hand the rank of $p_{n+1}$ does not change between $T_1$ and $T_2$, it is $d(T_1|_n, T_2|_n) = d(T_1,T_2) = 1$ and $|r_{T_1}(p_{n+1}) - r_{T_2}(p_{n+1})| = 1$ and therefore, $d(T_1|_n, T_2|_n) = 1 \leq 1 - 0 = d(T_1,T_2) - |r_{T_1}(p_{n+1}) - r_{T_2}(p_{n+1})|$.
+
+    Step($k \to k+1$):
+    Let $T_1$ and $T_2$ be trees on $n+1$ taxa with distance $d(T_1,T_2) = k$.
+    Let $T_2'$ be the last tree on a shortest path from $T_1$ to $T_2$, i.e. $d(T_1,T_2') = k$.
+    We can apply the induction hypothesis on $T_1$ and $T_2'$ and see that $d(T_1|_n, T_2'|_n) \leq d(T_1, T_2') - |r_{T_1}(p_{n+1}) - r_{T_2'}(p_{n+1})| := k - m$.
+    Now we differ two cases:
+    \begin{enumerate}
+        \item $p_{n+1}$ changes its rank between $T_2$ and $T_2'$ or is affected by the $\nni$ move is performed between $T_2$ and $T_2'$.
+
+        Then the trees $T_2|_n$ and $T_2'|_n$ are equal. \todo{Is this obivous?}
+        Therefore, it is $d(T_1|_n,T_2|_n) = d(T_1|_n, T_2'|_n) \leq k - m$.
+        It is either $|r_{T_1}(p_{n+1}) - r_{T_2}(p_{n+1})| = m+1$, $|r_{T_1}(p_{n+1}) - r_{T_2}(p_{n+1})| = m$, or $|r_{T_1}(p_{n+1}) - r_{T_2}(p_{n+1})| = m-1$.
+        In either case, we have $d(T_1|_n,T_2|_n) = d(T_1|_n, T_2'|_n) \leq k - m = (k+1) - (m+1) \leq d(T_1, T_2) - |r_{T_1}(p_{n+1}) - r_{T_2}(p_{n+1})|$.
+
+        \item $p_{n+1}$ has same rank in $T_2$ and $T_2'$.
+
+        In this case it is $d(T_1|_n,T_2|_n) \leq d(T_1|_n,T_2'|_n) + d(T_2'|_n, T_2|_n) \leq d(T_1|_n, T_2'|_n) + 1 \leq (k-m)+ 1$.
+        It also follows that $|r_{T_1}(p_{n+1}) - r_{T_2'}(p_{n+1})| = |r_{T_1}(p_{n+1}) - r_{T_2}(p_{n+1})|$ and therefore, we can conclude $d(T_1|_n,T_2|_n) \leq k-m+1 = k+1-m = d(T_1,T_2) -  |r_{T_1}(p_{n+1}) - r_{T_2}(p_{n+1})|$.
+    \end{enumerate}
+\end{proof}
+
 \begin{lemma}
@@
     We can compute a  shortest caterpillar path from $T$ to $T'$ using Bubble Sort.
-    In each loop $k=1, \ldots, n-2$ of the algorithm there can be at most $n-2-k$ exchanges of taxa.
+    In each loop $k=1, \ldots, n-2$ of the algorithm there can be at most $n-1-k$ exchanges of taxa.
     It follows that the maximum length of a path computed by Bubble Sort is $\sum\limits_{k=1}^{n-2} k = \frac{(n-1)(n-2)}{2}$.
@@
     Induction step: $n \to n+1$\\
-    Induction hypothesis: For trees $T$, $T'$ on $n$ taxa it is $d_c(T,T') = d(T, T') = \frac{(n-1)(n-2)}{2} = \Delta(\rnniu_n)$.\\
+    Induction hypothesis: For trees $R$, $R'$ on $n$ taxa it is $d_c(R,R') = d(R, R') = \frac{(n-1)(n-2)}{2} = \Delta(\rnniu_n)$.\\
     Let $T, T'$ be trees on $n+1$ taxa with caterpillar distance $\Delta(\rnniu_{n+1})$.
-    Since we can reach $T'$ from $T$ by using Bubble Sort where each pair of taxa that is compared in that algorithm exchanges, we can conclude that taxon $n+1$ must be in the cherry of $T'$.
-    It follows that there are on every path (not only caterpillar paths) from $T$ to $T'$ $n-1$ $\rnni$ moves needed to move the parent of $n+1$ down in the tree, since each $\rnni$ move can decrease the rank of an internal node by at most one.
+    When applying the Bubble Sort algorithm to compute a path between $T'$ and $T$, each pair of taxa that is compared in that algorithm exchanges, since there must be $\Delta(\rnniu)=\frac{(n-1)(n-2)}{2}$ moves between $T$ and $T'$.
+    So we can conclude that taxon $n+1$ must be in the cherry of $T'$.
     Let us consider the trees $T|_n$ to $T'|_n$ that result from $T$ and $T'$ by deleting taxon $n+1$.
-    When running Bubble Sort to compute a path from $T|_n$ to $T'|_n$ one can observe the same exchanges of taxa as on the caterpillar path from $T$ to $T'$, except for the $n-1$ ones that involve taxon $n+1$.
+    When running Bubble Sort to compute a path from $T|_n$ to $T'|_n$ one can observe the same exchanges of taxa as on the Bubble Sort path from $T$ to $T'$, except for the $n-1$ ones that involve taxon $n+1$.
     Hence, the caterpillar distance is $d_c(T|_n, T'|_n) = \frac{n(n-1)}{2} - (n-1)$ = $\frac{(n-1)(n-2)}{2}$.
@@
 
-    In general, deleting taxon $n+1$ and its parent in every tree on any path $p$ from $T$ to $T'$ results in a path $p'$ from $T|_n$ to $T'|n$.
-    Such a path $p'$ from $T|_n$ to $T'|n$ does not contain moves involving the parent of $n+1$.
-    Since the parent of $n+1$ need to move from the root of $T_1$ to the cherry of $T_2$ and each $\rnni$ move can change the rank of a node by at most one, $p'$ is at least $n-1$ $\rnni$ moves shorter than $p$.
-\todo{Really?}
-    If we assume that there is a path from $T$ to $T'$ of length less than $\Delta(\rnniu_{n+1}) = \frac{n(n-1)}{2}$, we could delete taxon $n+1$ and at least $n-1$ moves of it's parent and receive a path from $T|_n$ to $T'|_n$ of length less than $\frac{n(n-1)}{2} - (n-1)$ = $\frac{(n-1)(n-2)}{2}$.
+    With Lemma~\ref{lemma:distance_delete_taxon} we know that the distance between $T|_n$ and $T'|_n$ is at least $d(T|_n, T'|_n) \leq d(T,T') - |r_{T}(p_{n+1}) - r_{T'}(p_{n+1})|$.
+    We also know $|r_{T}(p_{n+1}) - r_{T'}(p_{n+1})| = n-1$
+    If we assume that there is a path from $T$ to $T'$ of length less than $\Delta(\rnniu_{n+1}) = \frac{n(n-1)}{2}$, it would follow that $d(T|_n, T'|_n) \leq d(T,T') - |r_{T}(p_{n+1}) - r_{T'}(p_{n+1})| < \frac{n(n-1)}{2} - (n-1) < \frac{(n-1)(n-2)}{2}$.
     So it would be $d(T|_n, T'|_n) < \Delta(\rnniu_n)$.
@@
 
-%This lemma might be central  for induction proofs in RNNI space
-%Is it correct and if it is, do we need to prove the statement, or is it rather obvious?
-%The lemma implies another algorithm to compute distances (it would not give shortest paths explicitly, unlike alg:find_path)
-\begin{lemma}
-    Let $T_1$ and $T_2$ be two trees on $n+1$ taxa.
-    Let $T_1|_n$ and $T_2|_n$ be these trees restricted to $n$ taxa, i.e. the same taxon $n+1$ is deleted from both trees $T_1$ and $T_2$.
-    Then it is $d(T_1|_n, T_2|_n) \leq d(T_1,T_2) - |r_{T_1}(p_{n+1}) - r_{T_2}(p_{n+1})|$ where $r_{T_i}(p_j)$ determines the rank of the parent of taxon $j$ in tree $T_i$.
-\end{lemma}
-
-\begin{proof}
-    Induction on $k = d(T_1,T_2)$:
-
-    Basis($k = 1$):
-    If rank of $p_{n+1}$ changes between $T_1$ and $T_2$, it is $d(T_1|_n, T_2|_n) = 0$ and $|r_{T_1}(p_{n+1}) - r_{T_2}(p_{n+1})| = 1$ and therefore, $d(T_1|_n, T_2|_n) = 0 \leq 1 - 1 = d(T_1,T_2) - |r_{T_1}(p_{n+1}) - r_{T_2}(p_{n+1})|$.
-    If on the other hand the rank of $p_{n+1}$ does not change between $T_1$ and $T_2$, it is $d(T_1|_n, T_2|_n) = d(T_1,T_2) = 1$ and $|r_{T_1}(p_{n+1}) - r_{T_2}(p_{n+1})| = 1$ and therefore, $d(T_1|_n, T_2|_n) = 1 \leq 1 - 0 = d(T_1,T_2) - |r_{T_1}(p_{n+1}) - r_{T_2}(p_{n+1})|$.
-
-    Step($k \to k+1$):
-    Let $T_1$ and $T_2$ be trees on $n+1$ taxa with distance $d(T_1,T_2) = k$.
-    Let $T_2'$ be the last tree on a shortest path from $T_1$ to $T_2$, i.e. $d(T_1,T_2') = k$.
-    We can apply the induction hypothesis on $T_1$ and $T_2'$ and see that $d(T_1|_n, T_2'|_n) \leq d(T_1, T_2') - |r_{T_1}(p_{n+1}) - r_{T_2'}(p_{n+1})| := k - m$.
-    Now we differ two cases:
-    \begin{enumerate}
-        \item $p_{n+1}$ changes its rank between $T_2$ and $T_2'$ or is affected by the $\nni$ move is performed between $T_2$ and $T_2'$.
-
-        Then the trees $T_2|_n$ and $T_2'|_n$ are equal. \todo{Is this obivous?}
-        Therefore, it is $d(T_1|_n,T_2|_n) = d(T_1|_n, T_2'|_n) \leq k - m$.
-        It is either $|r_{T_1}(p_{n+1}) - r_{T_2}(p_{n+1})| = m+1$, $|r_{T_1}(p_{n+1}) - r_{T_2}(p_{n+1})| = m$, or $|r_{T_1}(p_{n+1}) - r_{T_2}(p_{n+1})| = m-1$.
-        In either case, we have $d(T_1|_n,T_2|_n) = d(T_1|_n, T_2'|_n) \leq k - m = (k+1) - (m+1) \leq d(T_1, T_2) - |r_{T_1}(p_{n+1}) - r_{T_2}(p_{n+1})|$.
-
-        \item $p_{n+1}$ has same rank in $T_2$ and $T_2'$.
-
-        In this case it is $d(T_1|_n,T_2|_n) \leq d(T_1|_n,T_2'|_n) + d(T_2'|_n, T_2|_n) \leq d(T_1|_n, T_2'|_n) + 1 \leq (k-m)+ 1$.
-        It also follows that $|r_{T_1}(p_{n+1}) - r_{T_2'}(p_{n+1})| = |r_{T_1}(p_{n+1}) - r_{T_2}(p_{n+1})|$ and therefore, we can conclude $d(T_1|_n,T_2|_n) \leq k-m+1 = k+1-m = d(T_1,T_2) -  |r_{T_1}(p_{n+1}) - r_{T_2}(p_{n+1})|$.
-    \end{enumerate}
-\end{proof}
-
 \section{Partition lattice}

### 2019-06-25 paper.tex
number of ctrees with max dist from given ctree
lots of definitions

@@
 \usepackage{amsthm}
+\usepackage{amssymb}
 \usepackage{float}
@@
 
-%Do we restrict everything to RNNI? Make sure that we correctly distinguish between RNNI and RNNIu (for instance in the algorithm section)
+\todo{citation of definitions??}
 
-%ranks of internal node shall be from 1 to n, leaves have rank 0
+A \amph{rooted binary phylogenetic tree} a set $X$ is a pair $(t, \Phi)$ where $t$ is a rooted binary tree with leaf set $\mathcal{L}(t)$ and $\Phi:X \mapsto \mathcal{L}(T)$ a bijective mapping.
+\todo{do we need to define leaves and internal nodes, child, parent?}
+A \emph{ranked tree} is given by a pair of such a rooted binary phylogenetic tree $(t, \Phi)$ on $X = \{1, \ldots, n\}$ for $n \in \mathbb N$, and a rank function that maps the vertex set of $t$ onto the set $\{0,\ldots,n-1\}$ such that all leaves are assigned rank $0$ and each internal node has rank greater than its children.
+The number $n$ of taxa will remain fixed throughout the whole paper.
+For simplicity of notation we use the term \emph{tree} instead instead of ranked tree.
+Furthermore, we will identify leaves by their labels and might call them \emph{taxa} (singular: \emph{taxon}).
+For a tree $T$ a subtree consisting of an internal node with both children being leaves is called \emph{cherry}.
+An edge $e$ that has no leaf as child is an \emph{internal edge}.
+The set of taxa that are descending from an internal edge $e$ is called \emph{cluster}. \todo{OK?}
+In the literature, clusters often refer to sets of taxa descending from internal nodes.
+The set of clusters as defined above would then be recalled as non-trivial clusters.
+Given a subset $M$ of the set of taxa $X$, the \emph{most recent common ancestor} $mrca(M)$ of $M$ is the internal node with lowest rank that is ancestor of all taxa in $M$. \todo{OK?}
+For a single taxon $x$, $mrca(\{x\})$ is the parent of $x$ and will be denoted by $parent(x)$.
+Two trees are called \emph{isomorphic} if the underlying rooted phylogenetic trees are isomorphic and the isomorphism preserves ranks of the nodes.
 
-%MRCA
+The set of ranked trees builds the vertex set of the \emph{$\rnni$-graph}.
+We introduce two types of operations (\emph{$\rnni$ moves}) connecting trees $T$ and $R$ in that graph by an edge:
+The swap of the ranks of two internal nodes that have rank difference one is called \emph{rank move}.
+We say that there is an $\nni$ move between trees $T$ and $R$ if there are edges $e$ in $T$ and $f$ in $R$ that are bounded by internal nodes with rank difference one such that the graphs obtained by shrinking $e$ and $f$ to internal nodes are isomorphic.
+Trees that are connected by either a rank swap or an $\nni$ move are called \emph{$\rnni$ neighbours} or briefly neighbours.
+Note that the previously defined $\rnni$ graph is a metric space, which is why we may call it \emph{$\rnni$ space}.
 
-%cluster
+\todo{figure: rank + $\nni$ move}
 
-%leaf or taxon?
+The main purpose of this paper is understanding geometric properties of the $\rnni$ space.
+Therefore, further notations need to be introduced.
+If every internal node of a tree $T$ has at least one leaf as child, $T$ is called \emph{caterpillar tree}.
+A \emph{path} $p$ between trees $T_0$ and $T_m$ in $\rnni$ is a sequence $T_0, T_1, \ldots, T_m$ of trees such that there is an edge $\{T_i, T_{i+1}\}$ for all $i=0,\ldots,m-1$.
+For such a path $p$ we say the \emph{length} $|p|$ of $p$ is $m$.
+A path $p$ between trees $T$ and $R$ is a \emph{shortest path} if for every path $r$ between $T$ and $R$, $|p| \leq |r|$.
+The \emph{distance} $d(T,R)$ between trees $T$ and $R$ is the length of a shortest path between them.
+If $p$ is a path only consisting of caterpillar trees we call it \emph{caterpillar path}.
+For two caterpillar trees $T$ and $R$ a \emph{shortest caterpillar path} is a caterpillar path that is shortest among all caterpillar paths.
+The length of such a shortest caterpillar path is called \emph{caterpillar distance} and denoted by $d_c(T,R)$.
+A set $\mathbb S$ of trees is called \emph{convex} if for all pairs of trees $T,R \in \mathbb S$, there exists a shortest path from $T$ to $R$ such that every tree on $p$ is element of $\mathbb S$.
 
-%caterpillar path, shortest caterpillar path, caterpillar distance (def currently in proof of lemma:caterpillar_dist=diameter)
+%Is it sufficient to mention what RNNI stands for in the Introduction?
 
-%convexity (there is a shortest path within that set, not: all shortest path are within this set)
+%Restrict to RNNIu (See whether we can get same results for RNNI)
 
-\begin{definition}
-	Let $A$ be a pending subtree of the tree $T$.
-	We call the set $c(A)$ of taxa of $A$ the \emph{cluster} induced by $A$.
-\end{definition}
+%Definitions inside a tree: cherry, partition, edges represent partitions
+
+%If needed: diameter, radius (in the corresponding chapter?), induced triplet (in that chapter?)
+
+%delete definitions in the following chapters that are already covered here!
+
+%replace rank swap with rank move in the following!!
 
@@
 
-	\textbf{Rank swaps:}
-	Within rank swaps the labels of internal nodes do not change.
+	\textbf{Rank moves:}
+	Within rank moves the labels of internal nodes do not change.
 	For an example consider Figure~\ref{fig:rank_swap_internal_label}.
@@
 		\includegraphics[width=0.8\textwidth]{rank_swap_internal_label}
-		\caption{The labelling of the internal nodes that swap ranks (red) do not change within the rank swap.
+		\caption{The labelling of the internal nodes that swap ranks (red) do not change within the rank move.
 		}
@@
 	Within each $\rnni$ move the ranks of at most two internal nodes change by at most one each.
-	This is true for both rank swaps and $\nni$ moves, according to our definition of the labelling:
-	Within a rank swap two nodes exchange their ranks, while there are not necessarily changes of ranks when $\nni$ moves are performed.
+	This is true for both rank moves and $\nni$ moves, according to our definition of the labelling:
+	Within a rank move two nodes exchange their ranks, while there are not necessarily changes of ranks when $\nni$ moves are performed.
 
@@
 Introducing new cherries means introducing new internal nodes with new ranks.
-Since there are some rank swaps needed for an $\rnni$ path similar to the shortest $\nni$ path presented above, the path in $\rnni$ that breaks the split $1, \ldots, n | n+1, \ldots, 2n$ is no shortest path.
+Since there are some rank moves needed for an $\rnni$ path similar to the shortest $\nni$ path presented above, the path in $\rnni$ that breaks the split $1, \ldots, n | n+1, \ldots, 2n$ is no shortest path.
 % How do we assign the ranks to the new cherry nodes?
-% We could either create a path where $T'$ and $R'$ have distance equal to $d_{\rnni}(T|_{\{1, \ldots, n\}}, R|_{\{n+1, \ldots, 2n\}})$ and have a lot of additional rank swaps before and after that part OR
+% We could either create a path where $T'$ and $R'$ have distance equal to $d_{\rnni}(T|_{\{1, \ldots, n\}}, R|_{\{n+1, \ldots, 2n\}})$ and have a lot of additional rank moves before and after that part OR
 % Build $T'$ and $R'$ that contain all the cherries within the least number of moves possible (3n-3) and having a longer path between $T'$ and $R'$.
@@
 
+% With the knowledge that the set of caterpillar trees is convex in $\rnniu$, we can prove the Cluster Theorem restricted to caterpillar trees.
+
+\begin{conjecture}
+    Let $T_1$ and $T_2$ be caterpillar trees.
+    If $T_1$ and $T_2$ contain the same cluster $C$, then $C$ is present as cluster in every tree on every shortest path between $T_1$ and $T_2$.
+\todo{Cluster means non-trivial cluster -> def chapter?}
+\end{conjecture}
+
+% Following are some ideas that did not lead to major results so far:
 % idea for description of max distance caterpillar trees:
@@
 For caterpillar trees it is $(n-1)!$.
+The number of caterpillar trees with distance $\Delta(\rnniu)$ from a given caterpillar tree is $2^{n-2}$.
 
@@
 
+%If this is true, we can follow that radius = diameter
+
 \section{Partition lattice}

### 2019-06-25 paper.tex
minor on definitions

@@
 
-A \amph{rooted binary phylogenetic tree} a set $X$ is a pair $(t, \Phi)$ where $t$ is a rooted binary tree with leaf set $\mathcal{L}(t)$ and $\Phi:X \mapsto \mathcal{L}(T)$ a bijective mapping.
-\todo{do we need to define leaves and internal nodes, child, parent?}
-A \emph{ranked tree} is given by a pair of such a rooted binary phylogenetic tree $(t, \Phi)$ on $X = \{1, \ldots, n\}$ for $n \in \mathbb N$, and a rank function that maps the vertex set of $t$ onto the set $\{0,\ldots,n-1\}$ such that all leaves are assigned rank $0$ and each internal node has rank greater than its children.
-The number $n$ of taxa will remain fixed throughout the whole paper.
+A \amph{rooted binary phylogenetic tree} on a set $X$ is a pair $(t, \Phi)$ where $t$ is a rooted binary tree with leaf set $\mathcal{L}(t)$ and $\Phi:X \mapsto \mathcal{L}(T)$ a bijective mapping.
+\todo{do we need to define leaves and internal nodes, child, parent, subtree, cherry(defined below)?}
+From now on, we assume $X = \{1,\ldots,n\}$ for a fixed number $n \in \mathbb N$ of leaves.
+Within this paper we study \emph{ranked trees} that are determined by a pair consisting of a rooted binary phylogenetic tree $(t, \Phi)$ on $X = \{1, \ldots, n\}$ for $n \in \mathbb N$, and a rank function.
+A \emph{rank function} on $t$ maps the vertex set of $t$ onto the set $\{0,\ldots,n-1\}$ such that all leaves are assigned rank $0$ and each internal node has rank greater than its children.
 For simplicity of notation we use the term \emph{tree} instead instead of ranked tree.
@@
 An edge $e$ that has no leaf as child is an \emph{internal edge}.
-The set of taxa that are descending from an internal edge $e$ is called \emph{cluster}. \todo{OK?}
-In the literature, clusters often refer to sets of taxa descending from internal nodes.
-The set of clusters as defined above would then be recalled as non-trivial clusters.
-Given a subset $M$ of the set of taxa $X$, the \emph{most recent common ancestor} $mrca(M)$ of $M$ is the internal node with lowest rank that is ancestor of all taxa in $M$. \todo{OK?}
-For a single taxon $x$, $mrca(\{x\})$ is the parent of $x$ and will be denoted by $parent(x)$.
+Each internal edge defines a \emph{cluster} which is the set of taxa that are descending from the lower of both nodes bounding $e$. \todo{OK?}
+In the literature, clusters often refer to sets of taxa descending from internal nodes in which case clusters as defined above coincide with non-trivial clusters.
+Given a subset $S$ of the set of taxa $X$, the \emph{most recent common ancestor} $mrca(S)$ of $S$ is the internal node with lowest rank that is ancestor of all taxa in $S$. \todo{OK?}
+In particular, for a single taxon $x \in X$, $mrca(\{x\})$ is the parent of $x$ and will be denoted by $parent(x)$.
 Two trees are called \emph{isomorphic} if the underlying rooted phylogenetic trees are isomorphic and the isomorphism preserves ranks of the nodes.
 
-The set of ranked trees builds the vertex set of the \emph{$\rnni$-graph}.
-We introduce two types of operations (\emph{$\rnni$ moves}) connecting trees $T$ and $R$ in that graph by an edge:
+Now we are ready to define the tree space that is the subject of study of this paper, the $\rnni$ graph.
+The vertex set of the \emph{$\rnni$-graph} is given by the set of all ranked trees on $n$ taxa.
+We introduce two types of operations (\emph{$\rnni$ moves}) connecting trees $T$ and $R$ in this graph by an edge:
 The swap of the ranks of two internal nodes that have rank difference one is called \emph{rank move}.
 We say that there is an $\nni$ move between trees $T$ and $R$ if there are edges $e$ in $T$ and $f$ in $R$ that are bounded by internal nodes with rank difference one such that the graphs obtained by shrinking $e$ and $f$ to internal nodes are isomorphic.
-Trees that are connected by either a rank swap or an $\nni$ move are called \emph{$\rnni$ neighbours} or briefly neighbours.
+Trees that are connected by either a rank move or an $\nni$ move are called \emph{$\rnni$ neighbours} or briefly neighbours.
 Note that the previously defined $\rnni$ graph is a metric space, which is why we may call it \emph{$\rnni$ space}.
@@
 Therefore, further notations need to be introduced.
-If every internal node of a tree $T$ has at least one leaf as child, $T$ is called \emph{caterpillar tree}.
+If every internal node of a tree $T$ has at least one child that is a leaf, $T$ is called \emph{caterpillar tree}.
 A \emph{path} $p$ between trees $T_0$ and $T_m$ in $\rnni$ is a sequence $T_0, T_1, \ldots, T_m$ of trees such that there is an edge $\{T_i, T_{i+1}\}$ for all $i=0,\ldots,m-1$.
@@
 If $p$ is a path only consisting of caterpillar trees we call it \emph{caterpillar path}.
-For two caterpillar trees $T$ and $R$ a \emph{shortest caterpillar path} is a caterpillar path that is shortest among all caterpillar paths.
+For two caterpillar trees $T$ and $R$ a \emph{shortest caterpillar path} is a caterpillar path that is shortest among all caterpillar paths between $T$ and $R$.
 The length of such a shortest caterpillar path is called \emph{caterpillar distance} and denoted by $d_c(T,R)$.

### 2019-06-26 paper.tex
restructuring

@@
 
+%Introduce RNNI as ranked nearest neighbour interchange space, only ultrametric trees considered (extension could be mentioned in the end)
+
+%We use the exact algorithm to calculate all the distances for the examples given in this paper
+
 \section{Definitions}
@@
 A \emph{rank function} on $t$ maps the vertex set of $t$ onto the set $\{0,\ldots,n-1\}$ such that all leaves are assigned rank $0$ and each internal node has rank greater than its children.
+The rank of a node $v$ in a ranked tree will be denoted by $rank(v)$.
 For simplicity of notation we use the term \emph{tree} instead instead of ranked tree.
@@
 For a tree $T$ a subtree consisting of an internal node with both children being leaves is called \emph{cherry}.
-An edge $e$ that has no leaf as child is an \emph{internal edge}.
+An edge $e$ that has no leaf as child is an \emph{internal edge} and we say an edge $e=(u,v)$ has \emph{length} $l$ if ${|rank(v) - rank(u)| = l}$.
 Each internal edge defines a \emph{cluster} which is the set of taxa that are descending from the lower of both nodes bounding $e$. \todo{OK?}
@@
 In particular, for a single taxon $x \in X$, $mrca(\{x\})$ is the parent of $x$ and will be denoted by $parent(x)$.
+We write $mrca_T(S)$ to emphasise that we consider the most recent common ancestor of set $S$ in tree $T$.
 Two trees are called \emph{isomorphic} if the underlying rooted phylogenetic trees are isomorphic and the isomorphism preserves ranks of the nodes.
@@
 Trees that are connected by either a rank move or an $\nni$ move are called \emph{$\rnni$ neighbours} or briefly neighbours.
-Note that the previously defined $\rnni$ graph is a metric space, which is why we may call it \emph{$\rnni$ space}.
+Note that the previously defined $\rnni$ graph is a metric space, which is why we may call it \emph{$\rnni$ space} or just $\rnni$.
 
@@
 
-\section{Caterpillar Trees}
+\section{Geometry of $\rnni$}
+
+The following Algorithm~\ref{alg:find_path} which computes paths between pairs of trees in $\rnni$ is a central tool for our investigation of shortest paths in $\rnni$.
+
+\begin{algorithm}[H]
+\caption{FIND\_PATH($T,R$)}
+\label{alg:find_path}
+\begin{algorithmic}[1]
+	\STATE $\hat{T} := T$
+	\STATE $p := (T)$
+	\FOR {$i = 1, \dots, n-2$}
+		\STATE Let $C$ be the cluster associated with edge $(v,w)$ in $R$ with ${rank(v) = i < rank(w)}$
+		\WHILE {$rank(mrca_{\hat{T}}(C))>i$}
+			\STATE Update $\hat{T}$: Decrease $rank(mrca_{\hat{T}}(C))$ by an $\rnni$ move \label{alg:line:move_set_down}
+			\STATE $p = p+\hat{T}$
+		\ENDWHILE
+	\ENDFOR
+	\RETURN $p$
+\end{algorithmic}
+\end{algorithm}
+
+\todo{figure of an example?}
+
+One can easily prove that there is always exactly one $\rnni$ move on $T$ that decreases the rank of $mrca_T(C)$ as in line~\ref{alg:line:move_set_down} of Algorithm~\ref{alg:find_path}.
+It follows that Algorithm~\ref{alg:find_path} returns a path in $\rnni$ for any pair of trees $T,R$.
+Moreover, it gives an upper bound for the diameter of $\rnni$, which improves the one given in~\cite{Gavryushkin2017}:
+
+\begin{lemma}
+    It is $\Delta(\rnni) \leq \frac{(n-1)(n-2)}{2}$.
+\end{lemma}
+
+\begin{proof}
+    Let us consider the maximum length of a path given by Algorithm~\ref{alg:find_path}.
+    In worst case each of the $n-2$ clusters $C$ given by internal edges of $R$, represented by $i = 1, \dots, n-2$ in the algorithm, requires $n-1-i$ $\rnni$ moves for moving $mrca(C)$ down in $\hat{T}$.
+	Hence, the maximum length of a path computed by that algorithm for any pair of trees is less or equal to ${\sum\limits_{i = 1}^{n-2} i = \frac{(n-2)(n-1)}{2}}$.
+\end{proof}
+
+Lemma~\ref{lemma:distance_delete_taxon} is central for some of the following induction proofs. \todo{Is it correct and if it is, do we need to prove the statement, or is it rather obvious?}
+
+\todo{Change notations in this lemma+proof (less indices!!)}
+\begin{lemma}
+    Let $T_1$ and $T_2$ be two trees on $n+1$ taxa.
+    Let $T_1|_n$ and $T_2|_n$ be these trees restricted to $n$ taxa, i.e. the same taxon $n+1$ is deleted from both trees $T_1$ and $T_2$.
+    Then it is $d(T_1|_n, T_2|_n) \leq d(T_1,T_2) - |r_{T_1}(p_{n+1}) - r_{T_2}(p_{n+1})|$ where $r_{T_i}(p_j)$ determines the rank of the parent of taxon $j$ in tree $T_i$.
+    \label{lemma:distance_delete_taxon}
+\end{lemma}
+
+\begin{proof}
+    Induction on $k = d(T_1,T_2)$:
+
+    Basis($k = 1$):
+    If rank of $p_{n+1}$ changes between $T_1$ and $T_2$, it is $d(T_1|_n, T_2|_n) = 0$ and $|r_{T_1}(p_{n+1}) - r_{T_2}(p_{n+1})| = 1$ and therefore, $d(T_1|_n, T_2|_n) = 0 \leq 1 - 1 = d(T_1,T_2) - |r_{T_1}(p_{n+1}) - r_{T_2}(p_{n+1})|$.
+    If on the other hand the rank of $p_{n+1}$ does not change between $T_1$ and $T_2$, it is $d(T_1|_n, T_2|_n) = d(T_1,T_2) = 1$ and $|r_{T_1}(p_{n+1}) - r_{T_2}(p_{n+1})| = 1$ and therefore, $d(T_1|_n, T_2|_n) = 1 \leq 1 - 0 = d(T_1,T_2) - |r_{T_1}(p_{n+1}) - r_{T_2}(p_{n+1})|$.
+
+    Step($k \to k+1$):
+    Let $T_1$ and $T_2$ be trees on $n+1$ taxa with distance $d(T_1,T_2) = k$.
+    Let $T_2'$ be the last tree on a shortest path from $T_1$ to $T_2$, i.e. $d(T_1,T_2') = k$.
+    We can apply the induction hypothesis on $T_1$ and $T_2'$ and see that $d(T_1|_n, T_2'|_n) \leq d(T_1, T_2') - |r_{T_1}(p_{n+1}) - r_{T_2'}(p_{n+1})| := k - m$.
+    Now we differ two cases:
+    \begin{enumerate}
+        \item $p_{n+1}$ changes its rank between $T_2$ and $T_2'$ or is affected by the $\nni$ move is performed between $T_2$ and $T_2'$.
+
+        Then the trees $T_2|_n$ and $T_2'|_n$ are equal. \todo{Is this obivous?}
+        Therefore, it is $d(T_1|_n,T_2|_n) = d(T_1|_n, T_2'|_n) \leq k - m$.
+        It is either $|r_{T_1}(p_{n+1}) - r_{T_2}(p_{n+1})| = m+1$, $|r_{T_1}(p_{n+1}) - r_{T_2}(p_{n+1})| = m$, or $|r_{T_1}(p_{n+1}) - r_{T_2}(p_{n+1})| = m-1$.
+        In either case, we have $d(T_1|_n,T_2|_n) = d(T_1|_n, T_2'|_n) \leq k - m = (k+1) - (m+1) \leq d(T_1, T_2) - |r_{T_1}(p_{n+1}) - r_{T_2}(p_{n+1})|$.
+
+        \item $p_{n+1}$ has same rank in $T_2$ and $T_2'$.
+
+        In this case it is $d(T_1|_n,T_2|_n) \leq d(T_1|_n,T_2'|_n) + d(T_2'|_n, T_2|_n) \leq d(T_1|_n, T_2'|_n) + 1 \leq (k-m)+ 1$.
+        It also follows that $|r_{T_1}(p_{n+1}) - r_{T_2'}(p_{n+1})| = |r_{T_1}(p_{n+1}) - r_{T_2}(p_{n+1})|$ and therefore, we can conclude $d(T_1|_n,T_2|_n) \leq k-m+1 = k+1-m = d(T_1,T_2) -  |r_{T_1}(p_{n+1}) - r_{T_2}(p_{n+1})|$.
+    \end{enumerate}
+\end{proof}
+
+\subsection{The set of caterpillar trees}
+\label{section:caterpillar_convex}
+
+The set of caterpillar trees is of interested in $\rnni$ because shortest paths between such trees in $\rnni$ differ from those in classical $\nni$ space.
+Since $\nni$ moves in $\rnni$ are only allowed on edges of length one, the structure of shortest paths between caterpillar trees is different in $\nni$ and $\rnni$ as it is pointed out in Figure~\ref{fig:NNI_vs_RNNI}.
 
@@
 	\includegraphics[width=\textwidth]{NNI_vs_RNNI}
-	\caption{Paths between caterpillar trees $T_1$ and $T_2$: black -- $\nni$ path of length $5$; red -- $\rnni$ path of length $6$; blue -- $\rnni$ path of length $8$.}
-	\label{NNI_vs_RNNI}
+	\caption{Paths between caterpillar trees $T$ and $R$: black -- shortest path in classical $\nni$ (ignoring ranks) of length $5$; red -- shortest path in $\rnni$ of length $6$; blue -- path in $\rnni$ which is the natural extension of the shortest path in $\nni$ but has length $8$.}
+	\label{fig:NNI_vs_RNNI}
 \end{figure}
 
-\begin{conjecture}
-	The set of caterpillar trees is convex in $\rnni$ space.
-\end{conjecture}
+This observation already suggests that there is always a caterpillar path that is shortest path between two arbitrary caterpillar trees in $\rnni$.
+Indeed, our aim for this section is to prove Theorem~\ref{thm:caterpillar_convex} which states that the set of caterpillar trees is convex in $\rnni$.
+Note that this is not true for the $\nni$ space, which is clear from the example in Figure~\ref{fig:NNI_vs_RNNI}.
 
-This conjecture is computationally proven for up to $8$ taxa (maximum $\rnni$ graph we can compute).
-For a more restricted subset of trees in $\rnni$, namely caterpillar trees that have one cherry leaf in common, one can show convexity.
+% Shall we mention the data structure we use (list of sets) - this might be useful to see the connection to the partition lattice.
+One can see caterpillar trees as sequences of taxa, ordered non-decreasing according to the ranks of their parents.
+$\nni$ moves between caterpillar trees, which are the only possible $\rnni$ moves between caterpillar trees, coincide with exchanges of two neighboured elements in the sequence.
+In these sequences the first two taxa are somehow equivalent since they build the unique cherry.
+$\nni$ moves that change the cherry of a tree refer to exchanges of the first or the second element with the third element of the sequence while exchanges of the first two elements do not change the corresponding tree.
+With this observation we can provide an algorithm for computing shortest caterpillar paths by a simple modification of the commonly known \emph{Bubble Sort} algorithm~\cite{knuth1997}:
+
+Following the observations above, only the moves for the first three positions of a sequence with Bubble Sort need to be adapted for the computation of shortest caterpillar paths.
+Let us consider the first three elements $a,b,c$ which appear in this order in a sequence representing a caterpillar tree on a caterpillar path between trees $T$ and $R$.
+If the rank of $parent(c)$ is higher than the ranks of $parent(a)$ and $parent(b)$ in $R$, there are no swaps needed for these taxa.
+If just one of the two taxa $a$ and $b$ is on the other side of $c$ in $R$, this one swaps position with $c$ in the sequence.
+But if both $a$ and $b$ are on the other side of $c$ in $R$, the one that has the parent with highest rank of all these three taxa in $R$ swaps with $c$ at first.
+This makes sure that there is no additional swap of $a$ and $b$ needed on the path to $R$.
+
+Lemma~\ref{lemma:caterpillar_dist=diameter} regards caterpillar trees with distance $\frac{(n-1)(n-2)}{2}$ and will be needed for proving the convexity of the set of caterpillar trees in $\rnni$.
 
 \begin{lemma}
-	Let $T_1$ and $T_2$ be two caterpillar trees that both have a leaf labelled by $a$ as part of their cherry.
-	There is a shortest paths from $T_1$ to $T_2$ that is a caterpillar path.
-	\label{lemma:caterpillar_subset_convex}
+    Let $T$ and $T'$ be two caterpillar trees.
+    $T$ and $T'$ have distance $d(T,T') = \Delta(\rnniu) = \frac{(n-1)(n-2)}{2}$ if, and only if, they have caterpillar distance $d_c(T,T') = \Delta(\rnniu) = \frac{(n-1)(n-2)}{2}$.
+\todo{define $d_c$}
+    \label{lemma:caterpillar_dist=diameter}
 \end{lemma}
 
-\begin{figure}[h]
-	\centering
-	\includegraphics[width=0.3\textwidth]{label_internal_nodes}
-	\caption{The labelling of the caterpillar tree as introduced in the proof of Lemma~\ref{lemma:caterpillar_subset_convex} is unique due to its definition.
-	}
-	\label{label_internal_nodes}
-\end{figure}
+    % \todo{At the top of $T'$ it could also be $2,1$, or $t=1$}
+    %
+    % \begin{figure}[H]
+    %     \centering
+    %     \includegraphics[width=0.6\textwidth]{caterpillar_max_dist}
+    %     \caption{Caterpillar trees $T$ and $T'$ with distance $d(T,T') = \Delta(\rnniu)$}
+    %     \label{fig:caterpillar_max_distance}
+    % \end{figure}
 
 \begin{proof}
-	Let $T_1$ and $T_2$ be two caterpillar trees, both having a leaf labelled by $a$ as part of their cherry.
-	Let $p$ be a shortest path from $T_1$ to $T_2$.
-	We introduce a labelling of internal nodes of all trees on $p$ by first defining it for the caterpillar trees $T_1$ and $T_2$ and then explaining how the labelling is changed by $\rnni$ moves.
+    Let $d_c(T,T')$ denote the length of a shortest caterpillar path from $T$ to $T'$, that is a caterpillar path that is shortest among all caterpillar paths from $T$ to $T'$.
+    We need to prove that $d(T,T') = \Delta(\rnniu)$ if, and only if, $d_c(T,T') = \Delta(\rnniu)$.
 
-	In $T_1$ and $T_2$ the internal node of the cherry is labelled by the label of its child that is not $a$.
-	All other internal nodes shall have the same label as their leaf children.
-	From now on we will identify the leaf labelled by $x$ with $x$ and the internal node labelled by $x$ will be called $int_x$.
-	An example for a caterpillar tree with internal leaf labels is depicted in Figure~\ref{label_internal_nodes}.
-	In the following, changes of the internal labelling following $\rnni$ moves are defined:
+    Let $T$ and $T'$ have distance $d(T,T') = \Delta(\rnniu)$.
+    We can compute a  shortest caterpillar path from $T$ to $T'$ using Bubble Sort.
+    In each loop $k=1, \ldots, n-2$ of the algorithm there can be at most $n-1-k$ exchanges of taxa.
+    It follows that the maximum length of a path computed by Bubble Sort is $\sum\limits_{k=1}^{n-2} k = \frac{(n-1)(n-2)}{2}$.
+    Since there cannot be a caterpillar path shorter than $\frac{(n-1)(n-2)}{2}$ if it is $d(T,T') = \frac{(n-1)(n-2)}{2}$, it follows $d_c(T,T') = \Delta(\rnniu)$.
 
-	\textbf{Rank moves:}
-	Within rank moves the labels of internal nodes do not change.
-	For an example consider Figure~\ref{fig:rank_swap_internal_label}.
+    Let now $T$ and $T'$ have caterpillar distance $d_c(T,T') = \Delta(\rnniu)$.
+    We prove $d(T,T') = d_c(T,T')$ by induction on the number of taxa $n$ of $T$ and $T'$:
+    We assume without loss of generality that $T$ is the caterpillar tree $(\ldots ((1,2),3), \ldots, n)$.
+    Otherwise, we re-label $T$ and $T'$.
+    Let $T'$ be a caterpillar tree with distance $\Delta(\rnniu)$ to $T$.
 
-	\begin{figure}[H]
-		\centering
-		\includegraphics[width=0.8\textwidth]{rank_swap_internal_label}
-		\caption{The labelling of the internal nodes that swap ranks (red) do not change within the rank move.
-		}
-		\label{fig:rank_swap_internal_label}
-	\end{figure}
+    Induction basis: $n=3$\\
+    All trees have distance $1$ and are caterpillar trees.
 
-	\textbf{NNI move:}
-	Let there be an $\nni$ move on an edge $(int_x, int_y)$ where $rank(int_x) < rank(int_y)$ as in Figure~\ref{fig:nni_moves}.
+    Induction step: $n \to n+1$\\
+    Induction hypothesis: For trees $R$, $R'$ on $n$ taxa it is $d_c(R,R') = d(R, R') = \frac{(n-1)(n-2)}{2} = \Delta(\rnniu_n)$.\\
+    Let $T, T'$ be trees on $n+1$ taxa with caterpillar distance $\Delta(\rnniu_{n+1})$.
+    When applying the Bubble Sort algorithm to compute a path between $T'$ and $T$, each pair of taxa that is compared in that algorithm exchanges, since there must be $\Delta(\rnniu)=\frac{(n-1)(n-2)}{2}$ moves between $T$ and $T'$.
+    So we can conclude that taxon $n+1$ must be in the cherry of $T'$.
+    Let us consider the trees $T|_n$ to $T'|_n$ that result from $T$ and $T'$ by deleting taxon $n+1$.
+    When running Bubble Sort to compute a path from $T|_n$ to $T'|_n$ one can observe the same exchanges of taxa as on the Bubble Sort path from $T$ to $T'$, except for the $n-1$ ones that involve taxon $n+1$.
+    Hence, the caterpillar distance is $d_c(T|_n, T'|_n) = \frac{n(n-1)}{2} - (n-1)$ = $\frac{(n-1)(n-2)}{2}$.
+    So we can apply the induction hypothesis on $T|_n$ and $T'|_n$ and know that $d(T|_n,T'|_n) = \frac{(n-1)(n-2)}{2}$.
 
-	\begin{figure}[H]
-		\centering
-		\includegraphics[width=0.7\textwidth]{nni_moves_internal_label}
-		\caption{Two possible $\nni$ moves on edge $(int_x, int_y)$ where $x \in A, y \in B$.
-		}
-		\label{fig:nni_moves}
-	\end{figure}
+    With Lemma~\ref{lemma:distance_delete_taxon} we know that the distance between $T|_n$ and $T'|_n$ is at least $d(T|_n, T'|_n) \leq d(T,T') - |r_{T}(p_{n+1}) - r_{T'}(p_{n+1})|$.
+    We also know $|r_{T}(p_{n+1}) - r_{T'}(p_{n+1})| = n-1$
+    If we assume that there is a path from $T$ to $T'$ of length less than $\Delta(\rnniu_{n+1}) = \frac{n(n-1)}{2}$, it would follow that $d(T|_n, T'|_n) \leq d(T,T') - |r_{T}(p_{n+1}) - r_{T'}(p_{n+1})| < \frac{n(n-1)}{2} - (n-1) < \frac{(n-1)(n-2)}{2}$.
+    So it would be $d(T|_n, T'|_n) < \Delta(\rnniu_n)$.
+    Since this is a contradiction to $d(T|_n, T'|_n) = \Delta(\rnniu_n)$, it must be $d(T,T') = \Delta(\rnniu_{n+1}) = \frac{n(n-1)}{2} $.
 
-	The nodes incident to the edge after the $\nni$ move are labelled by $int_x$ and $int_y$ as well.
-	By default, the node that has lower rank than the other one is labelled by $int_x$, the other one by $int_y$.
-	But if $int_x$ is no ancestor of $x$ or $int_y$ is no ancestor of $y$, we swap the labels of the two new nodes, such that $rank(int_x) > rank(int_y)$.
-	An example of an $\nni$ move and the new labelling is provided in Figure~\ref{fig:nni_example_internal_label}.
-
-	\begin{figure}[H]
-		\centering
-		\includegraphics[width=0.8\textwidth]{nni_example_internal_label}
-		\caption{The first labelling after the $\nni$ move preserves the ranks of $int_4$ and $int_6$.
-		But then $int_6$ is not an ancestor of $6$ any more.
-		Therefore, $int_4$ and $int_6$ have to change their labels.
-		}
-		\label{fig:nni_example_internal_label}
-	\end{figure}
-
-	We proceed to show that $int_x$ and $int_y$ are ancestors of $x$ and $y$ after an $\nni$ move as described above, respectively.
-	If $x$ stays in the subtree rooted in the lower node of $e$, this does obviously hold.
-	We now turn to the case that $x$ is not in the subtree rooted in the lower node of $e$ after the $\nni$ move.
-	Then the labels of the two internal nodes change such that $rank(int_x) > rank(int_y)$.
-	This makes sure that $int_x$ is an ancestor of $x$.
-	Thus, it remains to prove that $int_y$ is ancestor of $y$.
-	In particular, we need to show that $y$ is not in the same subtree $A$ as $x$, whose root has the node labelled by $int_x$ as parent before and after the $\nni$ move.
-	Let us consider the tree before the $\nni$ move.
-	$A$ has $|A|$ leaves and $|A| - 1$ internal nodes.
-	These internal nodes are labelled by some leaves of $A$, because all internal nodes must be labelled by a descendant leaf.
-	We know that $x \in A$, and the parent of the root of $A$ is labelled by $x$.
-	All remaining $|A|-1$ leaf labels of $A$ appear as labels of the $|A|-1$ internal nodes of $A$.
-	From the fact that the internal node labelled by $y$ is not inside $A$ we can conclude that $y$ cannot be a leaf of $A$.
-	This proves that the labelling of the internal nodes is is well defined and allows a unique identification of internal nodes by their labels.
-
-	The task is now to prove that there is always a caterpillar path at least as short as the given path $p$.
-	Therefore, we use the following observation:
-	Within each $\rnni$ move the ranks of at most two internal nodes change by at most one each.
-	This is true for both rank moves and $\nni$ moves, according to our definition of the labelling:
-	Within a rank move two nodes exchange their ranks, while there are not necessarily changes of ranks when $\nni$ moves are performed.
-
-	On a caterpillar path the only possible moves are $\nni$ moves that change the ranks of two internal nodes by one each.
-	Therefore, we can construct a caterpillar path $r$ out of $p$:
-	every time two internal nodes $int_x$ and $int_y$ exchange ranks on $p$, they do so on the caterpillar path $r$ as well.
-	Since there are moves on $p$ that do not change the ranks of internal nodes, this caterpillar path has length $|r| \leq |p|$.
-	Hence, there is always a shortest path between $T_1$ and $T_2$ that is a caterpillar path.
 \end{proof}
 
-\section{Split Theorem}
+\begin{theorem}
+    The set of caterpillar trees is convex.
+    \label{thm:caterpillar_convex}
+\end{theorem}
+
+\begin{proof}
+    We prove the Theorem by backwards induction on the length $l$ of a shortest caterpillar path between two caterpillar trees $T$ and $T'$.
+
+    Induction basis: $l = \Delta(\rnniu) = \frac{(n-1)(n-2)}{2}$\\
+    With Lemma~\ref{lemma:caterpillar_dist=diameter} one can see that for each pair of caterpillar trees $T$, $T'$ with caterpillar distance $d_c(T,T') = \Delta(\rnniu)$, it is $d(T,T') = d_c(T,T')$, which means that there is a shortest path between $T$ and $T'$ that is a caterpillar path.
+
+    Induction step: $l+1 \to l$\\
+    Induction hypothesis: For each pair of caterpillar trees with caterpillar distance $l+1$ there is a shortest path between these trees that only consists of caterpillar trees, which means that the distance between these trees is $l+1$.
+    Let $T$ and $T'$ be caterpillar trees with caterpillar distance $l$.
+    Since $l < \Delta(\rnniu) = \frac{(n-1)(n-2)}{2}$, there is one loop $k$ in the Bubble Sort algorithm from $T'$ to $T$ where there are less than $n-2-k$ swaps of neighboured taxa.
+    This means that there is a taxon $a$ that has parent with rank $n-2-k$ in $T$ and that does not need to move all the way up starting at the cherry in loop $k$.
+    Therefore, there must be a taxon $b$ which has parent with rank one less than the parent of $a$ in the tree at the beginning of loop number $k$.
+    If we assume that loop $k$ is the first loop where a single taxon is not moved by $n-k-2$ $\rnni$ moves, it follows that $a$ and $b$ must be neighboured taxa in $T'$.
+    Let $T''$ be the caterpillar tree resulting from $T'$ by exchanging $a$ and $b$.
+    It follows that $T'$ has caterpillar distance $d_c(T,T'') = d_c(T,T') + 1 = l+1$ to $T$ and $d(T',T'') = 1$.
+    We can apply the induction hypothesis on $T$ and $T''$ and know that $d(T,T'') = l+1$.
+    If it was $d(T,T') < l$, it would hold $d(T,T'') \leq d(T,T') + d(T',T'') < l + 1$ which contradicts the induction hypothesis.
+    We can conclude that $d(T,T') = l$.
+
+\end{proof}
+
+\subsection{Diameter}
+
+In~\cite{Gavryushkin2017} upper and lower bound for the diameter of $\rnni$ are given.
+Following the results of section~\ref{section:caterpillar_convex} one can directly improve this result by giving the exact diameter of the $\rnni$ graph.
+
+\begin{lemma}
+	There are two caterpillar trees in $\rnniu$ with distance $\frac{(n-1)(n-2))}{2}$.
+	\label{conj:caterpillar_diameter}
+\end{lemma}
+
+\begin{proof}
+	Let $T = (( \dots (n,1)2)\dots)n-1)$ and $R = (( \dots (n,n-1)n-2)\dots)1)$.
+    According to Theorem~\ref{thm:caterpillar_convex}, there is a shortest path between $T$ and $R$ that is a caterpillar path.
+    We can compute a shortest caterpillar path by using the modified version of Bubble Sort as explained above. \todo{Do we want to name this algorithm?}
+    When running this algorithm on $T$ and $R$, taxon $i$ swaps $n-1-i$ times with its right neighbour.
+    Therefore, the length of a shortest caterpillar path is ${\sum\limits_{i=1}^{n-1}(n-1-i) = \sum\limits_{i=0}^{n-2}i = \frac{(n-1)(n-2)}{2}}$.
+\end{proof}
+
+\begin{corollary}
+    It is $\Delta(\rnniu) = \frac{(n-1)(n-2)}{2}$.
+\end{corollary}
+
+%Do we need this?
+
+% \begin{lemma}
+% 	Let $T_1$ and $T_2$ be two caterpillar trees that both have a leaf labelled by $a$ as part of their cherry.
+% 	There is a shortest paths from $T_1$ to $T_2$ that is a caterpillar path.
+% 	\label{lemma:caterpillar_subset_convex}
+% \end{lemma}
+%
+% \begin{figure}[h]
+% 	\centering
+% 	\includegraphics[width=0.3\textwidth]{label_internal_nodes}
+% 	\caption{The labelling of the caterpillar tree as introduced in the proof of Lemma~\ref{lemma:caterpillar_subset_convex} is unique due to its definition.
+% 	}
+% 	\label{label_internal_nodes}
+% \end{figure}
+%
+% \begin{proof}
+% 	Let $T_1$ and $T_2$ be two caterpillar trees, both having a leaf labelled by $a$ as part of their cherry.
+% 	Let $p$ be a shortest path from $T_1$ to $T_2$.
+% 	We introduce a labelling of internal nodes of all trees on $p$ by first defining it for the caterpillar trees $T_1$ and $T_2$ and then explaining how the labelling is changed by $\rnni$ moves.
+%
+% 	In $T_1$ and $T_2$ the internal node of the cherry is labelled by the label of its child that is not $a$.
+% 	All other internal nodes shall have the same label as their leaf children.
+% 	From now on we will identify the leaf labelled by $x$ with $x$ and the internal node labelled by $x$ will be called $int_x$.
+% 	An example for a caterpillar tree with internal leaf labels is depicted in Figure~\ref{label_internal_nodes}.
+% 	In the following, changes of the internal labelling following $\rnni$ moves are defined:
+%
+% 	\textbf{Rank moves:}
+% 	Within rank moves the labels of internal nodes do not change.
+% 	For an example consider Figure~\ref{fig:rank_swap_internal_label}.
+%
+% 	\begin{figure}[H]
+% 		\centering
+% 		\includegraphics[width=0.8\textwidth]{rank_swap_internal_label}
+% 		\caption{The labelling of the internal nodes that swap ranks (red) do not change within the rank move.
+% 		}
+% 		\label{fig:rank_swap_internal_label}
+% 	\end{figure}
+%
+% 	\textbf{NNI move:}
+% 	Let there be an $\nni$ move on an edge $(int_x, int_y)$ where $rank(int_x) < rank(int_y)$ as in Figure~\ref{fig:nni_moves}.
+%
+% 	\begin{figure}[H]
+% 		\centering
+% 		\includegraphics[width=0.7\textwidth]{nni_moves_internal_label}
+% 		\caption{Two possible $\nni$ moves on edge $(int_x, int_y)$ where $x \in A, y \in B$.
+% 		}
+% 		\label{fig:nni_moves}
+% 	\end{figure}
+%
+% 	The nodes incident to the edge after the $\nni$ move are labelled by $int_x$ and $int_y$ as well.
+% 	By default, the node that has lower rank than the other one is labelled by $int_x$, the other one by $int_y$.
+% 	But if $int_x$ is no ancestor of $x$ or $int_y$ is no ancestor of $y$, we swap the labels of the two new nodes, such that $rank(int_x) > rank(int_y)$.
+% 	An example of an $\nni$ move and the new labelling is provided in Figure~\ref{fig:nni_example_internal_label}.
+%
+% 	\begin{figure}[H]
+% 		\centering
+% 		\includegraphics[width=0.8\textwidth]{nni_example_internal_label}
+% 		\caption{The first labelling after the $\nni$ move preserves the ranks of $int_4$ and $int_6$.
+% 		But then $int_6$ is not an ancestor of $6$ any more.
+% 		Therefore, $int_4$ and $int_6$ have to change their labels.
+% 		}
+% 		\label{fig:nni_example_internal_label}
+% 	\end{figure}
+%
+% 	We proceed to show that $int_x$ and $int_y$ are ancestors of $x$ and $y$ after an $\nni$ move as described above, respectively.
+% 	If $x$ stays in the subtree rooted in the lower node of $e$, this does obviously hold.
+% 	We now turn to the case that $x$ is not in the subtree rooted in the lower node of $e$ after the $\nni$ move.
+% 	Then the labels of the two internal nodes change such that $rank(int_x) > rank(int_y)$.
+% 	This makes sure that $int_x$ is an ancestor of $x$.
+% 	Thus, it remains to prove that $int_y$ is ancestor of $y$.
+% 	In particular, we need to show that $y$ is not in the same subtree $A$ as $x$, whose root has the node labelled by $int_x$ as parent before and after the $\nni$ move.
+% 	Let us consider the tree before the $\nni$ move.
+% 	$A$ has $|A|$ leaves and $|A| - 1$ internal nodes.
+% 	These internal nodes are labelled by some leaves of $A$, because all internal nodes must be labelled by a descendant leaf.
+% 	We know that $x \in A$, and the parent of the root of $A$ is labelled by $x$.
+% 	All remaining $|A|-1$ leaf labels of $A$ appear as labels of the $|A|-1$ internal nodes of $A$.
+% 	From the fact that the internal node labelled by $y$ is not inside $A$ we can conclude that $y$ cannot be a leaf of $A$.
+% 	This proves that the labelling of the internal nodes is is well defined and allows a unique identification of internal nodes by their labels.
+%
+% 	The task is now to prove that there is always a caterpillar path at least as short as the given path $p$.
+% 	Therefore, we use the following observation:
+% 	Within each $\rnni$ move the ranks of at most two internal nodes change by at most one each.
+% 	This is true for both rank moves and $\nni$ moves, according to our definition of the labelling:
+% 	Within a rank move two nodes exchange their ranks, while there are not necessarily changes of ranks when $\nni$ moves are performed.
+%
+% 	On a caterpillar path the only possible moves are $\nni$ moves that change the ranks of two internal nodes by one each.
+% 	Therefore, we can construct a caterpillar path $r$ out of $p$:
+% 	every time two internal nodes $int_x$ and $int_y$ exchange ranks on $p$, they do so on the caterpillar path $r$ as well.
+% 	Since there are moves on $p$ that do not change the ranks of internal nodes, this caterpillar path has length $|r| \leq |p|$.
+% 	Hence, there is always a shortest path between $T_1$ and $T_2$ that is a caterpillar path.
+% \end{proof}
+
+\subsection{Split Theorem}
 
@@
 
-\section{Algorithms}
+\section{Ideas}
+\todo{Which ideas do we want to keep? Shall this become an 'open problems/future work' section?}
 
-We are now going to introduce different algorithms that computes paths in $\rnni$.
-\cite{Gavryushkin2017} gave an algorithm for computing the whole $\rnni$ space.
-We implemented this algorithm and are able to use it on trees with up to $8$ taxa.
-Therefore, we can compare the paths given by the algorithms presented in this chapter with the exact shortest path for trees with up to $8$ taxa.
-
-% Shall we mention the data structure we use (list of sets) - this might be useful to see the connection to the partition lattice.
-
-A modified version of the classic Bubble Sort \todo{cite} algorithm returns a caterpillar path that is shortest among all caterpillar paths between given caterpillar trees $T_1, T_2$.
-Let $T_1 = (\ldots((1,2)3) \ldots )n), T_2 = (\ldots((\sigma_1,\sigma_2)\sigma_3) \ldots )\sigma_n)$ be two caterpillar trees.
-They can be seen as sequences $(1,2,\ldots,n)$ and $(\sigma_1, \sigma_2, \ldots, \sigma_n)$ that need to be sorted.
-$\nni$ moves between caterpillar trees correspond to exchanging two neighboured elements in a sequence.
-The only exception of this are $\nni$ moves that change the cherry of a tree.
-The corresponding moves on sequences are exchanged of either the second and third element or the first and third element.
-On the other hand, there is no difference between two the trees corresponding to two sequences where only the first two taxa are exchanged.
-Since all pairs of taxa whose order in $T_1$ is different from the one in $T_2$ need to be exchanged by $\nni$ moves on a caterpillar path, Bubble Sort will give a shortest path among all caterpillar paths.
-Just the moves for the first three positions of the sequence must be adapted.
-Let us consider the first three elements $(a,b,c,\ldots)$ of a sequence representing a caterpillar tree on a caterpillar path between $T_1$ and $T_2$
-If the order of these elements is the same in $T_2$, there are no swaps needed for these taxa.
-If just one of the two taxa $a$ and $b$ is on the other side of $c$ in $T_2$, this one will change wih $c$.
-But if both $a$ and $b$ are on the other side of $c$ in $T_2$, the one which is the righmost of all three taxa in $T_2$ swaps with $c$ at first.
-This makes sure that there is no additional swap of $a$ and $b$ needed on the path to $T_2$.
-\todo{This paragraph is just a draft. Is it needed at all?}
-
-Following algorithm computes paths between two trees in $\rnni$:
-
-\begin{algorithm}[H]
-\caption{FIND\_PATH($T_1,T_2$)}
-\label{alg:find_path}
-\begin{algorithmic}[1]
-	\STATE $T := T_1$
-	\STATE $p := (T_1)$
-	\FOR {$r = 1, \dots, n-2$}
-		\STATE Let $S$ be the cluster associated with the node $v$ in $T_2$ with $rank(v) = r$
-		\WHILE {$rank(MRCA_T(S))>r$}
-			\STATE Update $T$: Decrease $rank(MRCA_T(S))$ by an $\rnni$ move \label{alg:line:move_set_down}
-			\STATE $p = p+T$
-		\ENDWHILE
-	\ENDFOR
-	\RETURN p
-\end{algorithmic}
-\end{algorithm}
-
-One can easily prove that there is always exactly one possible $\rnni$ move that decreases the rank of $MRCA_T(S)$ as in line~\ref{alg:line:move_set_down} of Algorithm~\ref{alg:find_path}.
-
-The proof of Theorem 7 in \cite{Gavryushkin2017} gives an algorithm to compute paths between trees in $\rnni$. %Do we want to mention that algorithm here?
-
-% Compare the two algorithms by sampling trees and check which algorithm gives shorter paths.
-
-% Also mention the exact algorithm for computing the whole RNNI space for up to 7 (?) taxa and check that Algorithm find_path gives correct distances there.
-% Maybe it's worth trying the other algorithm in these trees as well.
-
-Algorithm~\ref{alg:find_path} provides an upper bound for the diameter of $\rnniu$ space.
-It improves the bound $n^2 - 3n - \frac{5}{8}$ given in \cite{Gavryushkin2017}.
-
-\begin{theorem}
-	It is $\Delta(\rnniu) \leq \frac{1}{2}n^2-\frac{3}{2}n+1$
-\end{theorem}
-
-\begin{proof}
-	This result directly follows from Algorithm~\ref{alg:find_path}.
-	In the worst case each $r = 1, \dots, n-2$ requires $n-1-r$ $\rnni$ moves for moving the corresponding $MRCA$ down.
-	Hence, there is for each pair of trees a path of length less or equal to $\sum\limits_{i = 1}^{n-2} i = \frac{(n-2)(n-1)}{2} = \frac{1}{2}n^2-\frac{3}{2}n+1$.
-\end{proof}
-
-\begin{lemma}
-	There are two caterpillar trees in $\rnniu$ with distance $\frac{1}{2}n^2-\frac{3}{2}n+1$.
-	\label{conj:caterpillar_diameter}
-\end{lemma}
-
-\begin{proof}
-    For this proof we will give two caterpillar trees $T_1$ and $T_2$ on $n$ taxa with $d_{RNNI}(T_1,T_2) = \frac{1}{2}n^2-\frac{3}{2}n+1$.
-    \todo{$d$ or $d_{RNNI}$?}
-	Let $T_1 = (( \dots (n,1)2)\dots)n-1)$ and $T_2 = (( \dots (n,n-1)n-2)\dots)1)$.
-    According to Lemma~\ref{lemma:caterpillar_subset_convex}, there is a shortest path between $T_1$ and $T_2$ that only consists of caterpillar trees.
-    We can compute such a shortest caterpillar path by using a modified version of Bubble Sort.
-    When running this algorithm on $T_1$ and $T_2$ there are $n-1-i$ swaps of taxon $i$ with its right neighbour.
-    Therefore, the total number of these swaps, and therefore $d_{RNNI}(T_1,T_2)$, is $\sum\limits_{i=1}^{n-1}(n-1-i) = \sum\limits_{i=0}^{n-2}i = \frac{1}{2}n^2-\frac{3}{2}n+1$
-\end{proof}
-
-\begin{corollary}
-It is $\Delta(\rnniu) = \frac{(n-1)(n-2)}{2} = \frac{1}{2}n^2-\frac{3}{2}n+1$.
-\end{corollary}
-
-%This lemma might be central  for induction proofs in RNNI space
-%Is it correct and if it is, do we need to prove the statement, or is it rather obvious?
-%The lemma implies another algorithm to compute distances (it would not give shortest paths explicitly, unlike alg:find_path)
-\begin{lemma}
-    Let $T_1$ and $T_2$ be two trees on $n+1$ taxa.
-    Let $T_1|_n$ and $T_2|_n$ be these trees restricted to $n$ taxa, i.e. the same taxon $n+1$ is deleted from both trees $T_1$ and $T_2$.
-    Then it is $d(T_1|_n, T_2|_n) \leq d(T_1,T_2) - |r_{T_1}(p_{n+1}) - r_{T_2}(p_{n+1})|$ where $r_{T_i}(p_j)$ determines the rank of the parent of taxon $j$ in tree $T_i$.
-    \label{lemma:distance_delete_taxon}
-\end{lemma}
-
-\begin{proof}
-    Induction on $k = d(T_1,T_2)$:
-
-    Basis($k = 1$):
-    If rank of $p_{n+1}$ changes between $T_1$ and $T_2$, it is $d(T_1|_n, T_2|_n) = 0$ and $|r_{T_1}(p_{n+1}) - r_{T_2}(p_{n+1})| = 1$ and therefore, $d(T_1|_n, T_2|_n) = 0 \leq 1 - 1 = d(T_1,T_2) - |r_{T_1}(p_{n+1}) - r_{T_2}(p_{n+1})|$.
-    If on the other hand the rank of $p_{n+1}$ does not change between $T_1$ and $T_2$, it is $d(T_1|_n, T_2|_n) = d(T_1,T_2) = 1$ and $|r_{T_1}(p_{n+1}) - r_{T_2}(p_{n+1})| = 1$ and therefore, $d(T_1|_n, T_2|_n) = 1 \leq 1 - 0 = d(T_1,T_2) - |r_{T_1}(p_{n+1}) - r_{T_2}(p_{n+1})|$.
-
-    Step($k \to k+1$):
-    Let $T_1$ and $T_2$ be trees on $n+1$ taxa with distance $d(T_1,T_2) = k$.
-    Let $T_2'$ be the last tree on a shortest path from $T_1$ to $T_2$, i.e. $d(T_1,T_2') = k$.
-    We can apply the induction hypothesis on $T_1$ and $T_2'$ and see that $d(T_1|_n, T_2'|_n) \leq d(T_1, T_2') - |r_{T_1}(p_{n+1}) - r_{T_2'}(p_{n+1})| := k - m$.
-    Now we differ two cases:
-    \begin{enumerate}
-        \item $p_{n+1}$ changes its rank between $T_2$ and $T_2'$ or is affected by the $\nni$ move is performed between $T_2$ and $T_2'$.
-
-        Then the trees $T_2|_n$ and $T_2'|_n$ are equal. \todo{Is this obivous?}
-        Therefore, it is $d(T_1|_n,T_2|_n) = d(T_1|_n, T_2'|_n) \leq k - m$.
-        It is either $|r_{T_1}(p_{n+1}) - r_{T_2}(p_{n+1})| = m+1$, $|r_{T_1}(p_{n+1}) - r_{T_2}(p_{n+1})| = m$, or $|r_{T_1}(p_{n+1}) - r_{T_2}(p_{n+1})| = m-1$.
-        In either case, we have $d(T_1|_n,T_2|_n) = d(T_1|_n, T_2'|_n) \leq k - m = (k+1) - (m+1) \leq d(T_1, T_2) - |r_{T_1}(p_{n+1}) - r_{T_2}(p_{n+1})|$.
-
-        \item $p_{n+1}$ has same rank in $T_2$ and $T_2'$.
-
-        In this case it is $d(T_1|_n,T_2|_n) \leq d(T_1|_n,T_2'|_n) + d(T_2'|_n, T_2|_n) \leq d(T_1|_n, T_2'|_n) + 1 \leq (k-m)+ 1$.
-        It also follows that $|r_{T_1}(p_{n+1}) - r_{T_2'}(p_{n+1})| = |r_{T_1}(p_{n+1}) - r_{T_2}(p_{n+1})|$ and therefore, we can conclude $d(T_1|_n,T_2|_n) \leq k-m+1 = k+1-m = d(T_1,T_2) -  |r_{T_1}(p_{n+1}) - r_{T_2}(p_{n+1})|$.
-    \end{enumerate}
-\end{proof}
-
-\begin{lemma}
-    Let $T$ and $T'$ be two caterpillar trees.
-    $T$ and $T'$ have distance $d(T,T') = \Delta(\rnniu) = \frac{(n-1)(n-2)}{2}$ if, and only if, they have caterpillar distance $d_c(T,T') = \Delta(\rnniu) = \frac{(n-1)(n-2)}{2}$.
-\todo{define $d_c$}
-    \label{lemma:caterpillar_dist=diameter}
-\end{lemma}
-
-    % \todo{At the top of $T'$ it could also be $2,1$, or $t=1$}
-    %
-    % \begin{figure}[H]
-    %     \centering
-    %     \includegraphics[width=0.6\textwidth]{caterpillar_max_dist}
-    %     \caption{Caterpillar trees $T$ and $T'$ with distance $d(T,T') = \Delta(\rnniu)$}
-    %     \label{fig:caterpillar_max_distance}
-    % \end{figure}
-
-\begin{proof}
-    Let $d_c(T,T')$ denote the length of a shortest caterpillar path from $T$ to $T'$, that is a caterpillar path that is shortest among all caterpillar paths from $T$ to $T'$.
-    We need to prove that $d(T,T') = \Delta(\rnniu)$ if, and only if, $d_c(T,T') = \Delta(\rnniu)$.
-
-    Let $T$ and $T'$ have distance $d(T,T') = \Delta(\rnniu)$.
-    We can compute a  shortest caterpillar path from $T$ to $T'$ using Bubble Sort.
-    In each loop $k=1, \ldots, n-2$ of the algorithm there can be at most $n-1-k$ exchanges of taxa.
-    It follows that the maximum length of a path computed by Bubble Sort is $\sum\limits_{k=1}^{n-2} k = \frac{(n-1)(n-2)}{2}$.
-    Since there cannot be a caterpillar path shorter than $\frac{(n-1)(n-2)}{2}$ if it is $d(T,T') = \frac{(n-1)(n-2)}{2}$, it follows $d_c(T,T') = \Delta(\rnniu)$.
-
-    Let now $T$ and $T'$ have caterpillar distance $d_c(T,T') = \Delta(\rnniu)$.
-    We prove $d(T,T') = d_c(T,T')$ by induction on the number of taxa $n$ of $T$ and $T'$:
-    We assume without loss of generality that $T$ is the caterpillar tree $(\ldots ((1,2),3), \ldots, n)$.
-    Otherwise, we re-label $T$ and $T'$.
-    Let $T'$ be a caterpillar tree with distance $\Delta(\rnniu)$ to $T$.
-
-    Induction basis: $n=3$\\
-    All trees have distance $1$ and are caterpillar trees.
-
-    Induction step: $n \to n+1$\\
-    Induction hypothesis: For trees $R$, $R'$ on $n$ taxa it is $d_c(R,R') = d(R, R') = \frac{(n-1)(n-2)}{2} = \Delta(\rnniu_n)$.\\
-    Let $T, T'$ be trees on $n+1$ taxa with caterpillar distance $\Delta(\rnniu_{n+1})$.
-    When applying the Bubble Sort algorithm to compute a path between $T'$ and $T$, each pair of taxa that is compared in that algorithm exchanges, since there must be $\Delta(\rnniu)=\frac{(n-1)(n-2)}{2}$ moves between $T$ and $T'$.
-    So we can conclude that taxon $n+1$ must be in the cherry of $T'$.
-    Let us consider the trees $T|_n$ to $T'|_n$ that result from $T$ and $T'$ by deleting taxon $n+1$.
-    When running Bubble Sort to compute a path from $T|_n$ to $T'|_n$ one can observe the same exchanges of taxa as on the Bubble Sort path from $T$ to $T'$, except for the $n-1$ ones that involve taxon $n+1$.
-    Hence, the caterpillar distance is $d_c(T|_n, T'|_n) = \frac{n(n-1)}{2} - (n-1)$ = $\frac{(n-1)(n-2)}{2}$.
-    So we can apply the induction hypothesis on $T|_n$ and $T'|_n$ and know that $d(T|_n,T'|_n) = \frac{(n-1)(n-2)}{2}$.
-
-    With Lemma~\ref{lemma:distance_delete_taxon} we know that the distance between $T|_n$ and $T'|_n$ is at least $d(T|_n, T'|_n) \leq d(T,T') - |r_{T}(p_{n+1}) - r_{T'}(p_{n+1})|$.
-    We also know $|r_{T}(p_{n+1}) - r_{T'}(p_{n+1})| = n-1$
-    If we assume that there is a path from $T$ to $T'$ of length less than $\Delta(\rnniu_{n+1}) = \frac{n(n-1)}{2}$, it would follow that $d(T|_n, T'|_n) \leq d(T,T') - |r_{T}(p_{n+1}) - r_{T'}(p_{n+1})| < \frac{n(n-1)}{2} - (n-1) < \frac{(n-1)(n-2)}{2}$.
-    So it would be $d(T|_n, T'|_n) < \Delta(\rnniu_n)$.
-    Since this is a contradiction to $d(T|_n, T'|_n) = \Delta(\rnniu_n)$, it must be $d(T,T') = \Delta(\rnniu_{n+1}) = \frac{n(n-1)}{2} $.
-
-\end{proof}
-
-\begin{theorem}
-    The set of caterpillar trees is convex.
-\end{theorem}
-
-\begin{proof}
-    We prove the Theorem by backwards induction on the length $l$ of a shortest caterpillar path between two caterpillar trees $T$ and $T'$.
-
-    Induction basis: $l = \Delta(\rnniu) = \frac{(n-1)(n-2)}{2}$\\
-    With Lemma~\ref{lemma:caterpillar_dist=diameter} one can see that for each pair of caterpillar trees $T$, $T'$ with caterpillar distance $d_c(T,T') = \Delta(\rnniu)$, it is $d(T,T') = d_c(T,T')$, which means that there is a shortest path between $T$ and $T'$ that is a caterpillar path.
-
-    Induction step: $l+1 \to l$\\
-    Induction hypothesis: For each pair of caterpillar trees with caterpillar distance $l+1$ there is a shortest path between these trees that only consists of caterpillar trees, which means that the distance between these trees is $l+1$.
-    Let $T$ and $T'$ be caterpillar trees with caterpillar distance $l$.
-    Since $l < \Delta(\rnniu) = \frac{(n-1)(n-2)}{2}$, there is one loop $k$ in the Bubble Sort algorithm from $T'$ to $T$ where there are less than $n-2-k$ swaps of neighboured taxa.
-    This means that there is a taxon $a$ that has parent with rank $n-2-k$ in $T$ and that does not need to move all the way up starting at the cherry in loop $k$.
-    Therefore, there must be a taxon $b$ which has parent with rank one less than the parent of $a$ in the tree at the beginning of loop number $k$.
-    If we assume that loop $k$ is the first loop where a single taxon is not moved by $n-k-2$ $\rnni$ moves, it follows that $a$ and $b$ must be neighboured taxa in $T'$.
-    Let $T''$ be the caterpillar tree resulting from $T'$ by exchanging $a$ and $b$.
-    It follows that $T'$ has caterpillar distance $d_c(T,T'') = d_c(T,T') + 1 = l+1$ to $T$ and $d(T',T'') = 1$.
-    We can apply the induction hypothesis on $T$ and $T''$ and know that $d(T,T'') = l+1$.
-    If it was $d(T,T') < l$, it would hold $d(T,T'') \leq d(T,T') + d(T',T'') < l + 1$ which contradicts the induction hypothesis.
-    We can conclude that $d(T,T') = l$.
-
-\end{proof}
-
-% With the knowledge that the set of caterpillar trees is convex in $\rnniu$, we can prove the Cluster Theorem restricted to caterpillar trees.
-
-\begin{conjecture}
-    Let $T_1$ and $T_2$ be caterpillar trees.
-    If $T_1$ and $T_2$ contain the same cluster $C$, then $C$ is present as cluster in every tree on every shortest path between $T_1$ and $T_2$.
-\todo{Cluster means non-trivial cluster -> def chapter?}
-\end{conjecture}
-
-% Following are some ideas that did not lead to major results so far:
+Following are some ideas that did not lead to major results so far:
 % idea for description of max distance caterpillar trees:
@@
 
-\section{Partition lattice}
+\subsection{Partition lattice}
 

### 2019-06-26 paper.tex
shorten proof: length of path if taxon deleted from all trees

@@
 A \emph{rank function} on $t$ maps the vertex set of $t$ onto the set $\{0,\ldots,n-1\}$ such that all leaves are assigned rank $0$ and each internal node has rank greater than its children.
-The rank of a node $v$ in a ranked tree will be denoted by $rank(v)$.
+The rank of a node $v$ in a ranked tree will be denoted by $rank(v)$ or $rank_T(v)$ if it is not obvious that we consider tree $T$.
 For simplicity of notation we use the term \emph{tree} instead instead of ranked tree.
@@
 For a tree $T$ a subtree consisting of an internal node with both children being leaves is called \emph{cherry}.
-An edge $e$ that has no leaf as child is an \emph{internal edge} and we say an edge $e=(u,v)$ has \emph{length} $l$ if ${|rank(v) - rank(u)| = l}$.
+An edge $e$ that has no leaf as child is an \emph{internal edge} and we say an edge $e=(u,v)$ has \emph{length} $l$ if $|rank(v) - rank(u)| = l$.
 Each internal edge defines a \emph{cluster} which is the set of taxa that are descending from the lower of both nodes bounding $e$. \todo{OK?}
@@
 If every internal node of a tree $T$ has at least one child that is a leaf, $T$ is called \emph{caterpillar tree}.
-A \emph{path} $p$ between trees $T_0$ and $T_m$ in $\rnni$ is a sequence $T_0, T_1, \ldots, T_m$ of trees such that there is an edge $\{T_i, T_{i+1}\}$ for all $i=0,\ldots,m-1$.
+A \emph{path} $p$ between trees $T_0$ and $T_m$ in $\rnni$ is a sequence $T_0, T, \ldots, T_m$ of trees such that there is an edge $\{T_i, T_{i+1}\}$ for all $i=0,\ldots,m-1$.
 For such a path $p$ we say the \emph{length} $|p|$ of $p$ is $m$.
@@
 	\FOR {$i = 1, \dots, n-2$}
-		\STATE Let $C$ be the cluster associated with edge $(v,w)$ in $R$ with ${rank(v) = i < rank(w)}$
+		\STATE Let $C$ be the cluster associated with edge $(v,w)$ in $R$ with $rank(v) = i < rank(w)$
 		\WHILE {$rank(mrca_{\hat{T}}(C))>i$}
@@
     In worst case each of the $n-2$ clusters $C$ given by internal edges of $R$, represented by $i = 1, \dots, n-2$ in the algorithm, requires $n-1-i$ $\rnni$ moves for moving $mrca(C)$ down in $\hat{T}$.
-	Hence, the maximum length of a path computed by that algorithm for any pair of trees is less or equal to ${\sum\limits_{i = 1}^{n-2} i = \frac{(n-2)(n-1)}{2}}$.
+	Hence, the maximum length of a path computed by that algorithm for any pair of trees is less or equal to $\sum\limits_{i = 1}^{n-2} i = \frac{(n-2)(n-1)}{2}$.
 \end{proof}
 
-Lemma~\ref{lemma:distance_delete_taxon} is central for some of the following induction proofs. \todo{Is it correct and if it is, do we need to prove the statement, or is it rather obvious?}
+Lemma~\ref{lemma:distance_delete_taxon} is central for some of the following proofs using induction.
 
-\todo{Change notations in this lemma+proof (less indices!!)}
 \begin{lemma}
-    Let $T_1$ and $T_2$ be two trees on $n+1$ taxa.
-    Let $T_1|_n$ and $T_2|_n$ be these trees restricted to $n$ taxa, i.e. the same taxon $n+1$ is deleted from both trees $T_1$ and $T_2$.
-    Then it is $d(T_1|_n, T_2|_n) \leq d(T_1,T_2) - |r_{T_1}(p_{n+1}) - r_{T_2}(p_{n+1})|$ where $r_{T_i}(p_j)$ determines the rank of the parent of taxon $j$ in tree $T_i$.
+    Let $T$ and $R$ be two trees on $n+1$ taxa.
+    Let $T|_n$ and $R|_n$ be these trees restricted to $n$ taxa, i.e. the same taxon $n+1$ is deleted from both trees $T$ and $R$.
+    Then it is $d(T|_n, R|_n) \leq d(T,R) - |rank_T(parent(n+1)) - rank_R(parent(n+1})|$.
     \label{lemma:distance_delete_taxon}
@@
 \begin{proof}
-    Induction on $k = d(T_1,T_2)$:
-
-    Basis($k = 1$):
-    If rank of $p_{n+1}$ changes between $T_1$ and $T_2$, it is $d(T_1|_n, T_2|_n) = 0$ and $|r_{T_1}(p_{n+1}) - r_{T_2}(p_{n+1})| = 1$ and therefore, $d(T_1|_n, T_2|_n) = 0 \leq 1 - 1 = d(T_1,T_2) - |r_{T_1}(p_{n+1}) - r_{T_2}(p_{n+1})|$.
-    If on the other hand the rank of $p_{n+1}$ does not change between $T_1$ and $T_2$, it is $d(T_1|_n, T_2|_n) = d(T_1,T_2) = 1$ and $|r_{T_1}(p_{n+1}) - r_{T_2}(p_{n+1})| = 1$ and therefore, $d(T_1|_n, T_2|_n) = 1 \leq 1 - 0 = d(T_1,T_2) - |r_{T_1}(p_{n+1}) - r_{T_2}(p_{n+1})|$.
-
-    Step($k \to k+1$):
-    Let $T_1$ and $T_2$ be trees on $n+1$ taxa with distance $d(T_1,T_2) = k$.
-    Let $T_2'$ be the last tree on a shortest path from $T_1$ to $T_2$, i.e. $d(T_1,T_2') = k$.
-    We can apply the induction hypothesis on $T_1$ and $T_2'$ and see that $d(T_1|_n, T_2'|_n) \leq d(T_1, T_2') - |r_{T_1}(p_{n+1}) - r_{T_2'}(p_{n+1})| := k - m$.
-    Now we differ two cases:
-    \begin{enumerate}
-        \item $p_{n+1}$ changes its rank between $T_2$ and $T_2'$ or is affected by the $\nni$ move is performed between $T_2$ and $T_2'$.
-
-        Then the trees $T_2|_n$ and $T_2'|_n$ are equal. \todo{Is this obivous?}
-        Therefore, it is $d(T_1|_n,T_2|_n) = d(T_1|_n, T_2'|_n) \leq k - m$.
-        It is either $|r_{T_1}(p_{n+1}) - r_{T_2}(p_{n+1})| = m+1$, $|r_{T_1}(p_{n+1}) - r_{T_2}(p_{n+1})| = m$, or $|r_{T_1}(p_{n+1}) - r_{T_2}(p_{n+1})| = m-1$.
-        In either case, we have $d(T_1|_n,T_2|_n) = d(T_1|_n, T_2'|_n) \leq k - m = (k+1) - (m+1) \leq d(T_1, T_2) - |r_{T_1}(p_{n+1}) - r_{T_2}(p_{n+1})|$.
-
-        \item $p_{n+1}$ has same rank in $T_2$ and $T_2'$.
-
-        In this case it is $d(T_1|_n,T_2|_n) \leq d(T_1|_n,T_2'|_n) + d(T_2'|_n, T_2|_n) \leq d(T_1|_n, T_2'|_n) + 1 \leq (k-m)+ 1$.
-        It also follows that $|r_{T_1}(p_{n+1}) - r_{T_2'}(p_{n+1})| = |r_{T_1}(p_{n+1}) - r_{T_2}(p_{n+1})|$ and therefore, we can conclude $d(T_1|_n,T_2|_n) \leq k-m+1 = k+1-m = d(T_1,T_2) -  |r_{T_1}(p_{n+1}) - r_{T_2}(p_{n+1})|$.
-    \end{enumerate}
+    On every path from $T$ to $R$ there are at least $|rank_T(parent(n+1)) - rank_R(parent(n+1})|$ $\rnni$ moves that change the rank of $parent(n+1)$ because each $\rnni$ move can change the rank of an internal node by at most one.
+    In particular, this holds for a shortest path $p$ from $T$ to $R$.
+    Furthermore, deleting taxon $n+1$ and its parent $p_{n+1}$ from all trees on $p$ results in a path $p|_n$ from $T|_n$ to $R|_n$.
+    All moves that do not involve $p_{n+1}$ can be performed the same way and those that involved $p_{n+1}$ are just left out.
+    We thus have a path $p|_n$ from $T|_n$ to $R|_n$ that has at least $|rank_T(parent(n+1)) - rank_R(parent(n+1})|$ $\rnni$ moves less than $p$, which results in $|p|_n| \leq d(T,R) - |rank_T(parent(n+1)) - rank_R(parent(n+1})|$.
+    \todo{Is this proof sufficient? Otherwise: induction}
+    % Induction on $k = d(T,R)$:
+    %
+    % Basis($k = 1$):
+    % If rank of $parent(n+1)$ changes between $T$ and $R$, it is $d(T|_n, R|_n) = 0$ and $|rank_T(parent(n+1)) - rank_R(parent(n+1))| = 1$ and therefore, $d(T|_n, R|_n) = 0 \leq 1 - 1 = d(T,R) - |rank_T(parent(n+1)) - rank_R(parent(n+1))|$.
+    % If on the other hand the rank of $parent(n+1}$ does not change between $T$ and $R$, it is $d(T|_n, R|_n) = d(T,R) = 1$ and $|r_{T}(parent(n+1}) - r_{R}(parent(n+1})| = 1$ and therefore, $d(T|_n, R|_n) = 1 \leq 1 - 0 = d(T,R) - |r_{T}(parent(n+1}) - r_{R}(parent(n+1})|$.
+    %
+    % Step($k \to k+1$):
+    % Let $T$ and $R$ be trees on $n+1$ taxa with distance $d(T,R) = k$.
+    % Let $R'$ be the last tree on a shortest path from $T$ to $R$, i.e. $d(T,R') = k$.
+    % We can apply the induction hypothesis on $T$ and $R'$ and see that $d(T|_n, R'|_n) \leq d(T, R') - |r_{T}(parent(n+1}) - r_{R'}(parent(n+1})| := k - m$.
+    % Now we differ two cases:
+    % \begin{enumerate}
+    %     \item $parent(n+1}$ changes its rank between $R$ and $R'$ or is affected by the $\nni$ move is performed between $R$ and $R'$.
+    %
+    %     Then the trees $R|_n$ and $R'|_n$ are equal. \todo{Is this obivous?}
+    %     Therefore, it is $d(T|_n,R|_n) = d(T|_n, R'|_n) \leq k - m$.
+    %     It is either $|r_{T}(parent(n+1}) - r_{R}(parent(n+1})| = m+1$, $|r_{T}(parent(n+1}) - r_{R}(parent(n+1})| = m$, or $|r_{T}(parent(n+1}) - r_{R}(parent(n+1})| = m-1$.
+    %     In either case, we have $d(T|_n,R|_n) = d(T|_n, R'|_n) \leq k - m = (k+1) - (m+1) \leq d(T, R) - |r_{T}(parent(n+1}) - r_{R}(parent(n+1})|$.
+    %
+    %     \item $parent(n+1}$ has same rank in $R$ and $R'$.
+    %
+    %     In this case it is $d(T|_n,R|_n) \leq d(T|_n,R'|_n) + d(R'|_n, R|_n) \leq d(T|_n, R'|_n) + 1 \leq (k-m)+ 1$.
+    %     It also follows that $|r_{T}(parent(n+1}) - r_{R'}(parent(n+1})| = |r_{T}(parent(n+1}) - r_{R}(parent(n+1})|$ and therefore, we can conclude $d(T|_n,R|_n) \leq k-m+1 = k+1-m = d(T,R) -  |r_{T}(parent(n+1}) - r_{R}(parent(n+1})|$.
+    % \end{enumerate}
 \end{proof}
@@
 \begin{lemma}
-    Let $T$ and $T'$ be two caterpillar trees.
-    $T$ and $T'$ have distance $d(T,T') = \Delta(\rnniu) = \frac{(n-1)(n-2)}{2}$ if, and only if, they have caterpillar distance $d_c(T,T') = \Delta(\rnniu) = \frac{(n-1)(n-2)}{2}$.
-\todo{define $d_c$}
+    Let $T$ and $R$ be two caterpillar trees.
+    $T$ and $R$ have distance $d(T,R) = \Delta(\rnniu) = \frac{(n-1)(n-2)}{2}$ if, and only if, they have caterpillar distance $d_c(T,R) = \Delta(\rnniu) = \frac{(n-1)(n-2)}{2}$. \todo{define $d_c$}
     \label{lemma:caterpillar_dist=diameter}
@@
 
-    % \todo{At the top of $T'$ it could also be $2,1$, or $t=1$}
+    % \todo{At the top of $R$ it could also be $2,1$, or $t=1$}
     %
@@
     %     \includegraphics[width=0.6\textwidth]{caterpillar_max_dist}
-    %     \caption{Caterpillar trees $T$ and $T'$ with distance $d(T,T') = \Delta(\rnniu)$}
+    %     \caption{Caterpillar trees $T$ and $R$ with distance $d(T,R) = \Delta(\rnniu)$}
     %     \label{fig:caterpillar_max_distance}
@@
 \begin{proof}
-    Let $d_c(T,T')$ denote the length of a shortest caterpillar path from $T$ to $T'$, that is a caterpillar path that is shortest among all caterpillar paths from $T$ to $T'$.
-    We need to prove that $d(T,T') = \Delta(\rnniu)$ if, and only if, $d_c(T,T') = \Delta(\rnniu)$.
+    Let $d_c(T,R)$ denote the length of a shortest caterpillar path from $T$ to $R$, that is a caterpillar path that is shortest among all caterpillar paths from $T$ to $R$.
+    We need to prove that $d(T,R) = \Delta(\rnniu)$ if, and only if, $d_c(T,R) = \Delta(\rnniu)$.
 
-    Let $T$ and $T'$ have distance $d(T,T') = \Delta(\rnniu)$.
-    We can compute a  shortest caterpillar path from $T$ to $T'$ using Bubble Sort.
+    Let $T$ and $R$ have distance $d(T,R) = \Delta(\rnniu)$.
+    We can compute a  shortest caterpillar path from $T$ to $R$ using Bubble Sort.
     In each loop $k=1, \ldots, n-2$ of the algorithm there can be at most $n-1-k$ exchanges of taxa.
     It follows that the maximum length of a path computed by Bubble Sort is $\sum\limits_{k=1}^{n-2} k = \frac{(n-1)(n-2)}{2}$.
-    Since there cannot be a caterpillar path shorter than $\frac{(n-1)(n-2)}{2}$ if it is $d(T,T') = \frac{(n-1)(n-2)}{2}$, it follows $d_c(T,T') = \Delta(\rnniu)$.
+    Since there cannot be a caterpillar path shorter than $\frac{(n-1)(n-2)}{2}$ if it is $d(T,R) = \frac{(n-1)(n-2)}{2}$, it follows $d_c(T,R) = \Delta(\rnniu)$.
 
-    Let now $T$ and $T'$ have caterpillar distance $d_c(T,T') = \Delta(\rnniu)$.
-    We prove $d(T,T') = d_c(T,T')$ by induction on the number of taxa $n$ of $T$ and $T'$:
+    Let now $T$ and $R$ have caterpillar distance $d_c(T,R) = \Delta(\rnniu)$.
+    We prove $d(T,R) = d_c(T,R)$ by induction on the number of taxa $n$ of $T$ and $R$:
     We assume without loss of generality that $T$ is the caterpillar tree $(\ldots ((1,2),3), \ldots, n)$.
-    Otherwise, we re-label $T$ and $T'$.
-    Let $T'$ be a caterpillar tree with distance $\Delta(\rnniu)$ to $T$.
+    Otherwise, we re-label $T$ and $R$.
+    Let $R$ be a caterpillar tree with distance $\Delta(\rnniu)$ to $T$.
 
@@
     Induction hypothesis: For trees $R$, $R'$ on $n$ taxa it is $d_c(R,R') = d(R, R') = \frac{(n-1)(n-2)}{2} = \Delta(\rnniu_n)$.\\
-    Let $T, T'$ be trees on $n+1$ taxa with caterpillar distance $\Delta(\rnniu_{n+1})$.
-    When applying the Bubble Sort algorithm to compute a path between $T'$ and $T$, each pair of taxa that is compared in that algorithm exchanges, since there must be $\Delta(\rnniu)=\frac{(n-1)(n-2)}{2}$ moves between $T$ and $T'$.
-    So we can conclude that taxon $n+1$ must be in the cherry of $T'$.
-    Let us consider the trees $T|_n$ to $T'|_n$ that result from $T$ and $T'$ by deleting taxon $n+1$.
-    When running Bubble Sort to compute a path from $T|_n$ to $T'|_n$ one can observe the same exchanges of taxa as on the Bubble Sort path from $T$ to $T'$, except for the $n-1$ ones that involve taxon $n+1$.
-    Hence, the caterpillar distance is $d_c(T|_n, T'|_n) = \frac{n(n-1)}{2} - (n-1)$ = $\frac{(n-1)(n-2)}{2}$.
-    So we can apply the induction hypothesis on $T|_n$ and $T'|_n$ and know that $d(T|_n,T'|_n) = \frac{(n-1)(n-2)}{2}$.
+    Let $T, R$ be trees on $n+1$ taxa with caterpillar distance $\Delta(\rnniu_{n+1})$.
+    When applying the Bubble Sort algorithm to compute a path between $R$ and $T$, each pair of taxa that is compared in that algorithm exchanges, since there must be $\Delta(\rnniu)=\frac{(n-1)(n-2)}{2}$ moves between $T$ and $R$.
+    So we can conclude that taxon $n+1$ must be in the cherry of $R$.
+    Let us consider the trees $T|_n$ to $R|_n$ that result from $T$ and $R$ by deleting taxon $n+1$.
+    When running Bubble Sort to compute a path from $T|_n$ to $R|_n$ one can observe the same exchanges of taxa as on the Bubble Sort path from $T$ to $R$, except for the $n-1$ ones that involve taxon $n+1$.
+    Hence, the caterpillar distance is $d_c(T|_n, R|_n) = \frac{n(n-1)}{2} - (n-1)$ = $\frac{(n-1)(n-2)}{2}$.
+    So we can apply the induction hypothesis on $T|_n$ and $R|_n$ and know that $d(T|_n,R|_n) = \frac{(n-1)(n-2)}{2}$.
 
-    With Lemma~\ref{lemma:distance_delete_taxon} we know that the distance between $T|_n$ and $T'|_n$ is at least $d(T|_n, T'|_n) \leq d(T,T') - |r_{T}(p_{n+1}) - r_{T'}(p_{n+1})|$.
-    We also know $|r_{T}(p_{n+1}) - r_{T'}(p_{n+1})| = n-1$
-    If we assume that there is a path from $T$ to $T'$ of length less than $\Delta(\rnniu_{n+1}) = \frac{n(n-1)}{2}$, it would follow that $d(T|_n, T'|_n) \leq d(T,T') - |r_{T}(p_{n+1}) - r_{T'}(p_{n+1})| < \frac{n(n-1)}{2} - (n-1) < \frac{(n-1)(n-2)}{2}$.
-    So it would be $d(T|_n, T'|_n) < \Delta(\rnniu_n)$.
-    Since this is a contradiction to $d(T|_n, T'|_n) = \Delta(\rnniu_n)$, it must be $d(T,T') = \Delta(\rnniu_{n+1}) = \frac{n(n-1)}{2} $.
+    With Lemma~\ref{lemma:distance_delete_taxon} we know that the distance between $T|_n$ and $R|_n$ is at least $d(T|_n, R|_n) \leq d(T,R) - |r_{T}(parent(n+1}) - r_{R}(parent(n+1})|$.
+    We also know $|r_{T}(parent(n+1}) - r_{R}(parent(n+1})| = n-1$
+    If we assume that there is a path from $T$ to $R$ of length less than $\Delta(\rnniu_{n+1}) = \frac{n(n-1)}{2}$, it would follow that $d(T|_n, R|_n) \leq d(T,R) - |r_{T}(parent(n+1}) - r_{R}(parent(n+1})| < \frac{n(n-1)}{2} - (n-1) < \frac{(n-1)(n-2)}{2}$.
+    So it would be $d(T|_n, R|_n) < \Delta(\rnniu_n)$.
+    Since this is a contradiction to $d(T|_n, R|_n) = \Delta(\rnniu_n)$, it must be $d(T,R) = \Delta(\rnniu_{n+1}) = \frac{n(n-1)}{2} $.
 
@@
 \begin{proof}
-    We prove the Theorem by backwards induction on the length $l$ of a shortest caterpillar path between two caterpillar trees $T$ and $T'$.
+    We prove the Theorem by backwards induction on the length $l$ of a shortest caterpillar path between two caterpillar trees $T$ and $R$.
 
     Induction basis: $l = \Delta(\rnniu) = \frac{(n-1)(n-2)}{2}$\\
-    With Lemma~\ref{lemma:caterpillar_dist=diameter} one can see that for each pair of caterpillar trees $T$, $T'$ with caterpillar distance $d_c(T,T') = \Delta(\rnniu)$, it is $d(T,T') = d_c(T,T')$, which means that there is a shortest path between $T$ and $T'$ that is a caterpillar path.
+    With Lemma~\ref{lemma:caterpillar_dist=diameter} one can see that for each pair of caterpillar trees $T$, $R$ with caterpillar distance $d_c(T,R) = \Delta(\rnniu)$, it is $d(T,R) = d_c(T,R)$, which means that there is a shortest path between $T$ and $R$ that is a caterpillar path.
 
@@
     Induction hypothesis: For each pair of caterpillar trees with caterpillar distance $l+1$ there is a shortest path between these trees that only consists of caterpillar trees, which means that the distance between these trees is $l+1$.
-    Let $T$ and $T'$ be caterpillar trees with caterpillar distance $l$.
-    Since $l < \Delta(\rnniu) = \frac{(n-1)(n-2)}{2}$, there is one loop $k$ in the Bubble Sort algorithm from $T'$ to $T$ where there are less than $n-2-k$ swaps of neighboured taxa.
+    Let $T$ and $R$ be caterpillar trees with caterpillar distance $l$.
+    Since $l < \Delta(\rnniu) = \frac{(n-1)(n-2)}{2}$, there is one loop $k$ in the Bubble Sort algorithm from $R$ to $T$ where there are less than $n-2-k$ swaps of neighboured taxa.
     This means that there is a taxon $a$ that has parent with rank $n-2-k$ in $T$ and that does not need to move all the way up starting at the cherry in loop $k$.
     Therefore, there must be a taxon $b$ which has parent with rank one less than the parent of $a$ in the tree at the beginning of loop number $k$.
-    If we assume that loop $k$ is the first loop where a single taxon is not moved by $n-k-2$ $\rnni$ moves, it follows that $a$ and $b$ must be neighboured taxa in $T'$.
-    Let $T''$ be the caterpillar tree resulting from $T'$ by exchanging $a$ and $b$.
-    It follows that $T'$ has caterpillar distance $d_c(T,T'') = d_c(T,T') + 1 = l+1$ to $T$ and $d(T',T'') = 1$.
-    We can apply the induction hypothesis on $T$ and $T''$ and know that $d(T,T'') = l+1$.
-    If it was $d(T,T') < l$, it would hold $d(T,T'') \leq d(T,T') + d(T',T'') < l + 1$ which contradicts the induction hypothesis.
-    We can conclude that $d(T,T') = l$.
+    If we assume that loop $k$ is the first loop where a single taxon is not moved by $n-k-2$ $\rnni$ moves, it follows that $a$ and $b$ must be neighboured taxa in $R$.
+    Let $R'$ be the caterpillar tree resulting from $R$ by exchanging $a$ and $b$.
+    It follows that $R$ has caterpillar distance $d_c(T,R') = d_c(T,R) + 1 = l+1$ to $T$ and $d(R,R') = 1$.
+    We can apply the induction hypothesis on $T$ and $R'$ and know that $d(T,R') = l+1$.
+    If it was $d(T,R) < l$, it would hold $d(T,R') \leq d(T,R) + d(R,R') < l + 1$ which contradicts the induction hypothesis.
+    We can conclude that $d(T,R) = l$.
 
@@
     When running this algorithm on $T$ and $R$, taxon $i$ swaps $n-1-i$ times with its right neighbour.
-    Therefore, the length of a shortest caterpillar path is ${\sum\limits_{i=1}^{n-1}(n-1-i) = \sum\limits_{i=0}^{n-2}i = \frac{(n-1)(n-2)}{2}}$.
+    Therefore, the length of a shortest caterpillar path is \sum\limits_{i=1}^{n-1}(n-1-i) = \sum\limits_{i=0}^{n-2}i = \frac{(n-1)(n-2)}{2}.
 \end{proof}
@@
 % \begin{lemma}
-% 	Let $T_1$ and $T_2$ be two caterpillar trees that both have a leaf labelled by $a$ as part of their cherry.
-% 	There is a shortest paths from $T_1$ to $T_2$ that is a caterpillar path.
+% 	Let $T$ and $R$ be two caterpillar trees that both have a leaf labelled by $a$ as part of their cherry.
+% 	There is a shortest paths from $T$ to $R$ that is a caterpillar path.
 % 	\label{lemma:caterpillar_subset_convex}
@@
 % \begin{proof}
-% 	Let $T_1$ and $T_2$ be two caterpillar trees, both having a leaf labelled by $a$ as part of their cherry.
-% 	Let $p$ be a shortest path from $T_1$ to $T_2$.
-% 	We introduce a labelling of internal nodes of all trees on $p$ by first defining it for the caterpillar trees $T_1$ and $T_2$ and then explaining how the labelling is changed by $\rnni$ moves.
+% 	Let $T$ and $R$ be two caterpillar trees, both having a leaf labelled by $a$ as part of their cherry.
+% 	Let $p$ be a shortest path from $T$ to $R$.
+% 	We introduce a labelling of internal nodes of all trees on $p$ by first defining it for the caterpillar trees $T$ and $R$ and then explaining how the labelling is changed by $\rnni$ moves.
 %
-% 	In $T_1$ and $T_2$ the internal node of the cherry is labelled by the label of its child that is not $a$.
+% 	In $T$ and $R$ the internal node of the cherry is labelled by the label of its child that is not $a$.
 % 	All other internal nodes shall have the same label as their leaf children.
@@
 % 	Since there are moves on $p$ that do not change the ranks of internal nodes, this caterpillar path has length $|r| \leq |p|$.
-% 	Hence, there is always a shortest path between $T_1$ and $T_2$ that is a caterpillar path.
+% 	Hence, there is always a shortest path between $T$ and $R$ that is a caterpillar path.
 % \end{proof}
@@
 \begin{theorem}
-	There are trees $T_1,T_2$ in $\nni$ sharing a partition which is not shared by any intermediate tree on a shortest path from $T_1$ to $T_2$.
+	There are trees $T,R$ in $\nni$ sharing a partition which is not shared by any intermediate tree on a shortest path from $T$ to $R$.
 	\label{thm:split_nni}
@@
 The counterexample presented in~\cite{Li1996} works in $\nni$ is based on following observation:
-Let $T_1$ and $T_2$ be two caterpillar trees.
-If each taxon $a$ is replaced by a cherry $(a_1,a_2)$ in both trees $T_1$ and $T_2$, the resulting trees $T_1'$ and $T_2'$ have distance $d_{\nni}(T_1',T_2') = d_{\nni}(T_1,T_2)$ in the $\nni$ graph.
+Let $T$ and $R$ be two caterpillar trees.
+If each taxon $a$ is replaced by a cherry $(a_1,a_2)$ in both trees $T$ and $R$, the resulting trees $R$ and $R'$ have distance $d_{\nni}(R,R') = d_{\nni}(T,R)$ in the $\nni$ graph.
 When proving~\ref{thm:split_nni},~\cite{Li1996} give an example of two trees $T$ and $R$ on $2n$ taxa that consist of two caterpillar trees each that share the root as parent.
 The split given by the edges incident to the root is $1, \ldots, n | n+1, \ldots, 2n$ in both trees $T$ and $R$.
-For the given example there is a shortest path in $\nni$ where at first $T$ is transformed to a tree $T'$ with $n$ cherries where each cherry contains one taxon of $\{1, \ldots, n\}$ and one taxon of $\{n+1, \ldots, 2n\}$.
-Afterwards, $T'$ is tranformed to a tree $R'$ that contains the same cherries as $T'$.
+For the given example there is a shortest path in $\nni$ where at first $T$ is transformed to a tree $R$ with $n$ cherries where each cherry contains one taxon of $\{1, \ldots, n\}$ and one taxon of $\{n+1, \ldots, 2n\}$.
+Afterwards, $R$ is tranformed to a tree $R'$ that contains the same cherries as $R$.
 Finally, all cherries of $R'$ are resolved in order to end up in tree $R$.
-For proving that this path is shorter than any path preserving the split $1, \ldots, n | n+1, \ldots, 2n$, the fact that $d_{\nni}(T',R') = d_{\nni}(T|_{\{1, \ldots, n\}}, R|_{\{1, \ldots, n\}})$ is being used.
+For proving that this path is shorter than any path preserving the split $1, \ldots, n | n+1, \ldots, 2n$, the fact that $d_{\nni}(R,R') = d_{\nni}(T|_{\{1, \ldots, n\}}, R|_{\{1, \ldots, n\}})$ is being used.
 However, this does not work in $\rnni$.
@@
 % How do we assign the ranks to the new cherry nodes?
-% We could either create a path where $T'$ and $R'$ have distance equal to $d_{\rnni}(T|_{\{1, \ldots, n\}}, R|_{\{n+1, \ldots, 2n\}})$ and have a lot of additional rank moves before and after that part OR
-% Build $T'$ and $R'$ that contain all the cherries within the least number of moves possible (3n-3) and having a longer path between $T'$ and $R'$.
+% We could either create a path where $R$ and $R'$ have distance equal to $d_{\rnni}(T|_{\{1, \ldots, n\}}, R|_{\{n+1, \ldots, 2n\}})$ and have a lot of additional rank moves before and after that part OR
+% Build $R$ and $R'$ that contain all the cherries within the least number of moves possible (3n-3) and having a longer path between $R$ and $R'$.
 % Also anything in between these two options is possible. Are we able to disprove every option??
@@
 	\includegraphics[width=\textwidth]{splitthm_counterexample}
-	\caption{The split $123|45$ is present in $T_1$ and $T_2$, but the path at the top is a shortest path where none of the trees contains this split.
+	\caption{The split $123|45$ is present in $T$ and $R$, but the path at the top is a shortest path where none of the trees contains this split.
     However, this split is maintained on the path at the bottom that is a shortest path as well.}
@@
 Considering clusters rather than splits is more reasonable since two rooted trees have the same set of clusters if, and only if, they are isomorphic~\cite{steel2016}.
-The trees $T_1, T_2$ of counterexample~\ref{splitthm_counterexample} share the same set of induced splits, but have distinct sets of clusters.
+The trees $T, R$ of counterexample~\ref{splitthm_counterexample} share the same set of induced splits, but have distinct sets of clusters.
 
@@
 \begin{lemma}
-    Let $T_1$ and $T_2$ be caterpillar trees.
-    It is $d_c(T_1,T_2) = \Delta(\rnniu)$ if, and only if, $T_1$ and $T_2$ share no induced triplet.
+    Let $T$ and $R$ be caterpillar trees.
+    It is $d_c(T,R) = \Delta(\rnniu)$ if, and only if, $T$ and $R$ share no induced triplet.
     \todo{define induced triplets}
@@
     % Insert the proof here
-    Fix $T_1 = (\ldots(1,2) \ldots ,n)$, use induction on $n$ and bubble sort (taxon $n$ must be in cherry of $T_2$ if distance is max).
+    Fix $T = (\ldots(1,2) \ldots ,n)$, use induction on $n$ and bubble sort (taxon $n$ must be in cherry of $R$ if distance is max).
 \end{proof}
@@
 \begin{conjecture}
-    Let $T_1$ and $T_2$ be trees on $n$ taxa.
-    It is $d(T_1,T_2) = \Delta(\rnniu)$ if, and only if, $T_2$ can be produced from $T_1$ by applying Algorithm~\ref{alg:max_dist_tree}.
+    Let $T$ and $R$ be trees on $n$ taxa.
+    It is $d(T,R) = \Delta(\rnniu)$ if, and only if, $R$ can be produced from $T$ by applying Algorithm~\ref{alg:max_dist_tree}.
 \end{conjecture}

### 2019-06-26 paper.tex
proof: caterpillar distance = distance if max

@@
 
-A \amph{rooted binary phylogenetic tree} on a set $X$ is a pair $(t, \Phi)$ where $t$ is a rooted binary tree with leaf set $\mathcal{L}(t)$ and $\Phi:X \mapsto \mathcal{L}(T)$ a bijective mapping.
+A \emph{rooted binary phylogenetic tree} on a set $X$ is a pair $(t, \Phi)$ where $t$ is a rooted binary tree with leaf set $\mathcal{L}(t)$ and $\Phi:X \mapsto \mathcal{L}(T)$ a bijective mapping.
 \todo{do we need to define leaves and internal nodes, child, parent, subtree, cherry(defined below)?}
@@
     Let $T|_n$ and $R|_n$ be these trees restricted to $n$ taxa, i.e. the same taxon $n+1$ is deleted from both trees $T$ and $R$.
-    Then it is $d(T|_n, R|_n) \leq d(T,R) - |rank_T(parent(n+1)) - rank_R(parent(n+1})|$.
+    Then it is $d(T|_n, R|_n) \leq d(T,R) - |rank_T(parent(n+1)) - rank_R(parent(n+1))|$.
     \label{lemma:distance_delete_taxon}
@@
 \begin{proof}
-    On every path from $T$ to $R$ there are at least $|rank_T(parent(n+1)) - rank_R(parent(n+1})|$ $\rnni$ moves that change the rank of $parent(n+1)$ because each $\rnni$ move can change the rank of an internal node by at most one.
+    On every path from $T$ to $R$ there are at least $|rank_T(parent(n+1)) - rank_R(parent(n+1))|$ $\rnni$ moves that change the rank of $parent(n+1)$ because each $\rnni$ move can change the rank of an internal node by at most one.
     In particular, this holds for a shortest path $p$ from $T$ to $R$.
@@
     All moves that do not involve $p_{n+1}$ can be performed the same way and those that involved $p_{n+1}$ are just left out.
-    We thus have a path $p|_n$ from $T|_n$ to $R|_n$ that has at least $|rank_T(parent(n+1)) - rank_R(parent(n+1})|$ $\rnni$ moves less than $p$, which results in $|p|_n| \leq d(T,R) - |rank_T(parent(n+1)) - rank_R(parent(n+1})|$.
+    We thus have a path $p|_n$ from $T|_n$ to $R|_n$ that has at least $|rank_T(parent(n+1)) - rank_R(parent(n+1))|$ $\rnni$ moves less than $p$, which results in $|p|_n| \leq d(T,R) - |rank_T(parent(n+1)) - rank_R(parent(n+1))|$.
     \todo{Is this proof sufficient? Otherwise: induction}
@@
     Let $T$ and $R$ be two caterpillar trees.
-    $T$ and $R$ have distance $d(T,R) = \Delta(\rnniu) = \frac{(n-1)(n-2)}{2}$ if, and only if, they have caterpillar distance $d_c(T,R) = \Delta(\rnniu) = \frac{(n-1)(n-2)}{2}$. \todo{define $d_c$}
+    $T$ and $R$ have distance $d(T,R) = \frac{(n-1)(n-2)}{2}$ if, and only if, they have caterpillar distance $d_c(T,R) = \frac{(n-1)(n-2)}{2}$.
     \label{lemma:caterpillar_dist=diameter}
@@
 
-    % \todo{At the top of $R$ it could also be $2,1$, or $t=1$}
-    %
-    % \begin{figure}[H]
-    %     \centering
-    %     \includegraphics[width=0.6\textwidth]{caterpillar_max_dist}
-    %     \caption{Caterpillar trees $T$ and $R$ with distance $d(T,R) = \Delta(\rnniu)$}
-    %     \label{fig:caterpillar_max_distance}
-    % \end{figure}
-
 \begin{proof}
-    Let $d_c(T,R)$ denote the length of a shortest caterpillar path from $T$ to $R$, that is a caterpillar path that is shortest among all caterpillar paths from $T$ to $R$.
-    We need to prove that $d(T,R) = \Delta(\rnniu)$ if, and only if, $d_c(T,R) = \Delta(\rnniu)$.
-
-    Let $T$ and $R$ have distance $d(T,R) = \Delta(\rnniu)$.
-    We can compute a  shortest caterpillar path from $T$ to $R$ using Bubble Sort.
-    In each loop $k=1, \ldots, n-2$ of the algorithm there can be at most $n-1-k$ exchanges of taxa.
-    It follows that the maximum length of a path computed by Bubble Sort is $\sum\limits_{k=1}^{n-2} k = \frac{(n-1)(n-2)}{2}$.
+    Let $T$ and $R$ be caterpillar trees.
+    Let us first prove that if $d(T,R) = \frac{(n-1)(n-2)}{2}$, it follows $d_c(T,R) = \frac{(n-1)(n-2)}{2}$.
+    This easily follows from the modified Bubble Sort algorithm. \todo{name?}
+    In each loop $i=1, \ldots, n-2$ of the algorithm there can be at most $n-1-i$ exchanges of taxa.
+    It follows that the maximum length of a path computed by this algorithm is $\sum\limits_{k=1}^{n-2} k = \frac{(n-1)(n-2)}{2}$.
     Since there cannot be a caterpillar path shorter than $\frac{(n-1)(n-2)}{2}$ if it is $d(T,R) = \frac{(n-1)(n-2)}{2}$, it follows $d_c(T,R) = \Delta(\rnniu)$.
 
-    Let now $T$ and $R$ have caterpillar distance $d_c(T,R) = \Delta(\rnniu)$.
-    We prove $d(T,R) = d_c(T,R)$ by induction on the number of taxa $n$ of $T$ and $R$:
-    We assume without loss of generality that $T$ is the caterpillar tree $(\ldots ((1,2),3), \ldots, n)$.
-    Otherwise, we re-label $T$ and $R$.
-    Let $R$ be a caterpillar tree with distance $\Delta(\rnniu)$ to $T$.
+    For proving the other direction of the statement, let now $T$ and $R$ have caterpillar distance $d_c(T,R) =  \frac{(n-1)(n-2)}{2}$.
+    We prove $d(T,R) = \frac{(n-1)(n-2)}{2}$ by induction on the number of taxa $n$ of $T$ and $R$:
+    Without loss of generality we can assume that $T$ is the caterpillar tree $(\ldots ((1,2),3), \ldots, n)$.
+    Let $R$ be a caterpillar tree with distance $\frac{(n-1)(n-2)}{2}$ to $T$.
 
     Induction basis: $n=3$\\
-    All trees have distance $1$ and are caterpillar trees.
+    All three trees have distance $1$ from another and are caterpillar trees, hence it is $d(T,R) = d_c(T,R)$ for all trees $T,R$ on $n=3$ taxa.
 
     Induction step: $n \to n+1$\\
-    Induction hypothesis: For trees $R$, $R'$ on $n$ taxa it is $d_c(R,R') = d(R, R') = \frac{(n-1)(n-2)}{2} = \Delta(\rnniu_n)$.\\
-    Let $T, R$ be trees on $n+1$ taxa with caterpillar distance $\Delta(\rnniu_{n+1})$.
-    When applying the Bubble Sort algorithm to compute a path between $R$ and $T$, each pair of taxa that is compared in that algorithm exchanges, since there must be $\Delta(\rnniu)=\frac{(n-1)(n-2)}{2}$ moves between $T$ and $R$.
-    So we can conclude that taxon $n+1$ must be in the cherry of $R$.
+    Induction hypothesis: For trees $T, R$ on $n$ taxa it is $d(T, R) = \frac{n(n-1)}{2}$.\\
+    Let $T, R$ be trees on $n+1$ taxa with caterpillar distance $\frac{(n-1)(n-2)}{2}$.
+    Therefore, each of the $\frac{(n-1)(n-2)}{2}$ pairs of taxa that is considered when running modified Bubble Sort algorithm \todo{name} to compute a path $p$ from $R$ to $T$ exchanges.
+    From this we can conclude that taxon $n+1$, whose parent has highest rank in $T$, is part of the cherry of $R$.
     Let us consider the trees $T|_n$ to $R|_n$ that result from $T$ and $R$ by deleting taxon $n+1$.
-    When running Bubble Sort to compute a path from $T|_n$ to $R|_n$ one can observe the same exchanges of taxa as on the Bubble Sort path from $T$ to $R$, except for the $n-1$ ones that involve taxon $n+1$.
+    Bubble Sort \todo{name} computes a path from $T|_n$ to $R|_n$ that contains the same exchanges of taxa as $p$, except for the $n-1$ first ones on $p$ that involve taxon $n+1$.
     Hence, the caterpillar distance is $d_c(T|_n, R|_n) = \frac{n(n-1)}{2} - (n-1)$ = $\frac{(n-1)(n-2)}{2}$.
@@
 
-    With Lemma~\ref{lemma:distance_delete_taxon} we know that the distance between $T|_n$ and $R|_n$ is at least $d(T|_n, R|_n) \leq d(T,R) - |r_{T}(parent(n+1}) - r_{R}(parent(n+1})|$.
-    We also know $|r_{T}(parent(n+1}) - r_{R}(parent(n+1})| = n-1$
-    If we assume that there is a path from $T$ to $R$ of length less than $\Delta(\rnniu_{n+1}) = \frac{n(n-1)}{2}$, it would follow that $d(T|_n, R|_n) \leq d(T,R) - |r_{T}(parent(n+1}) - r_{R}(parent(n+1})| < \frac{n(n-1)}{2} - (n-1) < \frac{(n-1)(n-2)}{2}$.
-    So it would be $d(T|_n, R|_n) < \Delta(\rnniu_n)$.
-    Since this is a contradiction to $d(T|_n, R|_n) = \Delta(\rnniu_n)$, it must be $d(T,R) = \Delta(\rnniu_{n+1}) = \frac{n(n-1)}{2} $.
+    According to Lemma~\ref{lemma:distance_delete_taxon}, the distance between $T|_n$ and $R|_n$ is at least $d(T|_n, R|_n) \leq d(T,R) - |rank_T(parent(n+1)) - rank_R(parent(n+1))|$.
+    We also know $|rank_T(parent(n+1)) - rank_R(parent(n+1))| = n-1$
+    Suppose, contrary to our claim, that there is a path from $T$ to $R$ of length less than $\Delta(\rnniu_{n+1}) = \frac{n(n-1)}{2}$.
+    By the above, it follows that $d(T|_n, R|_n) \leq d(T,R) - |rank_T(parent(n+1)) - rank_R(parent(n+1))| < \frac{n(n-1)}{2} - (n-1) < \frac{(n-1)(n-2)}{2}$.
+    Since this is a contradiction to $d(T|_n, R|_n) = \frac{(n-1)(n-2)}{2}$, it is proven that $d(T,R) = \frac{n(n-1)}{2} $.
 
@@
     When running this algorithm on $T$ and $R$, taxon $i$ swaps $n-1-i$ times with its right neighbour.
-    Therefore, the length of a shortest caterpillar path is \sum\limits_{i=1}^{n-1}(n-1-i) = \sum\limits_{i=0}^{n-2}i = \frac{(n-1)(n-2)}{2}.
+    Therefore, the length of a shortest caterpillar path is $\sum\limits_{i=1}^{n-1}(n-1-i) = \sum\limits_{i=0}^{n-2}i = \frac{(n-1)(n-2)}{2}$.
 \end{proof}

### 2019-06-26 paper.tex
add figure RNNI moves definition
edits on proof: caterpillar convex

@@
 
-\todo{figure: rank + $\nni$ move}
+\begin{figure}[H]
+	\centering
+	\includegraphics[width=\textwidth]{RNNI}
+	\caption{Four trees connected by edges in the $\rnni$ graph. A rank move of $mrca(\{1,2\})$ and $mrca(\{3,4\})$ and all $\nni$ moves on the edge highlighted in red are illustrated}
+	\label{fig:NNI_vs_RNNI}
+\end{figure}
+
+\todo{label of taxa in figure shouldn't be numbers}
 
@@
 
-    Induction basis: $l = \Delta(\rnniu) = \frac{(n-1)(n-2)}{2}$\\
-    With Lemma~\ref{lemma:caterpillar_dist=diameter} one can see that for each pair of caterpillar trees $T$, $R$ with caterpillar distance $d_c(T,R) = \Delta(\rnniu)$, it is $d(T,R) = d_c(T,R)$, which means that there is a shortest path between $T$ and $R$ that is a caterpillar path.
+    Induction basis: $l = \frac{(n-1)(n-2)}{2}$\\
+    With Lemma~\ref{lemma:caterpillar_dist=diameter} one can see that for each pair of caterpillar trees $T$, $R$ with caterpillar distance $d_c(T,R) = \frac{(n-1)(n-2)}{2}$, it is $d(T,R) = d_c(T,R)$, which means that there is a shortest path between $T$ and $R$ that is a caterpillar path.
 
@@
     Let $T$ and $R$ be caterpillar trees with caterpillar distance $l$.
-    Since $l < \Delta(\rnniu) = \frac{(n-1)(n-2)}{2}$, there is one loop $k$ in the Bubble Sort algorithm from $R$ to $T$ where there are less than $n-2-k$ swaps of neighboured taxa.
-    This means that there is a taxon $a$ that has parent with rank $n-2-k$ in $T$ and that does not need to move all the way up starting at the cherry in loop $k$.
-    Therefore, there must be a taxon $b$ which has parent with rank one less than the parent of $a$ in the tree at the beginning of loop number $k$.
-    If we assume that loop $k$ is the first loop where a single taxon is not moved by $n-k-2$ $\rnni$ moves, it follows that $a$ and $b$ must be neighboured taxa in $R$.
-    Let $R'$ be the caterpillar tree resulting from $R$ by exchanging $a$ and $b$.
-    It follows that $R$ has caterpillar distance $d_c(T,R') = d_c(T,R) + 1 = l+1$ to $T$ and $d(R,R') = 1$.
-    We can apply the induction hypothesis on $T$ and $R'$ and know that $d(T,R') = l+1$.
-    If it was $d(T,R) < l$, it would hold $d(T,R') \leq d(T,R) + d(R,R') < l + 1$ which contradicts the induction hypothesis.
-    We can conclude that $d(T,R) = l$.
-
+    Since $l < \frac{(n-1)(n-2)}{2}$, there is one loop $k$ in the caterpillar path from $R$ to $T$ that is given by the Bubble Sort algorithm \todo{name} where less than $n-2-k$ pairs taxa swap positions.
+    It follows that there is a taxon $a$ that has parent with rank $n-2-k$ in $T$ and does not need to move all the way up starting at the cherry in loop $k$.
+    Let $b$ be the taxon whose parent has rank one less than the parent of $a$ in the tree at the beginning of loop number $k$.
+    We can assume without loss of generality that loop $k$ is the first loop in which the number of swaps is not maximal.
+    It follows that $a$ and $b$ must be neighboured taxa in $R$.
+    Let us no consider caterpillar tree $R'$ resulting from $R$ by exchanging taxa $a$ and $b$.
+    It is clear that $R'$ has caterpillar distance $d_c(T,R') = d_c(T,R) + 1 = l+1$ to $T$ and moreover, $d(R,R') = 1$.
+    Hence we can apply the induction hypothesis on $T$ and $R'$ and see that $d(T,R') = l+1$.
+    If it were $d(T,R) < l$, we would have $d(T,R') \leq d(T,R) + d(R,R') < l + 1$ which contradicts the induction hypothesis.
+    We can conclude $d(T,R) = l$, which completes the proof.
 \end{proof}
@@
 \end{lemma}
+
+\todo{This is already an implication of the results above. Shall we still keep a subsection for Diameter? In case we find something about the radius, this could be added here.}
 

### 2019-06-26 paper.tex
edits on split thm chapter

@@
 
+\newcommand{\np}{\mathcal{NP}}
 \newcommand{\nni}{\mathrm{NNI}}
@@
 \newcommand{\rnniu}{\mathrm{RNNIu}}
+\newcommand{\tbr}{\mathrm{TBR}}
+\newcommand{\rspr}{\mathrm{rSPR}}
 
@@
 Two trees are called \emph{isomorphic} if the underlying rooted phylogenetic trees are isomorphic and the isomorphism preserves ranks of the nodes.
+
+\todo{define splits and delete definition from split thm chapter!}
 
@@
 
-When analysing the complexity of the $\rnni$ space, the comparison to $\nni$ is an important step.
+%Shall we put this bit into the introduction?
+It is natural to try to relate the $\rnni$ graph to the $\nni$ as we already did in previous sections.
+Since computing distances in tree spaces is a main issue in most commonly known tree spaces as $\nni$~\cite{jiang2000}, $\rspr$~\cite{Bordewich2005}, or $\tbr$~\cite{allen2001subtree} is $\np$-hard, it would be desirable to find a tree space that is less complex.
+However, it is not known whether computing distances in $\rnni$ is $\np$-hard or not.
+The natural approach to solve this question is to compare $\rnni$ and $\nni$ graph to see whether the reason that $\nni$ is $\np$-hard transfers to $\rnni$.
+
 The proof of NP-completeness of $\nni$ in~\cite{jiang2000} is based on the fact that following theorem holds in $\nni$.
@@
 \begin{theorem}
-	There are trees $T,R$ in $\nni$ sharing a partition which is not shared by any intermediate tree on a shortest path from $T$ to $R$.
+	There are trees $T,R$ in $\nni$ sharing a cluster which is not shared by any intermediate tree on any shortest path from $T$ to $R$.
 	\label{thm:split_nni}
@@
 %This part is mainly for myself, not sure whether it is important to have it in the paper
-The counterexample presented in~\cite{Li1996} works in $\nni$ is based on following observation:
-Let $T$ and $R$ be two caterpillar trees.
-If each taxon $a$ is replaced by a cherry $(a_1,a_2)$ in both trees $T$ and $R$, the resulting trees $R$ and $R'$ have distance $d_{\nni}(R,R') = d_{\nni}(T,R)$ in the $\nni$ graph.
-When proving~\ref{thm:split_nni},~\cite{Li1996} give an example of two trees $T$ and $R$ on $2n$ taxa that consist of two caterpillar trees each that share the root as parent.
-The split given by the edges incident to the root is $1, \ldots, n | n+1, \ldots, 2n$ in both trees $T$ and $R$.
-For the given example there is a shortest path in $\nni$ where at first $T$ is transformed to a tree $R$ with $n$ cherries where each cherry contains one taxon of $\{1, \ldots, n\}$ and one taxon of $\{n+1, \ldots, 2n\}$.
-Afterwards, $R$ is tranformed to a tree $R'$ that contains the same cherries as $R$.
-Finally, all cherries of $R'$ are resolved in order to end up in tree $R$.
-For proving that this path is shorter than any path preserving the split $1, \ldots, n | n+1, \ldots, 2n$, the fact that $d_{\nni}(R,R') = d_{\nni}(T|_{\{1, \ldots, n\}}, R|_{\{1, \ldots, n\}})$ is being used.
+\todo{Shall we keep this explanation? The problem is that the $\rnni$ path depends on the assignment of ranks to the cherries produced on the path}
+The counterexample presented in~\cite{Li1996} works in $\nni$ because of the following:
+When proving Theorem~\ref{thm:split_nni},~\cite{Li1996} gives an example of two trees $T$ and $R$ on $2n$ taxa that consist of two caterpillar trees each that share the root as parent.
+The clusters given by the two edges incident to the root are $\{1, \ldots, n\}$ and $\{n+1, \ldots, 2n\}$ in both trees $T$ and $R$.
+Between these there is a shortest path in $\nni$ where at first $T$ is transformed to a tree $T'$ with $n$ cherries where each cherry contains one taxon of $\{1, \ldots, n\}$ and one taxon of $\{n+1, \ldots, 2n\}$.
+Afterwards, $T'$ is transformed to a tree $R'$ still containing the same cherries as $T'$ but in a way that resolving all these cherries of $R'$ results in $R$.
+For proving that this path is shorter than any path preserving the clusters $\{1, \ldots, n\}$ and $\{n+1, \ldots, 2n\}$, the fact that $d_{\nni}(R,R') = d_{\nni}(T|_{\{1, \ldots, n\}}, R|_{\{1, \ldots, n\}})$ where $T|_{\{1, \ldots, n\}}$ and $R|_{\{1, \ldots, n\}}$ are $T$ and $R$ restricted to taxa $\{1,\ldots,n\}$, respectively, is being used.
 However, this does not work in $\rnni$.
 Introducing new cherries means introducing new internal nodes with new ranks.
-Since there are some rank moves needed for an $\rnni$ path similar to the shortest $\nni$ path presented above, the path in $\rnni$ that breaks the split $1, \ldots, n | n+1, \ldots, 2n$ is no shortest path.
+Since there are lots of rank moves needed for an $\rnni$ path similar to the shortest $\nni$ path presented above, the path in $\rnni$ that breaks the clusters $\{1, \ldots, n\}$ and $\{n+1, \ldots, 2n\}$ is no shortest path.
 % How do we assign the ranks to the new cherry nodes?
@@
 
-Therefore, the following conjecture was raised~\cite{Gavryushkin2017}.
+Since the example given in~\cite{Li1996} that proves Theorem~\ref{thm:split_nni} in $\nni$ does not work in $\rnni$, the following conjecture was raised~\cite{Gavryushkin2017}.
 
@@
 
-However, we found a counterexample to this conjecture:
+Instead of using the notion of clusters as we do within this paper, the Split Theorem considers partitions of the taxon set $X$, so-called splits, received by deleting an edge of a tree and taking the taxon sets of the two resulting trees as partition of $X$.
+However, there is a simple counterexample to this conjecture:
 

### 2019-06-26 paper.tex
rm weak split thm, instead: cluster thm

@@
 \todo{define splits and delete definition from split thm chapter!}
+\todo{change definition of clusters: induced by vertices, not edges!}
 
@@
 	\caption{Four trees connected by edges in the $\rnni$ graph. A rank move of $mrca(\{1,2\})$ and $mrca(\{3,4\})$ and all $\nni$ moves on the edge highlighted in red are illustrated}
-	\label{fig:NNI_vs_RNNI}
+	\label{fig:RNNI}
 \end{figure}
@@
 Instead of using the notion of clusters as we do within this paper, the Split Theorem considers partitions of the taxon set $X$, so-called splits, received by deleting an edge of a tree and taking the taxon sets of the two resulting trees as partition of $X$.
-However, there is a simple counterexample to this conjecture:
+However, we found a simple counterexample to this conjecture:
 
@@
 
-Since the counterexample in Figure~\ref{splitthm_counterexample} clearly shows that the version of the Split Theorem stated in~\cite{Gavryushkin2017} does not hold, we will now claim an alternative to this conjecture.
-The alternative formulation is based on the observation that there is a shortest path preserving all splits between the trees of the counterexample of the Split Theorem, which is presented at the bottom of Figure~\ref{splitthm_counterexample}.
+Since the counterexample in Figure~\ref{splitthm_counterexample} clearly shows that the version of the Split Theorem stated in~\cite{Gavryushkin2017} does not hold, we will now claim an alternative to this conjecture, the \emph{Cluster Theorem}.
+Within the Cluster Theorem we are considering clusters instead of splits.
+This is motivated by the fact that rooted phylogenetic trees can uniquely be represented by sets of clusters~\ref{steel2016}, but not by sets of splits as they cannot define the position of the root.
+Moreover, a ranked tree can be uniquely determined by ordering the set of clusters increasingly according to the ranks of the internal nodes inducing those clusters.
 
-\begin{conjecture}[Weak Split Theorem]
-	For $\rnni$ graph the following statement holds:
-	if a partition of leaves given by an edge is present in two trees $T$ and $R$ then there exists a shortest path between $T$ and $R$ where this partition is present in every tree on this path.
-	\label{split_theorem_weak}
-\end{conjecture}
-
-\todo{Do we want to keep the weak version of the split thm? It is stronger than the Cluster Thm!}
-
+% \begin{conjecture}[Weak Split Theorem]
+% 	For $\rnni$ graph the following statement holds:
+% 	if a partition of leaves given by an edge is present in two trees $T$ and $R$ then there exists a shortest path between $T$ and $R$ where this partition is present in every tree on this path.
+% 	\label{split_theorem_weak}
+% \end{conjecture}
+%
+% \todo{Do we want to keep the weak version of the split thm? It is stronger than the Cluster Thm!}
+%
 % The weak Split Theorem (in the version as in Lena's thesis) is computationally proven to hold for trees with up to $6$ taxa ($\sim 38$ min running time on laptop).
-
-Another alternative formulation of the Split Theorem is the \emph{Cluster Theorem}, which considers clusters instead of splits.
-Considering clusters rather than splits is more reasonable since two rooted trees have the same set of clusters if, and only if, they are isomorphic~\cite{steel2016}.
-The trees $T, R$ of counterexample~\ref{splitthm_counterexample} share the same set of induced splits, but have distinct sets of clusters.
 
@@
 \end{conjecture}
+
+Note that the trees $T, R$ of counterexample~\ref{splitthm_counterexample} share the same set of induced splits, but have distinct sets of clusters.
+Furthermore, the example used in $\rnni$ to show that there are trees that share splits that are not present on any shortest path between these trees can be used to prove that the Cluster Theorem does not hold in $\nni$.
 

### 2019-06-26 paper.tex
change definition of cluster (now: induced by node, not by edge)

@@
 
-A \emph{rooted binary phylogenetic tree} on a set $X$ is a pair $(t, \Phi)$ where $t$ is a rooted binary tree with leaf set $\mathcal{L}(t)$ and $\Phi:X \mapsto \mathcal{L}(T)$ a bijective mapping.
+A \emph{rooted binary phylogenetic tree} on a set $X$ is a pair $(t, \Phi)$ where $t$ is a rooted binary tree with leaf set $\mathcal{L}(t)$ and $\Phi:X \to \mathcal{L}(T)$ a bijective mapping.
 \todo{do we need to define leaves and internal nodes, child, parent, subtree, cherry(defined below)?}
@@
 An edge $e$ that has no leaf as child is an \emph{internal edge} and we say an edge $e=(u,v)$ has \emph{length} $l$ if $|rank(v) - rank(u)| = l$.
-Each internal edge defines a \emph{cluster} which is the set of taxa that are descending from the lower of both nodes bounding $e$. \todo{OK?}
-In the literature, clusters often refer to sets of taxa descending from internal nodes in which case clusters as defined above coincide with non-trivial clusters.
+Each internal node $v$ gives a \emph{cluster} $C$ that is the set of taxa descending from this node. \todo{OK?}
+We then say that $v$ induces the cluster $C$.
+Note that all trees on $n$ taxa share the cluster $\{1,\ldots,n\}$ induced by the root.
 Given a subset $S$ of the set of taxa $X$, the \emph{most recent common ancestor} $mrca(S)$ of $S$ is the internal node with lowest rank that is ancestor of all taxa in $S$. \todo{OK?}
@@
 \todo{define splits and delete definition from split thm chapter!}
-\todo{change definition of clusters: induced by vertices, not edges!}
 
 Now we are ready to define the tree space that is the subject of study of this paper, the $\rnni$ graph.
-The vertex set of the \emph{$\rnni$-graph} is given by the set of all ranked trees on $n$ taxa.
+The vertex set of the \emph{$\rnni$ graph} is given by the set of all ranked trees on $n$ taxa.
 We introduce two types of operations (\emph{$\rnni$ moves}) connecting trees $T$ and $R$ in this graph by an edge:
@@
 	\FOR {$i = 1, \dots, n-2$}
-		\STATE Let $C$ be the cluster associated with edge $(v,w)$ in $R$ with $rank(v) = i < rank(w)$
+		\STATE Let $C$ be the cluster node $v$ in $R$ with $rank(v) = i$
 		\WHILE {$rank(mrca_{\hat{T}}(C))>i$}
@@
     Let us consider the maximum length of a path given by Algorithm~\ref{alg:find_path}.
-    In worst case each of the $n-2$ clusters $C$ given by internal edges of $R$, represented by $i = 1, \dots, n-2$ in the algorithm, requires $n-1-i$ $\rnni$ moves for moving $mrca(C)$ down in $\hat{T}$.
+    In worst case each of the $n-2$ clusters $C$ induced by internal nodes of $R$ with ranks $i = 1, \dots, n-2$, requires $n-1-i$ $\rnni$ moves to move $mrca(C)$ down in $\hat{T}$.
 	Hence, the maximum length of a path computed by that algorithm for any pair of trees is less or equal to $\sum\limits_{i = 1}^{n-2} i = \frac{(n-2)(n-1)}{2}$.
@@
 When proving Theorem~\ref{thm:split_nni},~\cite{Li1996} gives an example of two trees $T$ and $R$ on $2n$ taxa that consist of two caterpillar trees each that share the root as parent.
-The clusters given by the two edges incident to the root are $\{1, \ldots, n\}$ and $\{n+1, \ldots, 2n\}$ in both trees $T$ and $R$.
+The clusters given by the two nodes adjacent to the root are $\{1, \ldots, n\}$ and $\{n+1, \ldots, 2n\}$ in both trees $T$ and $R$.
 Between these there is a shortest path in $\nni$ where at first $T$ is transformed to a tree $T'$ with $n$ cherries where each cherry contains one taxon of $\{1, \ldots, n\}$ and one taxon of $\{n+1, \ldots, 2n\}$.
@@
 
-Instead of using the notion of clusters as we do within this paper, the Split Theorem considers partitions of the taxon set $X$, so-called splits, received by deleting an edge of a tree and taking the taxon sets of the two resulting trees as partition of $X$.
+Instead of using the notion of clusters as we do within this paper, the Split Theorem considers splits of the taxon set $X$.
 However, we found a simple counterexample to this conjecture:

### 2019-06-26 paper.tex
define splits + minor

@@
 For simplicity of notation we use the term \emph{tree} instead instead of ranked tree.
+
 Furthermore, we will identify leaves by their labels and might call them \emph{taxa} (singular: \emph{taxon}).
@@
 Note that all trees on $n$ taxa share the cluster $\{1,\ldots,n\}$ induced by the root.
+A \emph{split} is a bipartition of the taxon set $X$ into sets $A,B$ such that there is an edge $e$ whose deletion results in two trees with taxon sets $A$ and $B$, respectively.
+Such a split will be denoted by $A|B$ and we say that it is induced by edge $e$.
 Given a subset $S$ of the set of taxa $X$, the \emph{most recent common ancestor} $mrca(S)$ of $S$ is the internal node with lowest rank that is ancestor of all taxa in $S$. \todo{OK?}
@@
 We write $mrca_T(S)$ to emphasise that we consider the most recent common ancestor of set $S$ in tree $T$.
-Two trees are called \emph{isomorphic} if the underlying rooted phylogenetic trees are isomorphic and the isomorphism preserves ranks of the nodes.
-
-\todo{define splits and delete definition from split thm chapter!}
+Two trees are called \emph{isomorphic} if the underlying rooted phylogenetic trees are isomorphic and the isomorphism preserves ranks of the nodes
 
@@
 
-%Is it sufficient to mention what RNNI stands for in the Introduction?
-
-%Restrict to RNNIu (See whether we can get same results for RNNI)
-
-%Definitions inside a tree: cherry, partition, edges represent partitions
-
 %If needed: diameter, radius (in the corresponding chapter?), induced triplet (in that chapter?)
-
-%delete definitions in the following chapters that are already covered here!
-
-%replace rank swap with rank move in the following!!
 

### 2019-06-26 paper.tex
minor

@@
 %We use the exact algorithm to calculate all the distances for the examples given in this paper
+
+%Figure of ranked tree in introduction
 
@@
 
-The following Algorithm~\ref{alg:find_path} which computes paths between pairs of trees in $\rnni$ is a central tool for our investigation of shortest paths in $\rnni$.
+Algorithm~\ref{alg:find_path} computes paths between pairs of trees in $\rnni$ and will be a central tool for our investigation of shortest paths in $\rnni$.
 
@@
 
-\todo{figure of an example?}
+\todo{figure of an example? Brief explanation what algorithm does?}
 
@@
 This makes sure that there is no additional swap of $a$ and $b$ needed on the path to $R$.
+\todo{Pseudo-Code for algorithm?}
 
@@
 
-In~\cite{Gavryushkin2017} upper and lower bound for the diameter of $\rnni$ are given.
 Following the results of section~\ref{section:caterpillar_convex} one can directly improve this result by giving the exact diameter of the $\rnni$ graph.
@@
 Within the Cluster Theorem we are considering clusters instead of splits.
-This is motivated by the fact that rooted phylogenetic trees can uniquely be represented by sets of clusters~\ref{steel2016}, but not by sets of splits as they cannot define the position of the root.
+This is motivated by the fact that rooted phylogenetic trees can uniquely be represented by sets of clusters~\cite{steel2016}, but not by sets of splits as they cannot define the position of the root.
 Moreover, a ranked tree can be uniquely determined by ordering the set of clusters increasingly according to the ranks of the internal nodes inducing those clusters.
@@
     Let $T$ and $R$ be caterpillar trees.
-    It is $d_c(T,R) = \Delta(\rnniu)$ if, and only if, $T$ and $R$ share no induced triplet.
+    It is $d_c(T,R) = \Delta(\rnni)$ if, and only if, $T$ and $R$ share no induced triplet.
     \todo{define induced triplets}
@@
     Let $T$ and $R$ be trees on $n$ taxa.
-    It is $d(T,R) = \Delta(\rnniu)$ if, and only if, $R$ can be produced from $T$ by applying Algorithm~\ref{alg:max_dist_tree}.
+    It is $d(T,R) = \Delta(\rnni)$ if, and only if, $R$ can be produced from $T$ by applying Algorithm~\ref{alg:max_dist_tree}.
 \end{conjecture}
+
+If this conjecture is true, we have that $\Delta(\rnni) = rad(\rnni)$.
 
@@
 For caterpillar trees it is $(n-1)!$.
-The number of caterpillar trees with distance $\Delta(\rnniu)$ from a given caterpillar tree is $2^{n-2}$.
+The number of caterpillar trees with distance $\Delta(\rnni)$ from a given caterpillar tree is $2^{n-2}$.
 

### 2019-06-27 paper.tex
some housekeeping

@@
-\documentclass[11pt, a4paper]{article}
+\documentclass{amsart}
 
@@
 \usepackage{enumitem}
+\usepackage[foot]{amsaddr}
+\usepackage[hidelinks]{hyperref}
+\usepackage[style=authoryear,ibidtracker=false,uniquename=false,giveninits=true,terseinits=true,maxbibnames=5,backend=biber]{biblatex}
+\renewbibmacro{in:}{}
+\addbibresource{lit.bib}
 
@@
 \date{\today}
-\author{Lena Collienne, Alex Gavryushkin}
+\author{Lena Collienne}
+\email{lena.collienne@postgrad.otago.ac.nz}
+\author{Alex Gavryushkin}
+\email{alex@biods.org}
+\address{Department of Computer Science, University of Otago, New Zealand}
 
@@
 
+Although we endeavour to make this paper self-sufficient, we refer the reader to standard textbooks in phylogenetics for terminology \autocite{Felsenstein2004-of, Semple2003-nj, Steel2016-ye}.
+Unless is explicit, we follow these textbooks in our terminology.
+
 \section{Definitions}
-
-\todo{citation of definitions??}
 
@@
 We write $mrca_T(S)$ to emphasise that we consider the most recent common ancestor of set $S$ in tree $T$.
-Two trees are called \emph{isomorphic} if the underlying rooted phylogenetic trees are isomorphic and the isomorphism preserves ranks of the nodes
+Two trees are called \emph{isomorphic} if the underlying rooted phylogenetic trees are isomorphic and the isomorphism preserves ranks of the nodes.
 
@@
 It follows that Algorithm~\ref{alg:find_path} returns a path in $\rnni$ for any pair of trees $T,R$.
-Moreover, it gives an upper bound for the diameter of $\rnni$, which improves the one given in~\cite{Gavryushkin2017}:
+Moreover, it gives an upper bound for the diameter of $\rnni$, which improves the one given in \autocite{Gavryushkin2017}:
 
@@
 $\nni$ moves that change the cherry of a tree refer to exchanges of the first or the second element with the third element of the sequence while exchanges of the first two elements do not change the corresponding tree.
-With this observation we can provide an algorithm for computing shortest caterpillar paths by a simple modification of the commonly known \emph{Bubble Sort} algorithm~\cite{knuth1997}:
+With this observation we can provide an algorithm for computing shortest caterpillar paths by a simple modification of the commonly known \emph{Bubble Sort} algorithm \autocite{knuth1997}:
 
@@
 It is natural to try to relate the $\rnni$ graph to the $\nni$ as we already did in previous sections.
-Since computing distances in tree spaces is a main issue in most commonly known tree spaces as $\nni$~\cite{jiang2000}, $\rspr$~\cite{Bordewich2005}, or $\tbr$~\cite{allen2001subtree} is $\np$-hard, it would be desirable to find a tree space that is less complex.
+Since computing distances in tree spaces is a main issue in most commonly known tree spaces as $\nni$ \autocite{jiang2000}, $\rspr$ \autocite{Bordewich2005}, or $\tbr$ \autocite{allen2001subtree} is $\np$-hard, it would be desirable to find a tree space that is less complex.
 However, it is not known whether computing distances in $\rnni$ is $\np$-hard or not.
@@
 
-The proof of NP-completeness of $\nni$ in~\cite{jiang2000} is based on the fact that following theorem holds in $\nni$.
+The proof of NP-completeness of $\nni$ in \autocite{jiang2000} is based on the fact that following theorem holds in $\nni$.
 
@@
 \begin{proof}
-	See~\cite{Li1996}.
+	See \autocite{Li1996}.
 \end{proof}
 
-The proof presented in~\cite{Li1996} for Theorem~\ref{thm:split_nni} does not work for $\rnni$.
+The proof presented in \autocite{Li1996} for Theorem~\ref{thm:split_nni} does not work for $\rnni$.
 
@@
 \todo{Shall we keep this explanation? The problem is that the $\rnni$ path depends on the assignment of ranks to the cherries produced on the path}
-The counterexample presented in~\cite{Li1996} works in $\nni$ because of the following:
-When proving Theorem~\ref{thm:split_nni},~\cite{Li1996} gives an example of two trees $T$ and $R$ on $2n$ taxa that consist of two caterpillar trees each that share the root as parent.
+The counterexample presented in \autocite{Li1996} works in $\nni$ because of the following:
+When proving Theorem~\ref{thm:split_nni}, \autocite{Li1996} gives an example of two trees $T$ and $R$ on $2n$ taxa that consist of two caterpillar trees each that share the root as parent.
 The clusters given by the two nodes adjacent to the root are $\{1, \ldots, n\}$ and $\{n+1, \ldots, 2n\}$ in both trees $T$ and $R$.
@@
 
-Since the example given in~\cite{Li1996} that proves Theorem~\ref{thm:split_nni} in $\nni$ does not work in $\rnni$, the following conjecture was raised~\cite{Gavryushkin2017}.
+Since the example given in \autocite{Li1996} that proves Theorem~\ref{thm:split_nni} in $\nni$ does not work in $\rnni$, the following conjecture was raised \autocite{Gavryushkin2017}.
 
@@
 
-Since the counterexample in Figure~\ref{splitthm_counterexample} clearly shows that the version of the Split Theorem stated in~\cite{Gavryushkin2017} does not hold, we will now claim an alternative to this conjecture, the \emph{Cluster Theorem}.
+Since the counterexample in Figure~\ref{splitthm_counterexample} clearly shows that the version of the Split Theorem stated in \autocite{Gavryushkin2017} does not hold, we will now claim an alternative to this conjecture, the \emph{Cluster Theorem}.
 Within the Cluster Theorem we are considering clusters instead of splits.
-This is motivated by the fact that rooted phylogenetic trees can uniquely be represented by sets of clusters~\cite{steel2016}, but not by sets of splits as they cannot define the position of the root.
+This is motivated by the fact that rooted phylogenetic trees can uniquely be represented by sets of clusters \autocite{steel2016}, but not by sets of splits as they cannot define the position of the root.
 Moreover, a ranked tree can be uniquely determined by ordering the set of clusters increasingly according to the ranks of the internal nodes inducing those clusters.
@@
 
-\bibliographystyle{abbrv}
-\bibliography{lit}
+\printbibliography
 

### 2019-06-27 paper.tex
@leancoll please make sure trees are denoted by capitall letters throughout

@@
 
-\section{Introduction}
+% All papers have this section so no need for this heading -- let's wait until a journal pushes us to put it back :)
+%\section{Introduction}
 
@@
 
-A \emph{rooted binary phylogenetic tree} on a set $X$ is a pair $(t, \Phi)$ where $t$ is a rooted binary tree with leaf set $\mathcal{L}(t)$ and $\Phi:X \to \mathcal{L}(T)$ a bijective mapping.
-\todo{do we need to define leaves and internal nodes, child, parent, subtree, cherry(defined below)?}
+A \emph{rooted binary phylogenetic tree} on a set $X$ is a pair $(T, \Phi)$ where $T$ is a rooted binary tree with leaf set $\mathcal{L}(T)$ and $\Phi:X \to \mathcal{L}(T)$ a bijective mapping.
+\todo{do we need to define leaves and internal nodes, child, parent, subtree, cherry(defined below)?
+AG: Only if we use those terms.
+Can we use something like a glossary to automatically know what terminology we need to introduce?}
 From now on, we assume $X = \{1,\ldots,n\}$ for a fixed number $n \in \mathbb N$ of leaves.
-Within this paper we study \emph{ranked trees} that are determined by a pair consisting of a rooted binary phylogenetic tree $(t, \Phi)$ on $X = \{1, \ldots, n\}$ for $n \in \mathbb N$, and a rank function.
-A \emph{rank function} on $t$ maps the vertex set of $t$ onto the set $\{0,\ldots,n-1\}$ such that all leaves are assigned rank $0$ and each internal node has rank greater than its children.
+
+In this paper we study \emph{ranked trees} that are determined by a pair consisting of a rooted binary phylogenetic tree $(T, \Phi)$ on $X = \{1, \ldots, n\}$ for $n \in \mathbb N$, and a rank function.
+A \emph{rank function} on $T$ maps the vertex set of $T$ onto the set $\{0,\ldots,n-1\}$ such that all leaves are assigned rank $0$ and each internal node has rank greater than its children.
 The rank of a node $v$ in a ranked tree will be denoted by $rank(v)$ or $rank_T(v)$ if it is not obvious that we consider tree $T$.

### 2019-06-27 paper.tex
@lenacoll please make \mrca \parent etc. to use \mathrm or similar, throughout the paper

@@
 Furthermore, we will identify leaves by their labels and might call them \emph{taxa} (singular: \emph{taxon}).
-For a tree $T$ a subtree consisting of an internal node with both children being leaves is called \emph{cherry}.
+For a tree $T$ a subtree consisting of an internal node with both children being leaves is called a \emph{cherry}.
 An edge $e$ that has no leaf as child is an \emph{internal edge} and we say an edge $e=(u,v)$ has \emph{length} $l$ if $|rank(v) - rank(u)| = l$.
@@
 Given a subset $S$ of the set of taxa $X$, the \emph{most recent common ancestor} $mrca(S)$ of $S$ is the internal node with lowest rank that is ancestor of all taxa in $S$. \todo{OK?}
-In particular, for a single taxon $x \in X$, $mrca(\{x\})$ is the parent of $x$ and will be denoted by $parent(x)$.
-We write $mrca_T(S)$ to emphasise that we consider the most recent common ancestor of set $S$ in tree $T$.
+In particular, for a single taxon $x \in X$, $\mrca(\{x\})$ is the parent of $x$ and will be denoted by $\parent(x)$.
+We write $\mrca_T(S)$ to emphasise that we consider the most recent common ancestor of set $S$ in tree $T$.
 Two trees are called \emph{isomorphic} if the underlying rooted phylogenetic trees are isomorphic and the isomorphism preserves ranks of the nodes.

### 2019-06-27 paper.tex
authors + some definitions trim

@@
 \usepackage{enumitem}
+\usepackage[misc]{ifsym}
 \usepackage[foot]{amsaddr}
@@
 \date{\today}
-\author{Lena Collienne}
+\author{Lena Collienne\textsuperscript{1}}
 \email{lena.collienne@postgrad.otago.ac.nz}
-\author{Alex Gavryushkin}
+\address{\textsuperscript{1}Department of Computer Science, University of Otago, New Zealand}
+\author{Mareike Fischer\textsuperscript{2}}
+\email{email@mareikefischer.de}
+\address{\textsuperscript{2}Institute for Mathematics and Inofrmatics, University of Greifswald, Germany}
+\author{Alex Gavryushkin\textsuperscript{1, \Letter}}
 \email{alex@biods.org}
-\address{Department of Computer Science, University of Otago, New Zealand}
 
@@
 
-Furthermore, we will identify leaves by their labels and might call them \emph{taxa} (singular: \emph{taxon}).
-For a tree $T$ a subtree consisting of an internal node with both children being leaves is called a \emph{cherry}.
-An edge $e$ that has no leaf as child is an \emph{internal edge} and we say an edge $e=(u,v)$ has \emph{length} $l$ if $|rank(v) - rank(u)| = l$.
-Each internal node $v$ gives a \emph{cluster} $C$ that is the set of taxa descending from this node. \todo{OK?}
-We then say that $v$ induces the cluster $C$.
+We will identify leaves with their labels and also call them \emph{taxa} (singular: \emph{taxon}) following the systematic biology tradition.
+For a tree $T$, a subtree consisting of an internal node with both children being leaves is called a \emph{cherry}.
+An edge $e$ that has no leaf as a child is called an \emph{internal edge} and we say an edge $e=(u,v)$ has \emph{length} $l$ if $|rank(v) - rank(u)| = l$.
+Each internal node $v$ can be uniquely identified by its \emph{cluster} $C$, which is the set of taxa descending from the node.
 Note that all trees on $n$ taxa share the cluster $\{1,\ldots,n\}$ induced by the root.

### 2019-06-27 paper.tex
rm some definitions that are too standard

@@
 For a tree $T$, a subtree consisting of an internal node with both children being leaves is called a \emph{cherry}.
-An edge $e$ that has no leaf as a child is called an \emph{internal edge} and we say an edge $e=(u,v)$ has \emph{length} $l$ if $|rank(v) - rank(u)| = l$.
+An edge $e$ that has no leaf as a child is called an \emph{internal edge} and we say an edge $e=(u,v)$ has \emph{length} $l$ if $|\rank(v) - \rank(u)| = l$.
 Each internal node $v$ can be uniquely identified by its \emph{cluster} $C$, which is the set of taxa descending from the node.
+We also say that $v$ \emph{induces} the cluster $C$.
 Note that all trees on $n$ taxa share the cluster $\{1,\ldots,n\}$ induced by the root.
 A \emph{split} is a bipartition of the taxon set $X$ into sets $A,B$ such that there is an edge $e$ whose deletion results in two trees with taxon sets $A$ and $B$, respectively.
-Such a split will be denoted by $A|B$ and we say that it is induced by edge $e$.
-Given a subset $S$ of the set of taxa $X$, the \emph{most recent common ancestor} $mrca(S)$ of $S$ is the internal node with lowest rank that is ancestor of all taxa in $S$. \todo{OK?}
+Such a split will be denoted by $A|B$ and we say that it is \emph{induced} by edge $e$.
+Given a subset $S$ of the set of taxa $X$, the \emph{most recent common ancestor} of $S$, denoted by $\mrca(S)$, is the internal node of the lowest rank that is ancestral to all the taxa in $S$.
 In particular, for a single taxon $x \in X$, $\mrca(\{x\})$ is the parent of $x$ and will be denoted by $\parent(x)$.
@@
 
-Now we are ready to define the tree space that is the subject of study of this paper, the $\rnni$ graph.
+Now we are ready to define the tree space which is the subject of study in this paper, the $\rnni$ graph.
 The vertex set of the \emph{$\rnni$ graph} is given by the set of all ranked trees on $n$ taxa.
 We introduce two types of operations (\emph{$\rnni$ moves}) connecting trees $T$ and $R$ in this graph by an edge:
-The swap of the ranks of two internal nodes that have rank difference one is called \emph{rank move}.
-We say that there is an $\nni$ move between trees $T$ and $R$ if there are edges $e$ in $T$ and $f$ in $R$ that are bounded by internal nodes with rank difference one such that the graphs obtained by shrinking $e$ and $f$ to internal nodes are isomorphic.
-Trees that are connected by either a rank move or an $\nni$ move are called \emph{$\rnni$ neighbours} or briefly neighbours.
-Note that the previously defined $\rnni$ graph is a metric space, which is why we may call it \emph{$\rnni$ space} or just $\rnni$.
+A swap of the ranks of two internal nodes ($x$ and $y$) of subsequent ranks (|\rank(x) - \rank(y)| = 1) is called a \emph{rank move}.
+We say that trees $T$ and $R$ are \emph{connected} by an \emph{$\nni$ move} move if there exist internal edges $e$ in $T$ and $f$ in $R$ both of length one such that the graphs obtained by shrinking $e$ and $f$ to internal nodes are isomorphic.
+Trees that are connected by either a rank move or an $\nni$ move are called \emph{$\rnni$ neighbours} or \emph{adjacent in} $\rnni$.
+Note that thus defined $\rnni$ graph is a metric space with the standard shortest-path distance.
 
@@
 
-The main purpose of this paper is understanding geometric properties of the $\rnni$ space.
-Therefore, further notations need to be introduced.
-If every internal node of a tree $T$ has at least one child that is a leaf, $T$ is called \emph{caterpillar tree}.
-A \emph{path} $p$ between trees $T_0$ and $T_m$ in $\rnni$ is a sequence $T_0, T, \ldots, T_m$ of trees such that there is an edge $\{T_i, T_{i+1}\}$ for all $i=0,\ldots,m-1$.
-For such a path $p$ we say the \emph{length} $|p|$ of $p$ is $m$.
-A path $p$ between trees $T$ and $R$ is a \emph{shortest path} if for every path $r$ between $T$ and $R$, $|p| \leq |r|$.
-The \emph{distance} $d(T,R)$ between trees $T$ and $R$ is the length of a shortest path between them.
+The following definitions are less standard and required for the results obtained in this paper.
+If every internal node of a tree $T$ has at least one child that is a leaf, $T$ is called a \emph{caterpillar tree}.
 If $p$ is a path only consisting of caterpillar trees we call it \emph{caterpillar path}.
@@
 \section{Geometry of $\rnni$}
+
+This paper is an advance towards the general goal of understanding the geometry of the $\rnni$ space.
 

### 2019-06-27 paper.tex
a few general comments

@@
 This paper is an advance towards the general goal of understanding the geometry of the $\rnni$ space.
+%TODO
+In this section, we...
 
-Algorithm~\ref{alg:find_path} computes paths between pairs of trees in $\rnni$ and will be a central tool for our investigation of shortest paths in $\rnni$.
+We now present an efficient algorithm (Algorithm~\ref{alg:find_path}) for computing paths between pairs of trees in $\rnni$.
+This algorithm will be a central tool for our investigation of shortest paths in $\rnni$.
+
+%TODO
+The idea of the algorithm is the following...
 
@@
 
-\todo{figure of an example? Brief explanation what algorithm does?}
-
+\todo{This text comes out of the blue.
+Generally, before going into explanations like this it's a good idea to say what's ahead.}
 One can easily prove that there is always exactly one $\rnni$ move on $T$ that decreases the rank of $mrca_T(C)$ as in line~\ref{alg:line:move_set_down} of Algorithm~\ref{alg:find_path}.
 It follows that Algorithm~\ref{alg:find_path} returns a path in $\rnni$ for any pair of trees $T,R$.
+\todo{This should have been stated before, where you say what is going on.}
 Moreover, it gives an upper bound for the diameter of $\rnni$, which improves the one given in \autocite{Gavryushkin2017}:
@@
 \begin{lemma}
-    It is $\Delta(\rnni) \leq \frac{(n-1)(n-2)}{2}$.
+%TODO: here and generally, I prefer introducing definitions like this as you go but it's the matter of preference so up to you.
+$\Delta(\rnni) \leq \frac{(n-1)(n-2)}{2}$, where $\Delta$ is...
 \end{lemma}

### 2019-06-28 paper.tex
edit proof: bounding length of path when deleting taxon

@@
 \begin{proof}
-    On every path from $T$ to $R$ there are at least $|rank_T(parent(n+1)) - rank_R(parent(n+1))|$ $\rnni$ moves that change the rank of $parent(n+1)$ because each $\rnni$ move can change the rank of an internal node by at most one.
-    In particular, this holds for a shortest path $p$ from $T$ to $R$.
-    Furthermore, deleting taxon $n+1$ and its parent $p_{n+1}$ from all trees on $p$ results in a path $p|_n$ from $T|_n$ to $R|_n$.
-    All moves that do not involve $p_{n+1}$ can be performed the same way and those that involved $p_{n+1}$ are just left out.
-    We thus have a path $p|_n$ from $T|_n$ to $R|_n$ that has at least $|rank_T(parent(n+1)) - rank_R(parent(n+1))|$ $\rnni$ moves less than $p$, which results in $|p|_n| \leq d(T,R) - |rank_T(parent(n+1)) - rank_R(parent(n+1))|$.
-    \todo{Is this proof sufficient? Otherwise: induction}
+    Let $p$ be a shortest path from tree $T$ to $R$ and $p|_n$ the path resulting from deleting taxon $n+1$ from all trees on $p$.
+    This new path $p|_n$ is then a path from $T$ to $R$.
+    Showing that at least $\delta := |rank_T(parent(n+1)) - rank_R(parent(n+1))|$ $\rnni$ moves are deleted from $p$ to receive $p|_n$ results in $d(T|_n,R|_n) \leq |p|_n| \leq |p| - \delta = d(T,R) -\delta$ which proves the lemma.
+    By the following two observations one can easily see that this claim is true, indeed:
+    The rank of an internal node $v$ only changes within an $\rnni$ move if this move is a rank move of $v$ with another node or an $\nni$ move on an edge incident to $v$.
+    Furthermore, any $\rnni$ move can change the rank of an internal node by at most one.
+    Applying this to $parent(n+1)$ and knowing that the rank of this node differs by $\delta$ between $T$ and $R$ proves that $d(T|_n, R|_n) \leq d(T,R) - \delta$.
+
+    % Alternative: Induction:
+    %
     % Induction on $k = d(T,R)$:

### 2019-06-28 paper.tex
\mrca, \rank, \parent

@@
 \newcommand{\np}{\mathcal{NP}}
+
+\newcommand{\parent}{\mathrm{parent}}
+\newcommand{\mrca}{\mathrm{mrca}}
+\newcommand{\rank}{\mathrm{rank}}
 \newcommand{\nni}{\mathrm{NNI}}
@@
 A \emph{rank function} on $T$ maps the vertex set of $T$ onto the set $\{0,\ldots,n-1\}$ such that all leaves are assigned rank $0$ and each internal node has rank greater than its children.
-The rank of a node $v$ in a ranked tree will be denoted by $rank(v)$ or $rank_T(v)$ if it is not obvious that we consider tree $T$.
+The rank of a node $v$ in a ranked tree will be denoted by $\rank(v)$ or $\rank_T(v)$ if it is not obvious that we consider tree $T$.
 For simplicity of notation we use the term \emph{tree} instead instead of ranked tree.
@@
 Given a subset $S$ of the set of taxa $X$, the \emph{most recent common ancestor} of $S$, denoted by $\mrca(S)$, is the internal node of the lowest rank that is ancestral to all the taxa in $S$.
-In particular, for a single taxon $x \in X$, $\mrca(\{x\})$ is the parent of $x$ and will be denoted by $\parent(x)$.
+In particular, for a single taxon $x \in X$, $\mrca(\{x\})$ is the \emph{parent} of $x$ and will be denoted by $\parent(x)$.
 We write $\mrca_T(S)$ to emphasise that we consider the most recent common ancestor of set $S$ in tree $T$.
@@
 	\includegraphics[width=\textwidth]{RNNI}
-	\caption{Four trees connected by edges in the $\rnni$ graph. A rank move of $mrca(\{1,2\})$ and $mrca(\{3,4\})$ and all $\nni$ moves on the edge highlighted in red are illustrated}
+	\caption{Four trees connected by edges in the $\rnni$ graph. A rank move of $\mrca(\{1,2\})$ and $\mrca(\{3,4\})$ and all $\nni$ moves on the edge highlighted in red are illustrated}
 	\label{fig:RNNI}
@@
 	\FOR {$i = 1, \dots, n-2$}
-		\STATE Let $C$ be the cluster node $v$ in $R$ with $rank(v) = i$
-		\WHILE {$rank(mrca_{\hat{T}}(C))>i$}
-			\STATE Update $\hat{T}$: Decrease $rank(mrca_{\hat{T}}(C))$ by an $\rnni$ move \label{alg:line:move_set_down}
+		\STATE Let $C$ be the cluster node $v$ in $R$ with $\rank(v) = i$
+		\WHILE {$\rank(\mrca_{\hat{T}}(C))>i$}
+			\STATE Update $\hat{T}$: Decrease $\rank(\mrca_{\hat{T}}(C))$ by an $\rnni$ move \label{alg:line:move_set_down}
 			\STATE $p = p+\hat{T}$
@@
 Generally, before going into explanations like this it's a good idea to say what's ahead.}
-One can easily prove that there is always exactly one $\rnni$ move on $T$ that decreases the rank of $mrca_T(C)$ as in line~\ref{alg:line:move_set_down} of Algorithm~\ref{alg:find_path}.
+One can easily prove that there is always exactly one $\rnni$ move on $T$ that decreases the rank of $\mrca_T(C)$ as in line~\ref{alg:line:move_set_down} of Algorithm~\ref{alg:find_path}.
 It follows that Algorithm~\ref{alg:find_path} returns a path in $\rnni$ for any pair of trees $T,R$.
@@
     Let us consider the maximum length of a path given by Algorithm~\ref{alg:find_path}.
-    In worst case each of the $n-2$ clusters $C$ induced by internal nodes of $R$ with ranks $i = 1, \dots, n-2$, requires $n-1-i$ $\rnni$ moves to move $mrca(C)$ down in $\hat{T}$.
+    In worst case each of the $n-2$ clusters $C$ induced by internal nodes of $R$ with ranks $i = 1, \dots, n-2$, requires $n-1-i$ $\rnni$ moves to move $\mrca(C)$ down in $\hat{T}$.
 	Hence, the maximum length of a path computed by that algorithm for any pair of trees is less or equal to $\sum\limits_{i = 1}^{n-2} i = \frac{(n-2)(n-1)}{2}$.
@@
     Let $T|_n$ and $R|_n$ be these trees restricted to $n$ taxa, i.e. the same taxon $n+1$ is deleted from both trees $T$ and $R$.
-    Then it is $d(T|_n, R|_n) \leq d(T,R) - |rank_T(parent(n+1)) - rank_R(parent(n+1))|$.
+    Then it is $d(T|_n, R|_n) \leq d(T,R) - |\rank_T(\parent(n+1)) - \rank_R(\parent(n+1))|$.
     \label{lemma:distance_delete_taxon}
@@
     This new path $p|_n$ is then a path from $T$ to $R$.
-    Showing that at least $\delta := |rank_T(parent(n+1)) - rank_R(parent(n+1))|$ $\rnni$ moves are deleted from $p$ to receive $p|_n$ results in $d(T|_n,R|_n) \leq |p|_n| \leq |p| - \delta = d(T,R) -\delta$ which proves the lemma.
+    Showing that at least $\delta := |\rank_T(\parent(n+1)) - \rank_R(\parent(n+1))|$ $\rnni$ moves are deleted from $p$ to receive $p|_n$ results in $d(T|_n,R|_n) \leq |p|_n| \leq |p| - \delta = d(T,R) -\delta$ which proves the lemma.
     By the following two observations one can easily see that this claim is true, indeed:
@@
     Furthermore, any $\rnni$ move can change the rank of an internal node by at most one.
-    Applying this to $parent(n+1)$ and knowing that the rank of this node differs by $\delta$ between $T$ and $R$ proves that $d(T|_n, R|_n) \leq d(T,R) - \delta$.
+    Applying this to $\parent(n+1)$ and knowing that the rank of this node differs by $\delta$ between $T$ and $R$ proves that $d(T|_n, R|_n) \leq d(T,R) - \delta$.
 
@@
     % Basis($k = 1$):
-    % If rank of $parent(n+1)$ changes between $T$ and $R$, it is $d(T|_n, R|_n) = 0$ and $|rank_T(parent(n+1)) - rank_R(parent(n+1))| = 1$ and therefore, $d(T|_n, R|_n) = 0 \leq 1 - 1 = d(T,R) - |rank_T(parent(n+1)) - rank_R(parent(n+1))|$.
-    % If on the other hand the rank of $parent(n+1}$ does not change between $T$ and $R$, it is $d(T|_n, R|_n) = d(T,R) = 1$ and $|r_{T}(parent(n+1}) - r_{R}(parent(n+1})| = 1$ and therefore, $d(T|_n, R|_n) = 1 \leq 1 - 0 = d(T,R) - |r_{T}(parent(n+1}) - r_{R}(parent(n+1})|$.
+    % If rank of $\parent(n+1)$ changes between $T$ and $R$, it is $d(T|_n, R|_n) = 0$ and $|\rank_T(\parent(n+1)) - \rank_R(\parent(n+1))| = 1$ and therefore, $d(T|_n, R|_n) = 0 \leq 1 - 1 = d(T,R) - |\rank_T(\parent(n+1)) - \rank_R(\parent(n+1))|$.
+    % If on the other hand the rank of $\parent(n+1}$ does not change between $T$ and $R$, it is $d(T|_n, R|_n) = d(T,R) = 1$ and $|r_{T}(\parent(n+1}) - r_{R}(\parent(n+1})| = 1$ and therefore, $d(T|_n, R|_n) = 1 \leq 1 - 0 = d(T,R) - |r_{T}(\parent(n+1}) - r_{R}(\parent(n+1})|$.
     %
@@
     % Let $R'$ be the last tree on a shortest path from $T$ to $R$, i.e. $d(T,R') = k$.
-    % We can apply the induction hypothesis on $T$ and $R'$ and see that $d(T|_n, R'|_n) \leq d(T, R') - |r_{T}(parent(n+1}) - r_{R'}(parent(n+1})| := k - m$.
+    % We can apply the induction hypothesis on $T$ and $R'$ and see that $d(T|_n, R'|_n) \leq d(T, R') - |r_{T}(\parent(n+1}) - r_{R'}(\parent(n+1})| := k - m$.
     % Now we differ two cases:
     % \begin{enumerate}
-    %     \item $parent(n+1}$ changes its rank between $R$ and $R'$ or is affected by the $\nni$ move is performed between $R$ and $R'$.
+    %     \item $\parent(n+1}$ changes its rank between $R$ and $R'$ or is affected by the $\nni$ move is performed between $R$ and $R'$.
     %
@@
     %     Therefore, it is $d(T|_n,R|_n) = d(T|_n, R'|_n) \leq k - m$.
-    %     It is either $|r_{T}(parent(n+1}) - r_{R}(parent(n+1})| = m+1$, $|r_{T}(parent(n+1}) - r_{R}(parent(n+1})| = m$, or $|r_{T}(parent(n+1}) - r_{R}(parent(n+1})| = m-1$.
-    %     In either case, we have $d(T|_n,R|_n) = d(T|_n, R'|_n) \leq k - m = (k+1) - (m+1) \leq d(T, R) - |r_{T}(parent(n+1}) - r_{R}(parent(n+1})|$.
+    %     It is either $|r_{T}(\parent(n+1}) - r_{R}(\parent(n+1})| = m+1$, $|r_{T}(\parent(n+1}) - r_{R}(\parent(n+1})| = m$, or $|r_{T}(\parent(n+1}) - r_{R}(\parent(n+1})| = m-1$.
+    %     In either case, we have $d(T|_n,R|_n) = d(T|_n, R'|_n) \leq k - m = (k+1) - (m+1) \leq d(T, R) - |r_{T}(\parent(n+1}) - r_{R}(\parent(n+1})|$.
     %
-    %     \item $parent(n+1}$ has same rank in $R$ and $R'$.
+    %     \item $\parent(n+1}$ has same rank in $R$ and $R'$.
     %
     %     In this case it is $d(T|_n,R|_n) \leq d(T|_n,R'|_n) + d(R'|_n, R|_n) \leq d(T|_n, R'|_n) + 1 \leq (k-m)+ 1$.
-    %     It also follows that $|r_{T}(parent(n+1}) - r_{R'}(parent(n+1})| = |r_{T}(parent(n+1}) - r_{R}(parent(n+1})|$ and therefore, we can conclude $d(T|_n,R|_n) \leq k-m+1 = k+1-m = d(T,R) -  |r_{T}(parent(n+1}) - r_{R}(parent(n+1})|$.
+    %     It also follows that $|r_{T}(\parent(n+1}) - r_{R'}(\parent(n+1})| = |r_{T}(\parent(n+1}) - r_{R}(\parent(n+1})|$ and therefore, we can conclude $d(T|_n,R|_n) \leq k-m+1 = k+1-m = d(T,R) -  |r_{T}(\parent(n+1}) - r_{R}(\parent(n+1})|$.
     % \end{enumerate}
@@
 Let us consider the first three elements $a,b,c$ which appear in this order in a sequence representing a caterpillar tree on a caterpillar path between trees $T$ and $R$.
-If the rank of $parent(c)$ is higher than the ranks of $parent(a)$ and $parent(b)$ in $R$, there are no swaps needed for these taxa.
+If the rank of $\parent(c)$ is higher than the ranks of $\parent(a)$ and $\parent(b)$ in $R$, there are no swaps needed for these taxa.
 If just one of the two taxa $a$ and $b$ is on the other side of $c$ in $R$, this one swaps position with $c$ in the sequence.
@@
 
-    According to Lemma~\ref{lemma:distance_delete_taxon}, the distance between $T|_n$ and $R|_n$ is at least $d(T|_n, R|_n) \leq d(T,R) - |rank_T(parent(n+1)) - rank_R(parent(n+1))|$.
-    We also know $|rank_T(parent(n+1)) - rank_R(parent(n+1))| = n-1$
+    According to Lemma~\ref{lemma:distance_delete_taxon}, the distance between $T|_n$ and $R|_n$ is at least $d(T|_n, R|_n) \leq d(T,R) - |\rank_T(\parent(n+1)) - \rank_R(\parent(n+1))|$.
+    We also know $|\rank_T(\parent(n+1)) - \rank_R(\parent(n+1))| = n-1$
     Suppose, contrary to our claim, that there is a path from $T$ to $R$ of length less than $\Delta(\rnniu_{n+1}) = \frac{n(n-1)}{2}$.
-    By the above, it follows that $d(T|_n, R|_n) \leq d(T,R) - |rank_T(parent(n+1)) - rank_R(parent(n+1))| < \frac{n(n-1)}{2} - (n-1) < \frac{(n-1)(n-2)}{2}$.
+    By the above, it follows that $d(T|_n, R|_n) \leq d(T,R) - |\rank_T(\parent(n+1)) - \rank_R(\parent(n+1))| < \frac{n(n-1)}{2} - (n-1) < \frac{(n-1)(n-2)}{2}$.
     Since this is a contradiction to $d(T|_n, R|_n) = \frac{(n-1)(n-2)}{2}$, it is proven that $d(T,R) = \frac{n(n-1)}{2} $.
@@
 % 	\textbf{NNI move:}
-% 	Let there be an $\nni$ move on an edge $(int_x, int_y)$ where $rank(int_x) < rank(int_y)$ as in Figure~\ref{fig:nni_moves}.
+% 	Let there be an $\nni$ move on an edge $(int_x, int_y)$ where $\rank(int_x) < \rank(int_y)$ as in Figure~\ref{fig:nni_moves}.
 %
@@
 % 	By default, the node that has lower rank than the other one is labelled by $int_x$, the other one by $int_y$.
-% 	But if $int_x$ is no ancestor of $x$ or $int_y$ is no ancestor of $y$, we swap the labels of the two new nodes, such that $rank(int_x) > rank(int_y)$.
+% 	But if $int_x$ is no ancestor of $x$ or $int_y$ is no ancestor of $y$, we swap the labels of the two new nodes, such that $\rank(int_x) > \rank(int_y)$.
 % 	An example of an $\nni$ move and the new labelling is provided in Figure~\ref{fig:nni_example_internal_label}.
@@
 % 	We now turn to the case that $x$ is not in the subtree rooted in the lower node of $e$ after the $\nni$ move.
-% 	Then the labels of the two internal nodes change such that $rank(int_x) > rank(int_y)$.
+% 	Then the labels of the two internal nodes change such that $\rank(int_x) > \rank(int_y)$.
 % 	This makes sure that $int_x$ is an ancestor of $x$.

### 2019-06-28 paper.tex
add proposition: simple lower bound for distance

@@
 We introduce two types of operations (\emph{$\rnni$ moves}) connecting trees $T$ and $R$ in this graph by an edge:
-A swap of the ranks of two internal nodes ($x$ and $y$) of subsequent ranks (|\rank(x) - \rank(y)| = 1) is called a \emph{rank move}.
+A swap of the ranks of two internal nodes ($x$ and $y$) of subsequent ranks ($|\rank(x) - \rank(y)| = 1$) is called a \emph{rank move}.
 We say that trees $T$ and $R$ are \emph{connected} by an \emph{$\nni$ move} move if there exist internal edges $e$ in $T$ and $f$ in $R$ both of length one such that the graphs obtained by shrinking $e$ and $f$ to internal nodes are isomorphic.
@@
 \end{proof}
+
+%Following Proposition might not be needed(?)
+
+Related to Lemma~\ref{lemma:distance_delete_taxon} is Proposition~\ref{proposition:lower_bound_distance} which provides a lower bound for the distance of two trees.
+Using the same arguments as in the proof of Lemma~\ref{lemma:distance_delete_taxon}, it easily follows that the rank difference of parents of any taxon in two trees gives such a bound.
+
+\begin{proposition}
+    Given trees $T$ and $R$ with taxon $x$ that fulfils $|\rank_T(\parent(x)) - \rank_R(\parent(x))| = k$, it follows $d(T,R) \geq k$.
+    \label{proposition:lower_bound_distance}
+\end{proposition}
 

### 2019-06-28 paper.tex
adjust definitions to literature

@@
 In this paper we study \emph{ranked trees} that are determined by a pair consisting of a rooted binary phylogenetic tree $(T, \Phi)$ on $X = \{1, \ldots, n\}$ for $n \in \mathbb N$, and a rank function.
+In the literature ranked trees are often referred to as labelled history in the context of coalescent trees.
+The following notion differs from the definition of ranks in that context, since we do not use negative integers as ranks.
 A \emph{rank function} on $T$ maps the vertex set of $T$ onto the set $\{0,\ldots,n-1\}$ such that all leaves are assigned rank $0$ and each internal node has rank greater than its children.
@@
 We will identify leaves with their labels and also call them \emph{taxa} (singular: \emph{taxon}) following the systematic biology tradition.
-For a tree $T$, a subtree consisting of an internal node with both children being leaves is called a \emph{cherry}.
-An edge $e$ that has no leaf as a child is called an \emph{internal edge} and we say an edge $e=(u,v)$ has \emph{length} $l$ if $|\rank(v) - \rank(u)| = l$.
-Each internal node $v$ can be uniquely identified by its \emph{cluster} $C$, which is the set of taxa descending from the node.
+For a tree $T$, a pair of taxa adjacent to a common internal node called a \emph{cherry}.
+All nodes that are no leafs are called \emph{internal nodes} and edges $(u,v)$ are \emph{internal edge}, if both $u$ and $v$ are internal nodes.
+Our definition of ranked tree leads to a natural characterisation of edge lengths within ranked trees:
+integer $l$ denotes the \emph{length} of an edge $e=(u,v)$ if $|\rank(v) - \rank(u)| = l$.
+Each node $v$ can be uniquely identified by its \emph{cluster} $C$, which is the set of taxa descending from the node.
 We also say that $v$ \emph{induces} the cluster $C$.
@@
 Such a split will be denoted by $A|B$ and we say that it is \emph{induced} by edge $e$.
-Given a subset $S$ of the set of taxa $X$, the \emph{most recent common ancestor} of $S$, denoted by $\mrca(S)$, is the internal node of the lowest rank that is ancestral to all the taxa in $S$.
+Given a subset $S$ of the set of taxa $X$, the \emph{most recent common ancestor} (also known as \emph{least common ancestor}) of $S$, denoted by $\mrca(S)$, is the internal node of the lowest rank that is ancestral to all the taxa in $S$.
 In particular, for a single taxon $x \in X$, $\mrca(\{x\})$ is the \emph{parent} of $x$ and will be denoted by $\parent(x)$.
@@
 Now we are ready to define the tree space which is the subject of study in this paper, the $\rnni$ graph.
+Note that we will define the $\rnni$ graph on ultrametric trees where all leaves have the same rank $0$, and therefore refers to the $\rnniu$ graph as introduced in \autocite{Gavryushkin2017}.
 The vertex set of the \emph{$\rnni$ graph} is given by the set of all ranked trees on $n$ taxa.
@@
 We say that trees $T$ and $R$ are \emph{connected} by an \emph{$\nni$ move} move if there exist internal edges $e$ in $T$ and $f$ in $R$ both of length one such that the graphs obtained by shrinking $e$ and $f$ to internal nodes are isomorphic.
+The above definition follows \autocite{Gavryushkin2017} and is equivalent to the common definition of $\nni$ operations as given in \autocite{Steel2016-ye}.
 Trees that are connected by either a rank move or an $\nni$ move are called \emph{$\rnni$ neighbours} or \emph{adjacent in} $\rnni$.
@@
 \begin{algorithmic}[1]
-	\STATE $R$ only contains cherry taxa $(a,b)$ of $T$ and has internal node with rank $n-1$
+	\STATE $R$ only contains cherry taxa $a, b$ of $T$ and has internal node with rank $n-1$
     \STATE In list $L$ all taxa $\{1,\ldots,n\}\setminus\{a,b\}$ are sorted such that the ranks of their parents in $T$ do not decrease

### 2019-06-28 paper.tex
move some definitions to where they are needed

@@
 \newcommand{\np}{\mathcal{NP}}
-
 \newcommand{\parent}{\mathrm{parent}}
@@
 \newtheorem{corollary}[definition]{Corollary}
+\newtheorem{proposition}[definition]{Proposition}
 
@@
 A \emph{rank function} on $T$ maps the vertex set of $T$ onto the set $\{0,\ldots,n-1\}$ such that all leaves are assigned rank $0$ and each internal node has rank greater than its children.
-The rank of a node $v$ in a ranked tree will be denoted by $\rank(v)$ or $\rank_T(v)$ if it is not obvious that we consider tree $T$.
+The rank of a node $v$ in a ranked tree will be denoted by $\rank(v)$.
 For simplicity of notation we use the term \emph{tree} instead instead of ranked tree.
@@
 We will identify leaves with their labels and also call them \emph{taxa} (singular: \emph{taxon}) following the systematic biology tradition.
-For a tree $T$, a pair of taxa adjacent to a common internal node called a \emph{cherry}.
 All nodes that are no leafs are called \emph{internal nodes} and edges $(u,v)$ are \emph{internal edge}, if both $u$ and $v$ are internal nodes.
@@
 integer $l$ denotes the \emph{length} of an edge $e=(u,v)$ if $|\rank(v) - \rank(u)| = l$.
-Each node $v$ can be uniquely identified by its \emph{cluster} $C$, which is the set of taxa descending from the node.
+Each node $v$ can be uniquely identified by its \emph{cluster} $C$, which is the set of taxa descending from the node. \todo{move cluster+mrca to alg:find\_path}
 We also say that $v$ \emph{induces} the cluster $C$.
 Note that all trees on $n$ taxa share the cluster $\{1,\ldots,n\}$ induced by the root.
-A \emph{split} is a bipartition of the taxon set $X$ into sets $A,B$ such that there is an edge $e$ whose deletion results in two trees with taxon sets $A$ and $B$, respectively.
-Such a split will be denoted by $A|B$ and we say that it is \emph{induced} by edge $e$.
 Given a subset $S$ of the set of taxa $X$, the \emph{most recent common ancestor} (also known as \emph{least common ancestor}) of $S$, denoted by $\mrca(S)$, is the internal node of the lowest rank that is ancestral to all the taxa in $S$.
-In particular, for a single taxon $x \in X$, $\mrca(\{x\})$ is the \emph{parent} of $x$ and will be denoted by $\parent(x)$.
 We write $\mrca_T(S)$ to emphasise that we consider the most recent common ancestor of set $S$ in tree $T$.
@@
 The following definitions are less standard and required for the results obtained in this paper.
-If every internal node of a tree $T$ has at least one child that is a leaf, $T$ is called a \emph{caterpillar tree}.
-If $p$ is a path only consisting of caterpillar trees we call it \emph{caterpillar path}.
-For two caterpillar trees $T$ and $R$ a \emph{shortest caterpillar path} is a caterpillar path that is shortest among all caterpillar paths between $T$ and $R$.
-The length of such a shortest caterpillar path is called \emph{caterpillar distance} and denoted by $d_c(T,R)$.
 A set $\mathbb S$ of trees is called \emph{convex} if for all pairs of trees $T,R \in \mathbb S$, there exists a shortest path from $T$ to $R$ such that every tree on $p$ is element of $\mathbb S$.
@@
 	\FOR {$i = 1, \dots, n-2$}
-		\STATE Let $C$ be the cluster node $v$ in $R$ with $\rank(v) = i$
+		\STATE Let $C$ be the cluster induced by node $v$ in $R$ with $\rank(v) = i$
 		\WHILE {$\rank(\mrca_{\hat{T}}(C))>i$}
@@
 \begin{lemma}
-%TODO: here and generally, I prefer introducing definitions like this as you go but it's the matter of preference so up to you.
-$\Delta(\rnni) \leq \frac{(n-1)(n-2)}{2}$, where $\Delta$ is...
+    $\Delta(\rnni) \leq \frac{(n-1)(n-2)}{2}$, where $\Delta(G)$ denotes the diameter of graph $G$.
 \end{lemma}
@@
 
+Central for the following inductive proofs is the comparison of the length of a shortest paths $p$ on trees on $n+1$ taxa with a path on $n$ taxa induced by $p$.
+The following lemma bounds the length of these induced paths utilising the rank differences of certain internal nodes between start and end tree of $p$.
+Henceforth the notion $\parent(x)$ refers to the internal node that is parent of taxon $x$.
+
 \begin{lemma}
     Let $T$ and $R$ be two trees on $n+1$ taxa.
-    Let $T|_n$ and $R|_n$ be these trees restricted to $n$ taxa, i.e. the same taxon $n+1$ is deleted from both trees $T$ and $R$.
-    Then it is $d(T|_n, R|_n) \leq d(T,R) - |\rank_T(\parent(n+1)) - \rank_R(\parent(n+1))|$.
+    Let $T{\big|}_n$ and $R{\big|}_n$ be these trees restricted to taxa $1, \ldots, n$, i.e. taxon $n+1$ is deleted from $T$ and $R$ and the thereby arising node of degree two is suppressed.
+    Then it is $d(T{\big|}_n, R{\big|}_n) \leq d(T,R) - \delta$ where $\delta:= |\rank_T(\parent(n+1)) - \rank_R(\parent(n+1))|$ is the rank difference of the parent of taxon $x$ between trees $T$ and $R$.
     \label{lemma:distance_delete_taxon}
@@
 \begin{proof}
-    Let $p$ be a shortest path from tree $T$ to $R$ and $p|_n$ the path resulting from deleting taxon $n+1$ from all trees on $p$.
-    This new path $p|_n$ is then a path from $T$ to $R$.
-    Showing that at least $\delta := |\rank_T(\parent(n+1)) - \rank_R(\parent(n+1))|$ $\rnni$ moves are deleted from $p$ to receive $p|_n$ results in $d(T|_n,R|_n) \leq |p|_n| \leq |p| - \delta = d(T,R) -\delta$ which proves the lemma.
+    Let $p$ be a shortest path from tree $T$ to $R$ and $p{\big|}_n$ the path resulting from deleting taxon $n+1$ from all trees on $p$.
+    This new path $p{\big|}_n$ is then a path from $T$ to $R$.
+    Showing that at least $\delta$ $\rnni$ moves are deleted from $p$ to receive $p{\big|}_n$ results in $d(T{\big|}_n,R{\big|}_n) \leq |p{\big|}_n| \leq |p| - \delta = d(T,R) -\delta$ which proves the lemma.
     By the following two observations one can easily see that this claim is true, indeed:
@@
     Furthermore, any $\rnni$ move can change the rank of an internal node by at most one.
-    Applying this to $\parent(n+1)$ and knowing that the rank of this node differs by $\delta$ between $T$ and $R$ proves that $d(T|_n, R|_n) \leq d(T,R) - \delta$.
+    Applying this to $\parent(n+1)$ and knowing that the rank of this node differs by $\delta$ between $T$ and $R$ proves that $d(T{\big|}_n, R{\big|}_n) \leq d(T,R) - \delta$.
 
@@
     % Basis($k = 1$):
-    % If rank of $\parent(n+1)$ changes between $T$ and $R$, it is $d(T|_n, R|_n) = 0$ and $|\rank_T(\parent(n+1)) - \rank_R(\parent(n+1))| = 1$ and therefore, $d(T|_n, R|_n) = 0 \leq 1 - 1 = d(T,R) - |\rank_T(\parent(n+1)) - \rank_R(\parent(n+1))|$.
-    % If on the other hand the rank of $\parent(n+1}$ does not change between $T$ and $R$, it is $d(T|_n, R|_n) = d(T,R) = 1$ and $|r_{T}(\parent(n+1}) - r_{R}(\parent(n+1})| = 1$ and therefore, $d(T|_n, R|_n) = 1 \leq 1 - 0 = d(T,R) - |r_{T}(\parent(n+1}) - r_{R}(\parent(n+1})|$.
+    % If rank of $\parent(n+1)$ changes between $T$ and $R$, it is $d(T{\big|}_n, R{\big|}_n) = 0$ and $|\rank_T(\parent(n+1)) - \rank_R(\parent(n+1))| = 1$ and therefore, $d(T{\big|}_n, R{\big|}_n) = 0 \leq 1 - 1 = d(T,R) - |\rank_T(\parent(n+1)) - \rank_R(\parent(n+1))|$.
+    % If on the other hand the rank of $\parent(n+1}$ does not change between $T$ and $R$, it is $d(T{\big|}_n, R{\big|}_n) = d(T,R) = 1$ and $|r_{T}(\parent(n+1}) - r_{R}(\parent(n+1})| = 1$ and therefore, $d(T{\big|}_n, R{\big|}_n) = 1 \leq 1 - 0 = d(T,R) - |r_{T}(\parent(n+1}) - r_{R}(\parent(n+1})|$.
     %
@@
     % Let $R'$ be the last tree on a shortest path from $T$ to $R$, i.e. $d(T,R') = k$.
-    % We can apply the induction hypothesis on $T$ and $R'$ and see that $d(T|_n, R'|_n) \leq d(T, R') - |r_{T}(\parent(n+1}) - r_{R'}(\parent(n+1})| := k - m$.
+    % We can apply the induction hypothesis on $T$ and $R'$ and see that $d(T{\big|}_n, R'{\big|}_n) \leq d(T, R') - |r_{T}(\parent(n+1}) - r_{R'}(\parent(n+1})| := k - m$.
     % Now we differ two cases:
@@
     %
-    %     Then the trees $R|_n$ and $R'|_n$ are equal. \todo{Is this obivous?}
-    %     Therefore, it is $d(T|_n,R|_n) = d(T|_n, R'|_n) \leq k - m$.
+    %     Then the trees $R{\big|}_n$ and $R'{\big|}_n$ are equal. \todo{Is this obivous?}
+    %     Therefore, it is $d(T{\big|}_n,R{\big|}_n) = d(T{\big|}_n, R'{\big|}_n) \leq k - m$.
     %     It is either $|r_{T}(\parent(n+1}) - r_{R}(\parent(n+1})| = m+1$, $|r_{T}(\parent(n+1}) - r_{R}(\parent(n+1})| = m$, or $|r_{T}(\parent(n+1}) - r_{R}(\parent(n+1})| = m-1$.
-    %     In either case, we have $d(T|_n,R|_n) = d(T|_n, R'|_n) \leq k - m = (k+1) - (m+1) \leq d(T, R) - |r_{T}(\parent(n+1}) - r_{R}(\parent(n+1})|$.
+    %     In either case, we have $d(T{\big|}_n,R{\big|}_n) = d(T{\big|}_n, R'{\big|}_n) \leq k - m = (k+1) - (m+1) \leq d(T, R) - |r_{T}(\parent(n+1}) - r_{R}(\parent(n+1})|$.
     %
@@
     %
-    %     In this case it is $d(T|_n,R|_n) \leq d(T|_n,R'|_n) + d(R'|_n, R|_n) \leq d(T|_n, R'|_n) + 1 \leq (k-m)+ 1$.
-    %     It also follows that $|r_{T}(\parent(n+1}) - r_{R'}(\parent(n+1})| = |r_{T}(\parent(n+1}) - r_{R}(\parent(n+1})|$ and therefore, we can conclude $d(T|_n,R|_n) \leq k-m+1 = k+1-m = d(T,R) -  |r_{T}(\parent(n+1}) - r_{R}(\parent(n+1})|$.
+    %     In this case it is $d(T{\big|}_n,R{\big|}_n) \leq d(T{\big|}_n,R'{\big|}_n) + d(R'{\big|}_n, R{\big|}_n) \leq d(T{\big|}_n, R'{\big|}_n) + 1 \leq (k-m)+ 1$.
+    %     It also follows that $|r_{T}(\parent(n+1}) - r_{R'}(\parent(n+1})| = |r_{T}(\parent(n+1}) - r_{R}(\parent(n+1})|$ and therefore, we can conclude $d(T{\big|}_n,R{\big|}_n) \leq k-m+1 = k+1-m = d(T,R) -  |r_{T}(\parent(n+1}) - r_{R}(\parent(n+1})|$.
     % \end{enumerate}
@@
 
-The set of caterpillar trees is of interested in $\rnni$ because shortest paths between such trees in $\rnni$ differ from those in classical $\nni$ space.
+In this section we restrict our attention to the set of \emph{caterpillar trees} where each internal node has at least one child that is a leaf.
+This type of tree is of particular interest in $\rnni$ because shortest paths between such trees in $\rnni$ differ from those in classical $\nni$ space.
+This in combination with the importance of caterpillar trees in the proof of $\np$-hardness of $\nni$ distances in \autocite{jiang2000} suggests that understanding the geometry of paths between caterpillar trees helps investigating the complexity of computing $\rnni$ distances.
+Following notations will facilitate the study of shortest paths between caterpillar trees.
+A path $p$ between two caterpillar trees $T$ and $R$ that only consists of caterpillar trees is called \emph{caterpillar path}.
+If $p$ is shortest among all caterpillar paths between these trees we call it \emph{shortest caterpillar path}, and the length of $p$ will be denoted by $d_c(T,R)$ and is called \emph{caterpillar distance}.
+
+%TODO
+We are now going to....
+
 Since $\nni$ moves in $\rnni$ are only allowed on edges of length one, the structure of shortest paths between caterpillar trees is different in $\nni$ and $\rnni$ as it is pointed out in Figure~\ref{fig:NNI_vs_RNNI}.
@@
     From this we can conclude that taxon $n+1$, whose parent has highest rank in $T$, is part of the cherry of $R$.
-    Let us consider the trees $T|_n$ to $R|_n$ that result from $T$ and $R$ by deleting taxon $n+1$.
-    Bubble Sort \todo{name} computes a path from $T|_n$ to $R|_n$ that contains the same exchanges of taxa as $p$, except for the $n-1$ first ones on $p$ that involve taxon $n+1$.
-    Hence, the caterpillar distance is $d_c(T|_n, R|_n) = \frac{n(n-1)}{2} - (n-1)$ = $\frac{(n-1)(n-2)}{2}$.
-    So we can apply the induction hypothesis on $T|_n$ and $R|_n$ and know that $d(T|_n,R|_n) = \frac{(n-1)(n-2)}{2}$.
+    Let us consider the trees $T{\big|}_n$ to $R{\big|}_n$ that result from $T$ and $R$ by deleting taxon $n+1$.
+    Bubble Sort \todo{name} computes a path from $T{\big|}_n$ to $R{\big|}_n$ that contains the same exchanges of taxa as $p$, except for the $n-1$ first ones on $p$ that involve taxon $n+1$.
+    Hence, the caterpillar distance is $d_c(T{\big|}_n, R{\big|}_n) = \frac{n(n-1)}{2} - (n-1)$ = $\frac{(n-1)(n-2)}{2}$.
+    So we can apply the induction hypothesis on $T{\big|}_n$ and $R{\big|}_n$ and know that $d(T{\big|}_n,R{\big|}_n) = \frac{(n-1)(n-2)}{2}$.
 
-    According to Lemma~\ref{lemma:distance_delete_taxon}, the distance between $T|_n$ and $R|_n$ is at least $d(T|_n, R|_n) \leq d(T,R) - |\rank_T(\parent(n+1)) - \rank_R(\parent(n+1))|$.
+    According to Lemma~\ref{lemma:distance_delete_taxon}, the distance between $T{\big|}_n$ and $R{\big|}_n$ is at least $d(T{\big|}_n, R{\big|}_n) \leq d(T,R) - |\rank_T(\parent(n+1)) - \rank_R(\parent(n+1))|$.
     We also know $|\rank_T(\parent(n+1)) - \rank_R(\parent(n+1))| = n-1$
     Suppose, contrary to our claim, that there is a path from $T$ to $R$ of length less than $\Delta(\rnniu_{n+1}) = \frac{n(n-1)}{2}$.
-    By the above, it follows that $d(T|_n, R|_n) \leq d(T,R) - |\rank_T(\parent(n+1)) - \rank_R(\parent(n+1))| < \frac{n(n-1)}{2} - (n-1) < \frac{(n-1)(n-2)}{2}$.
-    Since this is a contradiction to $d(T|_n, R|_n) = \frac{(n-1)(n-2)}{2}$, it is proven that $d(T,R) = \frac{n(n-1)}{2} $.
+    By the above, it follows that $d(T{\big|}_n, R{\big|}_n) \leq d(T,R) - |\rank_T(\parent(n+1)) - \rank_R(\parent(n+1))| < \frac{n(n-1)}{2} - (n-1) < \frac{(n-1)(n-2)}{2}$.
+    Since this is a contradiction to $d(T{\big|}_n, R{\big|}_n) = \frac{(n-1)(n-2)}{2}$, it is proven that $d(T,R) = \frac{n(n-1)}{2} $.
 
@@
 Afterwards, $T'$ is transformed to a tree $R'$ still containing the same cherries as $T'$ but in a way that resolving all these cherries of $R'$ results in $R$.
-For proving that this path is shorter than any path preserving the clusters $\{1, \ldots, n\}$ and $\{n+1, \ldots, 2n\}$, the fact that $d_{\nni}(R,R') = d_{\nni}(T|_{\{1, \ldots, n\}}, R|_{\{1, \ldots, n\}})$ where $T|_{\{1, \ldots, n\}}$ and $R|_{\{1, \ldots, n\}}$ are $T$ and $R$ restricted to taxa $\{1,\ldots,n\}$, respectively, is being used.
+For proving that this path is shorter than any path preserving the clusters $\{1, \ldots, n\}$ and $\{n+1, \ldots, 2n\}$, the fact that $d_{\nni}(R,R') = d_{\nni}(T{\big|}_{\{1, \ldots, n\}}, R{\big|}_{\{1, \ldots, n\}})$ where $T{\big|}_{\{1, \ldots, n\}}$ and $R{\big|}_{\{1, \ldots, n\}}$ are $T$ and $R$ restricted to taxa $\{1,\ldots,n\}$, respectively, is being used.
 However, this does not work in $\rnni$.
@@
 % How do we assign the ranks to the new cherry nodes?
-% We could either create a path where $R$ and $R'$ have distance equal to $d_{\rnni}(T|_{\{1, \ldots, n\}}, R|_{\{n+1, \ldots, 2n\}})$ and have a lot of additional rank moves before and after that part OR
+% We could either create a path where $R$ and $R'$ have distance equal to $d_{\rnni}(T{\big|}_{\{1, \ldots, n\}}, R{\big|}_{\{n+1, \ldots, 2n\}})$ and have a lot of additional rank moves before and after that part OR
 % Build $R$ and $R'$ that contain all the cherries within the least number of moves possible (3n-3) and having a longer path between $R$ and $R'$.
@@
 
-Since the example given in \autocite{Li1996} that proves Theorem~\ref{thm:split_nni} in $\nni$ does not work in $\rnni$, the following conjecture was raised \autocite{Gavryushkin2017}.
+Since the example given in \autocite{Li1996} that proves Theorem~\ref{thm:split_nni} in $\nni$ does not work in $\rnni$, a contrary statement was conjectured in \autocite{Gavryushkin2017}.
+For understanding this conjecture it is important to realise the connection between edges in a tree and bipartitions of the taxon set of that tree.
+By deleting an edge $e$ of a tree $T$ one receives two trees on taxon sets $A$ and $B$ such that $\{A,B\}$ is a partition of the taxon set of $T$.
+Note that each edge induces such a bipartition, often written as $A|B$, and that both edges incident to the root of a tree induce the same partition.
+As these partitions are commonly referred to as \emph{splits}, the following conjecture was named \emph{Split Theorem} in \autocite{Gavryushkin2017}.
 
@@
 
-Instead of using the notion of clusters as we do within this paper, the Split Theorem considers splits of the taxon set $X$.
 However, we found a simple counterexample to this conjecture:
@@
 Within the Cluster Theorem we are considering clusters instead of splits.
-This is motivated by the fact that rooted phylogenetic trees can uniquely be represented by sets of clusters \autocite{steel2016}, but not by sets of splits as they cannot define the position of the root.
+This is motivated by the fact that rooted phylogenetic trees can uniquely be represented by sets of clusters \autocite{Steel2016-ye}, but not by sets of splits as they cannot define the position of the root.
 Moreover, a ranked tree can be uniquely determined by ordering the set of clusters increasingly according to the ranks of the internal nodes inducing those clusters.

### 2019-06-30 paper.tex
description alg: find path mrca

@@
 integer $l$ denotes the \emph{length} of an edge $e=(u,v)$ if $|\rank(v) - \rank(u)| = l$.
-Each node $v$ can be uniquely identified by its \emph{cluster} $C$, which is the set of taxa descending from the node. \todo{move cluster+mrca to alg:find\_path}
-We also say that $v$ \emph{induces} the cluster $C$.
-Note that all trees on $n$ taxa share the cluster $\{1,\ldots,n\}$ induced by the root.
-Given a subset $S$ of the set of taxa $X$, the \emph{most recent common ancestor} (also known as \emph{least common ancestor}) of $S$, denoted by $\mrca(S)$, is the internal node of the lowest rank that is ancestral to all the taxa in $S$.
-We write $\mrca_T(S)$ to emphasise that we consider the most recent common ancestor of set $S$ in tree $T$.
 Two trees are called \emph{isomorphic} if the underlying rooted phylogenetic trees are isomorphic and the isomorphism preserves ranks of the nodes.
@@
 
-%If needed: diameter, radius (in the corresponding chapter?), induced triplet (in that chapter?)
-
 \section{Geometry of $\rnni$}
@@
 %TODO
-The idea of the algorithm is the following...
+The idea of the algorithm is described in the following.
+Each node $v$ of a tree defines a \emph{cluster} $C$, which is the set of taxa descending from the $v$.
+We then say that $v$ \emph{induces} the cluster $C$.
+Note that the list of all clusters of a tree, if sorted increasingly according to the rank of the corresponding nodes, uniquely defines that tree.
+For computing a path from trees $T$ to $R$, we consider clusters $C_1, \ldots, C_{n-1}$ of $R$ where $C_i$ is induced by the node with rank $i$ for $i = 1, \ldots, n-1$.
+The so-called \emph{trivial clusters} that are induced by leaves and the root are the same in every tree and do therefore not need to be changed on any path between two trees.
+Therefore, the list of clusters $C_1, \ldots, C_{n-1}$, which is ordered increasingly according to the ranks of the corresponding nodes, uniquely defines $R$.
+Algorithm~\ref{alg:find_path} starts at tree $\hat{T} = T$ and constructs a path where the clusters of $R$ are constructed from bottom to top, using following procedure:
+For each $C_i$, find the \emph{most recent common ancestor} $\mrca_{\hat{T}}{C_i}$ (also known as \emph{least common ancestor}) of $C_i$ in current tree $\hat{T}$, that is the internal node with lowest rank that is ancestral to all the taxa in $C_i$.
+Update $\hat{T}$ by decreasing the rank of $\mrca_{\hat{T}}{C_i}$ by an $\rnni$ move until it has the same rank $i$ as in $R$.
+\todo{Do we need to explain why there is always an $\rnni$ move that moves the $\mrca$ down?}
+Since we consider the clusters of $R$ such that the ranks of the corresponding internal nodes in $R$ increase, $\mrca_{\hat{T}}{C_i}$ is the parent of cluster $C_i$ in $\hat{T}$ after these $\rnni$ moves.
+Thus, repeating this procedure for all clusters of $R$ results in a path between any pair of tress $T, R$, indeed.
+
+\todo{Keep the Pseudo-Code?}
 
@@
 
-\todo{This text comes out of the blue.
-Generally, before going into explanations like this it's a good idea to say what's ahead.}
-One can easily prove that there is always exactly one $\rnni$ move on $T$ that decreases the rank of $\mrca_T(C)$ as in line~\ref{alg:line:move_set_down} of Algorithm~\ref{alg:find_path}.
-It follows that Algorithm~\ref{alg:find_path} returns a path in $\rnni$ for any pair of trees $T,R$.
-\todo{This should have been stated before, where you say what is going on.}
-Moreover, it gives an upper bound for the diameter of $\rnni$, which improves the one given in \autocite{Gavryushkin2017}:
+This algorithm gives an upper bound for the diameter of $\rnni$, which improves the one given in \autocite{Gavryushkin2017}.
+Later we will see that this bound is tight.
 

### 2019-07-01 paper.tex
edit beginning of caterpillar section

@@
 
-%TODO
 The idea of the algorithm is described in the following.
@@
 This in combination with the importance of caterpillar trees in the proof of $\np$-hardness of $\nni$ distances in \autocite{jiang2000} suggests that understanding the geometry of paths between caterpillar trees helps investigating the complexity of computing $\rnni$ distances.
-Following notations will facilitate the study of shortest paths between caterpillar trees.
+In particular, we are going to prove that one can always find a shortest path between two caterpillar trees that contains caterpillar trees only, meaning that the set of caterpillar trees is convex in $\rnni$.
+We discovered this by computing caterpillar paths using an algorithm to compute shortest paths within the set of caterpillar trees and comparing the results with shortest path in $\rnni$.
+This algorithm will be explained in detail later in this section, after demonstrating why finding shortest paths between caterpillar trees builds the basis for understanding the difference between $\nni$ and $\rnni$.
+Further details on our computational results can be found in section~\ref{computation}.
+
+Before we start analysing shortest paths between caterpillar trees, we introduce further notations.
 A path $p$ between two caterpillar trees $T$ and $R$ that only consists of caterpillar trees is called \emph{caterpillar path}.
@@
 
-%TODO
-We are now going to....
-
-Since $\nni$ moves in $\rnni$ are only allowed on edges of length one, the structure of shortest paths between caterpillar trees is different in $\nni$ and $\rnni$ as it is pointed out in Figure~\ref{fig:NNI_vs_RNNI}.
+Let us compare shortest paths between caterpillar trees in $\nni$ with those in $\rnni$.
+The essential difference is that building a cherry and moving it may give a path that is shorter than a path where the taxa move separately in $\nni$ as it is illustrated out in Figure~\ref{fig:NNI_vs_RNNI}.
+There is only one move needed in $\nni$ to build and resolve a cherry, respectively, while the number of moves for moving the cherry is the same as for moving a single taxon.
+However, in $\rnni$ space, there are additional rank moves needed for the internal node of the cherry before an $\nni$ move resolves it, because $\nni$ moves are only allowed on edges of length one in $\rnni$.
+Therefore, a path where taxa move separately and no second cherry is built, is shortest in $\rnni$.
 
@@
 
-This observation already suggests that there is always a caterpillar path that is shortest path between two arbitrary caterpillar trees in $\rnni$.
-Indeed, our aim for this section is to prove Theorem~\ref{thm:caterpillar_convex} which states that the set of caterpillar trees is convex in $\rnni$.
+The above observation already suggests that there is always a caterpillar path that is shortest path between two arbitrary caterpillar trees in $\rnni$.
+Indeed, we are going to prove Theorem~\ref{thm:caterpillar_convex} within this section, which states that the set of caterpillar trees is convex in $\rnni$.
 Note that this is not true for the $\nni$ space, which is clear from the example in Figure~\ref{fig:NNI_vs_RNNI}.
 
-% Shall we mention the data structure we use (list of sets) - this might be useful to see the connection to the partition lattice.
 One can see caterpillar trees as sequences of taxa, ordered non-decreasing according to the ranks of their parents.
@@
 
+\section{Computational results}
+
 \section{Ideas}

### 2019-07-01 paper.tex
forgotten label
name for bubble sort on caterpillars: caterpillar sort

@@
 \newcommand{\rspr}{\mathrm{rSPR}}
+\newcommand{\csort}{\textsc{Caterpillar Sort}}
 
@@
 One can see caterpillar trees as sequences of taxa, ordered non-decreasing according to the ranks of their parents.
-$\nni$ moves between caterpillar trees, which are the only possible $\rnni$ moves between caterpillar trees, coincide with exchanges of two neighboured elements in the sequence.
+$\nni$ moves between caterpillar trees, which are the only possible $\rnni$ moves between those trees, coincide with exchanges of two neighboured elements in the sequence.
 In these sequences the first two taxa are somehow equivalent since they build the unique cherry.
 $\nni$ moves that change the cherry of a tree refer to exchanges of the first or the second element with the third element of the sequence while exchanges of the first two elements do not change the corresponding tree.
-With this observation we can provide an algorithm for computing shortest caterpillar paths by a simple modification of the commonly known \emph{Bubble Sort} algorithm \autocite{knuth1997}:
+With this observation we can provide an algorithm for computing shortest caterpillar paths by a simple modification of the commonly known \emph{Bubble Sort} algorithm \autocite{knuth1997}.
+Throughout this paper this algorithm will be called $\csort$.
 
-Following the observations above, only the moves for the first three positions of a sequence with Bubble Sort need to be adapted for the computation of shortest caterpillar paths.
+Following the observations above, only swaps of the first three elements of a sequence with Bubble Sort need to be adapted for the computation of shortest caterpillar paths.
 Let us consider the first three elements $a,b,c$ which appear in this order in a sequence representing a caterpillar tree on a caterpillar path between trees $T$ and $R$.
@@
     Let us first prove that if $d(T,R) = \frac{(n-1)(n-2)}{2}$, it follows $d_c(T,R) = \frac{(n-1)(n-2)}{2}$.
-    This easily follows from the modified Bubble Sort algorithm. \todo{name?}
+    This easily follows from algorithm $\csort$:
     In each loop $i=1, \ldots, n-2$ of the algorithm there can be at most $n-1-i$ exchanges of taxa.
@@
     Let $T, R$ be trees on $n+1$ taxa with caterpillar distance $\frac{(n-1)(n-2)}{2}$.
-    Therefore, each of the $\frac{(n-1)(n-2)}{2}$ pairs of taxa that is considered when running modified Bubble Sort algorithm \todo{name} to compute a path $p$ from $R$ to $T$ exchanges.
+    Therefore, each of the $\frac{(n-1)(n-2)}{2}$ pairs of taxa that is considered when running $\csort$ to compute a path $p$ from $R$ to $T$ exchanges.
     From this we can conclude that taxon $n+1$, whose parent has highest rank in $T$, is part of the cherry of $R$.
     Let us consider the trees $T{\big|}_n$ to $R{\big|}_n$ that result from $T$ and $R$ by deleting taxon $n+1$.
-    Bubble Sort \todo{name} computes a path from $T{\big|}_n$ to $R{\big|}_n$ that contains the same exchanges of taxa as $p$, except for the $n-1$ first ones on $p$ that involve taxon $n+1$.
+    $\csort$ computes a path from $T{\big|}_n$ to $R{\big|}_n$ that contains the same exchanges of taxa as $p$, except for the $n-1$ first ones on $p$ that involve taxon $n+1$.
     Hence, the caterpillar distance is $d_c(T{\big|}_n, R{\big|}_n) = \frac{n(n-1)}{2} - (n-1)$ = $\frac{(n-1)(n-2)}{2}$.
@@
     Let $T$ and $R$ be caterpillar trees with caterpillar distance $l$.
-    Since $l < \frac{(n-1)(n-2)}{2}$, there is one loop $k$ in the caterpillar path from $R$ to $T$ that is given by the Bubble Sort algorithm \todo{name} where less than $n-2-k$ pairs taxa swap positions.
+    Since $l < \frac{(n-1)(n-2)}{2}$, there is one loop $k$ in the caterpillar path from $R$ to $T$ that is given $\csort$ where less than $n-2-k$ pairs taxa swap positions.
     It follows that there is a taxon $a$ that has parent with rank $n-2-k$ in $T$ and does not need to move all the way up starting at the cherry in loop $k$.
@@
     According to Theorem~\ref{thm:caterpillar_convex}, there is a shortest path between $T$ and $R$ that is a caterpillar path.
-    We can compute a shortest caterpillar path by using the modified version of Bubble Sort as explained above. \todo{Do we want to name this algorithm?}
+    We can compute a shortest caterpillar path by using $\csort$ as explained above.
     When running this algorithm on $T$ and $R$, taxon $i$ swaps $n-1-i$ times with its right neighbour.
@@
 \section{Computational results}
+\label{computation}
 
@@
     % Insert the proof here
-    Fix $T = (\ldots(1,2) \ldots ,n)$, use induction on $n$ and bubble sort (taxon $n$ must be in cherry of $R$ if distance is max).
+    Fix $T = (\ldots(1,2) \ldots ,n)$, use induction on $n$ and $\csort$ (taxon $n$ must be in cherry of $R$ if distance is max).
 \end{proof}

### 2019-07-01 paper.tex
edit beginning of section geometry

@@
 This paper is an advance towards the general goal of understanding the geometry of the $\rnni$ space.
-%TODO
-In this section, we...
+In this section we are investigating the geometry by analysing the structure of shortest paths in $\rnni$.
+We will start by presenting an algorithm for computing paths in $\rnni$ (Algorithm~\ref{alg:find_path}) that turns out to be an essential tool for our studies.
+Using this algorithm, we are able to prove that there is always a shortest path between two caterpillar trees that only consists of caterpillar trees (Theorem~\ref{thm:caterpillar_convex}) in Section~\ref{section:caterpillar_convex}.
+In other words, the set of caterpillar trees is convex in $\rnni$.
+This distinguishes $\rnni$ from $\nni$ where this does not hold.
+One further result, presented in Section~\ref{section:diameter}, directly following from the proof of Theorem~\ref{thm:caterpillar_convex}, is the diameter of $\rnni$.
+Finally, we disprove the so-called Split Theorem, conjectured in \autocite{Gavryushkin2017}, and formulate an alternative version of it, the Cluster Theorem (Section~\ref{section:split_theorem}).
 
@@
 Later we will see that this bound is tight.
+However, we don't know how good Algorithm~\ref{alg:find_path} distances in $\rnni$.
+More details can be found in Section~\ref{section:computation}
 
@@
 This algorithm will be explained in detail later in this section, after demonstrating why finding shortest paths between caterpillar trees builds the basis for understanding the difference between $\nni$ and $\rnni$.
-Further details on our computational results can be found in section~\ref{computation}.
+Further details on our computational results can be found in Section~\ref{section:computation}.
 
@@
 \subsection{Diameter}
+\label{section:diameter}
 
@@
 \subsection{Split Theorem}
+\label{section:split_theorem}
 
@@
 \section{Computational results}
-\label{computation}
+\label{section:computation}
 

### 2019-07-01 paper.tex
edit beginning of section diameter

@@
 
-Following the results of section~\ref{section:caterpillar_convex} one can directly improve this result by giving the exact diameter of the $\rnni$ graph.
+With the results of Section~\ref{section:caterpillar_convex} we can easily provide the exact diameter of the $\rnni$ graph and hence, improve the bounds for the diameter given in \autocite{Gavryushkin2017}.
+Moreover, we can conclude from Lemma~\ref{lemma:caterpillar_diameter} that the maximum distance computed by Algorithm~\ref{alg:find_path} equals the diameter of $\rnni$.
+This quality of the algorithm suggests that it gives reasonable estimates of distances.
+For further details on the quality of that algorithm we refer the reader to Section~\ref{section:computation}.
 
@@
 	There are two caterpillar trees in $\rnniu$ with distance $\frac{(n-1)(n-2))}{2}$.
-	\label{conj:caterpillar_diameter}
+	\label{lemma:caterpillar_diameter}
 \end{lemma}
-
-\todo{This is already an implication of the results above. Shall we still keep a subsection for Diameter? In case we find something about the radius, this could be added here.}
 

### 2019-07-01 paper.tex
edits on split thm section

@@
 
-%Shall we put this bit into the introduction?
-It is natural to try to relate the $\rnni$ graph to the $\nni$ as we already did in previous sections.
-Since computing distances in tree spaces is a main issue in most commonly known tree spaces as $\nni$ \autocite{jiang2000}, $\rspr$ \autocite{Bordewich2005}, or $\tbr$ \autocite{allen2001subtree} is $\np$-hard, it would be desirable to find a tree space that is less complex.
+Within this section we will make a further step towards understanding the complexity of computing shortest paths in $\rnni$.
+Since the problem of computing distances in the most common tree spaces as $\nni$ \autocite{jiang2000}, $\rspr$ \autocite{Bordewich2005}, or $\tbr$ \autocite{allen2001subtree} is known to be $\np$-hard, it would be desirable to find a tree space that is less complex.
 However, it is not known whether computing distances in $\rnni$ is $\np$-hard or not.
 The natural approach to solve this question is to compare $\rnni$ and $\nni$ graph to see whether the reason that $\nni$ is $\np$-hard transfers to $\rnni$.
+In this section we will consider an important example of shortest paths in $\nni$ that build the key of the proof of its $\np$-hardness.
+Afterwards we see that this example cannot transferred to $\rnni$ which is why the so-called Split Theorem, which generalises the example, has been conjectured in \autocite{Gavryushkin2017}.
+However, we present a counterexample disproving the Split Theorem (Conjecture~\ref{conjecture:split_theorem}) in its original version and claiming the Cluster Theorem (Conjecture~\ref{conjecture:cluster_theorem}) as alternative version.
 
@@
 	\label{thm:split_nni}
+    %The formulation of this lemma is a bit different in Li1996.
+    %However, the lemma as stated here follows directly from Li1996.
 \end{theorem}
@@
 
-The proof presented in \autocite{Li1996} for Theorem~\ref{thm:split_nni} does not work for $\rnni$.
-
-%This part is mainly for myself, not sure whether it is important to have it in the paper
-\todo{Shall we keep this explanation? The problem is that the $\rnni$ path depends on the assignment of ranks to the cherries produced on the path}
-The counterexample presented in \autocite{Li1996} works in $\nni$ because of the following:
-When proving Theorem~\ref{thm:split_nni}, \autocite{Li1996} gives an example of two trees $T$ and $R$ on $2n$ taxa that consist of two caterpillar trees each that share the root as parent.
+In the proof presented in \autocite{Li1996} for Theorem~\ref{thm:split_nni} an example of two trees that share a cluster but have no shortest path that preserves this cluster is given.
+The reason that this proof does not work for $\rnni$ is similar to the reason that caterpillar trees are convex in $\rnni$, but not in $\nni$.
+Both trees $T$ and $R$ in the example of \autocite{Li1996} have $2n$ taxa and consist of two caterpillar trees that share the root as parent.
 The clusters given by the two nodes adjacent to the root are $\{1, \ldots, n\}$ and $\{n+1, \ldots, 2n\}$ in both trees $T$ and $R$.
-Between these there is a shortest path in $\nni$ where at first $T$ is transformed to a tree $T'$ with $n$ cherries where each cherry contains one taxon of $\{1, \ldots, n\}$ and one taxon of $\{n+1, \ldots, 2n\}$.
+There is a shortest path in $\nni$ where at first $T$ is transformed to a tree $T'$ with $n$ cherries where each cherry contains one taxon of $\{1, \ldots, n\}$ and one taxon of $\{n+1, \ldots, 2n\}$.
 Afterwards, $T'$ is transformed to a tree $R'$ still containing the same cherries as $T'$ but in a way that resolving all these cherries of $R'$ results in $R$.
@@
 However, this does not work in $\rnni$.
-Introducing new cherries means introducing new internal nodes with new ranks.
-Since there are lots of rank moves needed for an $\rnni$ path similar to the shortest $\nni$ path presented above, the path in $\rnni$ that breaks the clusters $\{1, \ldots, n\}$ and $\{n+1, \ldots, 2n\}$ is no shortest path.
-% How do we assign the ranks to the new cherry nodes?
+Similar to what we observed in the example depicted in Figure~\ref{fig:NNI_vs_RNNI}, there are lots of additional rank moves needed on a similar path in $\rnni$.
+Therefore, a path where the two caterpillar trees are sorted separately, is shorter in $\rnni$.
+\todo{(1) Do we want this explanation?
+(2) If yes, is this explanation sufficient (we never say how we assign ranks to $T$ and $R$)?}
+
+% \todo{Shall we keep this explanation? The problem is that the $\rnni$ path depends on the assignment of ranks to the cherries produced on the path}
+% The counterexample presented in \autocite{Li1996} works in $\nni$ because of the following:
+% When proving Theorem~\ref{thm:split_nni}, \autocite{Li1996} gives an example of two trees $T$ and $R$ on $2n$ taxa that consist of two caterpillar trees each that share the root as parent.
+% The clusters given by the two nodes adjacent to the root are $\{1, \ldots, n\}$ and $\{n+1, \ldots, 2n\}$ in both trees $T$ and $R$.
+% Between these there is a shortest path in $\nni$ where at first $T$ is transformed to a tree $T'$ with $n$ cherries where each cherry contains one taxon of $\{1, \ldots, n\}$ and one taxon of $\{n+1, \ldots, 2n\}$.
+% Afterwards, $T'$ is transformed to a tree $R'$ still containing the same cherries as $T'$ but in a way that resolving all these cherries of $R'$ results in $R$.
+% For proving that this path is shorter than any path preserving the clusters $\{1, \ldots, n\}$ and $\{n+1, \ldots, 2n\}$, the fact that $d_{\nni}(R,R') = d_{\nni}(T{\big|}_{\{1, \ldots, n\}}, R{\big|}_{\{1, \ldots, n\}})$ where $T{\big|}_{\{1, \ldots, n\}}$ and $R{\big|}_{\{1, \ldots, n\}}$ are $T$ and $R$ restricted to taxa $\{1,\ldots,n\}$, respectively, is being used.
+% However, this does not work in $\rnni$.
+% Introducing new cherries means introducing new internal nodes with new ranks.
+% Since there are lots of rank moves needed for an $\rnni$ path similar to the shortest $\nni$ path presented above, the path in $\rnni$ that breaks the clusters $\{1, \ldots, n\}$ and $\{n+1, \ldots, 2n\}$ is no shortest path.
+
+% ^How do we assign the ranks to the new cherry nodes?
 % We could either create a path where $R$ and $R'$ have distance equal to $d_{\rnni}(T{\big|}_{\{1, \ldots, n\}}, R{\big|}_{\{n+1, \ldots, 2n\}})$ and have a lot of additional rank moves before and after that part OR
@@
 	If a partition of leaves given by an edge is present in two trees $T$ and $R$ then the partition is presented in every tree on every shortest path between $T$ and $R$.
-	\label{split_theorem}
+	\label{conjecture:split_theorem}
 \end{conjecture}
@@
 	if two trees $T$ and $R$ contain the same cluster $C$, then $C$ is present as cluster in every tree on every shortest path between $T$ and $R$.
-	\label{cluster_theorem}
+	\label{conjecture:cluster_theorem}
 \end{conjecture}

### 2019-07-01 paper.tex
first draft section computation

@@
 \newcommand{\csort}{\textsc{Caterpillar Sort}}
+\newcommand{\findpath}{\textsc{FindPath}}
 
@@
 \section{Geometry of $\rnni$}
+\label{section:geometry}
 
@@
 
-We now present an efficient algorithm (Algorithm~\ref{alg:find_path}) for computing paths between pairs of trees in $\rnni$.
-This algorithm will be a central tool for our investigation of shortest paths in $\rnni$.
+We now present an efficient algorithm for computing paths between pairs of trees in $\rnni$.
+Throughout this paper we will denote this algorithm by $\findpath$.
+We will see that it becomes a central tool for our investigation of shortest paths.
 
@@
 Therefore, the list of clusters $C_1, \ldots, C_{n-1}$, which is ordered increasingly according to the ranks of the corresponding nodes, uniquely defines $R$.
-Algorithm~\ref{alg:find_path} starts at tree $\hat{T} = T$ and constructs a path where the clusters of $R$ are constructed from bottom to top, using following procedure:
+Starting at tree $\hat{T} = T$, $\findpath$ constructs a path where the clusters of $R$ are constructed from bottom to top, using following procedure:
 For each $C_i$, find the \emph{most recent common ancestor} $\mrca_{\hat{T}}{C_i}$ (also known as \emph{least common ancestor}) of $C_i$ in current tree $\hat{T}$, that is the internal node with lowest rank that is ancestral to all the taxa in $C_i$.
@@
 Thus, repeating this procedure for all clusters of $R$ results in a path between any pair of tress $T, R$, indeed.
-
-\todo{Keep the Pseudo-Code?}
-
-\begin{algorithm}[H]
-\caption{FIND\_PATH($T,R$)}
-\label{alg:find_path}
-\begin{algorithmic}[1]
-	\STATE $\hat{T} := T$
-	\STATE $p := (T)$
-	\FOR {$i = 1, \dots, n-2$}
-		\STATE Let $C$ be the cluster induced by node $v$ in $R$ with $\rank(v) = i$
-		\WHILE {$\rank(\mrca_{\hat{T}}(C))>i$}
-			\STATE Update $\hat{T}$: Decrease $\rank(\mrca_{\hat{T}}(C))$ by an $\rnni$ move \label{alg:line:move_set_down}
-			\STATE $p = p+\hat{T}$
-		\ENDWHILE
-	\ENDFOR
-	\RETURN $p$
-\end{algorithmic}
-\end{algorithm}
+Further details, including pseudo-code for $\findpath$, can be found in Section~\ref{section:computation_findpath}.
 
@@
 Later we will see that this bound is tight.
-However, we don't know how good Algorithm~\ref{alg:find_path} distances in $\rnni$.
-More details can be found in Section~\ref{section:computation}
+However, we don't know how good $\findpath$ approximates distances in $\rnni$. \todo{Can we say something about the approximation ratio?}
+More details can be found in Section~\ref{section:computation}.
 
@@
 This makes sure that there is no additional swap of $a$ and $b$ needed on the path to $R$.
-\todo{Pseudo-Code for algorithm?}
+\todo{Do we need to explain Bubble Sort?}
 
@@
 With the results of Section~\ref{section:caterpillar_convex} we can easily provide the exact diameter of the $\rnni$ graph and hence, improve the bounds for the diameter given in \autocite{Gavryushkin2017}.
-Moreover, we can conclude from Lemma~\ref{lemma:caterpillar_diameter} that the maximum distance computed by Algorithm~\ref{alg:find_path} equals the diameter of $\rnni$.
-This quality of the algorithm suggests that it gives reasonable estimates of distances.
+Moreover, we can conclude from Lemma~\ref{lemma:caterpillar_diameter} that the maximum distance computed by algorithm $\findpath$, which was dicussed introduced above, equals the diameter of $\rnni$.
+This suggests that it gives reasonable estimates of distances.
 For further details on the quality of that algorithm we refer the reader to Section~\ref{section:computation}.
@@
 \label{section:split_theorem}
+\todo{Split or Cluster?}
 
@@
 
+In \autocite{Gavryushkin2017} an efficient algorithm for computing the $\rnni$ graph has been introduced.
+We implemented this algorithm and could produce the graph for trees with up to $8$ taxa.
+This enables us to compute exact distances between trees of this size.
+It is in particular possible to compare the exact distances with distances we receive from algorithms we introduced previously and to verify the Cluster Theorem, which was raised in Section~\ref{section:split_theorem}, for small trees.
+All algorithms have been implemented with Python and are available online. \todo{link?}
+
+\subsection{Algorithm $\findpath$}
+\label{section:computation_findpath}
+
+In Section~\ref{section:geometry} we introduced the algorithm $\findpath$ for computing paths between arbitrary trees in $\rnni$.
+Therefore, we can use it as approximation for distances in $\rnni$.
+In this section we are providing pseudo-code for $\findpath$ (Algorithm~\ref{alg:find_path}) to provide further insights.
+We implemented and tested this algorithm for trees on up to $k$ taxa, for which it gives the exact $\rnni$ distance. \todo{$k = ?$}
+Furthermore, we saw in Section~\ref{section:diameter} that the maximum distance computed by $\findpath$ for an arbitrary pair of trees equals the diameter of $\rnni$.
+Thus, if the distance between two trees equals the diameter, $\findpath$ computes the correct distance.
+
+\begin{algorithm}[H]
+\caption{$\findpath$($T,R$)}
+\label{alg:find_path}
+\begin{algorithmic}[1]
+	\STATE $\hat{T} := T$
+	\STATE $p := (T)$
+	\FOR {$i = 1, \dots, n-2$}
+		\STATE Let $C$ be the cluster induced by node $v$ in $R$ with $\rank(v) = i$
+		\WHILE {$\rank(\mrca_{\hat{T}}(C))>i$}
+			\STATE Update $\hat{T}$: Decrease $\rank(\mrca_{\hat{T}}(C))$ by an $\rnni$ move \label{alg:line:move_set_down}
+			\STATE $p = p+\hat{T}$
+		\ENDWHILE
+	\ENDFOR
+	\RETURN $p$
+\end{algorithmic}
+\end{algorithm}
+
+\subsection{Connecting Computations and Proofs}
+\todo{How shall we structure this?}
+
+In Section~\ref{section:caterpillar_convex} we already indicated that computing distances between trees in $\rnni$ by using the algorithms introduced in previous sections has been important for understanding shortest paths.
+Moreover, these algorithm, in particular $\findpath$ and $\csort$, also became a substantial part of some of the in Section~\ref{section:geometry}.
+
+We can also substantiate the Cluster Theorem, which remains to be a conjecture, by proving it computationally for up to $k$ taxa. \todo{k=?, do we provide implementation?}
+Together with the analytical results of the previous section, this strongly supports our claim that the Cluster Theorem is true.
+
+%Max Distance Trees Algorithm (+ connection to MRCA algorithm)
+
 \section{Ideas}

### 2019-07-03 paper.tex
add alg:max_dist_path to computation chapter

@@
 We implemented and tested this algorithm for trees on up to $k$ taxa, for which it gives the exact $\rnni$ distance. \todo{$k = ?$}
-Furthermore, we saw in Section~\ref{section:diameter} that the maximum distance computed by $\findpath$ for an arbitrary pair of trees equals the diameter of $\rnni$.
-Thus, if the distance between two trees equals the diameter, $\findpath$ computes the correct distance.
 
@@
 \end{algorithm}
+
+Recall that the Section~\ref{section:diameter} revealed that the maximum distance computed by $\findpath$ for an arbitrary pair of trees equals the diameter of $\rnni$.
+Thus, if the distance between two trees equals the diameter, $\findpath$ computes the correct distance.
+Note that central for proving the convexity of the set of caterpillar trees in Section~\ref{section:caterpillar_convex} was Lemma~\ref{lemma:caterpillar_dist=diameter}.
+We therefore aim to find a property for any pair of trees, not only caterpillar trees, with maximum distance $\Delta(\rnni) = \frac{(n-1)(n-2)}{2}$.
+Following algorithm $\findpath$, we claim following conjecture:
+
+\begin{conjecture}
+    Let $T$ and $R$ be two trees.
+    It is $d(T,R) = \frac{(n-1)(n-2)}{2}$ if, and only if, $\findpath(T,R)$ computes a path of this length between $T$ and $R$.
+    \label{conjecture:findpath_maxdist_correct}
+\end{conjecture}
+
+Provided Conjecture~\ref{conjecture:findpath_maxdist_correct} holds, there is a simple algorithm for computing a tree $R$ with maximum distance from any given tree $T$.
+%TODO
+%Explain this algorithm, not only Pseudo-Code!
+
+\begin{algorithm}[H]
+\caption{MAX\_DISTANCE\_TREE($T$)}
+\label{alg:max_dist_tree}
+\begin{algorithmic}[1]
+	\STATE $R$ only contains cherry taxa $a, b$ of $T$ and has internal node with rank $n-1$
+    \STATE In list $L$ all taxa $\{1,\ldots,n\}\setminus\{a,b\}$ are sorted such that the ranks of their parents in $T$ do not decrease
+	\FOR {taxon $a$ in $L$}
+		\STATE Add an internal node as parent of $a$ on any edge in $R$ such that the new internal node has rank one less than the previously lowest internal node
+	\ENDFOR
+	\RETURN $R$
+\end{algorithmic}
+\end{algorithm}
+
+\begin{lemma}
+    Let $T$ and $R$ be trees.
+    Algorithm~\ref{alg:find_path} ($\findpath$) gives a path of length $\frac{(n-1)(n-2)}{2}$ between $T$ and $R$ if, and only if, $R$ can be generated from $T$ by applying Algorithm~\ref{alg:max_dist_tree}.
+    \label{lemma:max_dist_lemmas_equivalence}
+\end{lemma}
+
+According to Lemma~\ref{lemma:max_dist_lemmas_equivalence}, Conjecture~\ref{conjecture:findpath_maxdist_correct} is equivalence to the conjecture that two trees have maximum distance if, and only if, one can be obtained from the other by applying Algorithm~\ref{alg:max_dist_tree}.
+
+%TODO
+%proof (short explanation)
+
+If Conjecture~\ref{conjecture:findpath_maxdist_correct} is correct, it directly would follow that the radius of $\rnni$ is $rad(\rnni) = \Delta(\rnni) = \frac{(n-1)(n-2)}{2}$, where $rad(G) = \min\limits_u \max\limits_v d(u,v)$ for graph $G$ and nodes $u$ and $v$ in $G$.
 
@@
 
-\begin{conjecture}
-    Let $T$ and $R$ be trees on $n$ taxa.
-    It is $d(T,R) = \Delta(\rnni)$ if, and only if, $R$ can be produced from $T$ by applying Algorithm~\ref{alg:max_dist_tree}.
-\end{conjecture}
-
 If this conjecture is true, we have that $\Delta(\rnni) = rad(\rnni)$.
@@
 The number of caterpillar trees with distance $\Delta(\rnni)$ from a given caterpillar tree is $2^{n-2}$.
-
-\begin{algorithm}[H]
-\caption{MAX\_DISTANCE\_TREE($T$)}
-\label{alg:max_dist_tree}
-\begin{algorithmic}[1]
-	\STATE $R$ only contains cherry taxa $a, b$ of $T$ and has internal node with rank $n-1$
-    \STATE In list $L$ all taxa $\{1,\ldots,n\}\setminus\{a,b\}$ are sorted such that the ranks of their parents in $T$ do not decrease
-	\FOR {taxon $a$ in $L$}
-		\STATE Add an internal node as parent of $a$ on any edge in $R$ such that the new internal node has rank one less than the previously lowest internal node
-	\ENDFOR
-	\RETURN $R$
-\end{algorithmic}
-\end{algorithm}
-
-\begin{lemma}
-    Let $T$ and $R$ be trees.
-    Algorithm~\ref{alg:find_path} gives a path of length $\frac{(n-1)(n-2)}{2}$ between $T$ and $R$ if, and only if, $R$ can be generated from $T$ by applying Algorithm~\ref{alg:max_dist_tree}.
-\end{lemma}
 

### 2019-07-15 paper.tex
minor

@@
     Let $T$ and $R$ be caterpillar trees with caterpillar distance $l$.
-    Since $l < \frac{(n-1)(n-2)}{2}$, there is one loop $k$ in the caterpillar path from $R$ to $T$ that is given $\csort$ where less than $n-2-k$ pairs taxa swap positions.
+    We consider the shortest caterpillar path computed by $\csort$.
+    Since $l < \frac{(n-1)(n-2)}{2}$, there is one loop $k$ when running $\csort$ where less than $n-2-k$ pairs taxa swap positions.
     It follows that there is a taxon $a$ that has parent with rank $n-2-k$ in $T$ and does not need to move all the way up starting at the cherry in loop $k$.

### 2019-08-12 paper.tex
update citations, use paperpile

@@
 Now we are ready to define the tree space which is the subject of study in this paper, the $\rnni$ graph.
-Note that we will define the $\rnni$ graph on ultrametric trees where all leaves have the same rank $0$, and therefore refers to the $\rnniu$ graph as introduced in \autocite{Gavryushkin2017}.
+Note that we will define the $\rnni$ graph on ultrametric trees where all leaves have the same rank $0$, and therefore refers to the $\rnniu$ graph as introduced in \autocite{Gavryushkin2018-ol}.
 The vertex set of the \emph{$\rnni$ graph} is given by the set of all ranked trees on $n$ taxa.
@@
 We say that trees $T$ and $R$ are \emph{connected} by an \emph{$\nni$ move} move if there exist internal edges $e$ in $T$ and $f$ in $R$ both of length one such that the graphs obtained by shrinking $e$ and $f$ to internal nodes are isomorphic.
-The above definition follows \autocite{Gavryushkin2017} and is equivalent to the common definition of $\nni$ operations as given in \autocite{Steel2016-ye}.
+The above definition follows \autocite{Gavryushkin2018-ol} and is equivalent to the common definition of $\nni$ operations as given in \autocite{Steel2016-ye}.
 Trees that are connected by either a rank move or an $\nni$ move are called \emph{$\rnni$ neighbours} or \emph{adjacent in} $\rnni$.
@@
 One further result, presented in Section~\ref{section:diameter}, directly following from the proof of Theorem~\ref{thm:caterpillar_convex}, is the diameter of $\rnni$.
-Finally, we disprove the so-called Split Theorem, conjectured in \autocite{Gavryushkin2017}, and formulate an alternative version of it, the Cluster Theorem (Section~\ref{section:split_theorem}).
+Finally, we disprove the so-called Split Theorem, conjectured in \autocite{Gavryushkin2018-ol}, and formulate an alternative version of it, the Cluster Theorem (Section~\ref{section:split_theorem}).
 
@@
 
-This algorithm gives an upper bound for the diameter of $\rnni$, which improves the one given in \autocite{Gavryushkin2017}.
+This algorithm gives an upper bound for the diameter of $\rnni$, which improves the one given in \autocite{Gavryushkin2018-ol}.
 Later we will see that this bound is tight.
@@
 This type of tree is of particular interest in $\rnni$ because shortest paths between such trees in $\rnni$ differ from those in classical $\nni$ space.
-This in combination with the importance of caterpillar trees in the proof of $\np$-hardness of $\nni$ distances in \autocite{jiang2000} suggests that understanding the geometry of paths between caterpillar trees helps investigating the complexity of computing $\rnni$ distances.
+This in combination with the importance of caterpillar trees in the proof of $\np$-hardness of $\nni$ distances in \autocite{Dasgupta2000-xa} suggests that understanding the geometry of paths between caterpillar trees helps investigating the complexity of computing $\rnni$ distances.
 In particular, we are going to prove that one can always find a shortest path between two caterpillar trees that contains caterpillar trees only, meaning that the set of caterpillar trees is convex in $\rnni$.
@@
 $\nni$ moves that change the cherry of a tree refer to exchanges of the first or the second element with the third element of the sequence while exchanges of the first two elements do not change the corresponding tree.
-With this observation we can provide an algorithm for computing shortest caterpillar paths by a simple modification of the commonly known \emph{Bubble Sort} algorithm \autocite{knuth1997}.
+With this observation we can provide an algorithm for computing shortest caterpillar paths by a simple modification of the commonly known \emph{Bubble Sort} algorithm \autocite{Knuth1997-pi}.
 Throughout this paper this algorithm will be called $\csort$.
@@
     It follows that $a$ and $b$ must be neighboured taxa in $R$.
-    Let us no consider caterpillar tree $R'$ resulting from $R$ by exchanging taxa $a$ and $b$.
+    Let us now consider caterpillar tree $R'$ resulting from $R$ by exchanging taxa $a$ and $b$.
     It is clear that $R'$ has caterpillar distance $d_c(T,R') = d_c(T,R) + 1 = l+1$ to $T$ and moreover, $d(R,R') = 1$.
@@
 
-With the results of Section~\ref{section:caterpillar_convex} we can easily provide the exact diameter of the $\rnni$ graph and hence, improve the bounds for the diameter given in \autocite{Gavryushkin2017}.
+With the results of Section~\ref{section:caterpillar_convex} we can easily provide the exact diameter of the $\rnni$ graph and hence, improve the bounds for the diameter given in \autocite{Gavryushkin2018-ol}.
 Moreover, we can conclude from Lemma~\ref{lemma:caterpillar_diameter} that the maximum distance computed by algorithm $\findpath$, which was dicussed introduced above, equals the diameter of $\rnni$.
@@
 Within this section we will make a further step towards understanding the complexity of computing shortest paths in $\rnni$.
-Since the problem of computing distances in the most common tree spaces as $\nni$ \autocite{jiang2000}, $\rspr$ \autocite{Bordewich2005}, or $\tbr$ \autocite{allen2001subtree} is known to be $\np$-hard, it would be desirable to find a tree space that is less complex.
+Since the problem of computing distances in the most common tree spaces as $\nni$ \autocite{Dasgupta2000-xa}, $\rspr$ \autocite{Bordewich2005}, or $\tbr$ \autocite{Allen2001-ky} is known to be $\np$-hard, it would be desirable to find a tree space that is less complex.
 However, it is not known whether computing distances in $\rnni$ is $\np$-hard or not.
@@
 In this section we will consider an important example of shortest paths in $\nni$ that build the key of the proof of its $\np$-hardness.
-Afterwards we see that this example cannot transferred to $\rnni$ which is why the so-called Split Theorem, which generalises the example, has been conjectured in \autocite{Gavryushkin2017}.
+Afterwards we see that this example cannot transferred to $\rnni$ which is why the so-called Split Theorem, which generalises the example, has been conjectured in \autocite{Gavryushkin2018-ol}.
 However, we present a counterexample disproving the Split Theorem (Conjecture~\ref{conjecture:split_theorem}) in its original version and claiming the Cluster Theorem (Conjecture~\ref{conjecture:cluster_theorem}) as alternative version.
 
-The proof of NP-completeness of $\nni$ in \autocite{jiang2000} is based on the fact that following theorem holds in $\nni$.
+The proof of NP-completeness of $\nni$ in \autocite{Dasgupta2000-xa} is based on the fact that following theorem holds in $\nni$.
 
@@
 	\label{thm:split_nni}
-    %The formulation of this lemma is a bit different in Li1996.
-    %However, the lemma as stated here follows directly from Li1996.
+    %The formulation of this lemma is a bit different in Li1996-zw.
+    %However, the lemma as stated here follows directly from Li1996-zw.
 \end{theorem}
@@
 \begin{proof}
-	See \autocite{Li1996}.
+	See \autocite{Li1996-zw}.
 \end{proof}
 
-In the proof presented in \autocite{Li1996} for Theorem~\ref{thm:split_nni} an example of two trees that share a cluster but have no shortest path that preserves this cluster is given.
+In the proof presented in \autocite{Li1996-zw} for Theorem~\ref{thm:split_nni} an example of two trees that share a cluster but have no shortest path that preserves this cluster is given.
 The reason that this proof does not work for $\rnni$ is similar to the reason that caterpillar trees are convex in $\rnni$, but not in $\nni$.
-Both trees $T$ and $R$ in the example of \autocite{Li1996} have $2n$ taxa and consist of two caterpillar trees that share the root as parent.
+Both trees $T$ and $R$ in the example of \autocite{Li1996-zw} have $2n$ taxa and consist of two caterpillar trees that share the root as parent.
 The clusters given by the two nodes adjacent to the root are $\{1, \ldots, n\}$ and $\{n+1, \ldots, 2n\}$ in both trees $T$ and $R$.
@@
 % \todo{Shall we keep this explanation? The problem is that the $\rnni$ path depends on the assignment of ranks to the cherries produced on the path}
-% The counterexample presented in \autocite{Li1996} works in $\nni$ because of the following:
-% When proving Theorem~\ref{thm:split_nni}, \autocite{Li1996} gives an example of two trees $T$ and $R$ on $2n$ taxa that consist of two caterpillar trees each that share the root as parent.
+% The counterexample presented in \autocite{Li1996-zw} works in $\nni$ because of the following:
+% When proving Theorem~\ref{thm:split_nni}, \autocite{Li1996-zw} gives an example of two trees $T$ and $R$ on $2n$ taxa that consist of two caterpillar trees each that share the root as parent.
 % The clusters given by the two nodes adjacent to the root are $\{1, \ldots, n\}$ and $\{n+1, \ldots, 2n\}$ in both trees $T$ and $R$.
@@
 
-Since the example given in \autocite{Li1996} that proves Theorem~\ref{thm:split_nni} in $\nni$ does not work in $\rnni$, a contrary statement was conjectured in \autocite{Gavryushkin2017}.
+Since the example given in \autocite{Li1996-zw} that proves Theorem~\ref{thm:split_nni} in $\nni$ does not work in $\rnni$, a contrary statement was conjectured in \autocite{Gavryushkin2018-ol}.
 For understanding this conjecture it is important to realise the connection between edges in a tree and bipartitions of the taxon set of that tree.
@@
 Note that each edge induces such a bipartition, often written as $A|B$, and that both edges incident to the root of a tree induce the same partition.
-As these partitions are commonly referred to as \emph{splits}, the following conjecture was named \emph{Split Theorem} in \autocite{Gavryushkin2017}.
+As these partitions are commonly referred to as \emph{splits}, the following conjecture was named \emph{Split Theorem} in \autocite{Gavryushkin2018-ol}.
 
@@
 
-Since the counterexample in Figure~\ref{splitthm_counterexample} clearly shows that the version of the Split Theorem stated in \autocite{Gavryushkin2017} does not hold, we will now claim an alternative to this conjecture, the \emph{Cluster Theorem}.
+Since the counterexample in Figure~\ref{splitthm_counterexample} clearly shows that the version of the Split Theorem stated in \autocite{Gavryushkin2018-ol} does not hold, we will now claim an alternative to this conjecture, the \emph{Cluster Theorem}.
 Within the Cluster Theorem we are considering clusters instead of splits.
@@
 
-In \autocite{Gavryushkin2017} an efficient algorithm for computing the $\rnni$ graph has been introduced.
+In \autocite{Gavryushkin2018-ol} an efficient algorithm for computing the $\rnni$ graph has been introduced.
 We implemented this algorithm and could produce the graph for trees with up to $8$ taxa.

### 2019-08-12 paper.tex
some edits in section Geometry

@@
 We will identify leaves with their labels and also call them \emph{taxa} (singular: \emph{taxon}) following the systematic biology tradition.
-All nodes that are no leafs are called \emph{internal nodes} and edges $(u,v)$ are \emph{internal edge}, if both $u$ and $v$ are internal nodes.
+All nodes that are no leafs are called \emph{internal nodes} and an edge $(u,v)$ is an \emph{internal edge}, if both $u$ and $v$ are internal nodes.
 Our definition of ranked tree leads to a natural characterisation of edge lengths within ranked trees:
@@
 In this section we are investigating the geometry by analysing the structure of shortest paths in $\rnni$.
-We will start by presenting an algorithm for computing paths in $\rnni$ (Algorithm~\ref{alg:find_path}) that turns out to be an essential tool for our studies.
+We will start by presenting an algorithm for computing paths in $\rnni$ ($\findpath$) that turns out to be an essential tool for our studies.
 Using this algorithm, we are able to prove that there is always a shortest path between two caterpillar trees that only consists of caterpillar trees (Theorem~\ref{thm:caterpillar_convex}) in Section~\ref{section:caterpillar_convex}.
@@
 
-The idea of the algorithm is described in the following.
-Each node $v$ of a tree defines a \emph{cluster} $C$, which is the set of taxa descending from the $v$.
-We then say that $v$ \emph{induces} the cluster $C$.
+Before introducing the idea of this algorithm, we need further definitions.
+Each node $v$ of a tree defines a \emph{cluster} $C$, which is the set of taxa descending from $v$.
+We then say that $v$ \emph{induces} $C$.
+\emph{Trivial clusters} are clusters induced by leaves and the root of a tree.
 Note that the list of all clusters of a tree, if sorted increasingly according to the rank of the corresponding nodes, uniquely defines that tree.
-For computing a path from trees $T$ to $R$, we consider clusters $C_1, \ldots, C_{n-1}$ of $R$ where $C_i$ is induced by the node with rank $i$ for $i = 1, \ldots, n-1$.
-The so-called \emph{trivial clusters} that are induced by leaves and the root are the same in every tree and do therefore not need to be changed on any path between two trees.
-Therefore, the list of clusters $C_1, \ldots, C_{n-1}$, which is ordered increasingly according to the ranks of the corresponding nodes, uniquely defines $R$.
+For computing a path from trees $T$ to $R$ we consider clusters $C_1, \ldots, C_{n-1}$ of $R$ where $C_i$ is induced by the node with rank $i$ for $i = 1, \ldots, n-1$.
+Trivial clusters do not need to be changed on any path between two trees which is why they are not considered in the algorithm.
+This list $C_1, \ldots, C_{n-1}$ of clusters, which is ordered increasingly according to the ranks of their corresponding nodes, uniquely defines $R$.
 Starting at tree $\hat{T} = T$, $\findpath$ constructs a path where the clusters of $R$ are constructed from bottom to top, using following procedure:
-For each $C_i$, find the \emph{most recent common ancestor} $\mrca_{\hat{T}}{C_i}$ (also known as \emph{least common ancestor}) of $C_i$ in current tree $\hat{T}$, that is the internal node with lowest rank that is ancestral to all the taxa in $C_i$.
-Update $\hat{T}$ by decreasing the rank of $\mrca_{\hat{T}}{C_i}$ by an $\rnni$ move until it has the same rank $i$ as in $R$.
+For each $C_i$, find the \emph{most recent common ancestor} $\mrca_{\hat{T}}(C_i)$ (also known as \emph{least common ancestor}) of $C_i$ in current tree $\hat{T}$, that is the internal node with lowest rank that is ancestral to all the taxa in $C_i$.
+Update $\hat{T}$ by decreasing the rank of $\mrca_{\hat{T}}(C_i)$ with an $\rnni$ move until it has the same rank $i$ as in $R$.
 \todo{Do we need to explain why there is always an $\rnni$ move that moves the $\mrca$ down?}
-Since we consider the clusters of $R$ such that the ranks of the corresponding internal nodes in $R$ increase, $\mrca_{\hat{T}}{C_i}$ is the parent of cluster $C_i$ in $\hat{T}$ after these $\rnni$ moves.
+Since we consider the clusters of $R$ such that the ranks of the corresponding internal nodes in $R$ increase, $\mrca_{\hat{T}}(C_i)$ is the parent of cluster $C_i$ in $\hat{T}$ after these $\rnni$ moves.
 Thus, repeating this procedure for all clusters of $R$ results in a path between any pair of tress $T, R$, indeed.
@@
 \begin{lemma}
-    $\Delta(\rnni) \leq \frac{(n-1)(n-2)}{2}$, where $\Delta(G)$ denotes the diameter of graph $G$.
+    $\Delta(\rnni) \leq \frac{(n-1)(n-2)}{2}$, where $\Delta(G) = \max\limits_{u,v \in V(G)}d(u,v)$ denotes the diameter of graph $G$.
 \end{lemma}
@@
 \begin{proof}
-    Let us consider the maximum length of a path given by Algorithm~\ref{alg:find_path}.
-    In worst case each of the $n-2$ clusters $C$ induced by internal nodes of $R$ with ranks $i = 1, \dots, n-2$, requires $n-1-i$ $\rnni$ moves to move $\mrca(C)$ down in $\hat{T}$.
+    Let us consider the maximum length of a path computed by $\findpath$.
+    In worst case each of the $n-2$ clusters $C$ induced by internal nodes of $R$ with rank $i = 1, \dots, n-2$, requires $n-1-i$ $\rnni$ operations to move $\mrca(C)$ down in $\hat{T}$.
 	Hence, the maximum length of a path computed by that algorithm for any pair of trees is less or equal to $\sum\limits_{i = 1}^{n-2} i = \frac{(n-2)(n-1)}{2}$.
@@
 
-Lemma~\ref{lemma:distance_delete_taxon} is central for some of the following proofs using induction.
+Lemma~\ref{lemma:distance_delete_taxon} is important for some of the following proofs using induction.
 
-Central for the following inductive proofs is the comparison of the length of a shortest paths $p$ on trees on $n+1$ taxa with a path on $n$ taxa induced by $p$.
+In most of the following inductive proofs the length of a shortest paths $p$ on trees on $n+1$ taxa is compared with a path on $n$ taxa induced by $p$.
 The following lemma bounds the length of these induced paths utilising the rank differences of certain internal nodes between start and end tree of $p$.
@@
     By the following two observations one can easily see that this claim is true, indeed:
-    The rank of an internal node $v$ only changes within an $\rnni$ move if this move is a rank move of $v$ with another node or an $\nni$ move on an edge incident to $v$.
+    The rank of an internal node $v$ only changes within an $\rnni$ move if this move is a rank move of $v$ or an $\nni$ move on an edge incident to $v$.
     Furthermore, any $\rnni$ move can change the rank of an internal node by at most one.
@@
 Related to Lemma~\ref{lemma:distance_delete_taxon} is Proposition~\ref{proposition:lower_bound_distance} which provides a lower bound for the distance of two trees.
-Using the same arguments as in the proof of Lemma~\ref{lemma:distance_delete_taxon}, it easily follows that the rank difference of parents of any taxon in two trees gives such a bound.
+Using the same arguments as in the proof of Lemma~\ref{lemma:distance_delete_taxon}, it easily follows that the rank difference of parents of the same taxon but in two different trees gives such a bound.
 
@@
 
-If this conjecture is true, we have that $\Delta(\rnni) = rad(\rnni)$.
-
 According to Algorithm~\ref{alg:max_dist_tree}, the number of trees with maximum distance to a given tree $T$ depends on the number of cherries of $T$ and the ranks of their internal nodes.

### 2019-08-14 paper.tex
edits on section: caterpillar trees

@@
     $\Delta(\rnni) \leq \frac{(n-1)(n-2)}{2}$, where $\Delta(G) = \max\limits_{u,v \in V(G)}d(u,v)$ denotes the diameter of graph $G$.
+    \label{lemma:diameter_bound}
 \end{lemma}
@@
 In particular, we are going to prove that one can always find a shortest path between two caterpillar trees that contains caterpillar trees only, meaning that the set of caterpillar trees is convex in $\rnni$.
-We discovered this by computing caterpillar paths using an algorithm to compute shortest paths within the set of caterpillar trees and comparing the results with shortest path in $\rnni$.
+We discovered this by computing caterpillar paths using an algorithm to compute shortest paths within the set of caterpillar trees and comparing them with shortest path in $\rnni$.
 This algorithm will be explained in detail later in this section, after demonstrating why finding shortest paths between caterpillar trees builds the basis for understanding the difference between $\nni$ and $\rnni$.
@@
 
-Before we start analysing shortest paths between caterpillar trees, we introduce further notations.
+Before we start analysing shortest paths between caterpillar trees, we need to introduce further notations.
 A path $p$ between two caterpillar trees $T$ and $R$ that only consists of caterpillar trees is called \emph{caterpillar path}.
-If $p$ is shortest among all caterpillar paths between these trees we call it \emph{shortest caterpillar path}, and the length of $p$ will be denoted by $d_c(T,R)$ and is called \emph{caterpillar distance}.
+If $p$ is shortest among all caterpillar paths between these trees we call it \emph{shortest caterpillar path}, and the length of $p$ will be denoted by $d_c(T,R)$ and called \emph{caterpillar distance}.
 
-Let us compare shortest paths between caterpillar trees in $\nni$ with those in $\rnni$.
-The essential difference is that building a cherry and moving it may give a path that is shorter than a path where the taxa move separately in $\nni$ as it is illustrated out in Figure~\ref{fig:NNI_vs_RNNI}.
-There is only one move needed in $\nni$ to build and resolve a cherry, respectively, while the number of moves for moving the cherry is the same as for moving a single taxon.
-However, in $\rnni$ space, there are additional rank moves needed for the internal node of the cherry before an $\nni$ move resolves it, because $\nni$ moves are only allowed on edges of length one in $\rnni$.
+We now compare shortest paths between caterpillar trees in $\nni$ with those in $\rnni$.
+In $\nni$, building a cherry and moving it may give a path that is shorter than a path where the taxa move separately in as it is illustrated out in Figure~\ref{fig:NNI_vs_RNNI}.
+There is only one move needed to build and resolve a cherry, respectively, while the number of moves for moving the cherry is the same as for moving a single taxon in $\nni$.
+But in $\rnni$ space there are additional rank moves needed to change the rank of the internal node of the cherry before an $\nni$ move resolves it.
+This is due to the fact that $\nni$ moves are only allowed on edges of length one in $\rnni$.
 Therefore, a path where taxa move separately and no second cherry is built, is shortest in $\rnni$.
@@
 
-The above observation already suggests that there is always a caterpillar path that is shortest path between two arbitrary caterpillar trees in $\rnni$.
+The above observation suggests that there is always a caterpillar path that is shortest path between two arbitrary caterpillar trees in $\rnni$.
 Indeed, we are going to prove Theorem~\ref{thm:caterpillar_convex} within this section, which states that the set of caterpillar trees is convex in $\rnni$.
@@
 
+We next introduce a unique representation of caterpillar trees to show how one can use the classical \emph{Bubble Sort} algorithm \autocite{Knuth1997-pi} to compute caterpillar paths.
 One can see caterpillar trees as sequences of taxa, ordered non-decreasing according to the ranks of their parents.
 $\nni$ moves between caterpillar trees, which are the only possible $\rnni$ moves between those trees, coincide with exchanges of two neighboured elements in the sequence.
-In these sequences the first two taxa are somehow equivalent since they build the unique cherry.
-$\nni$ moves that change the cherry of a tree refer to exchanges of the first or the second element with the third element of the sequence while exchanges of the first two elements do not change the corresponding tree.
-With this observation we can provide an algorithm for computing shortest caterpillar paths by a simple modification of the commonly known \emph{Bubble Sort} algorithm \autocite{Knuth1997-pi}.
+In these sequences the first two taxa are somehow equivalent since they build the unique cherry which means they share their parent with rank one.
+Therefore, we need to consider $\nni$ moves that change the cherry of a tree separately.
+These moves refer to exchanges of the first or the second element with the third element of the sequence, while exchanges of the first two elements do not change the corresponding tree.
+With this observation we can provide an algorithm for computing shortest caterpillar paths by a simple modification of the commonly known Bubble Sort algorithm.
 Throughout this paper this algorithm will be called $\csort$.
@@
 Following the observations above, only swaps of the first three elements of a sequence with Bubble Sort need to be adapted for the computation of shortest caterpillar paths.
-Let us consider the first three elements $a,b,c$ which appear in this order in a sequence representing a caterpillar tree on a caterpillar path between trees $T$ and $R$.
+Let us consider the first three elements $a,b,c$, appearing in this order in a sequence representing a caterpillar tree on a caterpillar path between trees $T$ and $R$.
 If the rank of $\parent(c)$ is higher than the ranks of $\parent(a)$ and $\parent(b)$ in $R$, there are no swaps needed for these taxa.
-If just one of the two taxa $a$ and $b$ is on the other side of $c$ in $R$, this one swaps position with $c$ in the sequence.
-But if both $a$ and $b$ are on the other side of $c$ in $R$, the one that has the parent with highest rank of all these three taxa in $R$ swaps with $c$ at first.
-This makes sure that there is no additional swap of $a$ and $b$ needed on the path to $R$.
+If just one of the two taxa $a$ and $b$ is on the other side of $c$ in $R$, this taxon swaps position with $c$ in the sequence.
+And if both $a$ and $b$ are on the other side of $c$ in $R$, the taxon that has the parent with highest rank of all these three taxa in $R$ swaps with $c$ at first.
+This makes sure that there is no additional swap of $a$ and $b$ needed on the subsequent path to $R$.
 \todo{Do we need to explain Bubble Sort?}
@@
 Lemma~\ref{lemma:caterpillar_dist=diameter} regards caterpillar trees with distance $\frac{(n-1)(n-2)}{2}$ and will be needed for proving the convexity of the set of caterpillar trees in $\rnni$.
+Notice that we by Lemma~\ref{lemma:diameter_bound} $\frac{(n-1)(n-2)}{2}$ is the upper bound for the diameter of $\rnni$.
 
@@
     It follows that the maximum length of a path computed by this algorithm is $\sum\limits_{k=1}^{n-2} k = \frac{(n-1)(n-2)}{2}$.
-    Since there cannot be a caterpillar path shorter than $\frac{(n-1)(n-2)}{2}$ if it is $d(T,R) = \frac{(n-1)(n-2)}{2}$, it follows $d_c(T,R) = \Delta(\rnniu)$.
+    Since there cannot be a caterpillar path shorter than $\frac{(n-1)(n-2)}{2}$ if it is $d(T,R) = \frac{(n-1)(n-2)}{2}$, it follows $d_c(T,R) =  \frac{(n-1)(n-2)}{2}$.
 
@@
     We prove $d(T,R) = \frac{(n-1)(n-2)}{2}$ by induction on the number of taxa $n$ of $T$ and $R$:
-    Without loss of generality we can assume that $T$ is the caterpillar tree $(\ldots ((1,2),3), \ldots, n)$.
+    Without loss of generality we can assume that $T$ is the caterpillar tree $(\ldots ((1,2),3), \ldots, n)$ or $(\ldots ((1,2),3), \ldots, n+1)$ in the induction step.
     Let $R$ be a caterpillar tree with distance $\frac{(n-1)(n-2)}{2}$ to $T$.
@@
     Induction basis: $n=3$\\
-    All three trees have distance $1$ from another and are caterpillar trees, hence it is $d(T,R) = d_c(T,R)$ for all trees $T,R$ on $n=3$ taxa.
+    There are three trees on $3$ taxa that have distance $1$ from another and are caterpillar trees.
+    Hence it is $d(T,R) = d_c(T,R) = 1$ for all pairs of trees $T,R$ on $n=3$ taxa.
 
     Induction step: $n \to n+1$\\
-    Induction hypothesis: For trees $T, R$ on $n$ taxa it is $d(T, R) = \frac{n(n-1)}{2}$.\\
-    Let $T, R$ be trees on $n+1$ taxa with caterpillar distance $\frac{(n-1)(n-2)}{2}$.
-    Therefore, each of the $\frac{(n-1)(n-2)}{2}$ pairs of taxa that is considered when running $\csort$ to compute a path $p$ from $R$ to $T$ exchanges.
-    From this we can conclude that taxon $n+1$, whose parent has highest rank in $T$, is part of the cherry of $R$.
+    Induction hypothesis: For trees $T, R$ on $n$ taxa it is $d(T, R) = \frac{(n-1)(n-2)}{2}$.\\
+    Let $T, R$ be trees on $n+1$ taxa with caterpillar distance $\frac{n(n-1)}{2}$.
+    Therefore, each of the $\frac{n(n-1)}{2}$ pairs of taxa that is considered when running $\csort$ to compute a path $p$ from $R$ to $T$ exchanges.
+    From this we can conclude that taxon $n+1$, whose parent has highest rank in $T$, is part of the cherry of $R$. \todo{Do we need to explain this further (figure)?}
     Let us consider the trees $T{\big|}_n$ to $R{\big|}_n$ that result from $T$ and $R$ by deleting taxon $n+1$.
-    $\csort$ computes a path from $T{\big|}_n$ to $R{\big|}_n$ that contains the same exchanges of taxa as $p$, except for the $n-1$ first ones on $p$ that involve taxon $n+1$.
+    $\csort$ computes a path from $T{\big|}_n$ to $R{\big|}_n$ that contains the same exchanges of taxa as $p$, except for the $n-1$ ones on $p$ that involve taxon $n+1$.
     Hence, the caterpillar distance is $d_c(T{\big|}_n, R{\big|}_n) = \frac{n(n-1)}{2} - (n-1)$ = $\frac{(n-1)(n-2)}{2}$.
@@
     According to Lemma~\ref{lemma:distance_delete_taxon}, the distance between $T{\big|}_n$ and $R{\big|}_n$ is at least $d(T{\big|}_n, R{\big|}_n) \leq d(T,R) - |\rank_T(\parent(n+1)) - \rank_R(\parent(n+1))|$.
-    We also know $|\rank_T(\parent(n+1)) - \rank_R(\parent(n+1))| = n-1$
-    Suppose, contrary to our claim, that there is a path from $T$ to $R$ of length less than $\Delta(\rnniu_{n+1}) = \frac{n(n-1)}{2}$.
+    We also know $|\rank_T(\parent(n+1)) - \rank_R(\parent(n+1))| = n-1$.
+    Suppose, contrary to our claim, that there is a path from $T$ to $R$ of length less than $\frac{n(n-1)}{2}$.
     By the above, it follows that $d(T{\big|}_n, R{\big|}_n) \leq d(T,R) - |\rank_T(\parent(n+1)) - \rank_R(\parent(n+1))| < \frac{n(n-1)}{2} - (n-1) < \frac{(n-1)(n-2)}{2}$.
     Since this is a contradiction to $d(T{\big|}_n, R{\big|}_n) = \frac{(n-1)(n-2)}{2}$, it is proven that $d(T,R) = \frac{n(n-1)}{2} $.
-
 \end{proof}
@@
 \begin{proof}
-    We prove the Theorem by backwards induction on the length $l$ of a shortest caterpillar path between two caterpillar trees $T$ and $R$.
+    We prove this Theorem by backwards induction on the length $l$ of a shortest caterpillar path between two caterpillar trees $T$ and $R$.
+    Since we use backwards induction, we need to consider the maximum distance between any two caterpillar trees as base case, which is $\frac{(n-1)(n-2)}{2}$ as we saw above.
 
     Induction basis: $l = \frac{(n-1)(n-2)}{2}$\\
-    With Lemma~\ref{lemma:caterpillar_dist=diameter} one can see that for each pair of caterpillar trees $T$, $R$ with caterpillar distance $d_c(T,R) = \frac{(n-1)(n-2)}{2}$, it is $d(T,R) = d_c(T,R)$, which means that there is a shortest path between $T$ and $R$ that is a caterpillar path.
+    With Lemma~\ref{lemma:caterpillar_dist=diameter} one can see that for each pair of caterpillar trees $T$, $R$ with caterpillar distance $d_c(T,R) = \frac{(n-1)(n-2)}{2}$, it is $d(T,R) = d_c(T,R)$.
+    Therefore, there is a shortest path between $T$ and $R$ that is a caterpillar path.
 
@@
     We consider the shortest caterpillar path computed by $\csort$.
-    Since $l < \frac{(n-1)(n-2)}{2}$, there is one loop $k$ when running $\csort$ where less than $n-2-k$ pairs taxa swap positions.
-    It follows that there is a taxon $a$ that has parent with rank $n-2-k$ in $T$ and does not need to move all the way up starting at the cherry in loop $k$.
-    Let $b$ be the taxon whose parent has rank one less than the parent of $a$ in the tree at the beginning of loop number $k$.
-    We can assume without loss of generality that loop $k$ is the first loop in which the number of swaps is not maximal.
-    It follows that $a$ and $b$ must be neighboured taxa in $R$.
-    Let us now consider caterpillar tree $R'$ resulting from $R$ by exchanging taxa $a$ and $b$.
-    It is clear that $R'$ has caterpillar distance $d_c(T,R') = d_c(T,R) + 1 = l+1$ to $T$ and moreover, $d(R,R') = 1$.
-    Hence we can apply the induction hypothesis on $T$ and $R'$ and see that $d(T,R') = l+1$.
-    If it were $d(T,R) < l$, we would have $d(T,R') \leq d(T,R) + d(R,R') < l + 1$ which contradicts the induction hypothesis.
+    Since $l < \frac{(n-1)(n-2)}{2}$, there is one loop $k$ when running $\csort$ where less than $n-2-k$ taxon pairs swap positions.
+    We assume that this loop number $k$ is the first one in which this occurs.
+    This situation is illustrated in the top of Figure~\ref{fig:cat_convex_seq_example}.
+    Whenever the number of swaps of taxa or, in terms of sequences, neighboured elements in the sequence is maximum in a loop, one of the two first elements swaps position with $n-2-k$ subsequent elements of the sequence.
+    Since we assume that the number of swaps in loop $k$ is not maximal, this does not happen.
+    Let us assume that $a$ is the first element that is compared with its left neighbour in loop $k$ but does not swap with it.
+    Then the left neighbour $b$ of $a$ in $T$ is below $a$ in $R$ as well.
+    Therefore, the tree $T'$ which can be received from $T$ by exchanging $a$ and $b$ must have caterpillar distance $d_c(T',R) = d_c(T,R) + 1 = l + 1$ to $R$.
+    For an illustration of the $k^{th}$ loop when performing $\csort$ between $T$ and $R$ or $T'$ and $R$, respectively, consider Figure~\ref{fig:cat_convex_seq_example}.
+
+    \begin{figure}[H]
+    	\centering
+    	\includegraphics[width=\textwidth]{cat_convex_seq_example}
+    	\caption{Loop number $k$ of $\csort$ between $T$ and $R$ as described in proof of Theorem~\ref{thm:caterpillar_convex}.
+        The trees are represented as sequences that are being ordered by $\csort$.
+        Right before loop number $k$, $x_1, \ldots, x_{k-1}$ are already at the same position as in $R$.
+        At the top: sequences between $T$ and $R$ -- $x_k$ and $b$ swap in loop number $k$, but $x_k$ and $a$ not.
+        At the bottom: sequences between $T'$ and $R$ -- $a$ and $b$ swap in loop number $k$; $x_k$ and $b$ will swap later}
+    	\label{fig:cat_convex_seq_example}
+    \end{figure}
+
+    Since it is clear that $T'$ has caterpillar distance $d_c(T',R) = l+1$ to $T$, we can apply the induction hypothesis on $T'$ and $R$ and see that $d(T',R) = l+1$.
+    Furthermore, it is $d(T,T') = 1$.
+    If it now were $d(T,R) < l$, it would hold $d(T',R) \leq d(T,R) + d(T,T') < l + 1$ which contradicts the induction hypothesis.
     We can conclude $d(T,R) = l$, which completes the proof.

### 2019-08-14 paper.tex
minor on diameter section

@@
 
-With the results of Section~\ref{section:caterpillar_convex} we can easily provide the exact diameter of the $\rnni$ graph and hence, improve the bounds for the diameter given in \autocite{Gavryushkin2018-ol}.
-Moreover, we can conclude from Lemma~\ref{lemma:caterpillar_diameter} that the maximum distance computed by algorithm $\findpath$, which was dicussed introduced above, equals the diameter of $\rnni$.
-This suggests that it gives reasonable estimates of distances.
-For further details on the quality of that algorithm we refer the reader to Section~\ref{section:computation}.
+With the results of Section~\ref{section:caterpillar_convex} we can easily provide the exact diameter of the $\rnni$ graph and hence improve the bounds for the diameter given in \autocite{Gavryushkin2018-ol}.
+Moreover, we can conclude from Lemma~\ref{lemma:caterpillar_diameter} that the maximum distance computed by algorithm $\findpath$, introduced in Section~\ref{section:geometry}, equals the diameter of $\rnni$.
+This suggests that this algorithm computes reasonable estimates of distances.
+For further details on $\findpath$ we refer the reader to Section~\ref{section:computation}.
 
@@
     According to Theorem~\ref{thm:caterpillar_convex}, there is a shortest path between $T$ and $R$ that is a caterpillar path.
-    We can compute a shortest caterpillar path by using $\csort$ as explained above.
-    When running this algorithm on $T$ and $R$, taxon $i$ swaps $n-1-i$ times with its right neighbour.
+    We can compute a shortest caterpillar path by using $\csort$ as explained in Section~\ref{section:caterpillar_convex}.
+    When running this algorithm on $T$ and $R$, each taxon $i$ ($i \in \{1, \ldots, n\}$) swaps $n-1-i$ times with its right neighbour.
     Therefore, the length of a shortest caterpillar path is $\sum\limits_{i=1}^{n-1}(n-1-i) = \sum\limits_{i=0}^{n-2}i = \frac{(n-1)(n-2)}{2}$.

### 2019-08-14 paper.tex
minor on section Split Theorem

@@
 \newcommand{\tbr}{\mathrm{TBR}}
-\newcommand{\rspr}{\mathrm{rSPR}}
+\newcommand{\spr}{\mathrm{SPR}}
 \newcommand{\csort}{\textsc{Caterpillar Sort}}
@@
 Within this section we will make a further step towards understanding the complexity of computing shortest paths in $\rnni$.
-Since the problem of computing distances in the most common tree spaces as $\nni$ \autocite{Dasgupta2000-xa}, $\rspr$ \autocite{Bordewich2005}, or $\tbr$ \autocite{Allen2001-ky} is known to be $\np$-hard, it would be desirable to find a tree space that is less complex.
-However, it is not known whether computing distances in $\rnni$ is $\np$-hard or not.
-The natural approach to solve this question is to compare $\rnni$ and $\nni$ graph to see whether the reason that $\nni$ is $\np$-hard transfers to $\rnni$.
-In this section we will consider an important example of shortest paths in $\nni$ that build the key of the proof of its $\np$-hardness.
-Afterwards we see that this example cannot transferred to $\rnni$ which is why the so-called Split Theorem, which generalises the example, has been conjectured in \autocite{Gavryushkin2018-ol}.
+Since the problem of computing distances in the most common tree spaces as $\nni$ \autocite{Dasgupta2000-xa}, rooted $\spr$ \autocite{Bordewich2005-nx}, or $\tbr$ \autocite{Allen2001-ky} is known to be $\np$-hard, it is desirable to find a tree space that is less complex.
+So far it is not known whether computing distances in $\rnni$ is $\np$-hard or not.
+The natural approach to solve this question is to compare $\rnni$ and $\nni$ graph to see whether the proof of $\np$-hardness of $\nni$ can be transferred to $\rnni$.
+Within this section we will explain why this is not the case.
+Therefore we consider an example of shortest paths in $\nni$ that builds the key of the proof of its $\np$-hardness.
+We will see that this example cannot be transferred to $\rnni$.
+Following this, we will introduce the so-called Split Theorem, which generalises that example, and has been conjectured in \autocite{Gavryushkin2018-ol}.
 However, we present a counterexample disproving the Split Theorem (Conjecture~\ref{conjecture:split_theorem}) in its original version and claiming the Cluster Theorem (Conjecture~\ref{conjecture:cluster_theorem}) as alternative version.
 
-The proof of NP-completeness of $\nni$ in \autocite{Dasgupta2000-xa} is based on the fact that following theorem holds in $\nni$.
+The proof of NP-completeness of $\nni$ in \autocite{Dasgupta2000-xa} is based on the fact that Theorem~\ref{thm:split_nni} holds for $\nni$.
 
@@
 
-% ^How do we assign the ranks to the new cherry nodes?
+% ^How do we assign ranks to the new cherry nodes?
 % We could either create a path where $R$ and $R'$ have distance equal to $d_{\rnni}(T{\big|}_{\{1, \ldots, n\}}, R{\big|}_{\{n+1, \ldots, 2n\}})$ and have a lot of additional rank moves before and after that part OR
@@
 Since the example given in \autocite{Li1996-zw} that proves Theorem~\ref{thm:split_nni} in $\nni$ does not work in $\rnni$, a contrary statement was conjectured in \autocite{Gavryushkin2018-ol}.
-For understanding this conjecture it is important to realise the connection between edges in a tree and bipartitions of the taxon set of that tree.
+For understanding this conjecture it is important to realise the connection between edges in a tree and so-called splits, which are bipartitions of the taxon set of that tree:
 By deleting an edge $e$ of a tree $T$ one receives two trees on taxon sets $A$ and $B$ such that $\{A,B\}$ is a partition of the taxon set of $T$.
 Note that each edge induces such a bipartition, often written as $A|B$, and that both edges incident to the root of a tree induce the same partition.
-As these partitions are commonly referred to as \emph{splits}, the following conjecture was named \emph{Split Theorem} in \autocite{Gavryushkin2018-ol}.
+Since these partitions are commonly referred to as \emph{splits}, the following conjecture was named \emph{Split Theorem} in \autocite{Gavryushkin2018-ol}.
 
@@
 
-However, we found a simple counterexample to this conjecture:
+However, we found a simple counterexample to this conjecture, provided in Figure~\ref{fig:splitthm_counterexample}.
 
@@
 	\caption{The split $123|45$ is present in $T$ and $R$, but the path at the top is a shortest path where none of the trees contains this split.
-    However, this split is maintained on the path at the bottom that is a shortest path as well.}
-	\label{splitthm_counterexample}
+    However, this split is maintained on the path at the bottom, which is a shortest path as well.}
+	\label{fig:splitthm_counterexample}
 \end{figure}
 
-Since the counterexample in Figure~\ref{splitthm_counterexample} clearly shows that the version of the Split Theorem stated in \autocite{Gavryushkin2018-ol} does not hold, we will now claim an alternative to this conjecture, the \emph{Cluster Theorem}.
+Since the counterexample in Figure~\ref{fig:splitthm_counterexample} shows that the version of the Split Theorem stated in \autocite{Gavryushkin2018-ol} does not hold, we will now claim an alternative to this conjecture, the \emph{Cluster Theorem} (Conjecture~\ref{conjecture:cluster_theorem}).
 Within the Cluster Theorem we are considering clusters instead of splits.
 This is motivated by the fact that rooted phylogenetic trees can uniquely be represented by sets of clusters \autocite{Steel2016-ye}, but not by sets of splits as they cannot define the position of the root.
-Moreover, a ranked tree can be uniquely determined by ordering the set of clusters increasingly according to the ranks of the internal nodes inducing those clusters.
+Moreover, a ranked tree can be uniquely determined by ordering the clusters such that the ranks of the internal nodes inducing those clusters increase.
 
@@
 
-Note that the trees $T, R$ of counterexample~\ref{splitthm_counterexample} share the same set of induced splits, but have distinct sets of clusters.
-Furthermore, the example used in $\rnni$ to show that there are trees that share splits that are not present on any shortest path between these trees can be used to prove that the Cluster Theorem does not hold in $\nni$.
+Note that trees $T$ and $R$ of counterexample~\ref{fig:splitthm_counterexample} induce the same set of splits, but have distinct clusters.
+Furthermore, this counterexample can be used to prove that the Cluster Theorem does not hold in $\nni$.
 

### 2019-08-20 paper.tex
edit alg:max_dist_tree -- found counterexample for arbitrary trees

@@
 In \autocite{Gavryushkin2018-ol} an efficient algorithm for computing the $\rnni$ graph has been introduced.
-We implemented this algorithm and could produce the graph for trees with up to $8$ taxa.
+We implemented this algorithm and could produce the graph for trees with up to $8$ taxa. \todo{Shall we say more about the algorithm here?}
 This enables us to compute exact distances between trees of this size.
-It is in particular possible to compare the exact distances with distances we receive from algorithms we introduced previously and to verify the Cluster Theorem, which was raised in Section~\ref{section:split_theorem}, for small trees.
+In particular, this enables us to compare exact distances with distances we receive from algorithms introduced in previous sections and to verify the Cluster Theorem, which was raised in Section~\ref{section:split_theorem}, for small trees.
 All algorithms have been implemented with Python and are available online. \todo{link?}
@@
 Thus, if the distance between two trees equals the diameter, $\findpath$ computes the correct distance.
-Note that central for proving the convexity of the set of caterpillar trees in Section~\ref{section:caterpillar_convex} was Lemma~\ref{lemma:caterpillar_dist=diameter}.
-We therefore aim to find a property for any pair of trees, not only caterpillar trees, with maximum distance $\Delta(\rnni) = \frac{(n-1)(n-2)}{2}$.
-Following algorithm $\findpath$, we claim following conjecture:
+Following this observation, we claim following conjecture:
 
@@
 
-Provided Conjecture~\ref{conjecture:findpath_maxdist_correct} holds, there is a simple algorithm for computing a tree $R$ with maximum distance from any given tree $T$.
+Besides the diameter, the \emph{radius} of a graph $G$, which is defined by $rad(G) =  \min\limits_u \max\limits_v d(u,v)$ is a further interesting property of a graph.
+In Conjecture~\ref{conjecture:radius} we claim that the radius of $\rnni$ equals its diameter.
+Algorithm~\ref{alg:max_dist_tree} suggests this, since it computes for an arbitrary input tree $T$ a caterpillar tree $R$ with distance $d(T,R) = \frac{(n-1)(n-2)}{2}$ for $n \leq k$.
+\todo{$k = ?$, check computationally (or prove this -> see conjecture below)}
+
+\begin{conjecture}
+    It is $rad(\rnni) = \Delta(\rnni) = \frac{(n-1)(n-2)}{2}$.
+    \label{conjecture:radius}
+\end{conjecture}
+
 %TODO
-%Explain this algorithm, not only Pseudo-Code!
+%Explain the following algorithm, not only Pseudo-Code!
 
@@
 	\FOR {taxon $a$ in $L$}
-		\STATE Add an internal node as parent of $a$ on any edge in $R$ such that the new internal node has rank one less than the previously lowest internal node
+		\STATE Add an internal node as parent of $a$ on an edge in $R$ such that the new internal node has rank one less than the previously lowest internal node and $R$ is a caterpillar tree
 	\ENDFOR
@@
 
-\begin{lemma}
-    Let $T$ and $R$ be trees.
+\begin{conjecture}
+    Let $T$ be an arbitrary tree and $R$ a caterpillar tree.
     Algorithm~\ref{alg:find_path} ($\findpath$) gives a path of length $\frac{(n-1)(n-2)}{2}$ between $T$ and $R$ if, and only if, $R$ can be generated from $T$ by applying Algorithm~\ref{alg:max_dist_tree}.
-    \label{lemma:max_dist_lemmas_equivalence}
-\end{lemma}
-
-According to Lemma~\ref{lemma:max_dist_lemmas_equivalence}, Conjecture~\ref{conjecture:findpath_maxdist_correct} is equivalence to the conjecture that two trees have maximum distance if, and only if, one can be obtained from the other by applying Algorithm~\ref{alg:max_dist_tree}.
+    \label{conjecture:max_dist_caterpillar}
+\end{conjecture}
 
 %TODO
-%proof (short explanation)
-
-If Conjecture~\ref{conjecture:findpath_maxdist_correct} is correct, it directly would follow that the radius of $\rnni$ is $rad(\rnni) = \Delta(\rnni) = \frac{(n-1)(n-2)}{2}$, where $rad(G) = \min\limits_u \max\limits_v d(u,v)$ for graph $G$ and nodes $u$ and $v$ in $G$.
+%prove this conjecture! (<= would be sufficient for the conjecture above!)
 

### 2019-08-20 paper.tex
use alg:max_dist to prove radius=diameter

@@
     It is $\Delta(\rnniu) = \frac{(n-1)(n-2)}{2}$.
+    \label{corollary:diameter}
 \end{corollary}
@@
 Besides the diameter, the \emph{radius} of a graph $G$, which is defined by $rad(G) =  \min\limits_u \max\limits_v d(u,v)$ is a further interesting property of a graph.
-In Conjecture~\ref{conjecture:radius} we claim that the radius of $\rnni$ equals its diameter.
-Algorithm~\ref{alg:max_dist_tree} suggests this, since it computes for an arbitrary input tree $T$ a caterpillar tree $R$ with distance $d(T,R) = \frac{(n-1)(n-2)}{2}$ for $n \leq k$.
-\todo{$k = ?$, check computationally (or prove this -> see conjecture below)}
-
-\begin{conjecture}
-    It is $rad(\rnni) = \Delta(\rnni) = \frac{(n-1)(n-2)}{2}$.
-    \label{conjecture:radius}
-\end{conjecture}
+We will use Algorithm~\ref{alg:max_dist_tree} to show that in $\rnni$ the radius equals the diameter.
 
@@
 	\FOR {taxon $a$ in $L$}
-		\STATE Add an internal node as parent of $a$ on an edge in $R$ such that the new internal node has rank one less than the previously lowest internal node and $R$ is a caterpillar tree
+		\STATE Add an internal node as parent of $a$ on an edge in $R$ such that the new internal node has rank one less than the previously lowest internal node
 	\ENDFOR
@@
 
-\begin{conjecture}
-    Let $T$ be an arbitrary tree and $R$ a caterpillar tree.
-    Algorithm~\ref{alg:find_path} ($\findpath$) gives a path of length $\frac{(n-1)(n-2)}{2}$ between $T$ and $R$ if, and only if, $R$ can be generated from $T$ by applying Algorithm~\ref{alg:max_dist_tree}.
-    \label{conjecture:max_dist_caterpillar}
-\end{conjecture}
+\begin{lemma}
+    Let $T$ be a caterpillar tree.
+    A tree $R$ that is computed by applying Algorithm~\ref{alg:max_dist_tree} on $R$ has distance $d(T,R) = \Delta(\rnni)$ to $T$.
+    \label{lemma:max_dist_caterpillar}
+\end{lemma}
 
-%TODO
-%prove this conjecture! (<= would be sufficient for the conjecture above!)
+\begin{proof}
+    Recall that with Corollary~\ref{corollary:diameter} we know that $\Delta(\rnni) = \frac{(n-1)(n-2)}{2}$.
+    Within this proof we assume, without loss of generality, that it is $T = (( \dots (1,2)3)\dots)n)$
+    We prove the lemma by induction on the number of taxa $n$.
+
+    Induction basis: $n = 3$\\
+    Algorithm~\ref{alg:max_dist_tree} computes either $R = ((1,3),2)$ or $R'= ((2,3),1)$ when applied to $T = ((1,2),3)$.
+    Since it is $d(T,R) = d(T,R') = 1 = \Delta(\rnni_3)$, the lemma is true for trees on $n=3$ taxa.
+
+    Induction step: $n \to n+1$\\
+    Induction hypothesis: For a caterpillar tree $T$ on $n$ taxa Algorithm~\ref{alg:max_dist_tree} computes a tree $R$ with distance $d(T,R) = \frac{(n-1)(n-2)}{2}$.\\
+    Let us now  assume that $T$ is a tree on $n+1$ taxa.
+    In the list $L$ produced in Algorithm~\ref{alg:max_dist_tree}, when applied to $T$, taxon $n+1$ is the last one.
+    Hence $n+1$ is added to $R$ in the last step and therefore, it is in the lowest cherry of $R$ and has parent with rank $1$.
+    Because the parent of $n+1$ has rank $n$ in $T$, we can conclude from Lemma~\ref{lemma:distance_delete_taxon} that for $T$ and $R$ restricted to taxa $1,\ldots,n$ it holds: $d(T|_n,R|_n) \leq d(T,R) - (n-1)$.
+    It is easy to see that $R|_n$ can be received from $T|_n$ by applying Algorithm~\ref{alg:max_dist_tree}.
+    So we can follow from the induction hypothesis and Lemma~\ref{lemma:distance_delete_taxon} that it is $d(T,R) \geq d(T|_n,R|_n) + (n-1) = \frac{(n-1)(n-2)}{2} = \frac{n(n-1)}{2}$, which concludes the proof of the lemma.
+\end{proof}
+
+In Algorithm~\ref{alg:max_dist_tree} it is not uniquely determined, where a new taxon is added to the already existing tree.
+Instead, the attachment branch can be randomly chosen with the only restriction being that the rank of the new node needs to be below the ranks of already existing internal nodes.
+This is the reason why one can receive every tree topology on $n$ taxa when applying Algorithm~\ref{alg:max_dist_tree} onto a caterpillar tree.
+And because permuting the labels of the trees does not change the distance, there is for every tree $R$ on $n$ taxa a caterpillar tree $T$ with distance $d(T,R) = \Delta(\rnni) = \frac{(n-1)(n-2)}{2}$.
+This gives us Corollary~\ref{corollary:radius}.
+
+\begin{corollary}
+    The radius of the $\rnni$ graph is $rad(\rnni) = \Delta(\rnni) = \frac{(n-1)(n-2)}{2}$.
+    \label{corollary:radius}
+\end{corollary}
 

### 2019-08-21 paper.tex
restructure: diameter and radius section

@@
 \newcommand{\findpath}{\textsc{FindPath}}
+\newcommand{\mdtree}{\textsc{MDTree}}
 
@@
 
-\subsection{Diameter}
+\subsection{Diameter and radius}
 \label{section:diameter}
@@
 With the results of Section~\ref{section:caterpillar_convex} we can easily provide the exact diameter of the $\rnni$ graph and hence improve the bounds for the diameter given in \autocite{Gavryushkin2018-ol}.
-Moreover, we can conclude from Lemma~\ref{lemma:caterpillar_diameter} that the maximum distance computed by algorithm $\findpath$, introduced in Section~\ref{section:geometry}, equals the diameter of $\rnni$.
-This suggests that this algorithm computes reasonable estimates of distances.
-For further details on $\findpath$ we refer the reader to Section~\ref{section:computation}.
+Furthermore, we are able to show that the \emph{radius} of the $\rnni$ graph, which is defined as $rad(G) =  \min\limits_u \max\limits_v d(u,v)$ for a graph $G$, equals its diameter.
+Let us first consider the diameter of the $\rnni$ graph.
+With the results of Section~\ref{section:caterpillar_convex} is not hard to see that is $\Delta(\rnni) = \frac{(n-1)(n-2))}{2}$ as it is stated in Corollary~\ref{corollary:diameter}.
 
@@
     \label{corollary:diameter}
+\end{corollary}
+
+We now proceed to show that the radius of the $\rnni$ graph equals its diameter.
+For this aim an algorithm for constructing a tree $R$ with distance $\Delta(\rnni)$ from a given caterpillar tree $T$ will be used.
+This algorithm will be called $\mdtree$ (Maximum Distance Tree).
+We will prove that a tree $R = \mdtree(T)$ has distance $d(T,R) = \Delta(\rnni)$ to $T$ if $T$ is a caterpillar tree, after introducing the algorithm.
+
+$\mdtree$ works as follows:
+At first, all taxa of $T$ are sorted in a list $L$ such that the ranks of their parents are non-decreasing.
+If the input is a caterpillar tree, it follows that $L[0]$ and $L[1]$ are the cherry taxa of $T$ and $\parent(L[i])$ has rank $i$ in $T$ for all $i = 2, \ldots, n-1$.
+Now $R$ is built inductively as follows:
+Initially, $R$ only consists of a cherry with taxa $L[0]$ and $L[1]$ such that the parent of these taxa has rank $n-1$.
+All following taxa in $L$ are added to $R$ sequentially, according to their order in $L$, by creating a new internal node at one of the existing branches in $R$ such that this new node has rank one less than the previously lowest internal node in $R$.
+When this is done for all taxa in $L$, the resulting tree $R$ is an $\rnni$ tree.
+
+It is important to note that this algorithm can be applied to any input tree $T$, which does not need to be a caterpillar tree.
+However, the results provided in this section are only true for caterpillar trees.
+The pseudocode for $\mdtree$ can be found in Section~\ref{section:computation}.
+\todo{figure? + do we need the pseudocode?}
+
+\begin{lemma}
+    Let $T$ be a caterpillar tree.
+    A tree $R = \mdtree(T)$ has distance $d(T,R) = \Delta(\rnni)$ to $T$.
+    \label{lemma:max_dist_caterpillar}
+\end{lemma}
+
+\begin{proof}
+    Recall that with Corollary~\ref{corollary:diameter} we know that $\Delta(\rnni) = \frac{(n-1)(n-2)}{2}$.
+    Within this proof we assume, without loss of generality, that it is $T = (( \dots (1,2)3)\dots)n)$
+    We prove the lemma by induction on the number of taxa $n$.
+
+    Induction basis: $n = 3$\\
+    $\mdtree(T)$ results either in $R = ((1,3),2)$ or $R'= ((2,3),1)$ for $T = ((1,2),3)$.
+    Since it is $d(T,R) = d(T,R') = 1 = \Delta(\rnni_3)$, the lemma is true for trees on $n=3$ taxa.
+
+    Induction step: $n \to n+1$\\
+    Induction hypothesis: For a caterpillar tree $T$ on $n$ taxa $\mdtree$ computes a tree $R$ with distance $d(T,R) = \frac{(n-1)(n-2)}{2}$.\\
+    Let us now  assume that $T$ is a tree on $n+1$ taxa.
+    In the list $L$ produced in $\mdtree(T)$ the taxon $n+1$ is the last one, $L[n] = n+1$.
+    Hence $n+1$ is added to $R$ in the last step and therefore, it is in the lowest cherry of $R$ and has parent with rank $1$.
+    Because the parent of $n+1$ has rank $n$ in $T$, we can conclude from Lemma~\ref{lemma:distance_delete_taxon} that for $T$ and $R$ restricted to taxa $1,\ldots,n$ it holds: $d(T|_n,R|_n) \leq d(T,R) - (n-1)$.
+    It is easy to see that $R|_n$ can be received from $T|_n$ by applying $\mdtree$.
+    So we can follow from the induction hypothesis and Lemma~\ref{lemma:distance_delete_taxon} that it is $d(T,R) \geq d(T|_n,R|_n) + (n-1) = \frac{(n-1)(n-2)}{2} = \frac{n(n-1)}{2}$, which concludes the proof of the lemma.
+\end{proof}
+
+In $\mdtree$ it is not uniquely determined, where a new taxon is added to the already existing tree.
+Instead, the attachment branch can be randomly chosen with the only restriction being that the rank of the new node needs to be below the ranks of already existing internal nodes.
+Therefore the output tree $R$ of $\mdtree$ is not unique.
+Every tree topology on $n$ taxa can be resulting from $\mdtree$.
+And because permuting labels of trees does not change the distance between them, $\mdtree$ shows that there is for every tree $R$ on $n$ taxa a caterpillar tree $T$ with distance $d(T,R) = \Delta(\rnni) = \frac{(n-1)(n-2)}{2}$.
+This gives us Corollary~\ref{corollary:radius}.
+
+\begin{corollary}
+    The radius of the $\rnni$ graph is $rad(\rnni) = \Delta(\rnni) = \frac{(n-1)(n-2)}{2}$.
+    \label{corollary:radius}
 \end{corollary}
@@
 
-Recall that the Section~\ref{section:diameter} revealed that the maximum distance computed by $\findpath$ for an arbitrary pair of trees equals the diameter of $\rnni$.
+Recall that Section~\ref{section:diameter} revealed that the maximum distance computed by $\findpath$ for an arbitrary pair of trees equals the diameter of $\rnni$.
 Thus, if the distance between two trees equals the diameter, $\findpath$ computes the correct distance.
-Following this observation, we claim following conjecture:
+Following this observation, we claim Conjecture~\ref{conjecture:findpath_maxdist_correct}.
+Notice that one direction of the conjecture is already known to be true, it only remains open whether the distance of two trees is $\Delta(\rnni)$ indeed, if $\findpath$ computes a path of this length.
 
@@
 
-Besides the diameter, the \emph{radius} of a graph $G$, which is defined by $rad(G) =  \min\limits_u \max\limits_v d(u,v)$ is a further interesting property of a graph.
-We will use Algorithm~\ref{alg:max_dist_tree} to show that in $\rnni$ the radius equals the diameter.
-
-%TODO
-%Explain the following algorithm, not only Pseudo-Code!
+In order to find the radius of the $\rnni$ graph, the algorithm $\mdtree$ has been introduced in Section~\ref{section:diameter}.
+It computes a tree $R$ with distance $\Delta(\rnni)$ from a given caterpillar tree, as it has been proven in Lemma~\ref{lemma:max_dist_caterpillar}.
+Algorithm~\ref{alg:max_dist_tree} provides the pseudocode for this algorithm.
+It is important to note that $\mdtree$ can not only be applied onto caterpillar trees, as it is done in section~\ref{section:diameter}, but to any arbitrary tree.
+However, it does not necessarily produce a tree with maximum distance from a given tree.
 
 \begin{algorithm}[H]
-\caption{MAX\_DISTANCE\_TREE($T$)}
+\caption{$\mdtree(T)$}
 \label{alg:max_dist_tree}
@@
 \end{algorithm}
-
-\begin{lemma}
-    Let $T$ be a caterpillar tree.
-    A tree $R$ that is computed by applying Algorithm~\ref{alg:max_dist_tree} on $R$ has distance $d(T,R) = \Delta(\rnni)$ to $T$.
-    \label{lemma:max_dist_caterpillar}
-\end{lemma}
-
-\begin{proof}
-    Recall that with Corollary~\ref{corollary:diameter} we know that $\Delta(\rnni) = \frac{(n-1)(n-2)}{2}$.
-    Within this proof we assume, without loss of generality, that it is $T = (( \dots (1,2)3)\dots)n)$
-    We prove the lemma by induction on the number of taxa $n$.
-
-    Induction basis: $n = 3$\\
-    Algorithm~\ref{alg:max_dist_tree} computes either $R = ((1,3),2)$ or $R'= ((2,3),1)$ when applied to $T = ((1,2),3)$.
-    Since it is $d(T,R) = d(T,R') = 1 = \Delta(\rnni_3)$, the lemma is true for trees on $n=3$ taxa.
-
-    Induction step: $n \to n+1$\\
-    Induction hypothesis: For a caterpillar tree $T$ on $n$ taxa Algorithm~\ref{alg:max_dist_tree} computes a tree $R$ with distance $d(T,R) = \frac{(n-1)(n-2)}{2}$.\\
-    Let us now  assume that $T$ is a tree on $n+1$ taxa.
-    In the list $L$ produced in Algorithm~\ref{alg:max_dist_tree}, when applied to $T$, taxon $n+1$ is the last one.
-    Hence $n+1$ is added to $R$ in the last step and therefore, it is in the lowest cherry of $R$ and has parent with rank $1$.
-    Because the parent of $n+1$ has rank $n$ in $T$, we can conclude from Lemma~\ref{lemma:distance_delete_taxon} that for $T$ and $R$ restricted to taxa $1,\ldots,n$ it holds: $d(T|_n,R|_n) \leq d(T,R) - (n-1)$.
-    It is easy to see that $R|_n$ can be received from $T|_n$ by applying Algorithm~\ref{alg:max_dist_tree}.
-    So we can follow from the induction hypothesis and Lemma~\ref{lemma:distance_delete_taxon} that it is $d(T,R) \geq d(T|_n,R|_n) + (n-1) = \frac{(n-1)(n-2)}{2} = \frac{n(n-1)}{2}$, which concludes the proof of the lemma.
-\end{proof}
-
-In Algorithm~\ref{alg:max_dist_tree} it is not uniquely determined, where a new taxon is added to the already existing tree.
-Instead, the attachment branch can be randomly chosen with the only restriction being that the rank of the new node needs to be below the ranks of already existing internal nodes.
-This is the reason why one can receive every tree topology on $n$ taxa when applying Algorithm~\ref{alg:max_dist_tree} onto a caterpillar tree.
-And because permuting the labels of the trees does not change the distance, there is for every tree $R$ on $n$ taxa a caterpillar tree $T$ with distance $d(T,R) = \Delta(\rnni) = \frac{(n-1)(n-2)}{2}$.
-This gives us Corollary~\ref{corollary:radius}.
-
-\begin{corollary}
-    The radius of the $\rnni$ graph is $rad(\rnni) = \Delta(\rnni) = \frac{(n-1)(n-2)}{2}$.
-    \label{corollary:radius}
-\end{corollary}
 

### 2019-08-23 paper.tex
edits + additions in computation section

@@
 \begin{lemma}
-	There are two caterpillar trees in $\rnniu$ with distance $\frac{(n-1)(n-2))}{2}$.
+	There are two caterpillar trees in $\rnniu$ with distance $\frac{(n-1)(n-2)}{2}$.
 	\label{lemma:caterpillar_diameter}
@@
 This algorithm will be called $\mdtree$ (Maximum Distance Tree).
-We will prove that a tree $R = \mdtree(T)$ has distance $d(T,R) = \Delta(\rnni)$ to $T$ if $T$ is a caterpillar tree, after introducing the algorithm.
+After introducing this algorithm, we will prove that a tree $R = \mdtree(T)$ has distance $d(T,R) = \Delta(\rnni)$ to $T$ if $T$ is a caterpillar tree.
 
@@
 In \autocite{Gavryushkin2018-ol} an efficient algorithm for computing the $\rnni$ graph has been introduced.
-We implemented this algorithm and could produce the graph for trees with up to $8$ taxa. \todo{Shall we say more about the algorithm here?}
-This enables us to compute exact distances between trees of this size.
-In particular, this enables us to compare exact distances with distances we receive from algorithms introduced in previous sections and to verify the Cluster Theorem, which was raised in Section~\ref{section:split_theorem}, for small trees.
+We implemented this algorithm, which produces the graph for trees with up to $8$ taxa in reasonable time.
+\todo{Shall we say more about the algorithm here?
++ running time (in hours/minutes) on which computer?}
+This enables us to compute exact distances between trees with up to $8$ taxa.
+In particular, comparing exact distances with approximations as introduced in previous sections and verifying the Cluster Theorem of Section~\ref{section:split_theorem} can now be for small trees.
 All algorithms have been implemented with Python and are available online. \todo{link?}
+In the following, further details about the implementation of algorithms $\findpath$ and $\mdtree$ as well as a description of how to prove the Cluster Theorem will be given.
 
@@
 In Section~\ref{section:geometry} we introduced the algorithm $\findpath$ for computing paths between arbitrary trees in $\rnni$.
-Therefore, we can use it as approximation for distances in $\rnni$.
-In this section we are providing pseudo-code for $\findpath$ (Algorithm~\ref{alg:find_path}) to provide further insights.
-We implemented and tested this algorithm for trees on up to $k$ taxa, for which it gives the exact $\rnni$ distance. \todo{$k = ?$}
+Therefore, it approximates distances in $\rnni$.
+In this section we are providing pseudocode for $\findpath$ (Algorithm~\ref{alg:find_path}) to provide further insights.
+We implemented and tested this algorithm for trees on up to $k$ taxa, for which it gives the exact $\rnni$ distance. \todo{$k = ?$, How exactly did we test it?}
 
@@
 
-In order to find the radius of the $\rnni$ graph, the algorithm $\mdtree$ has been introduced in Section~\ref{section:diameter}.
+\subsection{Algorithm $\mdtree$}
+\label{section:computation_mdtree}
+
+For finding the radius of the $\rnni$ graph, algorithm $\mdtree$ has been introduced in Section~\ref{section:diameter}.
 It computes a tree $R$ with distance $\Delta(\rnni)$ from a given caterpillar tree, as it has been proven in Lemma~\ref{lemma:max_dist_caterpillar}.
 Algorithm~\ref{alg:max_dist_tree} provides the pseudocode for this algorithm.
-It is important to note that $\mdtree$ can not only be applied onto caterpillar trees, as it is done in section~\ref{section:diameter}, but to any arbitrary tree.
+It is important to note that $\mdtree$ can not only be applied onto caterpillar trees, as it is done in Section~\ref{section:diameter}, but to any arbitrary tree.
 However, it does not necessarily produce a tree with maximum distance from a given tree.
+
+%TODO
+% Shall we give an example where the distance is not max or simulate trees and compute histogram of distances approximated by $\findpath$?
 
@@
 
-\subsection{Connecting Computations and Proofs}
-\todo{How shall we structure this?}
+\subsection{Cluster Theorem}
+\label{section:computation_cluster_theorem}
 
-In Section~\ref{section:caterpillar_convex} we already indicated that computing distances between trees in $\rnni$ by using the algorithms introduced in previous sections has been important for understanding shortest paths.
-Moreover, these algorithm, in particular $\findpath$ and $\csort$, also became a substantial part of some of the in Section~\ref{section:geometry}.
+%TODO
+%How much detail do we need? Shall we have this as separate section or is it sufficient to mention that we checked the cluster theorem for up to $k$ taxa?
 
-We can also substantiate the Cluster Theorem, which remains to be a conjecture, by proving it computationally for up to $k$ taxa. \todo{k=?, do we provide implementation?}
-Together with the analytical results of the previous section, this strongly supports our claim that the Cluster Theorem is true.
+For proving the Cluster Theorem as stated in Conjecture~\ref{conjecture:cluster_theorem} for small trees, we implemented the algorithm for computing the $\rnni$ graph as stated above.
+If the graph is computed, one can get distances by using algorithms such as Floyd-Warshall~\autocite{Floyd1962-ew} or Dijkstra~\autocite{Dijkstra1959-ph}.
+However, the large number of vertices in $\rnni$ is an obstacle when it comes to computing distances.
+For instance the Floyd-Warshall algorithm needs to save a matrix containing all pairwise distances of vertices in the graph, which needs space quadratic in the number of vertices.
+Since there are $\frac{n!(n-1)!}{2^{n-1}}$ vertices in the $\rnni$ graph~\autocite{Gavryushkin2018-ol}, we already run out of space when considering $7$ taxa.
+Therefore, we cannot simply use this algorithm to compute all distances in $\rnni$.
+Instead of this we compute distances between certain pairs of trees and use the symmetry of $\rnni$ to draw conclusions about the remaining distances.
 
-%Max Distance Trees Algorithm (+ connection to MRCA algorithm)
+In the following we describe how the Cluster Theorem has computationally been proven for small trees.
+For this aim only a restricted subset of trees in $\rnni$ needs to be considered, that is trees that share a cluster.
+Furthermore, it is sufficient to consider clusters of the shape $\{1, \ldots, m\}$ for $2 \leq m \leq n-1$.
+This is due to the fact that the distance $d(T_1,R_2)$ and $d(T_2,R_2)$ between two pairs of trees $(T_1,R_1)$ and $(T_2,R_2)$ are equal if one can receive $T_2$ from $T_1$ by permuting the taxa the same way as it is necessary for receiving $R_2$ from $R_1$.
+Therefore, we compute for all $m = 2, \ldots, n-1$ exact $\rnni$ distances between all trees containing the cluster $\{1, \ldots, m\}$, using Dijkstra.
+Additionally, the subgraph of $\rnni$ only consisting of trees that have $\{1, \ldots, m\}$ as cluster is computed.
+All pairwise distances between the trees in the subgraph are computed, by using Floyd-Warshall, to compare them with the distances in $\rnni$.
+With this approach it is possible to show that the Cluster Theorem holds for up to $6$ taxa. \todo{Can we improve this?}
 
@@
 \begin{proof}
+    %TODO
     % Insert the proof here
@@
 
-According to Algorithm~\ref{alg:max_dist_tree}, the number of trees with maximum distance to a given tree $T$ depends on the number of cherries of $T$ and the ranks of their internal nodes.
-For caterpillar trees it is $(n-1)!$.
+According to Algorithm~\ref{alg:max_dist_tree}, the number of trees with maximum distance to a given caterpillar tree $T$ is $(n-1)!$.
 The number of caterpillar trees with distance $\Delta(\rnni)$ from a given caterpillar tree is $2^{n-2}$.
-
-%If this is true, we can follow that radius = diameter
 

### 2019-08-29 paper.tex
add idea: symmetry of FINDPATH

@@
 We implemented and tested this algorithm for trees on up to $k$ taxa, for which it gives the exact $\rnni$ distance. \todo{$k = ?$, How exactly did we test it?}
+The running time of $\findpath$ is quadratic in the number of taxa.
 
@@
 	\STATE $R$ only contains cherry taxa $a, b$ of $T$ and has internal node with rank $n-1$
-    \STATE In list $L$ all taxa $\{1,\ldots,n\}\setminus\{a,b\}$ are sorted such that the ranks of their parents in $T$ do not decrease
+    \STATE In list $L$ all taxa $\{1,\ldots,n\}\setminus\{a,b\}$ are ordered such that the ranks of their parents in $T$ do not decrease
 	\FOR {taxon $a$ in $L$}
@@
 \section{Ideas}
-\todo{Which ideas do we want to keep? Shall this become an 'open problems/future work' section?}
+\todo{This section is just a rough draft. Are there ideas we want to keep? Shall this become an 'open problems/future work' section?}
 
-Following are some ideas that did not lead to major results so far:
+The paths computed by $\findpath$ depend on the order of input trees $T$ and $R$:
+$\findpath(T,R)$ does not necessarily produce the reverse path of $\findpath(R,T)$.
+An example can be found in Figure~\ref{fig:findpath_not_symmetric}.
+However, simulations suggest that the lengths of the two paths computed by $\findpath$ for one pair of trees are equal.
+
+\begin{conjecture}
+    Let $T$ and $R$ be ranked trees.
+    The lengths of paths $\findpath(T,R)$ and $\findpath(R,T)$ are equal.
+\end{conjecture}
+
+\begin{figure}[H]
+	\centering
+	\includegraphics[width=0.7\textwidth]{findpath_not_symmetric}
+	\caption{Two paths computed by $\findpath$. At the top $\findpath(T,R)$, at the bottom $\findpath(R,T)$}
+	\label{fig:findpath_not_symmetric}
+\end{figure}
+
 % idea for description of max distance caterpillar trees:
-
 \begin{lemma}

### 2019-10-04 paper.tex
intro: headway

--- 2019-08-23 paper.tex
@@
 
-\title{The Ranked Nearest Neighbour Interchange space}
+\title{Geometry of Ranked Nearest Neighbour Interchange Space of Phylogenetic Trees}
 \date{\today}
@@
 \address{\textsuperscript{2}Institute for Mathematics and Inofrmatics, University of Greifswald, Germany}
+\author{David Bryant\textsuperscript{3}}
+\email{david.bryant@otago.ac.nz}
+\address{\textsuperscript{1}Department of Mathematics and Statistics, University of Otago, New Zealand}
 \author{Alex Gavryushkin\textsuperscript{1, \Letter}}
-\email{alex@biods.org}
+\email{\textsuperscript{\Letter}alex@biods.org}
 
@@
 
-% All papers have this section so no need for this heading -- let's wait until a journal pushes us to put it back :)
-%\section{Introduction}
+The nearest neighbour interchange ($\nni$) graph, defined on the set of phylogenetic trees with adjacency relation given by the interchange operation of two sister clades, has been known in mathematical biology literature for nearly 50 years \autocite{Robinson1971, Moore-Goodman-Barnabas1973}.
+Considered with the metric given by the length of the shortest path, this graph becomes a metric space and its geometry has been extensively studied \autocite{...}.
+In important property of the $\nni$ space is that the distance is NP-hard to compute \autocite{...}.
+As a result no algorithm exists to compute the distance in practical time \autocite{...Whidden may be}.
+[say a bit about tree search and sampling algorithms -- they are hard too]
 
-%Introduce RNNI as ranked nearest neighbour interchange space, only ultrametric trees considered (extension could be mentioned in the end)
+[a paragraph on RNNI, mention that only ultrametric trees considered (extension could be mentioned in the end)]
 
-%We use the exact algorithm to calculate all the distances for the examples given in this paper
+[Figure of ranked tree]
 
-%Figure of ranked tree in introduction
+[summary of results, including the approaximate algorithm that works exactly on small spaces (Lena: is it already in the text?)]
 

### 2019-10-04 paper.tex
fleshed out the intro a bit more + first def headway

@@
 \email{david.bryant@otago.ac.nz}
-\address{\textsuperscript{1}Department of Mathematics and Statistics, University of Otago, New Zealand}
+\address{\textsuperscript{3}Department of Mathematics and Statistics, University of Otago, New Zealand}
 \author{Alex Gavryushkin\textsuperscript{1, \Letter}}
@@
 The nearest neighbour interchange ($\nni$) graph, defined on the set of phylogenetic trees with adjacency relation given by the interchange operation of two sister clades, has been known in mathematical biology literature for nearly 50 years \autocite{Robinson1971, Moore-Goodman-Barnabas1973}.
-Considered with the metric given by the length of the shortest path, this graph becomes a metric space and its geometry has been extensively studied \autocite{...}.
+Considered with the metric given by the length of the shortest path (graph-distance), this graph becomes a metric space and its geometry has been extensively studied \autocite{...}.
 In important property of the $\nni$ space is that the distance is NP-hard to compute \autocite{...}.
 As a result no algorithm exists to compute the distance in practical time \autocite{...Whidden may be}.
-[say a bit about tree search and sampling algorithms -- they are hard too]
+A consequence is that tree search and sampling algorithms pose a significant challenge even for moderately sized trees \autocite{...}.
 
-[a paragraph on RNNI, mention that only ultrametric trees considered (extension could be mentioned in the end)]
+More recent advances in computational phylogenetics introduced various classes of molecular clock models \autocite{...} and made possible computational inference of phylogenetic time-trees \autocite{...}.
+However, mathematical challenges that come with this seemingly inessential change in parametrisation (genomic distance VS time distance) of trees have only recently been brought to the attention \autocite{Gavryushkin2016-uu}.
+\textcite{Gavryushkin2018-ol} have proposed an extension of the $\nni$ graph to the class of discrete time-trees.
+The simplest such extension introduces the $\rnni$ graph on the set of ranked phylogenetic trees.
+Considered with the graph-distance $\rnni$ becomes a metric space and inherits the geometric and algorithmic challenges that the $\nni$ space has been traditionally facing.
+Surprisingly, most of them cannot be settled by directly translating results or applying techniques developed for $\nni$ \autocite{Gavryushkin2018-ol}.
+
+In this paper, we consider the $\rnni$ space on the ranked phylogenetic trees with all taxa being of rank $0$.
+An example of such a tree is depicted in Figure~\ref{fig:...}.
+In the terminology of \autocite{Gavryushkin2018-ol}, the space considered in this paper is $\rnniu$ (see the Section~\ref{sex:...} for accurate definitions).
 
@@
 
-[summary of results, including the approaximate algorithm that works exactly on small spaces (Lena: is it already in the text?)]
-
-Although we endeavour to make this paper self-sufficient, we refer the reader to standard textbooks in phylogenetics for terminology \autocite{Felsenstein2004-of, Semple2003-nj, Steel2016-ye}.
-Unless is explicit, we follow these textbooks in our terminology.
+In line with the research programme proposed in \autocite{Gavryushkin2018-ol}, we investigate the geometry of the $\rnni$ space and also consider several computational versions of the problems.
+Specifically, in this paper we establish the exact radius and diameter of the space, [...
+-- Lena, please finish this including the approaximate algorithm that works exactly on small spaces (Lena: is it already in the text?)]
 
@@
 
-A \emph{rooted binary phylogenetic tree} on a set $X$ is a pair $(T, \Phi)$ where $T$ is a rooted binary tree with leaf set $\mathcal{L}(T)$ and $\Phi:X \to \mathcal{L}(T)$ a bijective mapping.
-\todo{do we need to define leaves and internal nodes, child, parent, subtree, cherry(defined below)?
-AG: Only if we use those terms.
-Can we use something like a glossary to automatically know what terminology we need to introduce?}
-From now on, we assume $X = \{1,\ldots,n\}$ for a fixed number $n \in \mathbb N$ of leaves.
+A \emph{ranked phylogenetic tree} is a pair consisting of a rooted binary phylogenetic tree $T$ on the set $X = \{1, \ldots, n\}$ of taxa for $n \in \mathbb N$, and a rank function that maps $X$ to $0$, all internal nodes to $\{1, \ldots, n-1\}$, and respects the partial order given by the tree.
 
-In this paper we study \emph{ranked trees} that are determined by a pair consisting of a rooted binary phylogenetic tree $(T, \Phi)$ on $X = \{1, \ldots, n\}$ for $n \in \mathbb N$, and a rank function.
 In the literature ranked trees are often referred to as labelled history in the context of coalescent trees.

### 2019-10-04 paper.tex
pass through the introducion -- @lenacoll you can now work on the Intro

@@
 An example of such a tree is depicted in Figure~\ref{fig:...}.
-In the terminology of \autocite{Gavryushkin2018-ol}, the space considered in this paper is $\rnniu$ (see the Section~\ref{sex:...} for accurate definitions).
+In the terminology of \autocite{Gavryushkin2018-ol}, the space considered in this paper is the space of ranked ultrametric phylogenetic trees $\rnniu$.
+We postpone accurate definitions until later in this section.
 
@@
 
-In line with the research programme proposed in \autocite{Gavryushkin2018-ol}, we investigate the geometry of the $\rnni$ space and also consider several computational versions of the problems.
+In line with the research programme proposed in \autocite{Gavryushkin2018-ol}, we investigate the geometry and algorithmic complexity of the $\rnni$ space.
 Specifically, in this paper we establish the exact radius and diameter of the space, [...
@@
 
-\section{Definitions}
+In the rest of introduction we formally introduce non-standard definitions necessary for this paper.
 
-A \emph{ranked phylogenetic tree} is a pair consisting of a rooted binary phylogenetic tree $T$ on the set $X = \{1, \ldots, n\}$ of taxa for $n \in \mathbb N$, and a rank function that maps $X$ to $0$, all internal nodes to $\{1, \ldots, n-1\}$, and respects the partial order given by the tree.
+A \emph{ranked phylogenetic tree} is a pair consisting of a rooted binary phylogenetic tree $T$ on the set $X = \{1, \ldots, n\}$ of taxa for $n \in \mathbb N$, and a (total) rank function $\rank$ that maps all elements of $X$ to $0$, all internal nodes of $T$ onto the set $\{1, \ldots, n-1\}$, and respects the partial order given by the tree.
+The latter means that if $x$ and $y$ are two internal nodes of $T$ such that there exists a path from a taxon $z \in X$ to $y$ which passes through $x$ then $\rank(x) < \rank(y)$.
+Ranked trees $(T_1, r_1)$ and $(T_2, r_2)$ are different if trees $T_1$ and $T_2$ are different or $T_1 = T_2$ and $\rank_1 \neq \rank_2$.
+Since all trees in this paper are ranked trees, we will abuse the notation and drop the ranking from the notation.
+We will also refer to a ranked tree simply by tree.
+For a tree $T$, we will use $\rank_T$ to refer to its ranking.
 
-In the literature ranked trees are often referred to as labelled history in the context of coalescent trees.
-The following notion differs from the definition of ranks in that context, since we do not use negative integers as ranks.
-A \emph{rank function} on $T$ maps the vertex set of $T$ onto the set $\{0,\ldots,n-1\}$ such that all leaves are assigned rank $0$ and each internal node has rank greater than its children.
-The rank of a node $v$ in a ranked tree will be denoted by $\rank(v)$.
-For simplicity of notation we use the term \emph{tree} instead instead of ranked tree.
-
-We will identify leaves with their labels and also call them \emph{taxa} (singular: \emph{taxon}) following the systematic biology tradition.
-All nodes that are no leafs are called \emph{internal nodes} and an edge $(u,v)$ is an \emph{internal edge}, if both $u$ and $v$ are internal nodes.
-Our definition of ranked tree leads to a natural characterisation of edge lengths within ranked trees:
-integer $l$ denotes the \emph{length} of an edge $e=(u,v)$ if $|\rank(v) - \rank(u)| = l$.
-Two trees are called \emph{isomorphic} if the underlying rooted phylogenetic trees are isomorphic and the isomorphism preserves ranks of the nodes.
+Our definition of a ranked tree implies a natural notion of the edge length.
+Indeed, a tree edge $(u,v)$ has length $l$ if $|\rank(v) - \rank(u)| = l$.
 
 Now we are ready to define the tree space which is the subject of study in this paper, the $\rnni$ graph.
-Note that we will define the $\rnni$ graph on ultrametric trees where all leaves have the same rank $0$, and therefore refers to the $\rnniu$ graph as introduced in \autocite{Gavryushkin2018-ol}.
-The vertex set of the \emph{$\rnni$ graph} is given by the set of all ranked trees on $n$ taxa.
-We introduce two types of operations (\emph{$\rnni$ moves}) connecting trees $T$ and $R$ in this graph by an edge:
-A swap of the ranks of two internal nodes ($x$ and $y$) of subsequent ranks ($|\rank(x) - \rank(y)| = 1$) is called a \emph{rank move}.
-We say that trees $T$ and $R$ are \emph{connected} by an \emph{$\nni$ move} move if there exist internal edges $e$ in $T$ and $f$ in $R$ both of length one such that the graphs obtained by shrinking $e$ and $f$ to internal nodes are isomorphic.
-The above definition follows \autocite{Gavryushkin2018-ol} and is equivalent to the common definition of $\nni$ operations as given in \autocite{Steel2016-ye}.
-Trees that are connected by either a rank move or an $\nni$ move are called \emph{$\rnni$ neighbours} or \emph{adjacent in} $\rnni$.
-Note that thus defined $\rnni$ graph is a metric space with the standard shortest-path distance.
 
+The vertex set of the $\rnni$ graph is the set of all ranked trees on $n$ taxa.
+We introduce two types of operation ($\rnni$ \emph{moves}) on trees (see Figure~\ref{fig:RNNI}) and say that two trees are adjacent in the $\rnni$ graph is they are connected by an operation of either type.
+The first type of operation is called a \emph{rank move} and defined by swapping the ranks of two internal nodes that are not adjacent in the tree.
+Formally, if $x$ and $y$ are such nodes of a tree $T$ that $|\rank_T(x) - \rank_T(y)| = 1$ and $(x, y)$ is not an edge in $T$ then the tree $R$ obtained from $T$ by only changing $\rank_R(x) = \rank_T(y)$ and $\rank_R(y) = \rank_T(x)$ is said to be obtained by a rank move.
+The second type of operation is called an $\nni$ \emph{move} and defined in the usual way, that is, two trees $T$ and $R$ are said to be connected by an $\nni$ move if there exist internal edges $e$ in $T$ and $f$ in $R$ both of length one such that the trees obtained by shrinking $e$ and $f$ to internal nodes coincide.
+
+\todo{Add a bit more space in all figures at the bottom to have at least a standard interval above he caption.}
+\todo{In this figure -- rm arrows (think how to place the RNNI edges), possibly avoid using colour, and highlight where the rank move is made (hollow nodes?)}
+\todo{Two of these trees are caterpillar trees.
+Do you think it would be helpful to draw them as caterpillars to prepare readers for the rest of the paper?}
 \begin{figure}[H]
@@
 	\includegraphics[width=\textwidth]{RNNI}
-	\caption{Four trees connected by edges in the $\rnni$ graph. A rank move of $\mrca(\{1,2\})$ and $\mrca(\{3,4\})$ and all $\nni$ moves on the edge highlighted in red are illustrated}
+	\caption{Two types of operation that define edges of the $\rnni$ graph.}
 	\label{fig:RNNI}
@@
 
-\todo{label of taxa in figure shouldn't be numbers}
-
-The following definitions are less standard and required for the results obtained in this paper.
-A set $\mathbb S$ of trees is called \emph{convex} if for all pairs of trees $T,R \in \mathbb S$, there exists a shortest path from $T$ to $R$ such that every tree on $p$ is element of $\mathbb S$.
+\todo{label of taxa in figure shouldn't be numbers\\
+AG: Well -- if you decide to change this, go through all the definitions VERY carefully to make sure we update everywhere.}
 
@@
 
-This paper is an advance towards the general goal of understanding the geometry of the $\rnni$ space.
-In this section we are investigating the geometry by analysing the structure of shortest paths in $\rnni$.
-We will start by presenting an algorithm for computing paths in $\rnni$ ($\findpath$) that turns out to be an essential tool for our studies.
-Using this algorithm, we are able to prove that there is always a shortest path between two caterpillar trees that only consists of caterpillar trees (Theorem~\ref{thm:caterpillar_convex}) in Section~\ref{section:caterpillar_convex}.
-In other words, the set of caterpillar trees is convex in $\rnni$.
+In this section we analyse shortest paths in $\rnni$.
+We will start by presenting an algorithm for computing paths in $\rnni$ ($\findpath$) that turns out to be an essential tool for understanding the geometry of $\rnni$.
+Specifically, this algorithm will enable us to prove that there is always a shortest path between two caterpillar trees that only consists of caterpillar trees (Theorem~\ref{thm:caterpillar_convex}) in Section~\ref{section:caterpillar_convex}.
+In other words, the set of caterpillar trees is \emph{convex} in $\rnni$.
+\todo{The rest of this paragraph should be carefully checked and be in the introduction.
+Generally, you might want to think a bit more about a better structure of this paper into chapters.}
 This distinguishes $\rnni$ from $\nni$ where this does not hold.
 One further result, presented in Section~\ref{section:diameter}, directly following from the proof of Theorem~\ref{thm:caterpillar_convex}, is the diameter of $\rnni$.
-Finally, we disprove the so-called Split Theorem, conjectured in \autocite{Gavryushkin2018-ol}, and formulate an alternative version of it, the Cluster Theorem (Section~\ref{section:split_theorem}).
+Finally, we disprove the so-called Strong Split Theorem, conjectured in \autocite{Gavryushkin2018-ol}.
+However, the Weak Split Theorem remains a conjecture.
+Furthermore, we propose an alternative (even weaker) version of the conjecture, the Cluster Theorem (Section~\ref{section:split_theorem}).
 
@@
 \begin{theorem}
-	The $\rnni$ space on $n$ taxa is isomorphic to the space of maximum chains of the partition lattice $\Pi_n$ where two maximum chains are connected by an edge if, and only if, they differ by exactly one partition.
+The $\rnni$ graph on $n$ taxa is isomorphic to the graph of maximal chains of the partition lattice $\Pi_n$ where two maximal chains are connected by an edge if and only if they differ by exactly one partition.
+The corresponding metric spaces are isometric.
 \end{theorem}

### 2019-10-04 paper.tex
pass through (with lots of comments) to the end of L1 proof

@@
 
-\title{Geometry of Ranked Nearest Neighbour Interchange Space of Phylogenetic Trees}
+\title[Ranked Nearest Neighbour Intarchange]{Geometry of Ranked Nearest Neighbour Interchange Space of Phylogenetic Trees}
 \date{\today}
@@
 -- Lena, please finish this including the approaximate algorithm that works exactly on small spaces (Lena: is it already in the text?)]
+The question of whether there exists a polynomial algorithms for computing the $\rnni$ distance remain an open problem.
 
@@
 In this section we analyse shortest paths in $\rnni$.
-We will start by presenting an algorithm for computing paths in $\rnni$ ($\findpath$) that turns out to be an essential tool for understanding the geometry of $\rnni$.
+We will start by presenting an efficient algorithm for computing (not necessarily shortest) paths in $\rnni$ ($\findpath$) that turns out to be an essential tool for understanding the geometry of $\rnni$.
 Specifically, this algorithm will enable us to prove that there is always a shortest path between two caterpillar trees that only consists of caterpillar trees (Theorem~\ref{thm:caterpillar_convex}) in Section~\ref{section:caterpillar_convex}.
@@
 
-We now present an efficient algorithm for computing paths between pairs of trees in $\rnni$.
-Throughout this paper we will denote this algorithm by $\findpath$.
-We will see that it becomes a central tool for our investigation of shortest paths.
-
-Before introducing the idea of this algorithm, we need further definitions.
+To introduce the $\findpath$ algorithm, we need the following definitions.
 Each node $v$ of a tree defines a \emph{cluster} $C$, which is the set of taxa descending from $v$.
 We then say that $v$ \emph{induces} $C$.
-\emph{Trivial clusters} are clusters induced by leaves and the root of a tree.
-Note that the list of all clusters of a tree, if sorted increasingly according to the rank of the corresponding nodes, uniquely defines that tree.
-For computing a path from trees $T$ to $R$ we consider clusters $C_1, \ldots, C_{n-1}$ of $R$ where $C_i$ is induced by the node with rank $i$ for $i = 1, \ldots, n-1$.
+\emph{Trivial clusters} are clusters induced by leaves (contain just one element each) and the root of the tree (contains all taxa).
+Note that the list of all
+\todo{I added `non-trivial' here because the clusters corresponding to leaves cannot be put on the list.
+Check that the indexing of $C_i$ is still correct.
+I think we need to be explicit whether we ever use trivial clusters and I suggest that we don't.}
+non-trivial
+clusters of a tree, sorted according to the rank of the corresponding node, defines the tree unambiguously.
+
+To compute a path from tree $T$ to $R$ we consider clusters $C_1, \ldots, C_{n-1}$ of $R$ (the destination tree) where $C_i$ is induced by the node with rank $i$ for $i = 1, \ldots, n-1$.
+\todo{In that case the number of clusters should be $n-2$.}
 Trivial clusters do not need to be changed on any path between two trees which is why they are not considered in the algorithm.
-This list $C_1, \ldots, C_{n-1}$ of clusters, which is ordered increasingly according to the ranks of their corresponding nodes, uniquely defines $R$.
-Starting at tree $\hat{T} = T$, $\findpath$ constructs a path where the clusters of $R$ are constructed from bottom to top, using following procedure:
-For each $C_i$, find the \emph{most recent common ancestor} $\mrca_{\hat{T}}(C_i)$ (also known as \emph{least common ancestor}) of $C_i$ in current tree $\hat{T}$, that is the internal node with lowest rank that is ancestral to all the taxa in $C_i$.
-Update $\hat{T}$ by decreasing the rank of $\mrca_{\hat{T}}(C_i)$ with an $\rnni$ move until it has the same rank $i$ as in $R$.
-\todo{Do we need to explain why there is always an $\rnni$ move that moves the $\mrca$ down?}
-Since we consider the clusters of $R$ such that the ranks of the corresponding internal nodes in $R$ increase, $\mrca_{\hat{T}}(C_i)$ is the parent of cluster $C_i$ in $\hat{T}$ after these $\rnni$ moves.
+First, set $\hat{T} = T$
+Then for each $i$, find the node that induces $C_i$ in $\hat T$ and denote this node by $\mrca_{\hat{T}}(C_i)$ (short for \emph{most recent common ancestor}).
+Update $\hat{T}$ by
+\todo{This presumes  that the rank is higher.
+Why?}
+decreasing
+the rank of $\mrca_{\hat{T}}(C_i)$ by performing $\rnni$ moves until it has the same rank $i$ as in $R$.
+\todo{Do we need to explain why there is always an $\rnni$ move that moves the $\mrca$ down?
+AG: We need to explain quite a bit here -- there  are many ways one can do those moves + comment above.}
+Since we consider the clusters of $R$ such that the ranks of the corresponding internal nodes in $R$ increase, $\mrca_{\hat{T}}(C_i)$ is
+\todo{What is the `parent of a cluster' -- please check that you follow definitions you introduce.}
+the parent of cluster
+$C_i$ in $\hat{T}$ after these $\rnni$ moves.
 Thus, repeating this procedure for all clusters of $R$ results in a path between any pair of tress $T, R$, indeed.
+\todo{This kind of structure is not ideal (see my comment on the paper structure above).
+If we talk about the algorithm here, all the things, including the pseudo-code should be here.
+Possible solution: have algorithmic stuff first and geometry second.
+If you do it like this, explain that we need the algorithms for the geometry section.}
 Further details, including pseudo-code for $\findpath$, can be found in Section~\ref{section:computation_findpath}.
 
-This algorithm gives an upper bound for the diameter of $\rnni$, which improves the one given in \autocite{Gavryushkin2018-ol}.
+This algorithm gives an upper bound for the diameter of $\rnni$, which
+\todo{Why is that?}
+improves
+the one given in \autocite{Gavryushkin2018-ol}.
+\todo{When}
 Later we will see that this bound is tight.
@@
 \begin{lemma}
-    $\Delta(\rnni) \leq \frac{(n-1)(n-2)}{2}$, where $\Delta(G) = \max\limits_{u,v \in V(G)}d(u,v)$ denotes the diameter of graph $G$.
+$\Delta(\rnni) \leq \frac{(n-1)(n-2)}{2}$, where $\Delta(\rnni)$ denotes the diameter of the $\rnni$ graph, that is, if $d(u, v)$ is the graph distance in $\rnni$ then $\Delta(\rnni) = \max \{d(u, v) \mid u, v \in \rnni\}$.
 \label{lemma:diameter_bound}
@@
 \begin{proof}
-    Let us consider the maximum length of a path computed by $\findpath$.
-    In worst case each of the $n-2$ clusters $C$ induced by internal nodes of $R$ with rank $i = 1, \dots, n-2$, requires $n-1-i$ $\rnni$ operations to move $\mrca(C)$ down in $\hat{T}$.
-	Hence, the maximum length of a path computed by that algorithm for any pair of trees is less or equal to $\sum\limits_{i = 1}^{n-2} i = \frac{(n-2)(n-1)}{2}$.
+Since the diameter is bounded from above the maximum length of a path computed by $\findpath$, it is enough to find this maximum.
+The worst-case execution of $\findpath$ would require each of the $n-2$ clusters $C$ induced by internal nodes of $R$ to be moved by $n - 1 - i$ $\rnni$ operations in $\hat{T}$.
+Hence, the maximum length of such a path is bounded by $\sum\limits_{i = 1}^{n-2} i = \frac{(n-2)(n-1)}{2}$.
 \end{proof}

### 2019-10-05 paper.tex
to the end of L2 proof -- @lenacoll please check this proof!

@@
 -- Lena, please finish this including the approaximate algorithm that works exactly on small spaces (Lena: is it already in the text?)]
-The question of whether there exists a polynomial algorithms for computing the $\rnni$ distance remain an open problem.
+The question of whether there exists a polynomial algorithm for computing the $\rnni$ distance remains an open problem.
 
@@
 The second type of operation is called an $\nni$ \emph{move} and defined in the usual way, that is, two trees $T$ and $R$ are said to be connected by an $\nni$ move if there exist internal edges $e$ in $T$ and $f$ in $R$ both of length one such that the trees obtained by shrinking $e$ and $f$ to internal nodes coincide.
+
+We use the notation $d(T, R)$ throughout the paper to denote the graph distance between trees in $\rnni$, that is, $d(T, R)$ is the length of a shortest $\rnni$ path between $T$ and $R$.
 
@@
 
+\todo{make sure trees are consistently denoted by upper case letters}
 \begin{lemma}
-$\Delta(\rnni) \leq \frac{(n-1)(n-2)}{2}$, where $\Delta(\rnni)$ denotes the diameter of the $\rnni$ graph, that is, if $d(u, v)$ is the graph distance in $\rnni$ then $\Delta(\rnni) = \max \{d(u, v) \mid u, v \in \rnni\}$.
+$\Delta(\rnni) \leq \frac{(n-1)(n-2)}{2}$, where $\Delta(\rnni)$ denotes the diameter of the $\rnni$ graph, that is, if $d(T, R)$ is the graph distance in $\rnni$ then $\Delta(\rnni) = \max \{d(T, R) \mid T, R \in \rnni\}$.
 \label{lemma:diameter_bound}
@@
 
-Lemma~\ref{lemma:distance_delete_taxon} is important for some of the following proofs using induction.
-
-In most of the following inductive proofs the length of a shortest paths $p$ on trees on $n+1$ taxa is compared with a path on $n$ taxa induced by $p$.
-The following lemma bounds the length of these induced paths utilising the rank differences of certain internal nodes between start and end tree of $p$.
-Henceforth the notion $\parent(x)$ refers to the internal node that is parent of taxon $x$.
+The following lemma relates distances between trees on $n$ and $n+1$ taxa and is an important tool for inductive arguments.
+We use the notion $\parent_T(x)$ to refer to the node adjacent to taxon $x$ in tree $T$.
+$T{\big|}_n$ denotes the restriction of tree $T$ to the set of taxa $\{1, \ldots, n\}$.
+In particular if $T$ is a tree on taxa $1, \ldots, n+1$ the tree $T{\big|}_n$ is obtained by deleting taxon $n+1$ and suppressing the thereby created node of degree two.
 
 \begin{lemma}
-    Let $T$ and $R$ be two trees on $n+1$ taxa.
-    Let $T{\big|}_n$ and $R{\big|}_n$ be these trees restricted to taxa $1, \ldots, n$, i.e. taxon $n+1$ is deleted from $T$ and $R$ and the thereby arising node of degree two is suppressed.
-    Then it is $d(T{\big|}_n, R{\big|}_n) \leq d(T,R) - \delta$ where $\delta:= |\rank_T(\parent(n+1)) - \rank_R(\parent(n+1))|$ is the rank difference of the parent of taxon $x$ between trees $T$ and $R$.
+Let $T$ and $R$ be two trees on taxa $\{1, \ldots, n+1\}$.
+Then $d(T{\big|}_n, R{\big|}_n) \leq d(T,R) - \delta$, where $\delta = |\rank_T(\parent_T(n+1)) - \rank_R(\parent_R(n+1))|$.
 \label{lemma:distance_delete_taxon}
@@
 \begin{proof}
+First observe that the rank of an internal node $v$ of a tree $T$ along an $\rnni$ path $p$ can only be changed by performing a rank move that involves $v$ or an $\nni$ move on an edge adjacent to $v$.
+Second observe that any $\rnni$ move can change the rank of an internal node by at most one.
+
 Let $p$ be a shortest path from tree $T$ to $R$ and $p{\big|}_n$ the path resulting from deleting taxon $n+1$ from all trees on $p$.
-    This new path $p{\big|}_n$ is then a path from $T$ to $R$.
-    Showing that at least $\delta$ $\rnni$ moves are deleted from $p$ to receive $p{\big|}_n$ results in $d(T{\big|}_n,R{\big|}_n) \leq |p{\big|}_n| \leq |p| - \delta = d(T,R) -\delta$ which proves the lemma.
-    By the following two observations one can easily see that this claim is true, indeed:
-    The rank of an internal node $v$ only changes within an $\rnni$ move if this move is a rank move of $v$ or an $\nni$ move on an edge incident to $v$.
-    Furthermore, any $\rnni$ move can change the rank of an internal node by at most one.
-    Applying this to $\parent(n+1)$ and knowing that the rank of this node differs by $\delta$ between $T$ and $R$ proves that $d(T{\big|}_n, R{\big|}_n) \leq d(T,R) - \delta$.
+This new path $p{\big|}_n$ is then a path from $T{\big|}_n$ to $R{\big|}_n$.
+Recall that $\delta$ is the difference in ranks between the parents of taxon $n+1$ in $T$ and $R$.
+\todo{This statement has to be checked very carefully and one or two sentences added to explain the `applying'.}
+Applying the two observations above we conclude that $|p{\big|}_n| \leq |p| - \delta$.
+Since $d(T{\big|}_n,R{\big|}_n) \leq |p{\big|}_n|$ and $|p| = d(T,R)$, the desired inequality follows.
+\end{proof}
 
@@
     % \end{enumerate}
-\end{proof}
 

### 2019-10-05 paper.tex
deleted the alternative proof of L2 that was commented out
to end of Prop3

@@
 
-    % Alternative: Induction:
-    %
-    % Induction on $k = d(T,R)$:
-    %
-    % Basis($k = 1$):
-    % If rank of $\parent(n+1)$ changes between $T$ and $R$, it is $d(T{\big|}_n, R{\big|}_n) = 0$ and $|\rank_T(\parent(n+1)) - \rank_R(\parent(n+1))| = 1$ and therefore, $d(T{\big|}_n, R{\big|}_n) = 0 \leq 1 - 1 = d(T,R) - |\rank_T(\parent(n+1)) - \rank_R(\parent(n+1))|$.
-    % If on the other hand the rank of $\parent(n+1}$ does not change between $T$ and $R$, it is $d(T{\big|}_n, R{\big|}_n) = d(T,R) = 1$ and $|r_{T}(\parent(n+1}) - r_{R}(\parent(n+1})| = 1$ and therefore, $d(T{\big|}_n, R{\big|}_n) = 1 \leq 1 - 0 = d(T,R) - |r_{T}(\parent(n+1}) - r_{R}(\parent(n+1})|$.
-    %
-    % Step($k \to k+1$):
-    % Let $T$ and $R$ be trees on $n+1$ taxa with distance $d(T,R) = k$.
-    % Let $R'$ be the last tree on a shortest path from $T$ to $R$, i.e. $d(T,R') = k$.
-    % We can apply the induction hypothesis on $T$ and $R'$ and see that $d(T{\big|}_n, R'{\big|}_n) \leq d(T, R') - |r_{T}(\parent(n+1}) - r_{R'}(\parent(n+1})| := k - m$.
-    % Now we differ two cases:
-    % \begin{enumerate}
-    %     \item $\parent(n+1}$ changes its rank between $R$ and $R'$ or is affected by the $\nni$ move is performed between $R$ and $R'$.
-    %
-    %     Then the trees $R{\big|}_n$ and $R'{\big|}_n$ are equal. \todo{Is this obivous?}
-    %     Therefore, it is $d(T{\big|}_n,R{\big|}_n) = d(T{\big|}_n, R'{\big|}_n) \leq k - m$.
-    %     It is either $|r_{T}(\parent(n+1}) - r_{R}(\parent(n+1})| = m+1$, $|r_{T}(\parent(n+1}) - r_{R}(\parent(n+1})| = m$, or $|r_{T}(\parent(n+1}) - r_{R}(\parent(n+1})| = m-1$.
-    %     In either case, we have $d(T{\big|}_n,R{\big|}_n) = d(T{\big|}_n, R'{\big|}_n) \leq k - m = (k+1) - (m+1) \leq d(T, R) - |r_{T}(\parent(n+1}) - r_{R}(\parent(n+1})|$.
-    %
-    %     \item $\parent(n+1}$ has same rank in $R$ and $R'$.
-    %
-    %     In this case it is $d(T{\big|}_n,R{\big|}_n) \leq d(T{\big|}_n,R'{\big|}_n) + d(R'{\big|}_n, R{\big|}_n) \leq d(T{\big|}_n, R'{\big|}_n) + 1 \leq (k-m)+ 1$.
-    %     It also follows that $|r_{T}(\parent(n+1}) - r_{R'}(\parent(n+1})| = |r_{T}(\parent(n+1}) - r_{R}(\parent(n+1})|$ and therefore, we can conclude $d(T{\big|}_n,R{\big|}_n) \leq k-m+1 = k+1-m = d(T,R) -  |r_{T}(\parent(n+1}) - r_{R}(\parent(n+1})|$.
-    % \end{enumerate}
-
-%Following Proposition might not be needed(?)
-
-Related to Lemma~\ref{lemma:distance_delete_taxon} is Proposition~\ref{proposition:lower_bound_distance} which provides a lower bound for the distance of two trees.
-Using the same arguments as in the proof of Lemma~\ref{lemma:distance_delete_taxon}, it easily follows that the rank difference of parents of the same taxon but in two different trees gives such a bound.
+Using the same arguemtn as in the proof of Lemma~\ref{lemma:distance_delete_taxon} we can establish the following Proposition~\ref{proposition:lower_bound_distance} that provides a lower bound on the distance between trees.
 
 \begin{proposition}
-    Given trees $T$ and $R$ with taxon $x$ that fulfils $|\rank_T(\parent(x)) - \rank_R(\parent(x))| = k$, it follows $d(T,R) \geq k$.
+If trees $T$ and $R$ and taxon $x$ are such that $|\rank_T(\parent_T(x)) - \rank_R(\parent_R(x))| = \delta$ then $d(T,R) \geq \delta$.
 \label{proposition:lower_bound_distance}
@@
 Let us consider the first three elements $a,b,c$, appearing in this order in a sequence representing a caterpillar tree on a caterpillar path between trees $T$ and $R$.
+\todo{Check throughout that $\parent$ is never used without tree name in the index.}
 If the rank of $\parent(c)$ is higher than the ranks of $\parent(a)$ and $\parent(b)$ in $R$, there are no swaps needed for these taxa.

### 2019-10-18 paper.tex
first skim through intro

@@
 The nearest neighbour interchange ($\nni$) graph, defined on the set of phylogenetic trees with adjacency relation given by the interchange operation of two sister clades, has been known in mathematical biology literature for nearly 50 years \autocite{Robinson1971, Moore-Goodman-Barnabas1973}.
-Considered with the metric given by the length of the shortest path (graph-distance), this graph becomes a metric space and its geometry has been extensively studied \autocite{...}.
-In important property of the $\nni$ space is that the distance is NP-hard to compute \autocite{...}.
+Considered with the metric given by the length of a shortest path (graph-distance), this graph becomes a metric space.
+Its geometry has been extensively studied \autocite{...}.
+An important property of the $\nni$ graph is that computing distance is NP-hard \autocite{...}.
 As a result no algorithm exists to compute the distance in practical time \autocite{...Whidden may be}.
@@
 
-More recent advances in computational phylogenetics introduced various classes of molecular clock models \autocite{...} and made possible computational inference of phylogenetic time-trees \autocite{...}.
-However, mathematical challenges that come with this seemingly inessential change in parametrisation (genomic distance VS time distance) of trees have only recently been brought to the attention \autocite{Gavryushkin2016-uu}.
+More recent advances in computational phylogenetics introduce various classes of molecular clock models \autocite{...} and made possible computational inference of phylogenetic time-trees \autocite{...}.
+However, mathematical challenges that come with this seemingly inessential change in parametrisation (genomic distance VS time distance) of trees have only recently been brought to attention \autocite{Gavryushkin2016-uu}.
 \textcite{Gavryushkin2018-ol} have proposed an extension of the $\nni$ graph to the class of discrete time-trees.
@@
 
-In this paper, we consider the $\rnni$ space on the ranked phylogenetic trees with all taxa being of rank $0$.
+In this paper, we consider the $\rnni$ space on the ranked phylogenetic trees with all taxa being of equal rank.
 An example of such a tree is depicted in Figure~\ref{fig:...}.
@@
 
-In the rest of introduction we formally introduce non-standard definitions necessary for this paper.
+In the rest of this chapter we formally introduce non-standard definitions necessary for this paper.
 
@@
 
-Using the same arguemtn as in the proof of Lemma~\ref{lemma:distance_delete_taxon} we can establish the following Proposition~\ref{proposition:lower_bound_distance} that provides a lower bound on the distance between trees.
+Using the same argument as in the proof of Lemma~\ref{lemma:distance_delete_taxon} we can establish the following Proposition~\ref{proposition:lower_bound_distance} that provides a lower bound on the distance between trees.
 

### 2019-10-18 paper.tex
rough restructuring

@@
 
+\todo{LC: move the following definition? - it's actually quite important not only for the algorithm, so we might want to have it in the introduction?}
+Each node $v$ of a tree defines a \emph{cluster} $C$, which is the set of taxa descending from $v$.
+We then say that $v$ \emph{induces} $C$.
+\emph{Trivial clusters} are clusters induced by leaves (contain just one element each) and the root of the tree (contains all taxa).
+Note that the list of all
+\todo{I added `non-trivial' here because the clusters corresponding to leaves cannot be put on the list.
+Check that the indexing of $C_i$ is still correct.
+I think we need to be explicit whether we ever use trivial clusters and I suggest that we don't.}
+non-trivial
+clusters of a tree, sorted according to the rank of the corresponding node, defines the tree unambiguously.
+
 Now we are ready to define the tree space which is the subject of study in this paper, the $\rnni$ graph.
@@
 
-\section{Geometry of $\rnni$}
-\label{section:geometry}
+\todo{LC: Explain structure of paper}
 
-In this section we analyse shortest paths in $\rnni$.
-We will start by presenting an efficient algorithm for computing (not necessarily shortest) paths in $\rnni$ ($\findpath$) that turns out to be an essential tool for understanding the geometry of $\rnni$.
-Specifically, this algorithm will enable us to prove that there is always a shortest path between two caterpillar trees that only consists of caterpillar trees (Theorem~\ref{thm:caterpillar_convex}) in Section~\ref{section:caterpillar_convex}.
-In other words, the set of caterpillar trees is \emph{convex} in $\rnni$.
-\todo{The rest of this paragraph should be carefully checked and be in the introduction.
-Generally, you might want to think a bit more about a better structure of this paper into chapters.}
-This distinguishes $\rnni$ from $\nni$ where this does not hold.
-One further result, presented in Section~\ref{section:diameter}, directly following from the proof of Theorem~\ref{thm:caterpillar_convex}, is the diameter of $\rnni$.
-Finally, we disprove the so-called Strong Split Theorem, conjectured in \autocite{Gavryushkin2018-ol}.
-However, the Weak Split Theorem remains a conjecture.
-Furthermore, we propose an alternative (even weaker) version of the conjecture, the Cluster Theorem (Section~\ref{section:split_theorem}).
+\section{Algorithms}
+\label{section:algorithms}
 
-To introduce the $\findpath$ algorithm, we need the following definitions.
-Each node $v$ of a tree defines a \emph{cluster} $C$, which is the set of taxa descending from $v$.
-We then say that $v$ \emph{induces} $C$.
-\emph{Trivial clusters} are clusters induced by leaves (contain just one element each) and the root of the tree (contains all taxa).
-Note that the list of all
-\todo{I added `non-trivial' here because the clusters corresponding to leaves cannot be put on the list.
-Check that the indexing of $C_i$ is still correct.
-I think we need to be explicit whether we ever use trivial clusters and I suggest that we don't.}
-non-trivial
-clusters of a tree, sorted according to the rank of the corresponding node, defines the tree unambiguously.
+\todo{LC: make sure that we move all definitions in this paper to the right place}
+
+Within this section we will two introduce algorithms, namely $\findpath$ and $\csort$ for computing paths between two given trees in $\rnni$.
+These will be used later to prove some geometric properties of the $\rnni$ graph.
+Additionally we will present the algorithm $\mdtree$, which computes a caterpillar tree with maximum distance
+\todo{LC: can we use the term max dist here?}
+from a given tree.
+
+% In \autocite{Gavryushkin2018-ol} an efficient algorithm for computing the $\rnni$ graph has been introduced.
+% We implemented this algorithm, which produces the graph for trees with up to $8$ taxa in reasonable time.
+\todo{Shall we say more about the algorithm here?
++ running time (in hours/minutes) on which computer?}
+% This enables us to compute exact distances between trees with up to $8$ taxa.
+% In particular, comparing exact distances with approximations as introduced in previous sections and verifying the Cluster Theorem of Section~\ref{section:split_theorem} can now be for small trees.
+% All algorithms have been implemented with Python and are available online. \todo{link?}
+% In the following, further details about the implementation of algorithms $\findpath$ and $\mdtree$ as well as a description of how to prove the Cluster Theorem will be given.
+
+\subsection{Algorithm $\findpath$}
+\label{section:alg_findpath}
+
+\todo{LC: Introducing words}
 
@@
 If you do it like this, explain that we need the algorithms for the geometry section.}
-Further details, including pseudo-code for $\findpath$, can be found in Section~\ref{section:computation_findpath}.
+
+\todo{LC: Change the references to sections}
+In Section~\ref{section:geometry} we introduced the algorithm $\findpath$ for computing paths between arbitrary trees in $\rnni$.
+Therefore, it approximates distances in $\rnni$.
+In this section we are providing pseudocode for $\findpath$ (Algorithm~\ref{alg:find_path}) to provide further insights.
+We implemented and tested this algorithm for trees on up to $k$ taxa, for which it gives the exact $\rnni$ distance. \todo{$k = ?$, How exactly did we test it?}
+The running time of $\findpath$ is quadratic in the number of taxa.
+
+\begin{algorithm}[H]
+\caption{$\findpath$($T,R$)}
+\label{alg:find_path}
+\begin{algorithmic}[1]
+	\STATE $\hat{T} := T$
+	\STATE $p := (T)$
+	\FOR {$i = 1, \dots, n-2$}
+		\STATE Let $C$ be the cluster induced by node $v$ in $R$ with $\rank(v) = i$
+		\WHILE {$\rank(\mrca_{\hat{T}}(C))>i$}
+			\STATE Update $\hat{T}$: Decrease $\rank(\mrca_{\hat{T}}(C))$ by an $\rnni$ move \label{alg:line:move_set_down}
+			\STATE $p = p+\hat{T}$
+		\ENDWHILE
+	\ENDFOR
+	\RETURN $p$
+\end{algorithmic}
+\end{algorithm}
+
+Recall that Section~\ref{section:diameter} revealed that the maximum distance computed by $\findpath$ for an arbitrary pair of trees equals the diameter of $\rnni$.
+Thus, if the distance between two trees equals the diameter, $\findpath$ computes the correct distance.
+Following this observation, we claim Conjecture~\ref{conjecture:findpath_maxdist_correct}.
+Notice that one direction of the conjecture is already known to be true, it only remains open whether the distance of two trees is $\Delta(\rnni)$ indeed, if $\findpath$ computes a path of this length.
+
+\begin{conjecture}
+    Let $T$ and $R$ be two trees.
+    It is $d(T,R) = \frac{(n-1)(n-2)}{2}$ if, and only if, $\findpath(T,R)$ computes a path of this length between $T$ and $R$.
+    \label{conjecture:findpath_maxdist_correct}
+\end{conjecture}
+
+\subsection{Algorithm $\cpath$}
+\label{section:alg_cpath}
+
+\todo{LC: move definitions of caterpillar stuff here}
+\todo{LC: pseudo code}
+
+Before we start analysing shortest paths between caterpillar trees, we need to introduce further notations.
+A path $p$ between two caterpillar trees $T$ and $R$ that only consists of caterpillar trees is called \emph{caterpillar path}.
+If $p$ is shortest among all caterpillar paths between these trees we call it \emph{shortest caterpillar path}, and the length of $p$ will be denoted by $d_c(T,R)$ and called \emph{caterpillar distance}.
+
+We next introduce a unique representation of caterpillar trees to show how one can use the classical \emph{Bubble Sort} algorithm \autocite{Knuth1997-pi} to compute caterpillar paths.
+One can see caterpillar trees as sequences of taxa, ordered non-decreasing according to the ranks of their parents.
+$\nni$ moves between caterpillar trees, which are the only possible $\rnni$ moves between those trees, coincide with exchanges of two neighboured elements in the sequence.
+In these sequences the first two taxa are somehow equivalent since they build the unique cherry which means they share their parent with rank one.
+Therefore, we need to consider $\nni$ moves that change the cherry of a tree separately.
+These moves refer to exchanges of the first or the second element with the third element of the sequence, while exchanges of the first two elements do not change the corresponding tree.
+With this observation we can provide an algorithm for computing shortest caterpillar paths by a simple modification of the commonly known Bubble Sort algorithm.
+Throughout this paper this algorithm will be called $\csort$.
+
+Following the observations above, only swaps of the first three elements of a sequence with Bubble Sort need to be adapted for the computation of shortest caterpillar paths.
+Let us consider the first three elements $a,b,c$, appearing in this order in a sequence representing a caterpillar tree on a caterpillar path between trees $T$ and $R$.
+\todo{Check throughout that $\parent$ is never used without tree name in the index.}
+If the rank of $\parent(c)$ is higher than the ranks of $\parent(a)$ and $\parent(b)$ in $R$, there are no swaps needed for these taxa.
+If just one of the two taxa $a$ and $b$ is on the other side of $c$ in $R$, this taxon swaps position with $c$ in the sequence.
+And if both $a$ and $b$ are on the other side of $c$ in $R$, the taxon that has the parent with highest rank of all these three taxa in $R$ swaps with $c$ at first.
+This makes sure that there is no additional swap of $a$ and $b$ needed on the subsequent path to $R$.
+\todo{Do we need to explain Bubble Sort?}
+
+\subsection{Algorithm $\mdtree$}
+\label{section:alg_mdtree}
+
+For finding the radius of the $\rnni$ graph, algorithm $\mdtree$ has been introduced in Section~\ref{section:diameter}.
+It computes a tree $R$ with distance $\Delta(\rnni)$ from a given caterpillar tree, as it has been proven in Lemma~\ref{lemma:max_dist_caterpillar}.
+Algorithm~\ref{alg:max_dist_tree} provides the pseudocode for this algorithm.
+It is important to note that $\mdtree$ can not only be applied onto caterpillar trees, as it is done in Section~\ref{section:diameter}, but to any arbitrary tree.
+However, it does not necessarily produce a tree with maximum distance from a given tree.
+
+%TODO
+% Shall we give an example where the distance is not max or simulate trees and compute histogram of distances approximated by $\findpath$?
+
+\begin{algorithm}[H]
+\caption{$\mdtree(T)$}
+\label{alg:max_dist_tree}
+\begin{algorithmic}[1]
+	\STATE $R$ only contains cherry taxa $a, b$ of $T$ and has internal node with rank $n-1$
+    \STATE In list $L$ all taxa $\{1,\ldots,n\}\setminus\{a,b\}$ are ordered such that the ranks of their parents in $T$ do not decrease
+	\FOR {taxon $a$ in $L$}
+		\STATE Add an internal node as parent of $a$ on an edge in $R$ such that the new internal node has rank one less than the previously lowest internal node
+	\ENDFOR
+	\RETURN $R$
+\end{algorithmic}
+\end{algorithm}
+
+\subsection{Cluster Theorem}
+\label{section:computation_cluster_theorem}
+
+\section{Geometry of $\rnni$}
+\label{section:geometry}
+
+In this section we analyse shortest paths in $\rnni$.
+We will start by presenting an efficient algorithm for computing (not necessarily shortest) paths in $\rnni$ ($\findpath$) that turns out to be an essential tool for understanding the geometry of $\rnni$.
+Specifically, this algorithm will enable us to prove that there is always a shortest path between two caterpillar trees that only consists of caterpillar trees (Theorem~\ref{thm:caterpillar_convex}) in Section~\ref{section:caterpillar_convex}.
+In other words, the set of caterpillar trees is \emph{convex} in $\rnni$.
+\todo{The rest of this paragraph should be carefully checked and be in the introduction.
+Generally, you might want to think a bit more about a better structure of this paper into chapters.}
+This distinguishes $\rnni$ from $\nni$ where this does not hold.
+One further result, presented in Section~\ref{section:diameter}, directly following from the proof of Theorem~\ref{thm:caterpillar_convex}, is the diameter of $\rnni$.
+Finally, we disprove the so-called Strong Split Theorem, conjectured in \autocite{Gavryushkin2018-ol}.
+However, the Weak Split Theorem remains a conjecture.
+Furthermore, we propose an alternative (even weaker) version of the conjecture, the Cluster Theorem (Section~\ref{section:split_theorem}).
 
@@
 However, we don't know how good $\findpath$ approximates distances in $\rnni$. \todo{Can we say something about the approximation ratio?}
-More details can be found in Section~\ref{section:computation}.
 
@@
 This algorithm will be explained in detail later in this section, after demonstrating why finding shortest paths between caterpillar trees builds the basis for understanding the difference between $\nni$ and $\rnni$.
-Further details on our computational results can be found in Section~\ref{section:computation}.
-
-Before we start analysing shortest paths between caterpillar trees, we need to introduce further notations.
-A path $p$ between two caterpillar trees $T$ and $R$ that only consists of caterpillar trees is called \emph{caterpillar path}.
-If $p$ is shortest among all caterpillar paths between these trees we call it \emph{shortest caterpillar path}, and the length of $p$ will be denoted by $d_c(T,R)$ and called \emph{caterpillar distance}.
 
@@
 Note that this is not true for the $\nni$ space, which is clear from the example in Figure~\ref{fig:NNI_vs_RNNI}.
-
-We next introduce a unique representation of caterpillar trees to show how one can use the classical \emph{Bubble Sort} algorithm \autocite{Knuth1997-pi} to compute caterpillar paths.
-One can see caterpillar trees as sequences of taxa, ordered non-decreasing according to the ranks of their parents.
-$\nni$ moves between caterpillar trees, which are the only possible $\rnni$ moves between those trees, coincide with exchanges of two neighboured elements in the sequence.
-In these sequences the first two taxa are somehow equivalent since they build the unique cherry which means they share their parent with rank one.
-Therefore, we need to consider $\nni$ moves that change the cherry of a tree separately.
-These moves refer to exchanges of the first or the second element with the third element of the sequence, while exchanges of the first two elements do not change the corresponding tree.
-With this observation we can provide an algorithm for computing shortest caterpillar paths by a simple modification of the commonly known Bubble Sort algorithm.
-Throughout this paper this algorithm will be called $\csort$.
-
-Following the observations above, only swaps of the first three elements of a sequence with Bubble Sort need to be adapted for the computation of shortest caterpillar paths.
-Let us consider the first three elements $a,b,c$, appearing in this order in a sequence representing a caterpillar tree on a caterpillar path between trees $T$ and $R$.
-\todo{Check throughout that $\parent$ is never used without tree name in the index.}
-If the rank of $\parent(c)$ is higher than the ranks of $\parent(a)$ and $\parent(b)$ in $R$, there are no swaps needed for these taxa.
-If just one of the two taxa $a$ and $b$ is on the other side of $c$ in $R$, this taxon swaps position with $c$ in the sequence.
-And if both $a$ and $b$ are on the other side of $c$ in $R$, the taxon that has the parent with highest rank of all these three taxa in $R$ swaps with $c$ at first.
-This makes sure that there is no additional swap of $a$ and $b$ needed on the subsequent path to $R$.
-\todo{Do we need to explain Bubble Sort?}
 
@@
 However, the results provided in this section are only true for caterpillar trees.
-The pseudocode for $\mdtree$ can be found in Section~\ref{section:computation}.
+The pseudocode for $\mdtree$ can be found in Section~\ref{section:alg_mdtree}.
 \todo{figure? + do we need the pseudocode?}
@@
 
-\section{Computational results}
-\label{section:computation}
-
-In \autocite{Gavryushkin2018-ol} an efficient algorithm for computing the $\rnni$ graph has been introduced.
-We implemented this algorithm, which produces the graph for trees with up to $8$ taxa in reasonable time.
-\todo{Shall we say more about the algorithm here?
-+ running time (in hours/minutes) on which computer?}
-This enables us to compute exact distances between trees with up to $8$ taxa.
-In particular, comparing exact distances with approximations as introduced in previous sections and verifying the Cluster Theorem of Section~\ref{section:split_theorem} can now be for small trees.
-All algorithms have been implemented with Python and are available online. \todo{link?}
-In the following, further details about the implementation of algorithms $\findpath$ and $\mdtree$ as well as a description of how to prove the Cluster Theorem will be given.
-
-\subsection{Algorithm $\findpath$}
-\label{section:computation_findpath}
-
-In Section~\ref{section:geometry} we introduced the algorithm $\findpath$ for computing paths between arbitrary trees in $\rnni$.
-Therefore, it approximates distances in $\rnni$.
-In this section we are providing pseudocode for $\findpath$ (Algorithm~\ref{alg:find_path}) to provide further insights.
-We implemented and tested this algorithm for trees on up to $k$ taxa, for which it gives the exact $\rnni$ distance. \todo{$k = ?$, How exactly did we test it?}
-The running time of $\findpath$ is quadratic in the number of taxa.
-
-\begin{algorithm}[H]
-\caption{$\findpath$($T,R$)}
-\label{alg:find_path}
-\begin{algorithmic}[1]
-	\STATE $\hat{T} := T$
-	\STATE $p := (T)$
-	\FOR {$i = 1, \dots, n-2$}
-		\STATE Let $C$ be the cluster induced by node $v$ in $R$ with $\rank(v) = i$
-		\WHILE {$\rank(\mrca_{\hat{T}}(C))>i$}
-			\STATE Update $\hat{T}$: Decrease $\rank(\mrca_{\hat{T}}(C))$ by an $\rnni$ move \label{alg:line:move_set_down}
-			\STATE $p = p+\hat{T}$
-		\ENDWHILE
-	\ENDFOR
-	\RETURN $p$
-\end{algorithmic}
-\end{algorithm}
-
-Recall that Section~\ref{section:diameter} revealed that the maximum distance computed by $\findpath$ for an arbitrary pair of trees equals the diameter of $\rnni$.
-Thus, if the distance between two trees equals the diameter, $\findpath$ computes the correct distance.
-Following this observation, we claim Conjecture~\ref{conjecture:findpath_maxdist_correct}.
-Notice that one direction of the conjecture is already known to be true, it only remains open whether the distance of two trees is $\Delta(\rnni)$ indeed, if $\findpath$ computes a path of this length.
-
-\begin{conjecture}
-    Let $T$ and $R$ be two trees.
-    It is $d(T,R) = \frac{(n-1)(n-2)}{2}$ if, and only if, $\findpath(T,R)$ computes a path of this length between $T$ and $R$.
-    \label{conjecture:findpath_maxdist_correct}
-\end{conjecture}
-
-\subsection{Algorithm $\mdtree$}
-\label{section:computation_mdtree}
-
-For finding the radius of the $\rnni$ graph, algorithm $\mdtree$ has been introduced in Section~\ref{section:diameter}.
-It computes a tree $R$ with distance $\Delta(\rnni)$ from a given caterpillar tree, as it has been proven in Lemma~\ref{lemma:max_dist_caterpillar}.
-Algorithm~\ref{alg:max_dist_tree} provides the pseudocode for this algorithm.
-It is important to note that $\mdtree$ can not only be applied onto caterpillar trees, as it is done in Section~\ref{section:diameter}, but to any arbitrary tree.
-However, it does not necessarily produce a tree with maximum distance from a given tree.
-
-%TODO
-% Shall we give an example where the distance is not max or simulate trees and compute histogram of distances approximated by $\findpath$?
-
-\begin{algorithm}[H]
-\caption{$\mdtree(T)$}
-\label{alg:max_dist_tree}
-\begin{algorithmic}[1]
-	\STATE $R$ only contains cherry taxa $a, b$ of $T$ and has internal node with rank $n-1$
-    \STATE In list $L$ all taxa $\{1,\ldots,n\}\setminus\{a,b\}$ are ordered such that the ranks of their parents in $T$ do not decrease
-	\FOR {taxon $a$ in $L$}
-		\STATE Add an internal node as parent of $a$ on an edge in $R$ such that the new internal node has rank one less than the previously lowest internal node
-	\ENDFOR
-	\RETURN $R$
-\end{algorithmic}
-\end{algorithm}
-
-\subsection{Cluster Theorem}
-\label{section:computation_cluster_theorem}
+\todo{LC: connect previous with following (has been in computation section before)}
 

### 2019-10-19 paper.tex
add lemma:mrca_move - one rnni move to decrease rank(mrca(C))

@@
 The latter means that if $x$ and $y$ are two internal nodes of $T$ such that there exists a path from a taxon $z \in X$ to $y$ which passes through $x$ then $\rank(x) < \rank(y)$.
-Ranked trees $(T_1, r_1)$ and $(T_2, r_2)$ are different if trees $T_1$ and $T_2$ are different or $T_1 = T_2$ and $\rank_1 \neq \rank_2$.
+Ranked trees $(T_1, \rank_1)$ and $(T_2, \rank_2)$ are different if trees $T_1$ and $T_2$ are different or $T_1 = T_2$ and $\rank_1 \neq \rank_2$.
 Since all trees in this paper are ranked trees, we will abuse the notation and drop the ranking from the notation.
@@
 Indeed, a tree edge $(u,v)$ has length $l$ if $|\rank(v) - \rank(u)| = l$.
-
-\todo{LC: move the following definition? - it's actually quite important not only for the algorithm, so we might want to have it in the introduction?}
-Each node $v$ of a tree defines a \emph{cluster} $C$, which is the set of taxa descending from $v$.
-We then say that $v$ \emph{induces} $C$.
-\emph{Trivial clusters} are clusters induced by leaves (contain just one element each) and the root of the tree (contains all taxa).
-Note that the list of all
-\todo{I added `non-trivial' here because the clusters corresponding to leaves cannot be put on the list.
-Check that the indexing of $C_i$ is still correct.
-I think we need to be explicit whether we ever use trivial clusters and I suggest that we don't.}
-non-trivial
-clusters of a tree, sorted according to the rank of the corresponding node, defines the tree unambiguously.
 
@@
 
-Within this section we will two introduce algorithms, namely $\findpath$ and $\csort$ for computing paths between two given trees in $\rnni$.
-These will be used later to prove some geometric properties of the $\rnni$ graph.
+Within this section we will first two introduce algorithms, namely $\findpath$ and $\csort$, for computing paths between two given trees in $\rnni$.
+These will be used later to prove some geometric properties of the $\rnni$ graph in section~\ref{section:geometry}.
 Additionally we will present the algorithm $\mdtree$, which computes a caterpillar tree with maximum distance
@@
 
-To compute a path from tree $T$ to $R$ we consider clusters $C_1, \ldots, C_{n-1}$ of $R$ (the destination tree) where $C_i$ is induced by the node with rank $i$ for $i = 1, \ldots, n-1$.
+\todo{LC: move the following definition? - it's actually quite important not only for the algorithm, so we might want to have it in the introduction?}
+Before introducing the algorithm, we need the following definitions.
+Each node $v$ of a tree defines a \emph{cluster} $C$, which is the set of taxa descending from $v$.
+We then say that $v$ \emph{induces} $C$.
+\emph{Trivial clusters} are clusters induced by leaves (contain just one element each) and the root of the tree (contains all taxa).
+Note that the list of all
+\todo{Check that the indexing of $C_i$ is still correct.}
+non-trivial clusters of a tree, sorted according to the rank of the corresponding node, defines the tree unambiguously.
+For the rest of this paper we will use the notion cluster and mean non-trivial clusters only because all trees on $n$ taxa share the trivial clusters.
+\todo{LC: so we ever use trivial clusters? If so, we need to add a comment here}
+
+To compute a path from tree $T$ to $R$ we consider clusters $C_1, \ldots, C_{n-2}$ of $R$ (the destination tree) where $C_i$ is induced by the node with rank $i$ for $i = 1, \ldots, n-2$.
 \todo{In that case the number of clusters should be $n-2$.}
-Trivial clusters do not need to be changed on any path between two trees which is why they are not considered in the algorithm.
 First, set $\hat{T} = T$
@@
 AG: We need to explain quite a bit here -- there  are many ways one can do those moves + comment above.}
-Since we consider the clusters of $R$ such that the ranks of the corresponding internal nodes in $R$ increase, $\mrca_{\hat{T}}(C_i)$ is
-\todo{What is the `parent of a cluster' -- please check that you follow definitions you introduce.}
-the parent of cluster
-$C_i$ in $\hat{T}$ after these $\rnni$ moves.
+Since we consider the clusters of $R$ such that the ranks of the corresponding internal nodes in $R$ increase, $\mrca_{\hat{T}}(C_i)$ is the node inducing cluster $C_i$ in $\hat{T}$ after these $\rnni$ moves.
 Thus, repeating this procedure for all clusters of $R$ results in a path between any pair of tress $T, R$, indeed.
@@
 If you do it like this, explain that we need the algorithms for the geometry section.}
+
+%BEGIN: From MareikesVisit
+
+\todo{LC: Is it worth introducing intervals for $\rnni$ moves for the proof of the following lemma?}
+
+\begin{lemma}
+    Let $T$ be a tree and $C \subset \{1, \ldots, n\}$ a subset of taxa such that $C$ is not induced by any node in $T$.
+    There is exactly one tree $T'$ that can be obtained from $T$ by one $\rnni$ move such that $\rank(\mrca_{T'}(C)) = \rank(\mrca_{T}(C)) -1 $.
+\end{lemma}
+
+\begin{proof}
+
+    \todo{LC: revise this whole proof}
+
+    Let $(C_1, \ldots, C_n)$ be the list representation of $T$ as introduced above.
+    \todo{LC: Introduce this representation above!}
+    Let $v:=mrca_T(C)$, let $C_{i+1}$ be the cluster induced by $v$, and let $w$ be the node inducing $C_i$.
+    We will distinguish the case that there is an edge $(v,w)$ in $T$ from the case that there is no such edge.
+
+    At first we assume that there is no edge $(v,w)$, which means $C_i \not \subset C_{i+1}$.
+    In this case the two sets $C_i, C_{i+1}$ can exchange by a rank swap.
+    Since $C \subset C_{i+1}$, this rank swap decreases the rank of $\mrca(C)$ by one and there is no $\nni$ move that could move $\mrca(C)$ down.
+
+    Let us now consider the case that $(v,w)$ is an edge in $T$.
+    The only move that could decrease the rank of $\mrca(C)$ is an $\nni$ move on the edge $(v,w)$.
+    Let $C_k$ and $C_j$ be the clusters induced by the children of $C_i$.
+    Then it is $C_k \cup C_j = C_i$.
+    There are two different $\nni$ moves possible on $(v,w)$.
+    $C_i$ can either be replaced by $C_j \cup (C_{i+1} \setminus C_i)$ or by $C_k \cup (C_{i+1} \setminus C_i)$.
+    To prove that there is always exactly one $\rnni$ moves that decreases the rank of $\mrca(C)$, we need to show that it is either $C \subset (C_j \cup (C_{i+1} \setminus C_i))$ or $C \subset (C_k \cup (C_{i+1} \setminus C_i))$.
+
+    Let $a \in C$.
+    Then it is $a \in C_{i+1} = mrca(C)$ and it follows $a \in C_{i+1} \setminus C_i$ or $a \in C_i$.
+    If $a \in C_{i+1} \setminus C_i$, $a$ is in the new set after the $\nni$ move.
+    So now we will assume that $a \in C_i$.
+    Then it is either $a \in C_j$ or $a \in C_k$.
+    Without loss of generality we assume that $a \in C_j$.
+
+    We need to prove that if $a \in C_j$, there is no $b \in C$ with $b \in C_k$.
+    This makes sure that it is either $C \subset (C_j \cup (C_{i+1} \setminus C_i))$ or $C \subset (C_k \cup (C_{i+1} \setminus C_i))$.
+    For contradiction we assume that $a \in C_j$ and $C \in S_k$.
+    We know that all sets that are subsets of $C$ already are at the same position in the current tree as in the destination tree $R$.
+    Therefore, we can conclude from $a \in C_j$ and $b \in C_k$ that $C_j \cup C_k = C$.
+    Hence, it holds $mrca(C) = C_i$.
+    This is a contradiction to our assumption that $C_{i+1} = mrca(C)$ in the current tree.
+    Therefore, it holds either $C \subset (C_j \cup (C_{i+1} \setminus C_i))$ or $C \subset (C_k \cup (C_{i+1} \setminus C_i))$.
+
+    Concluding, there is always an $\rnni$ move on the edge $(v,w)$ for $mrca(C) = C_{i+1}, C \subsetneq C_{i+1}$ that decreases the rank of $mrca(C)$.
+
+\end{proof}
 
@@
 \end{algorithm}
-
-\subsection{Cluster Theorem}
-\label{section:computation_cluster_theorem}
 

### 2019-10-21 paper.tex
first pass through new alg section - lots of todos

@@
 Within this section we will first two introduce algorithms, namely $\findpath$ and $\csort$, for computing paths between two given trees in $\rnni$.
-These will be used later to prove some geometric properties of the $\rnni$ graph in section~\ref{section:geometry}.
+We will use them later to prove some geometric properties of the $\rnni$ graph in section~\ref{section:geometry}.
 Additionally we will present the algorithm $\mdtree$, which computes a caterpillar tree with maximum distance
@@
 
-\subsection{Algorithm $\findpath$}
+\subsection{$\findpath$}
 \label{section:alg_findpath}
@@
 Note that the list of all
-\todo{Check that the indexing of $C_i$ is still correct.}
+\todo{Check that the indexing of $C_i$ is still correct. ($i=1,\ldots,n-2$)}
 non-trivial clusters of a tree, sorted according to the rank of the corresponding node, defines the tree unambiguously.
@@
 
-To compute a path from tree $T$ to $R$ we consider clusters $C_1, \ldots, C_{n-2}$ of $R$ (the destination tree) where $C_i$ is induced by the node with rank $i$ for $i = 1, \ldots, n-2$.
-\todo{In that case the number of clusters should be $n-2$.}
-First, set $\hat{T} = T$
-Then for each $i$, find the node that induces $C_i$ in $\hat T$ and denote this node by $\mrca_{\hat{T}}(C_i)$ (short for \emph{most recent common ancestor}).
-Update $\hat{T}$ by
-\todo{This presumes  that the rank is higher.
-Why?}
-decreasing
-the rank of $\mrca_{\hat{T}}(C_i)$ by performing $\rnni$ moves until it has the same rank $i$ as in $R$.
-\todo{Do we need to explain why there is always an $\rnni$ move that moves the $\mrca$ down?
-AG: We need to explain quite a bit here -- there  are many ways one can do those moves + comment above.}
-Since we consider the clusters of $R$ such that the ranks of the corresponding internal nodes in $R$ increase, $\mrca_{\hat{T}}(C_i)$ is the node inducing cluster $C_i$ in $\hat{T}$ after these $\rnni$ moves.
+We are now presenting an algorithm for computing paths in $\rnni$ between any two given trees.
+Let $C_1, \ldots, C_{n-2}$ be the clusters of $R$ (the destination tree) where $C_i$ is induced by the node with rank $i$ for $i = 1, \ldots, n-2$.
+Within every step of the algorithm we add one more tree to extend a path $p$, which will be the output path of $\findpath$.
+First, set $\hat{T} = T$ to be the first element in $p$.
+With $\hat T$ we denote the last tree of $p$ at each step in the algorithm.
+Consider the clusters $C_1, \ldots, C_{n-2}$ of $R$ in this order.
+When arriving at cluster $C_k$, find the node inducing the smallest cluster containing all elements of $C_k$ in $\hat T$.
+This node is denoted by $\mrca_{\hat{T}}(C_k)$ (short for \emph{most recent common ancestor}).
+\todo{LC: make sure you don't use mrca before this!}
+The rank of $\mrca_{\hat{T}}(C_k)$ in $\hat T$ is higher than the rank of the node inducing $C_k$ in $R$, because we consider $C_i$ for increasing $i$.
+As a result, all clusters $C_1, \ldots, C_{i-1}$ are induced by some nodes of $\hat T$ as they are in the destination tree $R$.
+Now update $\hat{T}$ by performing an $\rnni$ move that decreases the rank of $\mrca_{\hat{T}}(C_i)$.
+With Lemma~\ref{lemma:mrca_move} we know that there is always exactly one $\rnni$ move possible that decreases the rank of $\mrca_{\hat{T}}(C_i)$.
+Repeat this until $\mrca_{\hat{T}}(C_i)$ has the same rank $i$ in $\hat T$ as in $R$.
 Thus, repeating this procedure for all clusters of $R$ results in a path between any pair of tress $T, R$, indeed.
-\todo{This kind of structure is not ideal (see my comment on the paper structure above).
-If we talk about the algorithm here, all the things, including the pseudo-code should be here.
-Possible solution: have algorithmic stuff first and geometry second.
-If you do it like this, explain that we need the algorithms for the geometry section.}
 
@@
 
-\todo{LC: Is it worth introducing intervals for $\rnni$ moves for the proof of the following lemma?}
+\todo{LC: Work on the following proof! Is it worth introducing intervals for $\rnni$ moves for the proof of the following lemma?}
 
@@
     There is exactly one tree $T'$ that can be obtained from $T$ by one $\rnni$ move such that $\rank(\mrca_{T'}(C)) = \rank(\mrca_{T}(C)) -1 $.
+    \label{lemma:mrca_move}
 \end{lemma}
@@
 
-\todo{LC: Change the references to sections}
-In Section~\ref{section:geometry} we introduced the algorithm $\findpath$ for computing paths between arbitrary trees in $\rnni$.
-Therefore, it approximates distances in $\rnni$.
-In this section we are providing pseudocode for $\findpath$ (Algorithm~\ref{alg:find_path}) to provide further insights.
-We implemented and tested this algorithm for trees on up to $k$ taxa, for which it gives the exact $\rnni$ distance. \todo{$k = ?$, How exactly did we test it?}
-The running time of $\findpath$ is quadratic in the number of taxa.
+With $\findpath$ we can compute a path between two arbitrary trees in $\rnni$.
+The length of such a path can then be used as approximation for the distance between the trees.
+We implemented $\findpath$ in Python.
+\todo{LC: cite, link to github code?}
+\todo{LC: Shall we mention the exact algorithm at the beginning of this section already?}
+In~\autocite{Gavryushkin2018-ol} an algorithm for computing the $\rnni$ graph is being introduced.
+We implemented this algorithm
+\todo{LC: github?}
+and used
+\todo{LC: which algorithms did we use?}
+to compute exact distances between all pairs of trees on up to $7$ taxa.
+\todo{LC: Is 7 the actual number?}
+By comparing these distances with the distances given by $\findpath$, we find that $\findpath$ gives the correct distances for all pairs of trees on up to $7$ taxa.
+\todo{LS: Is it 7? How did we compute all distances with $\findpath$? - We used the symmetry of $\rnni$}
+The running time of $\findpath$, implemented as described by the pseudo-code in Algorithm~\ref{alg:find_path}, is quadratic in the number of taxa.
 
@@
 
-Recall that Section~\ref{section:diameter} revealed that the maximum distance computed by $\findpath$ for an arbitrary pair of trees equals the diameter of $\rnni$.
+In Section~\ref{section:diameter} we will see that for any pair of trees, $\findpath$ computes a path that is at most as long as the the diameter of $\rnni$.
 Thus, if the distance between two trees equals the diameter, $\findpath$ computes the correct distance.
-Following this observation, we claim Conjecture~\ref{conjecture:findpath_maxdist_correct}.
-Notice that one direction of the conjecture is already known to be true, it only remains open whether the distance of two trees is $\Delta(\rnni)$ indeed, if $\findpath$ computes a path of this length.
+\todo{LC: Can we say something about the approximation ratio here?}
 
-\begin{conjecture}
-    Let $T$ and $R$ be two trees.
-    It is $d(T,R) = \frac{(n-1)(n-2)}{2}$ if, and only if, $\findpath(T,R)$ computes a path of this length between $T$ and $R$.
-    \label{conjecture:findpath_maxdist_correct}
-\end{conjecture}
-
-\subsection{Algorithm $\cpath$}
+\subsection{$\csort$}
 \label{section:alg_cpath}
@@
 
-Before we start analysing shortest paths between caterpillar trees, we need to introduce further notations.
+Before we start introducing an algorithm for computing paths between caterpillar trees, we need to introduce further notations.
 A path $p$ between two caterpillar trees $T$ and $R$ that only consists of caterpillar trees is called \emph{caterpillar path}.
@@
 
-We next introduce a unique representation of caterpillar trees to show how one can use the classical \emph{Bubble Sort} algorithm \autocite{Knuth1997-pi} to compute caterpillar paths.
+\todo{LC: structure in the following paragraph!}
+We next introduce a unique way to represent of caterpillar trees as sequences of taxa.
+We will use this to show how one can use the classical \emph{Bubble Sort} algorithm \autocite{Knuth1997-pi} to compute caterpillar paths.
+This algorithm is denoted by $\csort$ and the pseudo-code is provided in Algorithm.
+\todo{LC: pseudo-code}
 One can see caterpillar trees as sequences of taxa, ordered non-decreasing according to the ranks of their parents.
+Note that the only two elements in such a sequence that have parents of equal rank are the first two elements, the cherry taxa $c_1$ and $c_2$.
+To make the sequence representation of caterpillar trees unique, we assume that for it is $c_1 < c_2$.
 $\nni$ moves between caterpillar trees, which are the only possible $\rnni$ moves between those trees, coincide with exchanges of two neighboured elements in the sequence.
-In these sequences the first two taxa are somehow equivalent since they build the unique cherry which means they share their parent with rank one.
-Therefore, we need to consider $\nni$ moves that change the cherry of a tree separately.
-These moves refer to exchanges of the first or the second element with the third element of the sequence, while exchanges of the first two elements do not change the corresponding tree.
-With this observation we can provide an algorithm for computing shortest caterpillar paths by a simple modification of the commonly known Bubble Sort algorithm.
-Throughout this paper this algorithm will be called $\csort$.
+The only exception are the two first three elements in the sequence.
+Corresponding to $\nni$ moves on the lowest internal edge are exchanges of either $c_1$ or $c_2$ with the third elements $c_3$ in the sequence.
+But a swap of $c_1$ and $c_2$ does not correspond to any move in $\rnni$.
+Therefore, we need to adapt Bubble Sort for the first three elements of the sequence as follows:
+\todo{LC: add the pseudocode and refer to the line where $\csort$ differs from Bubble Sort! (We assume that we are at the point where we do some changes only on the first positions of the sequence)}
+If the rank of $\parent_R(c_3)$ is higher than the ranks of both $\parent_R(c_1)$ and $\parent_R(c_2)$ in $R$, there are no swaps needed for these taxa.
+If just one of the two taxa $c_1$ or $c_2$ is on the other side of $c_3$ in $R$, this taxon swaps position with $c_3$ in the sequence.
+And if both $c_1$ and $c_2$ are on the other side of $c_3$ in $R$, the taxon that has the parent with higher rank in $R$ swaps with $c_3$.
+This makes sure that there is no additional swap of $c_1$ and $c_2$ needed on the subsequent path to $R$.
 
-Following the observations above, only swaps of the first three elements of a sequence with Bubble Sort need to be adapted for the computation of shortest caterpillar paths.
-Let us consider the first three elements $a,b,c$, appearing in this order in a sequence representing a caterpillar tree on a caterpillar path between trees $T$ and $R$.
 \todo{Check throughout that $\parent$ is never used without tree name in the index.}
-If the rank of $\parent(c)$ is higher than the ranks of $\parent(a)$ and $\parent(b)$ in $R$, there are no swaps needed for these taxa.
-If just one of the two taxa $a$ and $b$ is on the other side of $c$ in $R$, this taxon swaps position with $c$ in the sequence.
-And if both $a$ and $b$ are on the other side of $c$ in $R$, the taxon that has the parent with highest rank of all these three taxa in $R$ swaps with $c$ at first.
-This makes sure that there is no additional swap of $a$ and $b$ needed on the subsequent path to $R$.
-\todo{Do we need to explain Bubble Sort?}
 
-\subsection{Algorithm $\mdtree$}
+\subsection{$\mdtree$}
 \label{section:alg_mdtree}
 
-For finding the radius of the $\rnni$ graph, algorithm $\mdtree$ has been introduced in Section~\ref{section:diameter}.
-It computes a tree $R$ with distance $\Delta(\rnni)$ from a given caterpillar tree, as it has been proven in Lemma~\ref{lemma:max_dist_caterpillar}.
+The algorithm we are going to introduce here will be important for finding the radius of the $\rnni$ graph.
+In Section~\ref{section:diameter} we will prove that it computes a tree with maximum distance from a given caterpillar tree, which implies that the radius of $\rnni$ equals its diameter.
+Therefore, it is called $\mdtree$ (Maximum Distance Tree).
+\todo{LC: References to the corresponding proofs?}
+
+$\mdtree$ works as follows for an input tree $T$:
+At first, all taxa of $T$ are sorted in a list $L$ such that the ranks of their parents are non-decreasing.
+If the input is a caterpillar tree, it follows that $L[0]$ and $L[1]$ are the cherry taxa of $T$, in an arbitrary order, and $\parent(L[i])$ has rank $i$ in $T$ for all $i = 2, \ldots, n-1$.
+Note that this list $L$ does not give a unique representation of a tree.
+The output tree $R$ is built inductively as follows:
+Initially, $R$ only consists of a cherry with taxa $L[0]$ and $L[1]$ such that the parent of these taxa has rank $n-1$.
+All following taxa in $L$ are added to $R$ sequentially, according to their order in $L$, by creating a new internal node at one of the existing branches in $R$ such that this new node has rank one less than the previously lowest internal node in $R$.
+
 Algorithm~\ref{alg:max_dist_tree} provides the pseudocode for this algorithm.
-It is important to note that $\mdtree$ can not only be applied onto caterpillar trees, as it is done in Section~\ref{section:diameter}, but to any arbitrary tree.
-However, it does not necessarily produce a tree with maximum distance from a given tree.
+We will see in Section~\ref{section:diameter} that it computes trees with maximum distance form any given caterpillar tree.
+But if the given tree is an arbitrary tree, the output tree might not have maximum distance to the input tree.
 
-%TODO
-% Shall we give an example where the distance is not max or simulate trees and compute histogram of distances approximated by $\findpath$?
+\todo{LC: Shall we give an example where the distance is not max or simulate trees and compute histogram of distances approximated by $\findpath$?}
 
@@
 We now proceed to show that the radius of the $\rnni$ graph equals its diameter.
-For this aim an algorithm for constructing a tree $R$ with distance $\Delta(\rnni)$ from a given caterpillar tree $T$ will be used.
-This algorithm will be called $\mdtree$ (Maximum Distance Tree).
-After introducing this algorithm, we will prove that a tree $R = \mdtree(T)$ has distance $d(T,R) = \Delta(\rnni)$ to $T$ if $T$ is a caterpillar tree.
+Therefore we use the algorithm $\mdtree$ that has been introduced in section~\ref{section:alg_mdtree}.
 
-$\mdtree$ works as follows:
-At first, all taxa of $T$ are sorted in a list $L$ such that the ranks of their parents are non-decreasing.
-If the input is a caterpillar tree, it follows that $L[0]$ and $L[1]$ are the cherry taxa of $T$ and $\parent(L[i])$ has rank $i$ in $T$ for all $i = 2, \ldots, n-1$.
-Now $R$ is built inductively as follows:
-Initially, $R$ only consists of a cherry with taxa $L[0]$ and $L[1]$ such that the parent of these taxa has rank $n-1$.
-All following taxa in $L$ are added to $R$ sequentially, according to their order in $L$, by creating a new internal node at one of the existing branches in $R$ such that this new node has rank one less than the previously lowest internal node in $R$.
-When this is done for all taxa in $L$, the resulting tree $R$ is an $\rnni$ tree.
-
-It is important to note that this algorithm can be applied to any input tree $T$, which does not need to be a caterpillar tree.
 However, the results provided in this section are only true for caterpillar trees.

### 2019-10-23 paper.tex
once again through intro + alg sections

@@
 
+\todo{LC: mention some results here and how they make $\rnni$ differ from $\nni$}
+This distinguishes $\rnni$ from $\nni$ where this does not hold.
+One further result, presented in Section~\ref{section:diameter}, directly following from the proof of Theorem~\ref{thm:caterpillar_convex}, is the diameter of $\rnni$.
+Finally, we disprove the so-called Strong Split Theorem, conjectured in \autocite{Gavryushkin2018-ol}.
+However, the Weak Split Theorem remains a conjecture.
+Furthermore, we propose an alternative (even weaker) version of the conjecture, the Cluster Theorem (Section~\ref{section:split_theorem}).
+
 In this paper, we consider the $\rnni$ space on the ranked phylogenetic trees with all taxa being of equal rank.
@@
 
-A \emph{ranked phylogenetic tree} is a pair consisting of a rooted binary phylogenetic tree $T$ on the set $X = \{1, \ldots, n\}$ of taxa for $n \in \mathbb N$, and a (total) rank function $\rank$ that maps all elements of $X$ to $0$, all internal nodes of $T$ onto the set $\{1, \ldots, n-1\}$, and respects the partial order given by the tree.
-The latter means that if $x$ and $y$ are two internal nodes of $T$ such that there exists a path from a taxon $z \in X$ to $y$ which passes through $x$ then $\rank(x) < \rank(y)$.
+A \emph{ranked phylogenetic tree} is a pair consisting of a rooted binary phylogenetic tree $T$ on the set $X = \{1, \ldots, n\}$ of \emph{taxa} for $n \in \mathbb N$, and a (total) rank function $\rank$ that maps all elements of $X$ to $0$, all internal nodes of $T$ onto the set $\{1, \ldots, n-1\}$, and respects the partial order given by the tree.
+The latter means that if $u$ and $v$ are two internal nodes of $T$ such that there exists a path from a taxon $x \in X$ to the root which first passes through $u$ and then through $v$ then $\rank(u) < \rank(v)$.
 Ranked trees $(T_1, \rank_1)$ and $(T_2, \rank_2)$ are different if trees $T_1$ and $T_2$ are different or $T_1 = T_2$ and $\rank_1 \neq \rank_2$.
@@
 Our definition of a ranked tree implies a natural notion of the edge length.
-Indeed, a tree edge $(u,v)$ has length $l$ if $|\rank(v) - \rank(u)| = l$.
+Indeed, a tree edge $(u,v)$ of $T$ has length $l$ if $|\rank_T(v) - \rank_T(u)| = l$.
 
@@
 The vertex set of the $\rnni$ graph is the set of all ranked trees on $n$ taxa.
-We introduce two types of operation ($\rnni$ \emph{moves}) on trees (see Figure~\ref{fig:RNNI}) and say that two trees are adjacent in the $\rnni$ graph is they are connected by an operation of either type.
+We introduce two types of operation ($\rnni$ \emph{moves}) on trees (see Figure~\ref{fig:RNNI}) and say that two trees are adjacent in the $\rnni$ graph if they are connected by an operation of either type.
 The first type of operation is called a \emph{rank move} and defined by swapping the ranks of two internal nodes that are not adjacent in the tree.
-Formally, if $x$ and $y$ are such nodes of a tree $T$ that $|\rank_T(x) - \rank_T(y)| = 1$ and $(x, y)$ is not an edge in $T$ then the tree $R$ obtained from $T$ by only changing $\rank_R(x) = \rank_T(y)$ and $\rank_R(y) = \rank_T(x)$ is said to be obtained by a rank move.
+Formally, if $u$ and $v$ are such nodes of a tree $T$ that $|\rank_T(u) - \rank_T(v)| = 1$ and $(u, v)$ is not an edge in $T$ then the tree $R$ obtained from $T$ by only changing $\rank_R(u) = \rank_T(v)$ and $\rank_R(v) = \rank_T(u)$ is said to be obtained by a rank move.
 The second type of operation is called an $\nni$ \emph{move} and defined in the usual way, that is, two trees $T$ and $R$ are said to be connected by an $\nni$ move if there exist internal edges $e$ in $T$ and $f$ in $R$ both of length one such that the trees obtained by shrinking $e$ and $f$ to internal nodes coincide.
@@
 	\centering
-	\includegraphics[width=\textwidth]{RNNI}
+	\includegraphics[width=0.8\textwidth]{RNNI}
+    \vspace{2pt}
 	\caption{Two types of operation that define edges of the $\rnni$ graph.}
@@
 \end{figure}
-
-\todo{label of taxa in figure shouldn't be numbers\\
-AG: Well -- if you decide to change this, go through all the definitions VERY carefully to make sure we update everywhere.}
 
@@
 
-\todo{LC: make sure that we move all definitions in this paper to the right place}
-
-Within this section we will first two introduce algorithms, namely $\findpath$ and $\csort$, for computing paths between two given trees in $\rnni$.
+Within this section we will first introduce two algorithms, namely $\findpath$ (Section~\ref{section:alg_findpath}) and $\csort$ (Section~\ref{section:alg_csort}), for computing paths between two given trees in $\rnni$.
+The first one takes arbitrary trees as input while the latter one is restricted to trees with a certain tree shape.
 We will use them later to prove some geometric properties of the $\rnni$ graph in section~\ref{section:geometry}.
-Additionally we will present the algorithm $\mdtree$, which computes a caterpillar tree with maximum distance
-\todo{LC: can we use the term max dist here?}
-from a given tree.
-
-% In \autocite{Gavryushkin2018-ol} an efficient algorithm for computing the $\rnni$ graph has been introduced.
-% We implemented this algorithm, which produces the graph for trees with up to $8$ taxa in reasonable time.
-\todo{Shall we say more about the algorithm here?
-+ running time (in hours/minutes) on which computer?}
-% This enables us to compute exact distances between trees with up to $8$ taxa.
-% In particular, comparing exact distances with approximations as introduced in previous sections and verifying the Cluster Theorem of Section~\ref{section:split_theorem} can now be for small trees.
-% All algorithms have been implemented with Python and are available online. \todo{link?}
-% In the following, further details about the implementation of algorithms $\findpath$ and $\mdtree$ as well as a description of how to prove the Cluster Theorem will be given.
+Additionally we will present the algorithm $\mdtree$, which produces a tree that is far away from the given input tree.
+What we mean by far away depends on the input tree and will be discussed in Section~\ref{section:alg_mdtree}.
 
@@
 
-\todo{LC: Introducing words}
-
-\todo{LC: move the following definition? - it's actually quite important not only for the algorithm, so we might want to have it in the introduction?}
-Before introducing the algorithm, we need the following definitions.
+Before introducing the algorithm we need the following definitions.
 Each node $v$ of a tree defines a \emph{cluster} $C$, which is the set of taxa descending from $v$.
@@
 \emph{Trivial clusters} are clusters induced by leaves (contain just one element each) and the root of the tree (contains all taxa).
-Note that the list of all
-\todo{Check that the indexing of $C_i$ is still correct. ($i=1,\ldots,n-2$)}
-non-trivial clusters of a tree, sorted according to the rank of the corresponding node, defines the tree unambiguously.
+Note that the list of all non-trivial clusters of a tree, sorted according to the rank of the corresponding node, defines the tree unambiguously.
 For the rest of this paper we will use the notion cluster and mean non-trivial clusters only because all trees on $n$ taxa share the trivial clusters.
-\todo{LC: so we ever use trivial clusters? If so, we need to add a comment here}
 
-We are now presenting an algorithm for computing paths in $\rnni$ between any two given trees.
+We are now presenting $\findpath$ (Algorithm~\ref{alg:find_path}), an algorithm for computing paths between any two given trees in $\rnni$.
 Let $C_1, \ldots, C_{n-2}$ be the clusters of $R$ (the destination tree) where $C_i$ is induced by the node with rank $i$ for $i = 1, \ldots, n-2$.
-Within every step of the algorithm we add one more tree to extend a path $p$, which will be the output path of $\findpath$.
-First, set $\hat{T} = T$ to be the first element in $p$.
-With $\hat T$ we denote the last tree of $p$ at each step in the algorithm.
-Consider the clusters $C_1, \ldots, C_{n-2}$ of $R$ in this order.
-When arriving at cluster $C_k$, find the node inducing the smallest cluster containing all elements of $C_k$ in $\hat T$.
+$\findpath$ computes a path $p$ which initially consists of $T$ only and is extended stepwise.
+Let $\hat T$ denote the last tree of $p$ at any point in the algorithm.
+For extending $p$, which equals updating $\hat T$, the clusters $C_1, \ldots, C_{n-2}$ are considered in this order.
+When arriving at cluster $C_i$ the clusters in $\hat T$ are rearranged by some $\rnni$ moves until $C_i$ is induced by a node in $\hat T$.
+This works as follows.
+First find the node inducing the smallest cluster containing all elements of $C_k$ in $\hat T$.
 This node is denoted by $\mrca_{\hat{T}}(C_k)$ (short for \emph{most recent common ancestor}).
-\todo{LC: make sure you don't use mrca before this!}
 The rank of $\mrca_{\hat{T}}(C_k)$ in $\hat T$ is higher than the rank of the node inducing $C_k$ in $R$, because we consider $C_i$ for increasing $i$.
@@
 Now update $\hat{T}$ by performing an $\rnni$ move that decreases the rank of $\mrca_{\hat{T}}(C_i)$.
-With Lemma~\ref{lemma:mrca_move} we know that there is always exactly one $\rnni$ move possible that decreases the rank of $\mrca_{\hat{T}}(C_i)$.
-Repeat this until $\mrca_{\hat{T}}(C_i)$ has the same rank $i$ in $\hat T$ as in $R$.
-Thus, repeating this procedure for all clusters of $R$ results in a path between any pair of tress $T, R$, indeed.
+With Lemma~\ref{lemma:mrca_move} we know that there exists always exactly one $\rnni$ move that decreases the rank of $\mrca_{\hat{T}}(C_i)$.
+So we can perform these moves until $\mrca_{\hat{T}}(C_i)$ has the same rank $i$ in $\hat T$ as in $R$.
+Repeating this procedure for all clusters of $R$ results in a path from $T$ to $R$.
+
+\begin{algorithm}[H]
+\caption{$\findpath$($T,R$)}
+\label{alg:find_path}
+\begin{algorithmic}[1]
+	\STATE $\hat{T} := T$
+	\STATE $p := [\hat T]$
+	\FOR {$i = 1, \dots, n-2$}
+		\STATE Let $C_i$ be the cluster induced by the node with rank $i$ in $R$
+		\WHILE {$\rank(\mrca_{\hat{T}}(C_i))>i$}
+			\STATE Update $\hat{T}$: Decrease $\rank(\mrca_{\hat{T}}(C_i))$ by an $\rnni$ move \label{alg:line:move_set_down}
+			\STATE $p = p+\hat{T}$
+		\ENDWHILE
+	\ENDFOR
+	\RETURN $p$
+\end{algorithmic}
+\end{algorithm}
 
@@
 
-With $\findpath$ we can compute a path between two arbitrary trees in $\rnni$.
-The length of such a path can then be used as approximation for the distance between the trees.
-We implemented $\findpath$ in Python.
-\todo{LC: cite, link to github code?}
-\todo{LC: Shall we mention the exact algorithm at the beginning of this section already?}
-In~\autocite{Gavryushkin2018-ol} an algorithm for computing the $\rnni$ graph is being introduced.
-We implemented this algorithm
-\todo{LC: github?}
-and used
-\todo{LC: which algorithms did we use?}
-to compute exact distances between all pairs of trees on up to $7$ taxa.
-\todo{LC: Is 7 the actual number?}
-By comparing these distances with the distances given by $\findpath$, we find that $\findpath$ gives the correct distances for all pairs of trees on up to $7$ taxa.
-\todo{LS: Is it 7? How did we compute all distances with $\findpath$? - We used the symmetry of $\rnni$}
+Since $\findpath$ computes a path between two arbitrary trees in $\rnni$, we can use the length of such a path as approximation for the distance between the trees.
+A natural question is now how well this estimates the actual distance in $\rnni$.
+To answer this question for small trees, we implemented $\findpath$.
+\todo{LC: github - link}
+For computing the actual $\rnni$ distances we used the the algorithm for computing the $\rnni$ graph that was introduced in Section 3.3 of \autocite{Gavryushkin2018-ol}.
+With our implementation of this algorithm
+\todo{LC: gihub - link}
+we can compute the graph for trees with up to $8$ taxa in reasonable time. \todo{LC: Shall we say more about the algorithm here? + running time (in hours/minutes) on which computer?}
+We used
+\todo{LC: How do we compare the distances?}
+
 The running time of $\findpath$, implemented as described by the pseudo-code in Algorithm~\ref{alg:find_path}, is quadratic in the number of taxa.
-
-\begin{algorithm}[H]
-\caption{$\findpath$($T,R$)}
-\label{alg:find_path}
-\begin{algorithmic}[1]
-	\STATE $\hat{T} := T$
-	\STATE $p := (T)$
-	\FOR {$i = 1, \dots, n-2$}
-		\STATE Let $C$ be the cluster induced by node $v$ in $R$ with $\rank(v) = i$
-		\WHILE {$\rank(\mrca_{\hat{T}}(C))>i$}
-			\STATE Update $\hat{T}$: Decrease $\rank(\mrca_{\hat{T}}(C))$ by an $\rnni$ move \label{alg:line:move_set_down}
-			\STATE $p = p+\hat{T}$
-		\ENDWHILE
-	\ENDFOR
-	\RETURN $p$
-\end{algorithmic}
-\end{algorithm}
-
-In Section~\ref{section:diameter} we will see that for any pair of trees, $\findpath$ computes a path that is at most as long as the the diameter of $\rnni$.
-Thus, if the distance between two trees equals the diameter, $\findpath$ computes the correct distance.
-\todo{LC: Can we say something about the approximation ratio here?}
+\todo{LC: Where do we mention the running time?}
 
 \subsection{$\csort$}
-\label{section:alg_cpath}
+\label{section:alg_csort}
 
@@
 
-Before we start introducing an algorithm for computing paths between caterpillar trees, we need to introduce further notations.
+Before introducing an algorithm for computing paths between caterpillar trees, we need to introduce further notations.
+We aim to find paths between \emph{caterpillar trees} where each internal node has at least one child that is a leaf.
 A path $p$ between two caterpillar trees $T$ and $R$ that only consists of caterpillar trees is called \emph{caterpillar path}.
@@
 
-\todo{LC: structure in the following paragraph!}
-We next introduce a unique way to represent of caterpillar trees as sequences of taxa.
-We will use this to show how one can use the classical \emph{Bubble Sort} algorithm \autocite{Knuth1997-pi} to compute caterpillar paths.
-This algorithm is denoted by $\csort$ and the pseudo-code is provided in Algorithm.
+We next introduce a representation of caterpillar trees as sequences of taxa.
+We will use it to show how one can adapt the classical \emph{Bubble Sort} algorithm \autocite{Knuth1997-pi} to compute caterpillar paths.
 \todo{LC: pseudo-code}
-One can see caterpillar trees as sequences of taxa, ordered non-decreasing according to the ranks of their parents.
-Note that the only two elements in such a sequence that have parents of equal rank are the first two elements, the cherry taxa $c_1$ and $c_2$.
-To make the sequence representation of caterpillar trees unique, we assume that for it is $c_1 < c_2$.
-$\nni$ moves between caterpillar trees, which are the only possible $\rnni$ moves between those trees, coincide with exchanges of two neighboured elements in the sequence.
-The only exception are the two first three elements in the sequence.
-Corresponding to $\nni$ moves on the lowest internal edge are exchanges of either $c_1$ or $c_2$ with the third elements $c_3$ in the sequence.
-But a swap of $c_1$ and $c_2$ does not correspond to any move in $\rnni$.
-Therefore, we need to adapt Bubble Sort for the first three elements of the sequence as follows:
-\todo{LC: add the pseudocode and refer to the line where $\csort$ differs from Bubble Sort! (We assume that we are at the point where we do some changes only on the first positions of the sequence)}
-If the rank of $\parent_R(c_3)$ is higher than the ranks of both $\parent_R(c_1)$ and $\parent_R(c_2)$ in $R$, there are no swaps needed for these taxa.
-If just one of the two taxa $c_1$ or $c_2$ is on the other side of $c_3$ in $R$, this taxon swaps position with $c_3$ in the sequence.
-And if both $c_1$ and $c_2$ are on the other side of $c_3$ in $R$, the taxon that has the parent with higher rank in $R$ swaps with $c_3$.
-This makes sure that there is no additional swap of $c_1$ and $c_2$ needed on the subsequent path to $R$.
+Ordering the taxa of a caterpillar trees as a non-decreasing, according to the ranks of their parents, gives us a sequence of length $n$.
+Note that the only two elements in such a sequence that have parents of equal rank are the first two elements $c_1$ and $c_2$.
+We will call a pair of two taxa that share their parent \emph{cherry}.
+\todo{LC: pair of taxa or subtree (includes internal node and two edges)}
+To make the sequence representation of caterpillar trees unique, we assume that it is $c_1 < c_2$ for the first two elements in the sequence.
+
+In the following we explain how $\rnni$ moves between caterpillar trees work in the sequence representation.
+At first it is important to realise that the only possible $\rnni$ moves between caterpillar trees are $\nni$ moves, since there is no pair of internal nodes with rank difference one that is not connected by an edge.
+$\nni$ moves between caterpillar trees correspond to swaps of two neighboured elements in the sequence representation.
+The only exception of this are $\nni$ moves on the lowest internal edge, that is the edge connecting the internal nodes of rank one and two.
+There are two $\nni$ moves possible on this edge that result in caterpillar trees, while there is for every other edge only one $\nni$ move that results in a caterpillar tree.
+Corresponding to the $\nni$ moves on the lowest internal edge are exchanges of either the first or the second element with the third elements in the sequence representation.
+
+For computing shortest caterpillar paths we modify Bubble Sort as described in the following.
+We call this algorithm $\csort$ (Algorithm~\ref{alg:csort}).
+
+\todo{Description + Pseudo code $\csort$}
+% But a swap of $c_1$ and $c_2$ does not correspond to any move in $\rnni$.
+% Therefore, we need to adapt Bubble Sort for the first three elements of the sequence as follows:
+% \todo{LC: add the pseudocode and refer to the line where $\csort$ differs from Bubble Sort! (We assume that we are at the point where we do some changes only on the first positions of the sequence)}
+% If the rank of $\parent_R(c_3)$ is higher than the ranks of both $\parent_R(c_1)$ and $\parent_R(c_2)$ in $R$, there are no swaps needed for these taxa.
+% If just one of the two taxa $c_1$ or $c_2$ is on the other side of $c_3$ in $R$, this taxon swaps position with $c_3$ in the sequence.
+% And if both $c_1$ and $c_2$ are on the other side of $c_3$ in $R$, the taxon that has the parent with higher rank in $R$ swaps with $c_3$.
+% This makes sure that there is no additional swap of $c_1$ and $c_2$ needed on the subsequent path to $R$.
 
@@
 In this section we analyse shortest paths in $\rnni$.
-We will start by presenting an efficient algorithm for computing (not necessarily shortest) paths in $\rnni$ ($\findpath$) that turns out to be an essential tool for understanding the geometry of $\rnni$.
-Specifically, this algorithm will enable us to prove that there is always a shortest path between two caterpillar trees that only consists of caterpillar trees (Theorem~\ref{thm:caterpillar_convex}) in Section~\ref{section:caterpillar_convex}.
+Therefore, we will use the algorithms introduced in the previous section.
+They enable us to prove that there is always a shortest path between two caterpillar trees that only consists of caterpillar trees (Theorem~\ref{thm:caterpillar_convex}) in Section~\ref{section:caterpillar_convex}.
 In other words, the set of caterpillar trees is \emph{convex} in $\rnni$.
-\todo{The rest of this paragraph should be carefully checked and be in the introduction.
-Generally, you might want to think a bit more about a better structure of this paper into chapters.}
-This distinguishes $\rnni$ from $\nni$ where this does not hold.
-One further result, presented in Section~\ref{section:diameter}, directly following from the proof of Theorem~\ref{thm:caterpillar_convex}, is the diameter of $\rnni$.
-Finally, we disprove the so-called Strong Split Theorem, conjectured in \autocite{Gavryushkin2018-ol}.
-However, the Weak Split Theorem remains a conjecture.
-Furthermore, we propose an alternative (even weaker) version of the conjecture, the Cluster Theorem (Section~\ref{section:split_theorem}).
 

### 2019-10-23 paper.tex
add description + pseudo-code csort

@@
 
-We next introduce a representation of caterpillar trees as sequences of taxa.
+We next introduce a representation of caterpillar trees as lists of taxa.
 We will use it to show how one can adapt the classical \emph{Bubble Sort} algorithm \autocite{Knuth1997-pi} to compute caterpillar paths.
 \todo{LC: pseudo-code}
-Ordering the taxa of a caterpillar trees as a non-decreasing, according to the ranks of their parents, gives us a sequence of length $n$.
-Note that the only two elements in such a sequence that have parents of equal rank are the first two elements $c_1$ and $c_2$.
+Ordering the taxa of a caterpillar trees as a non-decreasing, according to the ranks of their parents, gives us a list of length $n$.
+Note that the only two elements in such a list that have parents of equal rank are the first two elements $c_1$ and $c_2$.
 We will call a pair of two taxa that share their parent \emph{cherry}.
 \todo{LC: pair of taxa or subtree (includes internal node and two edges)}
-To make the sequence representation of caterpillar trees unique, we assume that it is $c_1 < c_2$ for the first two elements in the sequence.
+To make the list representation of caterpillar trees unique, we assume that it is $c_1 < c_2$ for the first two elements in the list.
 
-In the following we explain how $\rnni$ moves between caterpillar trees work in the sequence representation.
+In the following we explain how $\rnni$ moves between caterpillar trees work in the list representation.
 At first it is important to realise that the only possible $\rnni$ moves between caterpillar trees are $\nni$ moves, since there is no pair of internal nodes with rank difference one that is not connected by an edge.
-$\nni$ moves between caterpillar trees correspond to swaps of two neighboured elements in the sequence representation.
+$\nni$ moves between caterpillar trees correspond to swaps of two neighboured elements in the list representation.
 The only exception of this are $\nni$ moves on the lowest internal edge, that is the edge connecting the internal nodes of rank one and two.
 There are two $\nni$ moves possible on this edge that result in caterpillar trees, while there is for every other edge only one $\nni$ move that results in a caterpillar tree.
-Corresponding to the $\nni$ moves on the lowest internal edge are exchanges of either the first or the second element with the third elements in the sequence representation.
+Corresponding to the $\nni$ moves on the lowest internal edge are exchanges of either the first or the second element with the third elements in the list representation.
 
@@
 \todo{Description + Pseudo code $\csort$}
-% But a swap of $c_1$ and $c_2$ does not correspond to any move in $\rnni$.
-% Therefore, we need to adapt Bubble Sort for the first three elements of the sequence as follows:
-% \todo{LC: add the pseudocode and refer to the line where $\csort$ differs from Bubble Sort! (We assume that we are at the point where we do some changes only on the first positions of the sequence)}
-% If the rank of $\parent_R(c_3)$ is higher than the ranks of both $\parent_R(c_1)$ and $\parent_R(c_2)$ in $R$, there are no swaps needed for these taxa.
-% If just one of the two taxa $c_1$ or $c_2$ is on the other side of $c_3$ in $R$, this taxon swaps position with $c_3$ in the sequence.
-% And if both $c_1$ and $c_2$ are on the other side of $c_3$ in $R$, the taxon that has the parent with higher rank in $R$ swaps with $c_3$.
-% This makes sure that there is no additional swap of $c_1$ and $c_2$ needed on the subsequent path to $R$.
+We consider the input trees $T$ and $R$ in their list representation $L_T$ and $L_R$ as described above and want to compute a path from $L_T$ to $L_R$.
+A path $p$ is built stepwise by adding a new list to the end of $p$.
+We call the last element of $p$ at any step $L_{\hat T}$ and when we say we update $L_{\hat T}$, we modify this list and append the new version of $L_{\hat T}$ to $p$.
+
+As in the Bubble Sort algorithm, we go through the list $L_{\hat T}$ and swap two elements if they appear in different orders in $L_{\hat T}$ and $L_R$.
+Each swap of elements is an update of $\hat T$ and extends the path $p$ by one element.
+We compare pairs of list elements from the start of the list until its end, and repeat this until the whole list is sorted, just as in classical Bubble Sort.
+Our algorithm $\csort$ only differs from Bubble Sort when considering the first three elements of the list.
+Since the first two elements refer to the cherry taxa of the caterpillar tree, both can exchange with the third element of the list as $\nni$ move.
+Therefore, we need to consider these moves separately to all other moves.
+Most importantly, we need to make sure that these moves are performed in a way that leads to a shortest caterpillar path.
+We can guarantee this, if the moves are done as described in the following.
+Let $c_1, c_2$ and $c_3$ be the first three elements of $L_{\hat T}$, in this order.
+If $c_1$ ($c_2$) is the last element of those three in $L_R$, we update $L_{\hat T}$ by swapping $c_1$ ($c_2$) and $c_3$.
+By doing this we make sure that there will be no point on $p$ where $c_1$ and $c_2$ swap positions, which would cause an additional $\nni$ move that can be avoided if we perform the moves as we describe them here.
+
+Because all moves between caterpillar trees are $\nni$ moves as described in the previous paragraph, all pairs of taxa that are not in the same order in $L_T$ and $L_R$ need to swap at some point on any path between $L_T$ and $L_R$.
+And since only elements that are direct neighbours in a sequence are allowed to swap, with the only exception being the first three elements, the path computed by $\csort$ is a shortest caterpillar path.
+\todo{LC: Do we need further explanation why $\csort$ gives a shortest caterpillar path?}
+
+\todo{LC: improve readability of pseudo-code}
+
+\begin{algorithm}[H]
+\caption{$\csort$($T,R$)}
+\label{alg:csort}
+\begin{algorithmic}[1]
+    \STATE $\hat T:= T$, $p = [\hat T]$
+    \STATE $L:=$ list representation of $\hat T$
+	% \STATE $L_{\hat T} :=$ list representation of $\hat T$, $L_R :=$ list representation of $R$
+    \FOR {$i=0,\ldots,n-1$}
+        \STATE $j = 0$ \COMMENT{Special case: cherry}
+        \IF {$L[0]$ is last element of $L[0],L[1],L[2]$ in list representation $R$}
+            \STATE Update $\hat T$: swap $L[0]$ and $L[2]$
+        \ELSIF {$L[1]$ is last element of $L[0],L[1],L[2]$ in list representation $R$}
+            \STATE Update $\hat T$: swap $L[1]$ and $L[2]$
+        \ENDIF
+        \STATE $j = 2$
+	    % \IF {$L_R^{-1}[L_{\hat T}[j]] > L_R^{-1}[L_{\hat T}[j+1]]$}
+        \IF {order of $L[j]$ and $L[j+1]$ different in $\hat T$ and $R$}
+			\STATE Update $\hat{T}$: Swap $L[j+1]$ and $L[j]$
+			\STATE $p = p+\hat{T}$
+		\ENDIF
+    \ENDFOR
+	\RETURN $p$
+\end{algorithmic}
+\end{algorithm}
 

### 2019-10-24 paper.tex
section caterpillar convexity

@@
 	\FOR {$i = 1, \dots, n-2$}
-		\STATE Let $C_i$ be the cluster induced by the node with rank $i$ in $R$
+		\STATE Let $C_i$ be the cluster induced by the node with rank $i$ in $R$ \label{alg:find_path:line:_cluster}
 		\WHILE {$\rank(\mrca_{\hat{T}}(C_i))>i$}
@@
 In other words, the set of caterpillar trees is \emph{convex} in $\rnni$.
+In Section~\ref{section:diameter} we will investigate diameter and radius of the $\rnni$ graph, before we discuss a conjecture, the so-called Split Theorem, which was raised in \autocite{Gavryushkin2018-ol}.
+We will give a counterexample to disprove their version of the Split Theorem and claim an alternative version.
 
-This algorithm gives an upper bound for the diameter of $\rnni$, which
-\todo{Why is that?}
-improves
-the one given in \autocite{Gavryushkin2018-ol}.
-\todo{When}
-Later we will see that this bound is tight.
-However, we don't know how good $\findpath$ approximates distances in $\rnni$. \todo{Can we say something about the approximation ratio?}
+Before we get into the details of these geometric properties of $\rnni$, we need to prove some statements that will be used a few times within this geometry section.
+At first we in Lemma~\ref{lemma:diameter_bound} an upper bound of the \emph{diameter} $\Delta(\rnni)$ of the $\rnni$ graph, which is defined as $\Delta(\RNNI) = \max \{d(T, R) \mid T, R \in \rnni\}$.
 
-\todo{make sure trees are consistently denoted by upper case letters}
 \begin{lemma}
-$\Delta(\rnni) \leq \frac{(n-1)(n-2)}{2}$, where $\Delta(\rnni)$ denotes the diameter of the $\rnni$ graph, that is, if $d(T, R)$ is the graph distance in $\rnni$ then $\Delta(\rnni) = \max \{d(T, R) \mid T, R \in \rnni\}$.
+$\Delta(\rnni) \leq \frac{(n-1)(n-2)}{2}$ for $n \geq 3$.
 \label{lemma:diameter_bound}
@@
 \begin{proof}
-Since the diameter is bounded from above the maximum length of a path computed by $\findpath$, it is enough to find this maximum.
-The worst-case execution of $\findpath$ would require each of the $n-2$ clusters $C$ induced by internal nodes of $R$ to be moved by $n - 1 - i$ $\rnni$ operations in $\hat{T}$.
+Since the diameter is bounded from above by the maximum length of a path computed by $\findpath$, it is enough to find this maximum.
+The worst-case execution of $\findpath$ would require each of the $n-2$ clusters $C_i$, considered in Line~\ref{alg:find_path:line:_cluster} of Algorithm~\ref{alg:find_path}, to be moved by $n-1-i$ $\rnni$ operations in $\hat{T}$.
 Hence, the maximum length of such a path is bounded by $\sum\limits_{i = 1}^{n-2} i = \frac{(n-2)(n-1)}{2}$.
@@
 In particular, we are going to prove that one can always find a shortest path between two caterpillar trees that contains caterpillar trees only, meaning that the set of caterpillar trees is convex in $\rnni$.
-We discovered this by computing caterpillar paths using an algorithm to compute shortest paths within the set of caterpillar trees and comparing them with shortest path in $\rnni$.
-This algorithm will be explained in detail later in this section, after demonstrating why finding shortest paths between caterpillar trees builds the basis for understanding the difference between $\nni$ and $\rnni$.
+We discovered this by computing caterpillar paths with $\csort$ and comparing them with shortest path in $\rnni$, computed by the exact algorithm as explained in Section~\ref{section:algorithm}.
+In the following we demonstrate why finding shortest paths between caterpillar trees builds the basis for understanding the difference between $\nni$ and $\rnni$.
 
-We now compare shortest paths between caterpillar trees in $\nni$ with those in $\rnni$.
+Therefore, we compare shortest paths between caterpillar trees in $\nni$ with those in $\rnni$.
 In $\nni$, building a cherry and moving it may give a path that is shorter than a path where the taxa move separately in as it is illustrated out in Figure~\ref{fig:NNI_vs_RNNI}.
@@
 This is due to the fact that $\nni$ moves are only allowed on edges of length one in $\rnni$.
-Therefore, a path where taxa move separately and no second cherry is built, is shortest in $\rnni$.
+Hence a path where taxa move separately and no second cherry is built, is shortest in $\rnni$.
 
@@
 	\includegraphics[width=\textwidth]{NNI_vs_RNNI}
+    \vspace{2pt}
+    \todo{LC: Change caption}
 	\caption{Paths between caterpillar trees $T$ and $R$: black -- shortest path in classical $\nni$ (ignoring ranks) of length $5$; red -- shortest path in $\rnni$ of length $6$; blue -- path in $\rnni$ which is the natural extension of the shortest path in $\nni$ but has length $8$.}
@@
 
-The above observation suggests that there is always a caterpillar path that is shortest path between two arbitrary caterpillar trees in $\rnni$.
+The above observation suggests that there is always a caterpillar path that is a shortest path between any two caterpillar trees in $\rnni$.
 Indeed, we are going to prove Theorem~\ref{thm:caterpillar_convex} within this section, which states that the set of caterpillar trees is convex in $\rnni$.
@@
 Lemma~\ref{lemma:caterpillar_dist=diameter} regards caterpillar trees with distance $\frac{(n-1)(n-2)}{2}$ and will be needed for proving the convexity of the set of caterpillar trees in $\rnni$.
-Notice that we by Lemma~\ref{lemma:diameter_bound} $\frac{(n-1)(n-2)}{2}$ is the upper bound for the diameter of $\rnni$.
+Notice that by Lemma~\ref{lemma:diameter_bound} $\frac{(n-1)(n-2)}{2}$ is the upper bound for the diameter of $\rnni$.
 
@@
     We prove $d(T,R) = \frac{(n-1)(n-2)}{2}$ by induction on the number of taxa $n$ of $T$ and $R$:
-    Without loss of generality we can assume that $T$ is the caterpillar tree $(\ldots ((1,2),3), \ldots, n)$ or $(\ldots ((1,2),3), \ldots, n+1)$ in the induction step.
-    Let $R$ be a caterpillar tree with distance $\frac{(n-1)(n-2)}{2}$ to $T$.
 
     Induction basis: $n=3$\\
-    There are three trees on $3$ taxa that have distance $1$ from another and are caterpillar trees.
-    Hence it is $d(T,R) = d_c(T,R) = 1$ for all pairs of trees $T,R$ on $n=3$ taxa.
+    There exist three trees on three taxa: $[1,2,3], [1,3,2]$ and $[2,3,1]$.
+    \todo{LC: Do we use list representation here? If so, we need to mention it}
+    They all have distance one from another and are caterpillar trees.
+    So it is $d(T,R) = d_c(T,R) = 1$ for all pairs of trees $T,R$ on $n=3$ taxa.
 
     Induction step: $n \to n+1$\\
-    Induction hypothesis: For trees $T, R$ on $n$ taxa it is $d(T, R) = \frac{(n-1)(n-2)}{2}$.\\
-    Let $T, R$ be trees on $n+1$ taxa with caterpillar distance $\frac{n(n-1)}{2}$.
-    Therefore, each of the $\frac{n(n-1)}{2}$ pairs of taxa that is considered when running $\csort$ to compute a path $p$ from $R$ to $T$ exchanges.
-    From this we can conclude that taxon $n+1$, whose parent has highest rank in $T$, is part of the cherry of $R$. \todo{Do we need to explain this further (figure)?}
-    Let us consider the trees $T{\big|}_n$ to $R{\big|}_n$ that result from $T$ and $R$ by deleting taxon $n+1$.
+    Induction hypothesis: For caterpillar trees $T, R$ on $n$ taxa with caterpillar distance $d_c(T,R) = \frac{(n-1)(n-2)}{2}$ it is $d(T,R) = \frac{(n-1)(n-2)}{2}$.\\
+    Without loss of generality we can assume that $T$ is the caterpillar tree $[1,2,3,\ldots,n+1]$.
+    Let $R$ be a caterpillar tree with distance $\frac{n(n-1)}{2}$ to $T$.
+    Then each of the $\frac{n(n-1)}{2}$ pairs of taxa that is considered when running $\csort$ to compute a path $p$ from $R$ to $T$ exchanges.
+    In particular, taxon $n+1$ exchanges with a taxon in each of the $n+1$ loops in $\csort$.
+    \todo{LC: Is this explanation sufficient?}
+    We can conclude that taxon $n+1$, whose parent has highest rank in $T$, is part of the cherry of $R$.
+    Let us now consider the trees $T{\big|}_n$ to $R{\big|}_n$ that result from $T$ and $R$ by deleting taxon $n+1$.
     $\csort$ computes a path from $T{\big|}_n$ to $R{\big|}_n$ that contains the same exchanges of taxa as $p$, except for the $n-1$ ones on $p$ that involve taxon $n+1$.
@@
 
-    According to Lemma~\ref{lemma:distance_delete_taxon}, the distance between $T{\big|}_n$ and $R{\big|}_n$ is at least $d(T{\big|}_n, R{\big|}_n) \leq d(T,R) - |\rank_T(\parent(n+1)) - \rank_R(\parent(n+1))|$.
-    We also know $|\rank_T(\parent(n+1)) - \rank_R(\parent(n+1))| = n-1$.
+    According to Lemma~\ref{lemma:distance_delete_taxon}, the distance between $T{\big|}_n$ and $R{\big|}_n$ is at least $d(T{\big|}_n, R{\big|}_n) \leq d(T,R) - |\rank_T(\parent_T(n+1)) - \rank_R(\parent_R(n+1))|$.
+    We also know $|\rank_T(\parent_T(n+1)) - \rank_R(\parent_R(n+1))| = n-1$.
     Suppose, contrary to our claim, that there is a path from $T$ to $R$ of length less than $\frac{n(n-1)}{2}$.
-    By the above, it follows that $d(T{\big|}_n, R{\big|}_n) \leq d(T,R) - |\rank_T(\parent(n+1)) - \rank_R(\parent(n+1))| < \frac{n(n-1)}{2} - (n-1) < \frac{(n-1)(n-2)}{2}$.
+    By the above, it follows that $d(T{\big|}_n, R{\big|}_n) \leq d(T,R) - |\rank_T(\parent_T(n+1)) - \rank_R(\parent_R(n+1))| < \frac{n(n-1)}{2} - (n-1) < \frac{(n-1)(n-2)}{2}$.
     Since this is a contradiction to $d(T{\big|}_n, R{\big|}_n) = \frac{(n-1)(n-2)}{2}$, it is proven that $d(T,R) = \frac{n(n-1)}{2} $.
@@
     We prove this Theorem by backwards induction on the length $l$ of a shortest caterpillar path between two caterpillar trees $T$ and $R$.
-    Since we use backwards induction, we need to consider the maximum distance between any two caterpillar trees as base case, which is $\frac{(n-1)(n-2)}{2}$ as we saw above.
+    Since we use backwards induction, the maximum distance between any two caterpillar trees, which is $\frac{(n-1)(n-2)}{2}$ (Lemma~\ref{lemma:diameter_bound}), is the base case.
 
@@
     We consider the shortest caterpillar path computed by $\csort$.
-    Since $l < \frac{(n-1)(n-2)}{2}$, there is one loop $k$ when running $\csort$ where less than $n-2-k$ taxon pairs swap positions.
-    We assume that this loop number $k$ is the first one in which this occurs.
+    Since $l < \frac{(n-1)(n-2)}{2}$, there must be one of the $n$ loops in $\csort$ where less than $n-2-k$ taxon pairs swap positions.
+    Let us consider loop number $k$ where this happens for the first time.
     This situation is illustrated in the top of Figure~\ref{fig:cat_convex_seq_example}.
-    Whenever the number of swaps of taxa or, in terms of sequences, neighboured elements in the sequence is maximum in a loop, one of the two first elements swaps position with $n-2-k$ subsequent elements of the sequence.
-    Since we assume that the number of swaps in loop $k$ is not maximal, this does not happen.
+    Whenever the number of swaps of taxa or, in terms of lists, neighboured elements in the list is maximum in a loop, one of the first two elements swaps position with $n-2-k$ subsequent elements of the list.
+    Since we assume that the number of swaps in loop $k$ is not maximal, there are less than $n-2-k$ such moves.
     Let us assume that $a$ is the first element that is compared with its left neighbour in loop $k$ but does not swap with it.
@@
 
+    \todo{LC: replace this figure}
     \begin{figure}[H]

### 2019-10-24 paper.tex
updates until end of geometry section

@@
 However, the Weak Split Theorem remains a conjecture.
-Furthermore, we propose an alternative (even weaker) version of the conjecture, the Cluster Theorem (Section~\ref{section:split_theorem}).
+Furthermore, we propose an alternative (even weaker) version of the conjecture, the Cluster Theorem (Section~\ref{section:cluster_theorem}).
 
@@
 	\STATE $R$ only contains cherry taxa $a, b$ of $T$ and has internal node with rank $n-1$
-    \STATE In list $L$ all taxa $\{1,\ldots,n\}\setminus\{a,b\}$ are ordered such that the ranks of their parents in $T$ do not decrease
+    \STATE In list $L$ all taxa $\{1,\ldots,n\}\setminus\{a,b\}$ are ordered such that the ranks of their parents in $T$ do not decrease \label{alg:mdtree_lineL}
 	\FOR {taxon $a$ in $L$}
-		\STATE Add an internal node as parent of $a$ on an edge in $R$ such that the new internal node has rank one less than the previously lowest internal node
+		\STATE Add an internal node as parent of $a$ on an edge in $R$ such that the new internal node has rank one less than the previously lowest internal node \label{alg:mdtree_lineAddTaxon}
 	\ENDFOR
@@
 \end{algorithm}
+
+An important property of $\mdtree$, which we will use later on in Section~\ref{section:diameter} is the fact that this algorithm is non-deterministic.
+This is because the list computed in Line~\ref{alg:mdtree_lineL} the and the attachment edge of the new taxon in the tree in Line~\ref{alg:mdtree_lineAddTaxon} are not uniquely determined.
+Therefore the output tree $R$ of $\mdtree$ is not unique.
 
@@
     	\centering
-    	\includegraphics[width=\textwidth]{cat_convex_seq_example}
-    	\caption{Loop number $k$ of $\csort$ between $T$ and $R$ as described in proof of Theorem~\ref{thm:caterpillar_convex}.
-        The trees are represented as sequences that are being ordered by $\csort$.
-        Right before loop number $k$, $x_1, \ldots, x_{k-1}$ are already at the same position as in $R$.
+        \begin{tabular}{|c|c|c|c|c|c|c|c|c|c|c|c|}
+            \hline
+            $x_{k+1}$ & $x_k$ & $\ldots$ & $b$ & $a$ & \multicolumn{3}{c|}{$\ldots$} & $x_n$ & $x_{k-1}$ & $\ldots$ & $x_1$\\
+            \hline
+            \multicolumn{12}{c}{}\\
+            \hline
+            $x_{k+1}$ & $x_{k+2}$ & $\ldots$ & $b$ & $x_k$ & $\ldots$ & $a$ & $\ldots$ & $x_n$ & $x_{k-1}$ & $\ldots$ & $x_1$\\
+            \hline
+            \multicolumn{12}{c}{}\\
+            \multicolumn{12}{c}{}\\
+            \hline
+            $x_{k+1}$ & $x_k$ & $\ldots$ & $a$ & $b$ & \multicolumn{3}{c|}{$\ldots$} & $x_n$ & $x_{k-1}$ & $\ldots$ & $x_1$\\
+            \hline
+            \multicolumn{12}{c}{}\\
+            \hline
+            $x_{k+1}$ & $x_{k+2}$ & $\ldots$ & $x_k$ & $b$ & $\ldots$ & $a$ & $\ldots$ & $x_n$ & $x_{k-1}$ & $\ldots$ & $x_1$\\
+            \hline
+        \end{tabular}
+        \vspace{2pt}
+    	\caption{Loop number $k$ of $\csort$ between $T$ and $R$ (in list representation) as described in proof of Theorem~\ref{thm:caterpillar_convex}.
+        The trees are represented as sequences that are being sorted by $\csort$.
+        Right before loop number $k$, taxa $x_1, \ldots, x_{k-1}$ are already at the same position as in $R$.
         At the top: sequences between $T$ and $R$ -- $x_k$ and $b$ swap in loop number $k$, but $x_k$ and $a$ not.
@@
 
-    Since it is clear that $T'$ has caterpillar distance $d_c(T',R) = l+1$ to $T$, we can apply the induction hypothesis on $T'$ and $R$ and see that $d(T',R) = l+1$.
+    Because $T'$ has caterpillar distance $d_c(T',R) = l+1$ to $T$, we can apply the induction hypothesis on $T'$ and $R$ and see that $d(T',R) = l+1$.
     Furthermore, it is $d(T,T') = 1$.
@@
 Furthermore, we are able to show that the \emph{radius} of the $\rnni$ graph, which is defined as $rad(G) =  \min\limits_u \max\limits_v d(u,v)$ for a graph $G$, equals its diameter.
-Let us first consider the diameter of the $\rnni$ graph.
-With the results of Section~\ref{section:caterpillar_convex} is not hard to see that is $\Delta(\rnni) = \frac{(n-1)(n-2))}{2}$ as it is stated in Corollary~\ref{corollary:diameter}.
+At first we consider the diameter of the $\rnni$ graph.
+By Lemma~\ref{lemma:diameter_bound} we already know an upper bound for the diameter.
+We will now show that this is not only a bound, but the exact diameter of $\rnni$.
 
@@
 \begin{proof}
-	Let $T = (( \dots (n,1)2)\dots)n-1)$ and $R = (( \dots (n,n-1)n-2)\dots)1)$.
+	Let $T = [1,n,2,3,\ldots,n-1]$ and $R = [n-1,n,n-2,n-3, \ldtos, 1]$ be two caterpillar trees.
     According to Theorem~\ref{thm:caterpillar_convex}, there is a shortest path between $T$ and $R$ that is a caterpillar path.
     We can compute a shortest caterpillar path by using $\csort$ as explained in Section~\ref{section:caterpillar_convex}.
-    When running this algorithm on $T$ and $R$, each taxon $i$ ($i \in \{1, \ldots, n\}$) swaps $n-1-i$ times with its right neighbour.
+    When running this algorithm on $T$ and $R$, each taxon $i$ ($i \in \{1, \ldots, n\}$) swaps $n-1-i$ times with its upper neighbour.
     Therefore, the length of a shortest caterpillar path is $\sum\limits_{i=1}^{n-1}(n-1-i) = \sum\limits_{i=0}^{n-2}i = \frac{(n-1)(n-2)}{2}$.
@@
 Therefore we use the algorithm $\mdtree$ that has been introduced in section~\ref{section:alg_mdtree}.
-
-However, the results provided in this section are only true for caterpillar trees.
-The pseudocode for $\mdtree$ can be found in Section~\ref{section:alg_mdtree}.
-\todo{figure? + do we need the pseudocode?}
+Note that the results provided in this section are only true if the input tree of $\mdtree$ is a caterpillar tree.
 
@@
     Recall that with Corollary~\ref{corollary:diameter} we know that $\Delta(\rnni) = \frac{(n-1)(n-2)}{2}$.
-    Within this proof we assume, without loss of generality, that it is $T = (( \dots (1,2)3)\dots)n)$
+    Within this proof we assume, without loss of generality, that it is $T = [1,2,3,\ldots,n]$.
     We prove the lemma by induction on the number of taxa $n$.
@@
     Induction basis: $n = 3$\\
-    $\mdtree(T)$ results either in $R = ((1,3),2)$ or $R'= ((2,3),1)$ for $T = ((1,2),3)$.
+    $\mdtree(T)$ results either in $R =[1,3,2]$ or $R'= [2,3,1]$ for $T = [1,2,3]$.
     Since it is $d(T,R) = d(T,R') = 1 = \Delta(\rnni_3)$, the lemma is true for trees on $n=3$ taxa.
@@
     Let us now  assume that $T$ is a tree on $n+1$ taxa.
-    In the list $L$ produced in $\mdtree(T)$ the taxon $n+1$ is the last one, $L[n] = n+1$.
-    Hence $n+1$ is added to $R$ in the last step and therefore, it is in the lowest cherry of $R$ and has parent with rank $1$.
+    In the list $L$ produced Line~\ref{alg:mdtree_lineL} in $\mdtree(T)$ (Algorithm~\ref{alg:max_dist_tree}) taxon $n+1$ is the last one, because its parent, the root, has the highest rank in $T$.
+    This means that this taxon $n+1$ is the last one added last to $R$ as in Line~\ref{alg:mdtree_lineAddTaxon} and therefore, it has parent with rank one in $R$.
     Because the parent of $n+1$ has rank $n$ in $T$, we can conclude from Lemma~\ref{lemma:distance_delete_taxon} that for $T$ and $R$ restricted to taxa $1,\ldots,n$ it holds: $d(T|_n,R|_n) \leq d(T,R) - (n-1)$.
-    It is easy to see that $R|_n$ can be received from $T|_n$ by applying $\mdtree$.
+    It is not hard to see that $R|_n$ can be received from $T|_n$ by applying $\mdtree$:
+    $R|_n$ can be constructed from input tree $T|_n$ in the same way as $R$ can be constructed from $T$, since the list $L$ computed in Line~\ref{alg:mdtree_lineL} of Algorithm~\ref{alf:max_dist_tree} for $T|_n$ is the same as the one for $T$, just the last element is missing.
     So we can follow from the induction hypothesis and Lemma~\ref{lemma:distance_delete_taxon} that it is $d(T,R) \geq d(T|_n,R|_n) + (n-1) = \frac{(n-1)(n-2)}{2} = \frac{n(n-1)}{2}$, which concludes the proof of the lemma.
@@
 
-In $\mdtree$ it is not uniquely determined, where a new taxon is added to the already existing tree.
-Instead, the attachment branch can be randomly chosen with the only restriction being that the rank of the new node needs to be below the ranks of already existing internal nodes.
-Therefore the output tree $R$ of $\mdtree$ is not unique.
-Every tree topology on $n$ taxa can be resulting from $\mdtree$.
-And because permuting labels of trees does not change the distance between them, $\mdtree$ shows that there is for every tree $R$ on $n$ taxa a caterpillar tree $T$ with distance $d(T,R) = \Delta(\rnni) = \frac{(n-1)(n-2)}{2}$.
+In Section~\ref{alg:max_dist_tree} we already discussed that $\mdtree$ is a non-deterministic algorithm.
+In fact, every tree topology on $n$ taxa can be resulting from $\mdtree$, since there is no restriction on where to add the new taxon in the already existing tree in Line~\ref{alg:mdtree_lineAddTaxon} of Algorithm~\ref{ald:max_dist_tree}.
+This means that for each caterpillar input tree $\mdtree$ can compute a tree with any tree topology that has maximum distance from the input tree.
+\todo{LC: Definition tree topology?}
+And because permuting labels of trees does not change the distance between them, there is for every tree $R$ on $n$ taxa a caterpillar tree $T$ with distance $d(T,R) = \Delta(\rnni) = \frac{(n-1)(n-2)}{2}$.
 This gives us Corollary~\ref{corollary:radius}.
@@
 
-\subsection{Split Theorem}
-\label{section:split_theorem}
+\subsection{Cluster Theorem}
+\label{section:cluster_theorem}
 \todo{Split or Cluster?}

### 2019-10-26 paper.tex
edits in cluster thm section, some todos

@@
 In the proof presented in \autocite{Li1996-zw} for Theorem~\ref{thm:split_nni} an example of two trees that share a cluster but have no shortest path that preserves this cluster is given.
+In Figure~\ref{fig:NNI_NP_proof} there is an illustration of this example.
 The reason that this proof does not work for $\rnni$ is similar to the reason that caterpillar trees are convex in $\rnni$, but not in $\nni$.
@@
 Afterwards, $T'$ is transformed to a tree $R'$ still containing the same cherries as $T'$ but in a way that resolving all these cherries of $R'$ results in $R$.
-For proving that this path is shorter than any path preserving the clusters $\{1, \ldots, n\}$ and $\{n+1, \ldots, 2n\}$, the fact that $d_{\nni}(R,R') = d_{\nni}(T{\big|}_{\{1, \ldots, n\}}, R{\big|}_{\{1, \ldots, n\}})$ where $T{\big|}_{\{1, \ldots, n\}}$ and $R{\big|}_{\{1, \ldots, n\}}$ are $T$ and $R$ restricted to taxa $\{1,\ldots,n\}$, respectively, is being used.
+For proving that this path is shorter than any path preserving the clusters $\{1, \ldots, n\}$ and $\{n+1, \ldots, 2n\}$, \autocite{Li1996-zw} use the fact that $d_{\nni}(R,R') = d_{\nni}(T{\big|}_{\{1, \ldots, n\}}, R{\big|}_{\{1, \ldots, n\}})$ where $T{\big|}_{\{1, \ldots, n\}}$ and $R{\big|}_{\{1, \ldots, n\}}$ are $T$ and $R$ restricted to taxa $\{1,\ldots,n\}$, respectively.
 However, this does not work in $\rnni$.
@@
 Therefore, a path where the two caterpillar trees are sorted separately, is shorter in $\rnni$.
-\todo{(1) Do we want this explanation?
-(2) If yes, is this explanation sufficient (we never say how we assign ranks to $T$ and $R$)?}
 
-% \todo{Shall we keep this explanation? The problem is that the $\rnni$ path depends on the assignment of ranks to the cherries produced on the path}
-% The counterexample presented in \autocite{Li1996-zw} works in $\nni$ because of the following:
-% When proving Theorem~\ref{thm:split_nni}, \autocite{Li1996-zw} gives an example of two trees $T$ and $R$ on $2n$ taxa that consist of two caterpillar trees each that share the root as parent.
-% The clusters given by the two nodes adjacent to the root are $\{1, \ldots, n\}$ and $\{n+1, \ldots, 2n\}$ in both trees $T$ and $R$.
-% Between these there is a shortest path in $\nni$ where at first $T$ is transformed to a tree $T'$ with $n$ cherries where each cherry contains one taxon of $\{1, \ldots, n\}$ and one taxon of $\{n+1, \ldots, 2n\}$.
-% Afterwards, $T'$ is transformed to a tree $R'$ still containing the same cherries as $T'$ but in a way that resolving all these cherries of $R'$ results in $R$.
-% For proving that this path is shorter than any path preserving the clusters $\{1, \ldots, n\}$ and $\{n+1, \ldots, 2n\}$, the fact that $d_{\nni}(R,R') = d_{\nni}(T{\big|}_{\{1, \ldots, n\}}, R{\big|}_{\{1, \ldots, n\}})$ where $T{\big|}_{\{1, \ldots, n\}}$ and $R{\big|}_{\{1, \ldots, n\}}$ are $T$ and $R$ restricted to taxa $\{1,\ldots,n\}$, respectively, is being used.
-% However, this does not work in $\rnni$.
-% Introducing new cherries means introducing new internal nodes with new ranks.
-% Since there are lots of rank moves needed for an $\rnni$ path similar to the shortest $\nni$ path presented above, the path in $\rnni$ that breaks the clusters $\{1, \ldots, n\}$ and $\{n+1, \ldots, 2n\}$ is no shortest path.
+\begin{figure}[H]
+	\centering
+	\includegraphics[width=\textwidth]{NNI_NP_proof}
+    \vspace{2pt}
+    \caption{Example of a shortest path between two trees $T$ and $R$ in $\nni$.
+    All leaves labelled with dots are labelled by taxa in $\{1, \ldots, n$, squares represent taxa in $\{n+1, \ldots, 2n\}$.
+    In \autocite{Li1996-zw} it is proven that there is such a labelling that ensures that there is no shortest path in $\nni$ where the clusters shared between $T$ and $R$ are preserved.}
+	\label{fig:NNI_NP_proof}
+\end{figure}
 
-% ^How do we assign ranks to the new cherry nodes?
-% We could either create a path where $R$ and $R'$ have distance equal to $d_{\rnni}(T{\big|}_{\{1, \ldots, n\}}, R{\big|}_{\{n+1, \ldots, 2n\}})$ and have a lot of additional rank moves before and after that part OR
-% Build $R$ and $R'$ that contain all the cherries within the least number of moves possible (3n-3) and having a longer path between $R$ and $R'$.
-% Also anything in between these two options is possible. Are we able to disprove every option??
-
-Since the example given in \autocite{Li1996-zw} that proves Theorem~\ref{thm:split_nni} in $\nni$ does not work in $\rnni$, a contrary statement was conjectured in \autocite{Gavryushkin2018-ol}.
-For understanding this conjecture it is important to realise the connection between edges in a tree and so-called splits, which are bipartitions of the taxon set of that tree:
-By deleting an edge $e$ of a tree $T$ one receives two trees on taxon sets $A$ and $B$ such that $\{A,B\}$ is a partition of the taxon set of $T$.
+Since the example given in \autocite{Li1996-zw}, which proves Theorem~\ref{thm:split_nni} in $\nni$, does not work in $\rnni$, a contrary statement was conjectured in \autocite{Gavryushkin2018-ol}.
+For understanding this conjecture it is important to realise the connection between edges in a tree and so-called \emph{splits}, which are bipartitions of the taxon set of a tree:
+By deleting an edge $e$ of a tree $T$ one receives two graphs with labels in $A$ and $B$, respectively, such that $\{A,B\}$ is a partition of the taxon set of $T$.
 Note that each edge induces such a bipartition, often written as $A|B$, and that both edges incident to the root of a tree induce the same partition.
-Since these partitions are commonly referred to as \emph{splits}, the following conjecture was named \emph{Split Theorem} in \autocite{Gavryushkin2018-ol}.
+Since these partitions are commonly referred to as splits, the following conjecture was named \emph{Split Theorem} in \autocite{Gavryushkin2018-ol}.
 
@@
 	\includegraphics[width=\textwidth]{splitthm_counterexample}
-	\caption{The split $123|45$ is present in $T$ and $R$, but the path at the top is a shortest path where none of the trees contains this split.
+    \vspace{2pt}
+	\caption{The split $123|45$ is present in $T$ and $R$, but the path at the top is a shortest path (computed by $\csort$) where none of the trees contains this split.
     However, this split is maintained on the path at the bottom, which is a shortest path as well.}
@@
 Moreover, a ranked tree can be uniquely determined by ordering the clusters such that the ranks of the internal nodes inducing those clusters increase.
-
-% \begin{conjecture}[Weak Split Theorem]
-% 	For $\rnni$ graph the following statement holds:
-% 	if a partition of leaves given by an edge is present in two trees $T$ and $R$ then there exists a shortest path between $T$ and $R$ where this partition is present in every tree on this path.
-% 	\label{split_theorem_weak}
-% \end{conjecture}
-%
-% \todo{Do we want to keep the weak version of the split thm? It is stronger than the Cluster Thm!}
-%
-% The weak Split Theorem (in the version as in Lena's thesis) is computationally proven to hold for trees with up to $6$ taxa ($\sim 38$ min running time on laptop).
+We already used this representation of trees in Section~\ref{section:algorithms}.
 
@@
 
-Note that trees $T$ and $R$ of counterexample~\ref{fig:splitthm_counterexample} induce the same set of splits, but have distinct clusters.
+Note that trees $T$ and $R$ in Figure~\ref{fig:splitthm_counterexample} induce the same set of splits, but have distinct clusters.
 Furthermore, this counterexample can be used to prove that the Cluster Theorem does not hold in $\nni$.
@@
 
-For proving the Cluster Theorem as stated in Conjecture~\ref{conjecture:cluster_theorem} for small trees, we implemented the algorithm for computing the $\rnni$ graph as stated above.
-If the graph is computed, one can get distances by using algorithms such as Floyd-Warshall~\autocite{Floyd1962-ew} or Dijkstra~\autocite{Dijkstra1959-ph}.
+\todo{LC: re-read the two following paragraphs and move them to correct position in paper (maybe some stuff should go to alg section?)}
+
+For providing further evidence that the Cluster Theorem holds in $\rnni$, we can show computationally that it holds for small trees with up to seven taxa.
+\todo{LC: We still need to check 7 taxa.}
+We are doing this in the following way.
+\todo{LC: Move the following paragraph to algorithms section?}
+At first we compute the $\rnni$ graph by using the algorithm introduced in \autocite{Gavryushkin2018-ol}, which we already discussed in Section~\ref{section:algorithms}.
+If the graph is computed, one can get distances by using algorithms such as Floyd-Warshall \autocite{Floyd1962-ew} or Dijkstra\autocite{Dijkstra1959-ph}.
 However, the large number of vertices in $\rnni$ is an obstacle when it comes to computing distances.
 For instance the Floyd-Warshall algorithm needs to save a matrix containing all pairwise distances of vertices in the graph, which needs space quadratic in the number of vertices.
-Since there are $\frac{n!(n-1)!}{2^{n-1}}$ vertices in the $\rnni$ graph~\autocite{Gavryushkin2018-ol}, we already run out of space when considering $7$ taxa.
+Since there are $\frac{n!(n-1)!}{2^{n-1}}$ vertices in the $\rnni$ graph \autocite{Gavryushkin2018-ol}, we already run out of space when considering seven taxa.
 Therefore, we cannot simply use this algorithm to compute all distances in $\rnni$.
@@
 Furthermore, it is sufficient to consider clusters of the shape $\{1, \ldots, m\}$ for $2 \leq m \leq n-1$.
-This is due to the fact that the distance $d(T_1,R_2)$ and $d(T_2,R_2)$ between two pairs of trees $(T_1,R_1)$ and $(T_2,R_2)$ are equal if one can receive $T_2$ from $T_1$ by permuting the taxa the same way as it is necessary for receiving $R_2$ from $R_1$.
-Therefore, we compute for all $m = 2, \ldots, n-1$ exact $\rnni$ distances between all trees containing the cluster $\{1, \ldots, m\}$, using Dijkstra.
-Additionally, the subgraph of $\rnni$ only consisting of trees that have $\{1, \ldots, m\}$ as cluster is computed.
+This is due to the fact that the distance $d(T_1,R_1)$ and $d(T_2,R_2)$ between two pairs of trees $(T_1,R_1)$ and $(T_2,R_2)$ are equal if one can receive $T_2$ from $T_1$ by permuting the taxa the same way as it is necessary for receiving $R_2$ from $R_1$.
+Therefore, we compute for all $m = 2, \ldots, n-1$ exact $\rnni$ distances between all trees containing the cluster $\{1, \ldots, m\}$, using Dijkstra's algorithm \autocite{Dijkstra1959-ph}.
+Additionally, the subgraph of $\rnni$ induced by the subset of trees that have $\{1, \ldots, m\}$ as cluster is computed.
+\todo{LC: Do we need to define induced subgraph?}
 All pairwise distances between the trees in the subgraph are computed, by using Floyd-Warshall, to compare them with the distances in $\rnni$.
-With this approach it is possible to show that the Cluster Theorem holds for up to $6$ taxa. \todo{Can we improve this?}
+With this approach it is possible to show that the Cluster Theorem holds for up to seven taxa. \todo{Can we improve this?}
 
@@
 	\includegraphics[width=0.7\textwidth]{findpath_not_symmetric}
+    \vspace{2pt}
 	\caption{Two paths computed by $\findpath$. At the top $\findpath(T,R)$, at the bottom $\findpath(R,T)$}

### 2019-10-27 paper.tex
proof lemma 1

@@
 	\FOR {$i = 1, \dots, n-2$}
-		\STATE Let $C_i$ be the cluster induced by the node with rank $i$ in $R$ \label{alg:find_path:line:_cluster}
-		\WHILE {$\rank(\mrca_{\hat{T}}(C_i))>i$}
-			\STATE Update $\hat{T}$: Decrease $\rank(\mrca_{\hat{T}}(C_i))$ by an $\rnni$ move \label{alg:line:move_set_down}
+		\STATE Let $C_i$ be the cluster induced by the node with rank $i$ in $R$ \label{alg:find_path:line:cluster}
+		\WHILE {$\rank_{\hat T}(\mrca_{\hat{T}}(C_i))>i$}
+			\STATE Update $\hat{T}$: Decrease $\rank_{\hat T}(\mrca_{\hat{T}}(C_i))$ by an $\rnni$ move \label{alg:findpath:line:move_set_down}
 			\STATE $p = p+\hat{T}$
@@
 
-%BEGIN: From MareikesVisit
-
 \todo{LC: Work on the following proof! Is it worth introducing intervals for $\rnni$ moves for the proof of the following lemma?}
@@
 \begin{lemma}
-    Let $T$ be a tree and $C \subset \{1, \ldots, n\}$ a subset of taxa such that $C$ is not induced by any node in $T$.
-    There is exactly one tree $T'$ that can be obtained from $T$ by one $\rnni$ move such that $\rank(\mrca_{T'}(C)) = \rank(\mrca_{T}(C)) -1 $.
+    Let $T$ and $R$ be two trees on which we apply $\findpath$ to compute a path from $T$ to $R$.
+    We assume that $\hat T \neq T,R$ is a tree on the path computed by this algorithm, which needs to be updated in Line~\ref{alg:findpath:line:move_set_down} of Algorithm~\ref{alg:find_path}.
+    Then there is always exactly one $\rnni$ move that can update $\hat T$ to a tree $T'$ such that $\mrca_{T'}(C_i) < \mrca_{\hat T}(C_i)$ for cluster $C_i$ as defined in Line~\ref{alg:find_path:line:cluster}.
     \label{lemma:mrca_move}
@@
 
-    Let $(C_1, \ldots, C_n)$ be the list representation of $T$ as introduced above.
-    \todo{LC: Introduce this representation above!}
-    Let $v:=mrca_T(C)$, let $C_{i+1}$ be the cluster induced by $v$, and let $w$ be the node inducing $C_i$.
-    We will distinguish the case that there is an edge $(v,w)$ in $T$ from the case that there is no such edge.
+    Let $T$ and $R$ be trees and $[C_1, \ldots, C_n]$ the cluster representation
+    \todo{LC: How do we call this representation? Introduce it in the beginning of alg chapter!}
+    of $R$ as introduced above.
+    Let $\hat T$ be a tree as in Line~\ref{alg:findpath:line:move_set_down} of Algorithm~\ref{alg:find_path} and $C_i$ the cluster defined in Line~\ref{alg:find_path:line:cluster}.
+    We want to prove that there is exactly one $\rnni$ move possible on $\hat T$ that results in a tree $T'$ where $\mrca_{T'}(C_i) < \mrca_{\hat T}(C_i)$ if $C_i$ is not induced by a node in $\hat T$ already.
 
-    At first we assume that there is no edge $(v,w)$, which means $C_i \not \subset C_{i+1}$.
-    In this case the two sets $C_i, C_{i+1}$ can exchange by a rank swap.
-    Since $C \subset C_{i+1}$, this rank swap decreases the rank of $\mrca(C)$ by one and there is no $\nni$ move that could move $\mrca(C)$ down.
+    To do this, we distinguish the following two cases:
+    Either there is an edge between the node $v = \mrca_{\hat T}(C_i)$ and the node $w$ with rank $rank_{\hat T}(w) = rank_{hat T}(v) - 1$, or there is no such edge.
 
-    Let us now consider the case that $(v,w)$ is an edge in $T$.
-    The only move that could decrease the rank of $\mrca(C)$ is an $\nni$ move on the edge $(v,w)$.
-    Let $C_k$ and $C_j$ be the clusters induced by the children of $C_i$.
-    Then it is $C_k \cup C_j = C_i$.
-    There are two different $\nni$ moves possible on $(v,w)$.
-    $C_i$ can either be replaced by $C_j \cup (C_{i+1} \setminus C_i)$ or by $C_k \cup (C_{i+1} \setminus C_i)$.
-    To prove that there is always exactly one $\rnni$ moves that decreases the rank of $\mrca(C)$, we need to show that it is either $C \subset (C_j \cup (C_{i+1} \setminus C_i))$ or $C \subset (C_k \cup (C_{i+1} \setminus C_i))$.
+    In the latter case, there is only one $\rnni$ move that could increase the rank of $v$.
+    That is a rank move swapping the ranks of $v$ and $w$.
 
-    Let $a \in C$.
-    Then it is $a \in C_{i+1} = mrca(C)$ and it follows $a \in C_{i+1} \setminus C_i$ or $a \in C_i$.
-    If $a \in C_{i+1} \setminus C_i$, $a$ is in the new set after the $\nni$ move.
-    So now we will assume that $a \in C_i$.
-    Then it is either $a \in C_j$ or $a \in C_k$.
-    Without loss of generality we assume that $a \in C_j$.
+    If there is an edge $(v,w)$ in $\hat T$, there are two $\nni$ moves possible on this edge, which could possibly change the rank of the most recent common ancestor of $C_i$.
+    We are now going to explain why only one of these two $\nni$ moves changes the rank of this most recent common ancestor.
+    Therefore it is important to see that the clusters $C_1, \ldots, C_{i-1}$ already appear in $\hat T$ as they do in $R$, which is a direct result from the algorithm.
+    It is also straight forward to see
+    \todo{LC: is it?}
+    that the $C_i$ either is the disjoint union $C_i = C_j \dot\cup C_k$ for $j,k < i$ or $C_i = C_j \dot\cup \{x\}$ for a taxon $x \in X \setminus (C_1 \cup \ldots \cup C_{i-1})$.
 
-    We need to prove that if $a \in C_j$, there is no $b \in C$ with $b \in C_k$.
-    This makes sure that it is either $C \subset (C_j \cup (C_{i+1} \setminus C_i))$ or $C \subset (C_k \cup (C_{i+1} \setminus C_i))$.
-    For contradiction we assume that $a \in C_j$ and $C \in S_k$.
-    We know that all sets that are subsets of $C$ already are at the same position in the current tree as in the destination tree $R$.
-    Therefore, we can conclude from $a \in C_j$ and $b \in C_k$ that $C_j \cup C_k = C$.
-    Hence, it holds $mrca(C) = C_i$.
-    This is a contradiction to our assumption that $C_{i+1} = mrca(C)$ in the current tree.
-    Therefore, it holds either $C \subset (C_j \cup (C_{i+1} \setminus C_i))$ or $C \subset (C_k \cup (C_{i+1} \setminus C_i))$.
+    Le us at first assume that $C_i = C_j \dot\cup C_k$ for some $j,k < i$.
+    The tree $\hat T$ looks like the tree depicted in Figure~\ref{fig:mrca_move} where $T_w^1$ and $T_w^2$ are the two subtrees rooted in the children of $w$ and $T_v$ is the subtree whose root is child of $v$ and does not include $w$.
+    Because it is $v = \mrca_{\hat T}(C_i)$, either $C_j$ or $C_k$ is a subset of the taxon set of $T_v$.
+    Without loss of generality we assume that it is $C_j$.
+    Since $C_k$ is already induced by a node in $\hat T$, all taxa in $C_k$ are either in $T_w^1$ or in $T_w^2$.
+    If there were taxa of $C_k$ in both subtrees $T_w^1$ and $T_w^2$, $w$ would be the node inducing $C_k$ and hence it would be $k = i-1$, which means that $v$ induces $C_i$.
+    Therefore, we know that $C_j$ is a subset of the taxon set of $T_v$ and $C_k$ is subset of the taxon set of one of the subtrees $T_w^1$ or $T_w^2$.
+    In Figure~\ref{fig:mrca_move} one can see the two moves possible on $(v,w)$.
+    Depending on which of the subtrees $T_w^1$ or $T_w^2$ includes the taxa of $C_k$, one can see that only one of the two $\nni$ moves results in a tree where the tank of the most recent common ancestor of $C_i$ is below its rank in $\hat T$.
 
-    Concluding, there is always an $\rnni$ move on the edge $(v,w)$ for $mrca(C) = C_{i+1}, C \subsetneq C_{i+1}$ that decreases the rank of $mrca(C)$.
+    Now we consider the case that $C_i = C_j \dot\cup \{x\}$ for a taxon $x \in X \setminus (C_1 \cup \ldots \cup C_{i-1})$.
+    For the same reason as above, $C_j$ is in subset of the taxon set of one of the subtrees $T_w^1, T_w^2$ or $T_v$ of $\hat T$ as depicted in Figure~\ref{fig:mrca_move}.
+    Similarly $x$ is taxon of one of these three subtrees in a way that either $C_j$ or $x$ is included in $T_v$, because $v = \mrca_{\hat T}(C_i)$.
+    So there is just one of the two subtrees $T_w^1$ and $T_w^2$ that neither includes $C_j$ nor $x$.
+    Therefore, the $\nni$ move that moves this subtree to the position where $T_v$ has been before, is the only $\nni$ move that decreases the rank of $\mrca_{\hat T}(C_i)$.
+
+    We can conclude that there is in any case exactly one $\rnni$ move that can be performed on $\hat T$ that decreases the rank of $\mrca_{\hat T}(C_i)$ in Line~\ref{alg:findpath:line:move_set_down} of Algorithm~\ref{alg:find_path}.
+
+    \begin{figure}[H]
+    	\centering
+    	\includegraphics[width=0.5\textwidth]{mrca_move}
+        \vspace{2pt}
+    	\caption{\todo{caption}}
+    	\label{fig:mrca_move}
+    \end{figure}
+
+    % Let $v:=\mrca_{\hat T}(C_i)$, let $C_{i+1}$ be the cluster induced by $v$, and let $w$ be the node inducing $C_i$.
+    % We will distinguish the case that there is an edge $(v,w)$ in $T$ from the case that there is no such edge.
+    %
+    % At first we assume that there is no edge $(v,w)$, which means $C_i \not \subset C_{i+1}$.
+    % In this case the two sets $C_i, C_{i+1}$ can exchange by a rank swap.
+    % Since $C \subset C_{i+1}$, this rank swap decreases the rank of $\mrca(C)$ by one and there is no $\nni$ move that could move $\mrca(C)$ down.
+    %
+    % Let us now consider the case that $(v,w)$ is an edge in $T$.
+    % The only move that could decrease the rank of $\mrca(C)$ is an $\nni$ move on the edge $(v,w)$.
+    % Let $C_k$ and $C_j$ be the clusters induced by the children of $C_i$.
+    % Then it is $C_k \cup C_j = C_i$.
+    % There are two different $\nni$ moves possible on $(v,w)$.
+    % $C_i$ can either be replaced by $C_j \cup (C_{i+1} \setminus C_i)$ or by $C_k \cup (C_{i+1} \setminus C_i)$.
+    % To prove that there is always exactly one $\rnni$ moves that decreases the rank of $\mrca(C)$, we need to show that it is either $C \subset (C_j \cup (C_{i+1} \setminus C_i))$ or $C \subset (C_k \cup (C_{i+1} \setminus C_i))$.
+    %
+    % Let $a \in C$.
+    % Then it is $a \in C_{i+1} = mrca(C)$ and it follows $a \in C_{i+1} \setminus C_i$ or $a \in C_i$.
+    % If $a \in C_{i+1} \setminus C_i$, $a$ is in the new set after the $\nni$ move.
+    % So now we will assume that $a \in C_i$.
+    % Then it is either $a \in C_j$ or $a \in C_k$.
+    % Without loss of generality we assume that $a \in C_j$.
+    %
+    % We need to prove that if $a \in C_j$, there is no $b \in C$ with $b \in C_k$.
+    % This makes sure that it is either $C \subset (C_j \cup (C_{i+1} \setminus C_i))$ or $C \subset (C_k \cup (C_{i+1} \setminus C_i))$.
+    % For contradiction we assume that $a \in C_j$ and $C \in S_k$.
+    % We know that all sets that are subsets of $C$ already are at the same position in the current tree as in the destination tree $R$.
+    % Therefore, we can conclude from $a \in C_j$ and $b \in C_k$ that $C_j \cup C_k = C$.
+    % Hence, it holds $mrca(C) = C_i$.
+    % This is a contradiction to our assumption that $C_{i+1} = mrca(C)$ in the current tree.
+    % Therefore, it holds either $C \subset (C_j \cup (C_{i+1} \setminus C_i))$ or $C \subset (C_k \cup (C_{i+1} \setminus C_i))$.
+    %
+    % Concluding, there is always an $\rnni$ move on the edge $(v,w)$ for $mrca(C) = C_{i+1}, C \subsetneq C_{i+1}$ that decreases the rank of $mrca(C)$.
 

### 2019-10-27 paper.tex
proof lemma3 + some typos

@@
 Before we get into the details of these geometric properties of $\rnni$, we need to prove some statements that will be used a few times within this geometry section.
-At first we in Lemma~\ref{lemma:diameter_bound} an upper bound of the \emph{diameter} $\Delta(\rnni)$ of the $\rnni$ graph, which is defined as $\Delta(\RNNI) = \max \{d(T, R) \mid T, R \in \rnni\}$.
+At first we in Lemma~\ref{lemma:diameter_bound} an upper bound of the \emph{diameter} $\Delta(\rnni)$ of the $\rnni$ graph, which is defined as $\Delta(\rnni) = \max \{d(T, R) \mid T, R \in \rnni\}$.
 
@@
 Since the diameter is bounded from above by the maximum length of a path computed by $\findpath$, it is enough to find this maximum.
-The worst-case execution of $\findpath$ would require each of the $n-2$ clusters $C_i$, considered in Line~\ref{alg:find_path:line:_cluster} of Algorithm~\ref{alg:find_path}, to be moved by $n-1-i$ $\rnni$ operations in $\hat{T}$.
+The worst-case execution of $\findpath$ would require each of the $n-2$ clusters $C_i$, considered in Line~\ref{alg:find_path:line:cluster} of Algorithm~\ref{alg:find_path}, to be moved by $n-1-i$ $\rnni$ operations in $\hat{T}$.
 Hence, the maximum length of such a path is bounded by $\sum\limits_{i = 1}^{n-2} i = \frac{(n-2)(n-1)}{2}$.
@@
 \begin{proof}
-First observe that the rank of an internal node $v$ of a tree $T$ along an $\rnni$ path $p$ can only be changed by performing a rank move that involves $v$ or an $\nni$ move on an edge adjacent to $v$.
-Second observe that any $\rnni$ move can change the rank of an internal node by at most one.
+First observe that the rank of the internal node $\parent_T(n+1)$ can only be changed by performing a rank move that involves $\parent_T(n+1)$ or an $\nni$ move on an edge adjacent to $\parent_T(n+1)$.
+Second observe that any $\rnni$ move can change the rank of $\parent_T(n+1)$ by at most one.
+So we can follow that the rank of $\parent_T(n+1)$ can be decreased by at most one by an $\rnni$ move on $T$.
 
@@
 This new path $p{\big|}_n$ is then a path from $T{\big|}_n$ to $R{\big|}_n$.
+The reason for this is that the only moves that are deleted from $p$ to receive $p{\big|}_n$ are those involving the parent of taxon $n+1$.
+All other moves stay the same and can be performed equally on a path from $T{\big|}_n$ to $R{\big|}_n$ as on a path from $T$ to $R$, because the relations of all other internal nodes are the same between both pairs of trees.
 Recall that $\delta$ is the difference in ranks between the parents of taxon $n+1$ in $T$ and $R$.
-\todo{This statement has to be checked very carefully and one or two sentences added to explain the `applying'.}
-Applying the two observations above we conclude that $|p{\big|}_n| \leq |p| - \delta$.
+With the knowledge that an $\rnni$ move on $T$ can change the rank of $\parent_T(n+1)$ by at most one, we can conclude that $|p{\big|}_n| \leq |p| - \delta$.
 Since $d(T{\big|}_n,R{\big|}_n) \leq |p{\big|}_n|$ and $|p| = d(T,R)$, the desired inequality follows.
@@
 In particular, we are going to prove that one can always find a shortest path between two caterpillar trees that contains caterpillar trees only, meaning that the set of caterpillar trees is convex in $\rnni$.
-We discovered this by computing caterpillar paths with $\csort$ and comparing them with shortest path in $\rnni$, computed by the exact algorithm as explained in Section~\ref{section:algorithm}.
+We discovered this by computing caterpillar paths with $\csort$ and comparing them with shortest path in $\rnni$, computed by the exact algorithm as explained in Section~\ref{section:algorithms}.
 In the following we demonstrate why finding shortest paths between caterpillar trees builds the basis for understanding the difference between $\nni$ and $\rnni$.
@@
 \begin{proof}
-	Let $T = [1,n,2,3,\ldots,n-1]$ and $R = [n-1,n,n-2,n-3, \ldtos, 1]$ be two caterpillar trees.
+	Let $T = [1,n,2,3,\ldots,n-1]$ and $R = [n-1,n,n-2,n-3, \ldots, 1]$ be two caterpillar trees.
     According to Theorem~\ref{thm:caterpillar_convex}, there is a shortest path between $T$ and $R$ that is a caterpillar path.
@@
     It is not hard to see that $R|_n$ can be received from $T|_n$ by applying $\mdtree$:
-    $R|_n$ can be constructed from input tree $T|_n$ in the same way as $R$ can be constructed from $T$, since the list $L$ computed in Line~\ref{alg:mdtree_lineL} of Algorithm~\ref{alf:max_dist_tree} for $T|_n$ is the same as the one for $T$, just the last element is missing.
+    $R|_n$ can be constructed from input tree $T|_n$ in the same way as $R$ can be constructed from $T$, since the list $L$ computed in Line~\ref{alg:mdtree_lineL} of Algorithm~\ref{alg:max_dist_tree} for $T|_n$ is the same as the one for $T$, just the last element is missing.
     So we can follow from the induction hypothesis and Lemma~\ref{lemma:distance_delete_taxon} that it is $d(T,R) \geq d(T|_n,R|_n) + (n-1) = \frac{(n-1)(n-2)}{2} = \frac{n(n-1)}{2}$, which concludes the proof of the lemma.
@@
 In Section~\ref{alg:max_dist_tree} we already discussed that $\mdtree$ is a non-deterministic algorithm.
-In fact, every tree topology on $n$ taxa can be resulting from $\mdtree$, since there is no restriction on where to add the new taxon in the already existing tree in Line~\ref{alg:mdtree_lineAddTaxon} of Algorithm~\ref{ald:max_dist_tree}.
+In fact, every tree topology on $n$ taxa can be resulting from $\mdtree$, since there is no restriction on where to add the new taxon in the already existing tree in Line~\ref{alg:mdtree_lineAddTaxon} of Algorithm~\ref{alg:max_dist_tree}.
 This means that for each caterpillar input tree $\mdtree$ can compute a tree with any tree topology that has maximum distance from the input tree.

### 2019-10-28 paper.tex
alg section

@@
 
-A \emph{ranked phylogenetic tree} is a pair consisting of a rooted binary phylogenetic tree $T$ on the set $X = \{1, \ldots, n\}$ of \emph{taxa} for $n \in \mathbb N$, and a (total) rank function $\rank$ that maps all elements of $X$ to $0$, all internal nodes of $T$ onto the set $\{1, \ldots, n-1\}$, and respects the partial order given by the tree.
+A \emph{ranked phylogenetic tree} is a pair consisting of a rooted binary phylogenetic tree $T$ on the set $X = \{1, \ldots, n\}$ of \emph{taxa} for $n \in \mathbb N$, and a (total) rank function $\rank$ that maps all leaves of $T$ to $0$, all internal nodes of $T$ onto the set $\{1, \ldots, n-1\}$, and respects the partial order given by the tree.
 The latter means that if $u$ and $v$ are two internal nodes of $T$ such that there exists a path from a taxon $x \in X$ to the root which first passes through $u$ and then through $v$ then $\rank(u) < \rank(v)$.
@@
 
-Our definition of a ranked tree implies a natural notion of the edge length.
-Indeed, a tree edge $(u,v)$ of $T$ has length $l$ if $|\rank_T(v) - \rank_T(u)| = l$.
+Our definition of a ranked tree implies a natural notion of the edge length.:
+A tree edge $(u,v)$ of $T$ has length $l$ if $|\rank_T(v) - \rank_T(u)| = l$.
 
@@
 
-We use the notation $d(T, R)$ throughout the paper to denote the graph distance between trees in $\rnni$, that is, $d(T, R)$ is the length of a shortest $\rnni$ path between $T$ and $R$.
+We use the notation $d(T, R)$ throughout the paper to denote the graph distance between trees in $\rnni$, that is, $d(T, R)$ is the length of a shortest $\rnni$ path between tree $T$ and $R$.
 
@@
     \vspace{2pt}
-	\caption{Two types of operation that define edges of the $\rnni$ graph.}
+	\caption{Two types of operation that define edges of the $\rnni$ graph.
+    The move at the top left is a rank move while the trees in the triangle are connected by $\nni$ moves.}
 	\label{fig:RNNI}
@@
 
-Within this section we will first introduce two algorithms, namely $\findpath$ (Section~\ref{section:alg_findpath}) and $\csort$ (Section~\ref{section:alg_csort}), for computing paths between two given trees in $\rnni$.
+Within this section we will at first described how we can compute exact distances between trees in the $\rnni$ graph (Section~\ref{section:alg_RNNI_graph}) by using results of \autocite{Gavryushkin2018-ol}.
+Afterwards, we introduce two algorithms, namely $\findpath$ (Section~\ref{section:alg_findpath}) and $\csort$ (Section~\ref{section:alg_csort}), for computing paths between two given trees in $\rnni$.
 The first one takes arbitrary trees as input while the latter one is restricted to trees with a certain tree shape.
@@
 Additionally we will present the algorithm $\mdtree$, which produces a tree that is far away from the given input tree.
-What we mean by far away depends on the input tree and will be discussed in Section~\ref{section:alg_mdtree}.
+What we mean by far away depends on the input tree and will be discussed after introducing this algorithm in Section~\ref{section:alg_mdtree}.
+
+\subsection{Computing the $\rnni$ graph}
+\label{section:alg_RNNI_graph}
+
+In \autocite{Gavryushkin2018-ol} an exact algorithm for computing the $\rnni$ graph has been introduced.
+We implemented this algorithm and used it for computing this graph for trees on up to $7$ taxa.
+\todo{LC: $7$ or $8$?}
+\todo{Extend this section}
 
@@
 For the rest of this paper we will use the notion cluster and mean non-trivial clusters only because all trees on $n$ taxa share the trivial clusters.
+We will refer to this list of clusters $n-2$ of a given tree on $n$ taxa as \emph{cluster representation} of the tree.
 
 We are now presenting $\findpath$ (Algorithm~\ref{alg:find_path}), an algorithm for computing paths between any two given trees in $\rnni$.
-Let $C_1, \ldots, C_{n-2}$ be the clusters of $R$ (the destination tree) where $C_i$ is induced by the node with rank $i$ for $i = 1, \ldots, n-2$.
+Let $[C_1, \ldots, C_{n-2}]$ be the cluster representation of $R$ (the destination tree).
+Then $C_i$ is induced by the node with rank $i$ for $i = 1, \ldots, n-2$.
 $\findpath$ computes a path $p$ which initially consists of $T$ only and is extended stepwise.
-Let $\hat T$ denote the last tree of $p$ at any point in the algorithm.
-For extending $p$, which equals updating $\hat T$, the clusters $C_1, \ldots, C_{n-2}$ are considered in this order.
-When arriving at cluster $C_i$ the clusters in $\hat T$ are rearranged by some $\rnni$ moves until $C_i$ is induced by a node in $\hat T$.
-This works as follows.
-First find the node inducing the smallest cluster containing all elements of $C_k$ in $\hat T$.
-This node is denoted by $\mrca_{\hat{T}}(C_k)$ (short for \emph{most recent common ancestor}).
-The rank of $\mrca_{\hat{T}}(C_k)$ in $\hat T$ is higher than the rank of the node inducing $C_k$ in $R$, because we consider $C_i$ for increasing $i$.
-As a result, all clusters $C_1, \ldots, C_{i-1}$ are induced by some nodes of $\hat T$ as they are in the destination tree $R$.
-Now update $\hat{T}$ by performing an $\rnni$ move that decreases the rank of $\mrca_{\hat{T}}(C_i)$.
-With Lemma~\ref{lemma:mrca_move} we know that there exists always exactly one $\rnni$ move that decreases the rank of $\mrca_{\hat{T}}(C_i)$.
-So we can perform these moves until $\mrca_{\hat{T}}(C_i)$ has the same rank $i$ in $\hat T$ as in $R$.
-Repeating this procedure for all clusters of $R$ results in a path from $T$ to $R$.
+In order to do this, the clusters $C_1, \ldots, C_{n-2}$ of $R$ are considered in that order.
+Let $T'$ be the last tree of $p$ when arriving at cluster $C_k$.
+We assume that the clusters $C_1, \ldots, C_{k-1}$ are induced by nodes of rank $1, \ldots, k-1$ in $T'$, respectively.
+In the first step it is $T' = T$ and $C_k = C_1$.
+The aim is now to find a sequence of $\rnni$ moves, starting at $T'$, that do not change the clusters $C_1, \ldots, C_{k-1}$, but result in a tree containing clusters $C_1, \ldots, C_{k}$.
+Therefore, we need to find the smallest cluster containing all elements of $C_k$ in $T'$.
+This node is denoted by $\mrca_{T'}(C_k)$ (short for \emph{most recent common ancestor}).
+The rank of $\mrca_{T'}(C_k)$ in $T'$ is higher than the rank of the node inducing $C_k$ in $R$, because $C_1, \ldots, C_{k-1}$ are already present in $T'$ and we assume that $C_k$ is not.
+Now perform an $\rnni$ move that decreases the rank of $\mrca_{T'}(C_k)$.
+With Lemma~\ref{lemma:mrca_move} we know that such an $\rnni$ move always exists if long as $T'$ does not induce the cluster $C_k$.
+So we can perform such moves until a tree $T''$ with $\mrca_{T''}(C_k)$ is reached that contains a node inducing $C_k$ that has the same rank $k$ in $T''$ as in $R$.
+Appending all trees on this path from $T'$ to $T''$ and repeating this procedure for all clusters of $R$ results in a path $p$ from $T$ to $R$.
 
@@
 \begin{algorithmic}[1]
-	\STATE $\hat{T} := T$
-	\STATE $p := [\hat T]$
+	\STATE $T' := T$
+	\STATE $p := [T']$
 	\FOR {$i = 1, \dots, n-2$}
 		\STATE Let $C_i$ be the cluster induced by the node with rank $i$ in $R$ \label{alg:find_path:line:cluster}
-		\WHILE {$\rank_{\hat T}(\mrca_{\hat{T}}(C_i))>i$}
-			\STATE Update $\hat{T}$: Decrease $\rank_{\hat T}(\mrca_{\hat{T}}(C_i))$ by an $\rnni$ move \label{alg:findpath:line:move_set_down}
-			\STATE $p = p+\hat{T}$
+		\WHILE {$\rank_{T'}(\mrca_{T'}(C_i))>i$}
+			\STATE Update $T'$: Decrease $\rank_{T'}(\mrca_{T'}(C_i))$ by an $\rnni$ move \label{alg:findpath:line:move_set_down}
+			\STATE $p = p+T'$
 		\ENDWHILE
@@
 
-Since $\findpath$ computes a path between two arbitrary trees in $\rnni$, we can use the length of such a path as approximation for the distance between the trees.
-A natural question is now how well this estimates the actual distance in $\rnni$.
+With Lemma~\ref{lemma:mrca_move} we can conclude that $\findpath$ is a deterministic algorithm.
+\todo{LC: move this up (before lemma1?)}
+Since it computes a path between two arbitrary trees in $\rnni$, we can use the length of such a path as approximation for the distance between the trees.
+A natural question is how well this estimates the actual distance in $\rnni$.
 To answer this question for small trees, we implemented $\findpath$.
@@
 
-The running time of $\findpath$, implemented as described by the pseudo-code in Algorithm~\ref{alg:find_path}, is quadratic in the number of taxa.
+The running time of $\findpath$, implemented as described by the pseudocode in Algorithm~\ref{alg:find_path}, is quadratic in the number of taxa.
 \todo{LC: Where do we mention the running time?}
@@
 
-Before introducing an algorithm for computing paths between caterpillar trees, we need to introduce further notations.
-We aim to find paths between \emph{caterpillar trees} where each internal node has at least one child that is a leaf.
+For introducing an algorithm to compute paths between caterpillar trees, we need further notations.
+In this section we aim to investigate properties of paths between \emph{caterpillar trees} where each internal node has at least one child that is a leaf.
 A path $p$ between two caterpillar trees $T$ and $R$ that only consists of caterpillar trees is called \emph{caterpillar path}.
@@
 
-We next introduce a representation of caterpillar trees as lists of taxa.
-We will use it to show how one can adapt the classical \emph{Bubble Sort} algorithm \autocite{Knuth1997-pi} to compute caterpillar paths.
-\todo{LC: pseudo-code}
-Ordering the taxa of a caterpillar trees as a non-decreasing, according to the ranks of their parents, gives us a list of length $n$.
-Note that the only two elements in such a list that have parents of equal rank are the first two elements $c_1$ and $c_2$.
+Let us now introduce a representation of caterpillar trees as lists of taxa.
+This representation is basically a simplification of the cluster representation for caterpillar trees and works as follows.
+For a caterpillar tree $T = [C_1, \ldots, C_{n-2}]$
+\todo{LC: Do we need to mention that we refer to a tree by this notation, or do we need to say that this is the cluster representation of a tree?}
+it is $C_i \subset C_{i+1}$ with $|C_i| = |C_{i+1}|$ for all $i = 1, \ldots, n-3$.
+Therefore, the list of taxa $L_T = [x_1, x_2, \ldots, x_n]$ with $x_1<x_2$ uniquely represents the caterpillar tree $T = [C_1, \ldots, C_{n-2}]$ with $C_i = \{x_1, \ldots, x_{i+1}\}$.
+In this list representing a caterpillar tree the taxa are ordered non-decreasing, according to the ranks of their parents.
+Note that the only two elements in such a list that have parents of equal rank are the first two elements $x_1$ and $x_2$.
+That is why we order them in the list such that $x_1 < x_2$ to make the list representation of caterpillar trees unique.
 We will call a pair of two taxa that share their parent \emph{cherry}.
-\todo{LC: pair of taxa or subtree (includes internal node and two edges)}
-To make the list representation of caterpillar trees unique, we assume that it is $c_1 < c_2$ for the first two elements in the list.
+We will use this list representation to show how one can adapt the classical \emph{Bubble Sort} algorithm \autocite{Knuth1997-pi} to compute caterpillar paths.
 
-In the following we explain how $\rnni$ moves between caterpillar trees work in the list representation.
+We are now going to explain what kind of $\rnni$ moves are possible between caterpillar trees and see how such a move changes the list representation of a tree.
 At first it is important to realise that the only possible $\rnni$ moves between caterpillar trees are $\nni$ moves, since there is no pair of internal nodes with rank difference one that is not connected by an edge.
@@
 There are two $\nni$ moves possible on this edge that result in caterpillar trees, while there is for every other edge only one $\nni$ move that results in a caterpillar tree.
-Corresponding to the $\nni$ moves on the lowest internal edge are exchanges of either the first or the second element with the third elements in the list representation.
+Corresponding to the $\nni$ moves on the lowest internal edge are exchanges of either the first or the second element with the third element in the list representation.
 
-For computing shortest caterpillar paths we modify Bubble Sort as described in the following.
+By using this knowledge we can compute shortest caterpillar paths by modifying Bubble Sort as described in the following.
 We call this algorithm $\csort$ (Algorithm~\ref{alg:csort}).
 
-\todo{Description + Pseudo code $\csort$}
-We consider the input trees $T$ and $R$ in their list representation $L_T$ and $L_R$ as described above and want to compute a path from $L_T$ to $L_R$.
-A path $p$ is built stepwise by adding a new list to the end of $p$.
-We call the last element of $p$ at any step $L_{\hat T}$ and when we say we update $L_{\hat T}$, we modify this list and append the new version of $L_{\hat T}$ to $p$.
+Let $T$ and $R$ be the input trees for $\csort$.
+We assume that they are given in their list representation $L_T$ and $L_R$ as described above and want to compute a path from $L_T$ to $L_R$.
+This path $p$ is built stepwise by adding a new list to the end of $p$.
+We call the last element of $p$ at any step $L_{T'}$, representing a tree $T'$ and when we say we update $L_{T'}$, we modify this list and append the new version of $L_{T'}$ to $p$.
 
-As in the Bubble Sort algorithm, we go through the list $L_{\hat T}$ and swap two elements if they appear in different orders in $L_{\hat T}$ and $L_R$.
-Each swap of elements is an update of $\hat T$ and extends the path $p$ by one element.
+As in the Bubble Sort algorithm, we go through the list $L_{T'}$ and swap two elements if they appear in different orders in $L_{T'}$ and $L_R$.
+Each swap of elements is an update of $T'$ and extends the path $p$ by one element.
 We compare pairs of list elements from the start of the list until its end, and repeat this until the whole list is sorted, just as in classical Bubble Sort.
 Our algorithm $\csort$ only differs from Bubble Sort when considering the first three elements of the list.
-Since the first two elements refer to the cherry taxa of the caterpillar tree, both can exchange with the third element of the list as $\nni$ move.
+As we discussed above, both the first and second element of $L_{T'}$ can exchange with its third element.
 Therefore, we need to consider these moves separately to all other moves.
 Most importantly, we need to make sure that these moves are performed in a way that leads to a shortest caterpillar path.
-We can guarantee this, if the moves are done as described in the following.
-Let $c_1, c_2$ and $c_3$ be the first three elements of $L_{\hat T}$, in this order.
-If $c_1$ ($c_2$) is the last element of those three in $L_R$, we update $L_{\hat T}$ by swapping $c_1$ ($c_2$) and $c_3$.
-By doing this we make sure that there will be no point on $p$ where $c_1$ and $c_2$ swap positions, which would cause an additional $\nni$ move that can be avoided if we perform the moves as we describe them here.
+We can guarantee this if the moves are done as described in the following.
+Let $x_1, x_2$ and $x_3$ be the first three elements of $L_{T'}$, in this order.
+If $x_1$ ($x_2$) is the last element of those three in $L_R$, we update $L_{T'}$ by swapping $x_1$ ($x_2$) and $x_3$.
+By doing this we make sure that there will be no point on $p$ where $x_1$ and $x_2$ swap positions, which would cause an additional $\nni$ move that can be avoided if we perform the moves as described.
 
-Because all moves between caterpillar trees are $\nni$ moves as described in the previous paragraph, all pairs of taxa that are not in the same order in $L_T$ and $L_R$ need to swap at some point on any path between $L_T$ and $L_R$.
+Because all moves between caterpillar trees are $\nni$ moves, all pairs of taxa that are not in the same order in $L_T$ and $L_R$ need to swap at some point on any path between $L_T$ and $L_R$.
 And since only elements that are direct neighbours in a sequence are allowed to swap, with the only exception being the first three elements, the path computed by $\csort$ is a shortest caterpillar path.
 \todo{LC: Do we need further explanation why $\csort$ gives a shortest caterpillar path?}
-
-\todo{LC: improve readability of pseudo-code}
 
@@
 \begin{algorithmic}[1]
-    \STATE $\hat T:= T$, $p = [\hat T]$
-    \STATE $L:=$ list representation of $\hat T$
+    \STATE $T':= T$, $p := [T']$
+    \STATE $L_{T'}:=$ list representation of $T'$, $L_R:=$ list representation of $R$
 	% \STATE $L_{\hat T} :=$ list representation of $\hat T$, $L_R :=$ list representation of $R$
-    \FOR {$i=0,\ldots,n-1$}
+    \FOR {$i=1,\ldots,n-2$} \label{alg:csort:line:loop}
         \STATE $j = 0$ \COMMENT{Special case: cherry}
-        \IF {$L[0]$ is last element of $L[0],L[1],L[2]$ in list representation $R$}
-            \STATE Update $\hat T$: swap $L[0]$ and $L[2]$
-        \ELSIF {$L[1]$ is last element of $L[0],L[1],L[2]$ in list representation $R$}
-            \STATE Update $\hat T$: swap $L[1]$ and $L[2]$
+        \IF {$L_{T'}[0]$ is last element of $L_{T'}[0],L_{T'}[1],L_{T'}[2]$ in $L_R$}
+            \STATE Update $T'$: swap $L_{T'}[0]$ and $L_{T'}[2]$
+        \ELSIF {$L_{T'}[1]$ is last element of $L_{T'}[0],L_{T'}[1],L_{T'}[2]$ in $L_R$}
+            \STATE Update $\hat T$: swap $L_{T'}[1]$ and $L_{T'}[2]$
         \ENDIF
@@
 	    % \IF {$L_R^{-1}[L_{\hat T}[j]] > L_R^{-1}[L_{\hat T}[j+1]]$}
-        \IF {order of $L[j]$ and $L[j+1]$ different in $\hat T$ and $R$}
-			\STATE Update $\hat{T}$: Swap $L[j+1]$ and $L[j]$
-			\STATE $p = p+\hat{T}$
-		\ENDIF
+        \WHILE {order of $L_{T'}[j]$ and $L_{T'}[j+1]$ different in $R$}
+			\STATE Update $T'$: Swap $L_{T'}[j]$ and $L_{T'}[j+1]$
+			\STATE $p = p+T', j = j + 1$
+		\ENDWHILE
     \ENDFOR
@@
 
-\todo{Check throughout that $\parent$ is never used without tree name in the index.}
+The running time of $\csort$ is the same as for Bubble Sort, that is quadratic in $n$.
+Within Section~\ref{section:caterpillar_convex} we will use $\csort$ for some proofs.
+Therefore it is important to notice that at the end of each loop $i$ starting in Line~\ref{alg:csort:line:loop} of Algorithm~\ref{alg:csort}, the last $i$ elements of $L_{T'}$ are the same as the ones in $L_R$.
+This is a well known property of Bubble Sort that $\csort$ inherits.
+\todo{LC: cite}
 
@@
 At first, all taxa of $T$ are sorted in a list $L$ such that the ranks of their parents are non-decreasing.
+This representation is somehow similar to the list representation of caterpillar trees, but it is ambiguous since we do not specify in which order taxa appear in the list that share their parent.
 If the input is a caterpillar tree, it follows that $L[0]$ and $L[1]$ are the cherry taxa of $T$, in an arbitrary order, and $\parent(L[i])$ has rank $i$ in $T$ for all $i = 2, \ldots, n-1$.
-Note that this list $L$ does not give a unique representation of a tree.
-The output tree $R$ is built inductively as follows:
+$\mdtree$ builds the output tree $R$ as follows:
 Initially, $R$ only consists of a cherry with taxa $L[0]$ and $L[1]$ such that the parent of these taxa has rank $n-1$.
@@
 
-Algorithm~\ref{alg:max_dist_tree} provides the pseudocode for this algorithm.
+Algorithm~\ref{alg:max_dist_tree} provides pseudocode for this algorithm.
 We will see in Section~\ref{section:diameter} that it computes trees with maximum distance form any given caterpillar tree.
@@
 
-An important property of $\mdtree$, which we will use later on in Section~\ref{section:diameter} is the fact that this algorithm is non-deterministic.
-This is because the list computed in Line~\ref{alg:mdtree_lineL} the and the attachment edge of the new taxon in the tree in Line~\ref{alg:mdtree_lineAddTaxon} are not uniquely determined.
-Therefore the output tree $R$ of $\mdtree$ is not unique.
+An important property of $\mdtree$, which we will use later on in Section~\ref{section:diameter}, is the fact that this algorithm is non-deterministic.
+Because the list computed in Line~\ref{alg:mdtree_lineL} the and the attachment edge of the new taxon in the tree in Line~\ref{alg:mdtree_lineAddTaxon} are not uniquely determined, the output tree $R$ of $\mdtree$ is not unique.
 

### 2019-10-29 paper.tex
another pass through geometry section + rm some todos

@@
 Specifically, in this paper we establish the exact radius and diameter of the space, [...
--- Lena, please finish this including the approaximate algorithm that works exactly on small spaces (Lena: is it already in the text?)]
+-- Lena, please finish this including the approximate algorithm that works exactly on small spaces (Lena: is it already in the text?)]
 The question of whether there exists a polynomial algorithm for computing the $\rnni$ distance remains an open problem.
@@
 
-In \autocite{Gavryushkin2018-ol} an exact algorithm for computing the $\rnni$ graph has been introduced.
-We implemented this algorithm and used it for computing this graph for trees on up to $7$ taxa.
-\todo{LC: $7$ or $8$?}
-\todo{Extend this section}
+In Section 3.3 of \autocite{Gavryushkin2018-ol} an exact algorithm for computing the $\rnni$ graph has been introduced.
+We implemented this algorithm and used it for computing this graph for trees on up to seven taxa.
+\todo{LC: $7$ or $8$? - I'll need to check this}
+Once the graph is computed, one can get distances by using algorithms such as Floyd-Warshall \autocite{Floyd1962-ew} or Dijkstra \autocite{Dijkstra1959-ph}.
+However, the large number of vertices in $\rnni$ is an obstacle when it comes to computing distances.
+Since there are $\frac{n!(n-1)!}{2^{n-1}}$ vertices in the $\rnni$ graph \autocite{Gavryushkin2018-ol}, we already run out of space when we want to save all pairwise distances for trees on seven taxa.
+Therefore, we compute distances between certain pairs of trees and use the symmetry of $\rnni$ to draw conclusions about the remaining distances.
 
@@
 
-\todo{LC: Work on the following proof! Is it worth introducing intervals for $\rnni$ moves for the proof of the following lemma?}
-
 \begin{lemma}
@@
 \begin{proof}
-
-    \todo{LC: revise this whole proof}
 
@@
 
-    % Let $v:=\mrca_{\hat T}(C_i)$, let $C_{i+1}$ be the cluster induced by $v$, and let $w$ be the node inducing $C_i$.
-    % We will distinguish the case that there is an edge $(v,w)$ in $T$ from the case that there is no such edge.
-    %
-    % At first we assume that there is no edge $(v,w)$, which means $C_i \not \subset C_{i+1}$.
-    % In this case the two sets $C_i, C_{i+1}$ can exchange by a rank swap.
-    % Since $C \subset C_{i+1}$, this rank swap decreases the rank of $\mrca(C)$ by one and there is no $\nni$ move that could move $\mrca(C)$ down.
-    %
-    % Let us now consider the case that $(v,w)$ is an edge in $T$.
-    % The only move that could decrease the rank of $\mrca(C)$ is an $\nni$ move on the edge $(v,w)$.
-    % Let $C_k$ and $C_j$ be the clusters induced by the children of $C_i$.
-    % Then it is $C_k \cup C_j = C_i$.
-    % There are two different $\nni$ moves possible on $(v,w)$.
-    % $C_i$ can either be replaced by $C_j \cup (C_{i+1} \setminus C_i)$ or by $C_k \cup (C_{i+1} \setminus C_i)$.
-    % To prove that there is always exactly one $\rnni$ moves that decreases the rank of $\mrca(C)$, we need to show that it is either $C \subset (C_j \cup (C_{i+1} \setminus C_i))$ or $C \subset (C_k \cup (C_{i+1} \setminus C_i))$.
-    %
-    % Let $a \in C$.
-    % Then it is $a \in C_{i+1} = mrca(C)$ and it follows $a \in C_{i+1} \setminus C_i$ or $a \in C_i$.
-    % If $a \in C_{i+1} \setminus C_i$, $a$ is in the new set after the $\nni$ move.
-    % So now we will assume that $a \in C_i$.
-    % Then it is either $a \in C_j$ or $a \in C_k$.
-    % Without loss of generality we assume that $a \in C_j$.
-    %
-    % We need to prove that if $a \in C_j$, there is no $b \in C$ with $b \in C_k$.
-    % This makes sure that it is either $C \subset (C_j \cup (C_{i+1} \setminus C_i))$ or $C \subset (C_k \cup (C_{i+1} \setminus C_i))$.
-    % For contradiction we assume that $a \in C_j$ and $C \in S_k$.
-    % We know that all sets that are subsets of $C$ already are at the same position in the current tree as in the destination tree $R$.
-    % Therefore, we can conclude from $a \in C_j$ and $b \in C_k$ that $C_j \cup C_k = C$.
-    % Hence, it holds $mrca(C) = C_i$.
-    % This is a contradiction to our assumption that $C_{i+1} = mrca(C)$ in the current tree.
-    % Therefore, it holds either $C \subset (C_j \cup (C_{i+1} \setminus C_i))$ or $C \subset (C_k \cup (C_{i+1} \setminus C_i))$.
-    %
-    % Concluding, there is always an $\rnni$ move on the edge $(v,w)$ for $mrca(C) = C_{i+1}, C \subsetneq C_{i+1}$ that decreases the rank of $mrca(C)$.
-
 \end{proof}
@@
 With Lemma~\ref{lemma:mrca_move} we can conclude that $\findpath$ is a deterministic algorithm.
-\todo{LC: move this up (before lemma1?)}
+Its running time is quadratic in the number of taxa.
 Since it computes a path between two arbitrary trees in $\rnni$, we can use the length of such a path as approximation for the distance between the trees.
@@
 \todo{LC: github - link}
-For computing the actual $\rnni$ distances we used the the algorithm for computing the $\rnni$ graph that was introduced in Section 3.3 of \autocite{Gavryushkin2018-ol}.
-With our implementation of this algorithm
-\todo{LC: gihub - link}
-we can compute the graph for trees with up to $8$ taxa in reasonable time. \todo{LC: Shall we say more about the algorithm here? + running time (in hours/minutes) on which computer?}
-We used
+For computing the actual $\rnni$ distances we used the the algorithm for computing the $\rnni$ graph as described in Section~\ref{section:alg_RNNI_graph}.
 \todo{LC: How do we compare the distances?}
-
-The running time of $\findpath$, implemented as described by the pseudocode in Algorithm~\ref{alg:find_path}, is quadratic in the number of taxa.
-\todo{LC: Where do we mention the running time?}
 
@@
 \label{section:alg_csort}
-
-\todo{LC: move definitions of caterpillar stuff here}
-\todo{LC: pseudo code}
 
@@
 This representation is basically a simplification of the cluster representation for caterpillar trees and works as follows.
-For a caterpillar tree $T = [C_1, \ldots, C_{n-2}]$
-\todo{LC: Do we need to mention that we refer to a tree by this notation, or do we need to say that this is the cluster representation of a tree?}
-it is $C_i \subset C_{i+1}$ with $|C_i| = |C_{i+1}|$ for all $i = 1, \ldots, n-3$.
+For a caterpillar tree with cluster representation $T = [C_1, \ldots, C_{n-2}]$ it is $C_i \subset C_{i+1}$ with $|C_i| = |C_{i+1}|$ for all $i = 1, \ldots, n-3$.
 Therefore, the list of taxa $L_T = [x_1, x_2, \ldots, x_n]$ with $x_1<x_2$ uniquely represents the caterpillar tree $T = [C_1, \ldots, C_{n-2}]$ with $C_i = \{x_1, \ldots, x_{i+1}\}$.
@@
 And since only elements that are direct neighbours in a sequence are allowed to swap, with the only exception being the first three elements, the path computed by $\csort$ is a shortest caterpillar path.
-\todo{LC: Do we need further explanation why $\csort$ gives a shortest caterpillar path?}
 
@@
 Therefore it is important to notice that at the end of each loop $i$ starting in Line~\ref{alg:csort:line:loop} of Algorithm~\ref{alg:csort}, the last $i$ elements of $L_{T'}$ are the same as the ones in $L_R$.
-This is a well known property of Bubble Sort that $\csort$ inherits.
-\todo{LC: cite}
+This is a well known property of Bubble Sort that $\csort$ inherits from it.
 
@@
 The algorithm we are going to introduce here will be important for finding the radius of the $\rnni$ graph.
-In Section~\ref{section:diameter} we will prove that it computes a tree with maximum distance from a given caterpillar tree, which implies that the radius of $\rnni$ equals its diameter.
+In Section~\ref{section:diameter} we will prove that it computes a tree with maximum distance from a given caterpillar tree (Lemma~\ref{lemma:max_dist_caterpillar}), which implies that the radius of $\rnni$ equals its diameter (Corollary~\ref{corollary:radius}).
 Therefore, it is called $\mdtree$ (Maximum Distance Tree).
-\todo{LC: References to the corresponding proofs?}
 
@@
 This representation is somehow similar to the list representation of caterpillar trees, but it is ambiguous since we do not specify in which order taxa appear in the list that share their parent.
-If the input is a caterpillar tree, it follows that $L[0]$ and $L[1]$ are the cherry taxa of $T$, in an arbitrary order, and $\parent(L[i])$ has rank $i$ in $T$ for all $i = 2, \ldots, n-1$.
+If the input is a caterpillar tree, it follows that $L[0]$ and $L[1]$ are the cherry taxa of $T$, in an arbitrary order, and the parent of taxon $L[i]$ has rank $i$ in $T$ for all $i = 2, \ldots, n-1$.
 $\mdtree$ builds the output tree $R$ as follows:
@@
 In other words, the set of caterpillar trees is \emph{convex} in $\rnni$.
-In Section~\ref{section:diameter} we will investigate diameter and radius of the $\rnni$ graph, before we discuss a conjecture, the so-called Split Theorem, which was raised in \autocite{Gavryushkin2018-ol}.
+In Section~\ref{section:diameter} we will investigate diameter and radius of the $\rnni$ graph.
+Interestingly, the \emph{diameter}, that is the maximum distance $\Delta(\rnni) = \max \{d(T, R) \mid T, R \in \rnni\}$ between any two trees in the graph, of $\rnni$ equals its \emph{radius} $rad(\rnni) =  \min\limits_T \max\limits_R d(T,R)$ in $\rnni$.
+After proving this, we are going to consider the so-called Split Theorem, which was raised in \autocite{Gavryushkin2018-ol}.
 We will give a counterexample to disprove their version of the Split Theorem and claim an alternative version.
@@
 Before we get into the details of these geometric properties of $\rnni$, we need to prove some statements that will be used a few times within this geometry section.
-At first we in Lemma~\ref{lemma:diameter_bound} an upper bound of the \emph{diameter} $\Delta(\rnni)$ of the $\rnni$ graph, which is defined as $\Delta(\rnni) = \max \{d(T, R) \mid T, R \in \rnni\}$.
 
 \begin{lemma}
-$\Delta(\rnni) \leq \frac{(n-1)(n-2)}{2}$ for $n \geq 3$.
+    It is $\Delta(\rnni) \leq \frac{(n-1)(n-2)}{2}$ for $n \geq 3$.
     \label{lemma:diameter_bound}
@@
 Since the diameter is bounded from above by the maximum length of a path computed by $\findpath$, it is enough to find this maximum.
-The worst-case execution of $\findpath$ would require each of the $n-2$ clusters $C_i$, considered in Line~\ref{alg:find_path:line:cluster} of Algorithm~\ref{alg:find_path}, to be moved by $n-1-i$ $\rnni$ operations in $\hat{T}$.
+The worst-case execution of $\findpath$ would require each of the $n-2$ clusters $C_i$, considered in Line~\ref{alg:find_path:line:cluster} of Algorithm~\ref{alg:find_path}, to be moved by $n-1-i$ $\rnni$ operations in $T'$.
+This can happen if each cluster $C_i$ has the root as $\mrca_{T'}(C_i)$.
+Then the rank of this most recent common ancestor must be changed from $n-1$ to $i$, which requires $n-1-i$ operations, according to Lemma~\ref{lemma:mrca_move}.
 Hence, the maximum length of such a path is bounded by $\sum\limits_{i = 1}^{n-2} i = \frac{(n-2)(n-1)}{2}$.
@@
 \begin{proof}
+Let $T$ and $R$ be trees as described in the lemma.
 First observe that the rank of the internal node $\parent_T(n+1)$ can only be changed by performing a rank move that involves $\parent_T(n+1)$ or an $\nni$ move on an edge adjacent to $\parent_T(n+1)$.
 Second observe that any $\rnni$ move can change the rank of $\parent_T(n+1)$ by at most one.
-So we can follow that the rank of $\parent_T(n+1)$ can be decreased by at most one by an $\rnni$ move on $T$.
+So we can follow that an $\rnni$ on $T$ move can decrease the rank of $\parent_T(n+1)$ by at most one.
 
-Let $p$ be a shortest path from tree $T$ to $R$ and $p{\big|}_n$ the path resulting from deleting taxon $n+1$ from all trees on $p$.
-This new path $p{\big|}_n$ is then a path from $T{\big|}_n$ to $R{\big|}_n$.
+Let $p$ be a shortest path from $T$ to $R$ and $p{\big|}_n$ the path resulting from deleting taxon $n+1$ from all trees on $p$.
+Then $p{\big|}_n$ is a path from $T{\big|}_n$ to $R{\big|}_n$.
 The reason for this is that the only moves that are deleted from $p$ to receive $p{\big|}_n$ are those involving the parent of taxon $n+1$.
@@
 
-Using the same argument as in the proof of Lemma~\ref{lemma:distance_delete_taxon} we can establish the following Proposition~\ref{proposition:lower_bound_distance} that provides a lower bound on the distance between trees.
+\todo{LC: We don't use the following Proposition anywhere in the paper. Do we want to keep it?}
+
+Using the same argument as in the proof of Lemma~\ref{lemma:distance_delete_taxon} we can establish Proposition~\ref{proposition:lower_bound_distance} that provides a lower bound on the distance between trees.
 
@@
 
-In this section we restrict our attention to the set of \emph{caterpillar trees} where each internal node has at least one child that is a leaf.
+In this section we restrict our attention to the set of caterpillar trees, which we will represent throughout this section in the simplified list representation as introduced in Section~\ref{section:alg_csort}.
 This type of tree is of particular interest in $\rnni$ because shortest paths between such trees in $\rnni$ differ from those in classical $\nni$ space.
@@
 We discovered this by computing caterpillar paths with $\csort$ and comparing them with shortest path in $\rnni$, computed by the exact algorithm as explained in Section~\ref{section:algorithms}.
+\todo{LC: Do we explain how we did this in the alg section? If not, we need to add an explanation here!}
 In the following we demonstrate why finding shortest paths between caterpillar trees builds the basis for understanding the difference between $\nni$ and $\rnni$.
@@
 There is only one move needed to build and resolve a cherry, respectively, while the number of moves for moving the cherry is the same as for moving a single taxon in $\nni$.
-But in $\rnni$ space there are additional rank moves needed to change the rank of the internal node of the cherry before an $\nni$ move resolves it.
+But in $\rnni$ space there are additional rank moves needed to change the rank of the internal node of the cherry before an $\nni$ move can resolve it.
 This is due to the fact that $\nni$ moves are only allowed on edges of length one in $\rnni$.
@@
     \vspace{2pt}
-    \todo{LC: Change caption}
-	\caption{Paths between caterpillar trees $T$ and $R$: black -- shortest path in classical $\nni$ (ignoring ranks) of length $5$; red -- shortest path in $\rnni$ of length $6$; blue -- path in $\rnni$ which is the natural extension of the shortest path in $\nni$ but has length $8$.}
+	\caption{Paths between caterpillar trees $T$ and $R$: The solid paths are paths in $\rnni$, the dashed one is a path that is only possible in the plain $\nni$ graph.
+    The upper path, which is the $\rnni$ path that is the natural extension of the shortest $\nni$ path, is longer than a shortest $\rnni$ path as the one at the bottom.}
 	\label{fig:NNI_vs_RNNI}
@@
 Lemma~\ref{lemma:caterpillar_dist=diameter} regards caterpillar trees with distance $\frac{(n-1)(n-2)}{2}$ and will be needed for proving the convexity of the set of caterpillar trees in $\rnni$.
-Notice that by Lemma~\ref{lemma:diameter_bound} $\frac{(n-1)(n-2)}{2}$ is the upper bound for the diameter of $\rnni$.
+Notice that $\frac{(n-1)(n-2)}{2}$ is the upper bound for the diameter of $\rnni$ by Lemma~\ref{lemma:diameter_bound}.
 
@@
     This easily follows from algorithm $\csort$:
-    In each loop $i=1, \ldots, n-2$ of the algorithm there can be at most $n-1-i$ exchanges of taxa.
-    It follows that the maximum length of a path computed by this algorithm is $\sum\limits_{k=1}^{n-2} k = \frac{(n-1)(n-2)}{2}$.
-    Since there cannot be a caterpillar path shorter than $\frac{(n-1)(n-2)}{2}$ if it is $d(T,R) = \frac{(n-1)(n-2)}{2}$, it follows $d_c(T,R) =  \frac{(n-1)(n-2)}{2}$.
+    In each loop $i=1, \ldots, n-2$ of Algorithm~\ref{alg:csort} there are at most $n-1-i$ exchanges of taxa.
+    This is due to the fact that after loop number $i$ the last $i$ elements in the list representation of the current tree on $p$ are sorted correctly as in the destination tree.
+    Hence there are no more exchanges of pairs of taxa needed within the last $i-1$ of the sequences in loop $i$, and so only $n-1-i$ swaps of elements might appear within the first $n-i$ elements of the list.
+    It follows that the maximum length of a path computed by $\csort$ is $\sum\limits_{k=1}^{n-2} k = \frac{(n-1)(n-2)}{2}$.
+    Since there cannot be a caterpillar path shorter than $d(T,R) = \frac{(n-1)(n-2)}{2}$, it follows $d_c(T,R) =  \frac{(n-1)(n-2)}{2}$.
 
@@
     Let $R$ be a caterpillar tree with distance $\frac{n(n-1)}{2}$ to $T$.
-    Then each of the $\frac{n(n-1)}{2}$ pairs of taxa that is considered when running $\csort$ to compute a path $p$ from $R$ to $T$ exchanges.
+    Then each of the $\frac{n(n-1)}{2}$ pairs of taxa that is considered when running $\csort$ to compute a path $p$ from $T$ to $R$ exchanges.
     In particular, taxon $n+1$ exchanges with a taxon in each of the $n+1$ loops in $\csort$.
@@
 
+    We are now using this observation to prove that the distance between $T$ and $R$ equals $\frac{n(n-1)}{2}$.
+    It is obvious that it gives an upper bound for the distance as it is $\Delta(\rnni_{n+1}) \leq \frac{n(n-1)}{2}$ for the $\rnni$ graph $\rnni_{n+1}$ on $n+1$ taxa by Lemma~\ref{lemma:diameter_bound}.
+    So we now need to prove that $\frac{n(n-1)}{2}$ is a lower bound for the distance between $T$ and $R$ as well.
     According to Lemma~\ref{lemma:distance_delete_taxon}, the distance between $T{\big|}_n$ and $R{\big|}_n$ is at least $d(T{\big|}_n, R{\big|}_n) \leq d(T,R) - |\rank_T(\parent_T(n+1)) - \rank_R(\parent_R(n+1))|$.
-    We also know $|\rank_T(\parent_T(n+1)) - \rank_R(\parent_R(n+1))| = n-1$.
+    We also know $|\rank_T(\parent_T(n+1)) - \rank_R(\pare nt_R(n+1))| = n-1$.
     Suppose, contrary to our claim, that there is a path from $T$ to $R$ of length less than $\frac{n(n-1)}{2}$.
@@
 
+\todo{LC: Work on the following proof, especially the figure and explanation of it}
+
 \begin{proof}
     We prove this Theorem by backwards induction on the length $l$ of a shortest caterpillar path between two caterpillar trees $T$ and $R$.
-    Since we use backwards induction, the maximum distance between any two caterpillar trees, which is $\frac{(n-1)(n-2)}{2}$ (Lemma~\ref{lemma:diameter_bound}), is the base case.
+    For this backwards induction the maximum caterpillar distance, which is $\frac{(n-1)(n-2)}{2}$ (Lemma~\ref{lemma:caterpillar_dist=diameter}), is the base case.
 
@@
     Induction hypothesis: For each pair of caterpillar trees with caterpillar distance $l+1$ there is a shortest path between these trees that only consists of caterpillar trees, which means that the distance between these trees is $l+1$.
+
     Let $T$ and $R$ be caterpillar trees with caterpillar distance $l$.
-    We consider the shortest caterpillar path computed by $\csort$.
-    Since $l < \frac{(n-1)(n-2)}{2}$, there must be one of the $n$ loops in $\csort$ where less than $n-2-k$ taxon pairs swap positions.
-    Let us consider loop number $k$ where this happens for the first time.
+    We consider the shortest caterpillar path from $T$ to $R$ computed by $\csort$.
+    Since $l < \frac{(n-1)(n-2)}{2}$, there must be one loop $k$ in $\csort$ where less than $n-2-k$ taxon pairs swap positions.
+    Let us assume that loop number $k$ is the first one where this happens.
     This situation is illustrated in the top of Figure~\ref{fig:cat_convex_seq_example}.
-    Whenever the number of swaps of taxa or, in terms of lists, neighboured elements in the list is maximum in a loop, one of the first two elements swaps position with $n-2-k$ subsequent elements of the list.
-    Since we assume that the number of swaps in loop $k$ is not maximal, there are less than $n-2-k$ such moves.
+    Whenever the number of swaps of taxa or, in terms of lists, neighboured elements in the list is maximum in a loop, one of the first two elements swaps position with $n-1-k$ subsequent elements of the list.
+    We already saw this in the proof of Lemma~\ref{lemma:caterpillar_dist=diameter}.
+    Since we assume that the number of swaps in loop $k$ is not maximal, there are less than $n-1-k$ such moves.
+    \todo{LC: From here on it is confusing}
     Let us assume that $a$ is the first element that is compared with its left neighbour in loop $k$ but does not swap with it.
@@
 
-With the results of Section~\ref{section:caterpillar_convex} we can easily provide the exact diameter of the $\rnni$ graph and hence improve the bounds for the diameter given in \autocite{Gavryushkin2018-ol}.
-Furthermore, we are able to show that the \emph{radius} of the $\rnni$ graph, which is defined as $rad(G) =  \min\limits_u \max\limits_v d(u,v)$ for a graph $G$, equals its diameter.
+With the results of Section~\ref{section:caterpillar_convex} we can easily get the exact diameter of the $\rnni$ graph and hence improve the bounds for the diameter given in Theorem 7 of \autocite{Gavryushkin2018-ol}.
 At first we consider the diameter of the $\rnni$ graph.
-By Lemma~\ref{lemma:diameter_bound} we already know an upper bound for the diameter.
+Lemma~\ref{lemma:diameter_bound} already gives an upper bound for the diameter.
 We will now show that this is not only a bound, but the exact diameter of $\rnni$.
@@
     We can compute a shortest caterpillar path by using $\csort$ as explained in Section~\ref{section:caterpillar_convex}.
-    When running this algorithm on $T$ and $R$, each taxon $i$ ($i \in \{1, \ldots, n\}$) swaps $n-1-i$ times with its upper neighbour.
+    When running this algorithm to compute a path from $T$ to $R$, each taxon $i$ ($i \in \{1, \ldots, n\}$) swaps $n-1-i$ times with its upper neighbour.
     Therefore, the length of a shortest caterpillar path is $\sum\limits_{i=1}^{n-1}(n-1-i) = \sum\limits_{i=0}^{n-2}i = \frac{(n-1)(n-2)}{2}$.
@@
 We now proceed to show that the radius of the $\rnni$ graph equals its diameter.
-Therefore we use the algorithm $\mdtree$ that has been introduced in section~\ref{section:alg_mdtree}.
+Therefore we use the algorithm $\mdtree$ that has been introduced in Section~\ref{section:alg_mdtree}.
 Note that the results provided in this section are only true if the input tree of $\mdtree$ is a caterpillar tree.
@@
     In the list $L$ produced Line~\ref{alg:mdtree_lineL} in $\mdtree(T)$ (Algorithm~\ref{alg:max_dist_tree}) taxon $n+1$ is the last one, because its parent, the root, has the highest rank in $T$.
-    This means that this taxon $n+1$ is the last one added last to $R$ as in Line~\ref{alg:mdtree_lineAddTaxon} and therefore, it has parent with rank one in $R$.
-    Because the parent of $n+1$ has rank $n$ in $T$, we can conclude from Lemma~\ref{lemma:distance_delete_taxon} that for $T$ and $R$ restricted to taxa $1,\ldots,n$ it holds: $d(T|_n,R|_n) \leq d(T,R) - (n-1)$.
+    This means that this taxon $n+1$ is the last one added to $R$ as in Line~\ref{alg:mdtree_lineAddTaxon} of this algorithm and therefore, it has parent with rank one in $R$.
+    Because the parent of $n+1$ has rank $n$ in $T$ and rank $1$ in $R$, we can conclude from Lemma~\ref{lemma:distance_delete_taxon} that for $T$ and $R$ restricted to taxa $1,\ldots,n$ it holds: $d(T|_n,R|_n) \leq d(T,R) - (n-1)$.
     It is not hard to see that $R|_n$ can be received from $T|_n$ by applying $\mdtree$:
@@
 In Section~\ref{alg:max_dist_tree} we already discussed that $\mdtree$ is a non-deterministic algorithm.
-In fact, every tree topology on $n$ taxa can be resulting from $\mdtree$, since there is no restriction on where to add the new taxon in the already existing tree in Line~\ref{alg:mdtree_lineAddTaxon} of Algorithm~\ref{alg:max_dist_tree}.
+In fact, the output tree of $\mdtree$ can have any tree topology
+\todo{LC: Do we need to define 'tree topology'?}
+, since there is no restriction on where to add the new taxon in the already existing tree in Line~\ref{alg:mdtree_lineAddTaxon} of Algorithm~\ref{alg:max_dist_tree}.
 This means that for each caterpillar input tree $\mdtree$ can compute a tree with any tree topology that has maximum distance from the input tree.
-\todo{LC: Definition tree topology?}
-And because permuting labels of trees does not change the distance between them, there is for every tree $R$ on $n$ taxa a caterpillar tree $T$ with distance $d(T,R) = \Delta(\rnni) = \frac{(n-1)(n-2)}{2}$.
+And because permuting labels of trees does not change the distance between them, there exists for every tree $R$ on $n$ taxa a caterpillar tree $T$ with distance $d(T,R) = \Delta(\rnni) = \frac{(n-1)(n-2)}{2}$.
 This gives us Corollary~\ref{corollary:radius}.
@@
 
-%Do we need this?
-
-% \begin{lemma}
-% 	Let $T$ and $R$ be two caterpillar trees that both have a leaf labelled by $a$ as part of their cherry.
-% 	There is a shortest paths from $T$ to $R$ that is a caterpillar path.
-% 	\label{lemma:caterpillar_subset_convex}
-% \end{lemma}
-%
-% \begin{figure}[h]
-% 	\centering
-% 	\includegraphics[width=0.3\textwidth]{label_internal_nodes}
-% 	\caption{The labelling of the caterpillar tree as introduced in the proof of Lemma~\ref{lemma:caterpillar_subset_convex} is unique due to its definition.
-% 	}
-% 	\label{label_internal_nodes}
-% \end{figure}
-%
-% \begin{proof}
-% 	Let $T$ and $R$ be two caterpillar trees, both having a leaf labelled by $a$ as part of their cherry.
-% 	Let $p$ be a shortest path from $T$ to $R$.
-% 	We introduce a labelling of internal nodes of all trees on $p$ by first defining it for the caterpillar trees $T$ and $R$ and then explaining how the labelling is changed by $\rnni$ moves.
-%
-% 	In $T$ and $R$ the internal node of the cherry is labelled by the label of its child that is not $a$.
-% 	All other internal nodes shall have the same label as their leaf children.
-% 	From now on we will identify the leaf labelled by $x$ with $x$ and the internal node labelled by $x$ will be called $int_x$.
-% 	An example for a caterpillar tree with internal leaf labels is depicted in Figure~\ref{label_internal_nodes}.
-% 	In the following, changes of the internal labelling following $\rnni$ moves are defined:
-%
-% 	\textbf{Rank moves:}
-% 	Within rank moves the labels of internal nodes do not change.
-% 	For an example consider Figure~\ref{fig:rank_swap_internal_label}.
-%
-% 	\begin{figure}[H]
-% 		\centering
-% 		\includegraphics[width=0.8\textwidth]{rank_swap_internal_label}
-% 		\caption{The labelling of the internal nodes that swap ranks (red) do not change within the rank move.
-% 		}
-% 		\label{fig:rank_swap_internal_label}
-% 	\end{figure}
-%
-% 	\textbf{NNI move:}
-% 	Let there be an $\nni$ move on an edge $(int_x, int_y)$ where $\rank(int_x) < \rank(int_y)$ as in Figure~\ref{fig:nni_moves}.
-%
-% 	\begin{figure}[H]
-% 		\centering
-% 		\includegraphics[width=0.7\textwidth]{nni_moves_internal_label}
-% 		\caption{Two possible $\nni$ moves on edge $(int_x, int_y)$ where $x \in A, y \in B$.
-% 		}
-% 		\label{fig:nni_moves}
-% 	\end{figure}
-%
-% 	The nodes incident to the edge after the $\nni$ move are labelled by $int_x$ and $int_y$ as well.
-% 	By default, the node that has lower rank than the other one is labelled by $int_x$, the other one by $int_y$.
-% 	But if $int_x$ is no ancestor of $x$ or $int_y$ is no ancestor of $y$, we swap the labels of the two new nodes, such that $\rank(int_x) > \rank(int_y)$.
-% 	An example of an $\nni$ move and the new labelling is provided in Figure~\ref{fig:nni_example_internal_label}.
-%
-% 	\begin{figure}[H]
-% 		\centering
-% 		\includegraphics[width=0.8\textwidth]{nni_example_internal_label}
-% 		\caption{The first labelling after the $\nni$ move preserves the ranks of $int_4$ and $int_6$.
-% 		But then $int_6$ is not an ancestor of $6$ any more.
-% 		Therefore, $int_4$ and $int_6$ have to change their labels.
-% 		}
-% 		\label{fig:nni_example_internal_label}
-% 	\end{figure}
-%
-% 	We proceed to show that $int_x$ and $int_y$ are ancestors of $x$ and $y$ after an $\nni$ move as described above, respectively.
-% 	If $x$ stays in the subtree rooted in the lower node of $e$, this does obviously hold.
-% 	We now turn to the case that $x$ is not in the subtree rooted in the lower node of $e$ after the $\nni$ move.
-% 	Then the labels of the two internal nodes change such that $\rank(int_x) > \rank(int_y)$.
-% 	This makes sure that $int_x$ is an ancestor of $x$.
-% 	Thus, it remains to prove that $int_y$ is ancestor of $y$.
-% 	In particular, we need to show that $y$ is not in the same subtree $A$ as $x$, whose root has the node labelled by $int_x$ as parent before and after the $\nni$ move.
-% 	Let us consider the tree before the $\nni$ move.
-% 	$A$ has $|A|$ leaves and $|A| - 1$ internal nodes.
-% 	These internal nodes are labelled by some leaves of $A$, because all internal nodes must be labelled by a descendant leaf.
-% 	We know that $x \in A$, and the parent of the root of $A$ is labelled by $x$.
-% 	All remaining $|A|-1$ leaf labels of $A$ appear as labels of the $|A|-1$ internal nodes of $A$.
-% 	From the fact that the internal node labelled by $y$ is not inside $A$ we can conclude that $y$ cannot be a leaf of $A$.
-% 	This proves that the labelling of the internal nodes is is well defined and allows a unique identification of internal nodes by their labels.
-%
-% 	The task is now to prove that there is always a caterpillar path at least as short as the given path $p$.
-% 	Therefore, we use the following observation:
-% 	Within each $\rnni$ move the ranks of at most two internal nodes change by at most one each.
-% 	This is true for both rank moves and $\nni$ moves, according to our definition of the labelling:
-% 	Within a rank move two nodes exchange their ranks, while there are not necessarily changes of ranks when $\nni$ moves are performed.
-%
-% 	On a caterpillar path the only possible moves are $\nni$ moves that change the ranks of two internal nodes by one each.
-% 	Therefore, we can construct a caterpillar path $r$ out of $p$:
-% 	every time two internal nodes $int_x$ and $int_y$ exchange ranks on $p$, they do so on the caterpillar path $r$ as well.
-% 	Since there are moves on $p$ that do not change the ranks of internal nodes, this caterpillar path has length $|r| \leq |p|$.
-% 	Hence, there is always a shortest path between $T$ and $R$ that is a caterpillar path.
-% \end{proof}
-
 \subsection{Cluster Theorem}
 \label{section:cluster_theorem}
-\todo{Split or Cluster?}
 
@@
 Afterwards, $T'$ is transformed to a tree $R'$ still containing the same cherries as $T'$ but in a way that resolving all these cherries of $R'$ results in $R$.
-For proving that this path is shorter than any path preserving the clusters $\{1, \ldots, n\}$ and $\{n+1, \ldots, 2n\}$, \autocite{Li1996-zw} use the fact that $d_{\nni}(R,R') = d_{\nni}(T{\big|}_{\{1, \ldots, n\}}, R{\big|}_{\{1, \ldots, n\}})$ where $T{\big|}_{\{1, \ldots, n\}}$ and $R{\big|}_{\{1, \ldots, n\}}$ are $T$ and $R$ restricted to taxa $\{1,\ldots,n\}$, respectively.
+For proving that this path is shorter than any path preserving the clusters $\{1, \ldots, n\}$ and $\{n+1, \ldots, 2n\}$, \autocite{Li1996-zw} use the fact that $d_{\nni}(T',R') = d_{\nni}(T{\big|}_n, R{\big|}_n)$ where $T{\big|}_n$ and $R{\big|}_n$ are $T$ and $R$ restricted to taxa $\{1,\ldots,n\}$, respectively.
 However, this does not work in $\rnni$.
-Similar to what we observed in the example depicted in Figure~\ref{fig:NNI_vs_RNNI}, there are lots of additional rank moves needed on a similar path in $\rnni$.
-Therefore, a path where the two caterpillar trees are sorted separately, is shorter in $\rnni$.
+Similar to what we observed in the example depicted in Figure~\ref{fig:NNI_vs_RNNI}, there are lots of additional rank moves needed on a path in $\rnni$ that similar to the one in $\nni$ as described above.
+Therefore, a path where the two caterpillar trees are sorted separately is shorter in $\rnni$.
 
@@
     \caption{Example of a shortest path between two trees $T$ and $R$ in $\nni$.
-    All leaves labelled with dots are labelled by taxa in $\{1, \ldots, n$, squares represent taxa in $\{n+1, \ldots, 2n\}$.
-    In \autocite{Li1996-zw} it is proven that there is such a labelling that ensures that there is no shortest path in $\nni$ where the clusters shared between $T$ and $R$ are preserved.}
+    All leaves labelled with dots are labelled by taxa in $\{1, \ldots, n\}$, squares represent taxa in $\{n+1, \ldots, 2n\}$.
+    In \autocite{Li1996-zw} it is proven that there is such a labelling that ensures that there is no shortest path in $\nni$ where the clusters shared between $T$ and $R$ are preserved.
+    Instead, the depicted path where $n$ cherries are built ad then sorted before being resolved is a shortest path}
 	\label{fig:NNI_NP_proof}
@@
 Since the example given in \autocite{Li1996-zw}, which proves Theorem~\ref{thm:split_nni} in $\nni$, does not work in $\rnni$, a contrary statement was conjectured in \autocite{Gavryushkin2018-ol}.
-For understanding this conjecture it is important to realise the connection between edges in a tree and so-called \emph{splits}, which are bipartitions of the taxon set of a tree:
+For understanding this conjecture it is important to know the connection between edges in a tree and so-called \emph{splits}, which are bipartitions of the taxon set of a tree:
 By deleting an edge $e$ of a tree $T$ one receives two graphs with labels in $A$ and $B$, respectively, such that $\{A,B\}$ is a partition of the taxon set of $T$.
@@
 Moreover, a ranked tree can be uniquely determined by ordering the clusters such that the ranks of the internal nodes inducing those clusters increase.
-We already used this representation of trees in Section~\ref{section:algorithms}.
+We already used this representation of trees for the algorithms in Section~\ref{section:algorithms}.
 
@@
 
-\todo{LC: connect previous with following (has been in computation section before)}
-
-%TODO
-%How much detail do we need? Shall we have this as separate section or is it sufficient to mention that we checked the cluster theorem for up to $k$ taxa?
-
-\todo{LC: re-read the two following paragraphs and move them to correct position in paper (maybe some stuff should go to alg section?)}
-
-For providing further evidence that the Cluster Theorem holds in $\rnni$, we can show computationally that it holds for small trees with up to seven taxa.
+For providing further evidence that the Cluster Theorem holds in $\rnni$, we can show computationally that it holds for the $\rnni$ graph on small trees with up to seven taxa.
 \todo{LC: We still need to check 7 taxa.}
 We are doing this in the following way.
-\todo{LC: Move the following paragraph to algorithms section?}
-At first we compute the $\rnni$ graph by using the algorithm introduced in \autocite{Gavryushkin2018-ol}, which we already discussed in Section~\ref{section:algorithms}.
-If the graph is computed, one can get distances by using algorithms such as Floyd-Warshall \autocite{Floyd1962-ew} or Dijkstra\autocite{Dijkstra1959-ph}.
-However, the large number of vertices in $\rnni$ is an obstacle when it comes to computing distances.
-For instance the Floyd-Warshall algorithm needs to save a matrix containing all pairwise distances of vertices in the graph, which needs space quadratic in the number of vertices.
-Since there are $\frac{n!(n-1)!}{2^{n-1}}$ vertices in the $\rnni$ graph \autocite{Gavryushkin2018-ol}, we already run out of space when considering seven taxa.
-Therefore, we cannot simply use this algorithm to compute all distances in $\rnni$.
-Instead of this we compute distances between certain pairs of trees and use the symmetry of $\rnni$ to draw conclusions about the remaining distances.
-
-In the following we describe how the Cluster Theorem has computationally been proven for small trees.
-For this aim only a restricted subset of trees in $\rnni$ needs to be considered, that is trees that share a cluster.
-Furthermore, it is sufficient to consider clusters of the shape $\{1, \ldots, m\}$ for $2 \leq m \leq n-1$.
-This is due to the fact that the distance $d(T_1,R_1)$ and $d(T_2,R_2)$ between two pairs of trees $(T_1,R_1)$ and $(T_2,R_2)$ are equal if one can receive $T_2$ from $T_1$ by permuting the taxa the same way as it is necessary for receiving $R_2$ from $R_1$.
-Therefore, we compute for all $m = 2, \ldots, n-1$ exact $\rnni$ distances between all trees containing the cluster $\{1, \ldots, m\}$, using Dijkstra's algorithm \autocite{Dijkstra1959-ph}.
-Additionally, the subgraph of $\rnni$ induced by the subset of trees that have $\{1, \ldots, m\}$ as cluster is computed.
-\todo{LC: Do we need to define induced subgraph?}
-All pairwise distances between the trees in the subgraph are computed, by using Floyd-Warshall, to compare them with the distances in $\rnni$.
-With this approach it is possible to show that the Cluster Theorem holds for up to seven taxa. \todo{Can we improve this?}
+We focus on the subset of trees that induce share a cluster.
+It is sufficient to consider clusters of the shape $\{1, \ldots, m\}$ for $2 \leq m \leq n-1$, because of the symmetry of the space:.
+distances $d(T_1,R_1)$ and $d(T_2,R_2)$ between two pairs of trees $(T_1,R_1)$ and $(T_2,R_2)$ are equal if one can receive $T_2$ from $T_1$ by permuting the taxa the same way as it is necessary for receiving $R_2$ from $R_1$.
+Therefore, we compute for all $m = 2, \ldots, n-1$ exact $\rnni$ distances between all pairs of trees containing the cluster $\{1, \ldots, m\}$, using Dijkstra's algorithm \autocite{Dijkstra1959-ph} on the $\rnni$ graph, which we can compute as explained in Section~\ref{section:alg_RNNI_graph}.
+Additionally, we compute the subgraph of $\rnni$ induced by the subset of trees that include the cluster $\{1, \ldots, m\}$.
+With the Floyd-Warshall algorithm we compute all pairwise distances between the trees in this subgraph and compare them with the distances in $\rnni$.
+With this approach it is possible to show that the Cluster Theorem holds for up to seven taxa.
+\todo{LC: We still need to check 7 taxa.}
 

### 2019-10-29 paper.tex
proof lemma 1

@@
 	\includegraphics[width=0.8\textwidth]{RNNI}
-    \vspace{2pt}
+    \vspace{12pt}
 	\caption{Two types of operation that define edges of the $\rnni$ graph.
@@
 
+The following lemma shows that $\findpath$ gives a path between the two input trees, indeed, by proving that there is always an $\rnni$ move possible as descibred in Line~\ref{ald:findpath:line:move_set_down} of Algorithm~\ref{alg:find_path}.
+
 \begin{lemma}
-    Let $T$ and $R$ be two trees on which we apply $\findpath$ to compute a path from $T$ to $R$.
-    We assume that $\hat T \neq T,R$ is a tree on the path computed by this algorithm, which needs to be updated in Line~\ref{alg:findpath:line:move_set_down} of Algorithm~\ref{alg:find_path}.
-    Then there is always exactly one $\rnni$ move that can update $\hat T$ to a tree $T'$ such that $\mrca_{T'}(C_i) < \mrca_{\hat T}(C_i)$ for cluster $C_i$ as defined in Line~\ref{alg:find_path:line:cluster}.
+    Let $T$ and $R = [C_1, \ldots, C_{n-2}]$ be two trees and $p:= \findpath (T,R)$ a path from $T$ to $R$.
+    Let $T' \neq T,R$ be a tree on $p$ such that $C_1, \ldots, C_{k-1}$ are induced by internal nodes of $T'$, but $C_k$ is not.
+    Then there is always exactly one $\rnni$ move on $T'$ that results in a tree $T''$ such that $\mrca_{T''}(C_i) < \mrca_{T'}(C_i)$.
     \label{lemma:mrca_move}
@@
 
-    Let $T$ and $R$ be trees and $[C_1, \ldots, C_n]$ the cluster representation
-    \todo{LC: How do we call this representation? Introduce it in the beginning of alg chapter!}
-    of $R$ as introduced above.
-    Let $\hat T$ be a tree as in Line~\ref{alg:findpath:line:move_set_down} of Algorithm~\ref{alg:find_path} and $C_i$ the cluster defined in Line~\ref{alg:find_path:line:cluster}.
-    We want to prove that there is exactly one $\rnni$ move possible on $\hat T$ that results in a tree $T'$ where $\mrca_{T'}(C_i) < \mrca_{\hat T}(C_i)$ if $C_i$ is not induced by a node in $\hat T$ already.
+    Let $T, R$ and $T'$ be trees as defined in the lemma and $[C_1, \ldots, C_{n-2}]$ the cluster representation of $R$.
+    We want to prove that there is exactly one $\rnni$ move on $T'$ that results in a tree $T''$ where $\mrca_{T''}(C_k) < \mrca_{T'}(C_k)$.
+    Notice, that the only $\rnni$ move on $T$ that could decrease the rank of $\mrca_{T'}(C_k)$ are $\nni$ movs on the edge between $\mrca_{T'}(C_k)$ and its child (if this edge has length one), or a rank move between $\mrca_{T'}(C_k)$ and the node with rank one less.
+    Therefore, we will now distinguish these two cases:
+    Either there is an edge $(v,w)$ between nodes $v = \mrca_{T'}(C_k)$ and $w$ with rank $rank_{T'}(w) = rank_{T'}(v) - 1$, or there is no such edge.
 
-    To do this, we distinguish the following two cases:
-    Either there is an edge between the node $v = \mrca_{\hat T}(C_i)$ and the node $w$ with rank $rank_{\hat T}(w) = rank_{hat T}(v) - 1$, or there is no such edge.
+    In the latter case the only $\rnni$ move that could increase the rank of $v$ is a rank move swapping the ranks of $v$ and $w$.
 
-    In the latter case, there is only one $\rnni$ move that could increase the rank of $v$.
-    That is a rank move swapping the ranks of $v$ and $w$.
+    If there is an edge $(v,w)$ in $T'$ there are two $\nni$ moves possible on this edge.
+    We are now going to explain why only one of these two $\nni$ moves changes the rank of this most recent common ancestor.
+    By definition we know that $C_1, \ldots, C_{k-1}$ are induced by the nodes of rank $1, \ldots, k-1$, respectively, in $T'$ already.
+    It is important to notice that $C_k$ either is the disjoint union $C_k = C_i \dot\cup C_j$ for $i,j < k$ or $C_k = C_i \dot\cup \{x\}$ for a taxon $x \in X \setminus (C_1 \cup \ldots \cup C_{k-1})$.
 
-    If there is an edge $(v,w)$ in $\hat T$, there are two $\nni$ moves possible on this edge, which could possibly change the rank of the most recent common ancestor of $C_i$.
-    We are now going to explain why only one of these two $\nni$ moves changes the rank of this most recent common ancestor.
-    Therefore it is important to see that the clusters $C_1, \ldots, C_{i-1}$ already appear in $\hat T$ as they do in $R$, which is a direct result from the algorithm.
-    It is also straight forward to see
-    \todo{LC: is it?}
-    that the $C_i$ either is the disjoint union $C_i = C_j \dot\cup C_k$ for $j,k < i$ or $C_i = C_j \dot\cup \{x\}$ for a taxon $x \in X \setminus (C_1 \cup \ldots \cup C_{i-1})$.
+    Le us at first assume that $C_k = C_i \dot\cup C_j$ for some $i,j < k$.
+    The tree $T'$ looks like the tree depicted in the left of Figure~\ref{fig:mrca_move} where $T_w^1$ and $T_w^2$ are the two subtrees rooted in the children of $w$ and $T_v$ is the subtree whose root is child of $v$ and does not include $w$.
+    Since we know that all clusters $C_1,\ ldots, C_{k-1}$ are induced by some nodes in $T'$, each of these clusters must be a subset of the taxon set of one of the three subtrees $T_w^1,T_w^2$, or $T_v$.
+    In particular, there can't be elements of the same cluster $C_m$ in more than one of these subtrees.
+    This and the fact that $v = \mrca_{T'}(C_i \dot\cup C_j)$ yields that either $C_i$ or $C_j$ is subset of the taxon set of $T_v$ and the other cluster is subset of the taxon set of either $T_w^1$ or $T_w^2$.
 
-    Le us at first assume that $C_i = C_j \dot\cup C_k$ for some $j,k < i$.
-    The tree $\hat T$ looks like the tree depicted in Figure~\ref{fig:mrca_move} where $T_w^1$ and $T_w^2$ are the two subtrees rooted in the children of $w$ and $T_v$ is the subtree whose root is child of $v$ and does not include $w$.
-    Because it is $v = \mrca_{\hat T}(C_i)$, either $C_j$ or $C_k$ is a subset of the taxon set of $T_v$.
-    Without loss of generality we assume that it is $C_j$.
-    Since $C_k$ is already induced by a node in $\hat T$, all taxa in $C_k$ are either in $T_w^1$ or in $T_w^2$.
-    If there were taxa of $C_k$ in both subtrees $T_w^1$ and $T_w^2$, $w$ would be the node inducing $C_k$ and hence it would be $k = i-1$, which means that $v$ induces $C_i$.
-    Therefore, we know that $C_j$ is a subset of the taxon set of $T_v$ and $C_k$ is subset of the taxon set of one of the subtrees $T_w^1$ or $T_w^2$.
-    In Figure~\ref{fig:mrca_move} one can see the two moves possible on $(v,w)$.
-    Depending on which of the subtrees $T_w^1$ or $T_w^2$ includes the taxa of $C_k$, one can see that only one of the two $\nni$ moves results in a tree where the tank of the most recent common ancestor of $C_i$ is below its rank in $\hat T$.
+    Without loss of generality we assume that $C_i$ is included in $T_v$.
+    Otherwise, change the indices $i$ and $j$.
+    Then $C_i$ is a subset of the taxon set of either $T_w^1$ or $T_w^2$.
+    Therefore, just one of the two $\nni$ moves on $(v,w)$, as depicted in Figure~\ref{fig:mrca_move}, moves the most recent common ancestor of $C_k = C_i \dot \cup C_j$ down.
 
-    Now we consider the case that $C_i = C_j \dot\cup \{x\}$ for a taxon $x \in X \setminus (C_1 \cup \ldots \cup C_{i-1})$.
-    For the same reason as above, $C_j$ is in subset of the taxon set of one of the subtrees $T_w^1, T_w^2$ or $T_v$ of $\hat T$ as depicted in Figure~\ref{fig:mrca_move}.
-    Similarly $x$ is taxon of one of these three subtrees in a way that either $C_j$ or $x$ is included in $T_v$, because $v = \mrca_{\hat T}(C_i)$.
+    Now we consider the case that $C_k = C_i \dot\cup \{x\}$ for a taxon $x \in X \setminus (C_1 \cup \ldots \cup C_{k-1})$.
+    For the same reason as above, $C_i$ is in subset of the taxon set of one of the subtrees $T_w^1, T_w^2$ or $T_v$ of $\hat T$ as depicted in Figure~\ref{fig:mrca_move}.
+    Similarly $x$ is taxon of one of these three subtrees in a way that either $C_i$ or $x$ is included in $T_v$, because $v = \mrca_{T'}(C_i \dot \cup \{x\})$.
     So there is just one of the two subtrees $T_w^1$ and $T_w^2$ that neither includes $C_j$ nor $x$.
-    Therefore, the $\nni$ move that moves this subtree to the position where $T_v$ has been before, is the only $\nni$ move that decreases the rank of $\mrca_{\hat T}(C_i)$.
+    As in the previous case, just one of the two $\nni$ moves on $(v,w)$ as illustrated in Figure~\ref{fig:mrca_move} decreases the rank of $\mrca_{T'}(C_i \dot \cup \{x\})$.
 
-    We can conclude that there is in any case exactly one $\rnni$ move that can be performed on $\hat T$ that decreases the rank of $\mrca_{\hat T}(C_i)$ in Line~\ref{alg:findpath:line:move_set_down} of Algorithm~\ref{alg:find_path}.
+    We can conclude that in any case there is exactly one $\rnni$ move possible on $T'$ that decreases the rank of $\mrca_{T'}(C_k)$ in Line~\ref{alg:findpath:line:move_set_down} of Algorithm~\ref{alg:find_path}.
 
@@
     	\includegraphics[width=0.5\textwidth]{mrca_move}
-        \vspace{2pt}
-    	\caption{\todo{caption}}
+        \vspace{12pt}
+        \caption{Two $\nni$ moves that are possible on the edge $(v,w)$ of the tree $T'$.
+        They result in the trees on the right as described in the proof of Lemma~\ref{lemma:mrca_move}.}
     	\label{fig:mrca_move}
@@
 	\includegraphics[width=\textwidth]{NNI_vs_RNNI}
-    \vspace{2pt}
+    \vspace{12pt}
 	\caption{Paths between caterpillar trees $T$ and $R$: The solid paths are paths in $\rnni$, the dashed one is a path that is only possible in the plain $\nni$ graph.
@@
         \end{tabular}
-        \vspace{2pt}
+        \vspace{12pt}
     	\caption{Loop number $k$ of $\csort$ between $T$ and $R$ (in list representation) as described in proof of Theorem~\ref{thm:caterpillar_convex}.
@@
 	\includegraphics[width=\textwidth]{NNI_NP_proof}
-    \vspace{2pt}
+    \vspace{12pt}
     \caption{Example of a shortest path between two trees $T$ and $R$ in $\nni$.
@@
 	\includegraphics[width=\textwidth]{splitthm_counterexample}
-    \vspace{2pt}
+    \vspace{12pt}
 	\caption{The split $123|45$ is present in $T$ and $R$, but the path at the top is a shortest path (computed by $\csort$) where none of the trees contains this split.
@@
 	\includegraphics[width=0.7\textwidth]{findpath_not_symmetric}
-    \vspace{2pt}
+    \vspace{12pt}
 	\caption{Two paths computed by $\findpath$. At the top $\findpath(T,R)$, at the bottom $\findpath(R,T)$}

### 2019-10-30 paper.tex
proof thm6

@@
 
-The following lemma shows that $\findpath$ gives a path between the two input trees, indeed, by proving that there is always an $\rnni$ move possible as descibred in Line~\ref{ald:findpath:line:move_set_down} of Algorithm~\ref{alg:find_path}.
+The following lemma shows that $\findpath$ gives a path between the two input trees, indeed, by proving that there is always an $\rnni$ move possible as described in Line~\ref{alg:findpath:line:move_set_down} of Algorithm~\ref{alg:find_path}.
 
@@
     According to Lemma~\ref{lemma:distance_delete_taxon}, the distance between $T{\big|}_n$ and $R{\big|}_n$ is at least $d(T{\big|}_n, R{\big|}_n) \leq d(T,R) - |\rank_T(\parent_T(n+1)) - \rank_R(\parent_R(n+1))|$.
-    We also know $|\rank_T(\parent_T(n+1)) - \rank_R(\pare nt_R(n+1))| = n-1$.
+    We also know $|\rank_T(\parent_T(n+1)) - \rank_R(\parent_R(n+1))| = n-1$.
     Suppose, contrary to our claim, that there is a path from $T$ to $R$ of length less than $\frac{n(n-1)}{2}$.
@@
 \begin{proof}
-    We prove this Theorem by backwards induction on the length $l$ of a shortest caterpillar path between two caterpillar trees $T$ and $R$.
+    We prove this Theorem by backwards induction on the caterpillar distance $l = d_c(T,R)$ between caterpillar trees $T$ and $R$.
     For this backwards induction the maximum caterpillar distance, which is $\frac{(n-1)(n-2)}{2}$ (Lemma~\ref{lemma:caterpillar_dist=diameter}), is the base case.
@@
     With Lemma~\ref{lemma:caterpillar_dist=diameter} one can see that for each pair of caterpillar trees $T$, $R$ with caterpillar distance $d_c(T,R) = \frac{(n-1)(n-2)}{2}$, it is $d(T,R) = d_c(T,R)$.
-    Therefore, there is a shortest path between $T$ and $R$ that is a caterpillar path.
+    This means that there is a shortest path between $T$ and $R$ that is a caterpillar path.
 
     Induction step: $l+1 \to l$\\
-    Induction hypothesis: For each pair of caterpillar trees with caterpillar distance $l+1$ there is a shortest path between these trees that only consists of caterpillar trees, which means that the distance between these trees is $l+1$.
+    Induction hypothesis: For each pair of caterpillar trees with caterpillar distance $l+1$ there is a shortest path between these trees that consists of caterpillar trees only, so the distance between these trees is $l+1$.
 
     Let $T$ and $R$ be caterpillar trees with caterpillar distance $l$.
-    We consider the shortest caterpillar path from $T$ to $R$ computed by $\csort$.
-    Since $l < \frac{(n-1)(n-2)}{2}$, there must be one loop $k$ in $\csort$ where less than $n-2-k$ taxon pairs swap positions.
-    Let us assume that loop number $k$ is the first one where this happens.
-    This situation is illustrated in the top of Figure~\ref{fig:cat_convex_seq_example}.
-    Whenever the number of swaps of taxa or, in terms of lists, neighboured elements in the list is maximum in a loop, one of the first two elements swaps position with $n-1-k$ subsequent elements of the list.
-    We already saw this in the proof of Lemma~\ref{lemma:caterpillar_dist=diameter}.
-    Since we assume that the number of swaps in loop $k$ is not maximal, there are less than $n-1-k$ such moves.
-    \todo{LC: From here on it is confusing}
-    Let us assume that $a$ is the first element that is compared with its left neighbour in loop $k$ but does not swap with it.
-    Then the left neighbour $b$ of $a$ in $T$ is below $a$ in $R$ as well.
-    Therefore, the tree $T'$ which can be received from $T$ by exchanging $a$ and $b$ must have caterpillar distance $d_c(T',R) = d_c(T,R) + 1 = l + 1$ to $R$.
-    For an illustration of the $k^{th}$ loop when performing $\csort$ between $T$ and $R$ or $T'$ and $R$, respectively, consider Figure~\ref{fig:cat_convex_seq_example}.
+    We consider the shortest caterpillar path between these trees computed by $\csort$.
+    From now on we will consider the trees $T$ and $R$, as well as all other caterpillar trees within this proof, in their list representation $L_T$ and $L_R$ as introduced in Section~\ref{section:alg_csort}.
 
-    \todo{LC: replace this figure}
-    \begin{figure}[H]
-    	\centering
-        \begin{tabular}{|c|c|c|c|c|c|c|c|c|c|c|c|}
-            \hline
-            $x_{k+1}$ & $x_k$ & $\ldots$ & $b$ & $a$ & \multicolumn{3}{c|}{$\ldots$} & $x_n$ & $x_{k-1}$ & $\ldots$ & $x_1$\\
-            \hline
-            \multicolumn{12}{c}{}\\
-            \hline
-            $x_{k+1}$ & $x_{k+2}$ & $\ldots$ & $b$ & $x_k$ & $\ldots$ & $a$ & $\ldots$ & $x_n$ & $x_{k-1}$ & $\ldots$ & $x_1$\\
-            \hline
-            \multicolumn{12}{c}{}\\
-            \multicolumn{12}{c}{}\\
-            \hline
-            $x_{k+1}$ & $x_k$ & $\ldots$ & $a$ & $b$ & \multicolumn{3}{c|}{$\ldots$} & $x_n$ & $x_{k-1}$ & $\ldots$ & $x_1$\\
-            \hline
-            \multicolumn{12}{c}{}\\
-            \hline
-            $x_{k+1}$ & $x_{k+2}$ & $\ldots$ & $x_k$ & $b$ & $\ldots$ & $a$ & $\ldots$ & $x_n$ & $x_{k-1}$ & $\ldots$ & $x_1$\\
-            \hline
+    Since $l < \frac{(n-1)(n-2)}{2}$, there must be one of the loops starting in Line~\ref{alg:csort:line:loop} of Algorithm~\ref{alg:csort} ($\csort$) where not every considered pair of taxa swaps positions.
+    Let us assume that loop number $k$ is the first one where the number of swaps is not maximal.
+    It might well be that this happens in the first loop $k=1$ already.
+    Since we assume that the number of swaps in loop $k$ is not maximal, there is a pair $a,b$ of elements that is compared in the algorithm but does not swap positions.
+    And because we consider the first time that this happens, $a$ is either one of the first two elements in the sequence, or it previously swapped with an element $c$ that has been between $a$ and $b$ before loop number $k$.
+
+    Let us first consider the latter case.
+    The sequence $L_{T'}$ at the time where $a$ and $b$ are compared but not swapped, looks as follows:
+
+    \vspace{12pt}
+    \begin{tabular}{c|c|c|c|c|c|c|c|c|}
+        \cline{2-9}
+        $L_{T'}$: & $\ldots$ & $c$ & $a$ & $b$ & $\ldots$& $x_{k-1}$ & $\ldots$ & $x_1$\\
+        \cline{2-9}
     \end{tabular}
     \vspace{12pt}
-    	\caption{Loop number $k$ of $\csort$ between $T$ and $R$ (in list representation) as described in proof of Theorem~\ref{thm:caterpillar_convex}.
-        The trees are represented as sequences that are being sorted by $\csort$.
-        Right before loop number $k$, taxa $x_1, \ldots, x_{k-1}$ are already at the same position as in $R$.
-        At the top: sequences between $T$ and $R$ -- $x_k$ and $b$ swap in loop number $k$, but $x_k$ and $a$ not.
-        At the bottom: sequences between $T'$ and $R$ -- $a$ and $b$ swap in loop number $k$; $x_k$ and $b$ will swap later}
-    	\label{fig:cat_convex_seq_example}
-    \end{figure}
 
-    Because $T'$ has caterpillar distance $d_c(T',R) = l+1$ to $T$, we can apply the induction hypothesis on $T'$ and $R$ and see that $d(T',R) = l+1$.
-    Furthermore, it is $d(T,T') = 1$.
-    If it now were $d(T,R) < l$, it would hold $d(T',R) \leq d(T,R) + d(T,T') < l + 1$ which contradicts the induction hypothesis.
-    We can conclude $d(T,R) = l$, which completes the proof.
+    Note that in this illustration of $L_{T'}$ the elements $x_1, \ldots, x_{k-1}$ are in the same place as they are in $L_R$.
+    However, it could be that $k=1$, which means that there are no such elements at the end of $L_{T'}$ that are already sorted the same way as in the destination tree.
+
+    The last move before comparing $a$ and $b$ is a swap of $a$ with an element $c$.
+    Since $a$ and $c$ swapped, but $a$ and $b$ did not, we can follow that $b$ and $c$ appear in $L_R$ in the same order as they do in $L_{T'}$.
+    Furthermore, $c$ and $b$ must have been neighbours in $L_T$, because we assume that the comparison of $a$ and $b$ is the first one where the compared elements are not swapped in the sequence.
+    Let now $L_{\hat T}$ be a list that equals $L_T$ with the only exception that $b$ and $c$ are swapped.
+    We can follow from our observations above, that the distance between $T$ and $\hat T$ is $d(T, \hat T) = 1$ and that $\csort$ needs one move more for a path from $\hat T$ to $R$ than from $T$ to $R$.
+    This results in $d_c(\hat T, R) = d(T,R) + 1$.
+
+    So we can apply the induction hypothesis on $\hat T$ and $R$ and see that $d(\hat T,R) = l+1$.
+    If it now were $d(T,R) < l$, it would hold $d(\hat T,R) \leq d(T,R) + d(T,\hat T) < l + 1$ which contradicts the induction hypothesis.
+    We can conclude $d(T,R) = l$, which completes the proof for the case that $a$ is not one of the first two elements of $L_{T'}$.
+
+    Let us now consider the other case:
+    of the two elements $a$ and $c$, that are compared but not swapped, one, say $a$, is either the first or the second one in $L_{T'}$.
+    Then the list looks like this:
+
+    \vspace{12pt}
+    \begin{tabular}{c|c|c|c|c|c|c|c|c|}
+        \cline{2-8}
+        $L_{T'}$: & $a$ & $b$ & $c$ & $\ldots$ & $x_{k-1}$ & $\ldots$ & $x_1$\\
+        \cline{2-8}
+    \end{tabular}
+    \vspace{12pt}
+
+    If there is no swap of either $a$ or $b$ with $c$, we can construct a tree $\hat T$ with caterpillar distance $d_c(\hat T, R) = l+1$ and $d(T,\hat T) = 1$ in a similar way as in the previous case.
+    In the list $L_T$ representing the start tree $T$, either $a$ or $b$ must be neighbour of $c$.
+    This is due to the fact that we are currently considering the first comparison of elements where there is no swap.
+    Without loss of generality we assume that $a$ is neighbour of $c$ in $L_T$.
+    Then an exchange of $a$ and $c$ in $L_T$ yields a list $L_{\hat T}$ representing a tree $\hat T$ with the desired properties.
+    As in the previous case we can then apply the induction hypothesis on $\hat T$ and conclude that there is a shortest path between $T$ and $R$ that only consists of caterpillar trees.
 \end{proof}

### 2019-10-30 paper.tex
minor

@@
 
-\todo{Add a bit more space in all figures at the bottom to have at least a standard interval above he caption.}
-\todo{In this figure -- rm arrows (think how to place the RNNI edges), possibly avoid using colour, and highlight where the rank move is made (hollow nodes?)}
-\todo{Two of these trees are caterpillar trees.
-Do you think it would be helpful to draw them as caterpillars to prepare readers for the rest of the paper?}
 \begin{figure}[H]
@@
 Afterwards, we introduce two algorithms, namely $\findpath$ (Section~\ref{section:alg_findpath}) and $\csort$ (Section~\ref{section:alg_csort}), for computing paths between two given trees in $\rnni$.
-The first one takes arbitrary trees as input while the latter one is restricted to trees with a certain tree shape.
-We will use them later to prove some geometric properties of the $\rnni$ graph in section~\ref{section:geometry}.
+The first one takes arbitrary trees as input while the latter one is restricted to trees with a certain tree topology.
+\todo{LC: Do we need to define tree topology?}
+We will use them later to prove some geometric properties of the $\rnni$ graph in Section~\ref{section:geometry}.
 Additionally we will present the algorithm $\mdtree$, which produces a tree that is far away from the given input tree.
@@
 
-    Le us at first assume that $C_k = C_i \dot\cup C_j$ for some $i,j < k$.
+    Let us at first assume that $C_k = C_i \dot\cup C_j$ for some $i,j < k$.
     The tree $T'$ looks like the tree depicted in the left of Figure~\ref{fig:mrca_move} where $T_w^1$ and $T_w^2$ are the two subtrees rooted in the children of $w$ and $T_v$ is the subtree whose root is child of $v$ and does not include $w$.
@@
 
-\todo{LC: Shall we give an example where the distance is not max or simulate trees and compute histogram of distances approximated by $\findpath$?}
-
 \begin{algorithm}[H]
@@
 In particular, we are going to prove that one can always find a shortest path between two caterpillar trees that contains caterpillar trees only, meaning that the set of caterpillar trees is convex in $\rnni$.
-We discovered this by computing caterpillar paths with $\csort$ and comparing them with shortest path in $\rnni$, computed by the exact algorithm as explained in Section~\ref{section:algorithms}.
+We discovered this by computing caterpillar paths with $\csort$ and comparing them with shortest path in $\rnni$, computed by the exact algorithm as explained in Section~\ref{section:alg_RNNI_graph}.
 \todo{LC: Do we explain how we did this in the alg section? If not, we need to add an explanation here!}
@@
 
+Within this chapter we will again use the list representation for caterpillar trees, which is a simplified version of the cluster representation of trees, as explained in Section~\ref{section:alg_csort}.
+
 \begin{lemma}
@@
     There exist three trees on three taxa: $[1,2,3], [1,3,2]$ and $[2,3,1]$.
-    \todo{LC: Do we use list representation here? If so, we need to mention it}
     They all have distance one from another and are caterpillar trees.
@@
     In particular, taxon $n+1$ exchanges with a taxon in each of the $n+1$ loops in $\csort$.
-    \todo{LC: Is this explanation sufficient?}
     We can conclude that taxon $n+1$, whose parent has highest rank in $T$, is part of the cherry of $R$.
@@
 \end{theorem}
-
-\todo{LC: Work on the following proof, especially the figure and explanation of it}
 

### 2019-11-03 rnni_convexity.tex
rename files
add citations + minor edits

--- 2019-10-30 paper.tex
@@
 \renewbibmacro{in:}{}
-\addbibresource{lit.bib}
+\addbibresource{rnni_convexity.bib}
 
@@
 
-The nearest neighbour interchange ($\nni$) graph, defined on the set of phylogenetic trees with adjacency relation given by the interchange operation of two sister clades, has been known in mathematical biology literature for nearly 50 years \autocite{Robinson1971, Moore-Goodman-Barnabas1973}.
+The nearest neighbour interchange ($\nni$) graph, defined on the set of phylogenetic trees with adjacency relation given by the interchange operation of two sister clades, has been known in mathematical biology literature for nearly 50 years \autocite{Robinson1971-ql,Moore1973-kk}.
 Considered with the metric given by the length of a shortest path (graph-distance), this graph becomes a metric space.
-Its geometry has been extensively studied \autocite{...}.
-An important property of the $\nni$ graph is that computing distance is NP-hard \autocite{...}.
-As a result no algorithm exists to compute the distance in practical time \autocite{...Whidden may be}.
+Its geometry has been extensively studied \autocite{Dasgupta2000-xa, Li1996-zw, Gordon2013-fw, De_Jong2016-al}.
+An important property of the $\nni$ graph is that computing distance is NP-hard \autocite{Dasgupta2000-xa}.
+As a result no algorithm exists to compute the distance in practical time.
 A consequence is that tree search and sampling algorithms pose a significant challenge even for moderately sized trees \autocite{...}.
 
-More recent advances in computational phylogenetics introduce various classes of molecular clock models \autocite{...} and made possible computational inference of phylogenetic time-trees \autocite{...}.
+More recent advances in computational phylogenetics introduce various classes of molecular clock models \autocite{Yoder2000-ks,Drummond2006-nl,Drummond2010-yf} and made possible computational inference of phylogenetic time-trees \autocite{...}.
 However, mathematical challenges that come with this seemingly inessential change in parametrisation (genomic distance VS time distance) of trees have only recently been brought to attention \autocite{Gavryushkin2016-uu}.
@@
 
-\todo{LC: mention some results here and how they make $\rnni$ differ from $\nni$}
-This distinguishes $\rnni$ from $\nni$ where this does not hold.
-One further result, presented in Section~\ref{section:diameter}, directly following from the proof of Theorem~\ref{thm:caterpillar_convex}, is the diameter of $\rnni$.
-Finally, we disprove the so-called Strong Split Theorem, conjectured in \autocite{Gavryushkin2018-ol}.
-However, the Weak Split Theorem remains a conjecture.
-Furthermore, we propose an alternative (even weaker) version of the conjecture, the Cluster Theorem (Section~\ref{section:cluster_theorem}).
-
-In this paper, we consider the $\rnni$ space on the ranked phylogenetic trees with all taxa being of equal rank.
-An example of such a tree is depicted in Figure~\ref{fig:...}.
+In this paper, we consider the $\rnni$ space on ranked phylogenetic trees with all taxa being of equal rank.
+An example of such a tree is depicted in Figure~\ref{fig:ranked_tree}.
 In the terminology of \autocite{Gavryushkin2018-ol}, the space considered in this paper is the space of ranked ultrametric phylogenetic trees $\rnniu$.
@@
 
-[Figure of ranked tree]
+\begin{figure}[H]
+	\centering
+	\includegraphics[width=0.25\textwidth]{ranked_tree}
+    \vspace{12pt}
+	\caption{A ranked tree with $6$ taxa.}
+	\label{fig:ranked_tree}
+\end{figure}
 
 In line with the research programme proposed in \autocite{Gavryushkin2018-ol}, we investigate the geometry and algorithmic complexity of the $\rnni$ space.
-Specifically, in this paper we establish the exact radius and diameter of the space, [...
--- Lena, please finish this including the approximate algorithm that works exactly on small spaces (Lena: is it already in the text?)]
+Specifically, in this paper we establish the exact radius and diameter of the space.
+We also show that the subset of caterpillar trees is convex in $\rnni$.
+Moreover, we can provide an approximate algorithm that computes exact distances for small trees.
 The question of whether there exists a polynomial algorithm for computing the $\rnni$ distance remains an open problem.
@@
 For a tree $T$, we will use $\rank_T$ to refer to its ranking.
+A ranked tree $T$ without its leaf labels will be called \emph{tree topology}.
 
@@
 The first one takes arbitrary trees as input while the latter one is restricted to trees with a certain tree topology.
-\todo{LC: Do we need to define tree topology?}
 We will use them later to prove some geometric properties of the $\rnni$ graph in Section~\ref{section:geometry}.
@@
 \todo{LC: github - link}
-For computing the actual $\rnni$ distances we used the the algorithm for computing the $\rnni$ graph as described in Section~\ref{section:alg_RNNI_graph}.
-\todo{LC: How do we compare the distances?}
+For computing the actual $\rnni$ distances we used the algorithm for computing the $\rnni$ graph as described in Section~\ref{section:alg_RNNI_graph}.
 
@@
 
-\todo{LC: We don't use the following Proposition anywhere in the paper. Do we want to keep it?}
-
-Using the same argument as in the proof of Lemma~\ref{lemma:distance_delete_taxon} we can establish Proposition~\ref{proposition:lower_bound_distance} that provides a lower bound on the distance between trees.
-
-\begin{proposition}
-If trees $T$ and $R$ and taxon $x$ are such that $|\rank_T(\parent_T(x)) - \rank_R(\parent_R(x))| = \delta$ then $d(T,R) \geq \delta$.
-\label{proposition:lower_bound_distance}
-\end{proposition}
-
 \subsection{The set of caterpillar trees}
@@
 In particular, we are going to prove that one can always find a shortest path between two caterpillar trees that contains caterpillar trees only, meaning that the set of caterpillar trees is convex in $\rnni$.
-We discovered this by computing caterpillar paths with $\csort$ and comparing them with shortest path in $\rnni$, computed by the exact algorithm as explained in Section~\ref{section:alg_RNNI_graph}.
-\todo{LC: Do we explain how we did this in the alg section? If not, we need to add an explanation here!}
+We discovered this by computing caterpillar paths with $\csort$ and comparing them with shortest path in $\rnni$, computed by the exact algorithm discussed in Section~\ref{section:alg_RNNI_graph}.
 In the following we demonstrate why finding shortest paths between caterpillar trees builds the basis for understanding the difference between $\nni$ and $\rnni$.
@@
 In Section~\ref{alg:max_dist_tree} we already discussed that $\mdtree$ is a non-deterministic algorithm.
-In fact, the output tree of $\mdtree$ can have any tree topology
-\todo{LC: Do we need to define 'tree topology'?}
-, since there is no restriction on where to add the new taxon in the already existing tree in Line~\ref{alg:mdtree_lineAddTaxon} of Algorithm~\ref{alg:max_dist_tree}.
+In fact, the output tree of $\mdtree$ can have any tree topology, since there is no restriction on where to add the new taxon in the already existing tree in Line~\ref{alg:mdtree_lineAddTaxon} of Algorithm~\ref{alg:max_dist_tree}.
 This means that for each caterpillar input tree $\mdtree$ can compute a tree with any tree topology that has maximum distance from the input tree.
@@
 \section{Ideas}
-\todo{This section is just a rough draft. Are there ideas we want to keep? Shall this become an 'open problems/future work' section?}
+
+[This section includes some ideas that we could possibly include in the paper]
+
+Using the same argument as in the proof of Lemma~\ref{lemma:distance_delete_taxon} we can establish Proposition~\ref{proposition:lower_bound_distance} that provides a lower bound on the distance between trees.
+
+\begin{proposition}
+If trees $T$ and $R$ and taxon $x$ are such that $|\rank_T(\parent_T(x)) - \rank_R(\parent_R(x))| = \delta$ then $d(T,R) \geq \delta$.
+\label{proposition:lower_bound_distance}
+\end{proposition}
 

### 2019-11-04 rnni_convexity.tex
pass through to Lemma 1 + rm unnecessary space (and possibly some necessary ones)

@@
 \usepackage[hidelinks]{hyperref}
+\usepackage{csquotes}
 \usepackage[style=authoryear,ibidtracker=false,uniquename=false,giveninits=true,terseinits=true,maxbibnames=5,backend=biber]{biblatex}
@@
 
-\maketitle
-
 \begin{abstract}
@@
 \end{abstract}
+
+\maketitle
 
@@
 As a result no algorithm exists to compute the distance in practical time.
-A consequence is that tree search and sampling algorithms pose a significant challenge even for moderately sized trees \autocite{...}.
+A consequence is that tree search and sampling algorithms pose a significant challenge even for moderately sized trees
+\todo{cite Chris Whidden}
+\autocite{}.
 
-More recent advances in computational phylogenetics introduce various classes of molecular clock models \autocite{Yoder2000-ks,Drummond2006-nl,Drummond2010-yf} and made possible computational inference of phylogenetic time-trees \autocite{...}.
+More recent advances in computational phylogenetics introduce various classes of molecular clock models \autocite{Yoder2000-ks,Drummond2006-nl,Drummond2010-yf} and made possible computational inference of phylogenetic time-trees
+\todo{cite MrBayes, BEAST, NextStrain}
+\autocite{}.
 However, mathematical challenges that come with this seemingly inessential change in parametrisation (genomic distance VS time distance) of trees have only recently been brought to attention \autocite{Gavryushkin2016-uu}.
-\textcite{Gavryushkin2018-ol} have proposed an extension of the $\nni$ graph to the class of discrete time-trees.
+These differences motivated \textcite{Gavryushkin2018-ol} to propose an extension of the $\nni$ graph to the class of discrete time-trees.
 The simplest such extension introduces the $\rnni$ graph on the set of ranked phylogenetic trees.
@@
 Specifically, in this paper we establish the exact radius and diameter of the space.
-We also show that the subset of caterpillar trees is convex in $\rnni$.
-Moreover, we can provide an approximate algorithm that computes exact distances for small trees.
+We also show that the subset of caterpillar trees is convex in $\rnni$, thus settling one of the open problems proposed in \autocite{Gavryushkin2018-ol}.
+Furthermore, we provide an approximate algorithm that computes exact distances for small trees.
 The question of whether there exists a polynomial algorithm for computing the $\rnni$ distance remains an open problem.
 
-In the rest of this chapter we formally introduce non-standard definitions necessary for this paper.
+In the rest of this chapter we formally introduce the terminology used in this paper.
 
-A \emph{ranked phylogenetic tree} is a pair consisting of a rooted binary phylogenetic tree $T$ on the set $X = \{1, \ldots, n\}$ of \emph{taxa} for $n \in \mathbb N$, and a (total) rank function $\rank$ that maps all leaves of $T$ to $0$, all internal nodes of $T$ onto the set $\{1, \ldots, n-1\}$, and respects the partial order given by the tree.
+A \emph{ranked phylogenetic tree} is a pair consisting of a rooted binary phylogenetic tree $T$ on the set $X = \{1, \ldots, n\}$ of \emph{taxa} for $n \in \mathbb N$, and a (total) rank function that maps all leaves of $T$ to $0$, all internal nodes of $T$ onto the set $\{1, \ldots, n-1\}$, and respects the partial order on the nodes given by the tree.
 The latter means that if $u$ and $v$ are two internal nodes of $T$ such that there exists a path from a taxon $x \in X$ to the root which first passes through $u$ and then through $v$ then $\rank(u) < \rank(v)$.
 Ranked trees $(T_1, \rank_1)$ and $(T_2, \rank_2)$ are different if trees $T_1$ and $T_2$ are different or $T_1 = T_2$ and $\rank_1 \neq \rank_2$.
-Since all trees in this paper are ranked trees, we will abuse the notation and drop the ranking from the notation.
-We will also refer to a ranked tree simply by tree.
+Since all trees in this paper are ranked trees, we will abuse the notation and drop the rank function from the notation.
+We will also be simply saying tress to mean ranked phylogenetic trees.
 For a tree $T$, we will use $\rank_T$ to refer to its ranking.
@@
 
-Our definition of a ranked tree implies a natural notion of the edge length.:
-A tree edge $(u,v)$ of $T$ has length $l$ if $|\rank_T(v) - \rank_T(u)| = l$.
+Our definition of a ranked tree implies a natural notion of the edge length -- we call $|\rank_T(v) - \rank_T(u)|$ the length of an edge $(u, v)$.
 
@@
 
-\todo{LC: Explain structure of paper}
+\todo[inline]{LC: Explain structure of paper}
 
@@
 
-Within this section we will at first described how we can compute exact distances between trees in the $\rnni$ graph (Section~\ref{section:alg_RNNI_graph}) by using results of \autocite{Gavryushkin2018-ol}.
-Afterwards, we introduce two algorithms, namely $\findpath$ (Section~\ref{section:alg_findpath}) and $\csort$ (Section~\ref{section:alg_csort}), for computing paths between two given trees in $\rnni$.
-The first one takes arbitrary trees as input while the latter one is restricted to trees with a certain tree topology.
-We will use them later to prove some geometric properties of the $\rnni$ graph in Section~\ref{section:geometry}.
-Additionally we will present the algorithm $\mdtree$, which produces a tree that is far away from the given input tree.
-What we mean by far away depends on the input tree and will be discussed after introducing this algorithm in Section~\ref{section:alg_mdtree}.
+We start this section with a discussion of the algorithm proposed in \autocite{Gavryushkin2018-ol} to compute the exact $\rnni$ distance between trees by pre-computing the $\rnni$ graph.
+Then we introduce two algorithms, $\findpath$ (Section~\ref{section:alg_findpath}) and $\csort$ (Section~\ref{section:alg_csort}), for computing paths between two given trees in $\rnni$.
+The first of these algorithms takes arbitrary trees as an input while the latter is restricted to trees with a certain tree topology.
+Both of these algorithms will become an important ingredient to establish several geometric properties of the $\rnni$ graph in Section~\ref{section:geometry}.
+Finally we will present the algorithm $\mdtree$, which produces a tree that is far away from the given input tree (see Section~\ref{section:alg_mdtree} for details).
+While this algorithm is interesting on its own (e.g.\ for simulation studies), we will again apply the algorithm to study the geometry of the $\rnni$ space.
 
@@
 
-In Section 3.3 of \autocite{Gavryushkin2018-ol} an exact algorithm for computing the $\rnni$ graph has been introduced.
+\textcite[Section 3.3]{Gavryushkin2018-ol} constructed an exact algorithm for computing the $\rnni$ graph.
+\todo{Do we make use of tanglegrams?}
+\todo{GitHub link?}
 We implemented this algorithm and used it for computing this graph for trees on up to seven taxa.
@@
 Since there are $\frac{n!(n-1)!}{2^{n-1}}$ vertices in the $\rnni$ graph \autocite{Gavryushkin2018-ol}, we already run out of space when we want to save all pairwise distances for trees on seven taxa.
+\todo{Is this the hint we're using tanglegrams?
+Either way this statement is too vague and requires more detail.
+Also, ``run out of space'' should be a bit less frank.
+Finally, think whether you want single out this paragraph as a special subsection?
+Either way you should say why you need this implementation.}
 Therefore, we compute distances between certain pairs of trees and use the symmetry of $\rnni$ to draw conclusions about the remaining distances.
@@
 \emph{Trivial clusters} are clusters induced by leaves (contain just one element each) and the root of the tree (contains all taxa).
+As all trees on $n$ taxa share all trivial clusters, we will use the notion cluster to mean non-trivial clusters.
 Note that the list of all non-trivial clusters of a tree, sorted according to the rank of the corresponding node, defines the tree unambiguously.
-For the rest of this paper we will use the notion cluster and mean non-trivial clusters only because all trees on $n$ taxa share the trivial clusters.
-We will refer to this list of clusters $n-2$ of a given tree on $n$ taxa as \emph{cluster representation} of the tree.
+We will refer to this list of $n-2$ clusters for a given tree on $n$ taxa as a \emph{cluster representation} of the tree.
 
-We are now presenting $\findpath$ (Algorithm~\ref{alg:find_path}), an algorithm for computing paths between any two given trees in $\rnni$.
-Let $[C_1, \ldots, C_{n-2}]$ be the cluster representation of $R$ (the destination tree).
-Then $C_i$ is induced by the node with rank $i$ for $i = 1, \ldots, n-2$.
+We are now ready to introduce $\findpath$ (Algorithm~\ref{alg:find_path}), an efficient algorithm for computing paths between any two given trees $T$ and $R$ in $\rnni$.
+Let $[C_1, \ldots, C_{n-2}]$ be the cluster representation of $R$ (the ``destination'' tree).
+Then $C_i$ is induced by the node of rank $i$ for $i = 1, \ldots, n-2$.
 $\findpath$ computes a path $p$ which initially consists of $T$ only and is extended stepwise.
-In order to do this, the clusters $C_1, \ldots, C_{n-2}$ of $R$ are considered in that order.
+In order to do this, the clusters $C_1, \ldots, C_{n-2}$ of $R$ are considered in their order.
 Let $T'$ be the last tree of $p$ when arriving at cluster $C_k$.
@@
 Now perform an $\rnni$ move that decreases the rank of $\mrca_{T'}(C_k)$.
-With Lemma~\ref{lemma:mrca_move} we know that such an $\rnni$ move always exists if long as $T'$ does not induce the cluster $C_k$.
-So we can perform such moves until a tree $T''$ with $\mrca_{T''}(C_k)$ is reached that contains a node inducing $C_k$ that has the same rank $k$ in $T''$ as in $R$.
+\todo{We haven't seen Lemma 1 yet.}
+With Lemma~\ref{lemma:mrca_move} we know that such an $\rnni$ move always exists as long as $T'$ does not induce the cluster $C_k$.
+So we can perform such moves until a tree $T''$ is reached with $\mrca_{T''}(C_k)$ inducing $C_k$ and having has the same rank $k$ in $T''$ as in $R$.
 Appending all trees on this path from $T'$ to $T''$ and repeating this procedure for all clusters of $R$ results in a path $p$ from $T$ to $R$.
@@
 \begin{proof}
-
 Let $T, R$ and $T'$ be trees as defined in the lemma and $[C_1, \ldots, C_{n-2}]$ the cluster representation of $R$.

### 2019-11-04 rnni_convexity.tex
better reference to Lemma 1

@@
 Now perform an $\rnni$ move that decreases the rank of $\mrca_{T'}(C_k)$.
-\todo{We haven't seen Lemma 1 yet.}
-With Lemma~\ref{lemma:mrca_move} we know that such an $\rnni$ move always exists as long as $T'$ does not induce the cluster $C_k$.
+The existence of such a move is not obvious but we will see later (Lemma~\ref{lemma:mrca_move}) that this kind of move always exists as long as $T'$ does not induce the cluster $C_k$.
 So we can perform such moves until a tree $T''$ is reached with $\mrca_{T''}(C_k)$ inducing $C_k$ and having has the same rank $k$ in $T''$ as in $R$.

### 2019-11-04 rnni_convexity.tex
Lemma 1 some tweeks + confused about the statement

@@
 \begin{lemma}
-Let $T$ and $R = [C_1, \ldots, C_{n-2}]$ be two trees and $p:= \findpath (T,R)$ a path from $T$ to $R$.
-Let $T' \neq T,R$ be a tree on $p$ such that $C_1, \ldots, C_{k-1}$ are induced by internal nodes of $T'$, but $C_k$ is not.
-Then there is always exactly one $\rnni$ move on $T'$ that results in a tree $T''$ such that $\mrca_{T''}(C_i) < \mrca_{T'}(C_i)$.
+Let $T$ and $R = [C_1, \ldots, C_{n-2}]$ be two trees and $p = \findpath (T,R)$.
+Let $T'$ be a tree on $p$ different from both $T$ and $R$ such that $C_1, \ldots, C_{k-1}$ are induced by internal nodes of $T'$, but $C_k$ is not.
+Then there exists exactly one $\rnni$ move on $T'$ that results in a tree $T''$ such that
+\todo{What's $i$?}
+$\mrca_{T''}(C_i) < \mrca_{T'}(C_i)$.
 \label{lemma:mrca_move}

### 2019-11-07 rnni_convexity.tex
pass to first mention of Bubble Sort

@@
 
+\todo{I suggest to rewrite this paragraph following the standard structure:\\
+\emph{Step $0$. $T_0 = T$, ...\\
+Step $k+1$. $T_{k+1}$ is defined...}}
 We are now ready to introduce $\findpath$ (Algorithm~\ref{alg:find_path}), an efficient algorithm for computing paths between any two given trees $T$ and $R$ in $\rnni$.
@@
 In order to do this, the clusters $C_1, \ldots, C_{n-2}$ of $R$ are considered in their order.
-Let $T'$ be the last tree of $p$ when arriving at cluster $C_k$.
+Let $T'$ be the last tree of $p$ when
+\todo{This term is not clear or defined.}
+arriving at cluster $C_k$.
 We assume that the clusters $C_1, \ldots, C_{k-1}$ are induced by nodes of rank $1, \ldots, k-1$ in $T'$, respectively.
@@
 Now perform an $\rnni$ move that decreases the rank of $\mrca_{T'}(C_k)$.
+\todo{I think it is obvious.
+I suggest to remove this Lemma and save it somewhere in case we get an too sensitive referee.
+Tag a commit?
+Or branch? -- whatever works.}
 The existence of such a move is not obvious but we will see later (Lemma~\ref{lemma:mrca_move}) that this kind of move always exists as long as $T'$ does not induce the cluster $C_k$.
@@
 
+\todo{$p = p+T'$: Do we really want to be updating this in every while loop?}
 \begin{algorithm}[H]
@@
 \begin{lemma}
-Let $T$ and $R = [C_1, \ldots, C_{n-2}]$ be two trees and $p = \findpath (T,R)$.
+Let $T$ and $R = [C_1, \ldots, C_{n-2}]$ be two trees and
+\todo{This lemma is to prove that FindPath is an (deterministic) algorithm, so we cannot relay on this.}
+$p = \findpath (T,R)$.
 Let $T'$ be a tree on $p$ different from both $T$ and $R$ such that $C_1, \ldots, C_{k-1}$ are induced by internal nodes of $T'$, but $C_k$ is not.
@@
 
-With Lemma~\ref{lemma:mrca_move} we can conclude that $\findpath$ is a deterministic algorithm.
-Its running time is quadratic in the number of taxa.
-Since it computes a path between two arbitrary trees in $\rnni$, we can use the length of such a path as approximation for the distance between the trees.
-A natural question is how well this estimates the actual distance in $\rnni$.
-To answer this question for small trees, we implemented $\findpath$.
+Note that the uniqueness of the $\rnni$ move in Line~\ref{alg:findpath:line:move_set_down} of Algorithm~\ref{alg:find_path} implies that the algorithm is deterministic.
+The running time of the algorithm is quadratic in the number of taxa.
+
+Because the algorithm returns a path between pairs of trees in $\rnni$, the length of the path approximates the $\rnni$ distance from above.
+A natural question then is how accurate this approximation is.
+To answer this question for small trees, we have implemented the algorithm $\findpath$
 \todo{LC: github - link}
-For computing the actual $\rnni$ distances we used the algorithm for computing the $\rnni$ graph as described in Section~\ref{section:alg_RNNI_graph}.
+\autocite{}
+and compared the results with the true distance found using the algorithm from Section~\ref{section:alg_RNNI_graph}.
 
@@
 
-For introducing an algorithm to compute paths between caterpillar trees, we need further notations.
-In this section we aim to investigate properties of paths between \emph{caterpillar trees} where each internal node has at least one child that is a leaf.
-A path $p$ between two caterpillar trees $T$ and $R$ that only consists of caterpillar trees is called \emph{caterpillar path}.
-If $p$ is shortest among all caterpillar paths between these trees we call it \emph{shortest caterpillar path}, and the length of $p$ will be denoted by $d_c(T,R)$ and called \emph{caterpillar distance}.
+In this section we aim to investigate properties of paths between \emph{caterpillar trees}, each internal node of which is adjacent to at least one leaf.
+A path $p$ between two caterpillar trees $T$ and $R$ that only consists of caterpillar trees is called a \emph{caterpillar path}.
+If $p$ is the shortest pass among all caterpillar paths between $T$ and $R$ we call it a \emph{shortest caterpillar path}, and denote the length of $p$ by $d_c(T,R)$, which we call the \emph{caterpillar distance}.
 
@@
 This representation is basically a simplification of the cluster representation for caterpillar trees and works as follows.
-For a caterpillar tree with cluster representation $T = [C_1, \ldots, C_{n-2}]$ it is $C_i \subset C_{i+1}$ with $|C_i| = |C_{i+1}|$ for all $i = 1, \ldots, n-3$.
+\todo{I don't follow these two sentences.}
+For a caterpillar tree with a cluster representation $T = [C_1, \ldots, C_{n-2}]$ it is $C_i \subset C_{i+1}$ with $|C_i| = |C_{i+1}|$ for all $i = 1, \ldots, n-3$.
 Therefore, the list of taxa $L_T = [x_1, x_2, \ldots, x_n]$ with $x_1<x_2$ uniquely represents the caterpillar tree $T = [C_1, \ldots, C_{n-2}]$ with $C_i = \{x_1, \ldots, x_{i+1}\}$.
@@
 Note that the only two elements in such a list that have parents of equal rank are the first two elements $x_1$ and $x_2$.
-That is why we order them in the list such that $x_1 < x_2$ to make the list representation of caterpillar trees unique.
-We will call a pair of two taxa that share their parent \emph{cherry}.
-We will use this list representation to show how one can adapt the classical \emph{Bubble Sort} algorithm \autocite{Knuth1997-pi} to compute caterpillar paths.
+That is why we order them in the list such that
+\todo{Are we assuming taxa is ordered?}
+$x_1 < x_2$
+to make the list representation of caterpillar trees unique.
+We will call a pair of two taxa that share their parent a \emph{cherry}.
 
-We are now going to explain what kind of $\rnni$ moves are possible between caterpillar trees and see how such a move changes the list representation of a tree.
-At first it is important to realise that the only possible $\rnni$ moves between caterpillar trees are $\nni$ moves, since there is no pair of internal nodes with rank difference one that is not connected by an edge.
-$\nni$ moves between caterpillar trees correspond to swaps of two neighboured elements in the list representation.
-The only exception of this are $\nni$ moves on the lowest internal edge, that is the edge connecting the internal nodes of rank one and two.
-There are two $\nni$ moves possible on this edge that result in caterpillar trees, while there is for every other edge only one $\nni$ move that results in a caterpillar tree.
-Corresponding to the $\nni$ moves on the lowest internal edge are exchanges of either the first or the second element with the third element in the list representation.
+%We are now going to explain what kind of $\rnni$ moves are possible between caterpillar trees and see how such a move changes the list representation of a tree.
+%At first it is important to realise that the only possible $\rnni$ moves between caterpillar trees are $\nni$ moves, since there is no pair of internal nodes with rank difference one that is not connected by an edge.
+%$\nni$ moves between caterpillar trees correspond to swaps of two neighboured elements in the list representation.
+%The only exception of this are $\nni$ moves on the lowest internal edge, that is the edge connecting the internal nodes of rank one and two.
+%There are two $\nni$ moves possible on this edge that result in caterpillar trees, while there is for every other edge only one $\nni$ move that results in a caterpillar tree.
+%Corresponding to the $\nni$ moves on the lowest internal edge are exchanges of either the first or the second element with the third element in the list representation.
 
-By using this knowledge we can compute shortest caterpillar paths by modifying Bubble Sort as described in the following.
-We call this algorithm $\csort$ (Algorithm~\ref{alg:csort}).
-
-Let $T$ and $R$ be the input trees for $\csort$.
-We assume that they are given in their list representation $L_T$ and $L_R$ as described above and want to compute a path from $L_T$ to $L_R$.
+The algorithm $\csort$ (Algorithm~\ref{alg:csort}) is a natural modification of the classical \emph{Bubble Sort} algorithm \autocite{Knuth1997-pi} to compute caterpillar paths.
+Let $T$ and $R$ be the input trees given by their list representations $L_T$ and $L_R$.
 This path $p$ is built stepwise by adding a new list to the end of $p$.

### 2019-11-14 rnni_convexity.tex
add citations

@@
 A consequence is that tree search and sampling algorithms pose a significant challenge even for moderately sized trees
-\todo{cite Chris Whidden}
-\autocite{}.
+\todo{LC: Alex, is this the citation you wanted?}
+\autocite{Whidden2018-tk}.
 
-More recent advances in computational phylogenetics introduce various classes of molecular clock models \autocite{Yoder2000-ks,Drummond2006-nl,Drummond2010-yf} and made possible computational inference of phylogenetic time-trees
-\todo{cite MrBayes, BEAST, NextStrain}
-\autocite{}.
+More recent advances in computational phylogenetics introduce various classes of molecular clock models \autocite{Yoder2000-ks,Drummond2006-nl,Drummond2010-yf} and made possible computational inference of phylogenetic time-trees \autocite{Ronquist2003-eq, Bouckaert2018-yr, Hadfield2018-xp}.
 However, mathematical challenges that come with this seemingly inessential change in parametrisation (genomic distance VS time distance) of trees have only recently been brought to attention \autocite{Gavryushkin2016-uu}.

### 2019-11-19 rnni_convexity.tex
introduction

@@
 
-More recent advances in computational phylogenetics introduce various classes of molecular clock models \autocite{Yoder2000-ks,Drummond2006-nl,Drummond2010-yf} and made possible computational inference of phylogenetic time-trees \autocite{Ronquist2003-eq, Bouckaert2018-yr, Hadfield2018-xp}.
+More recent advances in computational phylogenetics introduce various classes of molecular clock models \autocite{Yoder2000-ks,Drummond2006-nl,Drummond2010-yf} and made computational inference of phylogenetic time-trees possible \autocite{Ronquist2003-eq, Bouckaert2018-yr, Hadfield2018-xp}.
 However, mathematical challenges that come with this seemingly inessential change in parametrisation (genomic distance VS time distance) of trees have only recently been brought to attention \autocite{Gavryushkin2016-uu}.
@@
 In line with the research programme proposed in \autocite{Gavryushkin2018-ol}, we investigate the geometry and algorithmic complexity of the $\rnni$ space.
-Specifically, in this paper we establish the exact radius and diameter of the space.
-We also show that the subset of caterpillar trees is convex in $\rnni$, thus settling one of the open problems proposed in \autocite{Gavryushkin2018-ol}.
-Furthermore, we provide an approximate algorithm that computes exact distances for small trees.
+Specifically, in this paper we establish the exact radius and diameter of the space (Section~\ref{section:diameter}).
+We also show that the subset of caterpillar trees is convex in $\rnni$ (Section~\ref{section:caterpillar_convex}), thus settling one of the open problems proposed in \autocite{Gavryushkin2018-ol}.
+For establishing these geometric properties we are using algorithms that will be introduced in Section~\ref{section:algorithms}.
+We will in particular provide an approximate algorithm that computes exact distances for small trees.
 The question of whether there exists a polynomial algorithm for computing the $\rnni$ distance remains an open problem.
@@
 
-A \emph{ranked phylogenetic tree} is a pair consisting of a rooted binary phylogenetic tree $T$ on the set $X = \{1, \ldots, n\}$ of \emph{taxa} for $n \in \mathbb N$, and a (total) rank function that maps all leaves of $T$ to $0$, all internal nodes of $T$ onto the set $\{1, \ldots, n-1\}$, and respects the partial order on the nodes given by the tree.
+A \emph{ranked phylogenetic tree} is a pair consisting of a rooted binary phylogenetic tree $T$ on the set $X = \{1, \ldots, n\}$ of \emph{taxa} for $n \in \mathbb N$, and a (total) rank function that maps all leaves of $T$ to $0$, all internal nodes of $T$ onto elements of the set $\{1, \ldots, n-1\}$, and respects the partial order on the nodes given by the tree.
 The latter means that if $u$ and $v$ are two internal nodes of $T$ such that there exists a path from a taxon $x \in X$ to the root which first passes through $u$ and then through $v$ then $\rank(u) < \rank(v)$.
@@
 Since all trees in this paper are ranked trees, we will abuse the notation and drop the rank function from the notation.
-We will also be simply saying tress to mean ranked phylogenetic trees.
+We will also simply say trees to mean ranked phylogenetic trees.
 For a tree $T$, we will use $\rank_T$ to refer to its ranking.
@@
 
-Our definition of a ranked tree implies a natural notion of the edge length -- we call $|\rank_T(v) - \rank_T(u)|$ the length of an edge $(u, v)$.
+Our definition of a ranked tree implies a natural notion of edge length -- we call $|\rank_T(v) - \rank_T(u)|$ the length of an edge $(u, v)$.
 
@@
 The first type of operation is called a \emph{rank move} and defined by swapping the ranks of two internal nodes that are not adjacent in the tree.
-Formally, if $u$ and $v$ are such nodes of a tree $T$ that $|\rank_T(u) - \rank_T(v)| = 1$ and $(u, v)$ is not an edge in $T$ then the tree $R$ obtained from $T$ by only changing $\rank_R(u) = \rank_T(v)$ and $\rank_R(v) = \rank_T(u)$ is said to be obtained by a rank move.
+Formally, if $u$ and $v$ are nodes of a tree $T$ such that $|\rank_T(u) - \rank_T(v)| = 1$ and $(u, v)$ is not an edge in $T$ then the tree $R$ obtained from $T$ by only changing $\rank_R(u) = \rank_T(v)$ and $\rank_R(v) = \rank_T(u)$ is said to be obtained by a rank move.
 The second type of operation is called an $\nni$ \emph{move} and defined in the usual way, that is, two trees $T$ and $R$ are said to be connected by an $\nni$ move if there exist internal edges $e$ in $T$ and $f$ in $R$ both of length one such that the trees obtained by shrinking $e$ and $f$ to internal nodes coincide.
@@
 \caption{Two types of operation that define edges of the $\rnni$ graph.
-The move at the top left is a rank move while the trees in the triangle are connected by $\nni$ moves.}
+The rank move swaps the ranks of the two nodes in the tree on the left such that $\nni$ moves are possible on the edge between the two nodes marked by black dots.}
 \label{fig:RNNI}
 \end{figure}
-
-\todo[inline]{LC: Explain structure of paper}
 
@@
 \textcite[Section 3.3]{Gavryushkin2018-ol} constructed an exact algorithm for computing the $\rnni$ graph.
-\todo{Do we make use of tanglegrams?}
+\todo{Do we make use of tanglegrams? - LC: No}
 \todo{GitHub link?}

### 2019-11-19 rnni_convexity.tex
rm section for computing RNNI graph

@@
 Then we introduce two algorithms, $\findpath$ (Section~\ref{section:alg_findpath}) and $\csort$ (Section~\ref{section:alg_csort}), for computing paths between two given trees in $\rnni$.
-The first of these algorithms takes arbitrary trees as an input while the latter is restricted to trees with a certain tree topology.
-Both of these algorithms will become an important ingredient to establish several geometric properties of the $\rnni$ graph in Section~\ref{section:geometry}.
+These algorithms will become an important ingredient to establish several geometric properties of the $\rnni$ graph in Section~\ref{section:geometry}.
 Finally we will present the algorithm $\mdtree$, which produces a tree that is far away from the given input tree (see Section~\ref{section:alg_mdtree} for details).
@@
 
-\subsection{Computing the $\rnni$ graph}
-\label{section:alg_RNNI_graph}
-
-\textcite[Section 3.3]{Gavryushkin2018-ol} constructed an exact algorithm for computing the $\rnni$ graph.
+For comparing approximate distances computed by $\findpath$ (Section~\ref{section:alg_findpath}) with exact distances and for supporting Conjecture~\ref{conjecture:cluster_theorem}, we aim to compute exact distances in $\rnni$ for small trees.
+Therefore we will use the algorithm for computing the $\rnni$ graph that is given in \textcite[Section 3.3]{Gavryushkin2018-ol}.
 \todo{Do we make use of tanglegrams? - LC: No}
 \todo{GitHub link?}
-We implemented this algorithm and used it for computing this graph for trees on up to seven taxa.
-\todo{LC: $7$ or $8$? - I'll need to check this}
-Once the graph is computed, one can get distances by using algorithms such as Floyd-Warshall \autocite{Floyd1962-ew} or Dijkstra \autocite{Dijkstra1959-ph}.
+Once the graph is computed one can use common algorithms for computing shortest paths~\autocite{Floyd1962-ew,Dijkstra1959-ph}.
 However, the large number of vertices in $\rnni$ is an obstacle when it comes to computing distances.
-Since there are $\frac{n!(n-1)!}{2^{n-1}}$ vertices in the $\rnni$ graph \autocite{Gavryushkin2018-ol}, we already run out of space when we want to save all pairwise distances for trees on seven taxa.
-\todo{Is this the hint we're using tanglegrams?
-Either way this statement is too vague and requires more detail.
-Also, ``run out of space'' should be a bit less frank.
-Finally, think whether you want single out this paragraph as a special subsection?
-Either way you should say why you need this implementation.}
-Therefore, we compute distances between certain pairs of trees and use the symmetry of $\rnni$ to draw conclusions about the remaining distances.
+Since there are $\frac{n!(n-1)!}{2^{n-1}}$ vertices in the $\rnni$ graph \autocite{Gavryushkin2018-ol}, we are just able to use the $\rnni$ graph on up to six taxa for our analyses.
+We will get into further details when we discuss particular problems in Section~\ref{section:alg_findpath} and~\ref{section:cluster_theorem}.
 

### 2019-11-19 rnni_convexity.tex
update explanation of FINDPATH

@@
 
+Within this section we are presenting $\findpath$, an efficient algorithm for computing paths, and therefore approximating distances, between any given pair of trees in $\rnni$.
+As we will see in Section~\ref{section:geometry}, it is useful for gaining insights into geometric properties of the $\rnni$ graph.
+
 Before introducing the algorithm we need the following definitions.
@@
 
-\todo{I suggest to rewrite this paragraph following the standard structure:\\
-\emph{Step $0$. $T_0 = T$, ...\\
-Step $k+1$. $T_{k+1}$ is defined...}}
-We are now ready to introduce $\findpath$ (Algorithm~\ref{alg:find_path}), an efficient algorithm for computing paths between any two given trees $T$ and $R$ in $\rnni$.
+We are now ready to explain how $\findpath$ (Algorithm~\ref{alg:find_path}) computes a path between two input trees $T$ and $R$.
 Let $[C_1, \ldots, C_{n-2}]$ be the cluster representation of $R$ (the ``destination'' tree).
 Then $C_i$ is induced by the node of rank $i$ for $i = 1, \ldots, n-2$.
-$\findpath$ computes a path $p$ which initially consists of $T$ only and is extended stepwise.
-In order to do this, the clusters $C_1, \ldots, C_{n-2}$ of $R$ are considered in their order.
-Let $T'$ be the last tree of $p$ when
-\todo{This term is not clear or defined.}
-arriving at cluster $C_k$.
-We assume that the clusters $C_1, \ldots, C_{k-1}$ are induced by nodes of rank $1, \ldots, k-1$ in $T'$, respectively.
-In the first step it is $T' = T$ and $C_k = C_1$.
-The aim is now to find a sequence of $\rnni$ moves, starting at $T'$, that do not change the clusters $C_1, \ldots, C_{k-1}$, but result in a tree containing clusters $C_1, \ldots, C_{k}$.
-Therefore, we need to find the smallest cluster containing all elements of $C_k$ in $T'$.
-This node is denoted by $\mrca_{T'}(C_k)$ (short for \emph{most recent common ancestor}).
-The rank of $\mrca_{T'}(C_k)$ in $T'$ is higher than the rank of the node inducing $C_k$ in $R$, because $C_1, \ldots, C_{k-1}$ are already present in $T'$ and we assume that $C_k$ is not.
-Now perform an $\rnni$ move that decreases the rank of $\mrca_{T'}(C_k)$.
-\todo{I think it is obvious.
-I suggest to remove this Lemma and save it somewhere in case we get an too sensitive referee.
-Tag a commit?
-Or branch? -- whatever works.}
-The existence of such a move is not obvious but we will see later (Lemma~\ref{lemma:mrca_move}) that this kind of move always exists as long as $T'$ does not induce the cluster $C_k$.
-So we can perform such moves until a tree $T''$ is reached with $\mrca_{T''}(C_k)$ inducing $C_k$ and having has the same rank $k$ in $T''$ as in $R$.
-Appending all trees on this path from $T'$ to $T''$ and repeating this procedure for all clusters of $R$ results in a path $p$ from $T$ to $R$.
+$\findpath$ computes a path which initially consists of $T_0 = T$ only and is extended stepwise as described in the following.
+
+Step $0$. $p = [T_0]$.
+Consider the node in $T_0$ that induces the smallest cluster containing all elements of $C_1$.
+This node is denoted by $\mrca_{T_0}(C_1)$ (short for \emph{most recent common ancestor}).
+By performing an $\rnni$ move that decreases the rank of $\mrca_{T_0}(C_1)$, we receive a tree $T_1$ that is then appended to $p$.
+Now we repeatedly perform $\rnni$ moves that move the most recent common ancestor of $C_1$ down in order to extend $p$ to $p = [T_0, \ldots, T_l]$ for some $l$ such that $C_1$ is induced by the internal node of rank one in $T_l$.
+
+Step $k$. $p = [T_0, \ldots, T_l]$ where $l$ is the total number of trees that have been added to $p$ in the previous $k-1$ steps.
+Therefore the internal nodes in $T_l$ with ranks $1, \ldots, k-1$ induce clusters $C_1, \ldots, C_{k-1}$, respectively.
+Now the cluster of interest is $C_k$.
+As in the first step $p$ is extended by performing a series of $\rnni$ moves that decrease the rank of $\mrca_{T_l}(C_k)$ stepwise.
+Afterwards it is $p = [T_0, \ldots, T_{l'}]$ where the $k$ lowest internal nodes of $T_{l'}$ induce clusters $C_1, \ldots, C_{k}$.
+
+By following this procedure, the end tree of $p$ after step $n-2$ will be $R$, as it contains all clusters of $R$.
+It is important but easy to see that the output path of $\findpath$ is unique for a given pair of trees, which means that it is a deterministic algorithm.
+This is due to the fact that there exists exactly one $\rnni$ move that decreases the rank of the most recent common ancestor of a subset $C_k$ of the taxon set of a tree as long as $C_k$ is not induced by an internal node of that tree.
 

### 2019-11-19 rnni_convexity.tex
until end of section FINDPATH

@@
 
-\todo{$p = p+T'$: Do we really want to be updating this in every while loop?}
+\todo{$p = p+T'$: Do we really want to be updating this in every while loop? - LC: Yes, I guess so}
 \begin{algorithm}[H]
@@
 \STATE $p := [T']$
-\FOR {$i = 1, \dots, n-2$}
-\STATE Let $C_i$ be the cluster induced by the node with rank $i$ in $R$ \label{alg:find_path:line:cluster}
-\WHILE {$\rank_{T'}(\mrca_{T'}(C_i))>i$}
-\STATE Update $T'$: Decrease $\rank_{T'}(\mrca_{T'}(C_i))$ by an $\rnni$ move \label{alg:findpath:line:move_set_down}
+\FOR {$k = 1, \dots, n-2$}
+\STATE Let $C_k$ be the cluster induced by the node with rank $k$ in $R$ \label{alg:find_path:line:cluster}
+\WHILE {$\rank_{T'}(\mrca_{T'}(C_k))>k$}
+\STATE Update $T'$: Decrease $\rank_{T'}(\mrca_{T'}(C_k))$ by an $\rnni$ move \label{alg:findpath:line:move_set_down}
 \STATE $p = p+T'$
@@
 \begin{lemma}
-Let $T$ and $R = [C_1, \ldots, C_{n-2}]$ be two trees and
-\todo{This lemma is to prove that FindPath is an (deterministic) algorithm, so we cannot relay on this.}
-$p = \findpath (T,R)$.
-Let $T'$ be a tree on $p$ different from both $T$ and $R$ such that $C_1, \ldots, C_{k-1}$ are induced by internal nodes of $T'$, but $C_k$ is not.
-Then there exists exactly one $\rnni$ move on $T'$ that results in a tree $T''$ such that
-\todo{What's $i$?}
-$\mrca_{T''}(C_i) < \mrca_{T'}(C_i)$.
+Let $T = [C_1, \ldots, C_{k-1}, B_{k}, \ldots, B_{n-2}]$ and $R = [C_1, \ldots, C_{k-1}, C_k, \ldots, C_{n-2}]$ be two trees where $C_k \neq B_k, \ldots, B_{n-2}$.
+Then there exists exactly one $\rnni$ move on $T$ that results in a tree $T'$ such that $\mrca_{T'}(C_k) < \mrca_{T}(C_k)$.
 \label{lemma:mrca_move}
@@
 \begin{proof}
-Let $T, R$ and $T'$ be trees as defined in the lemma and $[C_1, \ldots, C_{n-2}]$ the cluster representation of $R$.
-We want to prove that there is exactly one $\rnni$ move on $T'$ that results in a tree $T''$ where $\mrca_{T''}(C_k) < \mrca_{T'}(C_k)$.
-Notice, that the only $\rnni$ move on $T$ that could decrease the rank of $\mrca_{T'}(C_k)$ are $\nni$ movs on the edge between $\mrca_{T'}(C_k)$ and its child (if this edge has length one), or a rank move between $\mrca_{T'}(C_k)$ and the node with rank one less.
+Let $T$ and $R$ be trees as defined in the lemma.
+Notice that the only $\rnni$ move on $T$ that could decrease the rank of $\mrca_T(C_k)$ are $\nni$ moves on the edge between $\mrca_T(C_k)$ and its child (if this edge has length one), or a rank move between $\mrca_T(C_k)$ and the node with rank one less.
 Therefore, we will now distinguish these two cases:
-Either there is an edge $(v,w)$ between nodes $v = \mrca_{T'}(C_k)$ and $w$ with rank $rank_{T'}(w) = rank_{T'}(v) - 1$, or there is no such edge.
+Either there is an edge $(v,w)$ between nodes $v = \mrca_T(C_k)$ and $w$ with rank $rank_T(w) = rank_T(v) - 1$, or there is no such edge.
 
-In the latter case the only $\rnni$ move that could increase the rank of $v$ is a rank move swapping the ranks of $v$ and $w$.
+In the latter case the only $\rnni$ move that could decrease the rank of $v$ is a rank move swapping the ranks of $v$ and $w$.
 
-If there is an edge $(v,w)$ in $T'$ there are two $\nni$ moves possible on this edge.
-We are now going to explain why only one of these two $\nni$ moves changes the rank of this most recent common ancestor.
-By definition we know that $C_1, \ldots, C_{k-1}$ are induced by the nodes of rank $1, \ldots, k-1$, respectively, in $T'$ already.
-It is important to notice that $C_k$ either is the disjoint union $C_k = C_i \dot\cup C_j$ for $i,j < k$ or $C_k = C_i \dot\cup \{x\}$ for a taxon $x \in X \setminus (C_1 \cup \ldots \cup C_{k-1})$.
+Let us now assume that there is an edge $(v,w)$ in $T$.
+Let $T_v$ be the subtree that is rooted in the child of $v$ that is not $w$ and let $T_{w_1}$ and $T_{w_2}$ be the subtrees rooted in the children of $w$.
+We are now going to explain why only one of these two $\nni$ moves changes the rank of $\mrca_T(C_k)$.
+It is important to notice that $C_k$ either is the disjoint union of two of the clusters $C_1, \ldots, C_{k-1}$ or the union of one of those clusters with a single taxon that is not included in any of those clusters.
+In general we can say that $C_k = S_1 \dot\cup S_2$ is the union of two subsets of taxon set of $T$ such each of these subsets is contained in one of the subtrees $T_v, T_{w_1}$, or $T_{w_2}$.
+In fact, $S_1$ and $S_2$ are contained in two different of these subtrees.
+Now it is easy to see that just one of the $\nni$ moves on $(v,w)$ (illustrated in Figure~\ref{fig:mrca_move}) decreases the rank of $\mrca_T(C_k)$.
 
-Let us at first assume that $C_k = C_i \dot\cup C_j$ for some $i,j < k$.
-The tree $T'$ looks like the tree depicted in the left of Figure~\ref{fig:mrca_move} where $T_w^1$ and $T_w^2$ are the two subtrees rooted in the children of $w$ and $T_v$ is the subtree whose root is child of $v$ and does not include $w$.
-Since we know that all clusters $C_1,\ ldots, C_{k-1}$ are induced by some nodes in $T'$, each of these clusters must be a subset of the taxon set of one of the three subtrees $T_w^1,T_w^2$, or $T_v$.
-In particular, there can't be elements of the same cluster $C_m$ in more than one of these subtrees.
-This and the fact that $v = \mrca_{T'}(C_i \dot\cup C_j)$ yields that either $C_i$ or $C_j$ is subset of the taxon set of $T_v$ and the other cluster is subset of the taxon set of either $T_w^1$ or $T_w^2$.
-
-Without loss of generality we assume that $C_i$ is included in $T_v$.
-Otherwise, change the indices $i$ and $j$.
-Then $C_i$ is a subset of the taxon set of either $T_w^1$ or $T_w^2$.
-Therefore, just one of the two $\nni$ moves on $(v,w)$, as depicted in Figure~\ref{fig:mrca_move}, moves the most recent common ancestor of $C_k = C_i \dot \cup C_j$ down.
-
-Now we consider the case that $C_k = C_i \dot\cup \{x\}$ for a taxon $x \in X \setminus (C_1 \cup \ldots \cup C_{k-1})$.
-For the same reason as above, $C_i$ is in subset of the taxon set of one of the subtrees $T_w^1, T_w^2$ or $T_v$ of $\hat T$ as depicted in Figure~\ref{fig:mrca_move}.
-Similarly $x$ is taxon of one of these three subtrees in a way that either $C_i$ or $x$ is included in $T_v$, because $v = \mrca_{T'}(C_i \dot \cup \{x\})$.
-So there is just one of the two subtrees $T_w^1$ and $T_w^2$ that neither includes $C_j$ nor $x$.
-As in the previous case, just one of the two $\nni$ moves on $(v,w)$ as illustrated in Figure~\ref{fig:mrca_move} decreases the rank of $\mrca_{T'}(C_i \dot \cup \{x\})$.
-
-We can conclude that in any case there is exactly one $\rnni$ move possible on $T'$ that decreases the rank of $\mrca_{T'}(C_k)$ in Line~\ref{alg:findpath:line:move_set_down} of Algorithm~\ref{alg:find_path}.
+% It is important to notice that $C_k$ either is the disjoint union $C_k = C_i \dot\cup C_j$ for $i,j < k$ or $C_k = C_i \dot\cup \{x\}$ for a taxon $x \in X \setminus (C_1 \cup \ldots \cup C_{k-1})$.
+%
+% Let us at first assume that $C_k = C_i \dot\cup C_j$ for some $i,j < k$.
+% The tree $T$ looks like the tree depicted in the left of Figure~\ref{fig:mrca_move} where $T_w^1$ and $T_w^2$ are the two subtrees rooted in the children of $w$ and $T_v$ is the subtree whose root is child of $v$ and does not include $w$.
+% Since we know that all clusters $C_1,\ ldots, C_{k-1}$ are induced by some nodes in $T$, each of these clusters must be a subset of the taxon set of one of the three subtrees $T_w^1,T_w^2$, or $T_v$.
+% In particular, there can't be elements of the same cluster $C_m$ in more than one of these subtrees.
+% This and the fact that $v = \mrca_{T}(C_i \dot\cup C_j)$ yields that either $C_i$ or $C_j$ is subset of the taxon set of $T_v$ and the other cluster is subset of the taxon set of either $T_w^1$ or $T_w^2$.
+%
+% Without loss of generality we assume that $C_i$ is included in $T_v$.
+% Otherwise, change the indices $i$ and $j$.
+% Then $C_i$ is a subset of the taxon set of either $T_w^1$ or $T_w^2$.
+% Therefore, just one of the two $\nni$ moves on $(v,w)$, as depicted in Figure~\ref{fig:mrca_move}, moves the most recent common ancestor of $C_k = C_i \dot \cup C_j$ down.
+%
+% Now we consider the case that $C_k = C_i \dot\cup \{x\}$ for a taxon $x \in X \setminus (C_1 \cup \ldots \cup C_{k-1})$.
+% For the same reason as above, $C_i$ is in subset of the taxon set of one of the subtrees $T_w^1, T_w^2$ or $T_v$ of $\hat T$ as depicted in Figure~\ref{fig:mrca_move}.
+% Similarly $x$ is taxon of one of these three subtrees in a way that either $C_i$ or $x$ is included in $T_v$, because $v = \mrca_{T'}(C_i \dot \cup \{x\})$.
+% So there is just one of the two subtrees $T_w^1$ and $T_w^2$ that neither includes $C_j$ nor $x$.
+% As in the previous case, just one of the two $\nni$ moves on $(v,w)$ as illustrated in Figure~\ref{fig:mrca_move} decreases the rank of $\mrca_{T}(C_i \dot \cup \{x\})$.
+%
+% We can conclude that in any case there is exactly one $\rnni$ move possible on $T$ that decreases the rank of $\mrca_{T}(C_k)$ in Line~\ref{alg:findpath:line:move_set_down} of Algorithm~\ref{alg:find_path}.
 
@@
 \vspace{12pt}
-\caption{Two $\nni$ moves that are possible on the edge $(v,w)$ of the tree $T'$.
-They result in the trees on the right as described in the proof of Lemma~\ref{lemma:mrca_move}.}
+\caption{Two $\nni$ moves that are possible on the edge $(v,w)$ of the tree $T'$ with subtrees $T_v, T_{w_1}, T_{w_2}$ as described in the proof of Lemma~\ref{lemma:mrca_move}.}
 \label{fig:mrca_move}
@@
 \autocite{}
-and compared the results with the true distance found using the algorithm from Section~\ref{section:alg_RNNI_graph}.
+and compared the results with the true distance found using the algorithm from Section~\ref{section:algorithms}.
+By doing this we could show that $\findpath$ computes correct distances for all trees on up to six taxa.
 

### 2019-11-20 rnni_convexity.tex
up to explanation csort

@@
 
-In this section we aim to investigate properties of paths between \emph{caterpillar trees}, each internal node of which is adjacent to at least one leaf.
+In this section we introduce an algorithm to compute paths between \emph{caterpillar trees}, which are trees where each internal node is adjacent to at least one leaf.
+As a result caterpillar trees have only one \emph{cherry}, which is a pair of taxa that share their parent.
+For introducing this algorithm we need the following definitions.
 A path $p$ between two caterpillar trees $T$ and $R$ that only consists of caterpillar trees is called a \emph{caterpillar path}.
-If $p$ is the shortest pass among all caterpillar paths between $T$ and $R$ we call it a \emph{shortest caterpillar path}, and denote the length of $p$ by $d_c(T,R)$, which we call the \emph{caterpillar distance}.
+If $p$ is the shortest path among all caterpillar paths between $T$ and $R$ we call it a \emph{shortest caterpillar path}, and denote the length of $p$ by $d_c(T,R)$, which we call the \emph{caterpillar distance}.
 
-Let us now introduce a representation of caterpillar trees as lists of taxa.
-This representation is basically a simplification of the cluster representation for caterpillar trees and works as follows.
-\todo{I don't follow these two sentences.}
-For a caterpillar tree with a cluster representation $T = [C_1, \ldots, C_{n-2}]$ it is $C_i \subset C_{i+1}$ with $|C_i| = |C_{i+1}|$ for all $i = 1, \ldots, n-3$.
-Therefore, the list of taxa $L_T = [x_1, x_2, \ldots, x_n]$ with $x_1<x_2$ uniquely represents the caterpillar tree $T = [C_1, \ldots, C_{n-2}]$ with $C_i = \{x_1, \ldots, x_{i+1}\}$.
-In this list representing a caterpillar tree the taxa are ordered non-decreasing, according to the ranks of their parents.
-Note that the only two elements in such a list that have parents of equal rank are the first two elements $x_1$ and $x_2$.
-That is why we order them in the list such that
-\todo{Are we assuming taxa is ordered?}
-$x_1 < x_2$
-to make the list representation of caterpillar trees unique.
-We will call a pair of two taxa that share their parent a \emph{cherry}.
+Let us now introduce a representation of caterpillar trees as lists of taxa, which is a simplification of the cluster representation for caterpillar trees and works as follows.
+In the list representing a caterpillar tree its taxa are ordered non-decreasing, according to the ranks of their parents.
+Note that the only two elements in such a list that have parents of equal rank are the first two elements $l_1$ and $l_2$.
+As we assume to have taxon set $X = \{1, \ldots, n\}$, we assume that they are order in the list such that $l_1 < l_2$, according to their ordering as natural numbers.
+This ensures that the list representation of caterpillar trees is unique.
 
@@
 
-The algorithm $\csort$ (Algorithm~\ref{alg:csort}) is a natural modification of the classical \emph{Bubble Sort} algorithm \autocite{Knuth1997-pi} to compute caterpillar paths.
-Let $T$ and $R$ be the input trees given by their list representations $L_T$ and $L_R$.
-This path $p$ is built stepwise by adding a new list to the end of $p$.
+The algorithm $\csort$ (Algorithm~\ref{alg:csort}) is a natural modification of the classical \emph{Bubble Sort} algorithm \autocite{Knuth1997-pi}.
+Let $T$ and $R$ be the input trees given by their list representations $L_T$ and $L_R$ as introduced above.
+The output path $p$ is built stepwise by adding a new list to the end of $p$.
 We call the last element of $p$ at any step $L_{T'}$, representing a tree $T'$ and when we say we update $L_{T'}$, we modify this list and append the new version of $L_{T'}$ to $p$.

### 2019-11-20 rnni_convexity.tex
update description csort

@@
 
-Let us now introduce a representation of caterpillar trees as lists of taxa, which is a simplification of the cluster representation for caterpillar trees and works as follows.
-In the list representing a caterpillar tree its taxa are ordered non-decreasing, according to the ranks of their parents.
-Note that the only two elements in such a list that have parents of equal rank are the first two elements $l_1$ and $l_2$.
-As we assume to have taxon set $X = \{1, \ldots, n\}$, we assume that they are order in the list such that $l_1 < l_2$, according to their ordering as natural numbers.
+The special topology of caterpillar trees gives a natural preorder of taxa that uniquely represents a caterpillar tree.
+That is, for a caterpillar tree $T = [\{x_1, x_2\}, \{x_1, x_2, x_3\}, \ldots, \{x_1, \ldots, x_n\}]$ the taxa are preordered according to the ranks of their parents such that $x_1 = x_2 \leq x_3 \leq \ldots \leq x_n$.
+From now on when we say that a taxon is below another one we refer to this preorder.
+Furthermore, we can use this preorder to create a list representation of caterpillar trees.
+For the tree $T$ above this list is $[x_1, x_2, \ldots, x_n]$.
+As we consider the taxon set $X = \{1, \ldots, n\}$, we assume that they are ordered in the list such that $x_1 < x_2$, according to their ordering as natural numbers.
 This ensures that the list representation of caterpillar trees is unique.
@@
 The algorithm $\csort$ (Algorithm~\ref{alg:csort}) is a natural modification of the classical \emph{Bubble Sort} algorithm \autocite{Knuth1997-pi}.
-Let $T$ and $R$ be the input trees given by their list representations $L_T$ and $L_R$ as introduced above.
-The output path $p$ is built stepwise by adding a new list to the end of $p$.
-We call the last element of $p$ at any step $L_{T'}$, representing a tree $T'$ and when we say we update $L_{T'}$, we modify this list and append the new version of $L_{T'}$ to $p$.
+A path $p$ between two given trees $T$ and $R$ is computed by sorting the list representations of $T$ and $R$ as it is done in Bubble Sort.
+The general idea is to update the list representing $T$ by considering it from beginning to end and swap two neighboured taxa if they appear in a different order in $R$ than in the current list.
+Note that each swap of taxa results in a new tree that is added to the path $p$.
+After passing the whole list the last element, which is the uppermost taxon, is at the same position as in $R$.
+By repeating this at most $n-1$ times all elements the resulting list represents $R$.
 
-As in the Bubble Sort algorithm, we go through the list $L_{T'}$ and swap two elements if they appear in different orders in $L_{T'}$ and $L_R$.
-Each swap of elements is an update of $T'$ and extends the path $p$ by one element.
-We compare pairs of list elements from the start of the list until its end, and repeat this until the whole list is sorted, just as in classical Bubble Sort.
-Our algorithm $\csort$ only differs from Bubble Sort when considering the first three elements of the list.
-As we discussed above, both the first and second element of $L_{T'}$ can exchange with its third element.
-Therefore, we need to consider these moves separately to all other moves.
-Most importantly, we need to make sure that these moves are performed in a way that leads to a shortest caterpillar path.
-We can guarantee this if the moves are done as described in the following.
-Let $x_1, x_2$ and $x_3$ be the first three elements of $L_{T'}$, in this order.
-If $x_1$ ($x_2$) is the last element of those three in $L_R$, we update $L_{T'}$ by swapping $x_1$ ($x_2$) and $x_3$.
-By doing this we make sure that there will be no point on $p$ where $x_1$ and $x_2$ swap positions, which would cause an additional $\nni$ move that can be avoided if we perform the moves as described.
+We use this main principle of Bubble Sort for computing shortest caterpillar paths with $\csort$, but there are special consideration needed for the first three taxa.
+The reason for this is that both $\nni$ moves on the lowest internal edge of a caterpillar tree result in caterpillar trees.
+For computing a shortest caterpillar path the moves on the first three taxa $x_1, x_2, x_3$ need to be done as described in the following.
+If $x_1$ ($x_2$) is the uppermost taxon of those three in $R$, we update the current tree by swapping $x_1$ ($x_2$) and $x_3$.
+By doing this we make sure that there will be no point on $p$ where $x_1$ and $x_2$ swap positions, as this is an additional $\nni$ move that can be avoided.
 
-Because all moves between caterpillar trees are $\nni$ moves, all pairs of taxa that are not in the same order in $L_T$ and $L_R$ need to swap at some point on any path between $L_T$ and $L_R$.
-And since only elements that are direct neighbours in a sequence are allowed to swap, with the only exception being the first three elements, the path computed by $\csort$ is a shortest caterpillar path.
+Because all moves between caterpillar trees are $\nni$ moves, all pairs of taxa that are not in the same order in $T$ and $R$ need to swap at some point on any path between these trees.
+And since only elements that are direct neighbours in a sequence are allowed to swap, with the only exception being the first three elements, the path computed by $\csort$ is a shortest caterpillar path indeed.
 
@@
 The running time of $\csort$ is the same as for Bubble Sort, that is quadratic in $n$.
-Within Section~\ref{section:caterpillar_convex} we will use $\csort$ for some proofs.
-Therefore it is important to notice that at the end of each loop $i$ starting in Line~\ref{alg:csort:line:loop} of Algorithm~\ref{alg:csort}, the last $i$ elements of $L_{T'}$ are the same as the ones in $L_R$.
-This is a well known property of Bubble Sort that $\csort$ inherits from it.
+Though simple, this algorithm will be crucial for some proofs in Section~\ref{section:caterpillar_convex}.
 

### 2019-11-20 rnni_convexity.tex
update mdtree

@@
 The algorithm we are going to introduce here will be important for finding the radius of the $\rnni$ graph.
-In Section~\ref{section:diameter} we will prove that it computes a tree with maximum distance from a given caterpillar tree (Lemma~\ref{lemma:max_dist_caterpillar}), which implies that the radius of $\rnni$ equals its diameter (Corollary~\ref{corollary:radius}).
-Therefore, it is called $\mdtree$ (Maximum Distance Tree).
+It is called $\mdtree$, which stands for Maximum Distance Tree, as it computes a tree with maximum distance if the input tree is a caterpillar tree.
+This will be discussed in Section~\ref{section:diameter}.
 
-$\mdtree$ works as follows for an input tree $T$:
-At first, all taxa of $T$ are sorted in a list $L$ such that the ranks of their parents are non-decreasing.
-This representation is somehow similar to the list representation of caterpillar trees, but it is ambiguous since we do not specify in which order taxa appear in the list that share their parent.
-If the input is a caterpillar tree, it follows that $L[0]$ and $L[1]$ are the cherry taxa of $T$, in an arbitrary order, and the parent of taxon $L[i]$ has rank $i$ in $T$ for all $i = 2, \ldots, n-1$.
-$\mdtree$ builds the output tree $R$ as follows:
-Initially, $R$ only consists of a cherry with taxa $L[0]$ and $L[1]$ such that the parent of these taxa has rank $n-1$.
-All following taxa in $L$ are added to $R$ sequentially, according to their order in $L$, by creating a new internal node at one of the existing branches in $R$ such that this new node has rank one less than the previously lowest internal node in $R$.
-
-Algorithm~\ref{alg:max_dist_tree} provides pseudocode for this algorithm.
-We will see in Section~\ref{section:diameter} that it computes trees with maximum distance form any given caterpillar tree.
-But if the given tree is an arbitrary tree, the output tree might not have maximum distance to the input tree.
+$\mdtree$ (Algorithm~\ref{alg:max_dist_tree}) works as follows:
+At first, all taxa of the input tree $T$ are sorted in a list $L = [l_1, \ldots, l_n]$ such that the ranks of their parents are non-decreasing.
+Though this representation appears similar to the list representation of caterpillar trees, it is ambiguous since we do not specify how to order cherry taxa.
+$\mdtree$ constructs an output tree $R$ as follows:
+Initially, $R$ only consists of a cherry with taxa $l_1$ and $l_2$ such that the parent of these taxa has rank $n-1$.
+The remaining taxa are added to $R$ sequentially, according to their order in $L$, by creating a new internal node at one of the existing branches in $R$ such that this new node has rank one less than the previously lowest internal node in $R$.
 
@@
 \begin{algorithmic}[1]
-\STATE $R$ only contains cherry taxa $a, b$ of $T$ and has internal node with rank $n-1$
-\STATE In list $L$ all taxa $\{1,\ldots,n\}\setminus\{a,b\}$ are ordered such that the ranks of their parents in $T$ do not decrease \label{alg:mdtree_lineL}
-\FOR {taxon $a$ in $L$}
-\STATE Add an internal node as parent of $a$ on an edge in $R$ such that the new internal node has rank one less than the previously lowest internal node \label{alg:mdtree_lineAddTaxon}
+\STATE Construct list $L=[l_1, \ldots, l_n]$ of all taxa, ordered such that the ranks of their parents in $T$ do not decrease \label{alg:mdtree_lineL}
+\STATE Build tree $R$ with only two taxa $l_1, l_2$ that are children of the root, an internal node with rank $n-1$
+\FOR {$i = 1, \ldots, n$}
+\STATE Add an internal node as parent of $l_i$ on an edge in $R$ such that the new internal node has rank one less than the previously lowest internal node \label{alg:mdtree_lineAddTaxon}
 \ENDFOR
@@
 
-An important property of $\mdtree$, which we will use later on in Section~\ref{section:diameter}, is the fact that this algorithm is non-deterministic.
-Because the list computed in Line~\ref{alg:mdtree_lineL} the and the attachment edge of the new taxon in the tree in Line~\ref{alg:mdtree_lineAddTaxon} are not uniquely determined, the output tree $R$ of $\mdtree$ is not unique.
+Both the list computed in Line~\ref{alg:mdtree_lineL} the and the attachment edge of the new taxon in the tree in Line~\ref{alg:mdtree_lineAddTaxon} are not uniquely determined within $\mdtree$.
+Therefore, this algorithm is non-deterministic which is an important observation needed in Section~\ref{section:diameter} where this algorithm is used for finding the radius of $\rnni$.
 

### 2019-11-20 rnni_convexity.tex
geometry up to caterpillars

@@
 In Section~\ref{section:diameter} we will investigate diameter and radius of the $\rnni$ graph.
-Interestingly, the \emph{diameter}, that is the maximum distance $\Delta(\rnni) = \max \{d(T, R) \mid T, R \in \rnni\}$ between any two trees in the graph, of $\rnni$ equals its \emph{radius} $rad(\rnni) = \min\limits_T \max\limits_R d(T,R)$ in $\rnni$.
+Interestingly, the \emph{diameter} of $\rnni$, that is the maximum distance $\Delta(\rnni) = \max \{d(T, R) \mid T, R \in \rnni\}$ between any two trees in the graph, equals its \emph{radius} $rad(\rnni) = \min\limits_T \max\limits_R d(T,R)$.
 After proving this, we are going to consider the so-called Split Theorem, which was raised in \autocite{Gavryushkin2018-ol}.
-We will give a counterexample to disprove their version of the Split Theorem and claim an alternative version.
+We will give a counterexample to disprove their version of the Split Theorem and claim an alternative version, the Cluster Theorem (Conjecture~\ref{conjecture:cluster_theorem}).
 
@@
 Since the diameter is bounded from above by the maximum length of a path computed by $\findpath$, it is enough to find this maximum.
-The worst-case execution of $\findpath$ would require each of the $n-2$ clusters $C_i$, considered in Line~\ref{alg:find_path:line:cluster} of Algorithm~\ref{alg:find_path}, to be moved by $n-1-i$ $\rnni$ operations in $T'$.
-This can happen if each cluster $C_i$ has the root as $\mrca_{T'}(C_i)$.
-Then the rank of this most recent common ancestor must be changed from $n-1$ to $i$, which requires $n-1-i$ operations, according to Lemma~\ref{lemma:mrca_move}.
-Hence, the maximum length of such a path is bounded by $\sum\limits_{i = 1}^{n-2} i = \frac{(n-2)(n-1)}{2}$.
+Let us assume that $T$ and $R = [C_1, \ldots, C_{n-2}]$ are trees for which $\findpath$ computes a path of maximum length.
+That means that if $T'$ is the last tree of $p$ at the beginning of loop number $i$ in the execution of $\findpath(T,R)$, $\mrca_{T'}(C_i)$ is the root.
+Thus, there are $n-1-i$ $\rnni$ operations needed to move $C_i$ to its correct position.
+Hence, the maximum length of a path computed by $\findpath$ is bounded by $\sum\limits_{i = 1}^{n-2} n-1-i = \frac{(n-2)(n-1)}{2}$.
 \end{proof}
@@
 $T{\big|}_n$ denotes the restriction of tree $T$ to the set of taxa $\{1, \ldots, n\}$.
-In particular if $T$ is a tree on taxa $1, \ldots, n+1$ the tree $T{\big|}_n$ is obtained by deleting taxon $n+1$ and suppressing the thereby created node of degree two.
+In particular if $T$ is a tree on taxa $\{1, \ldots, n+1\}$ the tree $T{\big|}_n$ is obtained by deleting taxon $n+1$ and suppressing the thereby created node of degree two.
 
@@
 Second observe that any $\rnni$ move can change the rank of $\parent_T(n+1)$ by at most one.
-So we can follow that an $\rnni$ on $T$ move can decrease the rank of $\parent_T(n+1)$ by at most one.
+So we can follow that an $\rnni$ move on $T$ can decrease the rank of $\parent_T(n+1)$ by at most one.
 

### 2019-11-20 rnni_convexity.tex
up to thm caterpillar convexity

@@
 
-In this section we restrict our attention to the set of caterpillar trees, which we will represent throughout this section in the simplified list representation as introduced in Section~\ref{section:alg_csort}.
+In this section we restrict our attention to the set of caterpillar trees.
+Throughout this section the simplified list representation for these trees as introduced in Section~\ref{section:alg_csort} will be used.
 This type of tree is of particular interest in $\rnni$ because shortest paths between such trees in $\rnni$ differ from those in classical $\nni$ space.
-This in combination with the importance of caterpillar trees in the proof of $\np$-hardness of $\nni$ distances in \autocite{Dasgupta2000-xa} suggests that understanding the geometry of paths between caterpillar trees helps investigating the complexity of computing $\rnni$ distances.
-In particular, we are going to prove that one can always find a shortest path between two caterpillar trees that contains caterpillar trees only, meaning that the set of caterpillar trees is convex in $\rnni$.
-We discovered this by computing caterpillar paths with $\csort$ and comparing them with shortest path in $\rnni$, computed by the exact algorithm discussed in Section~\ref{section:alg_RNNI_graph}.
+Following this observation and the importance of caterpillar trees in the proof of $\np$-hardness of $\nni$ distances in \autocite{Dasgupta2000-xa}, understanding the geometry of paths between caterpillar trees seems essential for investigating the complexity of computing $\rnni$ distances.
 In the following we demonstrate why finding shortest paths between caterpillar trees builds the basis for understanding the difference between $\nni$ and $\rnni$.
 
-Therefore, we compare shortest paths between caterpillar trees in $\nni$ with those in $\rnni$.
-In $\nni$, building a cherry and moving it may give a path that is shorter than a path where the taxa move separately in as it is illustrated out in Figure~\ref{fig:NNI_vs_RNNI}.
+Let us compare shortest paths between caterpillar trees in $\nni$ with those in $\rnni$.
+It can happen in $\nni$ that building a cherry and moving it may give a path that is shorter than a path where the taxa move separately in as it is illustrated out in Figure~\ref{fig:NNI_vs_RNNI}.
 There is only one move needed to build and resolve a cherry, respectively, while the number of moves for moving the cherry is the same as for moving a single taxon in $\nni$.
-But in $\rnni$ space there are additional rank moves needed to change the rank of the internal node of the cherry before an $\nni$ move can resolve it.
+In $\rnni$ space however, there are additional rank moves needed to change the rank of the internal node of the cherry before an $\nni$ move can resolve it.
 This is due to the fact that $\nni$ moves are only allowed on edges of length one in $\rnni$.
-Hence a path where taxa move separately and no second cherry is built, is shortest in $\rnni$.
+As a consequence a path where taxa move separately and no new cherry is built is shortest in $\rnni$.
 
@@
 \caption{Paths between caterpillar trees $T$ and $R$: The solid paths are paths in $\rnni$, the dashed one is a path that is only possible in the plain $\nni$ graph.
-The upper path, which is the $\rnni$ path that is the natural extension of the shortest $\nni$ path, is longer than a shortest $\rnni$ path as the one at the bottom.}
+The upper path is a natural extension of the shortest $\nni$ path in $\rnni$.
+It is longer than a shortest $\rnni$ path as the one at the bottom which consists of caterpillar trees only.}
 \label{fig:NNI_vs_RNNI}
@@
 
-The above observation suggests that there is always a caterpillar path that is a shortest path between any two caterpillar trees in $\rnni$.
-Indeed, we are going to prove Theorem~\ref{thm:caterpillar_convex} within this section, which states that the set of caterpillar trees is convex in $\rnni$.
+The above observation suggests that there is a caterpillar path that is a shortest path between any two caterpillar trees in $\rnni$.
+Indeed, we are going to prove within this section that the set of caterpillar trees is convex in $\rnni$ (Theorem~\ref{thm:caterpillar_convex}).
 Note that this is not true for the $\nni$ space, which is clear from the example in Figure~\ref{fig:NNI_vs_RNNI}.
 
-Lemma~\ref{lemma:caterpillar_dist=diameter} regards caterpillar trees with distance $\frac{(n-1)(n-2)}{2}$ and will be needed for proving the convexity of the set of caterpillar trees in $\rnni$.
+The following lemma regards caterpillar trees with distance $\frac{(n-1)(n-2)}{2}$ and will be needed for proving the convexity of the set of caterpillar trees in $\rnni$.
 Notice that $\frac{(n-1)(n-2)}{2}$ is the upper bound for the diameter of $\rnni$ by Lemma~\ref{lemma:diameter_bound}.
-
-Within this chapter we will again use the list representation for caterpillar trees, which is a simplified version of the cluster representation of trees, as explained in Section~\ref{section:alg_csort}.
 
@@
 This easily follows from algorithm $\csort$:
-In each loop $i=1, \ldots, n-2$ of Algorithm~\ref{alg:csort} there are at most $n-1-i$ exchanges of taxa.
-This is due to the fact that after loop number $i$ the last $i$ elements in the list representation of the current tree on $p$ are sorted correctly as in the destination tree.
-Hence there are no more exchanges of pairs of taxa needed within the last $i-1$ of the sequences in loop $i$, and so only $n-1-i$ swaps of elements might appear within the first $n-i$ elements of the list.
-It follows that the maximum length of a path computed by $\csort$ is $\sum\limits_{k=1}^{n-2} k = \frac{(n-1)(n-2)}{2}$.
+In each loop $i=1, \ldots, n-2$ of $\csort$ (Algorithm~\ref{alg:csort}) there are at most $n-1-i$ pairs of taxa that swap positions.
+This is due to the fact that after loop number $i$ the last $i$ elements in the list representation of the current tree on $p$ are in the same place as in the destination tree.
+As a result there are no further exchanges of pairs of taxa within the last $i-1$ taxa of the tree in loop $i$, which means that at most $n-1-i$ pairs of taxa exchange then.
+It follows that the maximum length of a path computed by $\csort$ is $\sum\limits_{i=1}^{n-2} n-1-i = \frac{(n-1)(n-2)}{2}$.
 Since there cannot be a caterpillar path shorter than $d(T,R) = \frac{(n-1)(n-2)}{2}$, it follows $d_c(T,R) = \frac{(n-1)(n-2)}{2}$.
@@
 Induction basis: $n=3$\\
-There exist three trees on three taxa: $[1,2,3], [1,3,2]$ and $[2,3,1]$.
+There are three trees on three taxa.
 They all have distance one from another and are caterpillar trees.
@@
 Let $R$ be a caterpillar tree with distance $\frac{n(n-1)}{2}$ to $T$.
-Then each of the $\frac{n(n-1)}{2}$ pairs of taxa that is considered when running $\csort$ to compute a path $p$ from $T$ to $R$ exchanges.
-In particular, taxon $n+1$ exchanges with a taxon in each of the $n+1$ loops in $\csort$.
-We can conclude that taxon $n+1$, whose parent has highest rank in $T$, is part of the cherry of $R$.
+It follows that in each loop of $\csort$ the maximum number of taxon swaps is reached which implies that every taxon will swap with taxon $n+1$ at some point.
+We can conclude that this taxon $n+1$ is part of the cherry of $R$.
 Let us now consider the trees $T{\big|}_n$ to $R{\big|}_n$ that result from $T$ and $R$ by deleting taxon $n+1$.
@@
 We are now using this observation to prove that the distance between $T$ and $R$ equals $\frac{n(n-1)}{2}$.
-It is obvious that it gives an upper bound for the distance as it is $\Delta(\rnni_{n+1}) \leq \frac{n(n-1)}{2}$ for the $\rnni$ graph $\rnni_{n+1}$ on $n+1$ taxa by Lemma~\ref{lemma:diameter_bound}.
-So we now need to prove that $\frac{n(n-1)}{2}$ is a lower bound for the distance between $T$ and $R$ as well.
+This obviously gives an upper bound for the distance as it is $\Delta(\rnni_{n+1}) \leq \frac{n(n-1)}{2}$ for the $\rnni$ graph on $n+1$ taxa by Lemma~\ref{lemma:diameter_bound}.
+It is left to prove that $\frac{n(n-1)}{2}$ is a lower bound for the distance between $T$ and $R$ as well.
 According to Lemma~\ref{lemma:distance_delete_taxon}, the distance between $T{\big|}_n$ and $R{\big|}_n$ is at least $d(T{\big|}_n, R{\big|}_n) \leq d(T,R) - |\rank_T(\parent_T(n+1)) - \rank_R(\parent_R(n+1))|$.

### 2019-11-20 rnni_convexity.tex
update diameter section

@@
 With the results of Section~\ref{section:caterpillar_convex} we can easily get the exact diameter of the $\rnni$ graph and hence improve the bounds for the diameter given in Theorem 7 of \autocite{Gavryushkin2018-ol}.
-At first we consider the diameter of the $\rnni$ graph.
-Lemma~\ref{lemma:diameter_bound} already gives an upper bound for the diameter.
-We will now show that this is not only a bound, but the exact diameter of $\rnni$.
+In the following we show that the upper bound of the diameter provided by Lemma~\ref{lemma:diameter_bound} actually is the exact diameter of $\rnni$.
 
 \begin{lemma}
-There are two caterpillar trees in $\rnniu$ with distance $\frac{(n-1)(n-2)}{2}$.
+There are two caterpillar trees in $\rnni$ with distance $\frac{(n-1)(n-2)}{2}$.
 \label{lemma:caterpillar_diameter}
@@
 \begin{proof}
-Let $T = [1,n,2,3,\ldots,n-1]$ and $R = [n-1,n,n-2,n-3, \ldots, 1]$ be two caterpillar trees.
+Let $T = [1,2,3,\ldots,n]$ and $R = [n-1,n,n-2,n-3, \ldots, 1]$ be two caterpillar trees.
 According to Theorem~\ref{thm:caterpillar_convex}, there is a shortest path between $T$ and $R$ that is a caterpillar path.
 We can compute a shortest caterpillar path by using $\csort$ as explained in Section~\ref{section:caterpillar_convex}.
-When running this algorithm to compute a path from $T$ to $R$, each taxon $i$ ($i \in \{1, \ldots, n\}$) swaps $n-1-i$ times with its upper neighbour.
-Therefore, the length of a shortest caterpillar path is $\sum\limits_{i=1}^{n-1}(n-1-i) = \sum\limits_{i=0}^{n-2}i = \frac{(n-1)(n-2)}{2}$.
+When running this algorithm to compute a path from $T$ to $R$, each taxon $i$ ($i = 1, \ldots, n-1$) swaps $n-1-i$ times with its upper neighbour.
+Therefore, the length of a shortest caterpillar path is $\sum\limits_{i=1}^{n-1}(n-1-i)  = \frac{(n-1)(n-2)}{2}$.
 \end{proof}
@@
 Let $T$ be a caterpillar tree.
-A tree $R = \mdtree(T)$ has distance $d(T,R) = \Delta(\rnni)$ to $T$.
+Any tree $R = \mdtree(T)$ has distance $d(T,R) = \Delta(\rnni)$ to $T$.
 \label{lemma:max_dist_caterpillar}
@@
 Let us now assume that $T$ is a tree on $n+1$ taxa.
-In the list $L$ produced Line~\ref{alg:mdtree_lineL} in $\mdtree(T)$ (Algorithm~\ref{alg:max_dist_tree}) taxon $n+1$ is the last one, because its parent, the root, has the highest rank in $T$.
+In the list $L$ computed in $\mdtree$ (Line~\ref{alg:mdtree_lineL} of Algorithm~\ref{alg:max_dist_tree}) taxon $n+1 $ is the last one as its parent, the root, has the highest rank in $T$.
 This means that this taxon $n+1$ is the last one added to $R$ as in Line~\ref{alg:mdtree_lineAddTaxon} of this algorithm and therefore, it has parent with rank one in $R$.
@@
 It is not hard to see that $R|_n$ can be received from $T|_n$ by applying $\mdtree$:
-$R|_n$ can be constructed from input tree $T|_n$ in the same way as $R$ can be constructed from $T$, since the list $L$ computed in Line~\ref{alg:mdtree_lineL} of Algorithm~\ref{alg:max_dist_tree} for $T|_n$ is the same as the one for $T$, just the last element is missing.
+$R|_n$ can be constructed from input tree $T|_n$ in the same way as $R$ can be constructed from $T$, since the list $L$ computed in Line~\ref{alg:mdtree_lineL} of Algorithm~\ref{alg:max_dist_tree} for $T|_n$ is the same as the one for $T$, but without the last element.
 So we can follow from the induction hypothesis and Lemma~\ref{lemma:distance_delete_taxon} that it is $d(T,R) \geq d(T|_n,R|_n) + (n-1) = \frac{(n-1)(n-2)}{2} = \frac{n(n-1)}{2}$, which concludes the proof of the lemma.
@@
 In Section~\ref{alg:max_dist_tree} we already discussed that $\mdtree$ is a non-deterministic algorithm.
-In fact, the output tree of $\mdtree$ can have any tree topology, since there is no restriction on where to add the new taxon in the already existing tree in Line~\ref{alg:mdtree_lineAddTaxon} of Algorithm~\ref{alg:max_dist_tree}.
+In fact, the output tree of $\mdtree$ can have any tree topology, since there is no restriction on where to add the new taxon in the already existing tree (Line~\ref{alg:mdtree_lineAddTaxon} of Algorithm~\ref{alg:max_dist_tree}).
 This means that for each caterpillar input tree $\mdtree$ can compute a tree with any tree topology that has maximum distance from the input tree.
-And because permuting labels of trees does not change the distance between them, there exists for every tree $R$ on $n$ taxa a caterpillar tree $T$ with distance $d(T,R) = \Delta(\rnni) = \frac{(n-1)(n-2)}{2}$.
-This gives us Corollary~\ref{corollary:radius}.
+And as permuting labels of trees does not change the distance between them, there is for every tree $R$ on $n$ taxa a caterpillar tree $T$ with distance $d(T,R) = \Delta(\rnni) = \frac{(n-1)(n-2)}{2}$.
+Following this, we get Corollary~\ref{corollary:radius}.
 

### 2019-11-21 rnni_convexity.tex
update section cluster thm

@@
 And as permuting labels of trees does not change the distance between them, there is for every tree $R$ on $n$ taxa a caterpillar tree $T$ with distance $d(T,R) = \Delta(\rnni) = \frac{(n-1)(n-2)}{2}$.
-Following this, we get Corollary~\ref{corollary:radius}.
+By this we obtain Corollary~\ref{corollary:radius}.
 
@@
 So far it is not known whether computing distances in $\rnni$ is $\np$-hard or not.
-The natural approach to solve this question is to compare $\rnni$ and $\nni$ graph to see whether the proof of $\np$-hardness of $\nni$ can be transferred to $\rnni$.
+The natural approach to solve this question is to compare $\rnni$ and $\nni$ graph to see whether the proof of $\np$-hardness of $\nni$ applies to $\rnni$ as well.
 Within this section we will explain why this is not the case.
-Therefore we consider an example of shortest paths in $\nni$ that builds the key of the proof of its $\np$-hardness.
-We will see that this example cannot be transferred to $\rnni$.
-Following this, we will introduce the so-called Split Theorem, which generalises that example, and has been conjectured in \autocite{Gavryushkin2018-ol}.
-However, we present a counterexample disproving the Split Theorem (Conjecture~\ref{conjecture:split_theorem}) in its original version and claiming the Cluster Theorem (Conjecture~\ref{conjecture:cluster_theorem}) as alternative version.
+Therefore we consider an example of shortest paths in $\nni$ that builds the key of the proof of its $\np$-hardness and explain why this example cannot be transferred to $\rnni$.
+Following this, we will introduce the so-called Split Theorem (Conjecture~\ref{conjecture:split_theorem}), which has been conjectured in \autocite{Gavryushkin2018-ol}.
+However, we present a counterexample disproving the Split Theorem in its original version and claiming the Cluster Theorem (Conjecture~\ref{conjecture:cluster_theorem}) as alternative version.
 
-The proof of NP-completeness of $\nni$ in \autocite{Dasgupta2000-xa} is based on the fact that Theorem~\ref{thm:split_nni} holds for $\nni$.
+The proof of $\np$-completeness of $\nni$ in \autocite{Dasgupta2000-xa} is based on the fact that Theorem~\ref{thm:split_nni} holds for $\nni$.
 
@@
 In the proof presented in \autocite{Li1996-zw} for Theorem~\ref{thm:split_nni} an example of two trees that share a cluster but have no shortest path that preserves this cluster is given.
-In Figure~\ref{fig:NNI_NP_proof} there is an illustration of this example.
-The reason that this proof does not work for $\rnni$ is similar to the reason that caterpillar trees are convex in $\rnni$, but not in $\nni$.
-Both trees $T$ and $R$ in the example of \autocite{Li1996-zw} have $2n$ taxa and consist of two caterpillar trees that share the root as parent.
+We describe this example in the following a simplified illustration of it can be found in Figure~\ref{fig:NNI_NP_proof}.
+In the example of \autocite{Li1996-zw} trees $T$ and $R$ on $2n$ taxa are considered where each of the trees consist of two caterpillar trees that share the root as parent.
 The clusters given by the two nodes adjacent to the root are $\{1, \ldots, n\}$ and $\{n+1, \ldots, 2n\}$ in both trees $T$ and $R$.
-There is a shortest path in $\nni$ where at first $T$ is transformed to a tree $T'$ with $n$ cherries where each cherry contains one taxon of $\{1, \ldots, n\}$ and one taxon of $\{n+1, \ldots, 2n\}$.
-Afterwards, $T'$ is transformed to a tree $R'$ still containing the same cherries as $T'$ but in a way that resolving all these cherries of $R'$ results in $R$.
+\autocite{Li1996-zw} prove that on a shortest path between $T$ and $R$ in $\nni$ at first $T$ is transformed to a tree $T'$ with $n$ cherries, each of which contains one taxon of $\{1, \ldots, n\}$ and one taxon of $\{n+1, \ldots, 2n\}$.
+Afterwards, $T'$ is transformed to a tree $R'$ containing the same cherries as $T'$ but in a way that resolving all these cherries of $R'$ results in $R$.
 For proving that this path is shorter than any path preserving the clusters $\{1, \ldots, n\}$ and $\{n+1, \ldots, 2n\}$, \autocite{Li1996-zw} use the fact that $d_{\nni}(T',R') = d_{\nni}(T{\big|}_n, R{\big|}_n)$ where $T{\big|}_n$ and $R{\big|}_n$ are $T$ and $R$ restricted to taxa $\{1,\ldots,n\}$, respectively.
-However, this does not work in $\rnni$.
-Similar to what we observed in the example depicted in Figure~\ref{fig:NNI_vs_RNNI}, there are lots of additional rank moves needed on a path in $\rnni$ that similar to the one in $\nni$ as described above.
-Therefore, a path where the two caterpillar trees are sorted separately is shorter in $\rnni$.
+For $\rnni$ there are lots of additional rank moves needed on a path that follows the principle of the shortest path in $\nni$ as described above.
+Similar to what we observed in the example depicted in Figure~\ref{fig:NNI_vs_RNNI}, this results in a path that is longer than a shortest $\rnni$ path.
+Therefore, a path where the two caterpillar subtrees of $T$ and $R$ are sorted separately is shorter in $\rnni$.
 
@@
 \caption{Example of a shortest path between two trees $T$ and $R$ in $\nni$.
-All leaves labelled with dots are labelled by taxa in $\{1, \ldots, n\}$, squares represent taxa in $\{n+1, \ldots, 2n\}$.
-In \autocite{Li1996-zw} it is proven that there is such a labelling that ensures that there is no shortest path in $\nni$ where the clusters shared between $T$ and $R$ are preserved.
-Instead, the depicted path where $n$ cherries are built ad then sorted before being resolved is a shortest path}
+All leaves labelled with dots represent taxa in $\{1, \ldots, n\}$, squares represent taxa in $\{n+1, \ldots, 2n\}$.
+In \autocite{Li1996-zw} it is proven that there is such a labelling that ensures that there is no shortest path in $\nni$ where the clusters (dots vs.
+squares) shared between $T$ and $R$ are preserved.
+Instead, the depicted path where $n$ cherries are built and then sorted before being resolved is a shortest path}
 \label{fig:NNI_NP_proof}
@@
 
-Since the example given in \autocite{Li1996-zw}, which proves Theorem~\ref{thm:split_nni} in $\nni$, does not work in $\rnni$, a contrary statement was conjectured in \autocite{Gavryushkin2018-ol}.
+Since the example given in \autocite{Li1996-zw}, which can be used to prove Theorem~\ref{thm:split_nni} in $\nni$, does not work in $\rnni$, a contrary statement was conjectured in \autocite{Gavryushkin2018-ol}.
 For understanding this conjecture it is important to know the connection between edges in a tree and so-called \emph{splits}, which are bipartitions of the taxon set of a tree:
@@
 
-However, we found a simple counterexample to this conjecture, provided in Figure~\ref{fig:splitthm_counterexample}.
+We found a simple counterexample to this conjecture, provided in Figure~\ref{fig:splitthm_counterexample}.
 
@@
 \caption{The split $123|45$ is present in $T$ and $R$, but the path at the top is a shortest path (computed by $\csort$) where none of the trees contains this split.
-However, this split is maintained on the path at the bottom, which is a shortest path as well.}
+On the path at the bottom, which is a shortest path as well, this split is maintained.}
 \label{fig:splitthm_counterexample}
@@
 This is motivated by the fact that rooted phylogenetic trees can uniquely be represented by sets of clusters \autocite{Steel2016-ye}, but not by sets of splits as they cannot define the position of the root.
-Moreover, a ranked tree can be uniquely determined by ordering the clusters such that the ranks of the internal nodes inducing those clusters increase.
-We already used this representation of trees for the algorithms in Section~\ref{section:algorithms}.
+Note that trees $T$ and $R$ in Figure~\ref{fig:splitthm_counterexample} induce the same set of splits, but share no cluster.
 
@@
 
-Note that trees $T$ and $R$ in Figure~\ref{fig:splitthm_counterexample} induce the same set of splits, but have distinct clusters.
-Furthermore, this counterexample can be used to prove that the Cluster Theorem does not hold in $\nni$.
+For providing further evidence that the Cluster Theorem holds in $\rnni$, we can show computationally that it holds for the $\rnni$ graph on small trees with up to six taxa.
+We did this by computing subgraph of $\rnni$ only containing trees that share a cluster and compare distances within this graph with distances in the whole $\rnni$ graph \autocite{}.
+\todo{LC: github link}
+We computed these graphs according to the algorithm of \autocite{Gavryushkin2018-ol}, discussed in Section~\ref{section:algorithms}.
 
-For providing further evidence that the Cluster Theorem holds in $\rnni$, we can show computationally that it holds for the $\rnni$ graph on small trees with up to seven taxa.
-\todo{LC: We still need to check 7 taxa.}
-We are doing this in the following way.
-We focus on the subset of trees that induce share a cluster.
-It is sufficient to consider clusters of the shape $\{1, \ldots, m\}$ for $2 \leq m \leq n-1$, because of the symmetry of the space:.
-distances $d(T_1,R_1)$ and $d(T_2,R_2)$ between two pairs of trees $(T_1,R_1)$ and $(T_2,R_2)$ are equal if one can receive $T_2$ from $T_1$ by permuting the taxa the same way as it is necessary for receiving $R_2$ from $R_1$.
-Therefore, we compute for all $m = 2, \ldots, n-1$ exact $\rnni$ distances between all pairs of trees containing the cluster $\{1, \ldots, m\}$, using Dijkstra's algorithm \autocite{Dijkstra1959-ph} on the $\rnni$ graph, which we can compute as explained in Section~\ref{section:alg_RNNI_graph}.
-Additionally, we compute the subgraph of $\rnni$ induced by the subset of trees that include the cluster $\{1, \ldots, m\}$.
-With the Floyd-Warshall algorithm we compute all pairwise distances between the trees in this subgraph and compare them with the distances in $\rnni$.
-With this approach it is possible to show that the Cluster Theorem holds for up to seven taxa.
-\todo{LC: We still need to check 7 taxa.}
+% \todo{LC: We still need to check 7 taxa.}
+% We are doing this in the following way.
+% We focus on the subset of trees that induce share a cluster.
+% It is sufficient to consider clusters of the shape $\{1, \ldots, m\}$ for $2 \leq m \leq n-1$, because of the symmetry of the space:.
+% distances $d(T_1,R_1)$ and $d(T_2,R_2)$ between two pairs of trees $(T_1,R_1)$ and $(T_2,R_2)$ are equal if one can receive $T_2$ from $T_1$ by permuting the taxa the same way as it is necessary for receiving $R_2$ from $R_1$.
+% Therefore, we compute for all $m = 2, \ldots, n-1$ exact $\rnni$ distances between all pairs of trees containing the cluster $\{1, \ldots, m\}$, using Dijkstra's algorithm \autocite{Dijkstra1959-ph} on the $\rnni$ graph, which we can compute as explained in Section~\ref{section:alg_RNNI_graph}.
+% Additionally, we compute the subgraph of $\rnni$ induced by the subset of trees that include the cluster $\{1, \ldots, m\}$.
+% With the Floyd-Warshall algorithm we compute all pairwise distances between the trees in this subgraph and compare them with the distances in $\rnni$.
+% With this approach it is possible to show that the Cluster Theorem holds for up to seven taxa.
+% \todo{LC: We still need to check 7 taxa.}
 

### 2019-11-21 rnni_convexity.tex
proof caterpillar convexity

@@
 We prove this Theorem by backwards induction on the caterpillar distance $l = d_c(T,R)$ between caterpillar trees $T$ and $R$.
-For this backwards induction the maximum caterpillar distance, which is $\frac{(n-1)(n-2)}{2}$ (Lemma~\ref{lemma:caterpillar_dist=diameter}), is the base case.
+For this backwards induction the maximum caterpillar distance $\frac{(n-1)(n-2)}{2}$ (Lemma~\ref{lemma:caterpillar_dist=diameter}) is the base case.
 
@@
 With Lemma~\ref{lemma:caterpillar_dist=diameter} one can see that for each pair of caterpillar trees $T$, $R$ with caterpillar distance $d_c(T,R) = \frac{(n-1)(n-2)}{2}$, it is $d(T,R) = d_c(T,R)$.
-This means that there is a shortest path between $T$ and $R$ that is a caterpillar path.
+Thus there is a shortest path between $T$ and $R$ that is a caterpillar path.
 
@@
 
-Since $l < \frac{(n-1)(n-2)}{2}$, there must be one of the loops starting in Line~\ref{alg:csort:line:loop} of Algorithm~\ref{alg:csort} ($\csort$) where not every considered pair of taxa swaps positions.
-Let us assume that loop number $k$ is the first one where the number of swaps is not maximal.
-It might well be that this happens in the first loop $k=1$ already.
-Since we assume that the number of swaps in loop $k$ is not maximal, there is a pair $a,b$ of elements that is compared in the algorithm but does not swap positions.
-And because we consider the first time that this happens, $a$ is either one of the first two elements in the sequence, or it previously swapped with an element $c$ that has been between $a$ and $b$ before loop number $k$.
+Since $l < \frac{(n-1)(n-2)}{2}$, there must be a loop in $\csort$ (Line~\ref{alg:csort:line:loop} of Algorithm~\ref{alg:csort}) where not every pair of taxa that is compared exchanges.
+Let us consider the first pair of taxa $a,b$ that is compared but does not swap positions, where $a$ is below $b$.
+Since the first time that this happens, $a$ is either one of the first two elements in the sequence, or it previously swapped with an element $c$ that has been between $a$ and $b$ before.
 
-Let us first consider the latter case.
-The sequence $L_{T'}$ at the time where $a$ and $b$ are compared but not swapped, looks as follows:
+We start with considering the latter case.
+The list representation of the tree at the point where $a$ and $b$ are compared but not swapped, looks as follows:
+\[[\ldots, c, a, b, \ldots, x_k, \ldots, x_1]\]
 
-\vspace{12pt}
-\begin{tabular}{c|c|c|c|c|c|c|c|c|}
-\cline{2-9}
-$L_{T'}$: & $\ldots$ & $c$ & $a$ & $b$ & $\ldots$& $x_{k-1}$ & $\ldots$ & $x_1$\\
-\cline{2-9}
-\end{tabular}
-\vspace{12pt}
+Note that here we assume that the elements $x_1, \ldots, x_k$ are in the same place as they are in the destination tree $R$.
+It could be that $k=1$, which means that there are no such elements.
 
-Note that in this illustration of $L_{T'}$ the elements $x_1, \ldots, x_{k-1}$ are in the same place as they are in $L_R$.
-However, it could be that $k=1$, which means that there are no such elements at the end of $L_{T'}$ that are already sorted the same way as in the destination tree.
-
-The last move before comparing $a$ and $b$ is a swap of $a$ with an element $c$.
-Since $a$ and $c$ swapped, but $a$ and $b$ did not, we can follow that $b$ and $c$ appear in $L_R$ in the same order as they do in $L_{T'}$.
-Furthermore, $c$ and $b$ must have been neighbours in $L_T$, because we assume that the comparison of $a$ and $b$ is the first one where the compared elements are not swapped in the sequence.
-Let now $L_{\hat T}$ be a list that equals $L_T$ with the only exception that $b$ and $c$ are swapped.
-We can follow from our observations above, that the distance between $T$ and $\hat T$ is $d(T, \hat T) = 1$ and that $\csort$ needs one move more for a path from $\hat T$ to $R$ than from $T$ to $R$.
+Since $a$ and $c$ swap, but $a$ and $b$ do not, we can follow that $b$ and $c$ appear in $R$ in the same order as they do in the current tree.
+Furthermore, $c$ and $b$ must have been neighbours in $T$, because we assume that the comparison of $a$ and $b$ is the first one where the compared elements are not swapped in the sequence.
+Let now $\hat T$ be a tree that equals $T$ with the only exception that $b$ and $c$ are exchanged.
+We can follow from our observations above that the distance between $T$ and $\hat T$ is $d(T, \hat T) = 1$ and that $\csort$ needs one move more for a path from $\hat T$ to $R$ than from $T$ to $R$.
 This results in $d_c(\hat T, R) = d(T,R) + 1$.
@@
 Let us now consider the other case:
-of the two elements $a$ and $c$, that are compared but not swapped, one, say $a$, is either the first or the second one in $L_{T'}$.
+of the two taxa that are compared but do not swap positions, one is part of the cherry current tree.
+Let $a$ and $b$ be the cherry taxa and $c$ their neighbour such that neither $a$ nor $b$ echanges with $c$.
 Then the list looks like this:
-
-\vspace{12pt}
-\begin{tabular}{c|c|c|c|c|c|c|c|c|}
-\cline{2-8}
-$L_{T'}$: & $a$ & $b$ & $c$ & $\ldots$ & $x_{k-1}$ & $\ldots$ & $x_1$\\
-\cline{2-8}
-\end{tabular}
-\vspace{12pt}
+\[[a, b, c, \ldots, x_k, \ldots, x_1]\]
 
 If there is no swap of either $a$ or $b$ with $c$, we can construct a tree $\hat T$ with caterpillar distance $d_c(\hat T, R) = l+1$ and $d(T,\hat T) = 1$ in a similar way as in the previous case.
-In the list $L_T$ representing the start tree $T$, either $a$ or $b$ must be neighbour of $c$.
+In the list representing the start tree $T$, either $a$ or $b$ must be neighbour of $c$.
 This is due to the fact that we are currently considering the first comparison of elements where there is no swap.
-Without loss of generality we assume that $a$ is neighbour of $c$ in $L_T$.
-Then an exchange of $a$ and $c$ in $L_T$ yields a list $L_{\hat T}$ representing a tree $\hat T$ with the desired properties.
+We can assume without loss of generality that $a$ is neighbour of $c$ in $T$.
+Then an exchange of $a$ and $c$ in $T$ yields a tree $\hat T$ with the desired properties.
 As in the previous case we can then apply the induction hypothesis on $\hat T$ and conclude that there is a shortest path between $T$ and $R$ that only consists of caterpillar trees.
@@
 \centering
-\includegraphics[width=\textwidth]{NNI_NP_proof}
+\includegraphics[width=.8\textwidth]{NNI_NP_proof}
 \vspace{12pt}
@@
 
-\section{Ideas}
+% \section{Ideas}
+%
+% [This section includes some ideas that we could possibly include in the paper]
+%
+% Using the same argument as in the proof of Lemma~\ref{lemma:distance_delete_taxon} we can establish Proposition~\ref{proposition:lower_bound_distance} that provides a lower bound on the distance between trees.
+%
+% \begin{proposition}
+% If trees $T$ and $R$ and taxon $x$ are such that $|\rank_T(\parent_T(x)) - \rank_R(\parent_R(x))| = \delta$ then $d(T,R) \geq \delta$.
+% \label{proposition:lower_bound_distance}
+% \end{proposition}
+%
+%
+% The paths computed by $\findpath$ depend on the order of input trees $T$ and $R$:
+% $\findpath(T,R)$ does not necessarily produce the reverse path of $\findpath(R,T)$.
+% An example can be found in Figure~\ref{fig:findpath_not_symmetric}.
+% However, simulations suggest that the lengths of the two paths computed by $\findpath$ for one pair of trees are equal.
+%
+% \begin{conjecture}
+% Let $T$ and $R$ be ranked trees.
+% The lengths of paths $\findpath(T,R)$ and $\findpath(R,T)$ are equal.
+% \end{conjecture}
+%
+% \begin{figure}[H]
+% \centering
+% \includegraphics[width=0.7\textwidth]{findpath_not_symmetric}
+% \vspace{12pt}
+% \caption{Two paths computed by $\findpath$. At the top $\findpath(T,R)$, at the bottom $\findpath(R,T)$}
+% \label{fig:findpath_not_symmetric}
+% \end{figure}
+%
+% % idea for description of max distance caterpillar trees:
+% \begin{lemma}
+% Let $T$ and $R$ be caterpillar trees.
+% It is $d_c(T,R) = \Delta(\rnni)$ if, and only if, $T$ and $R$ share no induced triplet.
+% \todo{define induced triplets}
+% \end{lemma}
+%
+% This Lemma does only hold for caterpillar trees and not for trees with more than one cherry.
+% Furthermore, triplets do not represent ranked trees uniquely as they do not contain information about the ranks of internal nodes (but they can uniquely define caterpillar trees as these only have one cherry)
+%
+% \begin{proof}
+% %TODO
+% % Insert the proof here
+% Fix $T = (\ldots(1,2) \ldots ,n)$, use induction on $n$ and $\csort$ (taxon $n$ must be in cherry of $R$ if distance is max).
+% \end{proof}
+%
+% According to Algorithm~\ref{alg:max_dist_tree}, the number of trees with maximum distance to a given caterpillar tree $T$ is $(n-1)!$.
+% The number of caterpillar trees with distance $\Delta(\rnni)$ from a given caterpillar tree is $2^{n-2}$.
+%
+%
+% \subsection{Partition lattice}
+%
+% % Define Partition Lattice $\Pi_n$ and max chains in that lattice!
+%
+% \begin{theorem}
+% The $\rnni$ graph on $n$ taxa is isomorphic to the graph of maximal chains of the partition lattice $\Pi_n$ where two maximal chains are connected by an edge if and only if they differ by exactly one partition.
+% The corresponding metric spaces are isometric.
+% \end{theorem}
 
-[This section includes some ideas that we could possibly include in the paper]
-
-Using the same argument as in the proof of Lemma~\ref{lemma:distance_delete_taxon} we can establish Proposition~\ref{proposition:lower_bound_distance} that provides a lower bound on the distance between trees.
-
-\begin{proposition}
-If trees $T$ and $R$ and taxon $x$ are such that $|\rank_T(\parent_T(x)) - \rank_R(\parent_R(x))| = \delta$ then $d(T,R) \geq \delta$.
-\label{proposition:lower_bound_distance}
-\end{proposition}
-
-The paths computed by $\findpath$ depend on the order of input trees $T$ and $R$:
-$\findpath(T,R)$ does not necessarily produce the reverse path of $\findpath(R,T)$.
-An example can be found in Figure~\ref{fig:findpath_not_symmetric}.
-However, simulations suggest that the lengths of the two paths computed by $\findpath$ for one pair of trees are equal.
-
-\begin{conjecture}
-Let $T$ and $R$ be ranked trees.
-The lengths of paths $\findpath(T,R)$ and $\findpath(R,T)$ are equal.
-\end{conjecture}
-
-\begin{figure}[H]
-\centering
-\includegraphics[width=0.7\textwidth]{findpath_not_symmetric}
-\vspace{12pt}
-\caption{Two paths computed by $\findpath$. At the top $\findpath(T,R)$, at the bottom $\findpath(R,T)$}
-\label{fig:findpath_not_symmetric}
-\end{figure}
-
-% idea for description of max distance caterpillar trees:
-\begin{lemma}
-Let $T$ and $R$ be caterpillar trees.
-It is $d_c(T,R) = \Delta(\rnni)$ if, and only if, $T$ and $R$ share no induced triplet.
-\todo{define induced triplets}
-\end{lemma}
-
-This Lemma does only hold for caterpillar trees and not for trees with more than one cherry.
-Furthermore, triplets do not represent ranked trees uniquely as they do not contain information about the ranks of internal nodes (but they can uniquely define caterpillar trees as these only have one cherry)
-
-\begin{proof}
-%TODO
-% Insert the proof here
-Fix $T = (\ldots(1,2) \ldots ,n)$, use induction on $n$ and $\csort$ (taxon $n$ must be in cherry of $R$ if distance is max).
-\end{proof}
-
-According to Algorithm~\ref{alg:max_dist_tree}, the number of trees with maximum distance to a given caterpillar tree $T$ is $(n-1)!$.
-The number of caterpillar trees with distance $\Delta(\rnni)$ from a given caterpillar tree is $2^{n-2}$.
-
-\subsection{Partition lattice}
-
-% Define Partition Lattice $\Pi_n$ and max chains in that lattice!
-
-\begin{theorem}
-The $\rnni$ graph on $n$ taxa is isomorphic to the graph of maximal chains of the partition lattice $\Pi_n$ where two maximal chains are connected by an edge if and only if they differ by exactly one partition.
-The corresponding metric spaces are isometric.
-\end{theorem}
+\section{Discussion}
 

### 2019-11-24 rnni_convexity.tex
first ideas abstract + discussion

@@
 \begin{abstract}
-
+Within this paper we consider a graph of ranked trees, that are rooted binary phylogenetic trees where internal nodes are ordered.
+The ranking of these internal nodes can be seen as a discretisation of trees where internal nodes are assigned divergence times.
+For phylogenetic reconstruction such as Markov Chain Monte Carlo (MCMC) methods, it is important to understand the geometry of tree spaces.
+This motivates our work, which focusses on understanding the geometry of the space of ranked trees, the so-called ranked nearest neighbour interchange ($\rnni$) graph.
+Moreover, we are interested in understanding the complexity of computing distances in this graph.
+We will see that even though there are similarities between the $\rnni$ graph on ranked trees and the classical nearest neighbour interchange ($\nni$) graph on unranked phylogenetic trees, some significant differences between these tree spaces suggest that there is a difference in the complexity of computing distances.
 \end{abstract}
@@
 \todo{LC: github link}
-We computed these graphs according to the algorithm of \autocite{Gavryushkin2018-ol}, discussed in Section~\ref{section:algorithms}.
+We computed these graphs according to the algorithm of \autocite{Gavryushkin2018-ol} as discussed in Section~\ref{section:algorithms}.
 
@@
 
+Within this paper we established some geometric properties of the $\rnni$ graph.
+We furthermore pointed out some differences between this graph on ranked trees and the well studied $\nni$ graph on unranked trees.
+This is of particular interest for analysing the complexity of computing distances in $\rnni$ as this problem is known to be $\np$-complete in $\nni$.
+Our finding that the set of caterpillar trees is convex in $\rnni$ (Theorem~\ref{thm:caterpillar_convex}) suggests that there is a difference in complexity between these two tree spaces.
+This leaves hope that computing distances in $\rnni$ is, contrary to this problem in $\nni$, not $\np$-complete.
+The question of how hard this problem actually is still remains open and will be subject of further research.
+
+Our work in this paper highlights the importance of algorithms for computing paths in $\rnni$ in order to prove geometric properties such as diameter and radius.
+One of the algorithms that has been introduced, $\findpath$, computed exact distances for all pairs of trees on up to six taxa.
+For future investigations of the $\rnni$ space this algorithm might be a useful tool.
+
 \printbibliography

### 2019-11-25 rnni_convexity.tex
missing linebreaks + reference

@@
 A consequence is that tree search and sampling algorithms pose a significant challenge even for moderately sized trees
-\todo{LC: Alex, is this the citation you wanted?}
-\autocite{Whidden2018-tk}.
+\autocite{Whidden2016-kl}.
 
@@
 
-Step $0$. $p = [T_0]$.
+Step $0$.
+$p = [T_0]$.
 Consider the node in $T_0$ that induces the smallest cluster containing all elements of $C_1$.
@@
 
-Step $k$. $p = [T_0, \ldots, T_l]$ where $l$ is the total number of trees that have been added to $p$ in the previous $k-1$ steps.
+Step $k$.
+$p = [T_0, \ldots, T_l]$ where $l$ is the total number of trees that have been added to $p$ in the previous $k-1$ steps.
 Therefore the internal nodes in $T_l$ with ranks $1, \ldots, k-1$ induce clusters $C_1, \ldots, C_{k-1}$, respectively.
@@
 All leaves labelled with dots represent taxa in $\{1, \ldots, n\}$, squares represent taxa in $\{n+1, \ldots, 2n\}$.
-In \autocite{Li1996-zw} it is proven that there is such a labelling that ensures that there is no shortest path in $\nni$ where the clusters (dots vs.
-squares) shared between $T$ and $R$ are preserved.
+In \autocite{Li1996-zw} it is proven that there is such a labelling that ensures that there is no shortest path in $\nni$ where the clusters (dots vs.\ squares) shared between $T$ and $R$ are preserved.
 Instead, the depicted path where $n$ cherries are built and then sorted before being resolved is a shortest path}
@@
 % \vspace{12pt}
-% \caption{Two paths computed by $\findpath$. At the top $\findpath(T,R)$, at the bottom $\findpath(R,T)$}
+% \caption{Two paths computed by $\findpath$.
+% At the top $\findpath(T,R)$, at the bottom $\findpath(R,T)$}
 % \label{fig:findpath_not_symmetric}

### 2019-11-25 rnni_convexity.tex
abstract pass

@@
 \begin{abstract}
-Within this paper we consider a graph of ranked trees, that are rooted binary phylogenetic trees where internal nodes are ordered.
-The ranking of these internal nodes can be seen as a discretisation of trees where internal nodes are assigned divergence times.
-For phylogenetic reconstruction such as Markov Chain Monte Carlo (MCMC) methods, it is important to understand the geometry of tree spaces.
-This motivates our work, which focusses on understanding the geometry of the space of ranked trees, the so-called ranked nearest neighbour interchange ($\rnni$) graph.
-Moreover, we are interested in understanding the complexity of computing distances in this graph.
-We will see that even though there are similarities between the $\rnni$ graph on ranked trees and the classical nearest neighbour interchange ($\nni$) graph on unranked phylogenetic trees, some significant differences between these tree spaces suggest that there is a difference in the complexity of computing distances.
+In this paper we study the graph of ranked phylogenetic trees where the adjacency relation is given by a local rearrangement of the tree structure.
+Our work is motivated by tree inference algorithms, such as maximum likelihood and Markov Chain Monte Carlo methods, where the geometry of the search space plays a central role for efficiency and practicality of the optimisation.
+We hence focus on understanding the geometry of the space (graph) of ranked trees, the so-called ranked nearest neighbour interchange ($\rnni$) graph.
+\todo{Geometry summary}
+Specifically, we...
+
+We also study the problem of computing distances between trees in this graph.
+Since the $\rnni$ graph is a generalisation of the classical nearest neighbour interchange ($\nni$) graph to ranked phylogenetic trees, we specifically focus on the properties that have already been established for $\nni$.
+Surprisingly, our results suggest that the complexity of computing distances in the two graphs is likely to be different.
+\todo{Algorithms summary}
+Specifically, we...
 \end{abstract}

### 2019-11-25 rnni_convexity.tex
pass to caterpillar sort + rm Lemma 1 proof details (some)

@@
 
+\sloppy
+
 \title[Ranked Nearest Neighbour Intarchange]{Geometry of Ranked Nearest Neighbour Interchange Space of Phylogenetic Trees}
@@
 We hence focus on understanding the geometry of the space (graph) of ranked trees, the so-called ranked nearest neighbour interchange ($\rnni$) graph.
-\todo{Geometry summary}
+%TODO
 Specifically, we...
@@
 Surprisingly, our results suggest that the complexity of computing distances in the two graphs is likely to be different.
-\todo{Algorithms summary}
+%TODO
 Specifically, we...
@@
 
-Now we are ready to define the tree space which is the subject of study in this paper, the $\rnni$ graph.
+Now we are ready to introduce the tree space which is the subject of study in this paper, the $\rnni$ graph.
 
@@
 
+\todo{Paper structure}
+
 \section{Algorithms}
@@
 
-We start this section with a discussion of the algorithm proposed in \autocite{Gavryushkin2018-ol} to compute the exact $\rnni$ distance between trees by pre-computing the $\rnni$ graph.
-Then we introduce two algorithms, $\findpath$ (Section~\ref{section:alg_findpath}) and $\csort$ (Section~\ref{section:alg_csort}), for computing paths between two given trees in $\rnni$.
-These algorithms will become an important ingredient to establish several geometric properties of the $\rnni$ graph in Section~\ref{section:geometry}.
-Finally we will present the algorithm $\mdtree$, which produces a tree that is far away from the given input tree (see Section~\ref{section:alg_mdtree} for details).
-While this algorithm is interesting on its own (e.g.\ for simulation studies), we will again apply the algorithm to study the geometry of the $\rnni$ space.
+In this section we introduce three algorithms, $\findpath$ (Section~\ref{section:alg_findpath}), $\csort$ (Section~\ref{section:alg_csort}), and $\mdtree$ (Section~\ref{section:alg_mdtree}).
+$\findpath$ computes a $\rnni$ path between trees.
+$\csort$ computes a shortest $\rnni$ path between caterpillar trees (also known as ladder trees, see later in this section for accurate definitions).
+$\mdtree$ maximises the dissimilarity to find a remote tree given an arbitrary $\rnni$ tree.
+While these algorithms are interesting on their own (e.g.\ in simulation studies), they will also form an important ingredient to study the geometry of the $\rnni$ space later in this paper.
 
-For comparing approximate distances computed by $\findpath$ (Section~\ref{section:alg_findpath}) with exact distances and for supporting Conjecture~\ref{conjecture:cluster_theorem}, we aim to compute exact distances in $\rnni$ for small trees.
-Therefore we will use the algorithm for computing the $\rnni$ graph that is given in \textcite[Section 3.3]{Gavryushkin2018-ol}.
-\todo{Do we make use of tanglegrams? - LC: No}
+The algorithm $\findpath$ can be used to approximate the $\rnni$ distance.
+We will study the accuracy of this approximation by computing the exact $\rnni$ distance for small trees.
+Specifically, we will use the algorithm for computing the complete $\rnni$ graph that is given in \autocite[their Section 3.3]{Gavryushkin2018-ol}.
 \todo{GitHub link?}
-Once the graph is computed one can use common algorithms for computing shortest paths~\autocite{Floyd1962-ew,Dijkstra1959-ph}.
-However, the large number of vertices in $\rnni$ is an obstacle when it comes to computing distances.
-Since there are $\frac{n!(n-1)!}{2^{n-1}}$ vertices in the $\rnni$ graph \autocite{Gavryushkin2018-ol}, we are just able to use the $\rnni$ graph on up to six taxa for our analyses.
-We will get into further details when we discuss particular problems in Section~\ref{section:alg_findpath} and~\ref{section:cluster_theorem}.
+After pre-computing the graph, we use the standard algorithms~\autocite{Floyd1962-ew,Dijkstra1959-ph} for computing shortest paths.
+However, the large number of vertices in $\rnni$ is an obstacle when it comes to computing distances for larger trees.
+Since there are $\frac{n!(n-1)!}{2^{n-1}}$ vertices in the $\rnni$ graph \autocite{Gavryushkin2018-ol}, we compute the $\rnni$ graph on up to six taxa for our analyses.
+This computational result will also be used later in the paper to support Conjecture~\ref{conjecture:cluster_theorem}, a modification of Conjecture~9 in \autocite{Gavryushkin2018-ol}, which we prove to be false in this paper.
 
@@
 
-Within this section we are presenting $\findpath$, an efficient algorithm for computing paths, and therefore approximating distances, between any given pair of trees in $\rnni$.
-As we will see in Section~\ref{section:geometry}, it is useful for gaining insights into geometric properties of the $\rnni$ graph.
+In this section we present $\findpath$, an efficient algorithm for computing paths, and therefore approximating distances, between trees in $\rnni$.
 
@@
 
-We are now ready to explain how $\findpath$ (Algorithm~\ref{alg:find_path}) computes a path between two input trees $T$ and $R$.
-Let $[C_1, \ldots, C_{n-2}]$ be the cluster representation of $R$ (the ``destination'' tree).
+Let $T$ and $R$ be arbitrary trees and $[C_1, \ldots, C_{n-2}]$ be the cluster representation of $R$ (the ``destination'' tree).
 Then $C_i$ is induced by the node of rank $i$ for $i = 1, \ldots, n-2$.
-$\findpath$ computes a path which initially consists of $T_0 = T$ only and is extended stepwise as described in the following.
+$\findpath$ computes a path by iteratively extending a sequence $p$ of trees starting from $T$.
+The algorithm terminates when $p$ is an $\rnni$ path from $T$ to $R$.
 
+\todo{Alex: go through these steps and proof of L1}
 Step $0$.
@@
 
-\todo{$p = p+T'$: Do we really want to be updating this in every while loop? - LC: Yes, I guess so}
 \begin{algorithm}[H]
@@
 
-The following lemma shows that $\findpath$ gives a path between the two input trees, indeed, by proving that there is always an $\rnni$ move possible as described in Line~\ref{alg:findpath:line:move_set_down} of Algorithm~\ref{alg:find_path}.
+The following lemma shows that $\findpath$ is correct, that is, the algorithm returns a path between the two input trees.
+To prove this we have to establish that there always exists a unique $\rnni$ move required in Line~\ref{alg:findpath:line:move_set_down} of the algorithm.
 
@@
 
-% It is important to notice that $C_k$ either is the disjoint union $C_k = C_i \dot\cup C_j$ for $i,j < k$ or $C_k = C_i \dot\cup \{x\}$ for a taxon $x \in X \setminus (C_1 \cup \ldots \cup C_{k-1})$.
-%
-% Let us at first assume that $C_k = C_i \dot\cup C_j$ for some $i,j < k$.
-% The tree $T$ looks like the tree depicted in the left of Figure~\ref{fig:mrca_move} where $T_w^1$ and $T_w^2$ are the two subtrees rooted in the children of $w$ and $T_v$ is the subtree whose root is child of $v$ and does not include $w$.
-% Since we know that all clusters $C_1,\ ldots, C_{k-1}$ are induced by some nodes in $T$, each of these clusters must be a subset of the taxon set of one of the three subtrees $T_w^1,T_w^2$, or $T_v$.
-% In particular, there can't be elements of the same cluster $C_m$ in more than one of these subtrees.
-% This and the fact that $v = \mrca_{T}(C_i \dot\cup C_j)$ yields that either $C_i$ or $C_j$ is subset of the taxon set of $T_v$ and the other cluster is subset of the taxon set of either $T_w^1$ or $T_w^2$.
-%
-% Without loss of generality we assume that $C_i$ is included in $T_v$.
-% Otherwise, change the indices $i$ and $j$.
-% Then $C_i$ is a subset of the taxon set of either $T_w^1$ or $T_w^2$.
-% Therefore, just one of the two $\nni$ moves on $(v,w)$, as depicted in Figure~\ref{fig:mrca_move}, moves the most recent common ancestor of $C_k = C_i \dot \cup C_j$ down.
-%
-% Now we consider the case that $C_k = C_i \dot\cup \{x\}$ for a taxon $x \in X \setminus (C_1 \cup \ldots \cup C_{k-1})$.
-% For the same reason as above, $C_i$ is in subset of the taxon set of one of the subtrees $T_w^1, T_w^2$ or $T_v$ of $\hat T$ as depicted in Figure~\ref{fig:mrca_move}.
-% Similarly $x$ is taxon of one of these three subtrees in a way that either $C_i$ or $x$ is included in $T_v$, because $v = \mrca_{T'}(C_i \dot \cup \{x\})$.
-% So there is just one of the two subtrees $T_w^1$ and $T_w^2$ that neither includes $C_j$ nor $x$.
-% As in the previous case, just one of the two $\nni$ moves on $(v,w)$ as illustrated in Figure~\ref{fig:mrca_move} decreases the rank of $\mrca_{T}(C_i \dot \cup \{x\})$.
-%
-% We can conclude that in any case there is exactly one $\rnni$ move possible on $T$ that decreases the rank of $\mrca_{T}(C_k)$ in Line~\ref{alg:findpath:line:move_set_down} of Algorithm~\ref{alg:find_path}.
-
 \begin{figure}[H]
@@
 \end{figure}
-
 \end{proof}
 
-Note that the uniqueness of the $\rnni$ move in Line~\ref{alg:findpath:line:move_set_down} of Algorithm~\ref{alg:find_path} implies that the algorithm is deterministic.
+Note that the uniqueness of the $\rnni$ move in Line~\ref{alg:findpath:line:move_set_down} of Algorithm~\ref{alg:find_path} also implies that the algorithm is deterministic.
 The running time of the algorithm is quadratic in the number of taxa.

### 2019-11-25 rnni_convexity.tex
last para of findpath

@@
 A natural question then is how accurate this approximation is.
-To answer this question for small trees, we have implemented the algorithm $\findpath$
+We have computationally shown that the algorithm $\findpath$ finds the correct distance for trees up to six taxa.
+Specifically, we have implemented $\findpath$
 \todo{LC: github - link}
 \autocite{}
-and compared the results with the true distance found using the algorithm from Section~\ref{section:algorithms}.
-By doing this we could show that $\findpath$ computes correct distances for all trees on up to six taxa.
+and compared the results with the true $\rnni$ distance as explained earlier in this section.
 

### 2019-11-26 rnni_convexity.tex
removed all the 'shortest caterpillar path' and similar terminology + todo to complete this fix

@@
 As a result caterpillar trees have only one \emph{cherry}, which is a pair of taxa that share their parent.
-For introducing this algorithm we need the following definitions.
-A path $p$ between two caterpillar trees $T$ and $R$ that only consists of caterpillar trees is called a \emph{caterpillar path}.
-If $p$ is the shortest path among all caterpillar paths between $T$ and $R$ we call it a \emph{shortest caterpillar path}, and denote the length of $p$ by $d_c(T,R)$, which we call the \emph{caterpillar distance}.
-
-The special topology of caterpillar trees gives a natural preorder of taxa that uniquely represents a caterpillar tree.
-That is, for a caterpillar tree $T = [\{x_1, x_2\}, \{x_1, x_2, x_3\}, \ldots, \{x_1, \ldots, x_n\}]$ the taxa are preordered according to the ranks of their parents such that $x_1 = x_2 \leq x_3 \leq \ldots \leq x_n$.
-From now on when we say that a taxon is below another one we refer to this preorder.
-Furthermore, we can use this preorder to create a list representation of caterpillar trees.
-For the tree $T$ above this list is $[x_1, x_2, \ldots, x_n]$.
-As we consider the taxon set $X = \{1, \ldots, n\}$, we assume that they are ordered in the list such that $x_1 < x_2$, according to their ordering as natural numbers.
-This ensures that the list representation of caterpillar trees is unique.
-
-%We are now going to explain what kind of $\rnni$ moves are possible between caterpillar trees and see how such a move changes the list representation of a tree.
-%At first it is important to realise that the only possible $\rnni$ moves between caterpillar trees are $\nni$ moves, since there is no pair of internal nodes with rank difference one that is not connected by an edge.
-%$\nni$ moves between caterpillar trees correspond to swaps of two neighboured elements in the list representation.
-%The only exception of this are $\nni$ moves on the lowest internal edge, that is the edge connecting the internal nodes of rank one and two.
-%There are two $\nni$ moves possible on this edge that result in caterpillar trees, while there is for every other edge only one $\nni$ move that results in a caterpillar tree.
-%Corresponding to the $\nni$ moves on the lowest internal edge are exchanges of either the first or the second element with the third element in the list representation.
+A path between two caterpillar trees that only consists of caterpillar trees is called a \emph{caterpillar path}.
+We can identify a caterpillar tree $T = [\{x_1, x_2\}, \{x_1, x_2, x_3\}, \ldots, \{x_1, \ldots, x_n\}]$ with the list of its taxa $[x_1, x_2, \ldots, x_n]$, assuming that $x_1 < x_2$ as natural numbers (recall that the set of taxa is $\{1, \ldots, n\}$).
 
@@
 Let $T$ and $R$ be two caterpillar trees.
-$T$ and $R$ have distance $d(T,R) = \frac{(n-1)(n-2)}{2}$ if, and only if, they have caterpillar distance $d_c(T,R) = \frac{(n-1)(n-2)}{2}$.
+Then $d(T,R) = \frac{(n-1)(n-2)}{2}$ if and only if $d_c(T,R) = \frac{(n-1)(n-2)}{2}$, where $d_c$ is the length of the shortest caterpillar path.
 \label{lemma:caterpillar_dist=diameter}
@@
 
-For proving the other direction of the statement, let now $T$ and $R$ have caterpillar distance $d_c(T,R) = \frac{(n-1)(n-2)}{2}$.
+For proving the other direction of the statement, assume that $d_c(T,R) = \frac{(n-1)(n-2)}{2}$.
 We prove $d(T,R) = \frac{(n-1)(n-2)}{2}$ by induction on the number of taxa $n$ of $T$ and $R$:
@@
 
-Induction step: $n \to n+1$\\
-Induction hypothesis: For caterpillar trees $T, R$ on $n$ taxa with caterpillar distance $d_c(T,R) = \frac{(n-1)(n-2)}{2}$ it is $d(T,R) = \frac{(n-1)(n-2)}{2}$.\\
+Induction step:\\
+Assume that $d_c(T,R) = \frac{(n-1)(n-2)}{2}$ implies $d(T,R) = \frac{(n-1)(n-2)}{2}$.\\
 Without loss of generality we can assume that $T$ is the caterpillar tree $[1,2,3,\ldots,n+1]$.
@@
 $\csort$ computes a path from $T{\big|}_n$ to $R{\big|}_n$ that contains the same exchanges of taxa as $p$, except for the $n-1$ ones on $p$ that involve taxon $n+1$.
-Hence, the caterpillar distance is $d_c(T{\big|}_n, R{\big|}_n) = \frac{n(n-1)}{2} - (n-1)$ = $\frac{(n-1)(n-2)}{2}$.
+Hence, $d_c(T{\big|}_n, R{\big|}_n) = \frac{n(n-1)}{2} - (n-1)$ = $\frac{(n-1)(n-2)}{2}$.
 So we can apply the induction hypothesis on $T{\big|}_n$ and $R{\big|}_n$ and know that $d(T{\big|}_n,R{\big|}_n) = \frac{(n-1)(n-2)}{2}$.
@@
 \begin{proof}
-We prove this Theorem by backwards induction on the caterpillar distance $l = d_c(T,R)$ between caterpillar trees $T$ and $R$.
-For this backwards induction the maximum caterpillar distance $\frac{(n-1)(n-2)}{2}$ (Lemma~\ref{lemma:caterpillar_dist=diameter}) is the base case.
+We prove this theorem by backwards induction on the length of a shortest caterpillar path $d_c(T,R) = l$ ($l$ is the induction parameter) between trees $T$ and $R$.
+It follows from the proof of Lemma~\ref{lemma:caterpillar_dist=diameter} that the maximum possible value for $l$ is $\frac{(n-1)(n-2)}{2}$, which forms the induction basis.
+\todo{Lena: Make sure we ALWAYS use this terminology}
+Lemma~\ref{lemma:caterpillar_dist=diameter} also implies in this case that $d(T, R) = d_c(T, R)$, hence a caterpillar path between $T$ and $R$ of length $d(T, R)$.
 
-Induction basis: $l = \frac{(n-1)(n-2)}{2}$\\
-With Lemma~\ref{lemma:caterpillar_dist=diameter} one can see that for each pair of caterpillar trees $T$, $R$ with caterpillar distance $d_c(T,R) = \frac{(n-1)(n-2)}{2}$, it is $d(T,R) = d_c(T,R)$.
-Thus there is a shortest path between $T$ and $R$ that is a caterpillar path.
-
-Induction step: $l+1 \to l$\\
-Induction hypothesis: For each pair of caterpillar trees with caterpillar distance $l+1$ there is a shortest path between these trees that consists of caterpillar trees only, so the distance between these trees is $l+1$.
-
-Let $T$ and $R$ be caterpillar trees with caterpillar distance $l$.
+Let $T$ and $R$ be caterpillar trees such that $d_c(T, R) = l$.
 We consider the shortest caterpillar path between these trees computed by $\csort$.
@@
 
-So we can apply the induction hypothesis on $\hat T$ and $R$ and see that $d(\hat T,R) = l+1$.
+The induction hypothesis for $\hat T$ and $R$ implies that $d(\hat T,R) = l+1$.
 If it now were $d(T,R) < l$, it would hold $d(\hat T,R) \leq d(T,R) + d(T,\hat T) < l + 1$ which contradicts the induction hypothesis.
-We can conclude $d(T,R) = l$, which completes the proof for the case that $a$ is not one of the first two elements of $L_{T'}$.
+We can conclude that $d(T,R) = l$, which completes the proof for the case that $a$ is not one of the first two elements of $L_{T'}$.
 
-Let us now consider the other case:
-of the two taxa that are compared but do not swap positions, one is part of the cherry current tree.
-Let $a$ and $b$ be the cherry taxa and $c$ their neighbour such that neither $a$ nor $b$ echanges with $c$.
+Let us now consider the other case: of the two taxa that are compared but do not swap positions, one is part of the cherry current tree.
+Let $a$ and $b$ be the cherry taxa and $c$ their neighbour such that neither $a$ nor $b$ exchanges with $c$.
 Then the list looks like this:
@@
 
-If there is no swap of either $a$ or $b$ with $c$, we can construct a tree $\hat T$ with caterpillar distance $d_c(\hat T, R) = l+1$ and $d(T,\hat T) = 1$ in a similar way as in the previous case.
+If there is no swap of either $a$ or $b$ with $c$, we can construct a tree $\hat T$ with $d_c(\hat T, R) = l+1$ and $d(T,\hat T) = 1$ in a similar way as in the previous case.
 In the list representing the start tree $T$, either $a$ or $b$ must be neighbour of $c$.

### 2019-11-26 rnni_convexity.tex
pass to MDTree

@@
 
+\todo{Lena: Think if we want to change the following two paragraphs to when we first pull $x_n$ up, then $x_{n-1}$, etc.}
 The algorithm $\csort$ (Algorithm~\ref{alg:csort}) is a natural modification of the classical \emph{Bubble Sort} algorithm \autocite{Knuth1997-pi}.
@@
 
-Because all moves between caterpillar trees are $\nni$ moves, all pairs of taxa that are not in the same order in $T$ and $R$ need to swap at some point on any path between these trees.
-And since only elements that are direct neighbours in a sequence are allowed to swap, with the only exception being the first three elements, the path computed by $\csort$ is a shortest caterpillar path indeed.
+The path $p$ returned by the algorithm $\csort$ is a shortest caterpillar path because every tree modification on $p$ reduces the number of inversions of taxa in $T$ and $R$.
+\todo{Lena: If you decide to change the algorithm, mention here that what we've implemented is different from what is described here.}
+Note that any algorithm with this property returns a shortest caterpillar path.
 
@@
 \STATE $L_{T'}:=$ list representation of $T'$, $L_R:=$ list representation of $R$
-% \STATE $L_{\hat T} :=$ list representation of $\hat T$, $L_R :=$ list representation of $R$
 \FOR {$i=1,\ldots,n-2$} \label{alg:csort:line:loop}
@@
 \STATE $j = 2$
-% \IF {$L_R^{-1}[L_{\hat T}[j]] > L_R^{-1}[L_{\hat T}[j+1]]$}
 \WHILE {order of $L_{T'}[j]$ and $L_{T'}[j+1]$ different in $R$}
@@
 The running time of $\csort$ is the same as for Bubble Sort, that is quadratic in $n$.
-Though simple, this algorithm will be crucial for some proofs in Section~\ref{section:caterpillar_convex}.
+
+It is not obvious that the path between $T$ and $R$ returned by $\csort$ has the least possible length among all $\rnni$ paths (not only caterpillar paths), that is, the length of $d(T, R)$.
+This fact will be established in Theorem~\ref{}.
 
@@
 $\mdtree$ constructs an output tree $R$ as follows:
-Initially, $R$ only consists of a cherry with taxa $l_1$ and $l_2$ such that the parent of these taxa has rank $n-1$.
+Initially, $R$ only consists two taxa $l_1$ and $l_2$, the most recent common ancestor of which will eventually be of rank $n - 1$.
 The remaining taxa are added to $R$ sequentially, according to their order in $L$, by creating a new internal node at one of the existing branches in $R$ such that this new node has rank one less than the previously lowest internal node in $R$.

### 2019-11-26 rnni_convexity.tex
update Fig2

@@
 \vspace{12pt}
-\caption{Two types of operation that define edges of the $\rnni$ graph.
-The rank move swaps the ranks of the two nodes in the tree on the left such that $\nni$ moves are possible on the edge between the two nodes marked by black dots.}
+\caption{Two types of operation that define edges of the $\rnni$ graph,
+The rank move swaps the rank of the two highlighted nodes and the $\nni$ moves are performed on the blue edges.}
 \label{fig:RNNI}

### 2019-11-26 rnni_convexity.tex
abstract + paper structure(introduction) + todos Alex (one as comment in abstract)

@@
 We hence focus on understanding the geometry of the space (graph) of ranked trees, the so-called ranked nearest neighbour interchange ($\rnni$) graph.
-%TODO
-Specifically, we...
+Specifically, we introduce some algorithms for computing paths in the $\rnni$ graph and use them to establish properties such as diameter and radius of this graph.
 
@@
 %TODO
-Specifically, we...
+% Alex: The proof for $\np$ hardness of $\nni$ uses trees that have two caterpillar trees as subtrees (See Fig5).
+% I'm not sure whether it's known that computing distances just between caterpillar trees is not polynomial in $\nni$.
+% Do I need to be more specific about this here?
+% And shall I use the term 'caterpillar trees' here?
+We prove that within a particular subset of trees distances can be computed in polynomial time in $\rnni$ while this is not known to be true in $\nni$.
 \end{abstract}
@@
 \begin{figure}[H]
+\todo{Alex: Is it OK to use 'shortest $\rnni$ paths' here?}
 \centering
@@
 
-\todo{Paper structure}
+The rest of this paper is structured as follows.
+Within the next section we introduce three algorithms for exploring the $\rnni$ graph.
+We use these algorithms in Section~\ref{section:geometry} for proving some geometrical properties of this graph, such as diameter and radius.
+Specifically, we will prove in Section~\ref{section:caterpillar_convex} that for a subset of $\rnni$ trees, distances can be computed in polynomial time.
+This already suggests that there is a difference in complexity of computing distances between the $\rnni$ graph and the classical $\nni$ graph.
+\todo{Alex: Is it OK to use 'shortest $\rnni$ paths' here?}
+We will discuss this further in Section~\ref{section:cluster_theorem} where we disprove a conjecture raised in \autocite{Gavryushkin2018-ol} and provide an alternative version, which makes a claim about the similarity of trees on shortest $\rnni$ paths.
 
@@
 It is not obvious that the path between $T$ and $R$ returned by $\csort$ has the least possible length among all $\rnni$ paths (not only caterpillar paths), that is, the length of $d(T, R)$.
-This fact will be established in Theorem~\ref{}.
+This fact will be established in Theorem~\ref{thm:caterpillar_convex}.
 

### 2019-11-28 rnni_convexity.tex
add github link to code

@@
 \begin{figure}[H]
-\todo{Alex: Is it OK to use 'shortest $\rnni$ paths' here?}
 \centering
@@
 Specifically, we will use the algorithm for computing the complete $\rnni$ graph that is given in \autocite[their Section 3.3]{Gavryushkin2018-ol}.
-\todo{GitHub link?}
 After pre-computing the graph, we use the standard algorithms~\autocite{Floyd1962-ew,Dijkstra1959-ph} for computing shortest paths.
@@
 This computational result will also be used later in the paper to support Conjecture~\ref{conjecture:cluster_theorem}, a modification of Conjecture~9 in \autocite{Gavryushkin2018-ol}, which we prove to be false in this paper.
+Our implementations can be found on GitHub in \autocite{Collienne2019}.
 
@@
 We have computationally shown that the algorithm $\findpath$ finds the correct distance for trees up to six taxa.
-Specifically, we have implemented $\findpath$
-\todo{LC: github - link}
-\autocite{}
-and compared the results with the true $\rnni$ distance as explained earlier in this section.
+Specifically, we have implemented $\findpath$ \autocite{Collienne2019} and compared the results with the true $\rnni$ distance as explained earlier in this section.
 
@@
 For providing further evidence that the Cluster Theorem holds in $\rnni$, we can show computationally that it holds for the $\rnni$ graph on small trees with up to six taxa.
-We did this by computing subgraph of $\rnni$ only containing trees that share a cluster and compare distances within this graph with distances in the whole $\rnni$ graph \autocite{}.
-\todo{LC: github link}
+We did this by computing subgraph of $\rnni$ only containing trees that share a cluster and compare distances within this graph with distances in the whole $\rnni$ graph \autocite{Collienne2019}.
 We computed these graphs according to the algorithm of \autocite{Gavryushkin2018-ol} as discussed in Section~\ref{section:algorithms}.

### 2019-11-28 rnni_convexity.tex
change csort algorithm

@@
 The algorithm $\csort$ (Algorithm~\ref{alg:csort}) is a natural modification of the classical \emph{Bubble Sort} algorithm \autocite{Knuth1997-pi}.
-A path $p$ between two given trees $T$ and $R$ is computed by sorting the list representations of $T$ and $R$ as it is done in Bubble Sort.
-The general idea is to update the list representing $T$ by considering it from beginning to end and swap two neighboured taxa if they appear in a different order in $R$ than in the current list.
-Note that each swap of taxa results in a new tree that is added to the path $p$.
-After passing the whole list the last element, which is the uppermost taxon, is at the same position as in $R$.
-By repeating this at most $n-1$ times all elements the resulting list represents $R$.
-
-We use this main principle of Bubble Sort for computing shortest caterpillar paths with $\csort$, but there are special consideration needed for the first three taxa.
-The reason for this is that both $\nni$ moves on the lowest internal edge of a caterpillar tree result in caterpillar trees.
-For computing a shortest caterpillar path the moves on the first three taxa $x_1, x_2, x_3$ need to be done as described in the following.
-If $x_1$ ($x_2$) is the uppermost taxon of those three in $R$, we update the current tree by swapping $x_1$ ($x_2$) and $x_3$.
-By doing this we make sure that there will be no point on $p$ where $x_1$ and $x_2$ swap positions, as this is an additional $\nni$ move that can be avoided.
+A path $p$ from $T$ to $R= [x_1, x_2, \ldots, x_n]$ is computed iteratively such that after $k$ steps the last $k$ taxa of $T$ and $R$ coincide.
+Specifically, in step $k$ ($k = 0, \ldots, n-3$) taxon $x_{n-k}$ is moved up to the position it has in $R$.
+Here we say that a taxon moves up when it swaps position with its neighboured taxon, which corresponds to an $\nni$ move.
+Notice that there might be more than just one such move per step, which means that $p$ can be extended by more than one tree in each step.
+After $n-2$ steps we reach $R$.
 
@@
 \todo{Lena: If you decide to change the algorithm, mention here that what we've implemented is different from what is described here.}
+\todo{Alex: I changed the algorithm.
+It shouldn't be too much work to change the implementation and the pseudo-code here.
+Do you want to keep the old version or shall I change everything to the new version?}
+\end{description}}
 Note that any algorithm with this property returns a shortest caterpillar path.

### 2019-11-28 rnni_convexity.tex
update subsection mdtree

@@
 
-\todo{Lena: Think if we want to change the following two paragraphs to when we first pull $x_n$ up, then $x_{n-1}$, etc.}
 The algorithm $\csort$ (Algorithm~\ref{alg:csort}) is a natural modification of the classical \emph{Bubble Sort} algorithm \autocite{Knuth1997-pi}.
@@
 Do you want to keep the old version or shall I change everything to the new version?}
-\end{description}}
 Note that any algorithm with this property returns a shortest caterpillar path.
@@
 
-The algorithm we are going to introduce here will be important for finding the radius of the $\rnni$ graph.
-It is called $\mdtree$, which stands for Maximum Distance Tree, as it computes a tree with maximum distance if the input tree is a caterpillar tree.
-This will be discussed in Section~\ref{section:diameter}.
+The algorithm $\mdtree$ (Maximum Distance Tree) will be an important tool for finding the radius of the $\rnni$ graph.
+It computes a tree with maximum distance if the input tree is a caterpillar tree, which we will prove in Section~\ref{section:diameter}.
 
@@
 At first, all taxa of the input tree $T$ are sorted in a list $L = [l_1, \ldots, l_n]$ such that the ranks of their parents are non-decreasing.
-Though this representation appears similar to the list representation of caterpillar trees, it is ambiguous since we do not specify how to order cherry taxa.
+This list does uniquely identify trees since we do not specify how to order taxa that share a parent.
 $\mdtree$ constructs an output tree $R$ as follows:
@@
 
-Both the list computed in Line~\ref{alg:mdtree_lineL} the and the attachment edge of the new taxon in the tree in Line~\ref{alg:mdtree_lineAddTaxon} are not uniquely determined within $\mdtree$.
-Therefore, this algorithm is non-deterministic which is an important observation needed in Section~\ref{section:diameter} where this algorithm is used for finding the radius of $\rnni$.
+Both the list computed in Line~\ref{alg:mdtree_lineL} and the attachment edge of the new taxon in the tree in Line~\ref{alg:mdtree_lineAddTaxon} are not uniquely determined within $\mdtree$.
+Therefore, this algorithm is non-deterministic, which is an important observation that we will use for finding the radius of $\rnni$ in Section~\ref{section:diameter}.
 

### 2019-11-29 rnni_convexity.tex
pass until caterpillar convexity

@@
 
-In this section we analyse shortest paths in $\rnni$.
-Therefore, we will use the algorithms introduced in the previous section.
+In this section we analyse shortest paths in $\rnni$ by using the algorithms introduced in the previous section.
 They enable us to prove that there is always a shortest path between two caterpillar trees that only consists of caterpillar trees (Theorem~\ref{thm:caterpillar_convex}) in Section~\ref{section:caterpillar_convex}.
@@
 Let us assume that $T$ and $R = [C_1, \ldots, C_{n-2}]$ are trees for which $\findpath$ computes a path of maximum length.
-That means that if $T'$ is the last tree of $p$ at the beginning of loop number $i$ in the execution of $\findpath(T,R)$, $\mrca_{T'}(C_i)$ is the root.
-Thus, there are $n-1-i$ $\rnni$ operations needed to move $C_i$ to its correct position.
+It follows that in each step of $\findpath$ the most recent common ancestor of the cluster $C_k$ considered in that step is the root.
+Thus, there are $n-1-k$ $\rnni$ operations needed to move $C_k$ to its correct position.
 Hence, the maximum length of a path computed by $\findpath$ is bounded by $\sum\limits_{i = 1}^{n-2} n-1-i = \frac{(n-2)(n-1)}{2}$.
@@
 \begin{proof}
-Let $T$ and $R$ be trees as described in the lemma.
 First observe that the rank of the internal node $\parent_T(n+1)$ can only be changed by performing a rank move that involves $\parent_T(n+1)$ or an $\nni$ move on an edge adjacent to $\parent_T(n+1)$.
@@
 Let $p$ be a shortest path from $T$ to $R$ and $p{\big|}_n$ the path resulting from deleting taxon $n+1$ from all trees on $p$.
-Then $p{\big|}_n$ is a path from $T{\big|}_n$ to $R{\big|}_n$.
-The reason for this is that the only moves that are deleted from $p$ to receive $p{\big|}_n$ are those involving the parent of taxon $n+1$.
-All other moves stay the same and can be performed equally on a path from $T{\big|}_n$ to $R{\big|}_n$ as on a path from $T$ to $R$, because the relations of all other internal nodes are the same between both pairs of trees.
+Then $p{\big|}_n$ is a path from $T{\big|}_n$ to $R{\big|}_n$ as the moves on $p$ that are left out to receive $p{\big|}_n$ are only ones involving the parent of taxon $n+1$, which do not affect the remaining $\rnni$ moves.
+
 Recall that $\delta$ is the difference in ranks between the parents of taxon $n+1$ in $T$ and $R$.
@@
 In this section we restrict our attention to the set of caterpillar trees.
-Throughout this section the simplified list representation for these trees as introduced in Section~\ref{section:alg_csort} will be used.
-This type of tree is of particular interest in $\rnni$ because shortest paths between such trees in $\rnni$ differ from those in classical $\nni$ space.
-Following this observation and the importance of caterpillar trees in the proof of $\np$-hardness of $\nni$ distances in \autocite{Dasgupta2000-xa}, understanding the geometry of paths between caterpillar trees seems essential for investigating the complexity of computing $\rnni$ distances.
-In the following we demonstrate why finding shortest paths between caterpillar trees builds the basis for understanding the difference between $\nni$ and $\rnni$.
+These trees are of particular interest because shortest paths between such trees in $\rnni$ differ from those in classical $\nni$ space, which we will show in the following.
+We will show later in this section that $\rnni$ distances between caterpillar trees can be computed in polynomial time, which underlines the importance of these trees once more and suggests that there is a difference in the complexity of computing distances in $\nni$ and $\rnni$.
+Throughout this section we represent these trees with lists as described in Section~\ref{section:alg_csort}.
 
 Let us compare shortest paths between caterpillar trees in $\nni$ with those in $\rnni$.
-It can happen in $\nni$ that building a cherry and moving it may give a path that is shorter than a path where the taxa move separately in as it is illustrated out in Figure~\ref{fig:NNI_vs_RNNI}.
-There is only one move needed to build and resolve a cherry, respectively, while the number of moves for moving the cherry is the same as for moving a single taxon in $\nni$.
+In $\nni$ building a cherry and moving it may result a path that is shorter than a path where taxa move separately as it is illustrated out in Figure~\ref{fig:NNI_vs_RNNI}.
+In this example only one move is needed to build and resolve a cherry, respectively, while the number of moves for moving the cherry is the same as for moving a single taxon in $\nni$.
 In $\rnni$ space however, there are additional rank moves needed to change the rank of the internal node of the cherry before an $\nni$ move can resolve it.
@@
 The upper path is a natural extension of the shortest $\nni$ path in $\rnni$.
-It is longer than a shortest $\rnni$ path as the one at the bottom which consists of caterpillar trees only.}
+It is longer than a shortest $\rnni$ path as the one at the bottom that consists of caterpillar trees only.}
 \label{fig:NNI_vs_RNNI}
@@
 Let $T$ and $R$ be two caterpillar trees.
-Then $d(T,R) = \frac{(n-1)(n-2)}{2}$ if and only if $d_c(T,R) = \frac{(n-1)(n-2)}{2}$, where $d_c$ is the length of the shortest caterpillar path.
+Then $d(T,R) = \frac{(n-1)(n-2)}{2}$ if and only if $d_c(T,R) = \frac{(n-1)(n-2)}{2}$, where $d_c$ is the length of a shortest caterpillar path.
 \label{lemma:caterpillar_dist=diameter}
@@
 \begin{proof}
-Let $T$ and $R$ be caterpillar trees.
-Let us first prove that if $d(T,R) = \frac{(n-1)(n-2)}{2}$, it follows $d_c(T,R) = \frac{(n-1)(n-2)}{2}$.
-This easily follows from algorithm $\csort$:
-In each loop $i=1, \ldots, n-2$ of $\csort$ (Algorithm~\ref{alg:csort}) there are at most $n-1-i$ pairs of taxa that swap positions.
-This is due to the fact that after loop number $i$ the last $i$ elements in the list representation of the current tree on $p$ are in the same place as in the destination tree.
-As a result there are no further exchanges of pairs of taxa within the last $i-1$ taxa of the tree in loop $i$, which means that at most $n-1-i$ pairs of taxa exchange then.
-It follows that the maximum length of a path computed by $\csort$ is $\sum\limits_{i=1}^{n-2} n-1-i = \frac{(n-1)(n-2)}{2}$.
+Let us explain that if $d(T,R) = \frac{(n-1)(n-2)}{2}$, we can follow $d_c(T,R) = \frac{(n-1)(n-2)}{2}$ by using $\csort$:
+In each step $i=0, \ldots, n-3$ of $\csort$ at most $n-2-i$ pairs of taxa swap positions.
+This is due to the fact that in worst case the taxon that is considered in step $k$ is moved from the cherry of the tree up to position $n-k$.
+It follows that the maximum length of a path computed by $\csort$ is $\sum\limits_{i=0}^{n-3} n-2-i = \frac{(n-1)(n-2)}{2}$.
 Since there cannot be a caterpillar path shorter than $d(T,R) = \frac{(n-1)(n-2)}{2}$, it follows $d_c(T,R) = \frac{(n-1)(n-2)}{2}$.
@@
 
-Induction basis: $n=3$\\
-There are three trees on three taxa.
-They all have distance one from another and are caterpillar trees.
-So it is $d(T,R) = d_c(T,R) = 1$ for all pairs of trees $T,R$ on $n=3$ taxa.
+For the induction basis we consider the $\rnni$ graph $n=3$ taxa which consists of three trees only.
+Since every of these trees is a caterpillar tree that is connected with both other trees by an edge, it is $d(T,R) = d_c(T,R) = 1$ for all pairs of trees $T,R$ on $n=3$ taxa.
 
-Induction step:\\
-Assume that $d_c(T,R) = \frac{(n-1)(n-2)}{2}$ implies $d(T,R) = \frac{(n-1)(n-2)}{2}$.\\
-Without loss of generality we can assume that $T$ is the caterpillar tree $[1,2,3,\ldots,n+1]$.
-Let $R$ be a caterpillar tree with distance $\frac{n(n-1)}{2}$ to $T$.
-It follows that in each loop of $\csort$ the maximum number of taxon swaps is reached which implies that every taxon will swap with taxon $n+1$ at some point.
-We can conclude that this taxon $n+1$ is part of the cherry of $R$.
+For the induction step we assume without loss of generality that $T$ is the caterpillar tree $[1,2,3,\ldots,n+1]$ and $R$ is a caterpillar tree with distance $\frac{n(n-1)}{2}$ to $T$.
+It follows that when applying $\csort$ onto $T,R$ the number of taxon swaps is maximum in each step, which implies that taxon $n+1$ is part of the cherry of $R$.
 Let us now consider the trees $T{\big|}_n$ to $R{\big|}_n$ that result from $T$ and $R$ by deleting taxon $n+1$.
-$\csort$ computes a path from $T{\big|}_n$ to $R{\big|}_n$ that contains the same exchanges of taxa as $p$, except for the $n-1$ ones on $p$ that involve taxon $n+1$.
+$\csort$ computes a path from $T{\big|}_n$ to $R{\big|}_n$ that contains the same $\nni$ moves as $p$, except for the $n-1$ ones on $p$ that involve taxon $n+1$.
 Hence, $d_c(T{\big|}_n, R{\big|}_n) = \frac{n(n-1)}{2} - (n-1)$ = $\frac{(n-1)(n-2)}{2}$.
 So we can apply the induction hypothesis on $T{\big|}_n$ and $R{\big|}_n$ and know that $d(T{\big|}_n,R{\big|}_n) = \frac{(n-1)(n-2)}{2}$.
+With Lemma~\ref{lemma:distance_delete_taxon} we also know that it is $d(T{\big|}_n, R{\big|}_n) \leq d(T,R) - |\rank_T(\parent_T(n+1)) - \rank_R(\parent_R(n+1))| = d(T,R) - (n-1)$, where the last equality follows from our observation above.
 
-We are now using this observation to prove that the distance between $T$ and $R$ equals $\frac{n(n-1)}{2}$.
-This obviously gives an upper bound for the distance as it is $\Delta(\rnni_{n+1}) \leq \frac{n(n-1)}{2}$ for the $\rnni$ graph on $n+1$ taxa by Lemma~\ref{lemma:diameter_bound}.
-It is left to prove that $\frac{n(n-1)}{2}$ is a lower bound for the distance between $T$ and $R$ as well.
-According to Lemma~\ref{lemma:distance_delete_taxon}, the distance between $T{\big|}_n$ and $R{\big|}_n$ is at least $d(T{\big|}_n, R{\big|}_n) \leq d(T,R) - |\rank_T(\parent_T(n+1)) - \rank_R(\parent_R(n+1))|$.
-We also know $|\rank_T(\parent_T(n+1)) - \rank_R(\parent_R(n+1))| = n-1$.
-Suppose, contrary to our claim, that there is a path from $T$ to $R$ of length less than $\frac{n(n-1)}{2}$.
-By the above, it follows that $d(T{\big|}_n, R{\big|}_n) \leq d(T,R) - |\rank_T(\parent_T(n+1)) - \rank_R(\parent_R(n+1))| < \frac{n(n-1)}{2} - (n-1) < \frac{(n-1)(n-2)}{2}$.
+For proving $d(T,R) = \frac{n(n-1)}{2}$ we suppose contrary to our claim that there is a path from $T$ to $R$ of length less than $\frac{n(n-1)}{2}$.
+By the above, it follows that $d(T{\big|}_n, R{\big|}_n) \leq d(T,R) - (n-1) < \frac{n(n-1)}{2} - (n-1) < \frac{(n-1)(n-2)}{2}$.
 Since this is a contradiction to $d(T{\big|}_n, R{\big|}_n) = \frac{(n-1)(n-2)}{2}$, it is proven that $d(T,R) = \frac{n(n-1)}{2} $.

### 2019-11-29 rnni_convexity.tex
update proof caterpillar convex

@@
 
-The above observation suggests that there is a caterpillar path that is a shortest path between any two caterpillar trees in $\rnni$.
+The above observation suggests that there is a caterpillar path of length $d(T,R)$ between any two caterpillar trees $T,R$ in $\rnni$.
 Indeed, we are going to prove within this section that the set of caterpillar trees is convex in $\rnni$ (Theorem~\ref{thm:caterpillar_convex}).
@@
 We prove this theorem by backwards induction on the length of a shortest caterpillar path $d_c(T,R) = l$ ($l$ is the induction parameter) between trees $T$ and $R$.
-It follows from the proof of Lemma~\ref{lemma:caterpillar_dist=diameter} that the maximum possible value for $l$ is $\frac{(n-1)(n-2)}{2}$, which forms the induction basis.
+We saw in the proof of Lemma~\ref{lemma:caterpillar_dist=diameter} that the maximum possible value for $l$ is $\frac{(n-1)(n-2)}{2}$, which forms the induction basis.
 \todo{Lena: Make sure we ALWAYS use this terminology}
-Lemma~\ref{lemma:caterpillar_dist=diameter} also implies in this case that $d(T, R) = d_c(T, R)$, hence a caterpillar path between $T$ and $R$ of length $d(T, R)$.
+Lemma~\ref{lemma:caterpillar_dist=diameter} also implies in this case that $d(T, R) = d_c(T, R)$, hence there is a caterpillar path between $T$ and $R$ of length $d(T, R)$.
 
-Let $T$ and $R$ be caterpillar trees such that $d_c(T, R) = l$.
-We consider the shortest caterpillar path between these trees computed by $\csort$.
-From now on we will consider the trees $T$ and $R$, as well as all other caterpillar trees within this proof, in their list representation $L_T$ and $L_R$ as introduced in Section~\ref{section:alg_csort}.
+Let $T$ and $R$ now be caterpillar trees such that $d_c(T, R) = l < \frac{(n-1)(n-2)}{2}$.
+For proving $d(T,R) = l$ we consider the shortest caterpillar path between these trees computed by $\csort$.
 
-Since $l < \frac{(n-1)(n-2)}{2}$, there must be a loop in $\csort$ (Line~\ref{alg:csort:line:loop} of Algorithm~\ref{alg:csort}) where not every pair of taxa that is compared exchanges.
-Let us consider the first pair of taxa $a,b$ that is compared but does not swap positions, where $a$ is below $b$.
-Since the first time that this happens, $a$ is either one of the first two elements in the sequence, or it previously swapped with an element $c$ that has been between $a$ and $b$ before.
+Since $l < \frac{(n-1)(n-2)}{2}$, there must be a step in $\csort$ where the taxon that is moved does not require the maximum number of moves, meaning that it is not part of the cherry at the beginning of this step.
+Let $x$ be the first taxon that has this property and is considered in step number $k$.
+Then there must be a taxon $y$ that is lower neighbour of $x$ in the tree at the beginning of step $k$.
+Since we assume that step $k$ is the first one that does not require the maximum number of $\nni$ moves, $y$ is lower neighbour of $x$ in $T$ as well.
 
-We start with considering the latter case.
-The list representation of the tree at the point where $a$ and $b$ are compared but not swapped, looks as follows:
-\[[\ldots, c, a, b, \ldots, x_k, \ldots, x_1]\]
+Let $\hat T$ be a tree that equals $T$ but with $x$ and $y$ being exchanged.
+Applying $\csort$ on $\hat T$ and $R$ results in almost the same path as the one computed by $\csort$ from $T$ to $R$, there is just one additional move in the beginning of step $k$ for swapping $x$ and $y$.
+It follows directly that $d(T, \hat T) = 1$ and $d_c(\hat T,R) = d_c(T,R) + 1 = l + 1$.
 
-Note that here we assume that the elements $x_1, \ldots, x_k$ are in the same place as they are in the destination tree $R$.
-It could be that $k=1$, which means that there are no such elements.
-
-Since $a$ and $c$ swap, but $a$ and $b$ do not, we can follow that $b$ and $c$ appear in $R$ in the same order as they do in the current tree.
-Furthermore, $c$ and $b$ must have been neighbours in $T$, because we assume that the comparison of $a$ and $b$ is the first one where the compared elements are not swapped in the sequence.
-Let now $\hat T$ be a tree that equals $T$ with the only exception that $b$ and $c$ are exchanged.
-We can follow from our observations above that the distance between $T$ and $\hat T$ is $d(T, \hat T) = 1$ and that $\csort$ needs one move more for a path from $\hat T$ to $R$ than from $T$ to $R$.
-This results in $d_c(\hat T, R) = d(T,R) + 1$.
-
-The induction hypothesis for $\hat T$ and $R$ implies that $d(\hat T,R) = l+1$.
-If it now were $d(T,R) < l$, it would hold $d(\hat T,R) \leq d(T,R) + d(T,\hat T) < l + 1$ which contradicts the induction hypothesis.
-We can conclude that $d(T,R) = l$, which completes the proof for the case that $a$ is not one of the first two elements of $L_{T'}$.
-
-Let us now consider the other case: of the two taxa that are compared but do not swap positions, one is part of the cherry current tree.
-Let $a$ and $b$ be the cherry taxa and $c$ their neighbour such that neither $a$ nor $b$ exchanges with $c$.
-Then the list looks like this:
-\[[a, b, c, \ldots, x_k, \ldots, x_1]\]
-
-If there is no swap of either $a$ or $b$ with $c$, we can construct a tree $\hat T$ with $d_c(\hat T, R) = l+1$ and $d(T,\hat T) = 1$ in a similar way as in the previous case.
-In the list representing the start tree $T$, either $a$ or $b$ must be neighbour of $c$.
-This is due to the fact that we are currently considering the first comparison of elements where there is no swap.
-We can assume without loss of generality that $a$ is neighbour of $c$ in $T$.
-Then an exchange of $a$ and $c$ in $T$ yields a tree $\hat T$ with the desired properties.
-As in the previous case we can then apply the induction hypothesis on $\hat T$ and conclude that there is a shortest path between $T$ and $R$ that only consists of caterpillar trees.
+Therefore, we can apply the induction hypothesis for $\hat T$ and $R$ and get $d(\hat T,R) = l+1$.
+If it now were $d(T,R) < l$, it would be $d(\hat T,R) \leq d(T,R) + d(T,\hat T) < l + 1$ which contradicts the induction hypothesis.
+We can conclude that $d(T,R) = l$, which completes the proof.
 \end{proof}

### 2019-12-01 rnni_convexity.tex
pass to cluster thm

@@
 
-With the results of Section~\ref{section:caterpillar_convex} we can easily get the exact diameter of the $\rnni$ graph and hence improve the bounds for the diameter given in Theorem 7 of \autocite{Gavryushkin2018-ol}.
-In the following we show that the upper bound of the diameter provided by Lemma~\ref{lemma:diameter_bound} actually is the exact diameter of $\rnni$.
+In this section we establish diameter and radius of the $\rnni$ graph.
+In Theorem 7 of \autocite{Gavryushkin2018-ol} an upper bound for the diameter is given, but we are able to give the exact diameter using the results of Section~\ref{section:caterpillar_convex}.
+In the following we prove the exact diameter of $\rnni$ before we investigate the radius of $\rnni$.
 
@@
 We can compute a shortest caterpillar path by using $\csort$ as explained in Section~\ref{section:caterpillar_convex}.
-When running this algorithm to compute a path from $T$ to $R$, each taxon $i$ ($i = 1, \ldots, n-1$) swaps $n-1-i$ times with its upper neighbour.
-Therefore, the length of a shortest caterpillar path is $\sum\limits_{i=1}^{n-1}(n-1-i)  = \frac{(n-1)(n-2)}{2}$.
+When running this algorithm to compute a path from $T$ to $R$, each taxon $i$ ($i = n, \ldots, 3$) swaps $i-2$ times with its upper neighbour.
+Therefore, the length of a shortest caterpillar path is $\sum\limits_{i=3}^{n}(i-2)  = \frac{(n-1)(n-2)}{2}$.
 \end{proof}
@@
 
-Induction basis: $n = 3$\\
-$\mdtree(T)$ results either in $R =[1,3,2]$ or $R'= [2,3,1]$ for $T = [1,2,3]$.
-Since it is $d(T,R) = d(T,R') = 1 = \Delta(\rnni_3)$, the lemma is true for trees on $n=3$ taxa.
+The lemma is true for the induction basis on $n=3$ taxa since all three trees have distance one, which equals diameter and radius in this case.
 
-Induction step: $n \to n+1$\\
-Induction hypothesis: For a caterpillar tree $T$ on $n$ taxa $\mdtree$ computes a tree $R$ with distance $d(T,R) = \frac{(n-1)(n-2)}{2}$.\\
-Let us now assume that $T$ is a tree on $n+1$ taxa.
-In the list $L$ computed in $\mdtree$ (Line~\ref{alg:mdtree_lineL} of Algorithm~\ref{alg:max_dist_tree}) taxon $n+1 $ is the last one as its parent, the root, has the highest rank in $T$.
-This means that this taxon $n+1$ is the last one added to $R$ as in Line~\ref{alg:mdtree_lineAddTaxon} of this algorithm and therefore, it has parent with rank one in $R$.
-Because the parent of $n+1$ has rank $n$ in $T$ and rank $1$ in $R$, we can conclude from Lemma~\ref{lemma:distance_delete_taxon} that for $T$ and $R$ restricted to taxa $1,\ldots,n$ it holds: $d(T|_n,R|_n) \leq d(T,R) - (n-1)$.
+Let us now assume for the induction step that $T$ is a tree on $n+1$ taxa.
+Taxon $n+1$ is the last one in the list $L$ computed in $\mdtree$ as its parent in $T$ is the root.
+It follows that this taxon $n+1$ is the last one added to $R$ in the algorithm, so its parent has rank one in $R$.
+With Lemma~\ref{lemma:distance_delete_taxon} we can follow that for $T$ and $R$ restricted to taxa $\{1, \ldots, n\}$ it holds $d(T|_n,R|_n) \leq d(T,R) - (n-1)$.
 It is not hard to see that $R|_n$ can be received from $T|_n$ by applying $\mdtree$:
-$R|_n$ can be constructed from input tree $T|_n$ in the same way as $R$ can be constructed from $T$, since the list $L$ computed in Line~\ref{alg:mdtree_lineL} of Algorithm~\ref{alg:max_dist_tree} for $T|_n$ is the same as the one for $T$, but without the last element.
+$R|_n$ can be constructed from input tree $T|_n$ in the same way as $R$ can be constructed from $T$, since the list $L$ computed for $T|_n$ in $\mdtree$ is the same as the one for $T$, but without the last element.
 So we can follow from the induction hypothesis and Lemma~\ref{lemma:distance_delete_taxon} that it is $d(T,R) \geq d(T|_n,R|_n) + (n-1) = \frac{(n-1)(n-2)}{2} = \frac{n(n-1)}{2}$, which concludes the proof of the lemma.
@@
 In Section~\ref{alg:max_dist_tree} we already discussed that $\mdtree$ is a non-deterministic algorithm.
-In fact, the output tree of $\mdtree$ can have any tree topology, since there is no restriction on where to add the new taxon in the already existing tree (Line~\ref{alg:mdtree_lineAddTaxon} of Algorithm~\ref{alg:max_dist_tree}).
-This means that for each caterpillar input tree $\mdtree$ can compute a tree with any tree topology that has maximum distance from the input tree.
-And as permuting labels of trees does not change the distance between them, there is for every tree $R$ on $n$ taxa a caterpillar tree $T$ with distance $d(T,R) = \Delta(\rnni) = \frac{(n-1)(n-2)}{2}$.
+In fact, the output tree of $\mdtree$ can have any tree topology, since in each step of the algorithm the next taxon can be added on any edge incident to a leaf in the tree.
+Therefore, if the input tree is a caterpillar tree, $\mdtree$ can compute a tree with any tree topology that has maximum distance from the input tree.
+And as permuting labels of trees does not change the distance between them, there exists for every tree $R$ on $n$ taxa a caterpillar tree $T$ with distance $d(T,R) = \Delta(\rnni) = \frac{(n-1)(n-2)}{2}$.
 By this we obtain Corollary~\ref{corollary:radius}.

### 2019-12-01 rnni_convexity.tex
pass section cluster thm

@@
 
-Within this section we will make a further step towards understanding the complexity of computing shortest paths in $\rnni$.
+In this section we will make a further step towards understanding the complexity of computing shortest paths in $\rnni$.
 Since the problem of computing distances in the most common tree spaces as $\nni$ \autocite{Dasgupta2000-xa}, rooted $\spr$ \autocite{Bordewich2005-nx}, or $\tbr$ \autocite{Allen2001-ky} is known to be $\np$-hard, it is desirable to find a tree space that is less complex.
@@
 The natural approach to solve this question is to compare $\rnni$ and $\nni$ graph to see whether the proof of $\np$-hardness of $\nni$ applies to $\rnni$ as well.
-Within this section we will explain why this is not the case.
-Therefore we consider an example of shortest paths in $\nni$ that builds the key of the proof of its $\np$-hardness and explain why this example cannot be transferred to $\rnni$.
-Following this, we will introduce the so-called Split Theorem (Conjecture~\ref{conjecture:split_theorem}), which has been conjectured in \autocite{Gavryushkin2018-ol}.
+We will explain why this is not the case by considering an example of shortest paths in $\nni$ that builds the key of the proof of its $\np$-hardness.
+After explaining why this example cannot be transferred to $\rnni$ we will introduce the so-called Split Theorem (Conjecture~\ref{conjecture:split_theorem}), which has been conjectured in \autocite{Gavryushkin2018-ol}.
 However, we present a counterexample disproving the Split Theorem in its original version and claiming the Cluster Theorem (Conjecture~\ref{conjecture:cluster_theorem}) as alternative version.
@@
 \label{thm:split_nni}
-%The formulation of this lemma is a bit different in Li1996-zw.
-%However, the lemma as stated here follows directly from Li1996-zw.
 \end{theorem}
@@
 
-In the proof presented in \autocite{Li1996-zw} for Theorem~\ref{thm:split_nni} an example of two trees that share a cluster but have no shortest path that preserves this cluster is given.
-We describe this example in the following a simplified illustration of it can be found in Figure~\ref{fig:NNI_NP_proof}.
-In the example of \autocite{Li1996-zw} trees $T$ and $R$ on $2n$ taxa are considered where each of the trees consist of two caterpillar trees that share the root as parent.
-The clusters given by the two nodes adjacent to the root are $\{1, \ldots, n\}$ and $\{n+1, \ldots, 2n\}$ in both trees $T$ and $R$.
-\autocite{Li1996-zw} prove that on a shortest path between $T$ and $R$ in $\nni$ at first $T$ is transformed to a tree $T'$ with $n$ cherries, each of which contains one taxon of $\{1, \ldots, n\}$ and one taxon of $\{n+1, \ldots, 2n\}$.
-Afterwards, $T'$ is transformed to a tree $R'$ containing the same cherries as $T'$ but in a way that resolving all these cherries of $R'$ results in $R$.
-For proving that this path is shorter than any path preserving the clusters $\{1, \ldots, n\}$ and $\{n+1, \ldots, 2n\}$, \autocite{Li1996-zw} use the fact that $d_{\nni}(T',R') = d_{\nni}(T{\big|}_n, R{\big|}_n)$ where $T{\big|}_n$ and $R{\big|}_n$ are $T$ and $R$ restricted to taxa $\{1,\ldots,n\}$, respectively.
-For $\rnni$ there are lots of additional rank moves needed on a path that follows the principle of the shortest path in $\nni$ as described above.
-Similar to what we observed in the example depicted in Figure~\ref{fig:NNI_vs_RNNI}, this results in a path that is longer than a shortest $\rnni$ path.
-Therefore, a path where the two caterpillar subtrees of $T$ and $R$ are sorted separately is shorter in $\rnni$.
+In the proof of Theorem~\ref{thm:split_nni} presented in \autocite{Li1996-zw} an example of two trees $T$ and $R$ that share clusters $C$ and $X \setminus C$ but have no shortest path that preserves these cluster is given.
+A simplified illustration of this example is provided in Figure~\ref{fig:NNI_NP_proof}.
+\autocite{Li1996-zw} prove that on a shortest path between $T$ and $R$ in $\nni$ at first $T$ is transformed to a tree $T'$ with $\frac{n}{2}$ cherries, each of which contains one taxon of each of the clusters $C$ and $X \setminus C$.
+Then $T'$ is sorted to a tree $R'$ containing the same cherries as $T'$ such that that resolving all these cherries of $R'$ results in $R$.
+Specifically, \autocite{Li1996-zw} prove that there is no shortest path between $T$ and $R$ in $\nni$ where $C$ and $X \setminus C$ are present in all trees on that path.
+
+The reason that this example cannot be transferred to $\rnni$ is that there are lots of additional rank moves needed on a path between $T$ ans $R$ similar to the one described above.
+As in the example depicted in Figure~\ref{fig:NNI_vs_RNNI}, adding rank moves to the shortest $\nni$ path from $T$ to $R$ results in a path that is longer than a shortest $\rnni$ path.
+Sorting the two caterpillar subtrees of $T$ and $R$ separately results in a shorter path in $\rnni$.
 
@@
 \caption{Example of a shortest path between two trees $T$ and $R$ in $\nni$.
-All leaves labelled with dots represent taxa in $\{1, \ldots, n\}$, squares represent taxa in $\{n+1, \ldots, 2n\}$.
-In \autocite{Li1996-zw} it is proven that there is such a labelling that ensures that there is no shortest path in $\nni$ where the clusters (dots vs.\ squares) shared between $T$ and $R$ are preserved.
-Instead, the depicted path where $n$ cherries are built and then sorted before being resolved is a shortest path}
+In \autocite{Li1996-zw} it is proven that there is a labelling that ensures that there is no shortest path in $\nni$ where the clusters (dots vs.\ squares) shared between $T$ and $R$ are preserved.
+Instead, the depicted path where $\frac{n}{2}$ cherries are built and then sorted before being resolved is a shortest path}
 \label{fig:NNI_NP_proof}
@@
 
-Since the example given in \autocite{Li1996-zw}, which can be used to prove Theorem~\ref{thm:split_nni} in $\nni$, does not work in $\rnni$, a contrary statement was conjectured in \autocite{Gavryushkin2018-ol}.
+Since the example of a shortest $\nni$ described above, which was used to prove Theorem~\ref{thm:split_nni} in \autocite{Li1996-zw}, does not work in $\rnni$, a contrary statement was conjectured in \autocite{Gavryushkin2018-ol}.
 For understanding this conjecture it is important to know the connection between edges in a tree and so-called \emph{splits}, which are bipartitions of the taxon set of a tree:
@@
 Since the counterexample in Figure~\ref{fig:splitthm_counterexample} shows that the version of the Split Theorem stated in \autocite{Gavryushkin2018-ol} does not hold, we will now claim an alternative to this conjecture, the \emph{Cluster Theorem} (Conjecture~\ref{conjecture:cluster_theorem}).
-Within the Cluster Theorem we are considering clusters instead of splits.
+The Cluster Theorem concerns clusters instead of splits.
 This is motivated by the fact that rooted phylogenetic trees can uniquely be represented by sets of clusters \autocite{Steel2016-ye}, but not by sets of splits as they cannot define the position of the root.
@@
 
-For providing further evidence that the Cluster Theorem holds in $\rnni$, we can show computationally that it holds for the $\rnni$ graph on small trees with up to six taxa.
-We did this by computing subgraph of $\rnni$ only containing trees that share a cluster and compare distances within this graph with distances in the whole $\rnni$ graph \autocite{Collienne2019}.
+For providing further evidence that the Cluster Theorem holds in $\rnni$, we could show computationally that it holds for the $\rnni$ graph on small trees with up to six taxa.
+We did this by computing subgraph of $\rnni$ only containing trees that share a cluster and compare distances within this subgraph with distances in the whole $\rnni$ graph \autocite{Collienne2019}.
 We computed these graphs according to the algorithm of \autocite{Gavryushkin2018-ol} as discussed in Section~\ref{section:algorithms}.

### 2019-12-01 rnni_convexity.tex
change terminology shortest path that is caterpillar path

@@
 In this section we analyse shortest paths in $\rnni$ by using the algorithms introduced in the previous section.
-They enable us to prove that there is always a shortest path between two caterpillar trees that only consists of caterpillar trees (Theorem~\ref{thm:caterpillar_convex}) in Section~\ref{section:caterpillar_convex}.
+They enable us to prove that between caterpillar trees $T$ and $R$ there is always a caterpillar path of length $d(T,R)$ (Theorem~\ref{thm:caterpillar_convex}) in Section~\ref{section:caterpillar_convex}.
 In other words, the set of caterpillar trees is \emph{convex} in $\rnni$.
@@
 We saw in the proof of Lemma~\ref{lemma:caterpillar_dist=diameter} that the maximum possible value for $l$ is $\frac{(n-1)(n-2)}{2}$, which forms the induction basis.
-\todo{Lena: Make sure we ALWAYS use this terminology}
 Lemma~\ref{lemma:caterpillar_dist=diameter} also implies in this case that $d(T, R) = d_c(T, R)$, hence there is a caterpillar path between $T$ and $R$ of length $d(T, R)$.
@@
 Let $T = [1,2,3,\ldots,n]$ and $R = [n-1,n,n-2,n-3, \ldots, 1]$ be two caterpillar trees.
-According to Theorem~\ref{thm:caterpillar_convex}, there is a shortest path between $T$ and $R$ that is a caterpillar path.
+According to Theorem~\ref{thm:caterpillar_convex}, there is a caterpillar path of length $d(T,R)$ between $T$ and $R$.
 We can compute a shortest caterpillar path by using $\csort$ as explained in Section~\ref{section:caterpillar_convex}.
 When running this algorithm to compute a path from $T$ to $R$, each taxon $i$ ($i = n, \ldots, 3$) swaps $i-2$ times with its upper neighbour.
-Therefore, the length of a shortest caterpillar path is $\sum\limits_{i=3}^{n}(i-2)  = \frac{(n-1)(n-2)}{2}$.
+Therefore, the length of a shortest caterpillar path, and therefore the distance between $T$ and $R$, is $\sum\limits_{i=3}^{n}(i-2)  = \frac{(n-1)(n-2)}{2}$.
 \end{proof}

### 2019-12-01 rnni_convexity.tex
update description findpath

@@
 
-\todo{Alex: go through these steps and proof of L1}
-Step $0$.
-$p = [T_0]$.
-Consider the node in $T_0$ that induces the smallest cluster containing all elements of $C_1$.
-This node is denoted by $\mrca_{T_0}(C_1)$ (short for \emph{most recent common ancestor}).
-By performing an $\rnni$ move that decreases the rank of $\mrca_{T_0}(C_1)$, we receive a tree $T_1$ that is then appended to $p$.
-Now we repeatedly perform $\rnni$ moves that move the most recent common ancestor of $C_1$ down in order to extend $p$ to $p = [T_0, \ldots, T_l]$ for some $l$ such that $C_1$ is induced by the internal node of rank one in $T_l$.
+\todo{Alex: go through the following two paragraphs and proof of L1}
+In step $k$ of $\findpath$ we consider the cluster $C_k$ for $k = 1, \ldots, n-2$.
+$p$ is extended in step $k$ by considering the last tree $T'$ on $p$ and finding the node in $T'$ that induces the smallest cluster containing all elements of $C_k$.
+This node is denoted by $\mrca_{T'}(C_k)$ (short for \emph{most recent common ancestor}).
+By performing an $\rnni$ move that decreases the rank of $\mrca_{T'}(C_k)$, we receive a tree that is then appended to $p$.
+Now we extend $p$ by repeatedly performing $\rnni$ moves on the last tree of $p$ that move the most recent common ancestor of $C_k$ down.
+In the end of step $k$, $C_k$ is induced by the internal node of rank $k$ in the last tree of $p$.
 
-Step $k$.
-$p = [T_0, \ldots, T_l]$ where $l$ is the total number of trees that have been added to $p$ in the previous $k-1$ steps.
-Therefore the internal nodes in $T_l$ with ranks $1, \ldots, k-1$ induce clusters $C_1, \ldots, C_{k-1}$, respectively.
-Now the cluster of interest is $C_k$.
-As in the first step $p$ is extended by performing a series of $\rnni$ moves that decrease the rank of $\mrca_{T_l}(C_k)$ stepwise.
-Afterwards it is $p = [T_0, \ldots, T_{l'}]$ where the $k$ lowest internal nodes of $T_{l'}$ induce clusters $C_1, \ldots, C_{k}$.
-
-By following this procedure, the end tree of $p$ after step $n-2$ will be $R$, as it contains all clusters of $R$.
+By following this procedure, the last of $p$ after step $n-2$ will be $R$, as it contains all clusters of $R$.
 It is important but easy to see that the output path of $\findpath$ is unique for a given pair of trees, which means that it is a deterministic algorithm.

### 2019-12-01 rnni_convexity.tex
new pseudocode csort

@@
 
-Note that the uniqueness of the $\rnni$ move in Line~\ref{alg:findpath:line:move_set_down} of Algorithm~\ref{alg:find_path} also implies that the algorithm is deterministic.
+Note that the uniqueness of the $\rnni$ move in Line~\ref{alg:findpath:line:move_set_down} of the algorithm $\findpath$ also implies that the algorithm is deterministic.
 The running time of the algorithm is quadratic in the number of taxa.
@@
 \begin{algorithmic}[1]
-\STATE $T':= T$, $p := [T']$
-\STATE $L_{T'}:=$ list representation of $T'$, $L_R:=$ list representation of $R$
-\FOR {$i=1,\ldots,n-2$} \label{alg:csort:line:loop}
-\STATE $j = 0$ \COMMENT{Special case: cherry}
-\IF {$L_{T'}[0]$ is last element of $L_{T'}[0],L_{T'}[1],L_{T'}[2]$ in $L_R$}
-\STATE Update $T'$: swap $L_{T'}[0]$ and $L_{T'}[2]$
-\ELSIF {$L_{T'}[1]$ is last element of $L_{T'}[0],L_{T'}[1],L_{T'}[2]$ in $L_R$}
-\STATE Update $\hat T$: swap $L_{T'}[1]$ and $L_{T'}[2]$
+\STATE $[x_1, x_2, \ldots, x_n] := R$, $T' := T$, $p = [T']$
+\FOR {$i=n, \ldots, 3$}
+    \FOR {$j=1, \ldots, i$}
+        \IF {$T'[j] = x_i$}
+            \STATE Update $T'$: swap $T'[j]$ and $T'[j+1]$
+            \STATE $p = p + T'$
         \ENDIF
-\STATE $j = 2$
-\WHILE {order of $L_{T'}[j]$ and $L_{T'}[j+1]$ different in $R$}
-\STATE Update $T'$: Swap $L_{T'}[j]$ and $L_{T'}[j+1]$
-\STATE $p = p+T', j = j + 1$
-\ENDWHILE
+    \ENDFOR
 \ENDFOR
@@
 \end{algorithm}
+
+% % Old version of algorithm csort that is closer to bubble sort
+% \begin{algorithm}[H]
+% \caption{$\csort$($T,R$)}
+% \label{alg:csort}
+% \begin{algorithmic}[1]
+% \STATE $T':= T$, $p := [T']$
+% \STATE $L_{T'}:=$ list representation of $T'$, $L_R:=$ list representation of $R$
+% \FOR {$i=1,\ldots,n-2$} \label{alg:csort:line:loop}
+% \STATE $j = 0$ \COMMENT{Special case: cherry}
+% \IF {$L_{T'}[0]$ is last element of $L_{T'}[0],L_{T'}[1],L_{T'}[2]$ in $L_R$}
+% \STATE Update $T'$: swap $L_{T'}[0]$ and $L_{T'}[2]$
+% \ELSIF {$L_{T'}[1]$ is last element of $L_{T'}[0],L_{T'}[1],L_{T'}[2]$ in $L_R$}
+% \STATE Update $\hat T$: swap $L_{T'}[1]$ and $L_{T'}[2]$
+% \ENDIF
+% \STATE $j = 2$
+% \WHILE {order of $L_{T'}[j]$ and $L_{T'}[j+1]$ different in $R$}
+% \STATE Update $T'$: Swap $L_{T'}[j]$ and $L_{T'}[j+1]$
+% \STATE $p = p+T', j = j + 1$
+% \ENDWHILE
+% \ENDFOR
+% \RETURN $p$
+% \end{algorithmic}
+% \end{algorithm}
 

### 2019-12-01 rnni_convexity.tex
github link csort

@@
 The path $p$ returned by the algorithm $\csort$ is a shortest caterpillar path because every tree modification on $p$ reduces the number of inversions of taxa in $T$ and $R$.
-\todo{Lena: If you decide to change the algorithm, mention here that what we've implemented is different from what is described here.}
-\todo{Alex: I changed the algorithm.
-It shouldn't be too much work to change the implementation and the pseudo-code here.
-Do you want to keep the old version or shall I change everything to the new version?}
-Note that any algorithm with this property returns a shortest caterpillar path.
+Our implementation of this algorithm can be found on GitHub \autocite{Collienne2019}.
 

### 2019-12-03 rnni_convexity.tex
update discussion

@@
 \vspace{12pt}
-\caption{Two types of operation that define edges of the $\rnni$ graph,
+\caption{Two types of operation that define edges of the $\rnni$ graph.
 The rank move swaps the rank of the two highlighted nodes and the $\nni$ moves are performed on the blue edges.}
@@
 
-Within this paper we established some geometric properties of the $\rnni$ graph.
-We furthermore pointed out some differences between this graph on ranked trees and the well studied $\nni$ graph on unranked trees.
-This is of particular interest for analysing the complexity of computing distances in $\rnni$ as this problem is known to be $\np$-complete in $\nni$.
-Our finding that the set of caterpillar trees is convex in $\rnni$ (Theorem~\ref{thm:caterpillar_convex}) suggests that there is a difference in complexity between these two tree spaces.
-This leaves hope that computing distances in $\rnni$ is, contrary to this problem in $\nni$, not $\np$-complete.
-The question of how hard this problem actually is still remains open and will be subject of further research.
+The aim of this paper was to gain better understanding of the $\rnni$ space of ranked phylogenetic trees.
+Besides proving some geometrical properties of $\rnni$, we also introduce some algorithms in this paper.
+Our work highlights the importance of these algorithms as they are important for proving geometric properties such as diameter and radius of the $\rnni$ graph.
+The most important algorithm we introduced might be $\findpath$, which computes exact distances for all pairs of trees on up to six taxa.
+This algorithm might be a useful tool for future investigations of the $\rnni$ space.
 
-Our work in this paper highlights the importance of algorithms for computing paths in $\rnni$ in order to prove geometric properties such as diameter and radius.
-One of the algorithms that has been introduced, $\findpath$, computed exact distances for all pairs of trees on up to six taxa.
-For future investigations of the $\rnni$ space this algorithm might be a useful tool.
+We used the algorithms introduced in Section~\ref{section:algorithms} to prove some geometric properties of the $\rnni$ graph such as diameter and radius.
+Moreover, we discovered that the set of caterpillar trees is convex in $\rnni$ (Theorem~\ref{thm:caterpillar_convex}).
+With this result and the algorithm $\csort$ for computing shortest caterpillar paths (Section~\ref{section:alg_csort}) we can compute $\rnni$ distances between caterpillar trees in polynomial time.
+
+This fact that computing $\rnni$ distances between caterpillar trees only requires polynomial running time already suggests that there is a difference in complexity of computing distances in $\rnni$ and $\nni$.
+In Section~\ref{section:cluster_theorem} we supported this claim further by discussing the proof of $\np$-completeness of computing distances in $\nni$ and showing that this proof does not work for $\rnni$.
+Specifically, we claimed the Cluster Theorem (Conjetcure~\ref{conjecture:cluster_theorem}), which does not hold in $\nni$, but does hold in the $\rnni$ for trees on up to six taxa \autocite{Collienne2019}.
+Though the Cluster Theorem remains a conjecture, the results of this paper suggest that it is true.
+
+\subsection{Open Problems}
+
+In Section~\ref{section:cluster_theorem} we claim that the Cluster Theorem holds for $\rnni$.
+We could show that the previous version of it, the Split Theorem as it was conjectured in \autocite{Gavryushkin2018-ol}, is not true.
+If the Cluster Theorem actually is true still remains an open question.
+Proving it will be a big step towards finding the complexity of computing $\rnni$ distances.
+
+This question of whether computing $\rnni$ distances is $\np$-hard is another problem for future research.
+Though we did not aim to answer this question in this paper, we made some progress towards solving it by establishing some geometric properties of the $\rnni$ graph.
+Comparing these results with what is known for the plain $\nni$ graph makes us assume that the complexity of computing shortest path in $\rnni$ might differ from $\nni$.
+We also suspect that the fact that computing distances between caterpillar trees can be done in polynomial time might be the key for further investigation of the complexity of computing $\rnni$ distances.
 

### 2019-12-03 rnni_convexity.tex
minor on discussion

@@
 The aim of this paper was to gain better understanding of the $\rnni$ space of ranked phylogenetic trees.
-Besides proving some geometrical properties of $\rnni$, we also introduce some algorithms in this paper.
-Our work highlights the importance of these algorithms as they are important for proving geometric properties such as diameter and radius of the $\rnni$ graph.
-The most important algorithm we introduced might be $\findpath$, which computes exact distances for all pairs of trees on up to six taxa.
+Before investigating geometrical properties of $\rnni$ we introduced some algorithms in Section~\ref{section:algorithms}).
+One of the most important algorithms we introduced here is $\findpath$, which can be used for approximating $\rnni$ distances.
+We could in particular show that this algorithm computes exact distances for all pairs of trees on up to six taxa.
 This algorithm might be a useful tool for future investigations of the $\rnni$ space.
@@
 
-This fact that computing $\rnni$ distances between caterpillar trees only requires polynomial running time already suggests that there is a difference in complexity of computing distances in $\rnni$ and $\nni$.
-In Section~\ref{section:cluster_theorem} we supported this claim further by discussing the proof of $\np$-completeness of computing distances in $\nni$ and showing that this proof does not work for $\rnni$.
-Specifically, we claimed the Cluster Theorem (Conjetcure~\ref{conjecture:cluster_theorem}), which does not hold in $\nni$, but does hold in the $\rnni$ for trees on up to six taxa \autocite{Collienne2019}.
-Though the Cluster Theorem remains a conjecture, the results of this paper suggest that it is true.
+This already suggests that there is a difference in complexity of computing distances in $\rnni$ and $\nni$, since in $\nni$ there is no subspace known for which computing distances only takes polynomial time.
+In Section~\ref{section:cluster_theorem} we supported the claim that computing $\rnni$ distances is not as hard as computing $\nni$ distances further by discussing the proof of $\np$-completeness of computing distances in $\nni$ and showing that this proof does not work for $\rnni$.
+Specifically, we conjectured the Cluster Theorem (Conjetcure~\ref{conjecture:cluster_theorem}), which does not hold in $\nni$.
+However, we strongly suspect that it holds in $\rnni$ and could prove that it does hold for trees on up to six taxa \autocite{Collienne2019}.
 
@@
 
-In Section~\ref{section:cluster_theorem} we claim that the Cluster Theorem holds for $\rnni$.
+In Section~\ref{section:cluster_theorem} we claimed that the Cluster Theorem holds for $\rnni$.
 We could show that the previous version of it, the Split Theorem as it was conjectured in \autocite{Gavryushkin2018-ol}, is not true.

### 2019-12-06 rnni_convexity.tex
abstract + paper structure

@@
 We hence focus on understanding the geometry of the space (graph) of ranked trees, the so-called ranked nearest neighbour interchange ($\rnni$) graph.
-Specifically, we introduce some algorithms for computing paths in the $\rnni$ graph and use them to establish properties such as diameter and radius of this graph.
-
-We also study the problem of computing distances between trees in this graph.
-Since the $\rnni$ graph is a generalisation of the classical nearest neighbour interchange ($\nni$) graph to ranked phylogenetic trees, we specifically focus on the properties that have already been established for $\nni$.
-Surprisingly, our results suggest that the complexity of computing distances in the two graphs is likely to be different.
-%TODO
-% Alex: The proof for $\np$ hardness of $\nni$ uses trees that have two caterpillar trees as subtrees (See Fig5).
-% I'm not sure whether it's known that computing distances just between caterpillar trees is not polynomial in $\nni$.
-% Do I need to be more specific about this here?
-% And shall I use the term 'caterpillar trees' here?
-We prove that within a particular subset of trees distances can be computed in polynomial time in $\rnni$ while this is not known to be true in $\nni$.
+For example, we find exact estimates of the radius and diameter of the space.
+Since the $\rnni$ graph is a generalisation of the classical nearest neighbour interchange ($\nni$) graph to ranked phylogenetic trees, we compare geometric and algorithmic properties of the two graphs.
+Surprisingly, we discover that both geometric and algorithmic properties of $\rnni$ and $\nni$ are quite different.
+For example, we establish convexity of certain natural subspaces in $\rnni$, which is not convex is $\nni$.
+Furthermore, our results suggest that the complexity of computing distances in the two graphs is different.
 \end{abstract}
@@
 
-The rest of this paper is structured as follows.
-Within the next section we introduce three algorithms for exploring the $\rnni$ graph.
-We use these algorithms in Section~\ref{section:geometry} for proving some geometrical properties of this graph, such as diameter and radius.
-Specifically, we will prove in Section~\ref{section:caterpillar_convex} that for a subset of $\rnni$ trees, distances can be computed in polynomial time.
-This already suggests that there is a difference in complexity of computing distances between the $\rnni$ graph and the classical $\nni$ graph.
-\todo{Alex: Is it OK to use 'shortest $\rnni$ paths' here?}
-We will discuss this further in Section~\ref{section:cluster_theorem} where we disprove a conjecture raised in \autocite{Gavryushkin2018-ol} and provide an alternative version, which makes a claim about the similarity of trees on shortest $\rnni$ paths.
+The rest of this paper is organised as follows.
+In Section~\ref{section:algorithms} we introduce three algorithms for exploring the $\rnni$ graph.
+We use these algorithms in Section~\ref{section:geometry} to establish several geometrical properties of this graph, such as its diameter and radius.
+Furthermore, we will prove in Section~\ref{section:caterpillar_convex} that distances can be computed in polynomial time in a subspace of the $\rnni$ space.
+This result suggests that the complexity of computing distances in $\rnni$ is different from that in the classical $\nni$ space.
+We will discuss this further in Section~\ref{section:cluster_theorem}, where we disprove a conjecture of \textcite{Gavryushkin2018-ol} and suggest its weaker version a plausible alternative.
 

### 2019-12-06 rnni_convexity.tex
typo
paper structure revised

@@
 Surprisingly, we discover that both geometric and algorithmic properties of $\rnni$ and $\nni$ are quite different.
-For example, we establish convexity of certain natural subspaces in $\rnni$, which is not convex is $\nni$.
+For example, we establish convexity of certain natural subspaces in $\rnni$, which are not convex is $\nni$.
 Furthermore, our results suggest that the complexity of computing distances in the two graphs is different.
@@
 Its geometry has been extensively studied \autocite{Dasgupta2000-xa, Li1996-zw, Gordon2013-fw, De_Jong2016-al}.
-An important property of the $\nni$ graph is that computing distance is NP-hard \autocite{Dasgupta2000-xa}.
+An important property of the $\nni$ graph is that computing distances is NP-hard \autocite{Dasgupta2000-xa}.
 As a result no algorithm exists to compute the distance in practical time.
@@
 We use these algorithms in Section~\ref{section:geometry} to establish several geometrical properties of this graph, such as its diameter and radius.
-Furthermore, we will prove in Section~\ref{section:caterpillar_convex} that distances can be computed in polynomial time in a subspace of the $\rnni$ space.
+Furthermore, we will establish in Section~\ref{section:caterpillar_convex} that a certain subspace of $\rnni$ is convex and design a polynomial-time algorithm for computing $\rnni$-distances in that subspace.
 This result suggests that the complexity of computing distances in $\rnni$ is different from that in the classical $\nni$ space.
 We will discuss this further in Section~\ref{section:cluster_theorem}, where we disprove a conjecture of \textcite{Gavryushkin2018-ol} and suggest its weaker version a plausible alternative.
+We will finish the paper with a discussion and open problems.
 

### 2019-12-06 rnni_convexity.tex
link to implementations in Algorithms section
minor working changes

@@
 This computational result will also be used later in the paper to support Conjecture~\ref{conjecture:cluster_theorem}, a modification of Conjecture~9 in \autocite{Gavryushkin2018-ol}, which we prove to be false in this paper.
-Our implementations can be found on GitHub in \autocite{Collienne2019}.
+
+%TODO: Make a zenodo dio for the repository before final publication
+We have implemented all these algorithms in Python and the source code is available at \autocite{Collienne2019}.
 
@@
 The algorithm $\csort$ (Algorithm~\ref{alg:csort}) is a natural modification of the classical \emph{Bubble Sort} algorithm \autocite{Knuth1997-pi}.
-A path $p$ from $T$ to $R= [x_1, x_2, \ldots, x_n]$ is computed iteratively such that after $k$ steps the last $k$ taxa of $T$ and $R$ coincide.
-Specifically, in step $k$ ($k = 0, \ldots, n-3$) taxon $x_{n-k}$ is moved up to the position it has in $R$.
-Here we say that a taxon moves up when it swaps position with its neighboured taxon, which corresponds to an $\nni$ move.
+A path $p$ from $T$ to $R= [x_1, x_2, \ldots, x_n]$ is computed iteratively so that after $k$ steps the last $k$ taxa of $T$ and $R$ coincide.
+Specifically, in step $k$ ($k = 0, \ldots, n-3$) the parent of taxon $x_{n-k}$ is moved up by $\nni$ moves to the position it has in $R$.
 Notice that there might be more than just one such move per step, which means that $p$ can be extended by more than one tree in each step.
@@
 The path $p$ returned by the algorithm $\csort$ is a shortest caterpillar path because every tree modification on $p$ reduces the number of inversions of taxa in $T$ and $R$.
-Our implementation of this algorithm can be found on GitHub \autocite{Collienne2019}.
+Our Python implementation of this algorithm can be accessed at \autocite{Collienne2019}.
 

### 2019-12-06 rnni_convexity.tex
rm old version of
discussion of algorithms should be in the beginning

@@
 While these algorithms are interesting on their own (e.g.\ in simulation studies), they will also form an important ingredient to study the geometry of the $\rnni$ space later in this paper.
+For example, $\mdtree$ will be an important tool for finding the radius of the $\rnni$ graph, as the algorithms computes a tree with maximum distance if the input tree is a caterpillar tree.
+We will prove this fact in Section~\ref{section:diameter}.
 
@@
 
-% % Old version of algorithm csort that is closer to bubble sort
-% \begin{algorithm}[H]
-% \caption{$\csort$($T,R$)}
-% \label{alg:csort}
-% \begin{algorithmic}[1]
-% \STATE $T':= T$, $p := [T']$
-% \STATE $L_{T'}:=$ list representation of $T'$, $L_R:=$ list representation of $R$
-% \FOR {$i=1,\ldots,n-2$} \label{alg:csort:line:loop}
-% \STATE $j = 0$ \COMMENT{Special case: cherry}
-% \IF {$L_{T'}[0]$ is last element of $L_{T'}[0],L_{T'}[1],L_{T'}[2]$ in $L_R$}
-% \STATE Update $T'$: swap $L_{T'}[0]$ and $L_{T'}[2]$
-% \ELSIF {$L_{T'}[1]$ is last element of $L_{T'}[0],L_{T'}[1],L_{T'}[2]$ in $L_R$}
-% \STATE Update $\hat T$: swap $L_{T'}[1]$ and $L_{T'}[2]$
-% \ENDIF
-% \STATE $j = 2$
-% \WHILE {order of $L_{T'}[j]$ and $L_{T'}[j+1]$ different in $R$}
-% \STATE Update $T'$: Swap $L_{T'}[j]$ and $L_{T'}[j+1]$
-% \STATE $p = p+T', j = j + 1$
-% \ENDWHILE
-% \ENDFOR
-% \RETURN $p$
-% \end{algorithmic}
-% \end{algorithm}
-
 The running time of $\csort$ is the same as for Bubble Sort, that is quadratic in $n$.
@@
 \label{section:alg_mdtree}
-
-The algorithm $\mdtree$ (Maximum Distance Tree) will be an important tool for finding the radius of the $\rnni$ graph.
-It computes a tree with maximum distance if the input tree is a caterpillar tree, which we will prove in Section~\ref{section:diameter}.
 

### 2019-12-06 rnni_convexity.tex
missing discription of 'mdtree' -- @leacoll please check
new 'mdtree' -- @lenacoll please make sure I haven't screwed everything up!

@@
 The path $p$ returned by the algorithm $\csort$ is a shortest caterpillar path because every tree modification on $p$ reduces the number of inversions of taxa in $T$ and $R$.
+
 Our Python implementation of this algorithm can be accessed at \autocite{Collienne2019}.
@@
 
-$\mdtree$ (Algorithm~\ref{alg:max_dist_tree}) works as follows:
-At first, all taxa of the input tree $T$ are sorted in a list $L = [l_1, \ldots, l_n]$ such that the ranks of their parents are non-decreasing.
-This list does uniquely identify trees since we do not specify how to order taxa that share a parent.
+The idea behind this algorithm is to efficiently return a tree as far away from a given tree as possible.
+As we will see in Section~\ref{section:diameter}, $\mdtree$ achieves this goal exactly for caterpillar trees.
+
+$\mdtree$ (Algorithm~\ref{alg:max_dist_tree}) works as follows.
+Given an arbitrary tree $T$, order its taxa in a list $L = [l_1, \ldots, l_n]$ so that the ranks of their parents are non-decreasing with respect to this order.
+Note that this list is not uniquely determined by the tree.
 $\mdtree$ constructs an output tree $R$ as follows:
-Initially, $R$ only consists two taxa $l_1$ and $l_2$, the most recent common ancestor of which will eventually be of rank $n - 1$.
-The remaining taxa are added to $R$ sequentially, according to their order in $L$, by creating a new internal node at one of the existing branches in $R$ such that this new node has rank one less than the previously lowest internal node in $R$.
+Initially, $R$ only consists of two taxa $l_1$ and $l_2$, the most recent common ancestor of which will eventually be of rank $n - 1$.
+The remaining taxa are iteratively added to $R$ in the reversed order of $L$ starting from $l_n$.
+At every iteration, the taxon is attached to a new internal node created on one of the existing branches in $R$ so that this new attachment node has rank one.
+Note that $R$ is not uniquely determined due to the non-deterministic choice of the attachment edge.
 
@@
 \begin{algorithmic}[1]
-\STATE Construct list $L=[l_1, \ldots, l_n]$ of all taxa, ordered such that the ranks of their parents in $T$ do not decrease \label{alg:mdtree_lineL}
-\STATE Build tree $R$ with only two taxa $l_1, l_2$ that are children of the root, an internal node with rank $n-1$
-\FOR {$i = 1, \ldots, n$}
-\STATE Add an internal node as parent of $l_i$ on an edge in $R$ such that the new internal node has rank one less than the previously lowest internal node \label{alg:mdtree_lineAddTaxon}
+\STATE Construct list $L=[l_1, \ldots, l_n]$ of all taxa ordered with respect to the ranks of their parents in $T$
+\label{alg:mdtree_lineL}
+\STATE Build tree $R$ with two taxa $l_1, l_2$ that are children of the root
+\FOR {$i = n, \ldots, 3$}
+\STATE Extend $R$ by attaching $l_i$ to a newly created node of rank $1$ on an arbitrarily chosen edge
+\label{alg:mdtree_lineAddTaxon}
 \ENDFOR
@@
 
-Both the list computed in Line~\ref{alg:mdtree_lineL} and the attachment edge of the new taxon in the tree in Line~\ref{alg:mdtree_lineAddTaxon} are not uniquely determined within $\mdtree$.
+Note that the list computed in Line~\ref{alg:mdtree_lineL} of $\mdtree$, as well as the attachment edge of the new taxon in Line~\ref{alg:mdtree_lineAddTaxon} are not uniquely determined.
 Therefore, this algorithm is non-deterministic, which is an important observation that we will use for finding the radius of $\rnni$ in Section~\ref{section:diameter}.
@@
 
-Before we get into the details of these geometric properties of $\rnni$, we need to prove some statements that will be used a few times within this geometry section.
+Before we get into details of these geometric properties of $\rnni$, we need to prove the following auxiliary results.
 

### 2019-12-06 rnni_convexity.tex
intro for Geometry of RNNI section

@@
 \newcommand{\mdtree}{\textsc{MDTree}}
+\newcommand{\rad}{\textsc{R}}
 
@@
 
-In this section we analyse shortest paths in $\rnni$ by using the algorithms introduced in the previous section.
-They enable us to prove that between caterpillar trees $T$ and $R$ there is always a caterpillar path of length $d(T,R)$ (Theorem~\ref{thm:caterpillar_convex}) in Section~\ref{section:caterpillar_convex}.
+The study of shortest paths in the $\rnni$ graph in this section is motivated by our aim to understand the geometry of $\rnni$ as a metric space.
+We will employ the algorithms developed in the previous section to aid our analysis of shortest paths.
+Those algorithms will, for example, enable us to prove that every pair of caterpillar trees $T$ and $R$ is connected by a caterpillar path of length $d(T,R)$ (Theorem~\ref{thm:caterpillar_convex} in Section~\ref{section:caterpillar_convex}).
 In other words, the set of caterpillar trees is \emph{convex} in $\rnni$.
-In Section~\ref{section:diameter} we will investigate diameter and radius of the $\rnni$ graph.
-Interestingly, the \emph{diameter} of $\rnni$, that is the maximum distance $\Delta(\rnni) = \max \{d(T, R) \mid T, R \in \rnni\}$ between any two trees in the graph, equals its \emph{radius} $rad(\rnni) = \min\limits_T \max\limits_R d(T,R)$.
-After proving this, we are going to consider the so-called Split Theorem, which was raised in \autocite{Gavryushkin2018-ol}.
-We will give a counterexample to disprove their version of the Split Theorem and claim an alternative version, the Cluster Theorem (Conjecture~\ref{conjecture:cluster_theorem}).
+In Section~\ref{section:diameter} we will investigate the diameter and radius of the $\rnni$ space.
+Interestingly, the \emph{diameter} of $\rnni$, that is the maximum distance $\Delta(\rnni) = \max \{d(T, R) \mid T, R \in \rnni\}$ between any two trees in the graph, equals its \emph{radius} $\rad(\rnni) = \min\limits_T \max\limits_R d(T,R)$.
+Next, we will consider the so-called Split Theorem \autocite{Gavryushkin2018-ol}, which states that every shared split of taxa is maintained along every shortest path.
+We will give a counterexample to the Split Theorem in $\rnni$ and consider the Weak Split Theorem that states that every shared split of taxa if maintained along a shortest path (see Conjecture~\ref{conjecture:cluster_theorem}).
 
-Before we get into details of these geometric properties of $\rnni$, we need to prove the following auxiliary results.
+We start with the following auxiliary results.
 
 \begin{lemma}
-It is $\Delta(\rnni) \leq \frac{(n-1)(n-2)}{2}$ for $n \geq 3$.
+$\Delta(\rnni) \leq \frac{(n-1)(n-2)}{2}$ for $n \geq 3$.
 \label{lemma:diameter_bound}
@@
 \begin{corollary}
-The radius of the $\rnni$ graph is $rad(\rnni) = \Delta(\rnni) = \frac{(n-1)(n-2)}{2}$.
+The radius of the $\rnni$ graph is $\rad(\rnni) = \Delta(\rnni) = \frac{(n-1)(n-2)}{2}$.
 \label{corollary:radius}

### 2019-12-08 rnni_convexity.tex
to fig nni_vs_rnni

@@
 Next, we will consider the so-called Split Theorem \autocite{Gavryushkin2018-ol}, which states that every shared split of taxa is maintained along every shortest path.
-We will give a counterexample to the Split Theorem in $\rnni$ and consider the Weak Split Theorem that states that every shared split of taxa if maintained along a shortest path (see Conjecture~\ref{conjecture:cluster_theorem}).
+We will give a counterexample to the Split Theorem in $\rnni$ and consider a variant of the Weak Split Theorem that states that every shared split of taxa is maintained along a shortest path.
+We will call this statement the Clade Theorem, as it is slightly more natural to work with ranked trees in terms clades rather than splits (see Conjecture~\ref{conjecture:cluster_theorem}).
 
@@
 It follows that in each step of $\findpath$ the most recent common ancestor of the cluster $C_k$ considered in that step is the root.
-Thus, there are $n-1-k$ $\rnni$ operations needed to move $C_k$ to its correct position.
-Hence, the maximum length of a path computed by $\findpath$ is bounded by $\sum\limits_{i = 1}^{n-2} n-1-i = \frac{(n-2)(n-1)}{2}$.
+Thus, there are $n-1-k$ $\rnni$ operations necessary to move $C_k$ to its correct position.
+Hence, the maximum length of a path computed by $\findpath$ is bounded by $\sum\limits_{i = 1}^{n-2} (n-1-i) = \frac{(n-2)(n-1)}{2}$.
 \end{proof}
@@
 Second observe that any $\rnni$ move can change the rank of $\parent_T(n+1)$ by at most one.
-So we can follow that an $\rnni$ move on $T$ can decrease the rank of $\parent_T(n+1)$ by at most one.
+Hence an $\rnni$ move on $T$ can decrease the rank of $\parent_T(n+1)$ by at most one.
 
 Let $p$ be a shortest path from $T$ to $R$ and $p{\big|}_n$ the path resulting from deleting taxon $n+1$ from all trees on $p$.
-Then $p{\big|}_n$ is a path from $T{\big|}_n$ to $R{\big|}_n$ as the moves on $p$ that are left out to receive $p{\big|}_n$ are only ones involving the parent of taxon $n+1$, which do not affect the remaining $\rnni$ moves.
-
+Then $p{\big|}_n$ is a path from $T{\big|}_n$ to $R{\big|}_n$.
 Recall that $\delta$ is the difference in ranks between the parents of taxon $n+1$ in $T$ and $R$.
-With the knowledge that an $\rnni$ move on $T$ can change the rank of $\parent_T(n+1)$ by at most one, we can conclude that $|p{\big|}_n| \leq |p| - \delta$.
+Combining this with the observations above, we conclude that $|p{\big|}_n| \leq |p| - \delta$.
 Since $d(T{\big|}_n,R{\big|}_n) \leq |p{\big|}_n|$ and $|p| = d(T,R)$, the desired inequality follows.
@@
 In this section we restrict our attention to the set of caterpillar trees.
-These trees are of particular interest because shortest paths between such trees in $\rnni$ differ from those in classical $\nni$ space, which we will show in the following.
-We will show later in this section that $\rnni$ distances between caterpillar trees can be computed in polynomial time, which underlines the importance of these trees once more and suggests that there is a difference in the complexity of computing distances in $\nni$ and $\rnni$.
-Throughout this section we represent these trees with lists as described in Section~\ref{section:alg_csort}.
+These trees are of particular interest because they are used to prove the computing the $\nni$ distance is an NP-hard problem.
+As we will see below, shortest paths between caterpillar trees in $\rnni$ differ from those in the classical $\nni$ space.
+We will show later in this section that $\rnni$ distances between caterpillar trees can be computed in polynomial time.
+Throughout this section we use the list representation of trees described in Section~\ref{section:alg_csort}.
 
-Let us compare shortest paths between caterpillar trees in $\nni$ with those in $\rnni$.
-In $\nni$ building a cherry and moving it may result a path that is shorter than a path where taxa move separately as it is illustrated out in Figure~\ref{fig:NNI_vs_RNNI}.
-In this example only one move is needed to build and resolve a cherry, respectively, while the number of moves for moving the cherry is the same as for moving a single taxon in $\nni$.
-In $\rnni$ space however, there are additional rank moves needed to change the rank of the internal node of the cherry before an $\nni$ move can resolve it.
-This is due to the fact that $\nni$ moves are only allowed on edges of length one in $\rnni$.
-As a consequence a path where taxa move separately and no new cherry is built is shortest in $\rnni$.
+Recall \autocite{Gavryushkin2018-ol} that to compute a shortest path between two caterpillar trees in $\nni$ sometimes requires first building a clade and then move the clade around the tree (see Figure~\ref{fig:NNI_vs_RNNI}).
+In $\rnni$ however, the necessity of additional rank moves invalidates this strategy.
+This is due to the fact that $\nni$ moves $\rnni$ are only allowed on edges of length one.
 
@@
 \vspace{12pt}
-\caption{Paths between caterpillar trees $T$ and $R$: The solid paths are paths in $\rnni$, the dashed one is a path that is only possible in the plain $\nni$ graph.
-The upper path is a natural extension of the shortest $\nni$ path in $\rnni$.
-It is longer than a shortest $\rnni$ path as the one at the bottom that consists of caterpillar trees only.}
+\caption{Paths between caterpillar trees $T$ and $R$: Solid arrows indicate paths in $\rnni$, dashed arrow is a move only possible in $\nni$.
+Bottom path is a shortest $\rnni$ path.}
 \label{fig:NNI_vs_RNNI}

### 2019-12-08 rnni_convexity.tex
to induction step of lemma caterpillar_dist

@@
 
-The above observation suggests that there is a caterpillar path of length $d(T,R)$ between any two caterpillar trees $T,R$ in $\rnni$.
-Indeed, we are going to prove within this section that the set of caterpillar trees is convex in $\rnni$ (Theorem~\ref{thm:caterpillar_convex}).
-Note that this is not true for the $\nni$ space, which is clear from the example in Figure~\ref{fig:NNI_vs_RNNI}.
+These observations motivate the following general result.
+Every pair of trees $T, R$ in $\rnni$ are connected by a caterpillar path of length $d(T,R)$, that is, the set of caterpillar trees is convex in $\rnni$ (Theorem~\ref{thm:caterpillar_convex}).
+Note that this is not true in the $\nni$ space (Figure~\ref{fig:NNI_vs_RNNI}).
 
-The following lemma regards caterpillar trees with distance $\frac{(n-1)(n-2)}{2}$ and will be needed for proving the convexity of the set of caterpillar trees in $\rnni$.
-Notice that $\frac{(n-1)(n-2)}{2}$ is the upper bound for the diameter of $\rnni$ by Lemma~\ref{lemma:diameter_bound}.
+We need the following lemma.
+Recall that $\frac{(n-1)(n-2)}{2}$ is the upper bound of the diameter of $\rnni$ by Lemma~\ref{lemma:diameter_bound}.
 
@@
 \begin{proof}
-Let us explain that if $d(T,R) = \frac{(n-1)(n-2)}{2}$, we can follow $d_c(T,R) = \frac{(n-1)(n-2)}{2}$ by using $\csort$:
-In each step $i=0, \ldots, n-3$ of $\csort$ at most $n-2-i$ pairs of taxa swap positions.
-This is due to the fact that in worst case the taxon that is considered in step $k$ is moved from the cherry of the tree up to position $n-k$.
+First note that $d(T,R) = \frac{(n-1)(n-2)}{2}$ implies $d_c(T,R) = \frac{(n-1)(n-2)}{2}$.Indeed, in each step $i=0, \ldots, n-3$ of $\csort$ at most $n-2-i$ pairs of taxa swap positions, as in the worst case the taxon considered in step $k$ is moved from the cherry of the tree up to the position $n-k$.
 It follows that the maximum length of a path computed by $\csort$ is $\sum\limits_{i=0}^{n-3} n-2-i = \frac{(n-1)(n-2)}{2}$.
-Since there cannot be a caterpillar path shorter than $d(T,R) = \frac{(n-1)(n-2)}{2}$, it follows $d_c(T,R) = \frac{(n-1)(n-2)}{2}$.
 
 For proving the other direction of the statement, assume that $d_c(T,R) = \frac{(n-1)(n-2)}{2}$.
-We prove $d(T,R) = \frac{(n-1)(n-2)}{2}$ by induction on the number of taxa $n$ of $T$ and $R$:
+We prove that $d(T,R) = \frac{(n-1)(n-2)}{2}$ by induction on the number of taxa $n$. The induction basis for $n=3$ taxa is trivial as the $\rnni$ graph is a triangle on three caterpillar trees.
 
-For the induction basis we consider the $\rnni$ graph $n=3$ taxa which consists of three trees only.
-Since every of these trees is a caterpillar tree that is connected with both other trees by an edge, it is $d(T,R) = d_c(T,R) = 1$ for all pairs of trees $T,R$ on $n=3$ taxa.
-
-For the induction step we assume without loss of generality that $T$ is the caterpillar tree $[1,2,3,\ldots,n+1]$ and $R$ is a caterpillar tree with distance $\frac{n(n-1)}{2}$ to $T$.
+For the induction step we assume without loss of generality that $T$ is the caterpillar tree $[1,2,3,\ldots,n+1]$ and $R$ is a caterpillar tree at distance $\frac{n(n-1)}{2}$ to $T$.
 It follows that when applying $\csort$ onto $T,R$ the number of taxon swaps is maximum in each step, which implies that taxon $n+1$ is part of the cherry of $R$.

### 2019-12-09 rnni_convexity.tex
proof of lemma 4 is problematic

@@
 For proving the other direction of the statement, assume that $d_c(T,R) = \frac{(n-1)(n-2)}{2}$.
-We prove that $d(T,R) = \frac{(n-1)(n-2)}{2}$ by induction on the number of taxa $n$. The induction basis for $n=3$ taxa is trivial as the $\rnni$ graph is a triangle on three caterpillar trees.
+We prove that $d(T,R) = \frac{(n-1)(n-2)}{2}$ by induction on the number of taxa $n$.
+The induction basis for $n=3$ taxa is trivial as the $\rnni$ graph is a triangle on three caterpillar trees in that case.
 
 For the induction step we assume without loss of generality that $T$ is the caterpillar tree $[1,2,3,\ldots,n+1]$ and $R$ is a caterpillar tree at distance $\frac{n(n-1)}{2}$ to $T$.
+\todo{We cannot assume that and the rest of the proof doesn't go through.}
 It follows that when applying $\csort$ onto $T,R$ the number of taxon swaps is maximum in each step, which implies that taxon $n+1$ is part of the cherry of $R$.
@@
 \begin{theorem}
-The set of caterpillar trees is convex.
+The set of caterpillar trees is convex in $\rnni$.
 \label{thm:caterpillar_convex}

### 2019-12-09 rnni_convexity.tex
Cluster theorem is not so much related to complexity
trying to fix Lemma 4 -- WiP

@@
 Next, we will consider the so-called Split Theorem \autocite{Gavryushkin2018-ol}, which states that every shared split of taxa is maintained along every shortest path.
-We will give a counterexample to the Split Theorem in $\rnni$ and consider a variant of the Weak Split Theorem that states that every shared split of taxa is maintained along a shortest path.
+We will give a counterexample to the Split Theorem in $\rnni$ and consider a variant of the Weak Split Theorem, which states that every shared split of taxa is maintained along a shortest path.
 We will call this statement the Clade Theorem, as it is slightly more natural to work with ranked trees in terms clades rather than splits (see Conjecture~\ref{conjecture:cluster_theorem}).
@@
 \begin{proof}
-First note that $d(T,R) = \frac{(n-1)(n-2)}{2}$ implies $d_c(T,R) = \frac{(n-1)(n-2)}{2}$.Indeed, in each step $i=0, \ldots, n-3$ of $\csort$ at most $n-2-i$ pairs of taxa swap positions, as in the worst case the taxon considered in step $k$ is moved from the cherry of the tree up to the position $n-k$.
-It follows that the maximum length of a path computed by $\csort$ is $\sum\limits_{i=0}^{n-3} n-2-i = \frac{(n-1)(n-2)}{2}$.
+First note that $d(T,R) = \frac{(n-1)(n-2)}{2}$ implies $d_c(T,R) = \frac{(n-1)(n-2)}{2}$.
+Indeed, in each step $i=0, \ldots, n-3$ of $\csort$ at most $n-2-i$ pairs of taxa swap positions, as in the worst case the taxon considered in step $k$ is moved from the cherry of the tree up to the position $n-k$.
+It follows that the maximum length of a path computed by $\csort$ is
+\begin{equation}
+\sum\limits_{i=0}^{n-3} n-2-i = \frac{(n-1)(n-2)}{2}.
+\label{csort_max}
+\end{equation}
 
@@
 
-For the induction step we assume without loss of generality that $T$ is the caterpillar tree $[1,2,3,\ldots,n+1]$ and $R$ is a caterpillar tree at distance $\frac{n(n-1)}{2}$ to $T$.
-\todo{We cannot assume that and the rest of the proof doesn't go through.}
-It follows that when applying $\csort$ onto $T,R$ the number of taxon swaps is maximum in each step, which implies that taxon $n+1$ is part of the cherry of $R$.
+For the induction step we assume without loss of generality that $T$ is the caterpillar tree $[1,2,3,\ldots,n+1]$ and $R$ is a caterpillar tree
+%TODO: this is an incorrect assumption -- we need to prove this!
+at distance $\frac{n(n-1)}{2}$ from $T$.
+Consider the path $p = \csort(T, R)$ of length
+Note that in this case taxon $n+1$ has to be in the cherry of $R$.
+Otherwise, $\csort$ would find a path between $T$ and $R$ shorter than $\frac{n(n-1)}{2}$, which can be shown as in (\ref{csort_max}) above.
+
 Let us now consider the trees $T{\big|}_n$ to $R{\big|}_n$ that result from $T$ and $R$ by deleting taxon $n+1$.
@@
 
+\todo{Why are we in the Geometry section then?}
 In this section we will make a further step towards understanding the complexity of computing shortest paths in $\rnni$.

### 2019-12-09 rnni_convexity.tex
fixed proof of lemma 4

@@
 
-For the induction step we assume without loss of generality that $T$ is the caterpillar tree $[1,2,3,\ldots,n+1]$ and $R$ is a caterpillar tree
-%TODO: this is an incorrect assumption -- we need to prove this!
-at distance $\frac{n(n-1)}{2}$ from $T$.
-Consider the path $p = \csort(T, R)$ of length
-Note that in this case taxon $n+1$ has to be in the cherry of $R$.
-Otherwise, $\csort$ would find a path between $T$ and $R$ shorter than $\frac{n(n-1)}{2}$, which can be shown as in (\ref{csort_max}) above.
+For the induction step we assume without loss of generality that $T$ is the caterpillar tree $[1, \ldots, n+1]$ and $R$ is a caterpillar tree such that $d_c(T, R) = \frac{n(n-1)}{2}$.
+Consider the path $p = \csort(T, R)$ of this length.
+Note that in this case taxon $n+1$ has to be in the cherry of $R$, as otherwise $\csort$ would find a path between $T$ and $R$ of length strictly less than $\frac{n(n-1)}{2}$, which can be shown by counting the number of moves as in (\ref{csort_max}) above.
 
-Let us now consider the trees $T{\big|}_n$ to $R{\big|}_n$ that result from $T$ and $R$ by deleting taxon $n+1$.
-$\csort$ computes a path from $T{\big|}_n$ to $R{\big|}_n$ that contains the same $\nni$ moves as $p$, except for the $n-1$ ones on $p$ that involve taxon $n+1$.
-Hence, $d_c(T{\big|}_n, R{\big|}_n) = \frac{n(n-1)}{2} - (n-1)$ = $\frac{(n-1)(n-2)}{2}$.
-So we can apply the induction hypothesis on $T{\big|}_n$ and $R{\big|}_n$ and know that $d(T{\big|}_n,R{\big|}_n) = \frac{(n-1)(n-2)}{2}$.
-With Lemma~\ref{lemma:distance_delete_taxon} we also know that it is $d(T{\big|}_n, R{\big|}_n) \leq d(T,R) - |\rank_T(\parent_T(n+1)) - \rank_R(\parent_R(n+1))| = d(T,R) - (n-1)$, where the last equality follows from our observation above.
-
-For proving $d(T,R) = \frac{n(n-1)}{2}$ we suppose contrary to our claim that there is a path from $T$ to $R$ of length less than $\frac{n(n-1)}{2}$.
-By the above, it follows that $d(T{\big|}_n, R{\big|}_n) \leq d(T,R) - (n-1) < \frac{n(n-1)}{2} - (n-1) < \frac{(n-1)(n-2)}{2}$.
-Since this is a contradiction to $d(T{\big|}_n, R{\big|}_n) = \frac{(n-1)(n-2)}{2}$, it is proven that $d(T,R) = \frac{n(n-1)}{2} $.
+Let us now consider the path $p\big|_n$ from $T\big|_n$ to $R\big|_n$, which is obtained by restricting every tree on $p$ to taxa $1, \ldots, n$ and removing identical trees.
+Observe that this path is shorter than $p$ by the number of moves that involve taxon $n+1$.
+Since this taxon is adjacent to the root in $T$ and part of the cherry in $R$, $\csort$ requires $n-1$ such moves.
+Hence, $d_c(T{\big|}_n, R{\big|}_n) = \frac{n(n-1)}{2} - (n-1)$ = $\frac{(n-1)(n-2)}{2} = d(T{\big|}_n,R{\big|}_n)$, by induction.
+Lemma~\ref{lemma:distance_delete_taxon} implies that $d(T{\big|}_n, R{\big|}_n) \leq d(T,R) - |\rank_T(\parent_T(n+1)) - \rank_R(\parent_R(n+1))| = d(T,R) - (n-1)$.
+So $\frac{(n-1)(n-2)}{2} \leq d(T,R) - (n-1)$, which implies that $d(T, R) \geq \frac{n(n-1)}{2}$.
+Thus $d(T, R) = \frac{n(n-1)}{2}$.
 \end{proof}

### 2019-12-09 rnni_convexity.tex
Cluster Theorem shown for 7 taxa. Update Authors.

@@
 \email{\textsuperscript{\Letter}alex@biods.org}
+\author{Kieran Elmes\textsuperscript{1}}
+\email{kelmes@cs.otago.ac.nz}
 
@@
 Specifically, we will use the algorithm for computing the complete $\rnni$ graph that is given in \autocite[their Section 3.3]{Gavryushkin2018-ol}.
-After pre-computing the graph, we use the standard algorithms~\autocite{Floyd1962-ew,Dijkstra1959-ph} for computing shortest paths.
+After pre-computing the graph, we use Seidel's algorithm~\autocite{seidel_all-pairs-shortest-path_1995} for computing distances.
 However, the large number of vertices in $\rnni$ is an obstacle when it comes to computing distances for larger trees.
-Since there are $\frac{n!(n-1)!}{2^{n-1}}$ vertices in the $\rnni$ graph \autocite{Gavryushkin2018-ol}, we compute the $\rnni$ graph on up to six taxa for our analyses.
+Since there are $\frac{n!(n-1)!}{2^{n-1}}$ vertices in the $\rnni$ graph \autocite{Gavryushkin2018-ol}, we compute the $\rnni$ graph on up to seven taxa for our analyses.
 This computational result will also be used later in the paper to support Conjecture~\ref{conjecture:cluster_theorem}, a modification of Conjecture~9 in \autocite{Gavryushkin2018-ol}, which we prove to be false in this paper.
@@
 %TODO: Make a zenodo dio for the repository before final publication
-We have implemented all these algorithms in Python and the source code is available at \autocite{Collienne2019}.
+We have implemented these algorithms in a combination of Python and C, and the source code is available at \autocite{Collienne2019}.
 
@@
 
-For providing further evidence that the Cluster Theorem holds in $\rnni$, we could show computationally that it holds for the $\rnni$ graph on small trees with up to six taxa.
+For providing further evidence that the Cluster Theorem holds in $\rnni$, we could show computationally that it holds for the $\rnni$ graph on small trees with up to seven taxa.
 We did this by computing subgraph of $\rnni$ only containing trees that share a cluster and compare distances within this subgraph with distances in the whole $\rnni$ graph \autocite{Collienne2019}.
@@
 Specifically, we conjectured the Cluster Theorem (Conjetcure~\ref{conjecture:cluster_theorem}), which does not hold in $\nni$.
-However, we strongly suspect that it holds in $\rnni$ and could prove that it does hold for trees on up to six taxa \autocite{Collienne2019}.
+However, we strongly suspect that it holds in $\rnni$ and could prove that it does hold for trees on up to seven taxa \autocite{Collienne2019}.
 

### 2019-12-09 rnni_convexity.tex
Theorem caterpillar_convex induction basis

--- 2019-12-09 rnni_convexity.tex
@@
 \begin{proof}
-We prove this theorem by backwards induction on the length of a shortest caterpillar path $d_c(T,R) = l$ ($l$ is the induction parameter) between trees $T$ and $R$.
-We saw in the proof of Lemma~\ref{lemma:caterpillar_dist=diameter} that the maximum possible value for $l$ is $\frac{(n-1)(n-2)}{2}$, which forms the induction basis.
-Lemma~\ref{lemma:caterpillar_dist=diameter} also implies in this case that $d(T, R) = d_c(T, R)$, hence there is a caterpillar path between $T$ and $R$ of length $d(T, R)$.
+We prove this theorem by backward induction on the length of a shortest caterpillar path $d_c(T,R) = l$ ($l$ is the induction parameter) between trees $T$ and $R$.
+The backward induction basis follows by Lemma~\ref{lemma:caterpillar_dist=diameter}, as the maximum possible value for $l$ is $\frac{(n-1)(n-2)}{2}$.
 

### 2019-12-10 rnni_convexity.tex
author order
fixed indexing errors in csort description

@@
 \address{\textsuperscript{1}Department of Computer Science, University of Otago, New Zealand}
+\author{Kieran Elmes\textsuperscript{1}}
+\email{kelmes@cs.otago.ac.nz}
 \author{Mareike Fischer\textsuperscript{2}}
@@
 \email{\textsuperscript{\Letter}alex@biods.org}
-\author{Kieran Elmes\textsuperscript{1}}
-\email{kelmes@cs.otago.ac.nz}
 
@@
 
-The algorithm $\csort$ (Algorithm~\ref{alg:csort}) is a natural modification of the classical \emph{Bubble Sort} algorithm \autocite{Knuth1997-pi}.
+The algorithm $\csort$ (Algorithm~\ref{alg:csort}) is a natural modification of the classical Bubble Sort algorithm \autocite{Knuth1997-pi}.
 A path $p$ from $T$ to $R= [x_1, x_2, \ldots, x_n]$ is computed iteratively so that after $k$ steps the last $k$ taxa of $T$ and $R$ coincide.
-Specifically, in step $k$ ($k = 0, \ldots, n-3$) the parent of taxon $x_{n-k}$ is moved up by $\nni$ moves to the position it has in $R$.
+Specifically, in step $k$ ($k = n, \ldots, 3$) the parent of taxon $x_{k}$ is moved up by $\nni$ moves to the position it has in $R$.
 Notice that there might be more than just one such move per step, which means that $p$ can be extended by more than one tree in each step.
-After $n-2$ steps we reach $R$.
 
@@
 \STATE $[x_1, x_2, \ldots, x_n] := R$, $T' := T$, $p = [T']$
-\FOR {$i=n, \ldots, 3$}
-    \FOR {$j=1, \ldots, i$}
-        \IF {$T'[j] = x_i$}
-            \STATE Update $T'$: swap $T'[j]$ and $T'[j+1]$
+\FOR {$k=n, \ldots, 3$}
+    \FOR {$i=1, \ldots, k$}
+        \IF {$T'[i] = x_k$}
+            \STATE Update $T'$: swap $T'[i]$ and $T'[i+1]$
             \STATE $p = p + T'$
@@
 
-Let $T$ and $R$ now be caterpillar trees such that $d_c(T, R) = l < \frac{(n-1)(n-2)}{2}$.
+Let $T$ and $R$ be caterpillar trees such that $d_c(T, R) = l < \frac{(n-1)(n-2)}{2}$.
 For proving $d(T,R) = l$ we consider the shortest caterpillar path between these trees computed by $\csort$.

### 2019-12-10 rnni_convexity.tex
error in proof of Th caterpillar_convexity. Will need to see if I can fix that...

@@
 Let $T$ and $R$ be caterpillar trees such that $d_c(T, R) = l < \frac{(n-1)(n-2)}{2}$.
-For proving $d(T,R) = l$ we consider the shortest caterpillar path between these trees computed by $\csort$.
-
-Since $l < \frac{(n-1)(n-2)}{2}$, there must be a step in $\csort$ where the taxon that is moved does not require the maximum number of moves, meaning that it is not part of the cherry at the beginning of this step.
-Let $x$ be the first taxon that has this property and is considered in step number $k$.
+Since $d_c(T, R) < \frac{(n-1)(n-2)}{2}$, there must be a step in $\csort(T, R)$ where the taxon that is moved up is not part of the cherry (of $T'$) at the beginning of this step.
+Let $x$ be the first taxon that has this property and is considered at step $k$.
 Then there must be a taxon $y$ that is lower neighbour of $x$ in the tree at the beginning of step $k$.
@@
 Applying $\csort$ on $\hat T$ and $R$ results in almost the same path as the one computed by $\csort$ from $T$ to $R$, there is just one additional move in the beginning of step $k$ for swapping $x$ and $y$.
-It follows directly that $d(T, \hat T) = 1$ and $d_c(\hat T,R) = d_c(T,R) + 1 = l + 1$.
+It follows directly that
+\todo{No.
+As $x$ and $y$ don't have to be neighbours in $T$.}
+$d(T, \hat T) = 1$ and $d_c(\hat T,R) = d_c(T,R) + 1 = l + 1$.
 

### 2019-12-10 rnni_convexity.tex
problem with csort

@@
 
+\todo{This is not enough, as sometimes one $\rnni$ move removes two inversions.}
 The path $p$ returned by the algorithm $\csort$ is a shortest caterpillar path because every tree modification on $p$ reduces the number of inversions of taxa in $T$ and $R$.

### 2019-12-10 rnni_convexity.tex
sketch to fix the argument that csort gives a shortest path

--- 2019-12-10 rnni_convexity.tex
@@
 
-The path $p$ returned by the algorithm $\csort$ is a shortest caterpillar path because every tree modification on $p$ reduces the number of inversions of taxa in $T$ and $R$.
-
-Our Python implementation of this algorithm can be accessed at \autocite{Collienne2019}.
-
 \begin{algorithm}[H]
@@
 \end{algorithm}
+
+\begin{lemma}
+$\csort(T, R)$ returns a shortest caterpillar path between $T$ and $R$.
+\end{lemma}
+
+\proof
+Let $p$ be an arbitrary caterpillar path.
+Every $\rnni$ move along $p$ can decrease the number of inversions of taxa in $T'$ and $R$ by either $1$ or $2$.
+A decrease by $2$ is only possible when $T' = [x, y, z, \ldots]$, $z<\min(x,y)$, and the tree after the move is $[z, \min(x, y), \max(x, y), \ldots]$.
+Hence $p$ is a shortest caterpillar path if every move along $p$ either is of this type or decreases the number of inversions by one.
+Note that $\csort(T, R)$ satisfies this property.
+\endproof
+
+Our Python implementation of this algorithm can be accessed at \autocite{Collienne2019}.
 

### 2019-12-10 rnni_convexity.tex
csort still needs details

@@
 A decrease by $2$ is only possible when $T' = [x, y, z, \ldots]$, $z<\min(x,y)$, and the tree after the move is $[z, \min(x, y), \max(x, y), \ldots]$.
+\todo{Explain why more moves that fix two inversion each are not possible.}
 Hence $p$ is a shortest caterpillar path if every move along $p$ either is of this type or decreases the number of inversions by one.
@@
 
-The running time of $\csort$ is the same as for Bubble Sort, that is quadratic in $n$.
+The running time of $\csort$ is quadratic in $n$.
 

### 2019-12-10 rnni_convexity.tex
fixed the argument that csort returns a shortest caterpillar path

--- 2019-12-10 rnni_convexity.tex
@@
 
-\todo{This is not enough, as sometimes one $\rnni$ move removes two inversions.}
-The path $p$ returned by the algorithm $\csort$ is a shortest caterpillar path because every tree modification on $p$ reduces the number of inversions of taxa in $T$ and $R$.
-
-Our Python implementation of this algorithm can be accessed at \autocite{Collienne2019}.
-
 \begin{algorithm}[H]
@@
 
-The running time of $\csort$ is the same as for Bubble Sort, that is quadratic in $n$.
+The running time of $\csort$ is quadratic in $n$.
+
+The path $p$ returned by $\csort$ is a shortest caterpillar path because every tree modification on $p$ reduces the number of inversions of taxa in $T$ and $R$.
+Indeed, no $\rnni$ move on caterpillar trees can reduce the number of taxa inversions by more than one, assuming that inversions with both taxa of a cherry are counted only once, that is, if both pairs of taxa $(x, z)$ and $(y, z)$ appear in opposite orders in $T$ and $R$ and $(x,y)$ is a cherry then they are counted as one inversion.
+Hence any caterpillar path every move along which reduced the number of inversions is a shortest caterpillar path.
 
@@
 This fact will be established in Theorem~\ref{thm:caterpillar_convex}.
+
+Our Python implementation of this algorithm can be accessed at \autocite{Collienne2019}.
 

### 2019-12-10 rnni_convexity.tex
ne need shorter sudo-codes
hope for thm:caterpillar_convex

@@
 
+\todo{This format takes up too much space.
+Why do we need END statements?}
 \begin{algorithm}[H]
@@
 \begin{algorithmic}[1]
-\STATE $T' := T$
-\STATE $p := [T']$
+\STATE $T' := T$, $p := [T']$
 \FOR {$k = 1, \dots, n-2$}
@@
 
-\begin{proof}
-We prove this theorem by backward induction on the length of a shortest caterpillar path $d_c(T,R) = l$ ($l$ is the induction parameter) between trees $T$ and $R$.
-The backward induction basis follows by Lemma~\ref{lemma:caterpillar_dist=diameter}, as the maximum possible value for $l$ is $\frac{(n-1)(n-2)}{2}$.
+\proof
+Note that it is suffice to prove that $d_c(T, R) = d(T, R)$ for an arbitrary pair of caterpillar trees $T$ and $R$.
+Lemma~\ref{lemma:caterpillar_dist=diameter} implies that if $T$ and $R$ are at the maximum possible distance, that is, $d_c(T, R) = \frac{(n-1)(n-2)}{2}$.
+Denote the latter number by $D$.
 
-Let $T$ and $R$ be caterpillar trees such that $d_c(T, R) = l < \frac{(n-1)(n-2)}{2}$.
-Since $d_c(T, R) < \frac{(n-1)(n-2)}{2}$, there must be a step in $\csort(T, R)$ where the taxon that is moved up is not part of the cherry (of $T'$) at the beginning of this step.
+Assume now that $d_c(T, R) = D - 1$.
+If we apply $\csort(T, R)$ in this case, there must be a step in the algorithm where the taxon that is moved up is not part of the cherry (of $T'$) at the beginning of this step.
 Let $x$ be the first taxon that has this property and is considered at step $k$.
-Then there must be a taxon $y$ that is lower neighbour of $x$ in the tree at the beginning of step $k$.
-Since we assume that step $k$ is the first one that does not require the maximum number of $\nni$ moves, $y$ is lower neighbour of $x$ in $T$ as well.
+Then there must be a taxon $y$ that is immediately preceding $x$ in the tree at the beginning of step $k$.
+Since $k$ is the first such step, $y$ must be preceding $x$ in tree $T$ as well.
+Consider a tree $\hat T$ that is obtained from $T$ by exchanging $x$ and $y$.
+Note that the caterpillar path from $\hat T$ to $R$ computed by $\csort$ passes $T$ and coincides with the path $\csort(T, R)$ from there on.
+Clearly, $d_c(\hat T, R) = D$.
 
-Let $\hat T$ be a tree that equals $T$ but with $x$ and $y$ being exchanged.
-Applying $\csort$ on $\hat T$ and $R$ results in almost the same path as the one computed by $\csort$ from $T$ to $R$, there is just one additional move in the beginning of step $k$ for swapping $x$ and $y$.
-It follows directly that
-\todo{No.
-As $x$ and $y$ don't have to be neighbours in $T$.}
-$d(T, \hat T) = 1$ and $d_c(\hat T,R) = d_c(T,R) + 1 = l + 1$.
+Assuming $d_c(T, R) < D$, we can iterate the construction above to find a tree $\hat T$ such that $d_c(\hat T, R) = D$ and the caterpillar path from $\hat T$ to $R$ computed by $\csort$ passes $T$ and coincides with the path $\csort(T, R)$ from there on.
+Lemma~\ref{lemma:caterpillar_dist=diameter} implies that $d(\hat T, R) = D$.
 
-Therefore, we can apply the induction hypothesis for $\hat T$ and $R$ and get $d(\hat T,R) = l+1$.
-If it now were $d(T,R) < l$, it would be $d(\hat T,R) \leq d(T,R) + d(T,\hat T) < l + 1$ which contradicts the induction hypothesis.
-We can conclude that $d(T,R) = l$, which completes the proof.
-\end{proof}
+Assume that $d(T, R) < d_c(T, R)$ and let $p$ be a path from $T$ to $R$ of length $d(T, R)$.
+Consider a path $q$ obtained by concatenating paths $\csort(\hat T, T)$ and $p$.
+Note that $q$ is a path from $\hat T$ to $R$ which is shorter than $\csort(\hat T, R)$.
+Indeed, the two paths coincide between $\hat T$ and $T$ and $q$ is shorter between $T$ and $R$.
+Since $d_c(\hat T, R) = D$, the existence of $q$ is a contradiction to Lemma~\ref{lemma:caterpillar_dist=diameter}.
+\endproof
 

### 2019-12-11 rnni_convexity.tex
RNNI diameter proof is missing!

@@
 \usepackage{algorithm}
-\usepackage{algorithmic}
+\usepackage[noend]{algorithmic}
 \usepackage{todonotes}
@@
 
-In this section we establish diameter and radius of the $\rnni$ graph.
-In Theorem 7 of \autocite{Gavryushkin2018-ol} an upper bound for the diameter is given, but we are able to give the exact diameter using the results of Section~\ref{section:caterpillar_convex}.
-In the following we prove the exact diameter of $\rnni$ before we investigate the radius of $\rnni$.
+\textcite[Theorem~7]{Gavryushkin2018-ol} gave an upper bound for the diameter of $\rnni$ space.
+In this section we improve that result and calculate the exact diameter and radius of the $\rnni$ space.
+To achieve this we will employ our results on caterpillar trees from the previous sections.
 
-\begin{lemma}
-There are two caterpillar trees in $\rnni$ with distance $\frac{(n-1)(n-2)}{2}$.
-\label{lemma:caterpillar_diameter}
-\end{lemma}
-
-\begin{proof}
-Let $T = [1,2,3,\ldots,n]$ and $R = [n-1,n,n-2,n-3, \ldots, 1]$ be two caterpillar trees.
-According to Theorem~\ref{thm:caterpillar_convex}, there is a caterpillar path of length $d(T,R)$ between $T$ and $R$.
-We can compute a shortest caterpillar path by using $\csort$ as explained in Section~\ref{section:caterpillar_convex}.
-When running this algorithm to compute a path from $T$ to $R$, each taxon $i$ ($i = n, \ldots, 3$) swaps $i-2$ times with its upper neighbour.
-Therefore, the length of a shortest caterpillar path, and therefore the distance between $T$ and $R$, is $\sum\limits_{i=3}^{n}(i-2)  = \frac{(n-1)(n-2)}{2}$.
-\end{proof}
+The results of previous section imply that $d([1,2,3,\ldots,n], [n-1,n,n-2,n-3, \ldots, 1]) = \frac{(n-1)(n-2)}{2}$, which is the maximum possible distance between caterpillar trees.
 
@@
 \end{corollary}
+
+\proof
+\todo{Why exactly is this true for arbitrary, not necessarily caterpillar, trees?}
+\endproof
 

### 2019-12-11 rnni_convexity.tex
added RNNI diameter proof + a few more tweaks (show keys too)

@@
 \usepackage{csquotes}
+\usepackage[notref,notcite]{showkeys}
 \usepackage[style=authoryear,ibidtracker=false,uniquename=false,giveninits=true,terseinits=true,maxbibnames=5,backend=biber]{biblatex}
@@
 \newtheorem{lemma}[definition]{Lemma}
-\newtheorem{corollary}[definition]{Corollary}
 \newtheorem{proposition}[definition]{Proposition}
@@
 We hence focus on understanding the geometry of the space (graph) of ranked trees, the so-called ranked nearest neighbour interchange ($\rnni$) graph.
-For example, we find exact estimates of the radius and diameter of the space.
+For example, we find exact estimates of the radius and diameter of the space, improving the best previously known estimates.
 Since the $\rnni$ graph is a generalisation of the classical nearest neighbour interchange ($\nni$) graph to ranked phylogenetic trees, we compare geometric and algorithmic properties of the two graphs.
@@
 
-\todo{This format takes up too much space.
-Why do we need END statements?}
 \begin{algorithm}[H]
@@
 
-\textcite[Theorem~7]{Gavryushkin2018-ol} gave an upper bound for the diameter of $\rnni$ space.
+\textcite[Theorem~7]{Gavryushkin2018-ol} gave an upper bound for the diameter of $\rnni$ space, which we recall is denoted by $\Delta(\rnni)$.
+They showed that $\Delta(\rnni) = n^2 - 3n - 5/8$.
 In this section we improve that result and calculate the exact diameter and radius of the $\rnni$ space.
-To achieve this we will employ our results on caterpillar trees from the previous sections.
 
-The results of previous section imply that $d([1,2,3,\ldots,n], [n-1,n,n-2,n-3, \ldots, 1]) = \frac{(n-1)(n-2)}{2}$, which is the maximum possible distance between caterpillar trees.
-
-\begin{corollary}
-It is $\Delta(\rnniu) = \frac{(n-1)(n-2)}{2}$.
-\label{corollary:diameter}
-\end{corollary}
+\begin{theorem}
+$\Delta(\rnni) = \frac{(n-1)(n-2)}{2}$.
+\label{thm:diameter}
+\end{theorem}
 
 \proof
-\todo{Why exactly is this true for arbitrary, not necessarily caterpillar, trees?}
+Since $d_c([1,2,3,\ldots,n], [n-1,n,n-2,n-3, \ldots, 1]) = \frac{(n-1)(n-2)}{2}$, the claim follows directly from from Theorem~\ref{thm:caterpillar_convex} and Lemma~\ref{lemma:diameter_bound}.
 \endproof
 
-We now proceed to show that the radius of the $\rnni$ graph equals its diameter.
-Therefore we use the algorithm $\mdtree$ that has been introduced in Section~\ref{section:alg_mdtree}.
-Note that the results provided in this section are only true if the input tree of $\mdtree$ is a caterpillar tree.
+We show in the rest of this section that the radius of $\rnni$ space coincides with its diameter.
+The main tool to establish this result is the algorithm $\mdtree$ from Section~\ref{section:alg_mdtree}.
+Recall that $\mdtree$ is a non-deterministic algorithm, which implies that the output tree is not uniquely defined by the input to the algorithm.
 
@@
 Let $T$ be a caterpillar tree.
-Any tree $R = \mdtree(T)$ has distance $d(T,R) = \Delta(\rnni)$ to $T$.
+Then $d(T, R) = \Delta(\rnni)$ for every tree $R$ such that $R = \mdtree(T)$.
 \label{lemma:max_dist_caterpillar}
@@
 \begin{proof}
-Recall that with Corollary~\ref{corollary:diameter} we know that $\Delta(\rnni) = \frac{(n-1)(n-2)}{2}$.
+Recall that with Theorem~\ref{thm:diameter} we know that $\Delta(\rnni) = \frac{(n-1)(n-2)}{2}$.
 Within this proof we assume, without loss of generality, that it is $T = [1,2,3,\ldots,n]$.
@@
 And as permuting labels of trees does not change the distance between them, there exists for every tree $R$ on $n$ taxa a caterpillar tree $T$ with distance $d(T,R) = \Delta(\rnni) = \frac{(n-1)(n-2)}{2}$.
-By this we obtain Corollary~\ref{corollary:radius}.
+By this we obtain Theorem~\ref{thm:radius}.
 
-\begin{corollary}
+\begin{theorem}
 The radius of the $\rnni$ graph is $\rad(\rnni) = \Delta(\rnni) = \frac{(n-1)(n-2)}{2}$.
-\label{corollary:radius}
-\end{corollary}
+\label{thm:radius}
+\end{theorem}
 

### 2019-12-11 rnni_convexity.tex
everything up to Cluster Theorem looks okay now

@@
 \newcommand{\mdtree}{\textsc{MDTree}}
-\newcommand{\rad}{\textsc{R}}
 
@@
 \STATE Build tree $R$ with two taxa $l_1, l_2$ that are children of the root
-\FOR {$i = n, \ldots, 3$}
+\FOR {$i = 3, \ldots, n$}
 \STATE Extend $R$ by attaching $l_i$ to a newly created node of rank $1$ on an arbitrarily chosen edge
@@
 In Section~\ref{section:diameter} we will investigate the diameter and radius of the $\rnni$ space.
-Interestingly, the \emph{diameter} of $\rnni$, that is the maximum distance $\Delta(\rnni) = \max \{d(T, R) \mid T, R \in \rnni\}$ between any two trees in the graph, equals its \emph{radius} $\rad(\rnni) = \min\limits_T \max\limits_R d(T,R)$.
+Interestingly, the \emph{diameter} of $\rnni$, that is the maximum distance $\Delta(\rnni) = \max \{d(T, R) \mid T, R \in \rnni\}$ between any two trees in the graph, equals its \emph{radius} defined as $\min\limits_T \max\limits_R d(T,R)$.
 Next, we will consider the so-called Split Theorem \autocite{Gavryushkin2018-ol}, which states that every shared split of taxa is maintained along every shortest path.
@@
 Note that it is suffice to prove that $d_c(T, R) = d(T, R)$ for an arbitrary pair of caterpillar trees $T$ and $R$.
-Lemma~\ref{lemma:caterpillar_dist=diameter} implies that if $T$ and $R$ are at the maximum possible distance, that is, $d_c(T, R) = \frac{(n-1)(n-2)}{2}$.
+Lemma~\ref{lemma:caterpillar_dist=diameter} implies that if $T$ and $R$ are at the maximal possible distance, that is, $d_c(T, R) = \frac{(n-1)(n-2)}{2}$.
 Denote the latter number by $D$.
@@
 \proof
-Since $d_c([1,2,3,\ldots,n], [n-1,n,n-2,n-3, \ldots, 1]) = \frac{(n-1)(n-2)}{2}$, the claim follows directly from from Theorem~\ref{thm:caterpillar_convex} and Lemma~\ref{lemma:diameter_bound}.
+Since $d_c([1\ldots,n], [n-1,n,n-2,n-3, \ldots, 1]) = \frac{(n-1)(n-2)}{2}$, the claim follows directly from from Theorem~\ref{thm:caterpillar_convex} and Lemma~\ref{lemma:diameter_bound}.
 \endproof
@@
 \begin{proof}
-Recall that with Theorem~\ref{thm:diameter} we know that $\Delta(\rnni) = \frac{(n-1)(n-2)}{2}$.
-Within this proof we assume, without loss of generality, that it is $T = [1,2,3,\ldots,n]$.
 We prove the lemma by induction on the number of taxa $n$.
+The induction basis for $n = 3$ is trivial.
 
-The lemma is true for the induction basis on $n=3$ taxa since all three trees have distance one, which equals diameter and radius in this case.
-
-Let us now assume for the induction step that $T$ is a tree on $n+1$ taxa.
-Taxon $n+1$ is the last one in the list $L$ computed in $\mdtree$ as its parent in $T$ is the root.
-It follows that this taxon $n+1$ is the last one added to $R$ in the algorithm, so its parent has rank one in $R$.
-With Lemma~\ref{lemma:distance_delete_taxon} we can follow that for $T$ and $R$ restricted to taxa $\{1, \ldots, n\}$ it holds $d(T|_n,R|_n) \leq d(T,R) - (n-1)$.
-It is not hard to see that $R|_n$ can be received from $T|_n$ by applying $\mdtree$:
-$R|_n$ can be constructed from input tree $T|_n$ in the same way as $R$ can be constructed from $T$, since the list $L$ computed for $T|_n$ in $\mdtree$ is the same as the one for $T$, but without the last element.
-So we can follow from the induction hypothesis and Lemma~\ref{lemma:distance_delete_taxon} that it is $d(T,R) \geq d(T|_n,R|_n) + (n-1) = \frac{(n-1)(n-2)}{2} = \frac{n(n-1)}{2}$, which concludes the proof of the lemma.
+Assume now that $T = [1, \ldots, n+1]$.
+In this case, the parent of taxon $n+1$ has rank $n$ in $T$ and rank $1$ in $R$.
+Lemma~\ref{lemma:distance_delete_taxon} then implies that
+\begin{equation}
+d(T\big|_n,R\big|_n) \leq d(T,R) - (n-1).
+\label{eqn:max_dist_cat_proof}
+\end{equation}
+Note that $R\big|_n = \mdtree (T\big|_n)$ for an appropriate choice of attachment points in the execution of $\mdtree$.
+By induction hypothesis and Theorem~\ref{thm:diameter}, $d(T\big|_n,R\big|_n) = \frac{(n-1)(n-2)}{2}$.
+So inequality~(\ref{eqn:max_dist_cat_proof}) becomes $d(T,R) \geq \frac{(n-1)(n-2)}{2} + n - 1 = \frac{n(n-1)}{2}$, which with Theorem~\ref{thm:diameter} gives the desired equality.
 \end{proof}
 
-In Section~\ref{alg:max_dist_tree} we already discussed that $\mdtree$ is a non-deterministic algorithm.
-In fact, the output tree of $\mdtree$ can have any tree topology, since in each step of the algorithm the next taxon can be added on any edge incident to a leaf in the tree.
-Therefore, if the input tree is a caterpillar tree, $\mdtree$ can compute a tree with any tree topology that has maximum distance from the input tree.
-And as permuting labels of trees does not change the distance between them, there exists for every tree $R$ on $n$ taxa a caterpillar tree $T$ with distance $d(T,R) = \Delta(\rnni) = \frac{(n-1)(n-2)}{2}$.
-By this we obtain Theorem~\ref{thm:radius}.
+The non-determinism of $\mdtree$ can be exploited to investigate what trees are at the maximal possible from a given caterpillar tree.
+In fact, those trees can have an arbitrary (non-ranked) topology!
+This is because in each step of the algorithm the next taxon can be added on any edge incident to a leaf in the running tree.
+Therefore, if the input tree is a caterpillar tree, $\mdtree$ can be used to find a tree of pre-defined topology that is at the maximal possible distance from the input tree.
+Because the distance between trees is invariant under permutations of taxa labels, we conclude that for every tree $R$ on $n$ taxa there exists a caterpillar tree $T$ such that $d(T,R) = \Delta(\rnni) = \frac{(n-1)(n-2)}{2}$.
+This proves the following Theorem~\ref{thm:radius}.
 
 \begin{theorem}
-The radius of the $\rnni$ graph is $\rad(\rnni) = \Delta(\rnni) = \frac{(n-1)(n-2)}{2}$.
+The radius and diameter of the $\rnni$ space coincide and are equal to $\frac{(n-1)(n-2)}{2}$.
 \label{thm:radius}

### 2019-12-11 rnni_convexity.tex
move NP-hardness things from Split Theorem to Discussion
rm commeted out tex with some ideas and importantly LATTICES -- @lenacoll please check you don't need these and add the lattices section back
discussion at the beginning of section:cluster_theorem

@@
 
-\todo{Why are we in the Geometry section then?}
-In this section we will make a further step towards understanding the complexity of computing shortest paths in $\rnni$.
-Since the problem of computing distances in the most common tree spaces as $\nni$ \autocite{Dasgupta2000-xa}, rooted $\spr$ \autocite{Bordewich2005-nx}, or $\tbr$ \autocite{Allen2001-ky} is known to be $\np$-hard, it is desirable to find a tree space that is less complex.
-So far it is not known whether computing distances in $\rnni$ is $\np$-hard or not.
-The natural approach to solve this question is to compare $\rnni$ and $\nni$ graph to see whether the proof of $\np$-hardness of $\nni$ applies to $\rnni$ as well.
-We will explain why this is not the case by considering an example of shortest paths in $\nni$ that builds the key of the proof of its $\np$-hardness.
-After explaining why this example cannot be transferred to $\rnni$ we will introduce the so-called Split Theorem (Conjecture~\ref{conjecture:split_theorem}), which has been conjectured in \autocite{Gavryushkin2018-ol}.
-However, we present a counterexample disproving the Split Theorem in its original version and claiming the Cluster Theorem (Conjecture~\ref{conjecture:cluster_theorem}) as alternative version.
+In this section we discussed the so-called Split Theorem, which states that every split shared between two trees is maintained along shortest paths between the trees.
+For example, Split Theorem is true in the $\spr$ space and false in $\nni$ \autocite{Li1996-zw}.
+Specifically, in $\nni$ there exist two trees $T$ and $R$, which share a split $A | B$, such that every tree on every shortest path between $T$ and $R$ does not have $A | B$.
 
-The proof of $\np$-completeness of $\nni$ in \autocite{Dasgupta2000-xa} is based on the fact that Theorem~\ref{thm:split_nni} holds for $\nni$.
+To construct this example, \textcite{Li1996-zw} used the idea illustrated in Figure~\ref{fig:NNI_NP_proof} and showed that for an appropriate permutation of the taxa in the two caterpillar subtrees of both $T$ and $R$, it is strictly shorter to merge the two caterpillar subtrees first, then sort them simultaneously, and then split back to two caterpillar subtrees, rather the two caterpillar subtrees independently.
 
-\begin{theorem}
-There are trees $T,R$ in $\nni$ sharing a cluster which is not shared by any intermediate tree on any shortest path from $T$ to $R$.
-\label{thm:split_nni}
-\end{theorem}
-
-\begin{proof}
-See \autocite{Li1996-zw}.
-\end{proof}
-
-In the proof of Theorem~\ref{thm:split_nni} presented in \autocite{Li1996-zw} an example of two trees $T$ and $R$ that share clusters $C$ and $X \setminus C$ but have no shortest path that preserves these cluster is given.
-A simplified illustration of this example is provided in Figure~\ref{fig:NNI_NP_proof}.
-\autocite{Li1996-zw} prove that on a shortest path between $T$ and $R$ in $\nni$ at first $T$ is transformed to a tree $T'$ with $\frac{n}{2}$ cherries, each of which contains one taxon of each of the clusters $C$ and $X \setminus C$.
-Then $T'$ is sorted to a tree $R'$ containing the same cherries as $T'$ such that that resolving all these cherries of $R'$ results in $R$.
-Specifically, \autocite{Li1996-zw} prove that there is no shortest path between $T$ and $R$ in $\nni$ where $C$ and $X \setminus C$ are present in all trees on that path.
-
-The reason that this example cannot be transferred to $\rnni$ is that there are lots of additional rank moves needed on a path between $T$ ans $R$ similar to the one described above.
-As in the example depicted in Figure~\ref{fig:NNI_vs_RNNI}, adding rank moves to the shortest $\nni$ path from $T$ to $R$ results in a path that is longer than a shortest $\rnni$ path.
-Sorting the two caterpillar subtrees of $T$ and $R$ separately results in a shorter path in $\rnni$.
+The reason that this example cannot be transferred to $\rnni$ is that additional rank moves are necessary to merge and then split the two caterpillar subtrees.
+Similarly to the example in Figure~\ref{fig:NNI_vs_RNNI}, adding rank moves to the shortest $\nni$ path from $T$ to $R$ results in a path that is no longer a shortest $\rnni$ path.
+Sorting the two caterpillar subtrees of $T$ and $R$ independently results in a shorter path in $\rnni$.
 
@@
 
-% \todo{LC: We still need to check 7 taxa.}
-% We are doing this in the following way.
-% We focus on the subset of trees that induce share a cluster.
-% It is sufficient to consider clusters of the shape $\{1, \ldots, m\}$ for $2 \leq m \leq n-1$, because of the symmetry of the space:.
-% distances $d(T_1,R_1)$ and $d(T_2,R_2)$ between two pairs of trees $(T_1,R_1)$ and $(T_2,R_2)$ are equal if one can receive $T_2$ from $T_1$ by permuting the taxa the same way as it is necessary for receiving $R_2$ from $R_1$.
-% Therefore, we compute for all $m = 2, \ldots, n-1$ exact $\rnni$ distances between all pairs of trees containing the cluster $\{1, \ldots, m\}$, using Dijkstra's algorithm \autocite{Dijkstra1959-ph} on the $\rnni$ graph, which we can compute as explained in Section~\ref{section:alg_RNNI_graph}.
-% Additionally, we compute the subgraph of $\rnni$ induced by the subset of trees that include the cluster $\{1, \ldots, m\}$.
-% With the Floyd-Warshall algorithm we compute all pairwise distances between the trees in this subgraph and compare them with the distances in $\rnni$.
-% With this approach it is possible to show that the Cluster Theorem holds for up to seven taxa.
-% \todo{LC: We still need to check 7 taxa.}
-
-% \section{Ideas}
-%
-% [This section includes some ideas that we could possibly include in the paper]
-%
-% Using the same argument as in the proof of Lemma~\ref{lemma:distance_delete_taxon} we can establish Proposition~\ref{proposition:lower_bound_distance} that provides a lower bound on the distance between trees.
-%
-% \begin{proposition}
-% If trees $T$ and $R$ and taxon $x$ are such that $|\rank_T(\parent_T(x)) - \rank_R(\parent_R(x))| = \delta$ then $d(T,R) \geq \delta$.
-% \label{proposition:lower_bound_distance}
-% \end{proposition}
-%
-%
-% The paths computed by $\findpath$ depend on the order of input trees $T$ and $R$:
-% $\findpath(T,R)$ does not necessarily produce the reverse path of $\findpath(R,T)$.
-% An example can be found in Figure~\ref{fig:findpath_not_symmetric}.
-% However, simulations suggest that the lengths of the two paths computed by $\findpath$ for one pair of trees are equal.
-%
-% \begin{conjecture}
-% Let $T$ and $R$ be ranked trees.
-% The lengths of paths $\findpath(T,R)$ and $\findpath(R,T)$ are equal.
-% \end{conjecture}
-%
-% \begin{figure}[H]
-% \centering
-% \includegraphics[width=0.7\textwidth]{findpath_not_symmetric}
-% \vspace{12pt}
-% \caption{Two paths computed by $\findpath$.
-% At the top $\findpath(T,R)$, at the bottom $\findpath(R,T)$}
-% \label{fig:findpath_not_symmetric}
-% \end{figure}
-%
-% % idea for description of max distance caterpillar trees:
-% \begin{lemma}
-% Let $T$ and $R$ be caterpillar trees.
-% It is $d_c(T,R) = \Delta(\rnni)$ if, and only if, $T$ and $R$ share no induced triplet.
-% \todo{define induced triplets}
-% \end{lemma}
-%
-% This Lemma does only hold for caterpillar trees and not for trees with more than one cherry.
-% Furthermore, triplets do not represent ranked trees uniquely as they do not contain information about the ranks of internal nodes (but they can uniquely define caterpillar trees as these only have one cherry)
-%
-% \begin{proof}
-% %TODO
-% % Insert the proof here
-% Fix $T = (\ldots(1,2) \ldots ,n)$, use induction on $n$ and $\csort$ (taxon $n$ must be in cherry of $R$ if distance is max).
-% \end{proof}
-%
-% According to Algorithm~\ref{alg:max_dist_tree}, the number of trees with maximum distance to a given caterpillar tree $T$ is $(n-1)!$.
-% The number of caterpillar trees with distance $\Delta(\rnni)$ from a given caterpillar tree is $2^{n-2}$.
-%
-%
-% \subsection{Partition lattice}
-%
-% % Define Partition Lattice $\Pi_n$ and max chains in that lattice!
-%
-% \begin{theorem}
-% The $\rnni$ graph on $n$ taxa is isomorphic to the graph of maximal chains of the partition lattice $\Pi_n$ where two maximal chains are connected by an edge if and only if they differ by exactly one partition.
-% The corresponding metric spaces are isometric.
-% \end{theorem}
-
 \section{Discussion}
 
+Since the problem of computing distances in the most common tree spaces as $\nni$ \autocite{Dasgupta2000-xa}, rooted $\spr$ \autocite{Bordewich2005-nx}, or $\tbr$ \autocite{Allen2001-ky} is known to be $\np$-hard, it is desirable to find a tree space that is less complex.
+So far it is not known whether computing distances in $\rnni$ is $\np$-hard or not.
+The natural approach to solve this question is to compare $\rnni$ and $\nni$ graph to see whether the proof of $\np$-hardness of $\nni$ applies to $\rnni$ as well.
+We will explain why this is not the case by considering an example of shortest paths in $\nni$ that builds the key of the proof of its $\np$-hardness.
+After explaining why this example cannot be transferred to $\rnni$ we will introduce the so-called Split Theorem (Conjecture~\ref{conjecture:split_theorem}), which has been conjectured in \autocite{Gavryushkin2018-ol}.
+However, we present a counterexample disproving the Split Theorem in its original version and claiming the Cluster Theorem (Conjecture~\ref{conjecture:cluster_theorem}) as alternative version.
+
+The proof of $\np$-completeness of $\nni$ in \autocite{Dasgupta2000-xa} is based on the fact that Theorem~\ref{thm:split_nni} holds for $\nni$.
 The aim of this paper was to gain better understanding of the $\rnni$ space of ranked phylogenetic trees.

### 2019-12-11 rnni_convexity.tex
made it to Discussion

@@
 
-The nearest neighbour interchange ($\nni$) graph, defined on the set of phylogenetic trees with adjacency relation given by the interchange operation of two sister clades, has been known in mathematical biology literature for nearly 50 years \autocite{Robinson1971-ql,Moore1973-kk}.
+The nearest neighbour interchange ($\nni$) graph, defined on the set of phylogenetic trees with adjacency relation given by the interchange operation of two sister clades (subtrees), has been known in mathematical biology literature for nearly 50 years \autocite{Robinson1971-ql,Moore1973-kk}.
 Considered with the metric given by the length of a shortest path (graph-distance), this graph becomes a metric space.
@@
 This result suggests that the complexity of computing distances in $\rnni$ is different from that in the classical $\nni$ space.
-We will discuss this further in Section~\ref{section:cluster_theorem}, where we disprove a conjecture of \textcite{Gavryushkin2018-ol} and suggest its weaker version a plausible alternative.
+We will discuss this further in Section~\ref{section:cluster_theorem}, where we disprove a conjecture of \textcite{Gavryushkin2018-ol} and suggest its weaker versions a plausible alternatives.
 We will finish the paper with a discussion and open problems.
@@
 We will study the accuracy of this approximation by computing the exact $\rnni$ distance for small trees.
-Specifically, we will use the algorithm for computing the complete $\rnni$ graph that is given in \autocite[their Section 3.3]{Gavryushkin2018-ol}.
-After pre-computing the graph, we use Seidel's algorithm~\autocite{seidel_all-pairs-shortest-path_1995} for computing distances.
+Specifically, we will use the algorithm for computing the complete $\rnni$ graph that is given in \autocite[Section 3.3]{Gavryushkin2018-ol}.
+After pre-computing the graph, we use Seidel's algorithm \autocite{seidel_all-pairs-shortest-path_1995} for computing distances.
 However, the large number of vertices in $\rnni$ is an obstacle when it comes to computing distances for larger trees.
@@
 
+Note that $\rnni$ space on trees with seven taxa is a challenging goal to achieve computationally.
+For example, \textcite{Whidden2016-kl} were able to achieve up to seven taxa in the much smaller $\rnni$ space, when it required considering pairs of trees.
 %TODO: Make a zenodo dio for the repository before final publication
-We have implemented these algorithms in a combination of Python and C, and the source code is available at \autocite{Collienne2019}.
+We have implemented all our algorithms in a combination of Python and C, and the source code is available at \autocite{Collienne2019}.
 
@@
 Next, we will consider the so-called Split Theorem \autocite{Gavryushkin2018-ol}, which states that every shared split of taxa is maintained along every shortest path.
+Recall that a \emph{split} is a bipartition of the set of taxa obtained by deleting an edge of the tree $T$, splits are denoted by $A|B$.
 We will give a counterexample to the Split Theorem in $\rnni$ and consider a variant of the Weak Split Theorem, which states that every shared split of taxa is maintained along a shortest path.
-We will call this statement the Clade Theorem, as it is slightly more natural to work with ranked trees in terms clades rather than splits (see Conjecture~\ref{conjecture:cluster_theorem}).
+We will also consider the so-called Clade Theorem as another weak version of the Split Theorem (see Conjecture~\ref{conjecture:cluster_theorem}).
 
@@
 
-Since the example of a shortest $\nni$ described above, which was used to prove Theorem~\ref{thm:split_nni} in \autocite{Li1996-zw}, does not work in $\rnni$, a contrary statement was conjectured in \autocite{Gavryushkin2018-ol}.
-For understanding this conjecture it is important to know the connection between edges in a tree and so-called \emph{splits}, which are bipartitions of the taxon set of a tree:
-By deleting an edge $e$ of a tree $T$ one receives two graphs with labels in $A$ and $B$, respectively, such that $\{A,B\}$ is a partition of the taxon set of $T$.
-Note that each edge induces such a bipartition, often written as $A|B$, and that both edges incident to the root of a tree induce the same partition.
-Since these partitions are commonly referred to as splits, the following conjecture was named \emph{Split Theorem} in \autocite{Gavryushkin2018-ol}.
+This argument motivated \textcite[Conjecture~9]{Gavryushkin2018-ol} to conjecture that every split shared between trees in $\rnni$ is maintained along every shortest path.
 
-\begin{conjecture}[Split Theorem]
-For the $\rnni$ graph the following statement holds:
-If a partition of leaves given by an edge is present in two trees $T$ and $R$ then the partition is presented in every tree on every shortest path between $T$ and $R$.
-\label{conjecture:split_theorem}
-\end{conjecture}
-
-We found a simple counterexample to this conjecture, provided in Figure~\ref{fig:splitthm_counterexample}.
+The following example in Figure~\ref{fig:splitthm_counterexample} shows that this form of the Split Theorem does now hold.
 
@@
 
-Since the counterexample in Figure~\ref{fig:splitthm_counterexample} shows that the version of the Split Theorem stated in \autocite{Gavryushkin2018-ol} does not hold, we will now claim an alternative to this conjecture, the \emph{Cluster Theorem} (Conjecture~\ref{conjecture:cluster_theorem}).
-The Cluster Theorem concerns clusters instead of splits.
-This is motivated by the fact that rooted phylogenetic trees can uniquely be represented by sets of clusters \autocite{Steel2016-ye}, but not by sets of splits as they cannot define the position of the root.
-Note that trees $T$ and $R$ in Figure~\ref{fig:splitthm_counterexample} induce the same set of splits, but share no cluster.
+Hence not every shortest path between trees in $\rnni$ has to maintain every shared split of taxa.
+However, \textbf{there exists} still another shortest path connecting the two trees, along which every shared split is maintained.
+This motivates the study of the following Weak Split Theorem in $\rnni$.
+
+\begin{conjecture}[Weak Split Theorem]
+For the $\rnni$ graph the following statement holds:
+If a partition of leaves given by an edge is present in two trees $T$ and $R$ then the partition is presented in every tree on every shortest path between $T$ and $R$.
+\label{conjecture:split_theorem}
+\end{conjecture}
+
+Another interesting property of our counterexample in Figure~\ref{fig:splitthm_counterexample} is that the set of taxa $\{1, 2, 3\}$, which is a part of the split in both $T$ and $R$, does not form a \textbf{cluster} in $R$, that is, there is no single node in the tree all descended taxa of which are exactly $1$, $2$, and $3$.
+Importantly, rooted phylogenetic trees can be uniquely represented by sets of clusters \autocite{Steel2016-ye}, but not by sets of splits as splits cannot define the position of the root.
+For example, trees $T$ and $R$ in Figure~\ref{fig:splitthm_counterexample} induce the same set of splits, but share no cluster.
+
+Furthermore, all our methods in this paper for computing shortest paths in special subspaces of $\rnni$ demonstrate that that shared clusters are maintained along shortest paths.
+
+These considerations motivate us to conjecture the following strong version of the Cluster Theorem.
 
 \begin{conjecture}[Cluster Theorem]
-For the $\rnni$ graph the following statement holds:
-if two trees $T$ and $R$ contain the same cluster $C$, then $C$ is present as cluster in every tree on every shortest path between $T$ and $R$.
+Let $T$ and $R$ be trees in $\rnni$ that share a cluster $C$.
+Then $C$ is present as a cluster in every tree on every shortest path between $T$ and $R$.
 \label{conjecture:cluster_theorem}
@@
 
-For providing further evidence that the Cluster Theorem holds in $\rnni$, we could show computationally that it holds for the $\rnni$ graph on small trees with up to seven taxa.
-We did this by computing subgraph of $\rnni$ only containing trees that share a cluster and compare distances within this subgraph with distances in the whole $\rnni$ graph \autocite{Collienne2019}.
-We computed these graphs according to the algorithm of \autocite{Gavryushkin2018-ol} as discussed in Section~\ref{section:algorithms}.
+We have computationally checked that this conjecture is true in $\rnni$ spaces with up to seven taxa ($n = 2, \ldots, 7$).
+%TODO
+Specifically, we computed subgraphs of $\rnni$ containing only trees with shared clusters and compared distances in this subgraph with distances in the whole $\rnni$ graph.
+We computed these graphs according to the algorithm from \autocite{Gavryushkin2018-ol} as discussed in Section~\ref{section:algorithms}.
+The source code of all these implementations can be openly accessed at \autocite{Collienne2019}.
 

### 2019-12-11 rnni_convexity.tex
Discussion, sitations, acknowledgements

@@
 \usepackage{csquotes}
-\usepackage[notref,notcite]{showkeys}
+%\usepackage[notref,notcite]{showkeys}
 \usepackage[style=authoryear,ibidtracker=false,uniquename=false,giveninits=true,terseinits=true,maxbibnames=5,backend=biber]{biblatex}
@@
 
-Since the problem of computing distances in the most common tree spaces as $\nni$ \autocite{Dasgupta2000-xa}, rooted $\spr$ \autocite{Bordewich2005-nx}, or $\tbr$ \autocite{Allen2001-ky} is known to be $\np$-hard, it is desirable to find a tree space that is less complex.
-So far it is not known whether computing distances in $\rnni$ is $\np$-hard or not.
-The natural approach to solve this question is to compare $\rnni$ and $\nni$ graph to see whether the proof of $\np$-hardness of $\nni$ applies to $\rnni$ as well.
-We will explain why this is not the case by considering an example of shortest paths in $\nni$ that builds the key of the proof of its $\np$-hardness.
-After explaining why this example cannot be transferred to $\rnni$ we will introduce the so-called Split Theorem (Conjecture~\ref{conjecture:split_theorem}), which has been conjectured in \autocite{Gavryushkin2018-ol}.
-However, we present a counterexample disproving the Split Theorem in its original version and claiming the Cluster Theorem (Conjecture~\ref{conjecture:cluster_theorem}) as alternative version.
+The problem of computing distances in all common phylogenetic graphs, including $\nni$ \autocite{Dasgupta2000-xa}, $\spr$ \autocite{Bordewich2005-nx}, and $\tbr$ \autocite{Allen2001-ky}, is $\np$-hard.
+Although $\np$-hardness does not necessarily imply practical impossibility, this so far has been the case for these phylogenetic algorithms with extremely few major advances \autocite{Whidden2010-bw}.
+$\nni$ is especially known for its algorithmic hardness \autocite{Whidden2018-fw}.
 
-The proof of $\np$-completeness of $\nni$ in \autocite{Dasgupta2000-xa} is based on the fact that Theorem~\ref{thm:split_nni} holds for $\nni$.
-The aim of this paper was to gain better understanding of the $\rnni$ space of ranked phylogenetic trees.
-Before investigating geometrical properties of $\rnni$ we introduced some algorithms in Section~\ref{section:algorithms}).
-One of the most important algorithms we introduced here is $\findpath$, which can be used for approximating $\rnni$ distances.
-We could in particular show that this algorithm computes exact distances for all pairs of trees on up to six taxa.
-This algorithm might be a useful tool for future investigations of the $\rnni$ space.
+Our results suggest a surprising advance in this field of $\np$-problems -- adding ranking information to trees simplifies some of the algorithms and makes some unintuitive counterexamples impossible.
+We hence continued the investigation of the $\rnni$ graph on ranked phylogenetic trees in this paper.
+We have taken the route of comparing the classic $\nni$ with the $\rnni$ graph, so have mostly been concentrating on questions that are settled in $\nni$.
+Specifically, we have considered a number of geometric properties, including the diameter, radius, convexity of caterpillar trees, and the Split Theorem.
+We have also justified that the technique of \textcite{Dasgupta2000-xa} to prove that distances in $\nni$ are $\np$-hard to compute cannot be directly applied in $\rnni$.
+This led us to the Cluster Theorem, a statement that does not hold in $\nni$.
+All our algorithms and methods developed in this paper witness in favour of the Cluster Theorem being true in $\rnni$.
+For example, we have checked the statement computationally for up to seven taxa \autocite{Collienne2019}.
+Although the Split Theorem is true in $\spr$, it is still $\np$-hard to compute distances in that graph.
+Hence Cluster Theorem alone cannot be used as an argument in favour of the distance problem having an efficient solution in $\rnni$.
+But the greedy nature of all our algorithms for computing shortest paths in $\rnni$ developed in this paper does provide such an argument.
 
-We used the algorithms introduced in Section~\ref{section:algorithms} to prove some geometric properties of the $\rnni$ graph such as diameter and radius.
-Moreover, we discovered that the set of caterpillar trees is convex in $\rnni$ (Theorem~\ref{thm:caterpillar_convex}).
-With this result and the algorithm $\csort$ for computing shortest caterpillar paths (Section~\ref{section:alg_csort}) we can compute $\rnni$ distances between caterpillar trees in polynomial time.
+The algorithms we developed and implemented in this paper have served as a main ingredient for our study of the geometry of $\rnni$ space.
+These algorithms are of interest on their own as well.
+For example, our $\findpath$ algorithm gives a good approximation of the $\rnni$ distance, performing exactly on small trees.
+We are not aware of the minimal number of taxa where this algorithm fails to return the correct distance.
+The $\mdtree$ algorithm produces trees as far away from a given tree as possible, a task of importance in simulation and model comparison studies.
+The $\csort$ algorithm efficiently computes the $\rnni$ distance and shortest paths between caterpillar trees exactly.
+To the best of our knowledge no such algorithm exists for $\nni$.
+This implies that the algorithmic complexity of comping distances in $\nni$ is higher than in $\rnni$.
 
-This already suggests that there is a difference in complexity of computing distances in $\rnni$ and $\nni$, since in $\nni$ there is no subspace known for which computing distances only takes polynomial time.
-In Section~\ref{section:cluster_theorem} we supported the claim that computing $\rnni$ distances is not as hard as computing $\nni$ distances further by discussing the proof of $\np$-completeness of computing distances in $\nni$ and showing that this proof does not work for $\rnni$.
-Specifically, we conjectured the Cluster Theorem (Conjetcure~\ref{conjecture:cluster_theorem}), which does not hold in $\nni$.
-However, we strongly suspect that it holds in $\rnni$ and could prove that it does hold for trees on up to seven taxa \autocite{Collienne2019}.
+The question of whether computing $\rnni$ distances is $\np$-hard is still an open problem.
 
-\subsection{Open Problems}
+\section*{Acknowledgements}
+We thank Charles Semple for useful discussions about the Cluster Theorem.
 
-In Section~\ref{section:cluster_theorem} we claimed that the Cluster Theorem holds for $\rnni$.
-We could show that the previous version of it, the Split Theorem as it was conjectured in \autocite{Gavryushkin2018-ol}, is not true.
-If the Cluster Theorem actually is true still remains an open question.
-Proving it will be a big step towards finding the complexity of computing $\rnni$ distances.
+We acknowledge support from the Royal Society of New Zealand through the Rutherford Discovery Fellowship awarded to AG.
+This work was partially supported by MBIE through the Endeavour Smart Ideas and Data Science Programmes grants.
 
-This question of whether computing $\rnni$ distances is $\np$-hard is another problem for future research.
-Though we did not aim to answer this question in this paper, we made some progress towards solving it by establishing some geometric properties of the $\rnni$ graph.
-Comparing these results with what is known for the plain $\nni$ graph makes us assume that the complexity of computing shortest path in $\rnni$ might differ from $\nni$.
-We also suspect that the fact that computing distances between caterpillar trees can be done in polynomial time might be the key for further investigation of the complexity of computing $\rnni$ distances.
+Part of this work was done while LC and AG were visiting Max Plank Institute for Evolutionary Biology, we appreciate their support.
 
 \printbibliography
-
 \end{document}

### 2019-12-12 rnni_convexity.tex
rewrote two paragraphs about FindPath

@@
 We then say that $v$ \emph{induces} $C$.
-\emph{Trivial clusters} are clusters induced by leaves (contain just one element each) and the root of the tree (contains all taxa).
-As all trees on $n$ taxa share all trivial clusters, we will use the notion cluster to mean non-trivial clusters.
-Note that the list of all non-trivial clusters of a tree, sorted according to the rank of the corresponding node, defines the tree unambiguously.
+All trees on $n$ leaves share trivial clusters, which are those induced by leaves and the root, so we exclude them from consideration and say cluster to mean a non-trivial cluster.
+A tree in unambiguously defined by the list of its clusters sorted according to the rank of the inducing nodes.
 We will refer to this list of $n-2$ clusters for a given tree on $n$ taxa as a \emph{cluster representation} of the tree.
@@
 The algorithm terminates when $p$ is an $\rnni$ path from $T$ to $R$.
-
-\todo{Alex: go through the following two paragraphs and proof of L1}
-In step $k$ of $\findpath$ we consider the cluster $C_k$ for $k = 1, \ldots, n-2$.
-$p$ is extended in step $k$ by considering the last tree $T'$ on $p$ and finding the node in $T'$ that induces the smallest cluster containing all elements of $C_k$.
+At step $k = 1, \ldots, n-2$ of $\findpath$ we consider the cluster $C_k$ and extend path $p$ as follows.
+We search for a node in the last tree $T'$ of $p$ from the previous step that induces the smallest cluster containing all elements of $C_k$.
 This node is denoted by $\mrca_{T'}(C_k)$ (short for \emph{most recent common ancestor}).
-By performing an $\rnni$ move that decreases the rank of $\mrca_{T'}(C_k)$, we receive a tree that is then appended to $p$.
-Now we extend $p$ by repeatedly performing $\rnni$ moves on the last tree of $p$ that move the most recent common ancestor of $C_k$ down.
-In the end of step $k$, $C_k$ is induced by the internal node of rank $k$ in the last tree of $p$.
-
-By following this procedure, the last of $p$ after step $n-2$ will be $R$, as it contains all clusters of $R$.
-It is important but easy to see that the output path of $\findpath$ is unique for a given pair of trees, which means that it is a deterministic algorithm.
-This is due to the fact that there exists exactly one $\rnni$ move that decreases the rank of the most recent common ancestor of a subset $C_k$ of the taxon set of a tree as long as $C_k$ is not induced by an internal node of that tree.
+We then extend $p$ by adding the tree obtained from $T'$ by performing the $\rnni$ move that decreases the rank of $\mrca_{T'}(C_k)$.
+We continue extending $p$ in this way by repeatedly performing $\rnni$ moves until the rank of $\mrca_{T'}(C_k) = k$, thus completing step $k$.
 
@@
 
-The following lemma shows that $\findpath$ is correct, that is, the algorithm returns a path between the two input trees.
-To prove this we have to establish that there always exists a unique $\rnni$ move required in Line~\ref{alg:findpath:line:move_set_down} of the algorithm.
+The following lemma shows that $\findpath$ is a correct deterministic algorithm.
 
 \begin{lemma}
-Let $T = [C_1, \ldots, C_{k-1}, B_{k}, \ldots, B_{n-2}]$ and $R = [C_1, \ldots, C_{k-1}, C_k, \ldots, C_{n-2}]$ be two trees where $C_k \neq B_k, \ldots, B_{n-2}$.
+Let $T = [C_1, \ldots, C_{k-1}, B_{k}, \ldots, B_{n-2}]$ and $R = [C_1, \ldots, C_{k-1}, C_k, \ldots, C_{n-2}]$ be two trees such that $C_k \neq B_k, \ldots, B_{n-2}$.
 Then there exists exactly one $\rnni$ move on $T$ that results in a tree $T'$ such that $\mrca_{T'}(C_k) < \mrca_{T}(C_k)$.

### 2019-12-12 rnni_convexity.tex
i think there is an error in lemma:mrca_move statement. fixed the error. hope the proof goes through

@@
 \usepackage{csquotes}
-%\usepackage[notref,notcite]{showkeys}
+\usepackage[notref,notcite]{showkeys}
 \usepackage[style=authoryear,ibidtracker=false,uniquename=false,giveninits=true,terseinits=true,maxbibnames=5,backend=biber]{biblatex}
@@
 \begin{lemma}
-Let $T = [C_1, \ldots, C_{k-1}, B_{k}, \ldots, B_{n-2}]$ and $R = [C_1, \ldots, C_{k-1}, C_k, \ldots, C_{n-2}]$ be two trees such that $C_k \neq B_k, \ldots, B_{n-2}$.
+Let $T = [C_1, \ldots, C_{k-1}, B_{k}, \ldots, B_{n-2}]$ and $R = [C_1, \ldots, C_{k-1}, C_k, \ldots, C_{n-2}]$ be two trees such that $C_k \neq B_k$.
 Then there exists exactly one $\rnni$ move on $T$ that results in a tree $T'$ such that $\mrca_{T'}(C_k) < \mrca_{T}(C_k)$.

### 2019-12-12 rnni_convexity.tex
nope proof has holes two...

@@
 
+\todo{We never use $R$ in this proof, so $B_k$'s are irrelevant in Lemma.}
 \begin{proof}
@@
 
-Note that the uniqueness of the $\rnni$ move in Line~\ref{alg:findpath:line:move_set_down} of the algorithm $\findpath$ also implies that the algorithm is deterministic.
-The running time of the algorithm is quadratic in the number of taxa.
+Clearly, the worst-case complexity of $\findpath$ is quadratic in the number of taxa.
 
@@
 A natural question then is how accurate this approximation is.
-We have computationally shown that the algorithm $\findpath$ finds the correct distance for trees up to six taxa.
-Specifically, we have implemented $\findpath$ \autocite{Collienne2019} and compared the results with the true $\rnni$ distance as explained earlier in this section.
+We have computationally shown that the algorithm $\findpath$ finds the correct distance for trees up to seven taxa \autocite{Collienne2019}.
 

### 2019-12-12 rnni_convexity.tex
I'm no longer sure about the lemma. Will have to go through this whole section again...

@@
 \begin{lemma}
-Let $T = [C_1, \ldots, C_{k-1}, B_{k}, \ldots, B_{n-2}]$ and $R = [C_1, \ldots, C_{k-1}, C_k, \ldots, C_{n-2}]$ be two trees such that $C_k \neq B_k$.
-Then there exists exactly one $\rnni$ move on $T$ that results in a tree $T'$ such that $\mrca_{T'}(C_k) < \mrca_{T}(C_k)$.
+Let $T = [C_1, \ldots, C_{k-1}, B_{k}, \ldots, B_{n-2}]$ be a tree and $C_k$ a cluster such that $\mrca_T(C_k) > k$.
+Then there exists exactly one $\rnni$ move on $T$ that results in a tree $T' = [C_1, \ldots, C_{k-1}, \ldots]$ with $\mrca_{T'}(C_k) < \mrca_{T}(C_k)$.
 \label{lemma:mrca_move}

### 2019-12-12 rnni_convexity.tex
fixed proof of FindPath correctness

@@
 \caption{$\findpath$($T,R$)}
-\label{alg:find_path}
 \begin{algorithmic}[1]
-\STATE $T' := T$, $p := [T']$
+\STATE $T' := T$, $p := [T']$, $[C_1, \ldots, C_{n-2}] := R$
 \FOR {$k = 1, \dots, n-2$}
-  \STATE Let $C_k$ be the cluster induced by the node with rank $k$ in $R$ \label{alg:find_path:line:cluster}
   \WHILE {$\rank_{T'}(\mrca_{T'}(C_k))>k$}
-    \STATE Update $T'$: Decrease $\rank_{T'}(\mrca_{T'}(C_k))$ by an $\rnni$ move \label{alg:findpath:line:move_set_down}
+    \STATE Update $T'$: Decrease $\rank_{T'}(\mrca_{T'}(C_k))$ by an $\rnni$ move
+    \label{alg:findpath:line:move_set_down}
     \STATE $p = p+T'$
@@
 
-The following lemma shows that $\findpath$ is a correct deterministic algorithm.
+\begin{proposition}
+$\findpath$ is a correct deterministic algorithm.
+\end{proposition}
 
-\begin{lemma}
-Let $T = [C_1, \ldots, C_{k-1}, B_{k}, \ldots, B_{n-2}]$ be a tree and $C_k$ a cluster such that $\mrca_T(C_k) > k$.
-Then there exists exactly one $\rnni$ move on $T$ that results in a tree $T' = [C_1, \ldots, C_{k-1}, \ldots]$ with $\mrca_{T'}(C_k) < \mrca_{T}(C_k)$.
-\label{lemma:mrca_move}
-\end{lemma}
+\proof
+It is suffice to show that the update operation (line~\ref{alg:findpath:line:move_set_down}) is well-defined, that is the $\rnni$ move that decreases the rank of $\mrca_{T'}(C_k)$ is unique.
 
-\todo{We never use $R$ in this proof, so $B_k$'s are irrelevant in Lemma.}
-\begin{proof}
-Let $T$ and $R$ be trees as defined in the lemma.
-Notice that the only $\rnni$ move on $T$ that could decrease the rank of $\mrca_T(C_k)$ are $\nni$ moves on the edge between $\mrca_T(C_k)$ and its child (if this edge has length one), or a rank move between $\mrca_T(C_k)$ and the node with rank one less.
-Therefore, we will now distinguish these two cases:
-Either there is an edge $(v,w)$ between nodes $v = \mrca_T(C_k)$ and $w$ with rank $rank_T(w) = rank_T(v) - 1$, or there is no such edge.
+Case $k = 1$.
+In this case $C_k$ is a cherry $(x, y)$.
+The node $v = \mrca_{T'}(x, y)$ has $\rank_{T'}(v) = r > 1$.
+Consider the node $w$ preceding $v$ in $T'$, that is, $\rank_{T'}(w) = r - 1$.
+If a rank move that swaps $v$ and $w$ is possible then this is the only move on $T'$ that can decrease the rank of $\mrca(x, y)$.
+If the rank move is impossible then there is an edge in $T'$ connecting $v$ and $w$.
+The only way to decrease the rank of $\mrca(x, y)$ then is to perform an $\nni$ move on that edge.
+It is not hard to see that out of two possible $\nni$ moves only one decreases the rank of $\mrca$.
 
-In the latter case the only $\rnni$ move that could decrease the rank of $v$ is a rank move swapping the ranks of $v$ and $w$.
-
-Let us now assume that there is an edge $(v,w)$ in $T$.
-Let $T_v$ be the subtree that is rooted in the child of $v$ that is not $w$ and let $T_{w_1}$ and $T_{w_2}$ be the subtrees rooted in the children of $w$.
-We are now going to explain why only one of these two $\nni$ moves changes the rank of $\mrca_T(C_k)$.
-It is important to notice that $C_k$ either is the disjoint union of two of the clusters $C_1, \ldots, C_{k-1}$ or the union of one of those clusters with a single taxon that is not included in any of those clusters.
-In general we can say that $C_k = S_1 \dot\cup S_2$ is the union of two subsets of taxon set of $T$ such each of these subsets is contained in one of the subtrees $T_v, T_{w_1}$, or $T_{w_2}$.
-In fact, $S_1$ and $S_2$ are contained in two different of these subtrees.
-Now it is easy to see that just one of the $\nni$ moves on $(v,w)$ (illustrated in Figure~\ref{fig:mrca_move}) decreases the rank of $\mrca_T(C_k)$.
-
-\begin{figure}[H]
-\centering
-\includegraphics[width=0.5\textwidth]{mrca_move}
-\vspace{12pt}
-\caption{Two $\nni$ moves that are possible on the edge $(v,w)$ of the tree $T'$ with subtrees $T_v, T_{w_1}, T_{w_2}$ as described in the proof of Lemma~\ref{lemma:mrca_move}.}
-\label{fig:mrca_move}
-\end{figure}
-\end{proof}
+Case $k > 1$.
+\begin{enumerate}
+\item $C_k = C_i \cup C_j$ for $i, j < k$.
+Suppress $C_i$ and $C_j$ in both $T'$ and $R$ to new taxa $c_i$ and $c_j$ respectively and proceed as in Case $k = 1$.
+\item $C_k = C_i \cup {x}$, where $x$ is a taxon not present in $C_1, \ldots, C_k$.
+Suppress $C_i$ in both $T'$ and $R$ to a new taxon $c_i$ and proceed as in Case $k = 1$.
+\item $C_k = \{x, y\}$ where both $x$ and $y$ are not in $C_1, \ldots, C_k$.
+Proceed as in Case $k = 1$.
+\end{enumerate}
+\endproof
 
@@
 
-The algorithm $\csort$ (Algorithm~\ref{alg:csort}) is a natural modification of the classical Bubble Sort algorithm \autocite{Knuth1997-pi}.
-A path $p$ from $T$ to $R= [x_1, x_2, \ldots, x_n]$ is computed iteratively so that after $k$ steps the last $k$ taxa of $T$ and $R$ coincide.
+The algorithm $\csort$ (Algorithm~\ref{alg:csort}) is a modification of the classical Bubble Sort algorithm \autocite{Knuth1997-pi}.
+A path $p$ from $T$ to $R= [x_1, \ldots, x_n]$ is computed iteratively so that after $k$ steps the last $k$ taxa of $T$ and $R$ coincide.
 Specifically, in step $k$ ($k = n, \ldots, 3$) the parent of taxon $x_{k}$ is moved up by $\nni$ moves to the position it has in $R$.

### 2019-12-15 rnni_convexity.tex
hide keys
refs for cluster representation + MBIE expanded

@@
 A tree in unambiguously defined by the list of its clusters sorted according to the rank of the inducing nodes.
-We will refer to this list of $n-2$ clusters for a given tree on $n$ taxa as a \emph{cluster representation} of the tree.
+We will refer to this list of $n-2$ clusters for a given tree on $n$ taxa as a \emph{cluster representation} of the tree \autocite{Gavryushkin2018-ol, Gavryushkin2016-uu, Semple2003-nj}.
 
@@
 We acknowledge support from the Royal Society of New Zealand through the Rutherford Discovery Fellowship awarded to AG.
-This work was partially supported by MBIE through the Endeavour Smart Ideas and Data Science Programmes grants.
+This work was partially supported by the Ministry of Business, Innovation, and Employment of New Zealand through the Endeavour Smart Ideas and Data Science Programmes grants.
 

### 2019-12-16 rnni_convexity.tex
problem with FindPath
corrected Mareike's affiliation English name
Weak Split Conjecture was full of typos

@@
 \email{email@mareikefischer.de}
-\address{\textsuperscript{2}Institute for Mathematics and Inofrmatics, University of Greifswald, Germany}
+\address{\textsuperscript{2}Institute of Mathematics and Computer Science, University of Greifswald, Germany}
 \author{David Bryant\textsuperscript{3}}
@@
 \begin{conjecture}[Weak Split Theorem]
-For the $\rnni$ graph the following statement holds:
-If a partition of leaves given by an edge is present in two trees $T$ and $R$ then the partition is presented in every tree on every shortest path between $T$ and $R$.
+If a split of taxa $A | B$ is shared by two trees $T$ and $R$ in $\rnni$ then there exists a shortest path $p$ between $T$ and $R$ such that all trees on $p$ share the split $A | B$.
 \label{conjecture:split_theorem}

### 2019-12-19 rnni_convexity.tex
Mareike's comments

@@
 We hence focus on understanding the geometry of the space (graph) of ranked trees, the so-called ranked nearest neighbour interchange ($\rnni$) graph.
-For example, we find exact estimates of the radius and diameter of the space, improving the best previously known estimates.
+For example, we find the radius and diameter of the space exactly, improving the best previously known estimates.
 Since the $\rnni$ graph is a generalisation of the classical nearest neighbour interchange ($\nni$) graph to ranked phylogenetic trees, we compare geometric and algorithmic properties of the two graphs.
@@
 An example of such a tree is depicted in Figure~\ref{fig:ranked_tree}.
-In the terminology of \autocite{Gavryushkin2018-ol}, the space considered in this paper is the space of ranked ultrametric phylogenetic trees $\rnniu$.
+In the terminology of \autocite{Gavryushkin2018-ol}, the space considered in this paper is the space of ranked ultrametric phylogenetic trees ($\rnniu$ in their notations).
 We postpone accurate definitions until later in this section.
@@
 Our definition of a ranked tree implies a natural notion of edge length -- we call $|\rank_T(v) - \rank_T(u)|$ the length of an edge $(u, v)$.
+We assume that edges of trees are undirected, so $(u, v) = (v, u)$.
 
@@
 In this section we introduce three algorithms, $\findpath$ (Section~\ref{section:alg_findpath}), $\csort$ (Section~\ref{section:alg_csort}), and $\mdtree$ (Section~\ref{section:alg_mdtree}).
-$\findpath$ computes a $\rnni$ path between trees.
+$\findpath$ computes an $\rnni$ path between trees.
 $\csort$ computes a shortest $\rnni$ path between caterpillar trees (also known as ladder trees, see later in this section for accurate definitions).
@@
 All trees on $n$ leaves share trivial clusters, which are those induced by leaves and the root, so we exclude them from consideration and say cluster to mean a non-trivial cluster.
-A tree in unambiguously defined by the list of its clusters sorted according to the rank of the inducing nodes.
+A tree is unambiguously defined by the list of its clusters sorted according to the rank of the inducing nodes.
 We will refer to this list of $n-2$ clusters for a given tree on $n$ taxa as a \emph{cluster representation} of the tree \autocite{Gavryushkin2018-ol, Gavryushkin2016-uu, Semple2003-nj}.
@@
 \proof
-It is suffice to show that the update operation (line~\ref{alg:findpath:line:move_set_down}) is well-defined, that is the $\rnni$ move that decreases the rank of $\mrca_{T'}(C_k)$ is unique.
+It suffices to show that the update operation (line~\ref{alg:findpath:line:move_set_down}) is well-defined, that is the $\rnni$ move that decreases the rank of $\mrca_{T'}(C_k)$ is unique.
 
 Case $k = 1$.
-In this case $C_k$ is a cherry $(x, y)$.
+In this case $C_k$ is a consists of two taxa $\{x, y\}$.
 The node $v = \mrca_{T'}(x, y)$ has $\rank_{T'}(v) = r > 1$.
@@
 In this section we restrict our attention to the set of caterpillar trees.
-These trees are of particular interest because they are used to prove the computing the $\nni$ distance is an NP-hard problem.
+These trees are of particular interest because they are used to prove that computing the $\nni$ distance is an NP-hard problem.
 As we will see below, shortest paths between caterpillar trees in $\rnni$ differ from those in the classical $\nni$ space.
@@
 
-Recall \autocite{Gavryushkin2018-ol} that to compute a shortest path between two caterpillar trees in $\nni$ sometimes requires first building a clade and then move the clade around the tree (see Figure~\ref{fig:NNI_vs_RNNI}).
+Recall \autocite{Gavryushkin2018-ol} that computing a shortest path between two caterpillar trees in $\nni$ sometimes requires first building a clade and then moving the clade around the tree (see Figure~\ref{fig:NNI_vs_RNNI}).
 In $\rnni$ however, the necessity of additional rank moves invalidates this strategy.
-This is due to the fact that $\nni$ moves $\rnni$ are only allowed on edges of length one.
+This is due to the fact that $\nni$ moves in $\rnni$ are only allowed on edges of length one.
 
@@
 \vspace{12pt}
-\caption{Paths between caterpillar trees $T$ and $R$: Solid arrows indicate paths in $\rnni$, dashed arrow is a move only possible in $\nni$.
-Bottom path is a shortest $\rnni$ path.}
+\caption{Paths between caterpillar trees $T$ and $R$: Solid arrows indicate paths in $\rnni$, the dashed arrow is a move only possible in $\nni$.
+The bottom path is a shortest $\rnni$ path.}
 \label{fig:NNI_vs_RNNI}
@@
 
-In this section we discussed the so-called Split Theorem, which states that every split shared between two trees is maintained along shortest paths between the trees.
+In this section we discuss the so-called Split Theorem, which states that every split shared between two trees is maintained along shortest paths between the trees.
 For example, Split Theorem is true in the $\spr$ space and false in $\nni$ \autocite{Li1996-zw}.
@@
 
-The following example in Figure~\ref{fig:splitthm_counterexample} shows that this form of the Split Theorem does now hold.
+The following example in Figure~\ref{fig:splitthm_counterexample} shows that this form of the Split Theorem does not hold.
 
@@
 
-Furthermore, all our methods in this paper for computing shortest paths in special subspaces of $\rnni$ demonstrate that that shared clusters are maintained along shortest paths.
+Furthermore, all our methods in this paper for computing shortest paths in special subspaces of $\rnni$ demonstrate that shared clusters are maintained along shortest paths.
 
@@
 We have computationally checked that this conjecture is true in $\rnni$ spaces with up to seven taxa ($n = 2, \ldots, 7$).
-%TODO
 Specifically, we computed subgraphs of $\rnni$ containing only trees with shared clusters and compared distances in this subgraph with distances in the whole $\rnni$ graph.
@@
 To the best of our knowledge no such algorithm exists for $\nni$.
-This implies that the algorithmic complexity of comping distances in $\nni$ is higher than in $\rnni$.
+This implies that the algorithmic complexity of computing distances in $\nni$ is higher than in $\rnni$.
 
@@
 
-We acknowledge support from the Royal Society of New Zealand through the Rutherford Discovery Fellowship awarded to AG.
-This work was partially supported by the Ministry of Business, Innovation, and Employment of New Zealand through the Endeavour Smart Ideas and Data Science Programmes grants.
+We acknowledge support from the Royal Society of New Zealand through the Rutherford Discovery Fellowship (RDF-UOO1702) awarded to AG.
+This work was partially supported by the Ministry of Business, Innovation, and Employment of New Zealand through the Endeavour Smart Ideas (CONT-61378-ENDSI-UOO) and Data Science Programmes grants.
+
+MF thanks the joint research project \textit{DIG-IT!} supported by the European Social Fund (ESF), reference: ESF/14-BM-A55-0017/19, and the Ministry of Education, Science, and Culture of Mecklenburg-Vorpommern, Germany, for funding parts of this work.
 

### 2019-12-19 rnni_convexity.tex
make it fit on 15 pages

@@
 
-The reason that this example cannot be transferred to $\rnni$ is that additional rank moves are necessary to merge and then split the two caterpillar subtrees.
+The reason this example cannot be transferred to $\rnni$ is that additional rank moves are necessary to merge and then split the two caterpillar subtrees.
 Similarly to the example in Figure~\ref{fig:NNI_vs_RNNI}, adding rank moves to the shortest $\nni$ path from $T$ to $R$ results in a path that is no longer a shortest $\rnni$ path.
@@
 
-The question of whether computing $\rnni$ distances is $\np$-hard is still an open problem.
+The question of whether computing $\rnni$ distances is $\np$-hard is still open.
 

### 2019-12-30 rnni_convexity.tex
Mareike came to Max Plank too!
add ranks fig 1
caption fig2: blue->dashed

@@
 \centering
-\includegraphics[width=0.25\textwidth]{ranked_tree}
+\includegraphics[width=0.4\textwidth]{ranked_tree}
 \vspace{12pt}
@@
 \caption{Two types of operation that define edges of the $\rnni$ graph.
-The rank move swaps the rank of the two highlighted nodes and the $\nni$ moves are performed on the blue edges.}
+The rank move swaps the rank of the two highlighted nodes and the $\nni$ moves are performed on the dashed edges.}
 \label{fig:RNNI}
@@
 
-Part of this work was done while LC and AG were visiting Max Plank Institute for Evolutionary Biology, we appreciate their support.
+Part of this work was done while AG, LC, and MF were visiting Max Plank Institute for Evolutionary Biology, we appreciate their support.
 

### 2019-12-30 rnni_convexity.tex
add leaf labels fig 4 (merge-sort-split)

@@
 \caption{Example of a shortest path between two trees $T$ and $R$ in $\nni$.
-In \autocite{Li1996-zw} it is proven that there is a labelling that ensures that there is no shortest path in $\nni$ where the clusters (dots vs.\ squares) shared between $T$ and $R$ are preserved.
+In \autocite{Li1996-zw} it is proven that there is a labelling that ensures that there is no shortest path in $\nni$ where the clusters $\{1, \ldots, m\}$ and $\{m+1, \ldots, n\}$ with $m = \frac{n}{2}$ shared between $T$ and $R$ are preserved.
 Instead, the depicted path where $\frac{n}{2}$ cherries are built and then sorted before being resolved is a shortest path}

### 2019-12-31 rnni_convexity.tex
delete notation for cherries

@@
 The path $p$ returned by $\csort$ is a shortest caterpillar path because every tree modification on $p$ reduces the number of inversions of taxa in $T$ and $R$.
-Indeed, no $\rnni$ move on caterpillar trees can reduce the number of taxa inversions by more than one, assuming that inversions with both taxa of a cherry are counted only once, that is, if both pairs of taxa $(x, z)$ and $(y, z)$ appear in opposite orders in $T$ and $R$ and $(x,y)$ is a cherry then they are counted as one inversion.
+Indeed, no $\rnni$ move on caterpillar trees can reduce the number of taxa inversions by more than one, assuming that inversions with both taxa of a cherry are counted only once, that is, if both pairs of taxa $(x, z)$ and $(y, z)$ appear in opposite orders in $T$ and $R$ and $x$ an $y$ build a cherry then they are counted as one inversion.
 Hence any caterpillar path every move along which reduced the number of inversions is a shortest caterpillar path.

### 2019-12-31 rnni_convexity.tex
rm trivial cluster from tree representations

@@
 A path between two caterpillar trees that only consists of caterpillar trees is called a \emph{caterpillar path}.
-We can identify a caterpillar tree $T = [\{x_1, x_2\}, \{x_1, x_2, x_3\}, \ldots, \{x_1, \ldots, x_n\}]$ with the list of its taxa $[x_1, x_2, \ldots, x_n]$, assuming that $x_1 < x_2$ as natural numbers (recall that the set of taxa is $\{1, \ldots, n\}$).
+We can identify a caterpillar tree $T = [\{x_1, x_2\}, \{x_1, x_2, x_3\}, \ldots, \{x_1, \ldots, x_{n-1}\}]$ with the list of its taxa $[x_1, x_2, \ldots, x_n]$, assuming that $x_1 < x_2$ as natural numbers (recall that the set of taxa is $\{1, \ldots, n\}$).
 

### 2020-01-03 rnni_convexity.tex
typo
some minor things throughout the whole paper

--- 2019-12-30 rnni_convexity.tex
@@
 This result suggests that the complexity of computing distances in $\rnni$ is different from that in the classical $\nni$ space.
-We will discuss this further in Section~\ref{section:cluster_theorem}, where we disprove a conjecture of \textcite{Gavryushkin2018-ol} and suggest its weaker versions a plausible alternatives.
+We will discuss this further in Section~\ref{section:cluster_theorem}, where we disprove a conjecture of \textcite{Gavryushkin2018-ol} and suggest weaker versions as plausible alternatives.
 We will finish the paper with a discussion and open problems.
@@
 While these algorithms are interesting on their own (e.g.\ in simulation studies), they will also form an important ingredient to study the geometry of the $\rnni$ space later in this paper.
-For example, $\mdtree$ will be an important tool for finding the radius of the $\rnni$ graph, as the algorithms computes a tree with maximum distance if the input tree is a caterpillar tree.
+For example, $\mdtree$ will be an important tool for finding the radius of the $\rnni$ graph, as the algorithm computes a tree with maximum distance if the input tree is a caterpillar tree.
 We will prove this fact in Section~\ref{section:diameter}.
@@
 
-Note that $\rnni$ space on trees with seven taxa is a challenging goal to achieve computationally.
-For example, \textcite{Whidden2016-kl} were able to achieve up to seven taxa in the much smaller $\rnni$ space, when it required considering pairs of trees.
+Note that computing the $\rnni$ space on trees with seven taxa is a challenging goal to achieve.
+For example, \textcite{Whidden2016-kl} were able to achieve up to seven taxa in the much smaller $\spr$ space, when it required considering pairs of trees.
 %TODO: Make a zenodo dio for the repository before final publication
@@
 This node is denoted by $\mrca_{T'}(C_k)$ (short for \emph{most recent common ancestor}).
-We then extend $p$ by adding the tree obtained from $T'$ by performing the $\rnni$ move that decreases the rank of $\mrca_{T'}(C_k)$.
+We then extend $p$ by adding the tree obtained from $T'$ by performing an $\rnni$ move that decreases the rank of $\mrca_{T'}(C_k)$.
 We continue extending $p$ in this way by repeatedly performing $\rnni$ moves until the rank of $\mrca_{T'}(C_k) = k$, thus completing step $k$.
@@
 Case $k = 1$.
-In this case $C_k$ is a consists of two taxa $\{x, y\}$.
+In this case $C_k$ consists of two taxa $\{x, y\}$.
 The node $v = \mrca_{T'}(x, y)$ has $\rank_{T'}(v) = r > 1$.
@@
 The only way to decrease the rank of $\mrca(x, y)$ then is to perform an $\nni$ move on that edge.
-It is not hard to see that out of two possible $\nni$ moves only one decreases the rank of $\mrca$.
+It is not hard to see that out of two possible $\nni$ moves only one decreases the rank of this $\mrca$.
 
@@
 A natural question then is how accurate this approximation is.
-We have computationally shown that the algorithm $\findpath$ finds the correct distance for trees up to seven taxa \autocite{Collienne2019}.
+We have computationally shown that the algorithm $\findpath$ finds the correct distance for trees up to six taxa \autocite{Collienne2019}.
 
@@
 A path between two caterpillar trees that only consists of caterpillar trees is called a \emph{caterpillar path}.
-We can identify a caterpillar tree $T = [\{x_1, x_2\}, \{x_1, x_2, x_3\}, \ldots, \{x_1, \ldots, x_n\}]$ with the list of its taxa $[x_1, x_2, \ldots, x_n]$, assuming that $x_1 < x_2$ as natural numbers (recall that the set of taxa is $\{1, \ldots, n\}$).
+We can identify a caterpillar tree $T = [\{x_1, x_2\}, \{x_1, x_2, x_3\}, \ldots, \{x_1, \ldots, x_{n-1}\}]$ with the list of its taxa $[x_1, x_2, \ldots, x_n]$, assuming that $x_1 < x_2$ (recall that the set of taxa is $\{1, \ldots, n\}$).
 
@@
 The path $p$ returned by $\csort$ is a shortest caterpillar path because every tree modification on $p$ reduces the number of inversions of taxa in $T$ and $R$.
-Indeed, no $\rnni$ move on caterpillar trees can reduce the number of taxa inversions by more than one, assuming that inversions with both taxa of a cherry are counted only once, that is, if both pairs of taxa $(x, z)$ and $(y, z)$ appear in opposite orders in $T$ and $R$ and $(x,y)$ is a cherry then they are counted as one inversion.
-Hence any caterpillar path every move along which reduced the number of inversions is a shortest caterpillar path.
+Indeed, no $\rnni$ move on caterpillar trees can reduce the number of taxa inversions by more than one, assuming that inversions with both taxa of a cherry are counted only once.
+That is, if both pairs of taxa $(x, z)$ and $(y, z)$ appear in opposite orders in $T$ and $R$ and $x$ an $y$ build a cherry then they are counted as one inversion.
+Hence any caterpillar path, every move along which reduced the number of inversions, is a shortest caterpillar path.
 
@@
 The idea behind this algorithm is to efficiently return a tree as far away from a given tree as possible.
-As we will see in Section~\ref{section:diameter}, $\mdtree$ achieves this goal exactly for caterpillar trees.
+As we will see in Section~\ref{section:diameter}, $\mdtree$ achieves this goal for caterpillar trees.
 
@@
 Initially, $R$ only consists of two taxa $l_1$ and $l_2$, the most recent common ancestor of which will eventually be of rank $n - 1$.
-The remaining taxa are iteratively added to $R$ in the reversed order of $L$ starting from $l_n$.
+The remaining taxa are iteratively added to $R$ reversing the order of $L$.
 At every iteration, the taxon is attached to a new internal node created on one of the existing branches in $R$ so that this new attachment node has rank one.
@@
 Next, we will consider the so-called Split Theorem \autocite{Gavryushkin2018-ol}, which states that every shared split of taxa is maintained along every shortest path.
-Recall that a \emph{split} is a bipartition of the set of taxa obtained by deleting an edge of the tree $T$, splits are denoted by $A|B$.
-We will give a counterexample to the Split Theorem in $\rnni$ and consider a variant of the Weak Split Theorem, which states that every shared split of taxa is maintained along a shortest path.
-We will also consider the so-called Clade Theorem as another weak version of the Split Theorem (see Conjecture~\ref{conjecture:cluster_theorem}).
+Recall that a \emph{split} is a bipartition of the set of taxa obtained by deleting an edge of the tree $T$.
+Splits are denoted by $A|B$.
+We will give a counterexample to the Split Theorem in $\rnni$ and consider a variant, the Weak Split Theorem, which states that every shared split of taxa is maintained along a shortest path.
+We will also consider the so-called Cluster Theorem as another weak version of the Split Theorem (see Conjecture~\ref{conjecture:cluster_theorem}).
 
@@
 
-Let $p$ be a shortest path from $T$ to $R$ and $p{\big|}_n$ the path resulting from deleting taxon $n+1$ from all trees on $p$.
+Let $p$ be a shortest path from $T$ to $R$ and $p{\big|}_n$ the path resulting from deleting taxon $n+1$ from all trees on $p$ and removing all identical trees.
 Then $p{\big|}_n$ is a path from $T{\big|}_n$ to $R{\big|}_n$.
@@
 In this section we restrict our attention to the set of caterpillar trees.
-These trees are of particular interest because they are used to prove that computing the $\nni$ distance is an NP-hard problem.
+These trees are of particular interest because they are used to prove that computing $\nni$ distances is an NP-hard problem.
 As we will see below, shortest paths between caterpillar trees in $\rnni$ differ from those in the classical $\nni$ space.
 We will show later in this section that $\rnni$ distances between caterpillar trees can be computed in polynomial time.
-Throughout this section we use the list representation of trees described in Section~\ref{section:alg_csort}.
+Throughout this section we use the list representation of caterpillar trees described in Section~\ref{section:alg_csort}.
 
@@
 These observations motivate the following general result.
-Every pair of trees $T, R$ in $\rnni$ are connected by a caterpillar path of length $d(T,R)$, that is, the set of caterpillar trees is convex in $\rnni$ (Theorem~\ref{thm:caterpillar_convex}).
+Every pair of caterpillar trees $T, R$ in $\rnni$ are connected by a caterpillar path of length $d(T,R)$, that is, the set of caterpillar trees is convex in $\rnni$ (Theorem~\ref{thm:caterpillar_convex}).
 Note that this is not true in the $\nni$ space (Figure~\ref{fig:NNI_vs_RNNI}).
 
-We need the following lemma.
-Recall that $\frac{(n-1)(n-2)}{2}$ is the upper bound of the diameter of $\rnni$ by Lemma~\ref{lemma:diameter_bound}.
+For proving the convexity of the set of caterpillar trees we need the following lemma.
+Recall that $\frac{(n-1)(n-2)}{2}$ is an upper bound of the diameter of $\rnni$ by Lemma~\ref{lemma:diameter_bound}.
 
@@
 
-For proving the other direction of the statement, assume that $d_c(T,R) = \frac{(n-1)(n-2)}{2}$.
+To prove the reverse, assume that $d_c(T,R) = \frac{(n-1)(n-2)}{2}$.
 We prove that $d(T,R) = \frac{(n-1)(n-2)}{2}$ by induction on the number of taxa $n$.
-The induction basis for $n=3$ taxa is trivial as the $\rnni$ graph is a triangle on three caterpillar trees in that case.
+The induction basis for $n=3$ taxa is trivial as the $\rnni$ graph is a triangle on three caterpillar trees in this case.
 
@@
 Since this taxon is adjacent to the root in $T$ and part of the cherry in $R$, $\csort$ requires $n-1$ such moves.
-Hence, $d_c(T{\big|}_n, R{\big|}_n) = \frac{n(n-1)}{2} - (n-1)$ = $\frac{(n-1)(n-2)}{2} = d(T{\big|}_n,R{\big|}_n)$, by induction.
+Hence, $d_c(T{\big|}_n, R{\big|}_n) = \frac{n(n-1)}{2} - (n-1)$ = $\frac{(n-1)(n-2)}{2} = d(T{\big|}_n,R{\big|}_n)$, by the induction hypothesis.
 Lemma~\ref{lemma:distance_delete_taxon} implies that $d(T{\big|}_n, R{\big|}_n) \leq d(T,R) - |\rank_T(\parent_T(n+1)) - \rank_R(\parent_R(n+1))| = d(T,R) - (n-1)$.
@@
 \proof
-Note that it is suffice to prove that $d_c(T, R) = d(T, R)$ for an arbitrary pair of caterpillar trees $T$ and $R$.
-Lemma~\ref{lemma:caterpillar_dist=diameter} implies that if $T$ and $R$ are at the maximal possible distance, that is, $d_c(T, R) = \frac{(n-1)(n-2)}{2}$.
+Note that it suffices to prove that $d_c(T, R) = d(T, R)$ for an arbitrary pair of caterpillar trees $T$ and $R$.
+Lemma~\ref{lemma:caterpillar_dist=diameter} implies that if $T$ and $R$ are at the maximal possible distance, it is $d_c(T, R) = \frac{(n-1)(n-2)}{2}$.
 Denote the latter number by $D$.
@@
 Note that $q$ is a path from $\hat T$ to $R$ which is shorter than $\csort(\hat T, R)$.
-Indeed, the two paths coincide between $\hat T$ and $T$ and $q$ is shorter between $T$ and $R$.
-Since $d_c(\hat T, R) = D$, the existence of $q$ is a contradiction to Lemma~\ref{lemma:caterpillar_dist=diameter}.
+Indeed, the two paths coincide between $\hat T$ and $T$, and $q$ is shorter between $T$ and $R$.
+Since $d_c(\hat T, R) = D$, the existence of $q$ is a contradiction of Lemma~\ref{lemma:caterpillar_dist=diameter}.
 \endproof
@@
 
-\textcite[Theorem~7]{Gavryushkin2018-ol} gave an upper bound for the diameter of $\rnni$ space, which we recall is denoted by $\Delta(\rnni)$.
+\textcite[Theorem~7]{Gavryushkin2018-ol} gave an upper bound for the diameter of $\rnni$ space, denoted by $\Delta(\rnni)$.
 They showed that $\Delta(\rnni) = n^2 - 3n - 5/8$.
@@
 \proof
-Since $d_c([1\ldots,n], [n-1,n,n-2,n-3, \ldots, 1]) = \frac{(n-1)(n-2)}{2}$, the claim follows directly from from Theorem~\ref{thm:caterpillar_convex} and Lemma~\ref{lemma:diameter_bound}.
+Since $d_c([1\ldots,n], [n-1,n,n-2,n-3, \ldots, 1]) = \frac{(n-1)(n-2)}{2}$, the claim follows directly from Theorem~\ref{thm:caterpillar_convex} and Lemma~\ref{lemma:diameter_bound}.
 \endproof
@@
 Note that $R\big|_n = \mdtree (T\big|_n)$ for an appropriate choice of attachment points in the execution of $\mdtree$.
-By induction hypothesis and Theorem~\ref{thm:diameter}, $d(T\big|_n,R\big|_n) = \frac{(n-1)(n-2)}{2}$.
-So inequality~(\ref{eqn:max_dist_cat_proof}) becomes $d(T,R) \geq \frac{(n-1)(n-2)}{2} + n - 1 = \frac{n(n-1)}{2}$, which with Theorem~\ref{thm:diameter} gives the desired equality.
+By the induction hypothesis and Theorem~\ref{thm:diameter}, $d(T\big|_n,R\big|_n) = \frac{(n-1)(n-2)}{2}$.
+So inequality~(\ref{eqn:max_dist_cat_proof}) becomes $d(T,R) \geq \frac{(n-1)(n-2)}{2} + n - 1 = \frac{n(n-1)}{2}$, which gives the desired equality with Theorem~\ref{thm:diameter}.
 \end{proof}
 
-The non-determinism of $\mdtree$ can be exploited to investigate what trees are at the maximal possible from a given caterpillar tree.
-In fact, those trees can have an arbitrary (non-ranked) topology!
+The non-determinism of $\mdtree$ can be exploited to investigate which trees are at the maximal possible distance from a given caterpillar tree.
+In fact, those trees can have an arbitrary (non-ranked) topology.
 This is because in each step of the algorithm the next taxon can be added on any edge incident to a leaf in the running tree.
-Therefore, if the input tree is a caterpillar tree, $\mdtree$ can be used to find a tree of pre-defined topology that is at the maximal possible distance from the input tree.
+Therefore, $\mdtree$ can be used to find a tree of pre-defined topology that is at the maximal possible distance from the input tree, in particular if the input is a caterpillar tree.
 Because the distance between trees is invariant under permutations of taxa labels, we conclude that for every tree $R$ on $n$ taxa there exists a caterpillar tree $T$ such that $d(T,R) = \Delta(\rnni) = \frac{(n-1)(n-2)}{2}$.
@@
 The reason this example cannot be transferred to $\rnni$ is that additional rank moves are necessary to merge and then split the two caterpillar subtrees.
-Similarly to the example in Figure~\ref{fig:NNI_vs_RNNI}, adding rank moves to the shortest $\nni$ path from $T$ to $R$ results in a path that is no longer a shortest $\rnni$ path.
+Similarly to the example in Figure~\ref{fig:NNI_vs_RNNI}, adding rank moves to the shortest $\nni$ path from $T$ to $R$ results in a path that is not a shortest $\rnni$ path.
 Sorting the two caterpillar subtrees of $T$ and $R$ independently results in a shorter path in $\rnni$.
@@
 
-The following example in Figure~\ref{fig:splitthm_counterexample} shows that this form of the Split Theorem does not hold.
+The example in Figure~\ref{fig:splitthm_counterexample} shows that this form of the Split Theorem does not hold.
 
@@
 Hence not every shortest path between trees in $\rnni$ has to maintain every shared split of taxa.
-However, \textbf{there exists} still another shortest path connecting the two trees, along which every shared split is maintained.
+However, there exists yet another shortest path connecting the two trees, along which every shared split is maintained.
 This motivates the study of the following Weak Split Theorem in $\rnni$.
@@
 
-Another interesting property of our counterexample in Figure~\ref{fig:splitthm_counterexample} is that the set of taxa $\{1, 2, 3\}$, which is a part of the split in both $T$ and $R$, does not form a \textbf{cluster} in $R$, that is, there is no single node in the tree all descended taxa of which are exactly $1$, $2$, and $3$.
+Another interesting property of our counterexample in Figure~\ref{fig:splitthm_counterexample} is that the set of taxa $\{1, 2, 3\}$, which is a part of the split in both $T$ and $R$, does not form a cluster in $R$, that is, there is no single node in the tree all descended taxa of which are exactly $1$, $2$, and $3$.
 Importantly, rooted phylogenetic trees can be uniquely represented by sets of clusters \autocite{Steel2016-ye}, but not by sets of splits as splits cannot define the position of the root.

### 2020-01-12 rnni_convexity.tex
typo
add Mike's suggestions

@@
 
-\newtheorem{definition}{Definition}
-\newtheorem{theorem}[definition]{Theorem}
-\newtheorem{conjecture}[definition]{Conjecture}
-\newtheorem{lemma}[definition]{Lemma}
-\newtheorem{proposition}[definition]{Proposition}
+\newtheorem*{definition}{Definition}
+\newtheorem{theorem}{Theorem}
+\newtheorem{conjecture}[theorem]{Conjecture}
+\newtheorem{lemma}[theorem]{Lemma}
+\newtheorem{proposition}[theorem]{Proposition}
 
@@
 
-\title[Ranked Nearest Neighbour Intarchange]{Geometry of Ranked Nearest Neighbour Interchange Space of Phylogenetic Trees}
+\title[Ranked Nearest Neighbour Interchange]{Geometry of Ranked Nearest Neighbour Interchange Space of Phylogenetic Trees}
 \date{\today}
@@
 Suppress $C_i$ and $C_j$ in both $T'$ and $R$ to new taxa $c_i$ and $c_j$ respectively and proceed as in Case $k = 1$.
-\item $C_k = C_i \cup {x}$, where $x$ is a taxon not present in $C_1, \ldots, C_k$.
+\item $C_k = C_i \cup \{x\}$, where $x$ is a taxon not present in $C_1, \ldots, C_k$.
 Suppress $C_i$ in both $T'$ and $R$ to a new taxon $c_i$ and proceed as in Case $k = 1$.
@@
 Let $T$ and $R$ be two trees on taxa $\{1, \ldots, n+1\}$.
-Then $d(T{\big|}_n, R{\big|}_n) \leq d(T,R) - \delta$, where $\delta = |\rank_T(\parent_T(n+1)) - \rank_R(\parent_R(n+1))|$.
+Then ${d(T{\big|}_n, R{\big|}_n) \leq d(T,R) - \delta}$, where $\delta = |\rank_T(\parent_T(n+1)) - \rank_R(\parent_R(n+1))|$.
 \label{lemma:distance_delete_taxon}
@@
 \textcite[Theorem~7]{Gavryushkin2018-ol} gave an upper bound for the diameter of $\rnni$ space, denoted by $\Delta(\rnni)$.
-They showed that $\Delta(\rnni) = n^2 - 3n - 5/8$.
+They showed that $\Delta(\rnni) \leq n^2 - 3n - 5/8$.
 In this section we improve that result and calculate the exact diameter and radius of the $\rnni$ space.
@@
 
-\begin{conjecture}[Weak Split Theorem]
-If a split of taxa $A | B$ is shared by two trees $T$ and $R$ in $\rnni$ then there exists a shortest path $p$ between $T$ and $R$ such that all trees on $p$ share the split $A | B$.
-\label{conjecture:split_theorem}
+\begin{definition}[Weak Split Theorem]
+If a split of taxa $A | B$ is shared by two trees $T$ and $R$ then there exists a shortest path $p$ between $T$ and $R$ such that all trees on $p$ share the split $A | B$.
+\end{definition}
+
+\begin{conjecture}
+\label{conjecture:weak_split_theorem}
+The Weak Split Theorem holds for $\rnni$.
 \end{conjecture}
@@
 
-\begin{conjecture}[Cluster Theorem]
-Let $T$ and $R$ be trees in $\rnni$ that share a cluster $C$.
+\begin{definition}[Cluster Theorem]
+Let $T$ and $R$ be trees that share a cluster $C$.
 Then $C$ is present as a cluster in every tree on every shortest path between $T$ and $R$.
+\end{definition}
+
+\begin{conjecture}
 \label{conjecture:cluster_theorem}
+The Cluster Theorem holds for $\rnni$.
 \end{conjecture}

### 2020-01-12 rnni_convexity.tex
thank Mike
edit def convexity

@@
 We will employ the algorithms developed in the previous section to aid our analysis of shortest paths.
-Those algorithms will, for example, enable us to prove that every pair of caterpillar trees $T$ and $R$ is connected by a caterpillar path of length $d(T,R)$ (Theorem~\ref{thm:caterpillar_convex} in Section~\ref{section:caterpillar_convex}).
-In other words, the set of caterpillar trees is \emph{convex} in $\rnni$.
+Those algorithms will, for example, enable us to prove that the set of caterpillar trees is convex in $\rnni$ (Theorem~\ref{thm:caterpillar_convex} in Section~\ref{section:caterpillar_convex}).
+We call a set of trees \emph{convex} if every pair of trees from this set is connected by a shortest path that stays within this set.
 In Section~\ref{section:diameter} we will investigate the diameter and radius of the $\rnni$ space.
@@
 \section*{Acknowledgements}
-We thank Charles Semple for useful discussions about the Cluster Theorem.
+We thank Charles Semple for useful discussions about the Cluster Theorem, and Mike Steel for his useful comments that improved this paper.
 

### 2020-01-12 rnni_convexity.tex
mv def convexity to introduction

@@
 We use the notation $d(T, R)$ throughout the paper to denote the graph distance between trees in $\rnni$, that is, $d(T, R)$ is the length of a shortest $\rnni$ path between tree $T$ and $R$.
+When analysing the geometry of $\rnni$, we will consider a subset of trees and see that it is \emph{convex} in $\rnni$, which means that every pair of trees from this subset is connected by a shortest path that stays within this subset.
 
@@
 Those algorithms will, for example, enable us to prove that the set of caterpillar trees is convex in $\rnni$ (Theorem~\ref{thm:caterpillar_convex} in Section~\ref{section:caterpillar_convex}).
-We call a set of trees \emph{convex} if every pair of trees from this set is connected by a shortest path that stays within this set.
 In Section~\ref{section:diameter} we will investigate the diameter and radius of the $\rnni$ space.

### 2020-01-12 rnni_convexity.tex
rewording of the convexity dfn

--- 2020-01-12 rnni_convexity.tex
@@
 We use the notation $d(T, R)$ throughout the paper to denote the graph distance between trees in $\rnni$, that is, $d(T, R)$ is the length of a shortest $\rnni$ path between tree $T$ and $R$.
+An important geometric property analysed in this paper is whether trees from a certain class are connected but shortest paths that do not leave the class.
+We say that a subset $S$ of trees is \emph{convex} if every pair of trees from $S$ is connected by a shortest path that completely stays within $S$.
 

### 2020-01-22 rnni_convexity.tex
finished checking findpath up to seven taxa
add David's suggestions

@@
 In this paper we study the graph of ranked phylogenetic trees where the adjacency relation is given by a local rearrangement of the tree structure.
-Our work is motivated by tree inference algorithms, such as maximum likelihood and Markov Chain Monte Carlo methods, where the geometry of the search space plays a central role for efficiency and practicality of the optimisation.
+Our work is motivated by tree inference algorithms, such as maximum likelihood and Markov Chain Monte Carlo methods, where the geometry of the search space plays a central role for efficiency and practicality of optimisation.
 We hence focus on understanding the geometry of the space (graph) of ranked trees, the so-called ranked nearest neighbour interchange ($\rnni$) graph.
-For example, we find the radius and diameter of the space exactly, improving the best previously known estimates.
+We find the radius and diameter of the space exactly, improving the best previously known estimates.
 Since the $\rnni$ graph is a generalisation of the classical nearest neighbour interchange ($\nni$) graph to ranked phylogenetic trees, we compare geometric and algorithmic properties of the two graphs.
 Surprisingly, we discover that both geometric and algorithmic properties of $\rnni$ and $\nni$ are quite different.
-For example, we establish convexity of certain natural subspaces in $\rnni$, which are not convex is $\nni$.
-Furthermore, our results suggest that the complexity of computing distances in the two graphs is different.
+For example, we establish convexity of certain natural subspaces in $\rnni$ which are not convex is $\nni$.
+Our results suggest that the complexity of computing distances in the two graphs is different.
 \end{abstract}
@@
 
-The nearest neighbour interchange ($\nni$) graph, defined on the set of phylogenetic trees with adjacency relation given by the interchange operation of two sister clades (subtrees), has been known in mathematical biology literature for nearly 50 years \autocite{Robinson1971-ql,Moore1973-kk}.
-Considered with the metric given by the length of a shortest path (graph-distance), this graph becomes a metric space.
-Its geometry has been extensively studied \autocite{Dasgupta2000-xa, Li1996-zw, Gordon2013-fw, De_Jong2016-al}.
-An important property of the $\nni$ graph is that computing distances is NP-hard \autocite{Dasgupta2000-xa}.
-As a result no algorithm exists to compute the distance in practical time.
-A consequence is that tree search and sampling algorithms pose a significant challenge even for moderately sized trees
+\section{Introduction}
+
+The nearest neighbour interchange ($\nni$) graph is defined on the set of phylogenetic trees with adjacency relation given by the interchange operation of two sister clades (subtrees).
+It has been known in mathematical biology literature for nearly 50 years \autocite{Robinson1971-ql,Moore1973-kk}.
+Its metric geometry has been extensively studied \autocite{Dasgupta2000-xa, Li1996-zw, Gordon2013-fw, De_Jong2016-al}.
+An important property of the $\nni$ graph is that computing distances is NP-hard \autocite{Dasgupta2000-xa}, a fact that goes partway towards explaining why tree search and sampling algorithms pose a significant challenge even for moderately sized trees
 \autocite{Whidden2016-kl}.
 
-More recent advances in computational phylogenetics introduce various classes of molecular clock models \autocite{Yoder2000-ks,Drummond2006-nl,Drummond2010-yf} and made computational inference of phylogenetic time-trees possible \autocite{Ronquist2003-eq, Bouckaert2018-yr, Hadfield2018-xp}.
-However, mathematical challenges that come with this seemingly inessential change in parametrisation (genomic distance VS time distance) of trees have only recently been brought to attention \autocite{Gavryushkin2016-uu}.
+Recent advances in computational phylogenetics have introduced various classes of molecular clock models \autocite{Yoder2000-ks,Drummond2006-nl,Drummond2010-yf} and made computational inference of phylogenetic time-trees possible \autocite{Ronquist2003-eq, Bouckaert2018-yr, Hadfield2018-xp}.
+However, the mathematical challenges resulting from this seemingly minor change in parametrisation (genomic distance VS time distance) of trees have only recently been brought to attention \autocite{Gavryushkin2016-uu}.
 These differences motivated \textcite{Gavryushkin2018-ol} to propose an extension of the $\nni$ graph to the class of discrete time-trees.
 The simplest such extension introduces the $\rnni$ graph on the set of ranked phylogenetic trees.
+Two rooted trees are considered adjacent if they differ by a single neighbour interchange or order change respecting that ranking. \todo{Can't read David's suggestion here.}
 Considered with the graph-distance $\rnni$ becomes a metric space and inherits the geometric and algorithmic challenges that the $\nni$ space has been traditionally facing.
@@
 
-In this paper, we consider the $\rnni$ space on ranked phylogenetic trees with all taxa being of equal rank.
+In this paper, we consider the $\rnni$ space on ranked phylogenetic trees which have all taxa being of equal rank zero.
 An example of such a tree is depicted in Figure~\ref{fig:ranked_tree}.
 In the terminology of \autocite{Gavryushkin2018-ol}, the space considered in this paper is the space of ranked ultrametric phylogenetic trees ($\rnniu$ in their notations).
-We postpone accurate definitions until later in this section.
+We postpone formal definitions until later in this section.
 
@@
 
-In line with the research programme proposed in \autocite{Gavryushkin2018-ol}, we investigate the geometry and algorithmic complexity of the $\rnni$ space.
+We investigate the geometry and algorithmic complexity of the $\rnni$ space.
 Specifically, in this paper we establish the exact radius and diameter of the space (Section~\ref{section:diameter}).
 We also show that the subset of caterpillar trees is convex in $\rnni$ (Section~\ref{section:caterpillar_convex}), thus settling one of the open problems proposed in \autocite{Gavryushkin2018-ol}.
-For establishing these geometric properties we are using algorithms that will be introduced in Section~\ref{section:algorithms}.
+For establishing these geometric properties we use algorithms that will be introduced in Section~\ref{section:algorithms}.
 We will in particular provide an approximate algorithm that computes exact distances for small trees.
@@
 The latter means that if $u$ and $v$ are two internal nodes of $T$ such that there exists a path from a taxon $x \in X$ to the root which first passes through $u$ and then through $v$ then $\rank(u) < \rank(v)$.
+This definition also implies that the rank of every internal node is distinct and every $i$ with $1 \leq i \leq n-1$ equals the rank of some node.
 Ranked trees $(T_1, \rank_1)$ and $(T_2, \rank_2)$ are different if trees $T_1$ and $T_2$ are different or $T_1 = T_2$ and $\rank_1 \neq \rank_2$.
@@
 
-Our definition of a ranked tree implies a natural notion of edge length -- we call $|\rank_T(v) - \rank_T(u)|$ the length of an edge $(u, v)$.
+Our definition of a ranked tree implies a natural notion of edge length -- we call the difference in rank $|\rank_T(v) - \rank_T(u)|$ the length of an edge $(u, v)$.
 We assume that edges of trees are undirected, so $(u, v) = (v, u)$.
@@
 The vertex set of the $\rnni$ graph is the set of all ranked trees on $n$ taxa.
-We introduce two types of operation ($\rnni$ \emph{moves}) on trees (see Figure~\ref{fig:RNNI}) and say that two trees are adjacent in the $\rnni$ graph if they are connected by an operation of either type.
+We introduce two types of operation ($\rnni$ \emph{moves}) on trees (see Figure~\ref{fig:RNNI}) and say that two trees are adjacent in the $\rnni$ graph if one can be obtained from the other by an operation of either type.
 The first type of operation is called a \emph{rank move} and defined by swapping the ranks of two internal nodes that are not adjacent in the tree.
@@
 
-We use the notation $d(T, R)$ throughout the paper to denote the graph distance between trees in $\rnni$, that is, $d(T, R)$ is the length of a shortest $\rnni$ path between tree $T$ and $R$.
+We use the notation $d(T, R)$ throughout the paper to denote the length of a shortest $\rnni$ path between tree $T$ and $R$.
 An important geometric property analysed in this paper is whether trees from a certain class are connected but shortest paths that do not leave the class.
@@
 $\findpath$ computes an $\rnni$ path between trees.
-$\csort$ computes a shortest $\rnni$ path between caterpillar trees (also known as ladder trees, see later in this section for accurate definitions).
-$\mdtree$ maximises the dissimilarity to find a remote tree given an arbitrary $\rnni$ tree.
+$\csort$ computes a shortest $\rnni$ path between caterpillar trees (also known as ladder trees, see later in this section for formal definitions).
+$\mdtree$ maximises the dissimilarity to find a remote tree from an arbitrary $\rnni$ tree.
 While these algorithms are interesting on their own (e.g.\ in simulation studies), they will also form an important ingredient to study the geometry of the $\rnni$ space later in this paper.
@@
 Note that computing the $\rnni$ space on trees with seven taxa is a challenging goal to achieve.
-For example, \textcite{Whidden2016-kl} were able to achieve up to seven taxa in the much smaller $\spr$ space, when it required considering pairs of trees.
+For example, \textcite{Whidden2016-kl} were able to compute the much smaller $\spr$ space on up to seven taxa and used this to compute curvature values for all pairs of trees within that space.
 %TODO: Make a zenodo dio for the repository before final publication
@@
 We then say that $v$ \emph{induces} $C$.
-All trees on $n$ leaves share trivial clusters, which are those induced by leaves and the root, so we exclude them from consideration and say cluster to mean a non-trivial cluster.
+All trees on $n$ leaves share trivial clusters, which are those induced by leaves and the root, so we exclude them from consideration and assume cluster means a non-trivial cluster.
 A tree is unambiguously defined by the list of its clusters sorted according to the rank of the inducing nodes.
@@
     \WHILE {$\rank_{T'}(\mrca_{T'}(C_k))>k$}
-    \STATE Update $T'$: Decrease $\rank_{T'}(\mrca_{T'}(C_k))$ by an $\rnni$ move
+        \IF {node $u$ with $\rank_{T'}(u) = rank_{T'}(\mrca_{T'}(C_k))-1$ is adjacent to $\mrca_{T'}(C_k)$}
+            \STATE $T'$ is $T'$ with $\rank_{T'}(\mrca_{T'}(C_k))$ decreased by an $\nni$ move
+        \ELSE
+            \STATE $T''$ is $T'$ with ranks of $u$ and $\mrca_{T'}(C_k)$ swapped
+        \ENDIF
         \label{alg:findpath:line:move_set_down}
+        \STATE $T' = T''$
         \STATE $p = p+T'$
@@
 A natural question then is how accurate this approximation is.
-We have computationally shown that the algorithm $\findpath$ finds the correct distance for trees up to six taxa \autocite{Collienne2019}.
+We have computationally shown that the algorithm $\findpath$ finds the correct distance for trees up to seven taxa \autocite{Collienne2019}.
 
@@
     \FOR {$i=1, \ldots, k$}
-        \IF {$T'[i] = x_k$}
-            \STATE Update $T'$: swap $T'[i]$ and $T'[i+1]$
+        \IF {taxon $T'[i]$ at position $i$ in $T'$ equals $x_k$}
+            \STATE $T''$ is $T'$ with $T'[i]$ and $T'[i+1]$ swapped
+            \STATE $T' = T''$
             \STATE $p = p + T'$
@@
 
-The path $p$ returned by $\csort$ is a shortest caterpillar path because every tree modification on $p$ reduces the number of inversions of taxa in $T$ and $R$.
+The path $p$ returned by $\csort$ is a shortest caterpillar path because every tree modification on $p$ reduces the number of inversions of taxa in $T$ and $R$, and no modification can reduce more than one insertion.
 Indeed, no $\rnni$ move on caterpillar trees can reduce the number of taxa inversions by more than one, assuming that inversions with both taxa of a cherry are counted only once.
@@
 We use the notion $\parent_T(x)$ to refer to the node adjacent to taxon $x$ in tree $T$.
-$T{\big|}_n$ denotes the restriction of tree $T$ to the set of taxa $\{1, \ldots, n\}$.
-In particular if $T$ is a tree on taxa $\{1, \ldots, n+1\}$ the tree $T{\big|}_n$ is obtained by deleting taxon $n+1$ and suppressing the thereby created node of degree two.
+Let $T{\big|}_n$ denote the restriction of tree $T$ to the set of taxa $\{1, \ldots, n\}$.
+In other words, if $T$ is a tree on taxa $\{1, \ldots, n+1\}$ the tree $T{\big|}_n$ is obtained by deleting taxon $n+1$ and suppressing the thereby created node of degree two, updating node rankings accordingly.
 
 \begin{lemma}
-Let $T$ and $R$ be two trees on taxa $\{1, \ldots, n+1\}$.
+Let $T$ and $R$ be two trees on taxa $\{1, \ldots, {n+1}\}$.
 Then ${d(T{\big|}_n, R{\big|}_n) \leq d(T,R) - \delta}$, where $\delta = |\rank_T(\parent_T(n+1)) - \rank_R(\parent_R(n+1))|$.
@@
 These trees are of particular interest because they are used to prove that computing $\nni$ distances is an NP-hard problem.
-As we will see below, shortest paths between caterpillar trees in $\rnni$ differ from those in the classical $\nni$ space.
+As we will see below, shortest paths between caterpillar trees in the $\rnni$ space differ from those in the classical $\nni$ space.
 We will show later in this section that $\rnni$ distances between caterpillar trees can be computed in polynomial time.
@@
 
-Recall \autocite{Gavryushkin2018-ol} that computing a shortest path between two caterpillar trees in $\nni$ sometimes requires first building a clade and then moving the clade around the tree (see Figure~\ref{fig:NNI_vs_RNNI}).
-In $\rnni$ however, the necessity of additional rank moves invalidates this strategy.
+\textcite{Gavryushkin2018-ol} showed that computing a shortest path between two caterpillar trees in the $\nni$ graph sometimes requires first building a clade and then moving the clade around the tree (see Figure~\ref{fig:NNI_vs_RNNI}).
+In the $\rnni$ graph however, the necessity of additional rank moves invalidates this strategy.
 This is due to the fact that $\nni$ moves in $\rnni$ are only allowed on edges of length one.
@@
 These observations motivate the following general result.
-Every pair of caterpillar trees $T, R$ in $\rnni$ are connected by a caterpillar path of length $d(T,R)$, that is, the set of caterpillar trees is convex in $\rnni$ (Theorem~\ref{thm:caterpillar_convex}).
+We will show that every pair of caterpillar trees $T, R$ in $\rnni$ are connected by a caterpillar path of length $d(T,R)$, that is, the set of caterpillar trees is convex in $\rnni$ (Theorem~\ref{thm:caterpillar_convex}).
 Note that this is not true in the $\nni$ space (Figure~\ref{fig:NNI_vs_RNNI}).
@@
 \begin{proof}
-First note that $d(T,R) = \frac{(n-1)(n-2)}{2}$ implies $d_c(T,R) = \frac{(n-1)(n-2)}{2}$.
-Indeed, in each step $i=0, \ldots, n-3$ of $\csort$ at most $n-2-i$ pairs of taxa swap positions, as in the worst case the taxon considered in step $k$ is moved from the cherry of the tree up to the position $n-k$.
+First note that $d(T,R) = \frac{(n-1)(n-2)}{2}$ implies $d_c(T,R) \geq \frac{(n-1)(n-2)}{2}$.
+In each step $i=0, \ldots, n-3$ of $\csort$ at most $n-2-i$ pairs of taxa swap positions, as in the worst case the taxon considered in step $k$ is moved from the cherry of the tree up to the position $n-k$.
 It follows that the maximum length of a path computed by $\csort$ is
 \begin{equation}
-\sum\limits_{i=0}^{n-3} n-2-i = \frac{(n-1)(n-2)}{2}.
+\sum\limits_{i=0}^{n-3} (n-2-i) = \frac{(n-1)(n-2)}{2},
 \label{csort_max}
@@
 
-To prove the reverse, assume that $d_c(T,R) = \frac{(n-1)(n-2)}{2}$.
+so $d_c(T,R) = \frac{(n-1)(n-2)}{2}$
+
+To prove the converse, assume that $d_c(T,R) = \frac{(n-1)(n-2)}{2}$.
 We prove that $d(T,R) = \frac{(n-1)(n-2)}{2}$ by induction on the number of taxa $n$.

### 2020-01-26 rnni_convexity.tex
split thm becomes split property + minor

@@
 The simplest such extension introduces the $\rnni$ graph on the set of ranked phylogenetic trees.
-Two rooted trees are considered adjacent if they differ by a single neighbour interchange or order change respecting that ranking. \todo{Can't read David's suggestion here.}
+Two rooted trees are considered adjacent if they differ by a single neighbour interchange or order change respecting that ranking (detailed description below).
 Considered with the graph-distance $\rnni$ becomes a metric space and inherits the geometric and algorithmic challenges that the $\nni$ space has been traditionally facing.
@@
 $\csort$ computes a shortest $\rnni$ path between caterpillar trees (also known as ladder trees, see later in this section for formal definitions).
-$\mdtree$ maximises the dissimilarity to find a remote tree from an arbitrary $\rnni$ tree.
+$\mdtree$ attempts to maximise the dissimilarity to find a remote tree from an arbitrary $\rnni$ tree.
 While these algorithms are interesting on their own (e.g.\ in simulation studies), they will also form an important ingredient to study the geometry of the $\rnni$ space later in this paper.
@@
 Interestingly, the \emph{diameter} of $\rnni$, that is the maximum distance $\Delta(\rnni) = \max \{d(T, R) \mid T, R \in \rnni\}$ between any two trees in the graph, equals its \emph{radius} defined as $\min\limits_T \max\limits_R d(T,R)$.
-Next, we will consider the so-called Split Theorem \autocite{Gavryushkin2018-ol}, which states that every shared split of taxa is maintained along every shortest path.
+Next, we will consider the so-called Split Property \autocite{Gavryushkin2018-ol}, that is, every shared split of taxa is maintained along every shortest path.
 Recall that a \emph{split} is a bipartition of the set of taxa obtained by deleting an edge of the tree $T$.
 Splits are denoted by $A|B$.
-We will give a counterexample to the Split Theorem in $\rnni$ and consider a variant, the Weak Split Theorem, which states that every shared split of taxa is maintained along a shortest path.
-We will also consider the so-called Cluster Theorem as another weak version of the Split Theorem (see Conjecture~\ref{conjecture:cluster_theorem}).
+We will give a counterexample to show that $\rnni$ does not have the Split Property and consider a variant, the Weak Split Property, which says that every shared split of taxa is maintained along a shortest path.
+We will also consider the so-called Cluster Property as another weak version of the Split Property (see Conjecture~\ref{conjecture:cluster_theorem}).
 
@@
 
-\subsection{Cluster Theorem}
+\subsection{Cluster Property}
 \label{section:cluster_theorem}
 
-In this section we discuss the so-called Split Theorem, which states that every split shared between two trees is maintained along shortest paths between the trees.
-For example, Split Theorem is true in the $\spr$ space and false in $\nni$ \autocite{Li1996-zw}.
+In this section we discuss the so-called Split Property, which states that every split shared between two trees is maintained along shortest paths between the trees.
+For example, the $\spr$ space has the Split Property while $\nni$ does not \autocite{Li1996-zw}.
 Specifically, in $\nni$ there exist two trees $T$ and $R$, which share a split $A | B$, such that every tree on every shortest path between $T$ and $R$ does not have $A | B$.
@@
 
-The example in Figure~\ref{fig:splitthm_counterexample} shows that this form of the Split Theorem does not hold.
+The example in Figure~\ref{fig:splitthm_counterexample} shows that this form of the Split Property does not hold.
 
@@
 However, there exists yet another shortest path connecting the two trees, along which every shared split is maintained.
-This motivates the study of the following Weak Split Theorem in $\rnni$.
+This motivates the study of the following Weak Split Property in $\rnni$.
 
-\begin{definition}[Weak Split Theorem]
+\begin{definition}[Weak Split Property]
 If a split of taxa $A | B$ is shared by two trees $T$ and $R$ then there exists a shortest path $p$ between $T$ and $R$ such that all trees on $p$ share the split $A | B$.
@@
 \label{conjecture:weak_split_theorem}
-The Weak Split Theorem holds for $\rnni$.
+The Weak Split Property holds for $\rnni$.
 \end{conjecture}
@@
 
-These considerations motivate us to conjecture the following strong version of the Cluster Theorem.
+These considerations motivate us to conjecture the following strong version of the Cluster Property.
 
-\begin{definition}[Cluster Theorem]
+\begin{definition}[Cluster Property]
 Let $T$ and $R$ be trees that share a cluster $C$.
@@
 \label{conjecture:cluster_theorem}
-The Cluster Theorem holds for $\rnni$.
+The Cluster Property holds for $\rnni$.
 \end{conjecture}
@@
 We have taken the route of comparing the classic $\nni$ with the $\rnni$ graph, so have mostly been concentrating on questions that are settled in $\nni$.
-Specifically, we have considered a number of geometric properties, including the diameter, radius, convexity of caterpillar trees, and the Split Theorem.
+Specifically, we have considered a number of geometric properties, including the diameter, radius, convexity of caterpillar trees, and the Split Property.
 We have also justified that the technique of \textcite{Dasgupta2000-xa} to prove that distances in $\nni$ are $\np$-hard to compute cannot be directly applied in $\rnni$.
-This led us to the Cluster Theorem, a statement that does not hold in $\nni$.
-All our algorithms and methods developed in this paper witness in favour of the Cluster Theorem being true in $\rnni$.
+This led us to the Cluster Property, a statement that does not hold in $\nni$.
+All our algorithms and methods developed in this paper witness in favour of the Cluster Property being true in $\rnni$.
 For example, we have checked the statement computationally for up to seven taxa \autocite{Collienne2019}.
-Although the Split Theorem is true in $\spr$, it is still $\np$-hard to compute distances in that graph.
-Hence Cluster Theorem alone cannot be used as an argument in favour of the distance problem having an efficient solution in $\rnni$.
+Although the Split Property is true in $\spr$, it is still $\np$-hard to compute distances in that graph.
+Hence Cluster Property alone cannot be used as an argument in favour of the distance problem having an efficient solution in $\rnni$.
 But the greedy nature of all our algorithms for computing shortest paths in $\rnni$ developed in this paper does provide such an argument.
@@
 \section*{Acknowledgements}
-We thank Charles Semple for useful discussions about the Cluster Theorem, and Mike Steel for his useful comments that improved this paper.
+We thank Charles Semple for useful discussions about the Cluster Property, and Mike Steel for his useful comments that improved this paper.
 

### 2020-01-26 rnni_convexity.tex
minor clarificants & wording tweaks

@@
 A natural question then is how accurate this approximation is.
-We have computationally shown that the algorithm $\findpath$ finds the correct distance for trees up to seven taxa \autocite{Collienne2019}.
+We have computationally shown that the algorithm $\findpath$ finds the correct distance for trees up to seven taxa \autocite{Collienne2019} and know of no larger counterexample.
 
@@
 
-The example in Figure~\ref{fig:splitthm_counterexample} shows that this form of the Split Property does not hold.
+The example in Figure~\ref{fig:splitthm_counterexample} shows that, in this form, the Split Property is not present.
 
@@
 \label{conjecture:weak_split_theorem}
-The Weak Split Property holds for $\rnni$.
+$\rnni$ has the Weak Split Property.
 \end{conjecture}
@@
 
-These considerations motivate us to conjecture the following strong version of the Cluster Property.
+These considerations motivate us to conjecture the following Cluster Property.
 
@@
 \label{conjecture:cluster_theorem}
-The Cluster Property holds for $\rnni$.
+$\rnni$ has the Cluster Property.
 \end{conjecture}
 
-We have computationally checked that this conjecture is true in $\rnni$ spaces with up to seven taxa ($n = 2, \ldots, 7$).
+We have found no counterexamples to this conjecture and have checked that it holds in $\rnni$ spaces with up to seven taxa ($n = 2, \ldots, 7$).
 Specifically, we computed subgraphs of $\rnni$ containing only trees with shared clusters and compared distances in this subgraph with distances in the whole $\rnni$ graph.
@@
 This led us to the Cluster Property, a statement that does not hold in $\nni$.
-All our algorithms and methods developed in this paper witness in favour of the Cluster Property being true in $\rnni$.
+All our algorithms and methods developed in this paper witness in favour of $\rnni$ having the Cluster Property.
 For example, we have checked the statement computationally for up to seven taxa \autocite{Collienne2019}.
-Although the Split Property is true in $\spr$, it is still $\np$-hard to compute distances in that graph.
-Hence Cluster Property alone cannot be used as an argument in favour of the distance problem having an efficient solution in $\rnni$.
+Although the Split Property is present in $\spr$, it is still $\np$-hard to compute distances in that graph.
+Hence the Cluster Property alone cannot be used as an argument in favour of the distance problem having an efficient solution in $\rnni$.
 But the greedy nature of all our algorithms for computing shortest paths in $\rnni$ developed in this paper does provide such an argument.

### 2020-01-29 rnni_convexity.tex
add partition lattice

--- 2020-01-26 rnni_convexity.tex
@@
 
+\subsection{Partition lattice}
+
+The \emph{partition lattice} on $\{1, \ldots, n\}$ is the lattice given by the partially ordered set $(\Pi_n,\leq)$, where $\Pi_n$ is the power set of $\{1, \ldots, n\}$ and $x \leq y$ if partition $x$ refines $y$.
+For simplification we will denote the partition lattice by $\Pi_n$.
+We assume that the partition lattice is ranked such that the partition into singletons has rank zero and the set $\{1, \ldots, n\}$ has rank $n-1$.
+This algebraic structure is related to the $\rnni$ graph in the following way.
+
+\begin{theorem}
+The $\rnni$ graph on $n$ taxa is isomorphic to the graph of maximal chains of the partition lattice $\Pi_n$ where two maximal chains are connected by an edge if and only if they differ by exactly one partition.
+The corresponding metric spaces are isometric.
+\end{theorem}
+
+\begin{proof}
+There is a one to one relation between ranked trees and maximum chains in a partition lattice.
+We can define a bijective mapping from the set of ranked trees to the set of maximum chains in $\Pi_n$ as follows.
+A tree $T$ maps onto a maximum chain $\mathcal{C}_T$ if the set in the partition of rank $i$ in $\mathcal{C}_T$ that is the union of two sets of the partition of rank $i-1$ in $\mathcal{C}_T$ is the cluster induced by the internal node of rank $i$ in $T$.
+
+An $\rnni$ move on a tree $T$ corresponds to changing exactly one partition on $\mathcal{C}_T$ as only one cluster changes by an $\rnni$ move on $T$.
+From this we can follow that the $\rnni$ graph and the metric space on maximum chains of the partition lattice with this operation are isometric.
+\end{proof}
+
 \section{Discussion}

### 2020-01-29 rnni_convexity.tex
add intro partition lattice + refer to it earlier

@@
 We will discuss this further in Section~\ref{section:cluster_theorem}, where we disprove a conjecture of \textcite{Gavryushkin2018-ol} and suggest weaker versions as plausible alternatives.
+In Section~\ref{section:partition_lattice} we establish a connection between the $\rnni$ graph and a classical algebraic structure, the partition lattice.
 We will finish the paper with a discussion and open problems.
@@
 We will also consider the so-called Cluster Property as another weak version of the Split Property (see Conjecture~\ref{conjecture:cluster_theorem}).
+In the final part of this geometry section we will discuss the relation between the $\rnni$ graph and the so-called partition lattice, a well known algebraic structure.
 
@@
 \subsection{Partition lattice}
+\label{section:partition_lattice}
+
+In this section we establish a connection between the $\rnni$ graph and a well known algebraic structure, the partition lattice.
+The relation between these two provides a new direction for further research and leads to a different view on problems of either structure.
 

### 2020-01-29 rnni_convexity.tex
add figure partition lattice

@@
 The \emph{partition lattice} on $\{1, \ldots, n\}$ is the lattice given by the partially ordered set $(\Pi_n,\leq)$, where $\Pi_n$ is the power set of $\{1, \ldots, n\}$ and $x \leq y$ if partition $x$ refines $y$.
-For simplification we will denote the partition lattice by $\Pi_n$.
+For simplification we will denote the partition lattice by $\Pi_n$, $\Pi_4$ is illustrated in Figure~\ref{fig:partition_lattice4}.
 We assume that the partition lattice is ranked such that the partition into singletons has rank zero and the set $\{1, \ldots, n\}$ has rank $n-1$.
@@
 \end{proof}
+
+\begin{figure}[H]
+\centering
+\includegraphics[width=\textwidth]{partition_lattice4}
+\vspace{12pt}
+\caption{The partition lattice $\Pi_4$ on $\{1,2,3,4\}$.
+The highlighted edges correspond to an $\rnni$ path from the tree represented by the leftmost chain to the rightmost one.}
+\label{fig:partition_lattice4}
+\end{figure}
 

### 2020-02-04 rnni_convexity.tex
from David: add note that c-dist is not just inversion dist (kendall tau)
Alex's suggestions

@@
 Indeed, no $\rnni$ move on caterpillar trees can reduce the number of taxa inversions by more than one, assuming that inversions with both taxa of a cherry are counted only once.
-That is, if both pairs of taxa $(x, z)$ and $(y, z)$ appear in opposite orders in $T$ and $R$ and $x$ an $y$ build a cherry then they are counted as one inversion.
+That is, if both pairs of taxa $(x, z)$ and $(y, z)$ appear in opposite orders in $T$ and $R$ and $x$ an $y$ form a cherry then they are counted as one inversion.
+Notice that because of this the distance between caterpillar trees does not coincide with the Kendall tau distance \autocite{Kendall1948-tx}, counting the number of pairs of elements that appear in different order in two permutations, even though the idea is very similar.
 Hence any caterpillar path, every move along which reduced the number of inversions, is a shortest caterpillar path.

### 2020-02-09 rnni_convexity.tex
add David's suggestions

@@
 %\usepackage[notref,notcite]{showkeys}
-\usepackage[style=authoryear,ibidtracker=false,uniquename=false,giveninits=true,terseinits=true,maxbibnames=5,backend=biber]{biblatex}
+\usepackage[style=authoryear,ibidtracker=false,uniquename=false,giveninits=true,terseinits=true,maxbibnames=5,backend=biber,date=year]{biblatex}
 \renewbibmacro{in:}{}
@@
 In this paper we study the graph of ranked phylogenetic trees where the adjacency relation is given by a local rearrangement of the tree structure.
-Our work is motivated by tree inference algorithms, such as maximum likelihood and Markov Chain Monte Carlo methods, where the geometry of the search space plays a central role for efficiency and practicality of optimisation.
+Our work is motivated by tree inference algorithms, such as maximum likelihood and Markov Chain Monte Carlo methods, where the geometry of the search space plays a central role for efficiency and practicality of optimisation and sampling.
 We hence focus on understanding the geometry of the space (graph) of ranked trees, the so-called ranked nearest neighbour interchange ($\rnni$) graph.
@@
 Recent advances in computational phylogenetics have introduced various classes of molecular clock models \autocite{Yoder2000-ks,Drummond2006-nl,Drummond2010-yf} and made computational inference of phylogenetic time-trees possible \autocite{Ronquist2003-eq, Bouckaert2018-yr, Hadfield2018-xp}.
-However, the mathematical challenges resulting from this seemingly minor change in parametrisation (genomic distance VS time distance) of trees have only recently been brought to attention \autocite{Gavryushkin2016-uu}.
+However, the mathematical challenges resulting from this seemingly minor change in parametrisation (genomic distance versus time distance) of trees have only recently been brought to attention \autocite{Gavryushkin2016-uu}.
 These differences motivated \textcite{Gavryushkin2018-ol} to propose an extension of the $\nni$ graph to the class of discrete time-trees.
@@
 Two rooted trees are considered adjacent if they differ by a single neighbour interchange or order change respecting that ranking (detailed description below).
-Considered with the graph-distance $\rnni$ becomes a metric space and inherits the geometric and algorithmic challenges that the $\nni$ space has been traditionally facing.
+As a metric space, the $\rnni$ graph inherits the geometric and algorithmic challenges that the $\nni$ space has been traditionally facing.
 Surprisingly, most of them cannot be settled by directly translating results or applying techniques developed for $\nni$ \autocite{Gavryushkin2018-ol}.
@@
 An example of such a tree is depicted in Figure~\ref{fig:ranked_tree}.
-In the terminology of \autocite{Gavryushkin2018-ol}, the space considered in this paper is the space of ranked ultrametric phylogenetic trees ($\rnniu$ in their notations).
+In the terminology of \autocite{Gavryushkin2018-ol}, the space considered in this paper is the space of ranked ultrametric phylogenetic trees ($\rnniu$ in their notation).
 We postpone formal definitions until later in this section.
@@
 
-We investigate the geometry and algorithmic complexity of the $\rnni$ space.
+We investigate the geometry and algorithmic complexity of the $\rnni$ metric space.
 Specifically, in this paper we establish the exact radius and diameter of the space (Section~\ref{section:diameter}).
-We also show that the subset of caterpillar trees is convex in $\rnni$ (Section~\ref{section:caterpillar_convex}), thus settling one of the open problems proposed in \autocite{Gavryushkin2018-ol}.
-For establishing these geometric properties we use algorithms that will be introduced in Section~\ref{section:algorithms}.
+We also show that the subset of caterpillar trees is convex in the $\rnni$ space (Section~\ref{section:caterpillar_convex}), thus settling one of the open problems proposed in \autocite{Gavryushkin2018-ol}.
+To establish these geometric properties we use algorithms that will be introduced in Section~\ref{section:algorithms}.
 We will in particular provide an approximate algorithm that computes exact distances for small trees.
@@
 
-In the rest of this chapter we formally introduce the terminology used in this paper.
+In the remainder of this section we formally introduce the terminology used in this paper.
 
@@
 This definition also implies that the rank of every internal node is distinct and every $i$ with $1 \leq i \leq n-1$ equals the rank of some node.
-Ranked trees $(T_1, \rank_1)$ and $(T_2, \rank_2)$ are different if trees $T_1$ and $T_2$ are different or $T_1 = T_2$ and $\rank_1 \neq \rank_2$.
+Ranked trees $(T_1, \rank_1)$ and $(T_2, \rank_2)$ are considered equal if $T_1 = T_2$ and $\rank_1 = \rank_2$.
 Since all trees in this paper are ranked trees, we will abuse the notation and drop the rank function from the notation.
@@
 We introduce two types of operation ($\rnni$ \emph{moves}) on trees (see Figure~\ref{fig:RNNI}) and say that two trees are adjacent in the $\rnni$ graph if one can be obtained from the other by an operation of either type.
-The first type of operation is called a \emph{rank move} and defined by swapping the ranks of two internal nodes that are not adjacent in the tree.
+The first type of operation is called a \emph{rank move} and defined by swapping the ranks of two internal nodes that are not adjacent in the tree but are consecutive in the ranking.
 Formally, if $u$ and $v$ are nodes of a tree $T$ such that $|\rank_T(u) - \rank_T(v)| = 1$ and $(u, v)$ is not an edge in $T$ then the tree $R$ obtained from $T$ by only changing $\rank_R(u) = \rank_T(v)$ and $\rank_R(v) = \rank_T(u)$ is said to be obtained by a rank move.
@@
 $\mdtree$ attempts to maximise the dissimilarity to find a remote tree from an arbitrary $\rnni$ tree.
-While these algorithms are interesting on their own (e.g.\ in simulation studies), they will also form an important ingredient to study the geometry of the $\rnni$ space later in this paper.
+While these algorithms are interesting on their own (e.g.\ in simulation studies), they will also form an important ingredient of our study of the geometry of the $\rnni$ space later in this paper.
 For example, $\mdtree$ will be an important tool for finding the radius of the $\rnni$ graph, as the algorithm computes a tree with maximum distance if the input tree is a caterpillar tree.
@@
 
-The algorithm $\findpath$ can be used to approximate the $\rnni$ distance.
-We will study the accuracy of this approximation by computing the exact $\rnni$ distance for small trees.
+The algorithm $\findpath$ provides an upper bound the $\rnni$ distance.
+We will study the accuracy of this approximation by also computing the exact $\rnni$ distance for small trees.
 Specifically, we will use the algorithm for computing the complete $\rnni$ graph that is given in \autocite[Section 3.3]{Gavryushkin2018-ol}.
@@
 
-Note that computing the $\rnni$ space on trees with seven taxa is a challenging goal to achieve.
+Note that computing the $\rnni$ space on trees with seven taxa is a challenging goal.
 For example, \textcite{Whidden2016-kl} were able to compute the much smaller $\spr$ space on up to seven taxa and used this to compute curvature values for all pairs of trees within that space.
@@
 
-In this section we present $\findpath$, an efficient algorithm for computing paths, and therefore approximating distances, between trees in $\rnni$.
+In this section we present $\findpath$, an efficient heuristic for computing short paths between trees in $\rnni$.
 
@@
 
-In this section we introduce an algorithm to compute paths between \emph{caterpillar trees}, which are trees where each internal node is adjacent to at least one leaf.
-As a result caterpillar trees have only one \emph{cherry}, which is a pair of taxa that share their parent.
+In this section we introduce an algorithm to compute $\rnni$ paths between \emph{caterpillar trees}, which are trees where each internal node is adjacent to at least one leaf.
+Caterpillar trees have only one \emph{cherry}, which is a pair of taxa that share their parent.
 A path between two caterpillar trees that only consists of caterpillar trees is called a \emph{caterpillar path}.
@@
 We computed these graphs according to the algorithm from \autocite{Gavryushkin2018-ol} as discussed in Section~\ref{section:algorithms}.
-The source code of all these implementations can be openly accessed at \autocite{Collienne2019}.
+The source code of all these implementations is available at \autocite{Collienne2019}.
 

### 2020-02-10 rnni_convexity.tex
quick look at partition lattices

@@
 We will also consider the so-called Cluster Property as another weak version of the Split Property (see Conjecture~\ref{conjecture:cluster_theorem}).
-In the final part of this geometry section we will discuss the relation between the $\rnni$ graph and the so-called partition lattice, a well known algebraic structure.
+In the final part of this geometry section we will discuss the relation between the $\rnni$ graph and the so-called partition lattice, a well-known algebraic structure.
 
@@
 
-In this section we establish a connection between the $\rnni$ graph and a well known algebraic structure, the partition lattice.
-The relation between these two provides a new direction for further research and leads to a different view on problems of either structure.
+In this section we establish a connection between the $\rnni$ graph and a well-known algebraic structure, the partition lattice.
+This connection provides a new direction for further research and translates results and open problems from the language of phylogenetics to the language of lattice theory.
 
-The \emph{partition lattice} on $\{1, \ldots, n\}$ is the lattice given by the partially ordered set $(\Pi_n,\leq)$, where $\Pi_n$ is the power set of $\{1, \ldots, n\}$ and $x \leq y$ if partition $x$ refines $y$.
-For simplification we will denote the partition lattice by $\Pi_n$, $\Pi_4$ is illustrated in Figure~\ref{fig:partition_lattice4}.
+The \emph{partition lattice} on $\{1, \ldots, n\}$ is the lattice given by the partially ordered set $(\Pi_n, \leq)$, where $\Pi_n$ is the power set of $\{1, \ldots, n\}$ and $x \leq y$ if partition $x$ refines $y$, which means...
+For simplification we will denote the partition lattice on $n$ elements by $\Pi_n$.
+$\Pi_4$ is illustrated in Figure~\ref{fig:partition_lattice4}.
 We assume that the partition lattice is ranked such that the partition into singletons has rank zero and the set $\{1, \ldots, n\}$ has rank $n-1$.

### 2020-02-10 rnni_convexity.tex
David's comment re Prop 1; the 'refines' definition; Figure 4 explanation

@@
 \FOR {$k = 1, \dots, n-2$}
+\label{alg:findpath:line:for_loop}
     \WHILE {$\rank_{T'}(\mrca_{T'}(C_k))>k$}
@@
 \proof
-It suffices to show that the update operation (line~\ref{alg:findpath:line:move_set_down}) is well-defined, that is the $\rnni$ move that decreases the rank of $\mrca_{T'}(C_k)$ is unique.
+It is not hard to see that the algorithm terminates with a path ending in $R$.
+Indeed, every iteration of the \textbf{for} loop (line~\ref{alg:findpath:line:for_loop}) is completed once $\rank_{T'}(\mrca_{T'}(C_k)) = k$, that is, the first $k$ clusters of $T'$ and $R$ coincide.
+Hence after $n-2$ such iteration the path has to arrive at $R$.
+
+To complete the proof, it suffices to show that the update operation (line~\ref{alg:findpath:line:move_set_down}) is well-defined, that is the $\rnni$ move that decreases the rank of $\mrca_{T'}(C_k)$ is unique.
 
@@
 
-The \emph{partition lattice} on $\{1, \ldots, n\}$ is the lattice given by the partially ordered set $(\Pi_n, \leq)$, where $\Pi_n$ is the power set of $\{1, \ldots, n\}$ and $x \leq y$ if partition $x$ refines $y$, which means...
+The \emph{partition lattice} on $\{1, \ldots, n\}$ is the lattice given by the partially ordered set $(\Pi_n, \leq)$, where $\Pi_n$ is the power set of $\{1, \ldots, n\}$ and $X \leq Y$ if partition $X$ refines $Y$, that is, $X \leq Y \Leftrightarrow (\forall x \in X)(\exists y \in Y) x \subseteq y$.
 For simplification we will denote the partition lattice on $n$ elements by $\Pi_n$.
 $\Pi_4$ is illustrated in Figure~\ref{fig:partition_lattice4}.
-We assume that the partition lattice is ranked such that the partition into singletons has rank zero and the set $\{1, \ldots, n\}$ has rank $n-1$.
-This algebraic structure is related to the $\rnni$ graph in the following way.
+We assume that a partition $X$ in the partition lattice $\Pi_n$ has rank $k$ if the number of elements in $X$ is $n-k$.
+The algebraic structure of $\Pi_n$ is related to the $\rnni$ graph on $n$-taxa trees in the following way.
 
@@
 The corresponding metric spaces are isometric.
+\label{thm:partition_lattice}
 \end{theorem}
@@
 
-An $\rnni$ move on a tree $T$ corresponds to changing exactly one partition on $\mathcal{C}_T$ as only one cluster changes by an $\rnni$ move on $T$.
-From this we can follow that the $\rnni$ graph and the metric space on maximum chains of the partition lattice with this operation are isometric.
+Note that this bijection is an isomorphism between the $\rnni$ graph and the graph of chains as in the theorem.
+Indeed, two chains are different by exactly one partition if and only if the corresponding trees are connected by an $\rnni$ move.
 \end{proof}
@@
 \end{figure}
+
+Figure~\ref{fig:partition_lattice4} is an illustration of the proof of Theorem~\ref{thm:partition_lattice}.
+The four chains indicated in bold correspond to the following $\rnni$ path.
+The leftmost chain corresponds to the caterpillar tree $[1, 2, 3, 4]$.
+First, the partition $\{1, 2, 3\} \{4\}$ is replaced with $\{1, 2\} \{3, 4\}$ and we get the chain corresponding to the tree $[\{1, 2\}, \{3, 4\}]$ (in the cluster representation), which is one $\rnni$ move away from the caterpillar tree.
+Second, the partition $\{1, 2\} \{3\} \{4\}$ is replaced with $\{1\} \{2\} \{3, 4\}$, which corresponds to the rank swap on the previous tree.
+Third, the partition $\{1, 2\} \{3, 4\}$ is replaced with $\{1\} \{2, 3, 4\}$ and we reach the caterpillar tree $[3, 4, 2, 1]$.
 

### 2020-07-03 rnni_convexity.tex
typo in FP pseudocode
add comments of SIDMA referee

@@
 \newcommand{\mdtree}{\textsc{MDTree}}
+\newcommand{\sidma}[1]{\todo[color=green]{#1}}
 
@@
 
+\sidma{Since this is the crux of the paper, expand, add images, and work through an example of the algorithm}
 Let $T$ and $R$ be arbitrary trees and $[C_1, \ldots, C_{n-2}]$ be the cluster representation of $R$ (the ``destination'' tree).
@@
 We continue extending $p$ in this way by repeatedly performing $\rnni$ moves until the rank of $\mrca_{T'}(C_k) = k$, thus completing step $k$.
+\sidma{Reword [line 5 of algorithm] and/or expand discussion in text to demonstrate what this is doing}
 
@@
         \IF {node $u$ with $\rank_{T'}(u) = rank_{T'}(\mrca_{T'}(C_k))-1$ is adjacent to $\mrca_{T'}(C_k)$}
-            \STATE $T'$ is $T'$ with $\rank_{T'}(\mrca_{T'}(C_k))$ decreased by an $\nni$ move
+            \STATE $T''$ is $T'$ with $\rank_{T'}(\mrca_{T'}(C_k))$ decreased by an $\nni$ move
         \ELSE
@@
 \proof
+\sidma{Remove the sentence "It is not hard to see that the algorithm..." or move to before the proof in the informal discussion of what the algorithm is doing.
+In the formal discussion (i.e.
+the proof), state why the algorithm terminates (i.e.
+expand on the rest of that first paragraph).}
 It is not hard to see that the algorithm terminates with a path ending in $R$.
@@
 
+\sidma{Include an argument about why line 5 makes sense in the algorithm, in the explanation of it before the proof, and here, in the proof.}
 To complete the proof, it suffices to show that the update operation (line~\ref{alg:findpath:line:move_set_down}) is well-defined, that is the $\rnni$ move that decreases the rank of $\mrca_{T'}(C_k)$ is unique.
 
+\sidma{Remove "it is not hard to see" and replace with images or written explanation. }
 Case $k = 1$.
@@
 
+\sidma{Fill in the missing words before starting the subcases.}
+\sidma{Define the term "supress" in terms of ranked trees.}
 Case $k > 1$.
@@
 Proceed as in Case $k = 1$.
+\sidma{The end of the proof is missing. Complete the proof. }
 \end{enumerate}
@@
 
+\sidma{While is highly likely that the worst-case complexity is quadratic, you need to demonstrate why that is the case.
+The simplest, and most convincing way would be to write this as a proposition and then go line-by-line detailing the running time.}
 Clearly, the worst-case complexity of $\findpath$ is quadratic in the number of taxa.
@@
 
+\sidma{Since these algorithms are central to your paper, write out the proofs of their running time and correctness.}
 The running time of $\csort$ is quadratic in $n$.
@@
 The path $p$ returned by $\csort$ is a shortest caterpillar path because every tree modification on $p$ reduces the number of inversions of taxa in $T$ and $R$, and no modification can reduce more than one insertion.
+\sidma{Define "number of inversions"}
 Indeed, no $\rnni$ move on caterpillar trees can reduce the number of taxa inversions by more than one, assuming that inversions with both taxa of a cherry are counted only once.
+\sidma{Define "appear in opposite orders"}
 That is, if both pairs of taxa $(x, z)$ and $(y, z)$ appear in opposite orders in $T$ and $R$ and $x$ an $y$ form a cherry then they are counted as one inversion.
 Notice that because of this the distance between caterpillar trees does not coincide with the Kendall tau distance \autocite{Kendall1948-tx}, counting the number of pairs of elements that appear in different order in two permutations, even though the idea is very similar.
+\sidma{Not sure why the statement starting with "Hence" about the shortest caterpillar path follows from the previous statements about Kendal tau distances. Explain the connection or reword}
 Hence any caterpillar path, every move along which reduced the number of inversions, is a shortest caterpillar path.
@@
 As we will see in Section~\ref{section:diameter}, $\mdtree$ achieves this goal for caterpillar trees.
+\sidma{ Explain or prove why the MDTree only produces caterpillars. Add images and work through the algorithm.}
 
@@
 Next, we will consider the so-called Split Property \autocite{Gavryushkin2018-ol}, that is, every shared split of taxa is maintained along every shortest path.
+\sidma{State whether you are including the root in your set of taxa.
+If not, then it would be good to explain why different rooted trees have the same set of splits (e.g., the rooted trees (1,(2,3)) and ((1,2),3) have the same split set) so that the later discussion on clusters is clearer.}
 Recall that a \emph{split} is a bipartition of the set of taxa obtained by deleting an edge of the tree $T$.
@@
 \end{lemma}
+\sidma{Include example and image for this lemma}
 
 \begin{proof}
+\sidma{This is an good motivation of the lemma and would be better as a lead-in to the proof, and then fill in a rigorous proof in its place.}
 Since the diameter is bounded from above by the maximum length of a path computed by $\findpath$, it is enough to find this maximum.
@@
 \begin{proof}
+\sidma{Second sentence runs on and does not make sense. What is a step? If it is a single move on the caterpillar tree, then moves one per step? Rewrite to make this clear.}
 First note that $d(T,R) = \frac{(n-1)(n-2)}{2}$ implies $d_c(T,R) \geq \frac{(n-1)(n-2)}{2}$.
@@
 \begin{proof}
-There is a one to one relation between ranked trees and maximum chains in a partition lattice.
+\sidma{Most of the proof is missing. Add in rest of proof.}
+There is a one-to-one relation between ranked trees and maximum chains in a partition lattice.
 We can define a bijective mapping from the set of ranked trees to the set of maximum chains in $\Pi_n$ as follows.

### 2020-07-03 rnni_convexity.tex
mark what has been moved

@@
 \newcommand{\sidma}[1]{\todo[color=green]{#1}}
+\newcommand{\moved}[1]{\todo[color=red]{start moved bock}{#1}\todo[color=red]{end moved block}}
 
@@
 
+\moved{
 \sidma{Since this is the crux of the paper, expand, add images, and work through an example of the algorithm}
@@
 The simplest, and most convincing way would be to write this as a proposition and then go line-by-line detailing the running time.}
-Clearly, the worst-case complexity of $\findpath$ is quadratic in the number of taxa.
+Clearly, the worst-case complexity of $\findpath$ is quadratic in the number of taxa.}
 
